### 2021-01-25 covid_nrs_place_of_death.Rmd
Rerunning analysis of open data on COVID-19 related deaths - started importing data & a basic plot

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)
library(opendatascot)
library(phsmethods)
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)
library(janitor)
library(ISOweek)  # for computing date from ISO week number + year
```

# Set up SPARQL query

Note: not needed if using `opendatascot`
To access the latest NRS data using their API, we first setup a SPARQL query.

```{r}
endpoint <- "http://statistics.gov.scot/sparql.json"
```

# Import data

```{r}
# checking out structure
ods_structure("deaths-involving-coronavirus-covid-19")
ods_schemes("deaths-involving-coronavirus-covid-19")
ods_print_query("deaths-involving-coronavirus-covid-19")

data_raw <- ods_dataset("deaths-involving-coronavirus-covid-19",geography = "sc")

# TODO: remove below bit, it's for comparison with previous results!
cv_data_raw <- read_csv("https://statistics.gov.scot/downloads/cube-table?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19")
cv_data_main <- cv_data_raw %>% 
  clean_names() %>% 
  filter(feature_code == "S92000003") %>% 
  mutate(date_code = str_replace(date_code, "w/c ", ""),
         date_code = ymd(date_code)) %>% 
  filter(!is.na(date_code)) %>% 
  select(date_code, sex, age, cause_of_death, location_of_death, measurement, value, 
         -feature_code, -units) %>% 
  arrange(date_code, sex, age, cause_of_death, location_of_death, measurement)
```

## Understanding how dates are coded

`refPeriod` refers to the week number, as defined by [ISO 8601](https://en.wikipedia.org/wiki/ISO_week_date), and can be computed from a date with `lubridate::isoweek()`; the first week of 2020 is the week starting with Monday 2019-12-30, ~~and the corresponding start dates for each week can be computed by adding weeks using `lubridate::dweeks()` to that date.~~ This is non-trivial, because the number of iso weeks in a year isn't constant. The `ISOweek::ISOweek2date()` function deals with this issue in a crude manner, and allows us to compute the date (w/c on date) from the year & week number. It relies on being provided a valid week number (some years have 52 and some have 53), but it's probably safe to assume that the NRS data has valid dates.

Invalid week numbers will produce a date in the next year: e.g. week 53 in 2021 would actually start in 2022, which makes it invalid. We can test for that by checking that the year provided is the same year in the date produced.

```{r}
date_first_week_2020 <- ymd("2019-12-30")
ISOweek::ISOweek2date(weekdate = "2019-W53-1")
ISOweek::ISOweek2date(weekdate = "2020-W01-1")
ISOweek::ISOweek2date(weekdate = "2020-W52-1")
ISOweek::ISOweek2date(weekdate = "2020-W53-1")
ISOweek::ISOweek2date(weekdate = "2021-W01-1")
ISOweek::ISOweek2date(weekdate = "2021-W53-1")

compute_start_date_from_week_number <- function(week_number, year_number) {
  ## this assumes a valid week number! There are 52 or 53 weeks in a year
  isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
  computed_dates <- ISOweek::ISOweek2date(isoweek_string)
  ## invalid weeks will result in a computed date in a year after the requested year...
  ## a valid week can result in a date that starts in the year before!
  if (any(year_number<year(ymd(computed_dates)))) stop("invalid week number provided!")
  return(computed_dates)
}

tibble(
    week_number = c(1,2,3,52,53)
  ) %>%
  mutate(
    date = compute_start_date_from_week_number(week_number = week_number, year_number = 2020),
    iso_week_from_date = isoweek(date)
  ) %>% knitr::kable(caption = "Examples of computing start date for each week, and computing isoweek from the dates!")
```

## Recoding dates

```{r}
data_main <- 
  data_raw %>%
  clean_names() %>%
  rename(cause_of_death = causeofdeath, location_of_death = locationofdeath) %>%  # neater names
  filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
  mutate(
    ref_year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = ref_year),
    value = as.numeric(value)  # convert to numeric!
  ) %>%
  select(-c(ref_area, measure_type))  # ref_area is Scotland for all records; measure_type is count

## check date conversion
data_main %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
```

# Visualisations

## Number of deaths by age group

```{r, fig.width=12, fig.height=9}
## graph format copied from: https://github.com/davidhen/nrs_covid_stats/

date_range_text <- glue::glue("Data from w/c {format(min(data_main$date_w_c), '%d %B %Y')} to w/c {format(max(data_main$date_w_c), '%d %B %Y')} (inclusive).")

data_main %>% 
  filter(sex == "all" & location_of_death == "all" & cause_of_death == "covid-19-related") %>% 
  filter(age != "all") %>% 
  group_by(age) %>% 
  summarise(n = sum(value)) %>%
  mutate(pct = round(n/sum(n), 3)) %>% 
  ggplot(aes(age, n)) +
  geom_col(fill = "#4477AA") +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80)) +
  theme_minimal() +
  labs(x = "",
       y = "",
       title = "Deaths Associated with COVID-19 by Age Group",
       subtitle = date_range_text,
       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate.")
```

##

### 2021-01-25 covid_nrs_place_of_death.Rmd
added past deaths so we can compute averages, min & max by week number. started explaining how some years have 52 and some 53 weeks

@@
 ```{r}
+# ods_all_datasets() %>% View()
+# ods_structure("deaths")
+
 # checking out structure
@@
   arrange(date_code, sex, age, cause_of_death, location_of_death, measurement)
+
+## load past deaths for calculating averages
+temp_file <- tempfile()
+past_deaths_source <- "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv"
+past_deaths_file <- curl_download(url = past_deaths_source, destfile = temp_file, quiet = FALSE)
+
+past_deaths <- 
+  read_csv(past_deaths_file, skip = 2) %>%
+  rename(year_of_death = 1) %>%  # rename first column
+  mutate(
+    place_of_death = if_else(is.na(`1`), true = year_of_death, NA_character_)
+  ) %>%
+  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
+  drop_na(`1`) %>%
+  pivot_longer(cols = all_of(as.character(1:53)), values_to = "value", names_to = "week_number") %>%
+  mutate(week_number=as.integer(week_number))
+
+past_deaths %>% filter(week_number %in% 52:53) %>% count(year_of_death, week_number)
+
+past_deaths_summarised <-
+  past_deaths %>%
+  filter(year_of_death!="average") %>%  # remove averages
+  group_by(year_of_death, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
+  filter(!all(value==0 & week_number==53)) %>%
+  ungroup %>%
+  group_by(week_number,place_of_death) %>%
+  summarise(across(value, .fns = list(mean=mean,min=min,max=max)),  # having removed 0s for week 53 where it didn't exist, we average over years for week 53 that had that week only
+    .groups = "drop"
+  ) %>% rename_all(~str_replace_all(.x, pattern = "value", replacement = "deaths"))
+
+# TODO: show that only 2015 had week 53 and deal with this somehow - combine next years week 1 and previous week 52 instead?
+week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks."
 ```
@@
 
+## Recoding place of death
+
+```{r}
+## recoding place of death
+data_main %>% count(location_of_death)
+past_deaths %>% count(place_of_death)
+
+# datapasta::tribble_paste(input_table = tibble(location_of_death = sort(unique(data_main$location_of_death)), place_of_death = sort(unique(past_deaths$place_of_death))))
+tbl_recoded_place_of_death <- tibble::tribble(
+  ~location_of_death,           ~place_of_death,   ~place_of_death_recoded,
+  "all",                        "All locations",   "All",
+  "care-home",                      "Care Home",   "Care home",
+  "-non-institution", "Home and other non-institution",   "Home",
+  "hospital",                       "Hospital",   "Hospital",
+  "other-institution",    "Other institution3",   "Other"
+     )
+```
+
+## Merged data
+
+```{r}
+## check whether there's an issue with week 52/53
+#
+
+merged_data <-
+  data_main %>%
+  left_join(tbl_recoded_place_of_death) %>%
+  left_join(past_deaths %>% mutate(sex = "all", age = "all", cause_of_death = "all-causes"),  # only merge for aggregated data b/c that's all we have!
+            by = c("place_of_death", "week_number", "sex", "age", "cause_of_death")) %>% 
+  select(-c(location_of_death,place_of_death))
+
+```
+
 # Visualisations
@@
 
-##
+## Deaths over time
+
+```{r, fig.width=12, fig.height=9}
+
+data_main %>% 
+  # filter(sex == "all" & location_of_death == "all" & cause_of_death == "covid-19-related") %>% 
+  # filter(age != "all") %>% 
+  group_by(age) %>% 
+  summarise(n = sum(value)) %>%
+  mutate(pct = round(n/sum(n), 3)) %>% 
+  ggplot(aes(age, n)) +
+  geom_col(fill = "#4477AA") +
+  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80)) +
+  theme_minimal() +
+  labs(x = "",
+       y = "",
+       title = "Deaths Associated with COVID-19 by Age Group",
+       subtitle = date_range_text,
+       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate.")
+```

### 2021-01-26 covid_nrs_place_of_death.Rmd
Re-doing figure 6 from previous analysis, and added some more data wrangling

@@
 
+We're importing the following datasets:
+
+* deaths involving COVID-19 from Scottish gov't open data service
+* weekly deaths grouped by location 2015-2019
+
+Although COVID-19 related deaths included a 5-year weekly average number of deaths, they don't include the minima or maxima for each week - this is why we also import past weekly deaths.
+
 ```{r}
@@
 
-past_deaths %>% filter(week_number %in% 52:53) %>% count(year_of_death, week_number)
+# past_deaths %>% filter(week_number %in% 52:53) %>% count(year_of_death, week_number)
 
@@
   filter(!all(value==0 & week_number==53)) %>%
+  (function (data_tbl) {
+    bind_rows(
+      data_tbl,
+      data_tbl %>% filter(year_of_death %in% 2016:2019 & week_number==52) %>% mutate(week_number=53)  # copy records for week 52 in years 2016-2019 and count them as if they were week 53
+    )
+  }) %>%
   ungroup %>%
@@
 
-# TODO: show that only 2015 had week 53 and deal with this somehow - combine next years week 1 and previous week 52 instead?
-week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks."
+week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
 ```
@@
   ~location_of_death,           ~place_of_death,   ~place_of_death_recoded,
-  "all",                        "All locations",   "All",
+  "hospital",                       "Hospital",   "Hospital",
   "care-home",                      "Care Home",   "Care home",
   "-non-institution", "Home and other non-institution",   "Home",
-  "hospital",                       "Hospital",   "Hospital",
-  "other-institution",    "Other institution3",   "Other"
-     )
+  "other-institution",    "Other institution3",   "Other",
+  "all",                        "All locations",   "All"
+     ) %>% mutate(place_of_death_recoded = factor(place_of_death_recoded, levels = place_of_death_recoded))
 ```
@@
 ```{r}
-## check whether there's an issue with week 52/53
-#
-
 merged_data <-
   data_main %>%
+  filter(cause_of_death!="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
+  pivot_wider(names_prefix = "deaths_", names_from=cause_of_death, values_from=value) %>%
+  clean_names() %>%
+  mutate(deaths_non_covid_related = deaths_all_causes - deaths_covid_19_related) %>%
   left_join(tbl_recoded_place_of_death) %>%
-  left_join(past_deaths %>% mutate(sex = "all", age = "all", cause_of_death = "all-causes"),  # only merge for aggregated data b/c that's all we have!
+  left_join(past_deaths_summarised %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+            by = c("place_of_death", "week_number", "sex", "age")) %>% 
+  select(-c(location_of_death,place_of_death)) %>%
+  mutate(
+    week_number_run_over = if_else(ref_year == 2021, 53 + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  )
+
+```
+
+### Check averages against previously computed data
+
+```{r}
+data_main %>%
+  left_join(tbl_recoded_place_of_death) %>%
+  left_join(past_deaths_summarised %>% mutate(sex = "all", age = "all", cause_of_death = "all-causes"),  # only merge for aggregated data b/c that's all we have!
             by = c("place_of_death", "week_number", "sex", "age", "cause_of_death")) %>%
-  select(-c(location_of_death,place_of_death))
-
+# merged_data %>%
+  filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years" | (place_of_death_recoded=="All" & !is.na(deaths_mean))) %>%
+  select(ref_year,week_number,cause_of_death,value,deaths_mean) %>%
+  mutate(value = if_else(cause_of_death=="all-causes", NA_real_, value)) %>%  # make NA the 2020/2021 value
+  pivot_longer(cols=c(value,deaths_mean), names_to="computation", values_to="average", values_drop_na=TRUE) %>%
+  select(-cause_of_death) %>%
+  pivot_wider(values_from=average,names_from=computation) %>%
+  rename(nrs_average = value, my_average = deaths_mean) %>%
+  filter(round(nrs_average)!=round(my_average)) %>% 
+  knitr::kable(caption = "Subset of weekly deaths where calculate mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file.
+Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019.
+For all the other datasets the mean deaths computed agree!")
 ```
@@
 date_range_text <- glue::glue("Data from w/c {format(min(data_main$date_w_c), '%d %B %Y')} to w/c {format(max(data_main$date_w_c), '%d %B %Y')} (inclusive).")
+
+source_nrs_text <- "Source: National Records of Scotland"
 
@@
        subtitle = date_range_text,
-       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate.")
+       caption = paste0(c(date_range_text,source_nrs_text), collapse="\n")
+  )
 ```
 
-## Deaths over time
+## Figure 6: Recorded deaths in Scotland, by week number & location
 
 ```{r, fig.width=12, fig.height=9}
+vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."
 
-data_main %>% 
-  # filter(sex == "all" & location_of_death == "all" & cause_of_death == "covid-19-related") %>% 
-  # filter(age != "all") %>% 
-  group_by(age) %>% 
-  summarise(n = sum(value)) %>%
-  mutate(pct = round(n/sum(n), 3)) %>% 
-  ggplot(aes(age, n)) +
-  geom_col(fill = "#4477AA") +
-  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80)) +
+removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data %>% filter(ref_year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data %>% filter(ref_year==2021) %>% .$week_number)} in 2021.")
+
+annotations_figure_6 <-
+# datapasta::tribble_paste(tbl_recoded_place_of_death %>% select(place_of_death_recoded))
+  tibble::tribble(
+    ~place_of_death_recoded, ~x, ~y, ~label,
+                 "Hospital", 35L, 800, "Hospital deaths dipped below historical minimum and remained low until Winter.",
+                 "Hospital", 35L, 250, "Shaded area shows range of deaths 2015-2019 dashed line shows weekly average",
+                "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
+                     "Home", 35L, 500, "The gap between historical\nhigh and both types of death\nat home was high throughout."
+                  )
+
+merged_data %>% 
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death_recoded %in% c("All","Other")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  # geom_line(aes(y = deaths_all_causes)) +
+  # geom_line(aes(y = deaths_non_covid_related), colour = "red") +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_grid(~place_of_death_recoded) +
   theme_minimal() +
-  labs(x = "",
-       y = "",
-       title = "Deaths Associated with COVID-19 by Age Group",
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
        subtitle = date_range_text,
-       caption = "Source: National Records of Scotland\nDeaths where COVID-19 mentioned on the death certificate.")
+       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n")
+       )
 ```

### 2021-01-27 covid_nrs_place_of_death.Rmd
fixed problem with place of death ordering in plot 6

@@
                      "Home", 35L, 500, "The gap between historical\nhigh and both types of death\nat home was high throughout."
-                  )
+                  ) %>% 
+  mutate(place_of_death_recoded = factor(place_of_death_recoded, levels = levels(tbl_recoded_place_of_death$place_of_death_recoded)))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
 
@@
   geom_line(aes(y = deaths, colour = deaths_type)) +
-  # geom_line(aes(y = deaths_all_causes)) +
-  # geom_line(aes(y = deaths_non_covid_related), colour = "red") +
   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +

### 2021-01-27 covid_nrs_place_of_death.Rmd
started saving plots

--- 2021-01-26 covid_nrs_place_of_death.Rmd
@@
                 "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
-                     "Home", 35L, 500, "The gap between historical\nhigh and both types of death\nat home was high throughout."
+                     "Home", 35L, 500, "The gap between historical high and both types of death at home was high throughout."
+                  ) %>% 
+  mutate(
+    place_of_death_recoded = factor(place_of_death_recoded, levels = levels(tbl_recoded_place_of_death$place_of_death_recoded)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
+    label = str_wrap(label, width = 40)
     )
 
+figure_6_location_of_death <-
   merged_data %>%
@@
        subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n")
+       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
        )
+
+map(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_6_location_of_death",.x), plot = figure_6_location_of_death, width = 12, height = 9, dpi = 300)
+)
+
+figure_6_location_of_death
 ```

### 2021-01-29 covid_nrs_place_of_death.Rmd
Added a lot of code to include historical figures broken down by HB, council area, sex, age, however, now I don't have weekly figures for HB and council area breakdowns, so those need included too

@@
 
-library(tidyverse)
-library(opendatascot)
-library(phsmethods)
+library(tidyverse)  # for tidy workflow
+library(opendatascot)  # importing data from ScotGov open data website
+library(phsmethods)  # methods for working with PHS data
 # library(SPARQL)  # taken care of by opendatascot
-library(lubridate)
-library(janitor)
+library(lubridate)  # dealing with dates
+library(janitor)  # for cleaning column names
 library(ISOweek)  # for computing date from ISO week number + year
@@
 
-We're importing the following datasets:
+This analysis is based off two sets of data:
 
-* deaths involving COVID-19 from Scottish gov't open data service
-* weekly deaths grouped by location 2015-2019
+1. Weekly deaths involving COVID-19 from Scottish gov't open data service
+  - this is imported using the `opendatascot` package in R, and connects to the data available at: [the Scottish Gov't open data   website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)
+2. Historical deaths 2015-2019, grouped by week & location
+  - These are available at the NRS website on [deaths involving COVID-19]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland), specifically under the 'Related Statistics' section, under the heading 'Ad-hoc queries'.
+  - We load the total deaths, but also grouped by:
+    - Health board
+    - Local Authority
+    - Sex
+    - Age
+    - ~~Ethnicity~~ Note: not sure this is available
 
-Although COVID-19 related deaths included a 5-year weekly average number of deaths, they don't include the minima or maxima for each week - this is why we also import past weekly deaths.
+Note that the weekly deaths due to COVID-19 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the other datasets.
+The same holds for the age & sex breakdown.
+
+## Load weekly COVID-19 related deaths
 
 ```{r}
+# checking out structure
+
 # ods_all_datasets() %>% View()
-# ods_structure("deaths")
+# ods_structure("deaths")  # can we import all deaths-related data this way?
+# ods_print_query("deaths")
 
-# checking out structure
-ods_structure("deaths-involving-coronavirus-covid-19")
-ods_schemes("deaths-involving-coronavirus-covid-19")
-ods_print_query("deaths-involving-coronavirus-covid-19")
+# ods_structure("deaths-involving-coronavirus-covid-19")
+# ods_schemes("deaths-involving-coronavirus-covid-19")
+# ods_print_query("deaths-involving-coronavirus-covid-19")
+ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003
 
-data_raw <- ods_dataset("deaths-involving-coronavirus-covid-19",geography = "sc")
+# TODO: split this into a scotland-wide dataset, and grouped by health board & council area
+covid_deaths_raw <- bind_rows(
+  ods_dataset("deaths-involving-coronavirus-covid-19",geography = "sc")#,
+  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la"),
+  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "hb")
+)
+```
 
-# TODO: remove below bit, it's for comparison with previous results!
-cv_data_raw <- read_csv("https://statistics.gov.scot/downloads/cube-table?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19")
-cv_data_main <- cv_data_raw %>% 
-  clean_names() %>% 
-  filter(feature_code == "S92000003") %>% 
-  mutate(date_code = str_replace(date_code, "w/c ", ""),
-         date_code = ymd(date_code)) %>% 
-  filter(!is.na(date_code)) %>% 
-  select(date_code, sex, age, cause_of_death, location_of_death, measurement, value, 
-         -feature_code, -units) %>% 
-  arrange(date_code, sex, age, cause_of_death, location_of_death, measurement)
+## Download historical deaths data
 
-## load past deaths for calculating averages
-temp_file <- tempfile()
-past_deaths_source <- "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv"
-past_deaths_file <- curl_download(url = past_deaths_source, destfile = temp_file, quiet = FALSE)
+These are all downloaded from the NRS website to a temporary file created by the `tempfile()` function.
 
-past_deaths <- 
-  read_csv(past_deaths_file, skip = 2) %>%
-  rename(year_of_death = 1) %>%  # rename first column
+```{r}
+## overall file - where did I get the link from? it's not obviously available from the website...
+past_deaths_overall_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
+                destfile = tempfile(),
+                quiet = FALSE)
+
+past_deaths_health_board_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
+                destfile = tempfile(),
+                quiet = FALSE)
+
+past_deaths_council_area_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
+                destfile = tempfile(),
+                quiet = FALSE)
+
+past_deaths_age_sex_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
+                destfile = tempfile(),
+                quiet = FALSE)
+```
+
+### Overall data
+
+```{r}
+past_deaths_overall <- 
+  read_csv(past_deaths_overall_file, skip = 2) %>%
+  rename(year = 1) %>%  # rename first column
   mutate(
-    place_of_death = if_else(is.na(`1`), true = year_of_death, NA_character_)
+    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
   ) %>%
@@
   drop_na(`1`) %>%
-  pivot_longer(cols = all_of(as.character(1:53)), values_to = "value", names_to = "week_number") %>%
+  pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
   mutate(week_number=as.integer(week_number))
 
-# past_deaths %>% filter(week_number %in% 52:53) %>% count(year_of_death, week_number)
+# past_deaths_overall %>% filter(week_number %in% 52:53) %>% count(year, week_number)
+```
 
-past_deaths_summarised <-
-  past_deaths %>%
-  filter(year_of_death!="average") %>%  # remove averages
-  group_by(year_of_death, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
-  filter(!all(value==0 & week_number==53)) %>%
-  (function (data_tbl) {
+### Grouped by health board, council area, and age & sex
+
+The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data.
+
+The total deaths by place of death don't add up between the overall deaths & health board. This may be due to small numbers being redacted?
+
+```{r, warning=FALSE}
+past_deaths_health_board <- 
+  readxl::read_excel(past_deaths_health_board_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  clean_names() %>%
+  rename(
+    week_number = week_of_occurrence,
+    place_of_death = location
+  ) %>%
+  mutate(
+    week_number = as.integer(week_number)
+  ) %>%
+  drop_na(week_number)  # remove empty rows & one row where it's just the copyright notice
+
+past_deaths_council_area <-
+  readxl::read_excel(past_deaths_council_area_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  clean_names() %>%
+  rename(
+    week_number = week,
+    place_of_death = location_of_death
+  ) %>%
+  mutate(
+    week_number = as.integer(week_number)
+  ) %>%
+  drop_na(week_number)  # remove empty rows & one row where it's just the copyright notice
+
+past_deaths_age_sex <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
+  readxl::read_excel(past_deaths_age_sex_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
+  mutate(
+    year = 2000L + as.integer(year),
+    week_number = as.integer(week_number)
+  ) %>%
+  drop_na(year)  # remove empty rows & one row where it's just the copyright notice
+
+## check that the results are the same as overall deaths
+map2_dfr(
+  .x = list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
+  .y = c("overall","health_board","council_area","age_sex"),
+  .f = function(data_tbl, name) {
+    data_tbl %>%
+      mutate(dataset=name,year=as.numeric(year)) %>%
+      filter(!is.na(year) & place_of_death!="All locations") %>%
+      group_by(dataset,year) %>%
+        summarise(total = sum(number_of_deaths))
+  }
+) %>% 
+  pivot_wider(names_from=year,values_from=total) %>%
+  knitr::kable(caption = "Number of deaths by year as calculated from different datasets.
+All datasets apart from the split by health board agree.
+The differences in the health board dataset likely reflect missing values and/or small numbers redacted from the dataset, which may have to do with the grouping of state hospital/non-NHS institutions.")
+```
+
+# Wrangling the data
+
+```{r}
+map2_dfr(
+  .x = list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
+  .y = c("overall","health_board","council_area","age_sex"),
+  .f = function(data_tbl, name) {
+    data_tbl %>%
+      mutate(dataset=name,year=as.numeric(year)) %>%
+      filter(year %in% 2016:2019 & week_number==53) %>%
+      count(dataset,year)
+  }
+) %>% 
+  pivot_wider(names_from=year,values_from=n) %>%
+  knitr::kable(caption = "Number of entries by dataset for week number 53 in years 2016-2019 (which only had 52 weeks).
+Only the 'overall' historical deaths dataset has week 53 entries for years with 52 weeks, and all the entries are for 0 deaths.")
+
+## helper function that creates week 53 data by taking week 52 in years with no year 53 and relabels it as 53
+## CAUTION: only use this for computing averages! otherwise it's making up data that wasnae there
+create_week_53_data_for_years_with_52_weeks <- function(data_tbl) {
+  real_data <-
+    data_tbl %>%
+    filter(!(year %in% 2016:2019 & week_number==53))  # remove any entries for week 53 in years with 52 weeks - in some datasets they existed but had 0 listed
+  fake_data <-
+    data_tbl %>%
+    filter(year %in% 2016:2019 & week_number==52) %>%
+    mutate(week_number=53)
+  return(
     bind_rows(
-      data_tbl,
-      data_tbl %>% filter(year_of_death %in% 2016:2019 & week_number==52) %>% mutate(week_number=53)  # copy records for week 52 in years 2016-2019 and count them as if they were week 53
+      real_data,
+      fake_data
     )
-  }) %>%
+  )
+}
+
+past_deaths_overall_summarised <-
+  past_deaths_overall %>%
+  filter(year!="average") %>%  # remove averages
+  group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
+  filter(!all(number_of_deaths==0 & week_number==53)) %>%
+  ungroup %>%
+  create_week_53_data_for_years_with_52_weeks %>%
   ungroup %>%
   group_by(week_number,place_of_death) %>% 
-  summarise(across(value, .fns = list(mean=mean,min=min,max=max)),  # having removed 0s for week 53 where it didn't exist, we average over years for week 53 that had that week only
+  summarise(across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),  # having removed 0s for week 53 where it didn't exist, we average over years for week 53 that had that week only
     .groups = "drop"
-  ) %>% rename_all(~str_replace_all(.x, pattern = "value", replacement = "deaths"))
+  ) %>% rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths"))
 
 week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
+
+# TODO: weekly averages may be too detailed for this level od subcategorising!
+
+historical_range_by_health_board <-
+  past_deaths_health_board %>%
+  create_week_53_data_for_years_with_52_weeks() %>%
+  group_by(
+    week_number, place_of_death, health_board
+  ) %>%
+  summarise(
+    deaths_mean = mean(number_of_deaths),
+    deaths_min = min(number_of_deaths),
+    deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+
+historical_range_by_council_area <-
+  past_deaths_council_area %>%
+  create_week_53_data_for_years_with_52_weeks() %>%
+  group_by(
+    week_number, place_of_death, council_area
+  ) %>%
+  summarise(
+    deaths_mean = mean(number_of_deaths),
+    deaths_min = min(number_of_deaths),
+    deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+
+historical_range_by_age_sex <-
+  past_deaths_age_sex %>%
+  create_week_53_data_for_years_with_52_weeks() %>%
+  group_by(
+    week_number, place_of_death, age, sex
+  ) %>%
+  summarise(
+    deaths_mean = mean(number_of_deaths),
+    deaths_min = min(number_of_deaths),
+    deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+
+historical_range_by_age <-
+  past_deaths_age_sex %>%
+  create_week_53_data_for_years_with_52_weeks() %>%
+  group_by(
+    week_number, place_of_death, age
+  ) %>%
+  summarise(
+    deaths_mean = mean(number_of_deaths),
+    deaths_min = min(number_of_deaths),
+    deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+
+historical_range_by_sex <-
+  past_deaths_age_sex %>%
+  create_week_53_data_for_years_with_52_weeks() %>%
+  group_by(
+    week_number, place_of_death, sex
+  ) %>%
+  summarise(
+    deaths_mean = mean(number_of_deaths),
+    deaths_min = min(number_of_deaths),
+    deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
 ```
@@
 ```{r}
-data_main <- 
-  data_raw %>%
+deaths_by_cause <- 
+  covid_deaths_raw %>%
   clean_names() %>%
-  rename(cause_of_death = causeofdeath, location_of_death = locationofdeath) %>%  # neater names
+  rename(
+    number_of_deaths = value,
+    place_of_death = location_of_death  # so it's the same as everywhere else!
+  ) %>%
   filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
   mutate(
-    ref_year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
+    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
     week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = ref_year),
-    value = as.numeric(value)  # convert to numeric!
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
   ) %>%
@@
 ## check date conversion
-data_main %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
+deaths_by_cause %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
 ```
@@
 
+Place of death was named differently between datasets (location/place) and the different places were also named slightly differently, so we find all of them and recode them to the same:
+
 ```{r}
-## recoding place of death
-data_main %>% count(location_of_death)
-past_deaths %>% count(place_of_death)
+unique_places_of_death_from_data <- map(
+  .x = list(deaths_by_cause, past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
+  .f = function(x) {x$place_of_death}
+  ) %>% unlist %>% unique %>% sort
 
-# datapasta::tribble_paste(input_table = tibble(location_of_death = sort(unique(data_main$location_of_death)), place_of_death = sort(unique(past_deaths$place_of_death))))
-tbl_recoded_place_of_death <- tibble::tribble(
-  ~location_of_death,           ~place_of_death,   ~place_of_death_recoded,
-  "hospital",                       "Hospital",   "Hospital",
-  "care-home",                      "Care Home",   "Care home",
-  "-non-institution", "Home and other non-institution",   "Home",
-  "other-institution",    "Other institution3",   "Other",
-  "all",                        "All locations",   "All"
-     ) %>% mutate(place_of_death_recoded = factor(place_of_death_recoded, levels = place_of_death_recoded))
+# datapasta::tribble_paste(input_table = tibble(place_of_death = unique_places_of_death_from_data))
+tbl_recoded_place_of_death <- 
+  tibble::tribble(
+                   ~place_of_death, ~place_of_death_recoded,
+                             "all", "All",
+                       "care-home", "Care home",
+                "-non-institution", "Home & other non-institution",
+               "other-institution", "Other",
+                        "hospital", "Hospital",
+                       "Care Home", "Care home",
+  "Home and other non-institution", "Home & other non-institution",
+                        "Hospital", "Hospital",
+              "Other institution3", "Other",
+                   "All locations", "All",
+          "Home / Non-institution", "Home & other non-institution",
+               "Other institution", "Other"
+  ) %>%
+  mutate(place_of_death_recoded = factor(place_of_death_recoded, levels = c("Hospital", "Care home", "Home & other non-institution", "Other", "All")))
+
+## helper function to recode place of death
+recode_place_of_death <- function(data_tbl) {
+  data_tbl %>%
+  left_join(x = ., y = tbl_recoded_place_of_death) %>% 
+  select(-place_of_death) %>%
+  distinct  # just in case?
+}
+
+## this is dumb, but it works!
+deaths_by_cause <- recode_place_of_death(deaths_by_cause)
+past_deaths_overall <- recode_place_of_death(past_deaths_overall)
+past_deaths_health_board <- recode_place_of_death(past_deaths_health_board)
+past_deaths_council_area <- recode_place_of_death(past_deaths_council_area)
+past_deaths_age_sex <- recode_place_of_death(past_deaths_age_sex)
+past_deaths_overall_summarised <- recode_place_of_death(past_deaths_overall_summarised)
+
+historical_range_by_age <- recode_place_of_death(historical_range_by_age)
+historical_range_by_age_sex <- recode_place_of_death(historical_range_by_age_sex)
+historical_range_by_sex <- recode_place_of_death(historical_range_by_sex)
+historical_range_by_council_area <- recode_place_of_death(historical_range_by_council_area)
+historical_range_by_health_board <- recode_place_of_death(historical_range_by_health_board)
 ```
 
-## Merged data
+## Merge weekly COVID/non-COVID-related deaths in 2020/2021 with historic death rates
+
+### Check weekly historic averages against NRS computed data
 
 ```{r}
-merged_data <-
-  data_main %>%
-  filter(cause_of_death!="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
-  pivot_wider(names_prefix = "deaths_", names_from=cause_of_death, values_from=value) %>%
-  clean_names() %>%
-  mutate(deaths_non_covid_related = deaths_all_causes - deaths_covid_19_related) %>%
-  left_join(tbl_recoded_place_of_death) %>%
-  left_join(past_deaths_summarised %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
-            by = c("place_of_death", "week_number", "sex", "age")) %>% 
-  select(-c(location_of_death,place_of_death)) %>%
-  mutate(
-    week_number_run_over = if_else(ref_year == 2021, 53 + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
-  )
-
-```
-
-### Check averages against previously computed data
-
-```{r}
-data_main %>%
-  left_join(tbl_recoded_place_of_death) %>%
-  left_join(past_deaths_summarised %>% mutate(sex = "all", age = "all", cause_of_death = "all-causes"),  # only merge for aggregated data b/c that's all we have!
-            by = c("place_of_death", "week_number", "sex", "age", "cause_of_death")) %>%
-# merged_data %>%
-  filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years" | (place_of_death_recoded=="All" & !is.na(deaths_mean))) %>%
-  select(ref_year,week_number,cause_of_death,value,deaths_mean) %>%
-  mutate(value = if_else(cause_of_death=="all-causes", NA_real_, value)) %>%  # make NA the 2020/2021 value
-  pivot_longer(cols=c(value,deaths_mean), names_to="computation", values_to="average", values_drop_na=TRUE) %>%
-  select(-cause_of_death) %>%
-  pivot_wider(values_from=average,names_from=computation) %>%
-  rename(nrs_average = value, my_average = deaths_mean) %>%
-  filter(round(nrs_average)!=round(my_average)) %>% 
+deaths_by_cause %>%
+  filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
+  select(week_number, year = year, place_of_death_recoded, nrs_computed_mean = number_of_deaths) %>%
+  left_join(past_deaths_overall_summarised) %>%
+  filter(round(nrs_computed_mean) != round(deaths_mean))
   knitr::kable(caption = "Subset of weekly deaths where calculate mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file.
@@
 
+### Merge figures for grouping by...
+
+```{r}
+## helpewr function to pivot all deaths and non-covid deaths into separate columns
+split_deaths_due_to_covid_and_not <- function(data_tbl) {
+  data_tbl %>%
+  filter(cause_of_death!="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
+  pivot_wider(names_prefix = "deaths_", names_from=cause_of_death, values_from=number_of_deaths) %>%
+  clean_names() %>%
+  mutate(deaths_non_covid_related = deaths_all_causes - deaths_covid_19_related)
+}
+
+## helper function to compute a "run-over" week number - weeks 54+ are in 2021
+compute_run_over_week_for_2021 <- function(data_tbl) {
+  data_tbl %>%
+  mutate(
+    week_number_run_over = if_else(year == 2021, 53 + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  )
+}
+
+merged_data_overall <-
+  deaths_by_cause %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(past_deaths_overall_summarised %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+            by = c("place_of_death_recoded", "week_number", "sex", "age")) %>% 
+  compute_run_over_week_for_2021
+
+# TODO: get health board & council area in ODS export!
+merged_deaths_by_health_board <-
+  deaths_by_cause %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(historical_range_by_health_board) %>%
+  compute_run_over_week_for_2021
+
+merged_deaths_by_council_area <-
+  deaths_by_cause %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(historical_range_by_council_area) %>%
+  compute_run_over_week_for_2021
+
+merged_deaths_by_age_sex <-
+  deaths_by_cause %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(historical_range_by_age_sex) %>%
+  compute_run_over_week_for_2021
+
+merged_deaths_by_age <-
+  deaths_by_cause %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(historical_range_by_age) %>%
+  compute_run_over_week_for_2021
+
+merged_deaths_by_sex <-
+  deaths_by_cause %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(historical_range_by_sex) %>%
+  compute_run_over_week_for_2021
+```
+
 # Visualisations
+
+## Common elements
+
+```{r}
+date_range_text <- glue::glue("Data from w/c {format(min(deaths_by_cause$date_w_c), '%d %B %Y')} to w/c {format(max(deaths_by_cause$date_w_c), '%d %B %Y')} (inclusive).")
+
+source_nrs_text <- "Source: National Records of Scotland"
+```
 
@@
 ## graph format copied from: https://github.com/davidhen/nrs_covid_stats/
-
-date_range_text <- glue::glue("Data from w/c {format(min(data_main$date_w_c), '%d %B %Y')} to w/c {format(max(data_main$date_w_c), '%d %B %Y')} (inclusive).")
-
-source_nrs_text <- "Source: National Records of Scotland"
-
-data_main %>% 
-  filter(sex == "all" & location_of_death == "all" & cause_of_death == "covid-19-related") %>% 
+deaths_by_cause %>% 
+  filter(sex == "all" & place_of_death == "all" & cause_of_death == "covid-19-related") %>% 
   filter(age != "all") %>% 
   group_by(age) %>% 
-  summarise(n = sum(value)) %>%
+  summarise(n = sum(number_of_deaths)) %>%
   mutate(pct = round(n/sum(n), 3)) %>% 
@@
 
-removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data %>% filter(ref_year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data %>% filter(ref_year==2021) %>% .$week_number)} in 2021.")
+removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data %>% filter(year==2021) %>% .$week_number)} in 2021.")
 
@@
     ~place_of_death_recoded, ~x, ~y, ~label,
-                 "Hospital", 35L, 800, "Hospital deaths dipped below historical minimum and remained low until Winter.",
-                 "Hospital", 35L, 250, "Shaded area shows range of deaths 2015-2019 dashed line shows weekly average",
+                 "Hospital", 20L, 850, "Shaded area shows range of 2015-2019 deaths; dashed line shows weekly average",
+                 "Hospital", 25L, 250, "Hospital deaths dipped below historical minimum and remained low until Winter.",
                 "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
@@
 
-figure_6_location_of_death <-
+# figure_6_location_of_death <-
   merged_data %>%
@@
 ```
+
+## Figure 7: deaths at home
+
+```{r}
+# TODO
+
+```
+
+## Figure 8: deaths at home, by health board
+
+```{r}
+# TODO: fix this once we have health board in weekly deaths!
+merged_deaths_by_health_board %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death_recoded == "Home & other non-institution") %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  # geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~health_board) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       # caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+```
+
+## Figure 10: deaths at home by sex
+
+```{r}
+
+```

### 2021-02-02 covid_nrs_place_of_death.Rmd
Reworking the script to reflect data availability from the two sources - doesn't run at the minute but just wanted to save changes before lunch

@@
 library(phsmethods)  # methods for working with PHS data
+library(readxl)  # for reading excel files
 # library(SPARQL)  # taken care of by opendatascot
@@
 
-# Import data
+# Import & wrangle data
 
-This analysis is based off two sets of data:
+We're importing two kinds of data:
 
-1. Weekly deaths involving COVID-19 from Scottish gov't open data service
-  - this is imported using the `opendatascot` package in R, and connects to the data available at: [the Scottish Gov't open data   website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)
-2. Historical deaths 2015-2019, grouped by week & location
-  - These are available at the NRS website on [deaths involving COVID-19]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland), specifically under the 'Related Statistics' section, under the heading 'Ad-hoc queries'.
-  - We load the total deaths, but also grouped by:
-    - Health board
-    - Local Authority
-    - Sex
-    - Age
-    - ~~Ethnicity~~ Note: not sure this is available
+1. Weekly deaths involving COVID-19 from Scottish gov't open data service, broken down by week & place of death (home or institution)
+2. Historical deaths 2015-2019, grouped by week & place of death
 
-Note that the weekly deaths due to COVID-19 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the other datasets.
-The same holds for the age & sex breakdown.
+For both of these, we'll also look at crosstabulated figures for:
+
+* Health board
+* Local Authority
+* Sex
+* Age
+* ~~Ethnicity~~ Note: not sure this is available
+
+There are two ways of obtaining the data:
+
+1.
+Using their SPARQL-based API ("data cube"), which can be accessed using the `opendatascot` package in R; the data can be seen on [the Scottish Gov't open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)
+2.
+There are a number of dataset, including the demographic/geographic crosstabulations available on the "Related Statistics" section of the [deaths involving COVID-19 website]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland)
+
+Note that the [*About* tab of the "data cube" interface](https://statistics.gov.scot/resource?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19) says the following about Health Board & Local Authority figures:
+
+>Health Board and Local Authority figures include non-residents.
+Deaths are allocated to areas based on the usual residence of the deceased.
+If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.
+
+This means we'll need to download those figures from the website rather than using the API
+
+Note that the weekly deaths due to COVID-19 in 2020 & 2021 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the weekly figures for individual years & the appropriate cross-tabulations.
+
+## Parsing week numbers
+
+In the `opendatascot`-imported data, `refPeriod` refers to the week number (called `week of occurrence` in the `.xlsx`/`.csv`-imported data has), as defined by [ISO 8601](https://en.wikipedia.org/wiki/ISO_week_date), and can be computed from a date with `lubridate::isoweek()`; the first week of 2020 is the week starting with Monday 2019-12-30, ~~and the corresponding start dates for each week can be computed by adding weeks using `lubridate::dweeks()` to that date.~~ This is non-trivial, because the number of iso weeks in a year isn't constant.
+The `ISOweek::ISOweek2date()` function deals with this issue in a crude manner, and allows us to compute the date (w/c on date) from the year & week number.
+It relies on being provided a valid week number (some years have 52 and some have 53), but it's probably safe to assume that the NRS data has valid dates.
+
+Invalid week numbers will produce a date in the next year: e.g.
+week 53 in 2021 would actually start in 2022, which makes it invalid.
+We can test for that by checking that the year provided is the same year in the date produced.
+
+We'll also compute a 'run-over' week number of 2021 in case we plot both years on a contibuous X-axis - in this case, week 1 of 2021 follows week 53 of 2020, so we start counting from 54 in 2021.
+
+```{r}
+date_first_week_2020 <- ymd("2019-12-30")
+ISOweek::ISOweek2date(weekdate = "2019-W53-1")
+ISOweek::ISOweek2date(weekdate = "2020-W01-1")
+ISOweek::ISOweek2date(weekdate = "2020-W52-1")
+ISOweek::ISOweek2date(weekdate = "2020-W53-1")
+ISOweek::ISOweek2date(weekdate = "2021-W01-1")
+ISOweek::ISOweek2date(weekdate = "2021-W53-1")
+
+compute_start_date_from_week_number <- function(week_number, year_number) {
+  ## this assumes a valid week number! There are 52 or 53 weeks in a year
+  isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
+  computed_dates <- ISOweek::ISOweek2date(isoweek_string)
+  ## invalid weeks will result in a computed date in a year after the requested year...
+  ## a valid week can result in a date that starts in the year before!
+  if (any(year_number<year(ymd(computed_dates)))) stop("invalid week number provided!")
+  return(computed_dates)
+}
+
+tibble(
+    week_number = c(1,2,3,52,53)
+  ) %>%
+  mutate(
+    date = compute_start_date_from_week_number(week_number = week_number, year_number = 2020),
+    iso_week_from_date = isoweek(date)
+  ) %>% knitr::kable(caption = "Examples of computing start date for each week, and computing isoweek from the dates!")
+```
 
 ## Load weekly COVID-19 related deaths
+
+~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above!
 
@@
 # ods_print_query("deaths-involving-coronavirus-covid-19")
-ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003
+# ods_print_query("deaths-involving-coronavirus-covid-19", geography = "la")
+# ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003
+# phsmethods::match_area("S92000003")  # "Scotland"
+
+make_human_readable_ref_area <- function(data_tbl, ref_area = refArea) {
+  enquoted_ref_area <- enquo(ref_area)
+  data_tbl %>%
+    mutate(
+      !!enquoted_ref_area := phsmethods::match_area(!!enquoted_ref_area)
+    )
+}
+# raw_covid_deaths <- list(
+#   scotland = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc"),
+#   health_board = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "hb"),
+#   local_authority = ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la")
+# ) %>%
+#   map(.x = ., .f = ~make_human_readable_ref_area(data_tbl = .x))
+
+weekly_deaths_by_age_sex <-
+  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
+  make_human_readable_ref_area %>%
+  clean_names() %>%
+  rename(
+    number_of_deaths = value,
+    place_of_death = location_of_death  # so it's the same as everywhere else!
+  ) %>%
+  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021
+  mutate(
+    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
+    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  ) %>%
+  select(-c(measure_type))  #  measure_type is count for all data
+
+weekly_deaths_by_la_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
+                destfile = tempfile(),
+                quiet = FALSE)
+
+weekly_deaths_by_la <- 
+  read_excel(weekly_deaths_by_la_file, sheet = 2, skip = 2) %>% 
+  clean_names %>%
+  rename(week_number = week_of_occurrence, ref_area = council, place_of_death = location_of_death, number_of_deaths = deaths) %>%
+  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
+  mutate(
+    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
+    week_number = parse_integer(str_sub(week_number, 4, 5)),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year)
+  )
+
+weekly_deaths_by_hb_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
+                destfile = tempfile(),
+                quiet = FALSE)
+
+weekly_deaths_by_hb <- 
+  read_excel(weekly_deaths_by_hb_file, sheet = 2, skip = 2) %>% 
+  clean_names %>%
+  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
+  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
+  mutate(
+    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
+    week_number = parse_integer(str_sub(week_number, 4, 5)),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year)
+  )
 
@@
 )
+
+ods_dataset("deaths-involving-coronavirus-covid-19")
+
 ```
@@
 ```{r}
+covid_deaths <-
+  raw_covid_deaths %>%
+  map(
+    .x = .,
+    .f = function(data_tbl) {
+      data_tbl %>%
+      clean_names() %>%
+      rename(
+        number_of_deaths = value,
+        place_of_death = location_of_death  # so it's the same as everywhere else!
+      ) %>%
+      filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
+      mutate(
+        year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
+        week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
+        date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+        number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
+      ) %>%
+      select(-c(measure_type))  #  measure_type is count for all data
+    }
+  )
+
+# TODO: remove this
 deaths_by_cause <- 
@@
 ## check date conversion
-deaths_by_cause %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
+covid_deaths$scotland %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
 ```
@@
 unique_places_of_death_from_data <- map(
-  .x = list(deaths_by_cause, past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
+  .x = c(covid_deaths,list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex)),
   .f = function(x) {x$place_of_death}
@@
 ## this is dumb, but it works!
-deaths_by_cause <- recode_place_of_death(deaths_by_cause)
+covid_deaths <- map(.x = covid_deaths, .f = ~recode_place_of_death(.x))
+# deaths_by_cause <- recode_place_of_death(deaths_by_cause)
 past_deaths_overall <- recode_place_of_death(past_deaths_overall)
@@
 ```{r}
-deaths_by_cause %>%
+covid_deaths$scotland %>%
   filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
@@
 merged_data_overall <-
-  deaths_by_cause %>%
+  covid_deaths$scotland %>%
   split_deaths_due_to_covid_and_not %>%
@@
 merged_deaths_by_health_board <-
-  deaths_by_cause %>%
+  covid_deaths$health_board %>%
   split_deaths_due_to_covid_and_not %>%
-  left_join(historical_range_by_health_board) %>%
+  # inner_join(historical_range_by_health_board, by = c("week_number","place_of_death_recoded","ref_area"="health_board")) %>%
+  inner_join(historical_range_by_health_board, by = c("week_number","ref_area"="health_board","place_of_death_recoded")) %>%
   compute_run_over_week_for_2021
@@
 
-removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data %>% filter(year==2021) %>% .$week_number)} in 2021.")
+removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
 
@@
 # figure_6_location_of_death <-
-  merged_data %>%
+  merged_data_overall %>%
   filter(sex == "all" & age == "all") %>%
@@
 ```{r}
-
+merged_deaths_by_sex %>%
+  filter(sex != "all" & age == "all") %>%
+  filter(place_of_death_recoded == "Home & other non-institution") %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  # geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~sex) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       # caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
 ```

### 2021-02-02 covid_nrs_place_of_death.Rmd
finally icnorporated the different datasets & cleaned them properly & completed implicit missing figures for accurate plotting. HB & LA plots work! Yay!

@@
 
+## Consistent pPlace of death coding
+
+Place of death was named differently between datasets (location/place) and the different places were also named slightly differently, so we find all of them and recode them to the same:
+
+```{r}
+order_of_place_of_death_levels <- c("Hospital", "Care home", "Home & other non-institution", "Other", "All")
+
+recode_place_of_death <- function(data_tbl) {
+  data_tbl %>%
+    mutate(place_of_death = case_when(
+      str_detect(place_of_death, pattern = regex(pattern = "all", ignore_case = TRUE)) ~ "All",
+      str_detect(place_of_death, pattern = regex(pattern = "care.*home", ignore_case = TRUE)) ~ "Care home",
+      str_detect(place_of_death, pattern = regex(pattern = "(non.*institut)|home", ignore_case = TRUE)) ~ "Home & other non-institution",
+      str_detect(place_of_death, pattern = regex(pattern = "hospital", ignore_case = TRUE)) ~ "Hospital",
+      str_detect(place_of_death, pattern = regex(pattern = "other", ignore_case = TRUE)) ~ "Other",
+      TRUE ~ NA_character_
+      ) %>%
+        factor(x = ., levels = order_of_place_of_death_levels)
+    )
+}
+```
+
+## Consistent Cause of death coding
+
+The `opendatascot`-imported data distinguishes "all causes" deaths from "covid-related" deaths - the non-covid-related deaths are the difference between those.
+In contrast, the other weekly datasets code cause of death as either covid-related or non-covid-related, in which case we need to compute all cause deaths as the sum of the two.
+
+We'll change the wording of these to a consistent format so it's easier to compute consistent columns.
+The below function will also compute non-covid deaths for data where all deaths and covid deaths are provided, and all deaths where covid- and non-covid-related deaths are provided.
+
+This involves pivotting the number of deaths variable into wide format based on the cause of death. Because of this, we need to deal with implicit missing values:
+
+The first COVID-19-related deaths were reported from week 12 onwards.
+In some of the datasets, a missing/omitted value represents 0 cases in that category, when a 0 should have been recorded.
+E.g.
+deaths in "Other institutions" are very low, and if a record is missing for that category, we should assume 0 deaths.
+
+```{r}
+## helper function to make consistently coded cause of death
+recode_cause_of_death <- function(data_tbl) {
+  data_tbl %>%
+    mutate(
+      cause_of_death = case_when(
+        cause_of_death == "all-causes" ~ "deaths_all_causes",
+        cause_of_death == "all-causes-average-of-corresponding-week-over-previous-5-years" ~ "deaths_nrs_past_average_all_causes",
+        cause_of_death == "covid-19-related" ~ "deaths_covid_related",
+        cause_of_death == "COVID-19 mentioned" ~ "deaths_covid_related",
+        cause_of_death == "Non-COVID-19" ~ "deaths_non_covid",
+        TRUE ~ NA_character_
+      )
+    )
+}
+
+## find which combinations of geographic area, cause of death, and place of death are missing and set the number of deaths to 0 for them
+complete_weekly_deaths_with_implicit_zeros <- function(data_tbl) {
+  data_tbl %>%
+  tidyr::complete(
+      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
+      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
+      fill = list(number_of_deaths = 0)
+      )
+}
+
+## covid death reporting started in week 12, 2020 - before then it doesn't make sense to distunguish between covid and non-covid deaths, and only the all cause death figure should be kept
+## by using the 'overrun' weeks we ensure that 2021 or later isn't affected
+## note: we're keeping all cause deaths!
+set_covid_and_non_covid_deaths_to_zero_before_reporting_started <- function(data_tbl) {
+  data_tbl %>%
+    mutate(
+      deaths_non_covid = if_else(
+        condition = week_number_run_over < 12,
+        true = NA_real_,
+        false = deaths_non_covid
+      ),
+      deaths_covid_related = if_else(
+        condition = week_number_run_over < 12,
+        true = NA_real_,
+        false = deaths_covid_related
+      )
+    )
+}
+
+# TODO: delete below function, it's too complicated and I've broken it down to the individual functions
+# recode_cause_of_death_and_compute_columns <- function(data_tbl) {
+#   potential_grouping_variables <- c("ref_area","sex","age","ref_period")
+#   grouping_variables <- potential_grouping_variables[which(potential_grouping_variables %in% names(data_tbl))]
+#   
+#   interim_data_tbl <- 
+#     data_tbl %>%
+#     recode_cause_of_death() %>%
+#     tidyr::complete(
+#       nesting(year, week_number, week_number_run_over, date_w_c),
+#       !!!rlang::syms(grouping_variables),
+#       cause_of_death, place_of_death,
+#       fill = list(number_of_deaths = 0)
+#       ) %>%
+#     pivot_wider(names_from = "cause_of_death", values_from = "number_of_deaths")
+#   
+#   ## compute non-covid deaths from all deaths minus covid deaths
+#   if (!"deaths_non_covid" %in% names(interim_data_tbl)) {
+#     interim_data_tbl <-
+#       interim_data_tbl %>%
+#       mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)
+#   }
+#   ## compute all deaths from non-covid deaths + covid deaths
+#   if (!"deaths_all_causes" %in% names(interim_data_tbl)) {
+#     interim_data_tbl <-
+#       interim_data_tbl %>%
+#       mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related)
+#   }
+#   
+#   interim_data_tbl <-
+#     interim_data_tbl %>%  # set as missing covid- & non-covid-related deaths before week 12
+#     mutate(
+#       deaths_non_covid = if_else(
+#         condition = week_number_run_over < 12,
+#         true = NA_real_,
+#         false = deaths_non_covid
+#       ),
+#       deaths_covid_related = if_else(
+#         condition = week_number_run_over < 12,
+#         true = NA_real_,
+#         false = deaths_covid_related
+#       )
+#     )
+#   
+#   return(interim_data_tbl)
+# }
+```
+
+## Consistent use of missing value
+
+```{r}
+replace_na_with_0_from_week_12 <- function(data_tbl) {
+  data_tbl %>%
+    mutate(
+      value = if_else(
+        condition = week_number_roll_over >=12 & is.na(number_of_deaths),
+        true = 0,
+        false = number_of_deaths
+      )
+    )
+}
+```
+
 ## Load weekly COVID-19 related deaths
@@
 
+## helper function that links the 9-character geography code to its name
 make_human_readable_ref_area <- function(data_tbl, ref_area = refArea) {
@@
 }
+
 # raw_covid_deaths <- list(
@@
 
-weekly_deaths_by_age_sex <-
+weekly_deaths_age_sex <-
   ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
@@
   clean_names() %>%
-  rename(
-    number_of_deaths = value,
-    place_of_death = location_of_death  # so it's the same as everywhere else!
-  ) %>%
-  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021
+  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
+  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
+  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
+  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
+  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
+  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
+  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
+  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
+  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
+  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021, we only want the weekly figures
   mutate(
@@
   ) %>%
-  select(-c(measure_type))  #  measure_type is count for all data
+  select(-c(measure_type)) %>%  #  measure_type is count for all data
+  recode_place_of_death() %>%
+  recode_cause_of_death() %>%
+  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
+  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing
 
-weekly_deaths_by_la_file <-
+weekly_deaths_la_file <-
   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
@@
 
-weekly_deaths_by_la <- 
-  read_excel(weekly_deaths_by_la_file, sheet = 2, skip = 2) %>% 
+weekly_deaths_la <- 
+  read_excel(weekly_deaths_la_file, sheet = 2, skip = 2) %>% 
   clean_names %>%
@@
     week_number = parse_integer(str_sub(week_number, 4, 5)),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year)
-  )
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  ) %>%
+  recode_cause_of_death() %>%
+  complete_weekly_deaths_with_implicit_zeros() %>%
+  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
+  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
+  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
+  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
 
-weekly_deaths_by_hb_file <-
+weekly_deaths_hb_file <-
   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
@@
 
-weekly_deaths_by_hb <- 
-  read_excel(weekly_deaths_by_hb_file, sheet = 2, skip = 2) %>% 
+weekly_deaths_hb <- 
+  read_excel(weekly_deaths_hb_file, sheet = 2, skip = 2) %>% 
   clean_names %>%
@@
     week_number = parse_integer(str_sub(week_number, 4, 5)),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year)
-  )
-
-# TODO: split this into a scotland-wide dataset, and grouped by health board & council area
-covid_deaths_raw <- bind_rows(
-  ods_dataset("deaths-involving-coronavirus-covid-19",geography = "sc")#,
-  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la"),
-  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "hb")
-)
-
-ods_dataset("deaths-involving-coronavirus-covid-19")
-
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  ) %>%
+  recode_cause_of_death() %>%
+  complete_weekly_deaths_with_implicit_zeros() %>%
+  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
+  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # make separate columns for causes of death figures
+  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
+  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
 ```
 
-## Download historical deaths data
-
-These are all downloaded from the NRS website to a temporary file created by the `tempfile()` function.
+### Checking week number coding
 
 ```{r}
+weekly_deaths_age_sex %>% 
+  count(ref_period, date_w_c, week_number, week_number_run_over) %>% 
+  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
+  select(-n) %>% 
+  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
+```
+
+## Load historical deaths data
+
+The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data.
+
+```{r, warning=FALSE}
 ## overall file - where did I get the link from? it's not obviously available from the website...
@@
 
-past_deaths_health_board_file <-
+past_deaths_hb_file <-
   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
@@
 
-past_deaths_council_area_file <-
+past_deaths_la_file <-
   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
@@
                 quiet = FALSE)
-```
 
-### Overall data
-
-```{r}
 past_deaths_overall <- 
@@
   pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
-  mutate(week_number=as.integer(week_number))
+  mutate(week_number=as.integer(week_number)) %>%
+  recode_place_of_death()
 
-# past_deaths_overall %>% filter(week_number %in% 52:53) %>% count(year, week_number)
-```
-
-### Grouped by health board, council area, and age & sex
-
-The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data.
-
-The total deaths by place of death don't add up between the overall deaths & health board. This may be due to small numbers being redacted?
-
-```{r, warning=FALSE}
-past_deaths_health_board <- 
+past_deaths_hb <- 
   readxl::read_excel(past_deaths_health_board_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
   clean_names() %>%
+  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
   rename(
@@
   ) %>%
-  drop_na(week_number)  # remove empty rows & one row where it's just the copyright notice
+  recode_place_of_death()
 
-past_deaths_council_area <-
+past_deaths_la <-
   readxl::read_excel(past_deaths_council_area_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
   clean_names() %>%
+  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
   rename(
@@
   ) %>%
-  drop_na(week_number)  # remove empty rows & one row where it's just the copyright notice
+  recode_place_of_death()
 
@@
   readxl::read_excel(past_deaths_age_sex_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
+  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
   mutate(
@@
   ) %>%
-  drop_na(year)  # remove empty rows & one row where it's just the copyright notice
+  recode_place_of_death()
 
@@
       group_by(dataset,year) %>%
-        summarise(total = sum(number_of_deaths))
+        summarise(total = sum(number_of_deaths), .groups = "drop")
   }
@@
 
-# Wrangling the data
+# Calculate historical averages, minima & maxima
 
-```{r}
+Due to the counting of weeks, most years have 52 weeks with some years (2015 & 2020) having 53.
+Therefore we need to decide how to deal with week 53 in 2020 since only 2015 has a corresponding weekly count.
+The solution used here is to calculate a weekly rate for week 53 from week 53 in 2015, and week 52 in 2016-2019 - this represents the "last week of the year", which includes December 31st.
+
+```{r, warning=FALSE}
 map2_dfr(
-  .x = list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
+  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_la, past_deaths_age_sex),
   .y = c("overall","health_board","council_area","age_sex"),
@@
 
-past_deaths_overall_summarised <-
+# TODO: clean up once these have been dealth with
+
+historical_average_overall <-
   past_deaths_overall %>%
@@
   group_by(week_number,place_of_death) %>% 
-  summarise(across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),  # having removed 0s for week 53 where it didn't exist, we average over years for week 53 that had that week only
+  summarise(
+    across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
     .groups = "drop"
@@
 
-historical_range_by_health_board <-
-  past_deaths_health_board %>%
+historical_average_hb <-
+  past_deaths_hb %>%
   create_week_53_data_for_years_with_52_weeks() %>%
@@
 
-historical_range_by_council_area <-
-  past_deaths_council_area %>%
+historical_average_la <-
+  past_deaths_la %>%
   create_week_53_data_for_years_with_52_weeks() %>%
@@
 
-historical_range_by_age_sex <-
+historical_average_age_sex <-
   past_deaths_age_sex %>%
@@
 
-historical_range_by_age <-
+historical_average_age <-
   past_deaths_age_sex %>%
@@
 
-historical_range_by_sex <-
+historical_average_sex <-
   past_deaths_age_sex %>%
@@
 
-## Understanding how dates are coded
-
-`refPeriod` refers to the week number, as defined by [ISO 8601](https://en.wikipedia.org/wiki/ISO_week_date), and can be computed from a date with `lubridate::isoweek()`; the first week of 2020 is the week starting with Monday 2019-12-30, ~~and the corresponding start dates for each week can be computed by adding weeks using `lubridate::dweeks()` to that date.~~ This is non-trivial, because the number of iso weeks in a year isn't constant.
-The `ISOweek::ISOweek2date()` function deals with this issue in a crude manner, and allows us to compute the date (w/c on date) from the year & week number.
-It relies on being provided a valid week number (some years have 52 and some have 53), but it's probably safe to assume that the NRS data has valid dates.
-
-Invalid week numbers will produce a date in the next year: e.g.
-week 53 in 2021 would actually start in 2022, which makes it invalid.
-We can test for that by checking that the year provided is the same year in the date produced.
+## Check weekly historic averages against NRS computed data
 
 ```{r}
-date_first_week_2020 <- ymd("2019-12-30")
-ISOweek::ISOweek2date(weekdate = "2019-W53-1")
-ISOweek::ISOweek2date(weekdate = "2020-W01-1")
-ISOweek::ISOweek2date(weekdate = "2020-W52-1")
-ISOweek::ISOweek2date(weekdate = "2020-W53-1")
-ISOweek::ISOweek2date(weekdate = "2021-W01-1")
-ISOweek::ISOweek2date(weekdate = "2021-W53-1")
-
-compute_start_date_from_week_number <- function(week_number, year_number) {
-  ## this assumes a valid week number! There are 52 or 53 weeks in a year
-  isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
-  computed_dates <- ISOweek::ISOweek2date(isoweek_string)
-  ## invalid weeks will result in a computed date in a year after the requested year...
-  ## a valid week can result in a date that starts in the year before!
-  if (any(year_number<year(ymd(computed_dates)))) stop("invalid week number provided!")
-  return(computed_dates)
-}
-
-tibble(
-    week_number = c(1,2,3,52,53)
-  ) %>%
-  mutate(
-    date = compute_start_date_from_week_number(week_number = week_number, year_number = 2020),
-    iso_week_from_date = isoweek(date)
-  ) %>% knitr::kable(caption = "Examples of computing start date for each week, and computing isoweek from the dates!")
-```
-
-## Recoding dates
-
-```{r}
-covid_deaths <-
-  raw_covid_deaths %>%
-  map(
-    .x = .,
-    .f = function(data_tbl) {
-      data_tbl %>%
-      clean_names() %>%
-      rename(
-        number_of_deaths = value,
-        place_of_death = location_of_death  # so it's the same as everywhere else!
-      ) %>%
-      filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
-      mutate(
-        year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
-        week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
-        date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-        number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
-      ) %>%
-      select(-c(measure_type))  #  measure_type is count for all data
-    }
-  )
-
-# TODO: remove this
-deaths_by_cause <- 
-  covid_deaths_raw %>%
-  clean_names() %>%
-  rename(
-    number_of_deaths = value,
-    place_of_death = location_of_death  # so it's the same as everywhere else!
-  ) %>%
-  filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
-  mutate(
-    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
-    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
-  ) %>%
-  select(-c(ref_area, measure_type))  # ref_area is Scotland for all records; measure_type is count
-
-## check date conversion
-covid_deaths$scotland %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
-```
-
-## Recoding place of death
-
-Place of death was named differently between datasets (location/place) and the different places were also named slightly differently, so we find all of them and recode them to the same:
-
-```{r}
-unique_places_of_death_from_data <- map(
-  .x = c(covid_deaths,list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex)),
-  .f = function(x) {x$place_of_death}
-  ) %>% unlist %>% unique %>% sort
-
-# datapasta::tribble_paste(input_table = tibble(place_of_death = unique_places_of_death_from_data))
-tbl_recoded_place_of_death <- 
-  tibble::tribble(
-                   ~place_of_death, ~place_of_death_recoded,
-                             "all", "All",
-                       "care-home", "Care home",
-                "-non-institution", "Home & other non-institution",
-               "other-institution", "Other",
-                        "hospital", "Hospital",
-                       "Care Home", "Care home",
-  "Home and other non-institution", "Home & other non-institution",
-                        "Hospital", "Hospital",
-              "Other institution3", "Other",
-                   "All locations", "All",
-          "Home / Non-institution", "Home & other non-institution",
-               "Other institution", "Other"
-  ) %>%
-  mutate(place_of_death_recoded = factor(place_of_death_recoded, levels = c("Hospital", "Care home", "Home & other non-institution", "Other", "All")))
-
-## helper function to recode place of death
-recode_place_of_death <- function(data_tbl) {
-  data_tbl %>%
-  left_join(x = ., y = tbl_recoded_place_of_death) %>% 
-  select(-place_of_death) %>%
-  distinct  # just in case?
-}
-
-## this is dumb, but it works!
-covid_deaths <- map(.x = covid_deaths, .f = ~recode_place_of_death(.x))
-# deaths_by_cause <- recode_place_of_death(deaths_by_cause)
-past_deaths_overall <- recode_place_of_death(past_deaths_overall)
-past_deaths_health_board <- recode_place_of_death(past_deaths_health_board)
-past_deaths_council_area <- recode_place_of_death(past_deaths_council_area)
-past_deaths_age_sex <- recode_place_of_death(past_deaths_age_sex)
-past_deaths_overall_summarised <- recode_place_of_death(past_deaths_overall_summarised)
-
-historical_range_by_age <- recode_place_of_death(historical_range_by_age)
-historical_range_by_age_sex <- recode_place_of_death(historical_range_by_age_sex)
-historical_range_by_sex <- recode_place_of_death(historical_range_by_sex)
-historical_range_by_council_area <- recode_place_of_death(historical_range_by_council_area)
-historical_range_by_health_board <- recode_place_of_death(historical_range_by_health_board)
-```
-
-## Merge weekly COVID/non-COVID-related deaths in 2020/2021 with historic death rates
-
-### Check weekly historic averages against NRS computed data
-
-```{r}
-covid_deaths$scotland %>%
-  filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
-  select(week_number, year = year, place_of_death_recoded, nrs_computed_mean = number_of_deaths) %>%
-  left_join(past_deaths_overall_summarised) %>%
-  filter(round(nrs_computed_mean) != round(deaths_mean))
-  knitr::kable(caption = "Subset of weekly deaths where calculate mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file.
+weekly_deaths_age_sex %>%
+  filter(sex=="all" & age=="all" & cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
+  select(week_number, year = year, place_of_death, nrs_computed_mean = number_of_deaths) %>%
+  left_join(historical_average_overall) %>%
+  filter(round(nrs_computed_mean) != round(deaths_mean)) %>%
+  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file.
 Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019.
@@
 
-### Merge figures for grouping by...
+# Merge weekly COVID/non-COVID-related deaths in 2020/2021 with historic death rates
 
 ```{r}
-## helpewr function to pivot all deaths and non-covid deaths into separate columns
-split_deaths_due_to_covid_and_not <- function(data_tbl) {
-  data_tbl %>%
-  filter(cause_of_death!="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
-  pivot_wider(names_prefix = "deaths_", names_from=cause_of_death, values_from=number_of_deaths) %>%
-  clean_names() %>%
-  mutate(deaths_non_covid_related = deaths_all_causes - deaths_covid_19_related)
-}
+merged_data_overall <-
+  weekly_deaths_age_sex %>%
+  filter(sex=="all" & age=="all") %>%
+  # covid_deaths$scotland %>%
+  split_deaths_due_to_covid_and_not %>%
+  left_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+            by = c("place_of_death", "week_number", "sex", "age"))
 
-## helper function to compute a "run-over" week number - weeks 54+ are in 2021
-compute_run_over_week_for_2021 <- function(data_tbl) {
-  data_tbl %>%
-  mutate(
-    week_number_run_over = if_else(year == 2021, 53 + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
-  )
-}
+# TODO: complete these for age & sex also
+merged_deaths_hb <-
+  weekly_deaths_hb %>%
+  inner_join(historical_average_hb, by = c("week_number","ref_area"="health_board","place_of_death"))
 
-merged_data_overall <-
-  covid_deaths$scotland %>%
-  split_deaths_due_to_covid_and_not %>%
-  left_join(past_deaths_overall_summarised %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
-            by = c("place_of_death_recoded", "week_number", "sex", "age")) %>% 
-  compute_run_over_week_for_2021
+merged_deaths_la <-
+  weekly_deaths_la %>%
+  inner_join(historical_average_la, by = c("week_number","ref_area"="council_area","place_of_death"))
 
-# TODO: get health board & council area in ODS export!
-merged_deaths_by_health_board <-
-  covid_deaths$health_board %>%
-  split_deaths_due_to_covid_and_not %>%
-  # inner_join(historical_range_by_health_board, by = c("week_number","place_of_death_recoded","ref_area"="health_board")) %>%
-  inner_join(historical_range_by_health_board, by = c("week_number","ref_area"="health_board","place_of_death_recoded")) %>%
-  compute_run_over_week_for_2021
-
-merged_deaths_by_council_area <-
-  deaths_by_cause %>%
-  split_deaths_due_to_covid_and_not %>%
-  left_join(historical_range_by_council_area) %>%
-  compute_run_over_week_for_2021
-
-merged_deaths_by_age_sex <-
-  deaths_by_cause %>%
-  split_deaths_due_to_covid_and_not %>%
-  left_join(historical_range_by_age_sex) %>%
-  compute_run_over_week_for_2021
-
-merged_deaths_by_age <-
-  deaths_by_cause %>%
-  split_deaths_due_to_covid_and_not %>%
-  left_join(historical_range_by_age) %>%
-  compute_run_over_week_for_2021
-
-merged_deaths_by_sex <-
-  deaths_by_cause %>%
-  split_deaths_due_to_covid_and_not %>%
-  left_join(historical_range_by_sex) %>%
-  compute_run_over_week_for_2021
+# merged_deaths_by_age_sex <-
+#   deaths_by_cause %>%
+#   split_deaths_due_to_covid_and_not %>%
+#   left_join(historical_range_by_age_sex) %>%
+#   compute_run_over_week_for_2021
+# 
+# merged_deaths_by_age <-
+#   deaths_by_cause %>%
+#   split_deaths_due_to_covid_and_not %>%
+#   left_join(historical_range_by_age) %>%
+#   compute_run_over_week_for_2021
+# 
+# merged_deaths_by_sex <-
+#   deaths_by_cause %>%
+#   split_deaths_due_to_covid_and_not %>%
+#   left_join(historical_range_by_sex) %>%
+#   compute_run_over_week_for_2021
 ```
@@
 # TODO: fix this once we have health board in weekly deaths!
-merged_deaths_by_health_board %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death_recoded == "Home & other non-institution") %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
+merged_deaths_hb %>%
+  # filter(sex == "all" & age == "all") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
                names_to = "deaths_type", values_to = "deaths") %>% 
@@
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~health_board) +
+  facet_wrap(~ref_area) +
   theme_minimal() +
@@
        )
+
+merged_deaths_la %>%
+  # filter(sex == "all" & age == "all") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  # geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~ref_area) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       # caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+
 ```

### 2021-02-03 covid_nrs_place_of_death.Rmd
added tidyr::complete to historical figures also - it was suspicious seeing no historical ranges dip down to 0!

@@
   ) %>%
-  recode_place_of_death()
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    health_board, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
 
@@
   ) %>%
-  recode_place_of_death()
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    council_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
 
@@
     week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
+    fill = list(number_of_deaths = 0)
   ) %>% 
@@
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~ref_area) +
+  facet_wrap(~ref_area, scales = "free_y") +
   theme_minimal() +

### 2021-02-03 covid_nrs_place_of_death.Rmd
added crown copyright notice; modifying age/sex data split in progress

@@
 ```
+
+# Copyright attribution
+
+All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)
 
@@
 ~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above!
+Even for the Scotland-wide data, place of death is collapsed to 'All' for the granular sex & age categories, actually.
 
@@
 
-weekly_deaths_age_sex <-
+weekly_deaths_overall <-
   ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
@@
   mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing
+
+weekly_deaths_sex_age_file <-
+  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
+                destfile = tempfile(),
+                quiet = FALSE)
+
+weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
+  readxl::read_excel(weekly_deaths_sex_age_file, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
+  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
+  mutate(
+    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
+    week_number = parse_integer(str_sub(week_number, 4, 5)),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  ) %>%
+  recode_cause_of_death() %>%
+  tidyr::complete(
+    nesting(week_number,week_number_run_over,date_w_c,year),
+    age,sex,place_of_death,cause_of_death,
+    fill = list(number_of_deaths = 0)
+  ) %>%
+  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
+  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
+  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
+  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
 
@@
 ```{r}
-weekly_deaths_age_sex %>% 
+weekly_deaths_overall %>% 
   count(ref_period, date_w_c, week_number, week_number_run_over) %>% 
@@
 
-past_deaths_age_sex_file <-
+past_deaths_sex_age_file <-
   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
@@
 past_deaths_hb <- 
-  readxl::read_excel(past_deaths_health_board_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  readxl::read_excel(past_deaths_hb_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
   clean_names() %>%
@@
 past_deaths_la <-
-  readxl::read_excel(past_deaths_council_area_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  readxl::read_excel(past_deaths_la_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
   clean_names() %>%
@@
 
-past_deaths_age_sex <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(past_deaths_age_sex_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
+past_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
+  readxl::read_excel(past_deaths_sex_age_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
   drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
@@
 map2_dfr(
-  .x = list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
-  .y = c("overall","health_board","council_area","age_sex"),
+  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
+  .y = c("overall","health_board","council_area","sex_age"),
   .f = function(data_tbl, name) {
@@
       mutate(dataset=name,year=as.numeric(year)) %>%
-      filter(!is.na(year) & place_of_death!="All locations") %>%
+      filter(!is.na(year) & place_of_death!="All") %>%
       group_by(dataset,year) %>%
@@
   knitr::kable(caption = "Number of deaths by year as calculated from different datasets.
-All datasets apart from the split by health board agree.
-The differences in the health board dataset likely reflect missing values and/or small numbers redacted from the dataset, which may have to do with the grouping of state hospital/non-NHS institutions.")
+The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other.
+The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other.
+Health Board and Local Authority figures include non-residents.
+Deaths are allocated to areas based on the usual residence of the deceased.
+If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
 ```
@@
 map2_dfr(
-  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_la, past_deaths_age_sex),
-  .y = c("overall","health_board","council_area","age_sex"),
+  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_la, past_deaths_sex_age),
+  .y = c("overall","health_board","council_area","sex_age"),
   .f = function(data_tbl, name) {
@@
 }
-
-# TODO: clean up once these have been dealth with
 
@@
 
-historical_average_age_sex <-
-  past_deaths_age_sex %>%
+historical_average_sex_age <-
+  past_deaths_sex_age %>%
   create_week_53_data_for_years_with_52_weeks() %>%
@@
 historical_average_age <-
-  past_deaths_age_sex %>%
+  past_deaths_sex_age %>%
   create_week_53_data_for_years_with_52_weeks() %>%
@@
 
-historical_average_sex <-
-  past_deaths_age_sex %>%
+historical_aversex_age <-
+  past_deaths_sex_age %>%
   create_week_53_data_for_years_with_52_weeks() %>%
@@
 ```{r}
-weekly_deaths_age_sex %>%
-  filter(sex=="all" & age=="all" & cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
-  select(week_number, year = year, place_of_death, nrs_computed_mean = number_of_deaths) %>%
+weekly_deaths_overall %>%
+  filter(sex=="all" & age=="all") %>%
   left_join(historical_average_overall) %>%
-  filter(round(nrs_computed_mean) != round(deaths_mean)) %>%
-  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file.
+  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean) %>%
+  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
+  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e.
+all other entries had the same figures!).
 Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019.
@@
 merged_data_overall <-
-  weekly_deaths_age_sex %>%
+  weekly_deaths_sex_age %>%
   filter(sex=="all" & age=="all") %>%
-  # covid_deaths$scotland %>%
-  split_deaths_due_to_covid_and_not %>%
-  left_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+  inner_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
             by = c("place_of_death", "week_number", "sex", "age"))
@@
 
-# merged_deaths_by_age_sex <-
+merged_data_sex_age <-
+  weekly_deaths_sex_age %>%
+  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))
+
+# merged_deaths_by_sex_age <-
 #   deaths_by_cause %>%
 #   split_deaths_due_to_covid_and_not %>%
-#   left_join(historical_range_by_age_sex) %>%
+#   left_join(historical_range_by_sex_age) %>%
 #   compute_run_over_week_for_2021

### 2021-02-03 covid_nrs_place_of_death.Rmd
finished data wrangling & modified plots to use new data; plots need work now!

@@
   pivot_wider(names_from=year,values_from=total) %>%
-  knitr::kable(caption = "Number of deaths by year as calculated from different datasets.
+  knitr::kable(caption = "Number of historical deaths by year as calculated from different datasets.
 The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other.
 The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other.
-Health Board and Local Authority figures include non-residents.
+This may be due to the following (verbatim from data description online): Health Board and Local Authority figures include non-residents.
 Deaths are allocated to areas based on the usual residence of the deceased.
@@
 
-historical_aversex_age <-
+historical_average_sex <-
   past_deaths_sex_age %>%
@@
 merged_data_overall <-
-  weekly_deaths_sex_age %>%
+  weekly_deaths_overall %>%
   filter(sex=="all" & age=="all") %>%
@@
 
-# TODO: complete these for age & sex also
 merged_deaths_hb <-
@@
 
-# merged_deaths_by_sex_age <-
-#   deaths_by_cause %>%
-#   split_deaths_due_to_covid_and_not %>%
-#   left_join(historical_range_by_sex_age) %>%
-#   compute_run_over_week_for_2021
-# 
-# merged_deaths_by_age <-
-#   deaths_by_cause %>%
-#   split_deaths_due_to_covid_and_not %>%
-#   left_join(historical_range_by_age) %>%
-#   compute_run_over_week_for_2021
-# 
-# merged_deaths_by_sex <-
-#   deaths_by_cause %>%
-#   split_deaths_due_to_covid_and_not %>%
-#   left_join(historical_range_by_sex) %>%
-#   compute_run_over_week_for_2021
+merged_data_sex <-
+  weekly_deaths_sex_age %>%
+  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
+  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop") %>%
+  inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))
+
+merged_data_age <-
+  weekly_deaths_sex_age %>%
+  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
+  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop") %>%
+  inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))
 ```
@@
 ```{r}
-date_range_text <- glue::glue("Data from w/c {format(min(deaths_by_cause$date_w_c), '%d %B %Y')} to w/c {format(max(deaths_by_cause$date_w_c), '%d %B %Y')} (inclusive).")
+date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")
 
@@
 ## graph format copied from: https://github.com/davidhen/nrs_covid_stats/
-deaths_by_cause %>% 
-  filter(sex == "all" & place_of_death == "all" & cause_of_death == "covid-19-related") %>% 
+weekly_deaths_overall %>% 
+  filter(sex == "all" & place_of_death == "All") %>% 
   filter(age != "all") %>%
   group_by(age) %>% 
-  summarise(n = sum(number_of_deaths)) %>%
+  summarise(n = sum(deaths_all_causes)) %>%
   mutate(pct = round(n/sum(n), 3)) %>% 
@@
   geom_col(fill = "#4477AA") +
-  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 80)) +
+  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 1000)) +
   theme_minimal() +
@@
 
-removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
+removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
 
 annotations_figure_6 <-
-# datapasta::tribble_paste(tbl_recoded_place_of_death %>% select(place_of_death_recoded))
+# datapasta::tribble_paste(merged_data_overall %>% select(place_of_death))
   tibble::tribble(
-    ~place_of_death_recoded, ~x, ~y, ~label,
+    ~place_of_death, ~x, ~y, ~label,
                  "Hospital", 20L, 850, "Shaded area shows range of 2015-2019 deaths; dashed line shows weekly average",
@@
                 "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
-                     "Home", 35L, 500, "The gap between historical high and both types of death at home was high throughout."
+                     "Home & other non-institution", 35L, 500, "The gap between historical high and both types of death at home was high throughout."
                   ) %>% 
   mutate(
-    place_of_death_recoded = factor(place_of_death_recoded, levels = levels(tbl_recoded_place_of_death$place_of_death_recoded)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
+    place_of_death = factor(place_of_death, levels = levels(merged_data_overall$place_of_death)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
     label = str_wrap(label, width = 40)
@@
 
-# figure_6_location_of_death <-
+figure_6_location_of_death <-
   merged_data_overall %>%
   filter(sex == "all" & age == "all") %>%
-  filter(!place_of_death_recoded %in% c("All","Other")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
+  filter(!place_of_death %in% c("All","Other")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
                names_to = "deaths_type", values_to = "deaths") %>% 
@@
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_grid(~place_of_death_recoded) +
+  facet_grid(~place_of_death) +
   theme_minimal() +
@@
 
-map(
+walk(
   .x = c(".pdf",".svg",".png"),
@@
 
-## Figure 7: deaths at home
+## Figure 7: deaths at home, all of Scotland
 
-```{r}
-# TODO
+```{r, fig.width=12, fig.height=9}
+# TODO: add start of lockdowns 1-3?
+# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
+figure_7_deaths_at_home <-
+  merged_data_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  # geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_grid(~place_of_death) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
 
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_7_deaths_at_home",.x), plot = figure_7_deaths_at_home, width = 12, height = 9, dpi = 300)
+)
+
+figure_7_deaths_at_home
 ```
@@
   filter(sex != "all" & age == "all") %>%
-  filter(place_of_death_recoded == "Home & other non-institution") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 

### 2021-02-03 covid_nrs_place_of_death.Rmd
fixed historical computations for age & sex 7

@@
 All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)
+
+# Context of lockdowns
+
+A recent timeline of COVID-19 events in Scotland is [available here](https://spice-spotlight.scot/2021/01/29/timeline-of-coronavirus-covid-19-in-scotland/).
+Including these might help in the interpreration fo the figures, though the figures are already pretty busy.
+This is a potential future TODO!
 
@@
 
+On 3 Februrary 2021 as I was updating the script, the weekly figures started distinguishing cause of death: COVID-19 can be a contributory factor, or an underlying cause of death.
+For the purposes of this study, COVID-19 deaths are in the minority, and we will disregard this distinction for now.
+
 ```{r}
@@
         cause_of_death == "COVID-19 mentioned" ~ "deaths_covid_related",
+        cause_of_death == "COVID-19 contributory factor" ~ "deaths_covid_related",
+        cause_of_death == "COVID-19 underlying cause" ~ "deaths_covid_related",
         cause_of_death == "Non-COVID-19" ~ "deaths_non_covid",
@@
 
+# TODO:  still too complicated to safely use as a function...
 ## find which combinations of geographic area, cause of death, and place of death are missing and set the number of deaths to 0 for them
-complete_weekly_deaths_with_implicit_zeros <- function(data_tbl) {
+# complete_weekly_deaths_with_implicit_zeros <- function(data_tbl) {
+#   data_tbl %>%
+#   tidyr::complete(
+#       nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
+#       ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
+#       fill = list(number_of_deaths = 0)
+#       )
+# }
+
+## calculate number of deaths after recoding cause of death -
+## recoding results in two entries for covid-related deaths, so we need to sum over them
+## this is achieved by grouping over all variables other than number of deaths
+recalculate_number_of_deaths_after_recoding_cause_of_death <- function(data_tbl) {
   data_tbl %>%
-  tidyr::complete(
-      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
-      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
-      fill = list(number_of_deaths = 0)
-      )
+  group_by_all %>% 
+  ungroup(number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")  # recompute the sum of deaths
 }
@@
   recode_cause_of_death() %>%
+  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # not needed here probably, but no harm in including it
   pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
@@
   recode_cause_of_death() %>%
-  tidyr::complete(
+  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
+  tidyr::complete(  # complete implicit missing entries and make them 0
     nesting(week_number,week_number_run_over,date_w_c,year),
@@
   recode_cause_of_death() %>%
-  complete_weekly_deaths_with_implicit_zeros() %>%
+  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
+  tidyr::complete(
+      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
+      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
+      fill = list(number_of_deaths = 0)
+      ) %>%
+  # complete_weekly_deaths_with_implicit_zeros() %>%
   recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
@@
   create_week_53_data_for_years_with_52_weeks() %>%
+  group_by_all %>%
+  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop") %>%  # recompute the sum of deaths over sexes
   group_by(
@@
   create_week_53_data_for_years_with_52_weeks() %>%
+  group_by_all %>%
+  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & age
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop") %>%  # recompute the sum of deaths over ages
   group_by(
@@
   weekly_deaths_sex_age %>%
+  filter(age!="all") %>%
   group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
@@
   weekly_deaths_sex_age %>%
+  filter(sex!="all") %>%
   group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
@@
        )
+```
 
+```{r}
 merged_deaths_la %>%
@@
 ```{r}
-merged_deaths_by_sex %>%
-  filter(sex != "all" & age == "all") %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
+# TODO: check the sex-disaggregated data, the historical figures look wrong
+merged_data_sex %>%
+  # filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
                names_to = "deaths_type", values_to = "deaths") %>% 
@@
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~sex) +
+  facet_grid(~sex) +
   theme_minimal() +
@@
        subtitle = date_range_text,
-       # caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
        colour = "Cause of death"

### 2021-02-03 covid_nrs_place_of_death.Rmd
fixed up lots of plots, could do with more details!

@@
 
-## Figure 6: Recorded deaths in Scotland, by week number & location
+## Recorded deaths in Scotland, by week number & location
 
@@
 
-annotations_figure_6 <-
+annotations_figure_place_of_death <-
 # datapasta::tribble_paste(merged_data_overall %>% select(place_of_death))
@@
 
-figure_6_location_of_death <-
+figure_place_of_death <-
   merged_data_overall %>%
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_6_location_of_death",.x), plot = figure_6_location_of_death, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0("./outputs/figure_place_of_death",.x), plot = figure_place_of_death, width = 12, height = 9, dpi = 300)
 )
 
-figure_6_location_of_death
+figure_place_of_death
 ```
 
-## Figure 7: deaths at home, all of Scotland
+## Deaths at home, all of Scotland
 
@@
 # TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
-figure_7_deaths_at_home <-
+figure_deaths_at_home <-
   merged_data_overall %>%
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_7_deaths_at_home",.x), plot = figure_7_deaths_at_home, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 12, height = 9, dpi = 300)
 )
 
-figure_7_deaths_at_home
+figure_deaths_at_home
 ```
 
-## Figure 8: deaths at home, by health board
+## Deaths at home, by health board
 
-```{r}
-# TODO: fix this once we have health board in weekly deaths!
+```{r, fig.width=12, fig.height=9}
+figure_deaths_at_home_by_health_board <-
   merged_deaths_hb %>%
-  # filter(sex == "all" & age == "all") %>%
   filter(place_of_death == "Home & other non-institution") %>%
@@
   geom_vline(xintercept = 54, linetype = "dotted") +
-  # geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
   scale_x_continuous(breaks = seq(0,60,by=10)) +
@@
        )
+
+figure_deaths_at_home_by_health_board
 ```
 
-```{r}
+## Deaths at home, by local authority
+
+```{r, fig.width=12, fig.height=9}
+figure_deaths_at_home_by_local_authority <-
   merged_deaths_la %>%
-  # filter(sex == "all" & age == "all") %>%
   filter(place_of_death == "Home & other non-institution") %>%
@@
 
+figure_deaths_at_home_by_local_authority
 ```
 
-## Figure 10: deaths at home by sex
+## Deaths at home by age
 
-```{r}
-# TODO: check the sex-disaggregated data, the historical figures look wrong
+```{r, fig.width=12, fig.height=9}
+removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_data_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_age %>% filter(year==2021) %>% .$week_number)} in 2021.")
+
+figure_deaths_at_home_by_age <-
+  merged_data_age %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  filter(!age %in% c('0','1-14')) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  # facet_wrap(~age, scales = "free_y") +
+  facet_wrap(~age) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(vertical_line_text,removed_under_15_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_age",.x), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300)
+)
+
+figure_deaths_at_home_by_age
+```
+
+## Deaths at home by sex
+
+```{r, fig.width=12, fig.height=9}
+figure_deaths_at_home_by_sex <-
   merged_data_sex %>%
-  # filter(sex == "all" & age == "all") %>%
+  mutate(sex = if_else(sex=="F", "Female", "Male")) %>%
   filter(place_of_death %in% c("Home & other non-institution")) %>%
@@
        )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_sex",.x), plot = figure_deaths_at_home_by_sex, width = 12, height = 9, dpi = 300)
+)
+
+figure_deaths_at_home_by_sex
 ```

### 2021-02-09 order_of_min_max_rolling_mean.R
working out how to compute rolling mean/moving average over historical figures & ranges

@@ -1,0 +1,47 @@
+## Script showing difference in how the order of min/max and rolling mean (moving average) operations produces different results
+## To show the variation in historical weekly figures, we show min & max values over a period of several years for each week
+## To produce a rolling mean of these values there are several options for the order in which these are computed:
+## A) rolling mean over weekly min/max values (the min/max having been computed over the period of years for each week)
+## B) Min/max over rolling average over weeks within each year - rolling mean is computed for each year separately, and then a min/max is computed over the years
+
+library(tidyverse)
+library(zoo)  # for rolling mean function
+
+set.seed(42)
+test_data <- tibble(year = rep(2015:2019, each = 4), week = rep(1:4, times = 5), n = as.integer(runif(n=20, min = 0, max = 100)))
+
+approach_A <-
+  test_data %>%
+  group_by(week) %>%
+  summarise(
+    weekly_mean = mean(n),
+    weekly_min = min(n), 
+    weekly_max = max(n),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    rolling_mean_A = zoo::rollmean(x=weekly_mean, k=2, fill=NA, align="right"),
+    rolling_mean_min_A = zoo::rollmean(x=weekly_min, k=2, fill=NA, align="right"),
+    rolling_mean_max_A = zoo::rollmean(x=weekly_max, k=2, fill=NA, align="right")
+  )
+
+approach_B <-
+  test_data %>%
+  group_by(year) %>%
+  mutate(
+    yearly_rolling_mean = zoo::rollmean(x=n, k=2, fill=NA, align="right")
+  ) %>%
+  ungroup %>%
+  group_by(week) %>%
+  summarise(
+    rolling_mean_B = mean(yearly_rolling_mean),
+    rolling_mean_min_B = min(yearly_rolling_mean), 
+    rolling_mean_max_B = max(yearly_rolling_mean),
+    .groups = "drop"
+  )
+
+approach_A
+approach_B
+## Note how the rolling mean is identical between the two, but approach A produces a wider range
+## This is because the order in approach B is mean -> range
+## Whereas in approach A it is range -> mean

### 2021-02-09 covid_nrs_place_of_death.Rmd
Trying to incorporate rolling mean & map data...

@@
 library(ISOweek)  # for computing date from ISO week number + year
+library(sf)  # for mapping
+library(maps)
 ```
@@
 
-## Consistent pPlace of death coding
+## Consistent Place of death coding
 
@@
 
+## Merging health boards
+
+Smaller health boards were merged... TODO!
+
+> "Note: Orkney, Shetland, & Western health boards were merged into 'Island HBs'; Borders & Dumfries and Galloway health boards were merged into 'B,D&G HB'"
+
 ## Load weekly COVID-19 related deaths
@@
 
+## Load map data with `sf`
+
+To get council areas in Scotland we need NUTS 2 level map data.q
+
+```{r}
+# TODO: learn how to use maps etc.
+zip_file <- tempfile()
+unzipped <- tempfile()
+
+## Scottish LAU 2011 level 1
+# curl_download(url = "https://borders.ukdataservice.ac.uk/ukborders/easy_download/prebuilt/shape/Scotland_laulevel1_2011.zip", destfile = zip_file, quiet = FALSE)  # download file from ONS
+# unzip(zipfile = zip_file, exdir = unzipped)
+# name <- list.files(unzipped, pattern=".shp")
+# shapefile <- st_read(file.path(unzipped, name))
+
+## below fails, saying cannot allocate vector of size XX MB/GB
+# curl_download(url = "https://opendata.arcgis.com/datasets/7f83b82ef6ce46d3a5635d371e8a3e7c_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D", destfile = zip_file, quiet = FALSE)  # download file from ONS
+
+# merged_deaths_la %>%
+#   left_join(shapefile,by=c("ref_area"="name")) %>%
+#   ggplot(aes(fill = deaths_mean, geometry = geometry)) +
+#   geom_sf()
+#
+
+# ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
+#   geom_polygon() +
+#   coord_map()
+
+```
+
+# Calculate data averaged over multiple weeks
+
+Once data are broken down by region there are very small figures on a weekly basis, so we can compute a moving average over several weeks instead.
+This is different to merging figures into 2-week chunks, for example, as information from consecutive weeks will overlap.
+
+The average will be computed to the "left" of the current value (a week's moving average is the average of values at this week + n-1 past weeks)
+
+```{r}
+weeks_to_average <- 2
+
+# TODO: implement rolling averages also - I think this needs to be applied at source tbls first (i.e. max of sum over two weeks rather than mean of maxes?)
+  # group_by(ref_area, sex, age, place_of_death) %>% 
+  # mutate(across(.cols = matches("deaths_"), .fns = ~zoo::rollmean(x = .x, k = 2, align = "right", fill = NA)))
+
+# weekly_deaths_sex_age
+# weekly_deaths_hb %>%
+#   group_by(year, week_number, )
+```
+
 # Calculate historical averages, minima & maxima
@@
 ```{r}
+## compute rolling average
+
 merged_data_overall <-
@@
 source_nrs_text <- "Source: National Records of Scotland"
+
+vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."
 ```
@@
 
-```{r, fig.width=12, fig.height=9}
+```{r, fig.width=12, fig.height=9, warning=FALSE}
 ## graph format copied from: https://github.com/davidhen/nrs_covid_stats/
@@
 
-```{r, fig.width=12, fig.height=9}
-vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."
-
+```{r, fig.width=12, fig.height=9, warning=FALSE}
 removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
@@
   geom_vline(xintercept = 54, linetype = "dotted") +
-  geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
+  geom_text(data = annotations_figure_place_of_death, aes(x=x,y=y, label=label)) +
   scale_x_continuous(breaks = seq(0,60,by=10)) +
@@
 
-```{r, fig.width=12, fig.height=9}
+```{r, fig.width=12, fig.height=9, warning=FALSE}
 # TODO: add start of lockdowns 1-3?
 # TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
-figure_deaths_at_home <-
+# figure_deaths_at_home <-
   merged_data_overall %>%
@@
 
-```{r, fig.width=12, fig.height=9}
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+island_health_boards_merged <- "Note: Orkney, Shetland, & Western health boards were merged into 'Island HBs'; Borders & Dumfries and Galloway health boards were merged into 'B,D&G HB'"
+vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"
+
 figure_deaths_at_home_by_health_board <-
@@
        subtitle = date_range_text,
-       # caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
        colour = "Cause of death"
@@
 
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_health_board",.x), plot = figure_deaths_at_home_by_health_board, width = 12, height = 9, dpi = 300)
+)
+
 figure_deaths_at_home_by_health_board
@@
 
-```{r, fig.width=12, fig.height=9}
+```{r, fig.width=12, fig.height=9, warning=FALSE}
 figure_deaths_at_home_by_local_authority <-
@@
 
+merged_deaths_la %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_local_authority",.x), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300)
+)
+
 figure_deaths_at_home_by_local_authority
@@
 
-```{r, fig.width=12, fig.height=9}
+```{r, fig.width=12, fig.height=9, warning=FALSE}
 removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_data_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_age %>% filter(year==2021) %>% .$week_number)} in 2021.")
@@
 
-```{r, fig.width=12, fig.height=9}
+```{r, fig.width=12, fig.height=9, warning=FALSE}
 figure_deaths_at_home_by_sex <-

### 2021-02-09 covid_nrs_place_of_death.Rmd
finally figured out moving averages - my approach now allows to specify how many weeks to average over, now the final test - plotting!

@@
   set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
+
+## compute weekly data for sex and age separately
+## sum over sex or age, respectively
+weekly_deaths_sex <-
+  weekly_deaths_sex_age %>%
+  filter(age!="all") %>%
+  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
+  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
+
+weekly_deaths_age <-
+  weekly_deaths_sex_age %>%
+  filter(sex!="all") %>%
+  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
+  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
 ```
@@
   drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
-  rename(
-    week_number = week_of_occurrence,
-    place_of_death = location
+  rename(  # same format as current data
+    week_number = matches("week"),
+    place_of_death = matches("location"),
+    ref_area = health_board
   ) %>%
@@
     nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    health_board, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
+    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
     fill = list(number_of_deaths = 0)
@@
   drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
-  rename(
-    week_number = week,
-    place_of_death = location_of_death
+  rename(  # same format as current data
+    week_number = matches("week"),
+    place_of_death = matches("location"),
+    ref_area = council_area
   ) %>%
@@
     nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    council_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
+    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
     fill = list(number_of_deaths = 0)
@@
 
+## compute weekly data for sex and age separately
+## sum over sex or age, respectively
+past_deaths_sex <-
+  past_deaths_sex_age %>%
+  group_by_all %>%
+  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
+
+past_deaths_age <-
+  past_deaths_sex_age %>%
+  group_by_all %>%
+  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
+```
+
+### Do we get the same number of deaths between datasets?
+
+```{r}
 ## check that the results are the same as overall deaths
@@
 
-```
-
-# Calculate data averaged over multiple weeks
-
-Once data are broken down by region there are very small figures on a weekly basis, so we can compute a moving average over several weeks instead.
-This is different to merging figures into 2-week chunks, for example, as information from consecutive weeks will overlap.
-
-The average will be computed to the "left" of the current value (a week's moving average is the average of values at this week + n-1 past weeks)
-
-```{r}
-weeks_to_average <- 2
-
-# TODO: implement rolling averages also - I think this needs to be applied at source tbls first (i.e. max of sum over two weeks rather than mean of maxes?)
-  # group_by(ref_area, sex, age, place_of_death) %>% 
-  # mutate(across(.cols = matches("deaths_"), .fns = ~zoo::rollmean(x = .x, k = 2, align = "right", fill = NA)))
-
-# weekly_deaths_sex_age
-# weekly_deaths_hb %>%
-#   group_by(year, week_number, )
 ```
@@
   group_by(
-    week_number, place_of_death, health_board
+    week_number, place_of_death, ref_area
   ) %>%
@@
   group_by(
-    week_number, place_of_death, council_area
+    week_number, place_of_death, ref_area
   ) %>%
@@
 historical_average_age <-
-  past_deaths_sex_age %>%
+  past_deaths_age %>%
   create_week_53_data_for_years_with_52_weeks() %>%
-  group_by_all %>%
-  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop") %>%  # recompute the sum of deaths over sexes
   group_by(
@@
 historical_average_sex <-
-  past_deaths_sex_age %>%
+  past_deaths_sex %>%
   create_week_53_data_for_years_with_52_weeks() %>%
-  group_by_all %>%
-  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & age
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop") %>%  # recompute the sum of deaths over ages
   group_by(
@@
 
+# Calculate data averaged over multiple weeks
+
+Once data are broken down by region there are very small figures on a weekly basis, so we can compute a moving average over several weeks instead.
+This is different to merging figures into 2-week chunks, for example, as information from consecutive weeks will overlap.
+
+The average will be computed to the "left" of the current value (a week's moving average is the average of values at this week + n-1 past weeks, or "right" aligned).
+
+Moving averages for historical data will be computed over the previously computed weekly means & ranges.
+Computing moving averages over historical ranges produces wider spread data than computing ranges of moving averages (see `order_of_min_max_rolling_mean.R`).
+
+```{r}
+weeks_to_average <- 2
+
+## compute moving average over weekly data, using average over k weeks
+## using historical data to compute the moving average in the first few weeks of current data
+## Note: historical data needs to be the non-summarised (raw) data!
+## to compute moving average in first few week(s) of year, we add in the previous years' data and code the weeks so that they retain the right order
+## i.e. in 2020 the week_number_run_over starts at 1, and the last week of 2019 is coded as 0, the 2nd last as -1, etc.
+## once the moving average has been computed, we remove the past data;
+## this approach works for 2021 data seamlessly, using the last weeks of 2020 for the moving average in 2021!
+compute_moving_average_for_current_data <- function(current_data, historical_data) {
+  bind_rows(
+    historical_data %>%  # add historical data and format it so that it matches current data
+      filter(year == 2019) %>%  # keep only last year before 2020
+      rename(deaths_all_causes = number_of_deaths) %>%  # in historical deaths there is no split between covid/non-covid so we rename to all cause deaths
+      mutate(week_number_run_over = -(52-week_number)),
+    current_data
+  ) %>%
+    arrange(week_number_run_over) %>%  # just in case the order got messed up, re-order by week_number_run_over for the computation
+    group_by_all() %>% ungroup(matches("^deaths_|week|date|year")) %>%  # clever way of getting grouping by ref_area, sex, age, whatever is available
+    mutate(across(
+        .cols = matches("^deaths_"), 
+        .fns = ~zoo::rollmean(.x, k=weeks_to_average, fill=NA, align="right"), 
+        .names = "ma{weeks_to_average}w_{.col}"
+        )
+    ) %>%
+    ungroup %>%
+    filter(year > 2019)  # keep only current data
+}
+
+rolling_deaths_hb <-
+  compute_moving_average_for_current_data(current_data = weekly_deaths_hb, historical_data = past_deaths_hb)
+
+rolling_deaths_la <-
+  compute_moving_average_for_current_data(current_data = weekly_deaths_la, historical_data = past_deaths_la)
+
+rolling_deaths_sex_age <-
+  compute_moving_average_for_current_data(current_data = weekly_deaths_sex_age, historical_data = past_deaths_sex_age)
+
+rolling_deaths_age <-
+  compute_moving_average_for_current_data(current_data = weekly_deaths_age, historical_data = past_deaths_age)
+
+rolling_deaths_sex <-
+  compute_moving_average_for_current_data(current_data = weekly_deaths_sex, historical_data = past_deaths_sex)
+
+## inspect the calculations  
+rolling_deaths_la %>% filter(week_number_run_over %in% c(1:3,52:54) & str_detect(ref_area,"Glasgow") & place_of_death=="Hospital") %>% select(year, week_number_run_over, ref_area, place_of_death, deaths_all_causes, ma2w_deaths_all_causes) %>% knitr::kable(caption="Showing how rolling average was computed: week 1 is average of week 52 of 2019 (not shown) and week 1 of 2020; week 2 is average of weeks 1 & 2, etc.")
+
+rolling_deaths_sex_age %>% filter(week_number_run_over %in% c(1:3,52:54) & str_detect(sex,"F") & age == "65-74" & place_of_death=="Hospital") %>% select(year, week_number_run_over, sex, age, place_of_death, deaths_all_causes, ma2w_deaths_all_causes) %>% knitr::kable(caption = "Showing how rolling average was computed: week 1 is average of week 52 of 2019 (not shown) and week 1 of 2020; week 2 is average of weeks 1 & 2, etc.")
+
+## helper function that takes historical data and computes a moving average over k weeks
+## with the additional trick of using the last week of all the years as the week prior to the first week
+## strictly speaking we should be using the last week(s) of 2014 to compute the moving average for the first week(s) in 2015, depending on k; and we shouldn't be using the last week of 2019 in the computation either
+## however, the overall effect of this is small since we're averaging over the entire 2015-2019 period
+## the inaccuracy of this approach increases with increasing k
+## the solution for fixing this is straightforward - just download 2014 data as well, and merge the "current" data and the historical data, and then perform the moving average calculation as usual
+compute_moving_average_for_historical_data <- function(historical_data) {
+  last_week_in_previous_year_required <- 53 + 2 - weeks_to_average  # this computes the last week needed from the previous year: if averaging over 2 weeks, we need week 3, if over 3 weeks, the lat week needed is 52, etc.
+  bind_rows(
+    historical_data %>%  # keep last week from all years and convert it to week 0
+      filter(week_number %in% last_week_in_previous_year_required:53) %>% 
+      mutate(week_number = week_number - 53),  # set week number for data in previous to negative numbers but retaining order 
+    historical_data
+  ) %>%
+  arrange(week_number) %>%  # sort in ascending order of week_number
+  group_by_all() %>% ungroup(matches("^deaths_"),week_number) %>%
+  mutate(across(
+      .cols = matches("^deaths_"), 
+      .fns = ~zoo::rollmean(.x, k=weeks_to_average, fill=NA, align="right"), 
+      .names = "ma{weeks_to_average}w_{.col}"
+      )
+  ) %>%
+  ungroup %>%
+  filter(week_number!=0)
+}
+
+historical_average_hb <- compute_moving_average_for_historical_data(historical_average_hb)
+historical_average_la <- compute_moving_average_for_historical_data(historical_average_la)
+historical_average_sex_age <- compute_moving_average_for_historical_data(historical_average_sex_age)
+historical_average_sex <- compute_moving_average_for_historical_data(historical_average_sex)
+historical_average_age <- compute_moving_average_for_historical_data(historical_average_age)
+```
+
 # Merge weekly COVID/non-COVID-related deaths in 2020/2021 with historic death rates
@@
 merged_deaths_hb <-
-  weekly_deaths_hb %>%
-  inner_join(historical_average_hb, by = c("week_number","ref_area"="health_board","place_of_death"))
+  # weekly_deaths_hb %>%
+  rolling_deaths_hb %>%  # use data with rolling movingincluded
+  inner_join(historical_average_hb, by = c("week_number","ref_area","place_of_death"))
 
 merged_deaths_la <-
-  weekly_deaths_la %>%
-  inner_join(historical_average_la, by = c("week_number","ref_area"="council_area","place_of_death"))
+  # weekly_deaths_la %>%
+  rolling_deaths_la %>%  # use data with rolling movingincluded
+  inner_join(historical_average_la, by = c("week_number","ref_area","place_of_death"))
 
 merged_data_sex_age <-
-  weekly_deaths_sex_age %>%
+  # weekly_deaths_sex_age %>%
+  rolling_deaths_sex_age %>%  # use data with rolling movingincluded
   inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))
@@
 merged_data_sex <-
-  weekly_deaths_sex_age %>%
-  filter(age!="all") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
-  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop") %>%
+  rolling_deaths_sex %>%
   inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))
@@
 merged_data_age <-
-  weekly_deaths_sex_age %>%
-  filter(sex!="all") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
-  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop") %>%
+  rolling_deaths_age %>%
   inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))

### 2021-02-09 covid_nrs_place_of_death.Rmd
added HB and LA data using 2week moving average; started an overall yearly look at data but need to include yearly range from historical data, plus a split of coivd-as underlying cause and one of factors in death

@@
 #   coord_map()
-
 ```
@@
 # TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
-# figure_deaths_at_home <-
+figure_deaths_at_home <-
   merged_data_overall %>%
@@
 figure_deaths_at_home_by_health_board
+```
+
+## Deaths at home with 2-week moving average, by health board
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+moving_average_text <- "All figures represent deaths averaged over past 2 weeks; this includes historical range and historical mean."
+
+figure_deaths_at_home_2week_average_by_health_board <-
+  merged_deaths_hb %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid), 
+               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = ma2w_deaths, colour = deaths_type)) +
+  geom_line(aes(y = ma2w_deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~ref_area, scales = "free_y") +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(moving_average_text, vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2week_average_by_health_board",.x), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300)
+)
+
+figure_deaths_at_home_2week_average_by_health_board
 ```
@@
 
+# merged_deaths_la %>%
+#   filter(place_of_death == "Home & other non-institution") %>%
+#   filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
+#   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+#                names_to = "deaths_type", values_to = "deaths") %>% 
+#   ggplot(aes(x = week_number_run_over)) +
+#   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+#   geom_line(aes(y = deaths, colour = deaths_type)) +
+#   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+#   geom_vline(xintercept = 54, linetype = "dotted") +
+#   scale_x_continuous(breaks = seq(0,60,by=10)) +
+#   facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
+#   theme_minimal() +
+#   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+#                       labels = c("All deaths", 
+#                                  "Non Covid-19 deaths"),
+#                       guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+#                       ) +
+#   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+#   labs(x = "Week number",
+#        y = "Deaths (N)",
+#        subtitle = date_range_text,
+#        caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
+#        colour = "Cause of death"
+#        )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_local_authority",.x), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300)
+)
+
+figure_deaths_at_home_by_local_authority
+```
+
+## Deaths at home with 2-week moving average, by local authority
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+figure_deaths_at_home_2week_average_by_local_authority <-
   merged_deaths_la %>%
   filter(place_of_death == "Home & other non-institution") %>%
-  filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
+  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid), 
+               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
   ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = ma2w_deaths, colour = deaths_type)) +
+  geom_line(aes(y = ma2w_deaths_mean), linetype = "dashed", size = 0.2) +
   geom_vline(xintercept = 54, linetype = "dotted") +
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
+  facet_wrap(~ref_area, scales = "free_y") +
   theme_minimal() +
@@
        subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(moving_average_text, vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
        colour = "Cause of death"
        )
+# 
+# merged_deaths_la %>%
+#   filter(place_of_death == "Home & other non-institution") %>%
+#   filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
+#   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+#                names_to = "deaths_type", values_to = "deaths") %>% 
+#   ggplot(aes(x = week_number_run_over)) +
+#   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+#   geom_line(aes(y = deaths, colour = deaths_type)) +
+#   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+#   geom_vline(xintercept = 54, linetype = "dotted") +
+#   scale_x_continuous(breaks = seq(0,60,by=10)) +
+#   facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
+#   theme_minimal() +
+#   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+#                       labels = c("All deaths", 
+#                                  "Non Covid-19 deaths"),
+#                       guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+#                       ) +
+#   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+#   labs(x = "Week number",
+#        y = "Deaths (N)",
+#        subtitle = date_range_text,
+#        caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
+#        colour = "Cause of death"
+#        )
 
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_local_authority",.x), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300)
-)
-
-figure_deaths_at_home_by_local_authority
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2week_average_by_local_authority",.x), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300)
+)
+
+figure_deaths_at_home_2week_average_by_local_authority
 ```
@@
 ```
+
+## Overall deaths at home 2020 by cause of death
+
+```{r}
+# TODO: add yearly range for all deaths
+# TODO: split data by covid-underlying cause of desath & covid-factor in death
+merged_data_overall %>%
+  filter(sex=="all", age=="all") %>%
+  filter(year==2020) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  group_by(ref_area, year) %>%
+  summarise(
+    deaths_all_causes = sum(deaths_all_causes, na.rm=TRUE),
+    deaths_covid_related = sum(deaths_covid_related, na.rm=TRUE),
+    deaths_non_covid = sum(deaths_non_covid, na.rm=TRUE)
+  )
+```

### 2021-02-09 covid_nrs_place_of_death.Rmd
added HB and LA data using 2week moving average; started an overall yearly look at data but need to include yearly range from historical data, plus a split of coivd-as underlying cause and one of factors in death

@@
 # TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
-figure_deaths_at_home <-
+# figure_deaths_at_home <-
   merged_data_overall %>%

### 2021-02-12 covid_nrs_place_of_death.Rmd
figured out how to keep track of data being updated - there's now a separate script that downloads file and keeps track of latest
now I need to modify the .Rmd to read the downloaded data instead!

--- 2021-02-09 covid_nrs_place_of_death.Rmd
@@
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
-output: html_document
+output: 
+  html_document:
+    code_folding: hide
+    toc: true
+    toc_float: true
 editor_options: 
   chunk_output_type: console
+params:
+  data_sources: '`r tibble()`'
 ---
@@
 All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)
+
+# Dates represent event registration, not occurrence!
+
+From NRS website:
+
+> All routine vital events information we publish is based on the date of registration, not the date on which the event occurred.
+Inevitably, of course, there are delays between the occurrence of an event and its registration.
 
@@
 ```{r}
+download_dir <- "./downloaded_data"
+if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist
+
 # checking out structure
@@
 
-To get council areas in Scotland we need NUTS 2 level map data.q
+TODO: figure out which boundaries in Scotland correspond to council areas!
+
+~~To get council areas in Scotland we need NUTS 2 level map data~~, and the most appropriate level of detail to use is likely the BUC - Ultra Generalised Clipped
+
+https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom
+
+https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom?geometry=-12.122%2C55.353%2C4.566%2C57.479
+
+health board boundaries April 2019, BUC
+https://opendata.arcgis.com/datasets/def40bdc98a9457aa108eb3a5fb052b1_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D
+
+getting error trying to use shapefile with ggplot:
+> Error in CPL_transform(x, crs, aoi, pipeline, reverse) : 
+  crs not found: is it missing?
+
+explicitly setting crs on shapefile fails,
+
+version fo sf 0.9-7:
+> Linking to GEOS 3.8.0, GDAL 3.0.4, PROJ 6.3.1
 
@@
 # curl_download(url = "https://borders.ukdataservice.ac.uk/ukborders/easy_download/prebuilt/shape/Scotland_laulevel1_2011.zip", destfile = zip_file, quiet = FALSE)  # download file from ONS
-# unzip(zipfile = zip_file, exdir = unzipped)
 # name <- list.files(unzipped, pattern=".shp")
@@
 
-# merged_deaths_la %>%
-#   left_join(shapefile,by=c("ref_area"="name")) %>%
+download_and_import_shapefile <- function(url) {
+  zip_file <- tempfile()
+  unzipped <- tempfile()
+  downloaded_zip <- curl_download(url = url, destfile = zip_file, quiet = FALSE) 
+  unzip(zipfile = zip_file, exdir = unzipped)
+  name <- list.files(unzipped, pattern=".shp$")
+  shapefile <- st_read(file.path(unzipped, name))
+  return(shapefile)
+}
+
+health_boards_shapefile_url <- "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip"
+la_shapefile_url <- "https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"
+
+shapefile_hb <- download_and_import_shapefile(url = health_boards_shapefile_url)
+shapefile_la <- download_and_import_shapefile(url = la_shapefile_url)
+
+# shapefile %>%
+#   ggplot(aes(geometry = geometry)) +
+#   geom_sf()
+
+# shapefile_la %>%
+#   inner_join(merged_deaths_la, by = c("lad19nm"="ref_area"))
+
+# shapefile_hb %>%
+#   inner_join(merged_deaths_hb, by = c("HBName"="ref_area"))
+
+# TODO: summarise data by LA/HB so it can be shown on a map
+# otherwise it loads forever because we're re-plotting the same regional boundary for each row!!
+
+merged_deaths_la %>%
+  filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
+  left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
+  ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
+  geom_sf()
+
+# ggplot(data = shapefile_hb, aes(geometry=geometry)) + geom_sf()
+
+# merged_deaths_hb %>%
+#   left_join(shapefile_hb,by=c("ref_area"="HBName")) %>%
 #   ggplot(aes(fill = deaths_mean, geometry = geometry)) +
 #   geom_sf()
-#
 
@@
 source_nrs_text <- "Source: National Records of Scotland"
+
+death_registrations_text <- "Note: all figures are based on date of registration."
 
@@
        subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
        colour = "Cause of death"
@@
 figure_deaths_at_home
+```
+
+## Deaths at home in 2020, all of Scotland
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+figure_deaths_at_home_2020 <-
+  merged_data_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  filter(year == 2020) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_grid(~place_of_death) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       # subtitle = date_range_text,
+       subtitle = "Data from 2020 shown, starting with w/c 30 Dec 2019.",
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2020",.x), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300)
+)
+
+figure_deaths_at_home_2020
 ```

### 2021-02-12 download_latest_data_and_render_reports.R

@@ -1,0 +1,87 @@
+
+# Overview ----------------------------------------------------------------
+
+## This function loads a table containing the filenames & urls of 
+## data sources, and then checks if newer data is available, downloading it in
+## the process. Once that's done, it will render one or more .Rmd reports using
+## the latest data.
+
+
+# Load packages -----------------------------------------------------------
+
+library(tidyverse)
+library(curl)
+library(knitr)
+library(rmarkdown)
+
+# Load table keeping track of latest data ---------------------------------
+
+file_data_sources <- "./table_of_data_sources.csv"
+download_dir <- "./downloaded_data"
+if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist
+
+if (file.exists(file_data_sources)) {
+  table_of_data_sources <- read_csv(file = "./table_of_data_sources.csv")
+} else {
+  table_of_data_sources <- tibble(
+    short_name = c("sex_age","la","hb"),
+    url = c(
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx"
+    )
+  ) %>% mutate(
+    filename = basename(url),
+    last_modified = ymd_hms("2020-01-01_01:01:01"),  # start with a date in the past
+    path_latest = NA_character_  
+  )
+}
+
+# Function to download latest data if available ---------------------------
+
+download_file_if_newer_available <- function(url, filename, last_modified, directory) {
+  todays_file <- curl_fetch_disk(url, path = tempfile())
+  if (last_modified < todays_file$modified) {
+    date_string <- strftime(todays_file$modified, format = "%F")  # prefix filename with last modified date
+    ## copy file to folder
+    filename_prefixed <- paste0(date_string,"_",filename)
+    file_path <- file.path(directory, filename_prefixed)
+    file.copy(from = todays_file$content, to = file_path, overwrite = TRUE)
+    ## set last_modified date & save
+    table_of_data_sources$last_modified[table_of_data_sources$filename == filename] <<- todays_file$modified
+    table_of_data_sources$path_latest[table_of_data_sources$filename == filename] <<- file_path
+    return(TRUE)
[…]

### 2021-02-12 download_latest_data_and_render_reports.R
added historical data to checking & downloading latest file pipeline. need to modify the .rmd to actually use the latest data instead of downloading now!

@@ -17,11 +17,16 @@ library(rmarkdown)
 # Load table keeping track of latest data ---------------------------------
 
 file_data_sources <- "./table_of_data_sources.csv"
+file_historical_data_sources <- "./table_of_historical_data_sources.csv"
+
 download_dir <- "./downloaded_data"
 if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist
 
+historical_dir <- "./historical_data"
+if (!dir.exists(historical_dir)) dir.create(historical_dir)  # create download dir if it doesn't exist
+
 if (file.exists(file_data_sources)) {
-  table_of_data_sources <- read_csv(file = "./table_of_data_sources.csv")
+  table_of_data_sources <- read_csv(file = file_data_sources)
 } else {
   table_of_data_sources <- tibble(
     short_name = c("sex_age","la","hb"),
@@ -37,42 +42,75 @@ if (file.exists(file_data_sources)) {
   )
 }
 
+if (file.exists(file_historical_data_sources)) {
+  table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
+} else {
+  table_of_historical_data_sources <- tibble(
+    short_name = c("overall","sex_age","la","hb"),
+    url = c(
+      "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx"
+    )
+  ) %>% mutate(
+    filename = basename(url),
+    last_modified = ymd_hms("2020-01-01_01:01:01"),  # start with a date in the past
+    path_latest = NA_character_  
+  )
+}
+
 # Function to download latest data if available ---------------------------
 
-download_file_if_newer_available <- function(url, filename, last_modified, directory) {
+download_file_if_newer_available <- function(url, filename, last_modified, directory, record_tbl) {
   todays_file <- curl_fetch_disk(url, path = tempfile())
+  message(paste0("Checking if we have latest file for: ",filename))
+  message(paste0("Latest file was last modified at: ",todays_file$modified))
[…]

### 2021-02-12 covid_nrs_place_of_death.Rmd
modified .rmd to read & use file paths specified in the tables of msot recently updated data & historical data

@@
   data_sources: '`r tibble()`'
+  historical_data_sources: '`r tibble()`'
 ---
@@
 * Age
-* ~~Ethnicity~~ Note: not sure this is available
+* ~~Ethnicity~~ *Note: not sure this is available*
 
@@
 1.
-Using their SPARQL-based API ("data cube"), which can be accessed using the `opendatascot` package in R; the data can be seen on [the Scottish Gov't open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)
+~~Using their SPARQL-based API ("data cube"), which can be accessed using the `opendatascot` package in R; the data can be seen on [the Scottish Gov't open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)~~
+    - Update: it's difficult to keep track of data updates using the API - it's simpler to query the .xlsx/.csv files (next bullet point) for when they were last modified and download the weekly update once it's been released!
 2.
-There are a number of dataset, including the demographic/geographic crosstabulations available on the "Related Statistics" section of the [deaths involving COVID-19 website]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland)
+There are a number of datasets, including the demographic/geographic crosstabulations available on the "Related Statistics" section of the [deaths involving COVID-19 website]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland), under the *Ad-hoc queries section*.
+They are updated weekly, and also contain historical figures.
 
@@
 
-This means we'll need to download those figures from the website rather than using the API
+This means we'll need to download those figures from the website rather than using the API!
 
 Note that the weekly deaths due to COVID-19 in 2020 & 2021 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the weekly figures for individual years & the appropriate cross-tabulations.
+
+```{r}
+# "unpack" the data source tables
+table_of_data_sources <- params$data_sources
+table_of_historical_data_sources <- params$historical_data_sources
+
+table_of_data_sources %>%
+  kable(caption = "Latest data used in this report.")
+```
 
@@
 
+files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
+
 # checking out structure
@@
 
-weekly_deaths_sex_age_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
-                destfile = tempfile(),
-                quiet = FALSE)
+# weekly_deaths_sex_age_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
 
 weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(weekly_deaths_sex_age_file, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
+  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
   drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
@@
 
-weekly_deaths_la_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
-                destfile = tempfile(),
-                quiet = FALSE)
+# 
+# weekly_deaths_la_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
 
 weekly_deaths_la <- 
-  read_excel(weekly_deaths_la_file, sheet = 2, skip = 2) %>% 
+  read_excel(files_weekly_deaths$la, sheet = 2, skip = 2) %>% 
   clean_names %>%
@@
 
-weekly_deaths_hb_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
-                destfile = tempfile(),
-                quiet = FALSE)
+# weekly_deaths_hb_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
 
 weekly_deaths_hb <- 
-  read_excel(weekly_deaths_hb_file, sheet = 2, skip = 2) %>% 
+  read_excel(files_weekly_deaths$hb, sheet = 2, skip = 2) %>% 
   clean_names %>%
@@
 ```{r, warning=FALSE}
+files_historical_deaths <- set_names(x = as.list(table_of_historical_data_sources$path_latest), nm = table_of_historical_data_sources$short_name)
+
 ## overall file - where did I get the link from? it's not obviously available from the website...
-past_deaths_overall_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
-                destfile = tempfile(),
-                quiet = FALSE)
-
-past_deaths_hb_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
-                destfile = tempfile(),
-                quiet = FALSE)
-
-past_deaths_la_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
-                destfile = tempfile(),
-                quiet = FALSE)
-
-past_deaths_sex_age_file <-
-  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
-                destfile = tempfile(),
-                quiet = FALSE)
+# past_deaths_overall_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
+# 
+# past_deaths_hb_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
+# 
+# past_deaths_la_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
+# 
+# past_deaths_sex_age_file <-
+#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
+#                 destfile = tempfile(),
+#                 quiet = FALSE)
 
 past_deaths_overall <- 
-  read_csv(past_deaths_overall_file, skip = 2) %>%
+  read_csv(files_historical_deaths$overall, skip = 2) %>%
   rename(year = 1) %>%  # rename first column
@@
 past_deaths_hb <- 
-  readxl::read_excel(past_deaths_hb_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  readxl::read_excel(files_historical_deaths$hb, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
   clean_names() %>%
@@
 past_deaths_la <-
-  readxl::read_excel(past_deaths_la_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  readxl::read_excel(files_historical_deaths$la, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
   clean_names() %>%
@@
 past_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(past_deaths_sex_age_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
+  readxl::read_excel(files_historical_deaths$sex_age, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
   drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice

### 2021-02-12 covid_nrs_place_of_death.Rmd
fixed & included map data & my first map plot - ratio of home death 2020 to historical mean. need to configure size of text to plot etc.

@@
 library(sf)  # for mapping
-library(maps)
 ```
@@
 
-TODO: figure out which boundaries in Scotland correspond to council areas!
-
-~~To get council areas in Scotland we need NUTS 2 level map data~~, and the most appropriate level of detail to use is likely the BUC - Ultra Generalised Clipped
-
-https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom
-
-https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom?geometry=-12.122%2C55.353%2C4.566%2C57.479
-
-health board boundaries April 2019, BUC
-https://opendata.arcgis.com/datasets/def40bdc98a9457aa108eb3a5fb052b1_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D
-
-getting error trying to use shapefile with ggplot:
-> Error in CPL_transform(x, crs, aoi, pipeline, reverse) : 
-  crs not found: is it missing?
-
-explicitly setting crs on shapefile fails,
-
-version fo sf 0.9-7:
-> Linking to GEOS 3.8.0, GDAL 3.0.4, PROJ 6.3.1
-
-```{r}
-# TODO: learn how to use maps etc.
-zip_file <- tempfile()
-unzipped <- tempfile()
-
-## Scottish LAU 2011 level 1
-# curl_download(url = "https://borders.ukdataservice.ac.uk/ukborders/easy_download/prebuilt/shape/Scotland_laulevel1_2011.zip", destfile = zip_file, quiet = FALSE)  # download file from ONS
-# name <- list.files(unzipped, pattern=".shp")
-# shapefile <- st_read(file.path(unzipped, name))
-
-## below fails, saying cannot allocate vector of size XX MB/GB
-# curl_download(url = "https://opendata.arcgis.com/datasets/7f83b82ef6ce46d3a5635d371e8a3e7c_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D", destfile = zip_file, quiet = FALSE)  # download file from ONS
-
-download_and_import_shapefile <- function(url) {
-  zip_file <- tempfile()
-  unzipped <- tempfile()
-  downloaded_zip <- curl_download(url = url, destfile = zip_file, quiet = FALSE) 
-  unzip(zipfile = zip_file, exdir = unzipped)
-  name <- list.files(unzipped, pattern=".shp$")
-  shapefile <- st_read(file.path(unzipped, name))
-  return(shapefile)
-}
-
-health_boards_shapefile_url <- "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip"
-la_shapefile_url <- "https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"
-
-shapefile_hb <- download_and_import_shapefile(url = health_boards_shapefile_url)
-shapefile_la <- download_and_import_shapefile(url = la_shapefile_url)
-
-# shapefile %>%
-#   ggplot(aes(geometry = geometry)) +
+I previously downloaded shapefiles for:
+
+* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
+* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
+  - Note: this is a fairly high resolution, don't need this much detail!
+  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)
+
+```{r}
+# shapefile_hb <- st_read(dsn = "./map_data/hb")  # too detailed!
+shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
+shapefile_la <- st_read(dsn = "./map_data/la")
+
+# merged_deaths_la %>%
+#   filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
+#   left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
+#   ggplot(aes(fill = deaths_all_causes, geometry = geometry, group=ref_area)) +
 #   geom_sf()
-
-# shapefile_la %>%
-#   inner_join(merged_deaths_la, by = c("lad19nm"="ref_area"))
-
-# shapefile_hb %>%
-#   inner_join(merged_deaths_hb, by = c("HBName"="ref_area"))
-
-# TODO: summarise data by LA/HB so it can be shown on a map
-# otherwise it loads forever because we're re-plotting the same regional boundary for each row!!
-
-merged_deaths_la %>%
-  filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
-  left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
-  ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
-  geom_sf()
-
-# ggplot(data = shapefile_hb, aes(geometry=geometry)) + geom_sf()
-
+# 
 # merged_deaths_hb %>%
-#   left_join(shapefile_hb,by=c("ref_area"="HBName")) %>%
-#   ggplot(aes(fill = deaths_mean, geometry = geometry)) +
-#   geom_sf()
-
+#   filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
+#   left_join(shapefile_hb_uk_wide,by=c("ref_area"="name")) %>%
+#   ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
+#   geom_sf() +
+#   theme(aspect.ratio = 1.3)
+
+# ggplot(data = shapefile_hb_uk_wide, aes(geometry=geometry)) + geom_sf()
+
+## simple UK map
 # ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
@@
   )
+
+compute_historical_annual_average_deaths <- function(data_tbl, ...) {
+  grouping_vars <- enquos(...)
+  data_tbl %>%
+    group_by(year, place_of_death, !!!grouping_vars) %>%
+    summarise(  # calculate annual totals first...
+    number_of_deaths = sum(number_of_deaths), 
+    .groups = "keep"
+  ) %>%
+  ungroup(year) %>%  # remove year as grouping to average over year now
+  summarise(
+    annual_deaths_mean = mean(number_of_deaths),
+    annual_deaths_min = min(number_of_deaths),
+    annual_deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+}
+
+historical_annual_average_hb <-
+  past_deaths_hb %>%
+  compute_historical_annual_average_deaths(ref_area)
+
+historical_annual_average_la <-
+  past_deaths_la %>%
+  compute_historical_annual_average_deaths(ref_area)
+
+historical_annual_average_sex_age <-
+  past_deaths_sex_age %>%
+  compute_historical_annual_average_deaths(sex, age)
+
+historical_annual_average_sex <-
+  past_deaths_sex %>%
+  compute_historical_annual_average_deaths(sex)
+
+historical_annual_average_age <-
+  past_deaths_age %>%
+  compute_historical_annual_average_deaths(age)
 ```
@@
 ## compute rolling average
-
-merged_data_overall <-
+merged_deaths_overall <-
   weekly_deaths_overall %>%
@@
 merged_deaths_hb <-
-  # weekly_deaths_hb %>%
   rolling_deaths_hb %>%  # use data with rolling movingincluded
@@
 merged_deaths_la <-
-  # weekly_deaths_la %>%
   rolling_deaths_la %>%  # use data with rolling movingincluded
@@
 
-merged_data_sex_age <-
-  # weekly_deaths_sex_age %>%
+merged_deaths_sex_age <-
   rolling_deaths_sex_age %>%  # use data with rolling movingincluded
@@
 
-merged_data_sex <-
+merged_deaths_sex <-
   rolling_deaths_sex %>%
@@
 
-merged_data_age <-
+merged_deaths_age <-
   rolling_deaths_age %>%
@@
 
+# Summarise annual deaths for 2020 & 2021
+
+For visualising data using maps, we need to compute data for the entire year.
+Note: the non-covid and covid-related deaths in 2020 don't add up to all cause deaths due to the covid status only starting in week 12!
+
+```{r}
+compute_total_deaths_in_year <- function(data_tbl, ...) {
+  grouping_vars <- enquos(...)
+  data_tbl %>%
+    group_by(year, place_of_death, !!!grouping_vars) %>%
+    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%
+    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)
+}
+
+compute_ratio_of_annual_deaths_to_historical <- function(data_tbl) {
+  data_tbl %>%
+    mutate(
+      ratio_annual_deaths_to_historical_mean = deaths_all_causes / annual_deaths_mean,
+      ratio_annual_deaths_to_historical_min = deaths_all_causes / annual_deaths_min,
+      ratio_annual_deaths_to_historical_max = deaths_all_causes / annual_deaths_max
+    )
+}
+
+total_deaths_la <-
+  merged_deaths_la %>%
+  compute_total_deaths_in_year(ref_area) %>%
+  left_join(historical_annual_average_la, by = c("place_of_death", "ref_area")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_hb <-
+  merged_deaths_hb %>%
+  compute_total_deaths_in_year(ref_area) %>%
+  left_join(historical_annual_average_hb, by = c("place_of_death", "ref_area")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_sex <-
+  merged_deaths_sex %>%
+  compute_total_deaths_in_year(sex) %>%
+  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_age <-
+  merged_deaths_age %>%
+  compute_total_deaths_in_year(age) %>%
+  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+```
+
 # Visualisations
@@
 date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")
+
+only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(min(merged_deaths_age$date_w_c))}.")
 
@@
 
+## Total deaths in 2020 by age group & place of death
+
+```{r, fig.width=12, fig.height=7, warning=FALSE}
+bar_and_error_bars_2020_text <- "Bars are 2020 all-cause deaths, point and bars shows annual mean and range 2015-2019"
+excluded_other_place_2020 <- glue::glue("Excluding records with 'Other' place of death, N={total_deaths_age %>% filter(year==2020 & place_of_death=='Other') %>% .$deaths_all_causes %>% sum}")
+
+light_shaded_area_shows_difference <- "Lighter area in the top of bar shows difference between 2020 deaths and historical maximum 2015-2019"
+
+# TODO: make the distinction max vs 2020 figures clearer! maybe annotate?
+
+figure_annual_deaths_by_age_group_2020 <-
+  total_deaths_age %>% 
+  filter(year==2020) %>%
+  filter(place_of_death != "Other") %>%
+  ggplot(aes(x = age, y = deaths_all_causes)) +
+  # geom_col(fill = "#4477AA", alpha = 0.5, colour = "grey10") +
+  geom_col(fill = "grey20", colour = "grey10", alpha = 0.5) +
+  geom_col(aes(x=age, y=annual_deaths_max), fill = "#4477AA", colour = "grey10", alpha = 0.5) +  # historical maximum over the actual figure
+  geom_point(aes(x=age, y=annual_deaths_mean)) +
+  geom_errorbar(aes(x = age, ymin = annual_deaths_min, ymax = annual_deaths_max)) +
+  facet_wrap(~place_of_death) +
+  scale_y_continuous(labels = function(x) format(x,big.mark=",")) +
+  theme_minimal() +
+  labs(x = "Age",
+       y = "Deaths (N)",
+       title = "Deaths in 2020 by age group & place of death",
+       subtitle = paste0(c(only_2020_text,light_shaded_area_shows_difference),collapse="\n"),
+       caption = paste0(c(excluded_other_place_2020,bar_and_error_bars_2020_text,death_registrations_text,source_nrs_text), collapse="\n")
+  )
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_annual_deaths_by_age_group_2020",.x), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300)
+)
+
+figure_annual_deaths_by_age_group_2020
+```
+
 ## Recorded deaths in Scotland, by week number & location
@@
 ```{r, fig.width=12, fig.height=9, warning=FALSE}
-removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
+removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
 
 annotations_figure_place_of_death <-
-# datapasta::tribble_paste(merged_data_overall %>% select(place_of_death))
+# datapasta::tribble_paste(merged_deaths_overall %>% select(place_of_death))
   tibble::tribble(
@@
   mutate(
-    place_of_death = factor(place_of_death, levels = levels(merged_data_overall$place_of_death)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
+    place_of_death = factor(place_of_death, levels = levels(merged_deaths_overall$place_of_death)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
     label = str_wrap(label, width = 40)
@@
 figure_place_of_death <-
-  merged_data_overall %>%
+  merged_deaths_overall %>%
   filter(sex == "all" & age == "all") %>%
@@
 figure_deaths_at_home <-
-  merged_data_overall %>%
+  merged_deaths_overall %>%
   filter(sex == "all" & age == "all") %>%
@@
 figure_deaths_at_home_2020 <-
-  merged_data_overall %>%
+  merged_deaths_overall %>%
   filter(sex == "all" & age == "all") %>%
@@
 ```{r, fig.width=12, fig.height=9, warning=FALSE}
-removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_data_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_age %>% filter(year==2021) %>% .$week_number)} in 2021.")
+removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_deaths_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_age %>% filter(year==2021) %>% .$week_number)} in 2021.")
 
 figure_deaths_at_home_by_age <-
-  merged_data_age %>%
+  merged_deaths_age %>%
   filter(place_of_death %in% c("Home & other non-institution")) %>%
@@
 figure_deaths_at_home_by_sex <-
-  merged_data_sex %>%
+  merged_deaths_sex %>%
   mutate(sex = if_else(sex=="F", "Female", "Male")) %>%
@@
 # TODO: split data by covid-underlying cause of desath & covid-factor in death
-merged_data_overall %>%
+merged_deaths_overall %>%
   filter(sex=="all", age=="all") %>%
@@
 ```
+
+# Map-based visualisations
+
+## Ratio of 2020 home deaths to historical home deaths by LA
+
+```{r}
+areas_relatively_higher_number_home_deaths <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
+home_deaths_ratio <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
+
+## max/min used for setting scale of fill colour so it's consistent between plots!
+maximum_ratio_between_hb_and_la_home_deaths <-
+  bind_rows(
+    total_deaths_la,
+    total_deaths_hb
+  ) %>%
+  filter(year==2020 & place_of_death == "Home & other non-institution") %>%
+  .$ratio_annual_deaths_to_historical_mean %>%
+  max
+
+minimum_ratio_between_hb_and_la_home_deaths <-
+  bind_rows(
+    total_deaths_la,
+    total_deaths_hb
+  ) %>%
+  filter(year==2020 & place_of_death == "Home & other non-institution") %>%
+  .$ratio_annual_deaths_to_historical_mean %>%
+  min %>%
+  round(digits=1)
+
+# TODO: work out the size of these plots
+
+figure_map_ratio_home_deaths_2020_to_historical_la <-
+  total_deaths_la %>%
+  filter(
+    year == 2020 & place_of_death == "Home & other non-institution"
+  ) %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths, maximum_ratio_between_hb_and_la_home_deaths)) +
+  theme_minimal() +
+  theme(axis.text = element_blank()) +
+  labs(x="", y="",
+       subtitle = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority",
+       caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
+       fill = "Ratio"
+       ) +
+  NULL
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_map_ratio_home_deaths_2020_to_historical_la",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 12, height = 15, dpi = 300)
+)
+
+figure_map_ratio_home_deaths_2020_to_historical_hb <-
+  total_deaths_hb %>%
+  filter(
+    year == 2020 & place_of_death == "Home & other non-institution"
+  ) %>%
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths, maximum_ratio_between_hb_and_la_home_deaths)) +
+  theme_minimal() +
+  theme(axis.text = element_blank(), aspect.ratio = 1.4) +
+  labs(x="", y="",
+       subtitle = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board",
+       caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
+       fill = "Ratio"
+       ) +
+  NULL
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0("./outputs/figure_map_ratio_home_deaths_2020_to_historical_hb",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 12, height = 15, dpi = 300)
+)
+
+```

### 2021-02-12 download_latest_data_and_render_reports.R

@@ -112,17 +112,43 @@ for (x in 1:nrow(table_of_historical_data_sources)) {
 write_csv(table_of_data_sources, file = file_data_sources)
 write_csv(table_of_historical_data_sources, file = file_historical_data_sources)
 
-# Pass latest data to analysis function -----------------------------------
+# Download map data -------------------------------------------------------
+
+dir_map_data <- "./map_data"
+if (!dir.exists(dir_map_data)) dir.create(dir_map_data)
+
+download_shapefile <- function(url, directory, subdirectory) {
+  zip_file <- tempfile()
+  downloaded_zip <- curl_download(url = url, destfile = zip_file, quiet = FALSE) 
+  unzip(zipfile = zip_file, exdir = file.path(directory, subdirectory), overwrite = TRUE)
+  return(TRUE)
+}
+
+health_boards_shapefile_url <- "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip"  # massive resolution
+health_boards_uk_wide_url <- "https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip"
+# https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
+la_shapefile_url <- "https://opendata.arcgis.com/datasets/910f48f3c4b3400aa9eb0af9f8989bbe_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"
+
+
+download_shapefile(url = health_boards_shapefile_url, directory = dir_map_data, subdirectory = "hb")
+download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")
+download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
+
+# Render report with latest data ------------------------------------------
 
 dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
 
+## this line is also useful for testing - just run it and it pre-populates the
+## params list for the rmd!
+params <- list( # pass the table of data sources
+  data_sources = table_of_data_sources,
+  historical_data_sources = table_of_historical_data_sources
+)
+
 render(
   input = "./covid_nrs_place_of_death.Rmd",
-  output_format = paste0("covid_nrs_place_of_death_",today()),
+  output_file = paste0("covid_nrs_place_of_death_",today()),
[…]

### 2021-02-15 covid_nrs_place_of_death.Rmd
modified script to create an outputs folder based on the latest data everytime its run so hat rerunning the script weekly will produce new figures;
also made small modifications to map-based plots

@@
 
+## Set up output directory
+
+Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:
+
+```{r}
+latest_data_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
+
+dir_outputs <- paste0("./outputs/",latest_data_date)
+if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
+```
+
 ## Common elements
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_annual_deaths_by_age_group_2020",.x), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_annual_deaths_by_age_group_2020",.x), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_place_of_death",.x), plot = figure_place_of_death, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_place_of_death",.x), plot = figure_place_of_death, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2020",.x), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020",.x), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_health_board",.x), plot = figure_deaths_at_home_by_health_board, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_health_board",.x), plot = figure_deaths_at_home_by_health_board, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2week_average_by_health_board",.x), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_health_board",.x), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_local_authority",.x), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_local_authority",.x), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2week_average_by_local_authority",.x), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_local_authority",.x), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_age",.x), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_age",.x), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300)
 )
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_by_sex",.x), plot = figure_deaths_at_home_by_sex, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_sex",.x), plot = figure_deaths_at_home_by_sex, width = 12, height = 9, dpi = 300)
 )
@@
 
-## Ratio of 2020 home deaths to historical home deaths by LA
+## Ratio of 2020 home deaths to historical home deaths by LA / HB
 
@@
   round(digits=1)
-
+```
+
+### LA
+
+```{r, fig.height=12, fig.width=8}
 # TODO: work out the size of these plots
-
 figure_map_ratio_home_deaths_2020_to_historical_la <-
@@
   labs(x="", y="",
-       subtitle = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority",
+       title = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority",
+       subtitle = "A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.",
        caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_map_ratio_home_deaths_2020_to_historical_la",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 12, height = 15, dpi = 300)
-)
-
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 9, height = 12, dpi = 300)
+)
+
+figure_map_ratio_home_deaths_2020_to_historical_la
+```
+
+### HB
+
+```{r, fig.height=12, fig.width=8}
+# TODO: work out the size of these plots
 figure_map_ratio_home_deaths_2020_to_historical_hb <-
@@
   labs(x="", y="",
-       subtitle = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board",
+       title = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board",
+       subtitle = "A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.",
        caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
@@
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0("./outputs/figure_map_ratio_home_deaths_2020_to_historical_hb",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 12, height = 15, dpi = 300)
-)
-
-```
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 9, height = 12, dpi = 300)
+)
+
+figure_map_ratio_home_deaths_2020_to_historical_hb
+```

### 2021-02-15 download_latest_data_and_render_reports.R

@@ -10,6 +10,7 @@
 # Load packages -----------------------------------------------------------
 
 library(tidyverse)
+library(lubridate)  # dealing with dates
 library(curl)
 library(knitr)
 library(rmarkdown)
@@ -150,5 +151,6 @@ render(
   input = "./covid_nrs_place_of_death.Rmd",
   output_file = paste0("covid_nrs_place_of_death_",today()),
   output_dir = dir_reports,
-  params = params
+  params = params, 
+  envir = new.env()
 )

### 2021-02-18 covid_nrs_place_of_death.Rmd
started incorporating covid case counts, scope for lots of calculations; need to normalise health board names between datasets

@@
   historical_data_sources: '`r tibble()`'
+  case_trends_data_sources: '`r tibble()`'
 ---
@@
 endpoint <- "http://statistics.gov.scot/sparql.json"
+endpoint <- "http://statistics.gov.scot/sparql"
 ```
@@
 table_of_historical_data_sources <- params$historical_data_sources
+table_of_case_trends_data_sources <- params$case_trends_data_sources
 
@@
 
-# ods_all_datasets() %>% View()
-# ods_structure("deaths")  # can we import all deaths-related data this way?
+ods_all_datasets() %>% View()
+ods_structure("deaths")  # can we import all deaths-related data this way?
+ods_structure("coronavirus-covid-19-management-information")
+
+ods_dataset("coronavirus-covid-19-management-information", geography = "hb") %>%
+  filter(variable == "testing-new-cases-reported")
+
+ods_dataset("coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported")
+ods_dataset("coronavirus-covid-19-management-information", geography = "la", variable = "testing-new-cases-reported")
+ods_dataset("coronavirus-covid-19-management-information", variable = "testing-new-cases-reported")
+
+ods_print_query(dataset = "coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported") %>% dput
+ods_print_query(dataset = "coronavirus-covid-19-management-information", variable = "testing-new-cases-reported") %>% dput
+
+modified_query <- "PREFIX qb: <http://purl.org/linked-data/cube#>                 PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>                 PREFIX xsd: <http://www.w3.org/2001/XMLSchema#> select   ?refArea ?refPeriod ?measureType ?variable ?value where { ?data qb:dataSet <http://statistics.gov.scot/data/coronavirus-covid-19-management-information>.
+?data <http://purl.org/linked-data/sdmx/2009/dimension#refArea> ?refArea.
+?data <http://purl.org/linked-data/sdmx/2009/dimension#refPeriod> ?refPeriod.
+?data <http://purl.org/linked-data/cube#measureType> ?measureType.
+?data <http://statistics.gov.scot/def/dimension/variable> ?variable.
+?data ?measureType ?value.
+filter (regex(str(?variable ), 'testing-new-cases-reported'))  } order by ?refPeriod ?refArea"
+
+modified_query %>%
+  opendatascot:::ods_query_database(endpoint, query = .) %>%
+  opendatascot:::pre_process_data()
+
+query <- ""
+
+library(ckanr)
+ckanr_setup(url = "https://www.opendata.nhs.scot/")
+# package_list(as = "table")
+# tag_list(as = "table")
+package_search(q = 'daily', rows = 2, as = 'table') %>% str
+
+download_data_from_opendata_via_ckan_api <- function(resource_id) {
+  ckan <- ckanr::src_ckan("https://www.opendata.nhs.scot/")
+  dplyr::tbl(src = ckan$con, from = resource_id_daily_case_trends_la, ) %>% 
+  as_tibble()
+}
+
+resource_id_daily_case_trends_la <- "427f9a25-db22-4014-a3bc-893b68243055"
+
+resource_show(id = resource_id_daily_case_trends_la, as = "table")
+
 # ods_print_query("deaths")
@@
 ```{r}
-weekly_deaths_overall %>% 
-  count(ref_period, date_w_c, week_number, week_number_run_over) %>% 
+weekly_deaths_sex_age %>% 
+  count(year, date_w_c, week_number, week_number_run_over) %>% 
   slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
@@
 
+## Load COVID case counts by LA/HB
+
+```{r}
+files_case_counts <- set_names(x = as.list(table_of_case_trends_data_sources$path_latest), nm = table_of_case_trends_data_sources$short_name)
+
+cases_hb <- read_csv(
+  files_case_counts$hb, 
+  col_types = 
+    cols(
+    .default = col_double(),
+    `_id` = col_skip(),
+    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
+    Date = col_date(format = "%Y%m%d"),
+    HB = col_character(),
+    HBName = col_character(),
+    HospitalAdmissionsQF = col_character(),
+    ICUAdmissionsQF = col_character()
+    ),
+  ) %>% clean_names() %>%
+  rename(ref_area = hb_name) %>%
+  mutate(
+    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8))
+    )
+
+cases_la <- read_csv(
+  file = files_case_counts$la, 
+  col_types = 
+    cols(
+    .default = col_double(),
+    `_id` = col_skip(),
+    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
+    Date = col_date(format = "%Y%m%d"),
+    CA = col_character(),
+    CAName = col_character()
+  )
+) %>% clean_names() %>%
+  rename(ref_area = ca_name) %>%
+  mutate(
+    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8))
+    )
+```
+
+# Calculate cumulative case counts at weekly deaths
+
+Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks.
+This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
+However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!
+
+```{r}
+weekly_covid_cases_hb <-
+  cases_hb %>%
+  group_by(year, week_number, ref_area) %>%
+  summarise(
+    weekly_positive = sum(daily_positive),
+    cumulative_positive = max(cumulative_positive),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
+    )
+
+```
+
 # Calculate historical averages, minima & maxima
@@
 
-# Summarise annual deaths for 2020 & 2021
+# Annual deaths for 2020 & 2021
 

### 2021-02-18 download_latest_data_and_render_reports.R

@@ -19,6 +19,7 @@ library(rmarkdown)
 
 file_data_sources <- "./table_of_data_sources.csv"
 file_historical_data_sources <- "./table_of_historical_data_sources.csv"
+file_case_trends_data_sources <- "./table_of_case_trends_data_sources.csv"
 
 download_dir <- "./downloaded_data"
 if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist
@@ -61,6 +62,24 @@ if (file.exists(file_historical_data_sources)) {
   )
 }
 
+
+if (file.exists(file_case_trends_data_sources)) {
+  table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)
+} else {
+  table_of_case_trends_data_sources <- tibble(
+    short_name = c("la","hb"),
+    url = c(
+      "https://www.opendata.nhs.scot/datastore/dump/427f9a25-db22-4014-a3bc-893b68243055?bom=True",
+      "https://www.opendata.nhs.scot/datastore/dump/2dd8534b-0a6f-4744-9253-9565d62f96c2?bom=True"
+    )
+  ) %>% mutate(
+    filename = c("trend_ca.csv","trend_hb.csv"),
+    # last_modified = ymd_hms("2020-01-01_01:01:01"),  # start with a date in the past
+    path_latest = NA_character_  
+  )
+}
+
+
 # Function to download latest data if available ---------------------------
 
 download_file_if_newer_available <- function(url, filename, last_modified, directory, record_tbl) {
@@ -81,6 +100,17 @@ download_file_if_newer_available <- function(url, filename, last_modified, direc
     message("We already have the most recent file! No download needed.")
   }
   message("...")
+  return(record_tbl)
+}
+
+download_file_regardless <- function(url, filename, directory, record_tbl) {
+  todays_date_string <- today()
+  todays_filename <- paste0(todays_date_string, "_", filename)
+  file_path <- file.path(directory,todays_filename)
+  message(paste0("Downloading file: ",todays_filename))
+  curl_download(url = url, destfile = file_path, quiet = FALSE)
+  message(paste0("Downloaded file to: ",file_path))
+  record_tbl$path_latest[which(record_tbl$filename == filename)] <- file_path
   return(record_tbl)
 }
 
@@ -108,10 +138,24 @@ for (x in 1:nrow(table_of_historical_data_sources)) {
     )
 }
 
[…]

### 2021-02-18 covid_nrs_place_of_death.Rmd
normalised HB/LA names between datasets and added some notes

@@
 
+I've also attempted loading the case counts using `ckanr` using the CKAN API from the opendata.nhs.scot website, but gave that up in favour of using their `.csv` links instead to download daily-updated data.
+
+Note about weekly LA/HB data:
+
+> From 14/12/2020 confirmed positive cases are now allocated to an area based on the postcode recorded in the testing system.
+Previously cases were allocated to local areas mainly based on their usual residential address recorded in the Community Health Index (CHI) database.
+This change will impact on newly reported cases on this day.
+
+> Please note that the updated methodology for determining COVID-19 related ICU admissions has been applied to the historic data.
+The updated definition is as follows: A patient who has tested positive for COVID at any time in the 21 days prior to admission to ICU, or who have tested positive from the date of admission up to and including the date of ICU discharge.
+
+> In addition, on 30th October it came to our attention that there is an ongoing issue when linking ICU data to the lab data for COVID-19 test results.
+Any COVID-19 positive patients with a missing CHI that had a first positive test in the community are unable to be linked to the ICU data.
+As a result the COVID-19 positive numbers could be underreported by a small number.
+
+`crude_rate_positive`: Crude rate of total positive cases (cumulative) per 100,000 population.
+
 ```{r}
@@
 
-ods_dataset("coronavirus-covid-19-management-information", geography = "hb") %>%
-  filter(variable == "testing-new-cases-reported")
-
-ods_dataset("coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported")
-ods_dataset("coronavirus-covid-19-management-information", geography = "la", variable = "testing-new-cases-reported")
-ods_dataset("coronavirus-covid-19-management-information", variable = "testing-new-cases-reported")
-
-ods_print_query(dataset = "coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported") %>% dput
-ods_print_query(dataset = "coronavirus-covid-19-management-information", variable = "testing-new-cases-reported") %>% dput
-
-modified_query <- "PREFIX qb: <http://purl.org/linked-data/cube#>                 PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>                 PREFIX xsd: <http://www.w3.org/2001/XMLSchema#> select   ?refArea ?refPeriod ?measureType ?variable ?value where { ?data qb:dataSet <http://statistics.gov.scot/data/coronavirus-covid-19-management-information>.
+# ods_dataset("coronavirus-covid-19-management-information", geography = "hb") %>%
+#   filter(variable == "testing-new-cases-reported")
+# 
+# ods_dataset("coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported")
+# ods_dataset("coronavirus-covid-19-management-information", geography = "la", variable = "testing-new-cases-reported")
+# ods_dataset("coronavirus-covid-19-management-information", variable = "testing-new-cases-reported")
+# 
+# ods_print_query(dataset = "coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported") %>% dput
+# ods_print_query(dataset = "coronavirus-covid-19-management-information", variable = "testing-new-cases-reported") %>% dput
+#   
+# modified_query <- "PREFIX qb: <http://purl.org/linked-data/cube#>                 PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>                 PREFIX xsd: <http://www.w3.org/2001/XMLSchema#> select   ?refArea ?refPeriod ?measureType ?variable ?value where { ?data qb:dataSet <http://statistics.gov.scot/data/coronavirus-covid-19-management-information>.
 ?data <http://purl.org/linked-data/sdmx/2009/dimension#refArea> ?refArea.
@@
 filter (regex(str(?variable ), 'testing-new-cases-reported'))  } order by ?refPeriod ?refArea"
-
-modified_query %>%
-  opendatascot:::ods_query_database(endpoint, query = .) %>%
-  opendatascot:::pre_process_data()
-
-query <- ""
-
-library(ckanr)
-ckanr_setup(url = "https://www.opendata.nhs.scot/")
-# package_list(as = "table")
-# tag_list(as = "table")
-package_search(q = 'daily', rows = 2, as = 'table') %>% str
-
-download_data_from_opendata_via_ckan_api <- function(resource_id) {
-  ckan <- ckanr::src_ckan("https://www.opendata.nhs.scot/")
-  dplyr::tbl(src = ckan$con, from = resource_id_daily_case_trends_la, ) %>% 
-  as_tibble()
-}
-
-resource_id_daily_case_trends_la <- "427f9a25-db22-4014-a3bc-893b68243055"
-
-resource_show(id = resource_id_daily_case_trends_la, as = "table")
+# 
+# modified_query %>%
+#   opendatascot:::ods_query_database(endpoint, query = .) %>%
+#   opendatascot:::pre_process_data()
+# 
+# query <- ""
+
+# 
+# library(ckanr)
+# ckanr_setup(url = "https://www.opendata.nhs.scot/")
+# # package_list(as = "table")
+# # tag_list(as = "table")
+# package_search(q = 'daily', rows = 2, as = 'table') %>% str
+# 
+# 
+# download_data_from_opendata_via_ckan_api <- function(resource_id) {
+#   ckan <- ckanr::src_ckan("https://www.opendata.nhs.scot/")
+#   dplyr::tbl(src = ckan$con, from = resource_id_daily_case_trends_la, ) %>% 
+#   as_tibble()
+# }
+# 
+# resource_id_daily_case_trends_la <- "427f9a25-db22-4014-a3bc-893b68243055"
+# 
+# resource_show(id = resource_id_daily_case_trends_la, as = "table")
 
@@
     year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
-    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8))
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
+    ref_area = phsmethods::match_area(hb)  # convert to standard HB names
     )
@@
     year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
-    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8))
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
+    ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
     )
@@
 
+weekly_covid_cases_la <-
+  cases_la %>%
+  group_by(year, week_number, ref_area) %>%
+  summarise(
+    weekly_positive = sum(daily_positive),
+    cumulative_positive = max(cumulative_positive),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
+    )
 ```
@@
 ```
+
+# Comparing geographic areas by case counts
+
+We can use the yearly cumulative case count to group geographic areas by how many cases they've had
+
+```{r}
+
+```

### 2021-02-19 covid_nrs_place_of_death.Rmd
added small changes to home deaths all scotland plot to improve it; also added computation of annual ratios of all deaths to past deaths; started looking at number of covid cases by region and splitting regions by cases

@@
 library(sf)  # for mapping
+library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
+library(patchwork)  # for assembling plots
 ```
@@
 This is a potential future TODO!
+
+# Set up output directory
+
+Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:
+
+```{r}
+latest_data_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
+
+dir_outputs <- paste0("./outputs/",latest_data_date)
+if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
+```
 
@@
       mutate(dataset=name,year=as.numeric(year)) %>%
-      filter(!is.na(year) & place_of_death!="All") %>%
+      filter(!is.na(year) & place_of_death!="All") %>%  # to remove summary values from the data
       group_by(dataset,year) %>%
@@
     cumulative_positive = max(cumulative_positive),
+    crude_rate_positive_per100k = max(crude_rate_positive),
     .groups = "drop"
@@
     )
+```
+
+# Comparing geographic areas by case counts
+
+We can use the yearly cumulative case count to group geographic areas by how many cases they've had by the end of 2020.
+
+```{r}
+cumulative_case_2020_based_grouping_la <-
+  weekly_covid_cases_la %>% 
+  filter(ref_area!="Scotland") %>%
+  filter(year == 2020) %>% 
+  filter(week_number == max(week_number)) %>%  # keep last week
+  select(ref_area, cumulative_positive, crude_rate_positive_per100k)
+
+cumulative_case_2020_based_grouping_hb <-
+  weekly_covid_cases_hb %>% filter(year == 2020) %>% 
+  filter(ref_area!="Scotland") %>%
+  filter(year == 2020) %>% 
+  filter(week_number == max(week_number)) %>%  # keep last week 
+  select(ref_area, cumulative_positive, crude_rate_positive_per100k)
+
+k_means_clustering_cases_and_per100k_2020_hb <- map(
+  .x = 2:4,
+  .f = ~kmeans(x=cumulative_case_2020_based_grouping_hb %>% select(cumulative_positive, crude_rate_positive_per100k), centers = .x, nstart = 1e3)
+) %>% set_names(nm=paste0("k",2:4))
+
+clusters_hb <-
+  bind_cols(
+    cumulative_case_2020_based_grouping_hb %>% select(ref_area),
+    map_dfc(.x = k_means_clustering_cases_and_per100k_2020_hb, .f = function(x) return(x$cluster))
+  )
+
+cumulative_case_2020_based_grouping_hb <-
+  cumulative_case_2020_based_grouping_hb %>%
+  left_join(clusters_hb, by="ref_area") %>%
+  mutate(
+    quartiles_cases = ntile(cumulative_positive, n=4),
+    quartiles_per100k = ntile(crude_rate_positive_per100k, n=4)
+  )
+
+# TODO: apply to LA also!
+
+hb_cases_plots <- list()
+
+hb_cases_plots$cumulative_cases_vs_per100k <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
+  geom_point() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases",
+    y="COVID cases per 100,000 population",
+    title="Cases vs. population rate"
+  ) +
+  NULL
+
+hb_cases_plots$cumulative_cases_vs_per100k_log10 <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
+  geom_point() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases (log)",
+    y="COVID cases per 100,000 population (log)",
+    title="Cases vs. population rate (log scale axes)"
+  ) +
+  scale_x_log10() +
+  scale_y_log10() +
+  NULL
+
+hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_cases))) +
+  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
+  scale_colour_viridis_d() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases",
+    y="COVID cases per 100,000 population",
+    title="Cases vs. population rate"
+  ) +
+  NULL
+
+hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_per100k))) +
+  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
+  scale_colour_viridis_d() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases",
+    y="COVID cases per 100,000 population",
+    title="Cases vs. population rate",
+    colour = "Quartile (Cases per 100,000)"
+  ) +
+  NULL
+
+hb_cases_plots$cumulative_cases_vs_per100k_k2 <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k2))) +
+  geom_point(size=3, aes(colour=factor(k2))) +
+  scale_colour_viridis_d() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases",
+    y="COVID cases per 100,000 population",
+    title="Cases vs. population rate"
+  ) +
+  NULL
+
+hb_cases_plots$cumulative_cases_vs_per100k_k3 <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k3))) +
+  geom_point(size=3, aes(colour=factor(k3))) +
+  scale_colour_viridis_d() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases",
+    y="COVID cases per 100,000 population",
+    title="Cases vs. population rate"
+  ) +
+  NULL
+
+hb_cases_plots$cumulative_cases_vs_per100k_k4 <-
+  cumulative_case_2020_based_grouping_hb %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k4))) +
+  geom_point(size=3, aes(colour=factor(k4))) +
+  scale_colour_viridis_d() +
+  geom_label_repel() +
+  theme_minimal() +
+  labs(
+    x="COVID cases",
+    y="COVID cases per 100,000 population",
+    title="Cases vs. population rate"
+  ) +
+  NULL
+
+hb_cases_plots_assembled <- patchwork::wrap_plots(hb_cases_plots, ncol = 2)
+
+total_deaths_hb %>% filter(year==2020 & str_starts(place_of_death, pattern="Home")) %>%
+  left_join(cumulative_case_2020_based_grouping_hb, by="ref_area") %>%
+  ggplot(data=., aes(x=cumulative_positive, y=deaths_all_causes, label=ref_area)) +
+  geom_point() +
+  geom_label_repel() +
+  theme_minimal() +
+  NULL
+
+# k means clustering
+hb_clustered <- kmeans(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland") %>% select(2:3), centers = 4, nstart = 2500)
+bind_cols(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland"), cluster=hb_clustered$cluster) %>%
+  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, fill=cluster, colour=cluster)) +
+    geom_point() +
+    geom_label_repel() +
+    theme_minimal() +
+    NULL
+
 ```
@@
 
-# Calculate data averaged over multiple weeks
+# Calculate moving averages (rolling mean) over multiple weeks
 
@@
 
+# Comparisons of annual rates with current rates
+
+```{r}
+## this is fairly ad-hoc... but it works!
+table_comparing_rates <-
+  bind_rows(
+    ## all available data
+    merged_data_overall %>%
+      filter(age == "all" &
+               sex == "all" &
+               str_starts(place_of_death, "Home")) %>% group_by(year, week_number) %>%
+      summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% left_join(
+        past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
+                                         year != "average") %>% group_by(week_number) %>% summarise(
+                                           average_weekly_deaths = sum(number_of_deaths) / 5,
+                                           .groups = "drop"
+                                         ),
+        by = "week_number"
+      ) %>% summarise(
+        period = "all available data",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        past_annual_deaths = sum(average_weekly_deaths)
+      )
+    ,
+    ## 2020 only
+    bind_cols(
+      merged_data_overall %>%
+        filter(
+          age == "all" &
+            sex == "all" & str_starts(place_of_death, "Home") & year == 2020
+        ) %>%
+        summarise(
+          period = "2020",
+          deaths_all_causes = sum(deaths_all_causes),
+          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
+        ),
+      past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
+                                       year != "average") %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
+    ),
+    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
+    bind_cols(
+      merged_data_overall %>%
+        filter(
+          age == "all" &
+            sex == "all" &
+            str_starts(place_of_death, "Home") &
+            year == 2020 & week_number >= 12
+        ) %>%
+        summarise(
+          period = "2020, from wk12",
+          deaths_all_causes = sum(deaths_all_causes),
+          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
+        ),
+      past_deaths_overall %>% filter(
+        week_number >= 12 &
+          str_starts(place_of_death, "Home") &
+          year != "average"
+      ) %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
+    )
+  ) %>%
+  mutate(
+    ratio_deaths_all_causes_to_historical = deaths_all_causes / past_annual_deaths,
+    ratio_non_covid_deaths_to_historical = deaths_non_covid / past_annual_deaths,
+    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
+  )
+
+write_csv(x = table_comparing_rates, file = file.path(dir_outputs, "annual_death_rates_comparison_with_historical_deaths.csv"))
+
+table_comparing_rates %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
+```
+
 # Visualisations
 
-## Set up output directory
-
-Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:
-
-```{r}
-latest_data_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
-
-dir_outputs <- paste0("./outputs/",latest_data_date)
-if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
-```
-
 ## Common elements
@@
 ```{r}
+theme_set(theme_minimal(base_size = 16) +
+            theme(panel.grid.minor = element_blank(),
+                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
+                  axis.title.x = element_text(margin = margin(20, 0, 0, 0))
+            ))
+
 date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")
+
+dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."
 
@@
 
-## Deaths at home, all of Scotland
+## All of Scotland, deaths at home
 
@@
 # TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
+total_deaths_at_home_in_period <- glue::glue("Total deaths at home N={format(merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death %in% c('Home & other non-institution')) %>% .$deaths_all_causes %>% sum, big.mark=',')}")
+
 figure_deaths_at_home <-
@@
   geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
   geom_vline(xintercept = 54, linetype = "dotted") +
   scale_x_continuous(breaks = seq(0,60,by=10)) +
+  ylim(0, 600) +
   facet_grid(~place_of_death) +
@@
                       ) +
+  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
@@
        subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
        colour = "Cause of death"
@@
 
+## figure for a bit less than half A4 page is approx 14*11cm
 walk(
   .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 12, height = 9, dpi = 300)
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 14, height = 11, units = "cm", dpi = 300)
 )
@@
 
-# Comparing geographic areas by case counts
-
-We can use the yearly cumulative case count to group geographic areas by how many cases they've had
-
-```{r}
-
-```
+```{r}
+
+```

### 2021-02-19 covid_nrs_place_of_death.Rmd
started tinkering with dual axis plot for case counts & weekly deaths. Probably better to split into two plots side-by-side, otherwise quite clutterd!

@@
 
-# Calculate cumulative case counts at weekly deaths
+# Calculate case counts: weekly counts to match weekly death reporting
 
@@
     )
+
+## extract Scotland-wide data and remove from HB split
+weekly_covid_cases_overall <- weekly_covid_cases_hb %>% filter(ref_area=="Scotland")
+weekly_covid_cases_hb <- weekly_covid_cases_hb %>% filter(ref_area!="Scotland")
 
@@
     group_by(year, place_of_death, !!!grouping_vars) %>%
-    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%
-    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)
+    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%  # compute sums of all types of death
+    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)  # compute a smaller number of all-cause deaths since covid started (over 2020, covid- and non-covid-related deaths don't add up to all-cause deaths because pre-covid deaths are included also!)
 }
@@
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  ylim(0, 600) +
+  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
   facet_grid(~place_of_death) +
@@
 
+# Case counts & home deaths - comparing plots
+
+## Scotland-wide figure
+
+```{r}
+# TODO: wokr out how to include both... two panels? two axes on same plot?
+
+ratio_max_weekly_deaths_to_cases <- 600 / 16e3
+
+# figure_home_deaths_and_cases_scotland <-
+  merged_deaths_overall %>%
+  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
+  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  geom_line(aes(y = weekly_positive * ratio_max_weekly_deaths_to_cases), colour="purple") +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly deaths")) +
+  # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  theme_minimal() +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       )
+
+# ## figure for a bit less than half A4 page is approx 14*11cm
+# walk(
+#   .x = c(".pdf",".svg",".png"),
+#   .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 14, height = 11, units = "cm", dpi = 300)
+# )
+# 
+# figure_deaths_at_home
+```
+
 # Map-based visualisations

### 2021-02-22 covid_nrs_place_of_death.Rmd
modified the scripts not to pass params to .Rmd but instead to load the latest data csv instead

@@
   chunk_output_type: console
-params:
-  data_sources: '`r tibble()`'
-  historical_data_sources: '`r tibble()`'
-  case_trends_data_sources: '`r tibble()`'
 ---
@@
 ```{r}
-endpoint <- "http://statistics.gov.scot/sparql.json"
+# endpoint <- "http://statistics.gov.scot/sparql.json"
 endpoint <- "http://statistics.gov.scot/sparql"
@@
 
-# Set up output directory
+# Load info about latest data
+
+A separate script checks for updated data and downloads it, also updating these files with the locations of the downloaded data:
+
+```{r}
+file_data_sources <- "./table_of_data_sources.csv"
+file_historical_data_sources <- "./table_of_historical_data_sources.csv"
+file_case_trends_data_sources <- "./table_of_case_trends_data_sources.csv"
+
+table_of_data_sources <- read_csv(file = file_data_sources)
+table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
+table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)
+
+table_of_data_sources %>%
+  kable(caption = "Latest data used in this report.")
+```
+
+## Set up output directory
 
@@
 ```{r}
-# "unpack" the data source tables
-table_of_data_sources <- params$data_sources
-table_of_historical_data_sources <- params$historical_data_sources
-table_of_case_trends_data_sources <- params$case_trends_data_sources
-
-table_of_data_sources %>%
-  kable(caption = "Latest data used in this report.")
+# TODO: remove this section, params not used anymore
+# # "unpack" the data source tables
+# table_of_data_sources <- params$data_sources
+# table_of_historical_data_sources <- params$historical_data_sources
+# table_of_case_trends_data_sources <- params$case_trends_data_sources
 ```
@@
 
-ods_all_datasets() %>% View()
-ods_structure("deaths")  # can we import all deaths-related data this way?
-ods_structure("coronavirus-covid-19-management-information")
+# ods_all_datasets() %>% View()
+# ods_structure("deaths")  # can we import all deaths-related data this way?
+# ods_structure("coronavirus-covid-19-management-information")
 
@@
 
-total_deaths_hb %>% filter(year==2020 & str_starts(place_of_death, pattern="Home")) %>%
-  left_join(cumulative_case_2020_based_grouping_hb, by="ref_area") %>%
-  ggplot(data=., aes(x=cumulative_positive, y=deaths_all_causes, label=ref_area)) +
-  geom_point() +
-  geom_label_repel() +
-  theme_minimal() +
-  NULL
-
-# k means clustering
-hb_clustered <- kmeans(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland") %>% select(2:3), centers = 4, nstart = 2500)
-bind_cols(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland"), cluster=hb_clustered$cluster) %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, fill=cluster, colour=cluster)) +
-    geom_point() +
-    geom_label_repel() +
-    theme_minimal() +
-    NULL
-
+hb_cases_plots_assembled
 ```
@@
     ## all available data
-    merged_data_overall %>%
+    merged_deaths_overall %>%
       filter(age == "all" &
@@
     bind_cols(
-      merged_data_overall %>%
+      merged_deaths_overall %>%
         filter(
@@
 source_nrs_text <- "Source: National Records of Scotland"
+source_phs_text <- "Source: Public Health Scotland"
 
@@
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly deaths")) +
+  scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly cases")) +
   # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
@@
 
+  merged_deaths_overall %>%
+  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
+  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
+  ggplot(aes(x=week_number_run_over, y=weekly_positive)) +
+  geom_line() +
+  theme_minimal() +
+  scale_y_continuous(labels = function(x) format(x, big.mark=",")) +
+  labs(
+    x = "Week number", y = "COVID-19 positive cases (N)",
+    subtitle = date_range_text,
+    caption = paste0(c(vertical_line_text, source_phs_text), collapse="\n")
+  )
+
 # ## figure for a bit less than half A4 page is approx 14*11cm

### 2021-02-22 download_latest_data_and_render_reports.R

@@ -184,18 +184,19 @@ download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, su
 dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
 
+# TODO: remove this section, params no longer used
 ## this line is also useful for testing - just run it and it pre-populates the
 ## params list for the rmd!
-params <- list( # pass the table of data sources
-  data_sources = table_of_data_sources,
-  historical_data_sources = table_of_historical_data_sources,
-  case_trends_data_sources = table_of_case_trends_data_sources
-)
+# params <- list( # pass the table of data sources
+#   data_sources = table_of_data_sources,
+#   historical_data_sources = table_of_historical_data_sources,
+#   case_trends_data_sources = table_of_case_trends_data_sources
+# )
 
 render(
   input = "./covid_nrs_place_of_death.Rmd",
   output_file = paste0("covid_nrs_place_of_death_",today()),
   output_dir = dir_reports,
-  params = params, 
+  # params = params, 
   envir = new.env()
 )

### 2021-02-22 covid_nrs_place_of_death.Rmd
added visualisation of proportion of deaths by place over time

@@
 
+historical_annual_average_overall <-
+  past_deaths_overall %>%
+  compute_historical_annual_average_deaths()
+
 historical_annual_average_hb <-
@@
   left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_overall <-
+  merged_deaths_overall %>%
+  compute_total_deaths_in_year() %>%
+  left_join(historical_annual_average_overall, by = c("place_of_death")) %>%
   compute_ratio_of_annual_deaths_to_historical
@@
 
+# Comparing historical rates of home death to total death to 2020/2021
+
+```{r}
+proportions_of_historical_deaths_by_place <-
+  past_deaths_overall %>% 
+  group_by(year, place_of_death) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%
+  filter(place_of_death != "All" & !is.na(as.integer(year))) %>%
+  group_by(year) %>%
+  mutate(
+    total_deaths = sum(number_of_deaths)
+  ) %>%
+  ungroup %>%
+  mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))
+
+proportions_of_current_deaths_by_place <-
+  total_deaths_overall %>%
+  select(year, place_of_death, number_of_deaths = deaths_all_causes) %>%
+  filter(place_of_death != "All") %>%
+  group_by(year) %>%
+  mutate(
+    total_deaths = sum(number_of_deaths)
+  ) %>%
+  ungroup %>%
+  mutate(proportion_of_total = number_of_deaths / total_deaths)
+
+merged_proportions_of_deaths_by_place <-
+  bind_rows(
+    proportions_of_historical_deaths_by_place,
+    proportions_of_current_deaths_by_place
+  )
+```
+
 # Visualisations
@@
 only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(min(merged_deaths_age$date_w_c))}.")
+only_2021_text <- glue::glue("Data for 2021 only include weeks 1-{format(max(merged_deaths_age$week_number[merged_deaths_age$year==2021]))} (w/c {format(max(merged_deaths_age$date_w_c))}).")
 
@@
 ```
+
+# Proportion home deaths over time
+
+```{r}
+figure_proportion_home_deaths_over_time <-
+  merged_proportions_of_deaths_by_place %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point() +
+  geom_line(size = 1) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  scale_x_continuous(breaks = 2015:2021) +
+  scale_y_continuous(label = scales::percent) +
+  scale_colour_viridis_d(option="C") +
+  theme_minimal() +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(
+    x = "Year", y= "Proportion of yearly deaths (%)",
+    colour = "",
+    subtitle = paste0("* ",only_2021_text),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time",.x), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
+)
+
+figure_proportion_home_deaths_over_time
+```
+
+# Number of deaths visualisations
 
@@
   )
+
 ```
@@
 
-# figure_home_deaths_and_cases_scotland <-
+figure_weekly_deaths_to_cases_part_1 <-
   merged_deaths_overall %>%
@@
   geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  geom_line(aes(y = weekly_positive * ratio_max_weekly_deaths_to_cases), colour="purple") +
+  # geom_line(aes(y = weekly_positive * ratio_max_weekly_deaths_to_cases), colour="purple") +
   geom_vline(xintercept = 54, linetype = "dotted") +
   scale_x_continuous(breaks = seq(0,60,by=10)) +
-  scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly cases")) +
+  # scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly cases")) +
   # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
@@
                       ) +
-  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
+  # annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
@@
 
+figure_weekly_deaths_to_cases_part_2 <-
   merged_deaths_overall %>%
@@
   ggplot(aes(x=week_number_run_over, y=weekly_positive)) +
-  geom_line() +
-  theme_minimal() +
+  geom_line(size=1) +
+  geom_vline(xintercept = 54, linetype = "dotted") +
+  theme_minimal() +
+  facet_grid(~ref_area) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
   scale_y_continuous(labels = function(x) format(x, big.mark=",")) +
@@
     x = "Week number", y = "COVID-19 positive cases (N)",
-    subtitle = date_range_text,
-    caption = paste0(c(vertical_line_text, source_phs_text), collapse="\n")
-  )
-
-# ## figure for a bit less than half A4 page is approx 14*11cm
-# walk(
-#   .x = c(".pdf",".svg",".png"),
-#   .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 14, height = 11, units = "cm", dpi = 300)
-# )
-# 
-# figure_deaths_at_home
+    subtitle = "Note the different scale for positive cases!",
+    caption = paste0(c(source_phs_text), collapse="\n")
+  )
+
+figure_weekly_deaths_to_cases <- wrap_plots(figure_weekly_deaths_to_cases_part_1,figure_weekly_deaths_to_cases_part_2, ncol = 1)
+
+## figure for a bit less than half A4 page is approx 14*11cm
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_weekly_deaths_to_cases",.x), plot = figure_weekly_deaths_to_cases, width = 15, height = 15, units = "cm", dpi = 300)
+)
+
+figure_weekly_deaths_to_cases
 ```

### 2021-02-22 covid_nrs_place_of_death.Rmd
Added plots of covid cases per 100,000 on a map by HB/LA - they match the home deaths it seems!

@@
 This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
+
 However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!
+
+Note also that the `crude_rate_positive_per100k` represents the cumulative rate!
 
@@
 ```{r}
-areas_relatively_higher_number_home_deaths <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
-home_deaths_ratio <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
+areas_relatively_higher_number_home_deaths_text <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
+home_deaths_ratio_text_text <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
 
@@
   labs(x="", y="",
-       title = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority",
-       subtitle = "A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.",
-       caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
+       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority", width = 50),
+       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
        fill = "Ratio"
@@
   labs(x="", y="",
-       title = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board",
-       subtitle = "A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.",
-       caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
+       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board", width=50),
+       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
        fill = "Ratio"
@@
 
-```{r}
-
-```
+## Per capita cases of COVID-19 by region
+
+```{r}
+areas_relatively_higher_number_cases <- "Lighter areas represent areas with relatively higher number of COVID-19 cases per 100,000 population in 2020"
+```
+
+### LA
+
+```{r, fig.height=12, fig.width=8}
+figure_map_per_capita_covid_cases_2020_la <-
+  weekly_covid_cases_la %>%
+  filter(year == 2020 & week_number == 53) %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B") +
+  theme_minimal() +
+  theme(axis.text = element_blank()) +
+  labs(x="", y="",
+       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Local Authority",width=50),
+       subtitle = "",
+       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
+       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
+       ) +
+  NULL
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la",.x), plot = figure_map_per_capita_covid_cases_2020_la, width = 9, height = 12, dpi = 300)
+)
+
+figure_map_per_capita_covid_cases_2020_la
+```
+
+### HB
+
+```{r, fig.height=12, fig.width=8}
+figure_map_per_capita_covid_cases_2020_hb <-
+  weekly_covid_cases_hb %>%
+  filter(year == 2020 & week_number == 53) %>%
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B") +
+  theme_minimal() +
+  theme(axis.text = element_blank(), aspect.ratio = 1.4) +
+  labs(x="", y="",
+       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
+       subtitle = "",
+       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
+       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
+       ) +
+  NULL
+
+walk(
+  .x = c(".pdf",".svg",".png"),
+  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb",.x), plot = figure_map_per_capita_covid_cases_2020_hb, width = 9, height = 12, dpi = 300)
+)
+
+figure_map_per_capita_covid_cases_2020_hb
+```

### 2021-02-25 covid_nrs_place_of_death.Rmd
Changed to use calibri font, but broken most outputs in process. fix display of pdfs! (by embedding calibri font, I assume?)

@@
 areas_relatively_higher_number_home_deaths_text <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
-home_deaths_ratio_text_text <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
+home_deaths_ratio_text <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
 

### 2021-02-25 download_latest_data_and_render_reports.R

@@ -14,6 +14,7 @@ library(lubridate)  # dealing with dates
 library(curl)
 library(knitr)
 library(rmarkdown)
+library(extrafont)
 
 # Load table keeping track of latest data ---------------------------------
 
@@ -179,19 +180,14 @@ download_shapefile(url = health_boards_shapefile_url, directory = dir_map_data, 
 download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")
 download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
 
+# Import Calibri font -----------------------------------------------------
+
+if (!any(fonttable()$FullName=="Calibri")) font_import(pattern = "calibri", prompt = FALSE)  # only run first time if Calibri not yet imported
+
 # Render report with latest data ------------------------------------------
 
 dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
-
-# TODO: remove this section, params no longer used
-## this line is also useful for testing - just run it and it pre-populates the
-## params list for the rmd!
-# params <- list( # pass the table of data sources
-#   data_sources = table_of_data_sources,
-#   historical_data_sources = table_of_historical_data_sources,
-#   case_trends_data_sources = table_of_case_trends_data_sources
-# )
 
 render(
   input = "./covid_nrs_place_of_death.Rmd",

### 2021-03-02 covid_nrs_place_of_death.Rmd
added a few more plots & amended existing plots

--- 2021-02-22 covid_nrs_place_of_death.Rmd
@@
 library(patchwork)  # for assembling plots
+library(extrafont)  # for working with fonts
 ```
@@
 table_of_data_sources %>%
-  kable(caption = "Latest data used in this report.")
+  knitr::kable(caption = "Latest data used in this report.")
 ```
@@
   isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
-  computed_dates <- ISOweek::ISOweek2date(isoweek_string)
+  # computed_dates <- ISOweek::ISOweek2date(isoweek_string)
+  computed_dates <- rep(NA_Date_,times = length(week_number))
+  computed_dates[which(!is.na(week_number))] <- ISOweek::ISOweek2date(isoweek_string[which(!is.na(week_number))])  # return NA instead of failing with ISOweek!
   ## invalid weeks will result in a computed date in a year after the requested year...
   ## a valid week can result in a date that starts in the year before!
-  if (any(year_number<year(ymd(computed_dates)))) stop("invalid week number provided!")
+  if (any(year_number<year(ymd(computed_dates)), na.rm = TRUE)) stop("invalid week number provided!")
   return(computed_dates)
@@
   geom_label_repel() +
-  theme_minimal() +
+  # theme_minimal() +
   labs(
@@
 ```{r}
-theme_set(theme_minimal(base_size = 16) +
+loadfonts()  # load imported fonts
+# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
+
+# theme_set(theme_minimal(base_size = 16) +
+theme_set(theme_minimal(base_size = 12) +
             theme(panel.grid.minor = element_blank(),
                   axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
-                  axis.title.x = element_text(margin = margin(20, 0, 0, 0))
+                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
+                  text=element_text(family="Calibri")
             ))
 
+## constants for use in graphs & annotations
+weeks_available_2021 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2021])
+most_recent_date_available_2021 <- max(merged_deaths_age$date_w_c)
+earliest_date <- min(merged_deaths_age$date_w_c)
+
 date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")
@@
 
-only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(min(merged_deaths_age$date_w_c))}.")
-only_2021_text <- glue::glue("Data for 2021 only include weeks 1-{format(max(merged_deaths_age$week_number[merged_deaths_age$year==2021]))} (w/c {format(max(merged_deaths_age$date_w_c))}).")
+only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
+only_2021_text <- glue::glue("Data for 2021 only include weeks 1-{format(weeks_available_2021)} (w/c {format(most_recent_date_available_2021)}).")
 
@@
 vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."
+
+```
+
+## Function to save files to multiple formats
+
+Saving to multiple formats is easy enough - the trouble is with saving PDF files with custom fonts, since they need to be embedded in the .pdf file. This requires:
+
+* telling R where to load the font from, using `extrafont::font_import()` - this was done in a previous step
+* registering fonts with R using `extrafont::loadfonts()` - note: this isn't strictly necessary, since once imported, `extrafont` keeps track of fonts to load when the `extrafont` package is loaded
+* ~~telling R where the `gs` executable is, i.e. the installation of ghostscript on the current machine~~
+* ~~embedding the fonts in the `.pdf` files once they've been written, using `extrafont::embed_fonts()`~~
+* instead of manually embedding fonts (which resulted in garbled text on my setup!), just use `device=device_cairo` as an argument to `ggsave()` when saving `.pdf` files and it just works
+
+```{r}
+save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {
+
+walk(
+  .x = extensions,
+  .f = ~ggsave(filename = paste0(filename, .x), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (.x==".pdf") cairo_pdf else NULL)
+)
+
+# TODO: remove below bit, not needed when using device=device_cairo
+# if (any(str_detect(string = tolower(extensions), pattern = "pdf"))) {
+#   pdf_filename <- paste0(filename, extensions[which(any(str_detect(string = tolower(extensions), pattern = "pdf")))])
+#   embed_fonts(file = pdf_filename)  # overwrites file with one with font embedded
+# }
+# invisible(filename)  # return invisibly
+}
+
 ```
@@
   scale_colour_viridis_d(option="C") +
-  theme_minimal() +
+  # theme_minimal() +
   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time",.x), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
 
@@
   geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 1000)) +
-  theme_minimal() +
+  # theme_minimal() +
   labs(x = "",
@@
   )
-
 ```
@@
   scale_y_continuous(labels = function(x) format(x,big.mark=",")) +
-  theme_minimal() +
+  # theme_minimal() +
   labs(x = "Age",
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_annual_deaths_by_age_group_2020",.x), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_annual_deaths_by_age_group_2020"), extensions = c(".pdf",".svg",".png"), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300, units = "in")
 
@@
 
-## Recorded deaths in Scotland, by week number & location
+## Recorded deaths in Scotland, by week number & location, with annotations (after DH's plots)
 
@@
 
-figure_place_of_death <-
+figure_place_of_death_after_dh <-
   merged_deaths_overall %>%
@@
   facet_grid(~place_of_death) +
-  theme_minimal() +
+  # theme_minimal() +
   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_place_of_death",.x), plot = figure_place_of_death, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death_after_dh"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death_after_dh, width = 12, height = 9, dpi = 300, units = "in")
+
+figure_place_of_death_after_dh
+```
+
+### Place of death with historic range, updated version
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
+annotations_figure_place_of_death_explaining_historic_range <- tribble(
+   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
+  "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
+  "curve", "Hospital", NA, 29, 830, 32, 550,
+  "text", "Hospital", "Historic mean", 22, 800, NA, NA,
+  "curve", "Hospital", NA, 22, 780, 26, 510
+) %>% 
+  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+
+figure_place_of_death <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
+  geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death, width = 28, height = 14, dpi = 300, units = "cm")
 
@@
 
-## All of Scotland, deaths at home
+## All of Scotland, deaths at home, for policy briefing
 
@@
 
-figure_deaths_at_home <-
+figure_deaths_at_home_policy_briefing <-
   merged_deaths_overall %>%
@@
                       ) +
-  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
+  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 21, y = 575, hjust=0) +
   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
@@
        colour = "Cause of death"
-       )
+       ) +
+  NULL
 
 ## figure for a bit less than half A4 page is approx 14*11cm
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 14, height = 11, units = "cm", dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_policy_briefing"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_policy_briefing, width = 14, height = 11, units = "cm", dpi = 300)
+
+figure_deaths_at_home_policy_briefing
+```
+
+## All of Scotland, deaths at home - date axis
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+figure_deaths_at_home <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60"), plot.margin = margin(5,15,5,5)) +
+  labs(x = "Date w/c",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 10, year_number = 2020), y = 100, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 10, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 14, year_number = 2020), y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic mean", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  ## annotate year
+  geom_text(aes(x=ymd("2021-01-07"), y=30, label = "2021"), hjust = 0, colour = "grey40") +
+  NULL
+
+## figure for a bit less than half A4 page is approx 14*11cm
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home, width = 18, height = 15, units = "cm", dpi = 300)
 
 figure_deaths_at_home
+```
+
+## All of Scotland, deaths at home - week number
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+figure_deaths_at_home_wwek_number <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = week_number_run_over)) +
+  geom_vline(xintercept = 53.5, linetype = "dotted", colour="grey40", show.legend = FALSE) +  # line should be between last week 2020 and 1st week 2021
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  geom_text(aes(x = 10, y = 100, label = "Historic range (min & max)"), colour = "grey20") +
+  geom_curve(aes(x = 10, xend = 14, y = 110, yend = 250), curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  geom_text(aes(x = 20, y = 50, label = "Historic mean"), colour = "grey20") +
+  geom_curve(aes(x = 20, xend = 25, y = 60, yend = 270), curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  ## annotate year
+  geom_text(aes(x=54, y=30, label = "2021"), hjust = 0, colour = "grey40") +
+  NULL
+
+## figure for a bit less than half A4 page is approx 14*11cm
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_wwek_number"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_wwek_number, width = 18, height = 15, units = "cm", dpi = 300)
+
+figure_deaths_at_home_wwek_number
 ```
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020",.x), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300, units = "in")
 
 figure_deaths_at_home_2020
+```
+
+## Deaths in hospital, date axis
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+figure_deaths_in_hospital <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Hospital")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60"), plot.margin = margin(5,15,5,5)) +
+  labs(x = "Date w/c",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  annotate(geom = "text", x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), y = 850, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom = "curve", x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 32, year_number = 2020), y = 830, yend = 550, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), y = 800, label = "Historic mean", colour = "grey20") +
+  annotate(geom = "curve", x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 26, year_number = 2020), y = 780, yend = 510, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  ## annotate year
+  geom_text(aes(x=ymd("2021-01-07"), y=30, label = "2021"), hjust = 0, colour = "grey40") +
+  NULL
+
+## figure for a bit less than half A4 page is approx 14*11cm
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_in_hospital"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_in_hospital, width = 18, height = 15, units = "cm", dpi = 300)
+
+figure_deaths_in_hospital
 ```
@@
   facet_wrap(~ref_area, scales = "free_y") +
-  theme_minimal() +
+  # theme_minimal() +
   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_health_board",.x), plot = figure_deaths_at_home_by_health_board, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_health_board"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_health_board, width = 12, height = 9, dpi = 300, units = "in")
 
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_health_board",.x), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_health_board"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300, units = "in")
 
@@
 
-# merged_deaths_la %>%
-#   filter(place_of_death == "Home & other non-institution") %>%
-#   filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
-#   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-#                names_to = "deaths_type", values_to = "deaths") %>% 
-#   ggplot(aes(x = week_number_run_over)) +
-#   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-#   geom_line(aes(y = deaths, colour = deaths_type)) +
-#   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-#   geom_vline(xintercept = 54, linetype = "dotted") +
-#   scale_x_continuous(breaks = seq(0,60,by=10)) +
-#   facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
-#   theme_minimal() +
-#   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-#                       labels = c("All deaths", 
-#                                  "Non Covid-19 deaths"),
-#                       guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-#                       ) +
-#   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-#   labs(x = "Week number",
-#        y = "Deaths (N)",
-#        subtitle = date_range_text,
-#        caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
-#        colour = "Cause of death"
-#        )
-
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_local_authority",.x), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_local_authority"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300, units = "in")
 
@@
        )
-# 
-# merged_deaths_la %>%
-#   filter(place_of_death == "Home & other non-institution") %>%
-#   filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
-#   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-#                names_to = "deaths_type", values_to = "deaths") %>% 
-#   ggplot(aes(x = week_number_run_over)) +
-#   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-#   geom_line(aes(y = deaths, colour = deaths_type)) +
-#   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-#   geom_vline(xintercept = 54, linetype = "dotted") +
-#   scale_x_continuous(breaks = seq(0,60,by=10)) +
-#   facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
-#   theme_minimal() +
-#   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-#                       labels = c("All deaths", 
-#                                  "Non Covid-19 deaths"),
-#                       guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-#                       ) +
-#   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-#   labs(x = "Week number",
-#        y = "Deaths (N)",
-#        subtitle = date_range_text,
-#        caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
-#        colour = "Cause of death"
-#        )
-
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_local_authority",.x), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300)
-)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_local_authority"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300, units = "in")
 
@@
   facet_wrap(~age) +
-  theme_minimal() +
+  # theme_minimal() +
   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_age",.x), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_age"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300, units = "in")
 
@@
   facet_grid(~sex) +
-  theme_minimal() +
+  # theme_minimal() +
   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_sex",.x), plot = figure_deaths_at_home_by_sex, width = 12, height = 9, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_sex"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_sex, width = 12, height = 9, dpi = 300, units = "in")
 
@@
   geom_vline(xintercept = 54, linetype = "dotted") +
-  theme_minimal() +
+  # theme_minimal() +
   facet_grid(~ref_area) +
@@
 ## figure for a bit less than half A4 page is approx 14*11cm
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_weekly_deaths_to_cases",.x), plot = figure_weekly_deaths_to_cases, width = 15, height = 15, units = "cm", dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_weekly_deaths_to_cases"), extensions = c(".pdf",".svg",".png"), plot = figure_weekly_deaths_to_cases,width = 15, height = 15, units = "cm", dpi = 300)
 
@@
 ```{r}
-areas_relatively_higher_number_home_deaths <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
-home_deaths_ratio <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
+areas_relatively_higher_number_home_deaths_text <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
+home_deaths_ratio_text <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
 
@@
   ) %>%
+  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
   left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
-  ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
+  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
+  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths, maximum_ratio_between_hb_and_la_home_deaths)) +
-  theme_minimal() +
-  theme(axis.text = element_blank()) +
+  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
   labs(x="", y="",
-       title = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority",
-       subtitle = "A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.",
-       caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
-       fill = "Ratio"
+       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority", width = 50),
+       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       # fill = "Ratio"
+       fill = "Home death increase"
        ) +
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 9, height = 12, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 9, height = 12, dpi = 300, units = "in")
 
@@
   ) %>%
+  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
   left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
-  ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
+  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
+  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths, maximum_ratio_between_hb_and_la_home_deaths)) +
-  theme_minimal() +
-  theme(axis.text = element_blank(), aspect.ratio = 1.4) +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
   labs(x="", y="",
-       title = "Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board",
-       subtitle = "A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.",
-       caption = paste0(c(areas_relatively_higher_number_home_deaths, home_deaths_ratio, source_nrs_text), collapse="\n"),
-       fill = "Ratio"
+       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board", width=50),
+       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       # fill = "Ratio"
+       fill = "Home death increase"
        ) +
@@
 
-walk(
-  .x = c(".pdf",".svg",".png"),
-  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb",.x), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 9, height = 12, dpi = 300)
-)
+save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 9, height = 12, dpi = 300, units = "in")
 
@@
 
-```{r}
-
-```
+## Per capita cases of COVID-19 by region
+
+```{r}
+areas_relatively_higher_number_cases <- "Lighter areas represent areas with relatively higher number of COVID-19 cases per 100,000 population in 2020"
+```
+
+### LA
+
+```{r, fig.height=12, fig.width=8}
+figure_map_per_capita_covid_cases_2020_la <-
+  weekly_covid_cases_la %>%
+  filter(year == 2020 & week_number == 53) %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B") +
+  # theme_minimal() +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  labs(x="", y="",
+       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Local Authority",width=50),
+       subtitle = "",
+       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
+       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
+       ) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la, width = 9, height = 12, dpi = 300, units = "in")
+
+figure_map_per_capita_covid_cases_2020_la
+```
+
+### HB
+
+```{r, fig.height=12, fig.width=8}
+figure_map_per_capita_covid_cases_2020_hb <-
+  weekly_covid_cases_hb %>%
+  filter(year == 2020 & week_number == 53) %>%
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B") +
+  # theme_minimal() +
+  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
+  labs(x="", y="",
+       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
+       subtitle = "",
+       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
+       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
+       ) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb, width = 9, height = 12, dpi = 300, units = "in")
+
+figure_map_per_capita_covid_cases_2020_hb
+```
+
+## Merged maps of home death ratios & covid cases
+
+### HB
+
+```{r}
+# TODO: work out where all the extra space comes from around plot & legends
+
+figure_map_home_deaths_ratios_2020_and_per_capita_cases <-
+  wrap_plots(
+    figure_map_ratio_home_deaths_2020_to_historical_hb + 
+      labs(
+        caption = source_nrs_text,   # keep source captions only
+        fill = str_wrap("Home deaths increase over historic mean",width = 20),
+        subtitle = NULL,
+        title = str_wrap("Increased home deaths 2020", width = 20),
+        x = NULL,
+        y = NULL
+        ) +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+    ,
+      # theme(legend.position = "left"),
+    figure_map_per_capita_covid_cases_2020_hb + 
+      labs(
+        caption = source_phs_text,
+        title = str_wrap("COVID cases 2020", width = 20),
+        x = NULL,
+        y = NULL
+        ) + 
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+    ,
+    nrow = 1  # force to same row  
+  )
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_home_deaths_ratios_2020_and_per_capita_cases"), extensions = c(".pdf",".svg",".png"), plot = figure_map_home_deaths_ratios_2020_and_per_capita_cases, width = 20, height = 10, dpi = 300, units = "cm")
+```

### 2021-03-03 covid_nrs_place_of_death.Rmd
updated the table of annual deahts & comparing with past rates

@@
 
+# Constants about latest data available
+
+```{r}
+## constants for use in graphs & annotations
+weeks_available_2021 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2021])
+most_recent_date_available_2021 <- max(merged_deaths_age$date_w_c)
+earliest_date <- min(merged_deaths_age$date_w_c)
+```
+
 # Calculate case counts: weekly counts to match weekly death reporting
@@
 ```{r}
-## this is fairly ad-hoc... but it works!
-table_comparing_rates <-
+table_current_deaths_compared_to_past <-
   bind_rows(
@@
       filter(age == "all" &
-               sex == "all" &
-               str_starts(place_of_death, "Home")) %>% group_by(year, week_number) %>%
+               sex == "all"
+             ) %>% 
+      group_by(year, place_of_death, week_number) %>%
       summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+      left_join(
+        past_deaths_overall %>% filter(year != "average") %>% group_by(week_number, place_of_death) %>% summarise(
+                                           average_weekly_deaths = sum(number_of_deaths) / 5,
+                                           .groups = "drop"
+                                         ),
+        by = c("week_number", "place_of_death")
+      ) %>% 
+      group_by(place_of_death) %>%
+      summarise(
+        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
         deaths_all_causes = sum(deaths_all_causes),
@@
         deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
         .groups = "drop"
-      ) %>% left_join(
-        past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
-                                         year != "average") %>% group_by(week_number) %>% summarise(
-                                           average_weekly_deaths = sum(number_of_deaths) / 5,
-                                           .groups = "drop"
-                                         ),
-        by = "week_number"
-      ) %>% summarise(
-        period = "all available data",
-        deaths_all_causes = sum(deaths_all_causes),
-        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-        past_annual_deaths = sum(average_weekly_deaths)
       )
@@
     ## 2020 only
-    bind_cols(
     merged_deaths_overall %>%
-        filter(
-          age == "all" &
-            sex == "all" & str_starts(place_of_death, "Home") & year == 2020
-        ) %>%
+      filter(age == "all" &
+               sex == "all" &
+               year == 2020
+             ) %>% 
+      group_by(place_of_death) %>%
       summarise(
@@
         deaths_all_causes = sum(deaths_all_causes),
-          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
-        ),
-      past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
-                                       year != "average") %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
-    ),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_overall %>% 
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+      by = "place_of_death"
+    )
+    ,
     ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
-    bind_cols(
     merged_deaths_overall %>%
@@
           sex == "all" &
-            str_starts(place_of_death, "Home") &
           year == 2020 & week_number >= 12
       ) %>%
+      group_by(place_of_death) %>%
       summarise(
-          period = "2020, from wk12",
+        period = "2020, week 12 - week 53",
         deaths_all_causes = sum(deaths_all_causes),
-          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
-        ),
-      past_deaths_overall %>% filter(
-        week_number >= 12 &
-          str_starts(place_of_death, "Home") &
-          year != "average"
-      ) %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      past_deaths_overall %>% 
+        filter(week_number >= 12 & year != "average") %>%
+        group_by(place_of_death) %>%
+        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
+      by = "place_of_death"
     )
@@
   mutate(
-    ratio_deaths_all_causes_to_historical = deaths_all_causes / past_annual_deaths,
-    ratio_non_covid_deaths_to_historical = deaths_non_covid / past_annual_deaths,
+    ratio_deaths_all_causes_to_historical = deaths_all_causes / mean_annual_deaths_2015_2019,
+    ratio_non_covid_deaths_to_historical = deaths_non_covid / mean_annual_deaths_2015_2019,
     proportion_covid_deaths = deaths_covid_related / deaths_all_causes
@@
 
-write_csv(x = table_comparing_rates, file = file.path(dir_outputs, "annual_death_rates_comparison_with_historical_deaths.csv"))
-
-table_comparing_rates %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
+write_csv(x = table_current_deaths_compared_to_past, file = file.path(dir_outputs, "annual_death_rates_comparison_with_historical_deaths.csv"))
+
+table_current_deaths_compared_to_past %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
 ```
@@
             ))
-
-## constants for use in graphs & annotations
-weeks_available_2021 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2021])
-most_recent_date_available_2021 <- max(merged_deaths_age$date_w_c)
-earliest_date <- min(merged_deaths_age$date_w_c)
 

### 2021-03-08 covid_nrs_place_of_death.Rmd
changed it so that we don't re-download shapefiles every time;
fixed order of .rmd script;

added a cumulative deaths graph, needs more work to be clear

@@
 
-# Constants about latest data available
-
-```{r}
-## constants for use in graphs & annotations
-weeks_available_2021 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2021])
-most_recent_date_available_2021 <- max(merged_deaths_age$date_w_c)
-earliest_date <- min(merged_deaths_age$date_w_c)
-```
-
 # Calculate case counts: weekly counts to match weekly death reporting
@@
   compute_ratio_of_annual_deaths_to_historical
+```
+
+# Constants about latest data available
+
+```{r}
+## constants for use in graphs & annotations
+weeks_available_2021 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2021])
+most_recent_date_available_2021 <- max(merged_deaths_age$date_w_c)
+earliest_date <- min(merged_deaths_age$date_w_c)
 ```
@@
 
+# Cumulative deaths
+
+Note: I've left this unfinished, it doesn't tell a better story than the weekly figures!
+
+```{r}
+text_non_covid_deaths_equivalent_to_all_cause_before_wk12 <- "Note: all-cause deaths and non-covid deaths were considered equivalent before 2020, week 12"
+
+figure_cumulative_deaths_by_place <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(deaths_non_covid = if_else(condition = is.na(deaths_non_covid), true = deaths_all_causes, false = deaths_non_covid)) %>%  # replace NA in non-covid deaths with all cause deaths - before covid, non-covid deaths are equivalent to all deaths
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>%
+  mutate(deaths_type = case_when(
+    deaths_type == "deaths_all_causes" ~ "All-cause deaths",
+    deaths_type == "deaths_non_covid" ~ "Non-COVID deaths",
+    TRUE ~ NA_character_
+  )) %>%
+  group_by(place_of_death, deaths_type) %>%
+  mutate(
+    cumulative_deaths = cumsum(replace_na(deaths, replace = 0)),
+    cumulative_historical_mean = cumsum(replace_na(deaths_mean, replace = 0)),
+    cumulative_historical_min = cumsum(replace_na(deaths_min, replace = 0)),
+    cumulative_historical_max = cumsum(replace_na(deaths_max, replace = 0))
+    ) %>%
+  ungroup %>%
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  geom_ribbon(aes(ymin = cumulative_historical_min, ymax = cumulative_historical_max, fill = place_of_death), alpha = 0.4) +
+  geom_line(aes(y = cumulative_historical_mean, colour = place_of_death), linetype = "dashed", alpha = 0.8, show.legend = FALSE) +
+  geom_line(aes(y = cumulative_deaths, colour = place_of_death)) +
+  facet_wrap(~deaths_type) +
+  scale_colour_viridis_d(option = "C") +
+  scale_fill_viridis_d(option = "C") +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Cumulative deaths (N)",
+       title = "Cumulative deaths by place of death",
+       # subtitle = paste0(c(only_2020_text,light_shaded_area_shows_difference),collapse="\n"),
+       caption = paste0(c(text_non_covid_deaths_equivalent_to_all_cause_before_wk12,death_registrations_text,source_nrs_text), collapse="\n")
+  ) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_cumulative_deaths_by_place"), extensions = c(".pdf",".svg",".png"), plot = figure_cumulative_deaths_by_place, width = 15, height = 12, dpi = 300, units = "cm")
+
+figure_cumulative_deaths_by_place
+```
+
 # Number of deaths visualisations
-
-## Number of deaths by age group
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-## graph format copied from: https://github.com/davidhen/nrs_covid_stats/
-weekly_deaths_overall %>% 
-  filter(sex == "all" & place_of_death == "All") %>% 
-  filter(age != "all") %>%
-  group_by(age) %>% 
-  summarise(n = sum(deaths_all_causes)) %>%
-  mutate(pct = round(n/sum(n), 3)) %>% 
-  ggplot(aes(age, n)) +
-  geom_col(fill = "#4477AA") +
-  geom_text(aes(label = scales::percent(pct, accuracy = 0.1), y = n + 1000)) +
-  # theme_minimal() +
-  labs(x = "",
-       y = "",
-       title = "Deaths Associated with COVID-19 by Age Group",
-       subtitle = date_range_text,
-       caption = paste0(c(date_range_text,source_nrs_text), collapse="\n")
-  )
-```
 
@@
   scale_y_continuous(labels = function(x) format(x,big.mark=",")) +
-  # theme_minimal() +
   labs(x = "Age",

### 2021-03-08 download_latest_data_and_render_reports.R

@@ -175,10 +175,10 @@ health_boards_uk_wide_url <- "https://github.com/tomwhite/covid-19-uk-data/files
 # https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
 la_shapefile_url <- "https://opendata.arcgis.com/datasets/910f48f3c4b3400aa9eb0af9f8989bbe_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"
 
-
-download_shapefile(url = health_boards_shapefile_url, directory = dir_map_data, subdirectory = "hb")
-download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")
-download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
+# TODO: these have been downloaded so don't need to be downloaded again... change to check if they exist already first?
+# download_shapefile(url = health_boards_shapefile_url, directory = dir_map_data, subdirectory = "hb")
+# download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")
+# download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
 
 # Import Calibri font -----------------------------------------------------
 

### 2021-03-10 covid_nrs_place_of_death.Rmd
added a few more plots, and split off sections for paper to a separate file

@@
 library(extrafont)  # for working with fonts
+library(openxlsx)  # for creating xlsx files
 ```
@@
 
+## Weekly deaths including past weekly deaths
+
+For this dataset, we simply `bind_rows` weekly deaths by year and place of death with past weekly deaths by year and location.
+Note that past data for years with 52 weeks includes an entry of 0 deaths for week 53, so we'll remove them, otherwise the date computation is thrown off!
+
+```{r}
+weekly_deaths_current_and_past <-
+  bind_rows(
+    weekly_deaths_overall %>%
+    filter(sex=="all" & age=="all" & place_of_death!="All") %>%  # keep only aggregated values over sex & age
+    select(year, week_number, week_number_run_over, place_of_death, matches("deaths"), -deaths_nrs_past_average_all_causes)
+    ,
+    past_deaths_overall %>%
+      rename(deaths_all_causes = number_of_deaths) %>%
+      filter(place_of_death!="All" & year!="average") %>%  # remove aggregated entries
+      mutate(year = as.integer(year)) %>%
+      filter(!(year %in% 2016:2019 & week_number==53))  # remove week 53 entries from years 2016-2019
+  ) %>%
+  mutate(date_w_c = compute_start_date_from_week_number(week_number = week_number, year_number = year))
+```
+
 # Annual deaths for 2020 & 2021
@@
 
+## Including non-COVID deaths
+
+Note: non-covid deaths are defined as all-cause deaths minus covid-related deaths.
+Otherwise the number of non-covid deaths only includes non-covid deaths once covid-related deaths started being reported in 2020 (week 12).
+
+```{r}
+proportions_of_current_non_covid_deaths_by_place <-
+  total_deaths_overall %>%
+  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+  select(year, place_of_death, number_of_deaths = deaths_non_covid) %>%
+  filter(place_of_death != "All") %>%
+  group_by(year) %>%
+  mutate(
+    total_deaths = sum(number_of_deaths)
+  ) %>%
+  ungroup %>%
+  mutate(proportion_of_total = number_of_deaths / total_deaths)
+
+merged_proportions_of_non_covid_deaths_by_place <-
+  bind_rows(
+    proportions_of_historical_deaths_by_place,  # all historical deaths are non-covid!
+    proportions_of_current_non_covid_deaths_by_place
+  )
+```
+
 # Visualisations
@@
 
-# theme_set(theme_minimal(base_size = 16) +
 theme_set(theme_minimal(base_size = 12) +
@@
   .x = extensions,
-  .f = ~ggsave(filename = paste0(filename, .x), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (.x==".pdf") cairo_pdf else NULL)
+  .f = ~function(ext) {
+    if (tolower(ext)==".pdf") device <- cairo_pdf
+
+  ggsave(filename = paste0(filename, .x), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (.x==".pdf") cairo_pdf else NULL)
+  }
 )
@@
 text_non_covid_deaths_equivalent_to_all_cause_before_wk12 <- "Note: all-cause deaths and non-covid deaths were considered equivalent before 2020, week 12"
+
+text_annotation_cunulative_all_cause_deaths <- "Hospital all-cause deaths were within the historical range; home & care home deaths exceeded historical ranges."
+
+# TODO: these figures over- and underestimate yearly cumulative deaths (because they use weekly min/max, as opposed to min/max of cumulative deaths)
 
@@
   left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
-  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
   ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
   theme(axis.text = element_blank(), panel.grid = element_blank()) +
   labs(x="", y="",
-       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority", width = 50),
-       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
-       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       # fill = "Ratio"
-       fill = "Home death increase"
+       title = str_wrap("Increased home deaths in 2020 relative to historical mean 2015-2019, by Local Authority", width=50),
+       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease"
        ) +
@@
 
-save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 9, height = 12, dpi = 300, units = "in")
+save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 12, height = 14, dpi = 300, units = "cm")
 
@@
 
+#### Grayscale version of the same
+
+```{r}
+figure_map_ratio_home_deaths_2020_to_historical_la_bw <-
+  figure_map_ratio_home_deaths_2020_to_historical_la +
+  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) scales::percent(x, accuracy = 1)) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la_bw, width = 12, height = 14, dpi = 300, units = "cm")
+```
+
 ### HB
@@
 ```{r, fig.height=12, fig.width=8}
-# TODO: work out the size of these plots
 figure_map_ratio_home_deaths_2020_to_historical_hb <-
@@
   scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
-  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
+  theme(
+    axis.text = element_blank(), 
+    panel.grid = element_blank(), 
+    aspect.ratio = 1.4,  # needed for Scotland HB plot
+    plot.margin = margin(5,5,5,5),
+    # legend.position = c(1.2,0.4)  # this was manually tweaked until just right  # problem - edge gets cut
+    ) +
   labs(x="", y="",
-       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board", width=50),
-       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
-       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       title = str_wrap("Increased home deaths in 2020 relative to historical mean 2015-2019, by Health Board", width=50),
+       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(source_nrs_text), collapse="\n"),
        # fill = "Ratio"
-       fill = "Home death increase"
+       fill = "Home death\nincrease"
        ) +
@@
 
-save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 9, height = 12, dpi = 300, units = "in")
+save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 12, height = 14, dpi = 300, units = "cm")
 
 figure_map_ratio_home_deaths_2020_to_historical_hb
+```
+
+#### Grayscale version of the same
+
+```{r}
+figure_map_ratio_home_deaths_2020_to_historical_hb_bw <-
+  figure_map_ratio_home_deaths_2020_to_historical_hb +
+  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) scales::percent(x, accuracy = 1)) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb_bw, width = 12, height = 14, dpi = 300, units = "cm")
 ```
@@
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(option = "B") +
-  # theme_minimal() +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
   theme(axis.text = element_blank(), panel.grid = element_blank()) +
   labs(x="", y="",
-       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Local Authority",width=50),
+       title = str_wrap("Total COVID-19 cases in 2020 per 100,000 population, by Local Authority",width=50),
        subtitle = "",
-       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
-       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
+       caption = paste0(c(source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
        ) +
@@
 
-save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la, width = 9, height = 12, dpi = 300, units = "in")
+save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la, width = 12, height = 14, dpi = 300, units = "cm")
 
 figure_map_per_capita_covid_cases_2020_la
+```
+
+#### Grayscale version of the same
+
+```{r}
+figure_map_per_capita_covid_cases_2020_la_bw <-
+  figure_map_per_capita_covid_cases_2020_la +
+  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) format(x, big.mark=",")) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la_bw, width = 12, height = 14, dpi = 300, units = "cm")
 ```
@@
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(option = "B") +
-  # theme_minimal() +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
   theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
   labs(x="", y="",
-       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
+       title = str_wrap("Total cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
        subtitle = "",
-       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
-       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
+       caption = paste0(c(source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
        ) +
@@
 
-save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb, width = 9, height = 12, dpi = 300, units = "in")
+save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb, width = 12, height = 14, dpi = 300, units = "cm")
 
 figure_map_per_capita_covid_cases_2020_hb
+```
+
+#### Grayscale version of the same
+
+```{r}
+figure_map_per_capita_covid_cases_2020_hb_bw <-
+  figure_map_per_capita_covid_cases_2020_hb +
+  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) format(x, big.mark=",")) +
+  NULL
+
+save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb_bw, width = 12, height = 14, dpi = 300, units = "cm")
 ```
@@
 ```
+
+# Save most recent workspace for further analyses
+
+```{r}
+save.image(file = "./workspace.RData", safe = TRUE)
+```
+
+# Print session info
+
+```{r}
+sessionInfo()
+```

### 2021-03-10 paper_reporting_death_at_home_increase.Rmd
created a separate script for producing tables & figures for paper

---
title: "Tables & figures for paper reporting increase of home deaths 2020"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(scico)  # for continuous grayscale palette
```

# Load workspace

This was previously saved!

```{r}
load(file = "./workspace.RData")
```

# Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
dir_paper <- paste0("./outputs/paper_",latest_data_date)
if (!dir.exists(dir_paper)) dir.create(dir_paper)
```

# Tables

Note: non-covid deaths are defined as all-cause deaths minus covid-related deaths. Otherwise the number of non-covid deaths only includes non-covid deaths once covid-related deaths started being reported in 2020 (week 12).

## Number & proportion of deaths by year & place

```{r}
table_of_all_cause_deaths_by_place_of_death <-
  merged_proportions_of_deaths_by_place %>%
  rename(n = number_of_deaths, prop = proportion_of_total) %>%
  pivot_wider(names_from = c(place_of_death), values_from = c(n, prop), names_sort = FALSE) %>%
  arrange(desc(year)) %>%
  relocate(1,2,3,7,4,8,5,9,6,10)  # manually reorder columns

table_of_non_covid_deaths_by_place_of_death <-
  merged_proportions_of_non_covid_deaths_by_place %>%
  rename(n = number_of_deaths, prop = proportion_of_total) %>%
  pivot_wider(names_from = c(place_of_death), values_from = c(n, prop), names_sort = FALSE) %>%
  arrange(desc(year)) %>%
  relocate(1,2,3,7,4,8,5,9,6,10)  # manually reorder columns

table_1_for_paper <-
  bind_rows(
    table_of_all_cause_deaths_by_place_of_death %>% mutate(cause = "All Deaths") %>% relocate(cause),
    table_of_non_covid_deaths_by_place_of_death %>% mutate(cause = "Non-covid Deaths") %>% relocate(cause)
  )

write.xlsx(x = table_1_for_paper, file = paste0(dir_paper,"/number_of_deaths_by_year_place_pre_edit.xlsx"))
```

# Visualisations

## Common elements

```{r}
## modify theme with relevant font
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))
```

## Charts: number of deaths by week & place of death

### Showing historical trends as well as 2020

```{r}
## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
# annotations_figure_place_of_death_explaining_historic_range <- tribble(
#    ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
#   "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
#   "curve", "Hospital", NA, 29, 830, 32, 550,
#   "text", "Hospital", "Historic mean", 22, 800, NA, NA,
#   "curve", "Hospital", NA, 22, 780, 26, 510
# ) %>% 
#   mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
#   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)

figure_past_place_of_death <-
  weekly_deaths_current_and_past %>%
  filter(!place_of_death %in% c("All","Other") & year < 2020) %>%  # remove Other place of death & keep past years only
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  # pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               # names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number, colour = year, y = deaths_all_causes, group = year)) +
  # annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
  # scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
  # geom_ribbon(data = merged_deaths_overall %>% filter(year==2020 & !place_of_death %in% c("All","Other")), aes(x = week_number, ymin = deaths_min, ymax = deaths_max), inherit.aes = FALSE, fill = "grey70", alpha = 0.5) +
  geom_line(aes(colour = as.factor(year))) +
  scale_colour_viridis_d() +
  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  # scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
  #                     labels = c("All deaths", 
  #                                "Non Covid-19 deaths"),
  #                     guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
  #                     ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = "Date w/c",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,"Other place of death not shown.",source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  # geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
  # geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
  # annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
  NULL

figure1b <- 
  figure_past_place_of_death +
  labs(title=NULL)

save_output_file(filename = paste0(dir_paper,"/figure1b"), extensions = c(".pdf",".wmf",".png"), plot = figure1b, width = 28, height = 14, dpi = 300, units = "cm")
```

### Grayscale plot of weekly deaths w/ historical range for 2020

```{r}
figure1 <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all" & year == 2020) %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(fill_text = "Historical range") %>%  # added for ease of using aes(fill) later
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid, deaths_mean), 
               names_to = "deaths_type", values_to = "deaths") %>%
  mutate(deaths_type = factor(x = deaths_type, levels = c("deaths_all_causes", "deaths_non_covid", "deaths_mean"), labels = c("All deaths", "Non Covid-19 deaths", "Historical mean"))) %>%
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = fill_text), alpha = 0.5) +
  geom_line(aes(y = deaths, linetype = deaths_type, colour = deaths_type)) +
  scale_colour_manual(values = c("black", "black", "gray30")) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  ylim(0, NA) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_fill_manual(values = c("grey70")) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.box.spacing = unit(0, "cm"),
    plot.caption = element_text(size = 10, colour = "gray60")
    ) +
  labs(x = "Week number",
       y = "Deaths (N)",
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n")
       ) +
  NULL

## figure for a bit less than half A4 page is approx 14*11cm
save_output_file(filename = paste0(dir_paper,"/figure1"), extensions = c(".pdf",".svg",".png"), plot = figure1, width = 28, height = 11, units = "cm", dpi = 300)
```

## Maps: home death increase & case counts

These are figure 2!

### Individual plots, titles removed

```{r}
figure2a_bw <-
  figure_map_ratio_home_deaths_2020_to_historical_hb_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(title = NULL, x = NULL, y = NULL)

figure2b_bw <-
  figure_map_per_capita_covid_cases_2020_hb_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) format(x, big.mark=",")) +
  labs(title = NULL, x = NULL, y = NULL)

figure2c_bw <-
  figure_map_ratio_home_deaths_2020_to_historical_la_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(title = NULL, x = NULL, y = NULL)

figure2d_bw <-
  figure_map_per_capita_covid_cases_2020_la_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) format(x, big.mark=",")) +
  labs(title = NULL, x = NULL, y = NULL)

save_output_file(filename = paste0(dir_paper,"/figure2a_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2a_bw, width = 12, height = 14, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2b_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2b_bw, width = 12, height = 14, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2c_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2c_bw, width = 12, height = 14, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2d_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2d_bw, width = 12, height = 14, dpi = 300, units = "cm")

```

### Combined plots

```{r}
figure2ab_bw <-
  wrap_plots(
    list(
      figure2a_bw + ggtitle("a)"), 
      figure2b_bw + ggtitle("b)")
      ), 
    nrow = 1)

figure2cd_bw <-
  wrap_plots(
    list(
      figure2c_bw + ggtitle("c)"), 
      figure2d_bw + ggtitle("d)")
      ), 
    nrow = 1)

save_output_file(filename = paste0(dir_paper,"/figure2ab_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2ab_bw, width = 20, height = 12, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2cd_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2cd_bw, width = 20, height = 12, dpi = 300, units = "cm")
```

### 2021-03-10 download_latest_data_and_render_reports.R

@@ -196,3 +196,11 @@ render(
   # params = params, 
   envir = new.env()
 )
+
+render(
+  input = "./paper_reporting_death_at_home_increase.Rmd",
+  output_file = paste0("paper_reporting_death_at_home_increase_",today()),
+  output_dir = dir_reports,
+  envir = new.env()
+)
+

### 2021-03-11 covid_nrs_place_of_death.Rmd
several updates:
* fixed the saving of workspace file in the first .Rmd script;
* added a separate script for the visualisations for the blog, to which I'll move the relevant tables & graphs gradually
* added a table showing percentage increase of home deaths by age group - this should be a function, really, so we can compare across age,x sex, la, and hb
* added some things to the paper .rmd

@@
 library(openxlsx)  # for creating xlsx files
+library(knitr)  # for displaying tables in rmd
 ```
@@
   .x = extensions,
-  .f = ~function(ext) {
-    if (tolower(ext)==".pdf") device <- cairo_pdf
-
-  ggsave(filename = paste0(filename, .x), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (.x==".pdf") cairo_pdf else NULL)
+  .f = function(ext) {
+  # if (tolower(ext)==".pdf") device <- cairo_pdf
+  ggsave(filename = paste0(filename, ext), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (tolower(ext)==".pdf") cairo_pdf else NULL)
   }
@@
 figure_deaths_at_home_by_age
+```
+
+### Percent increase in home deaths over historic average, by age
+
+```{r}
+merged_deaths_age %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  group_by(year, place_of_death, age, week_number) %>%
+  summarise(
+    deaths_all_causes = sum(deaths_all_causes),
+    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+    .groups = "drop"
+  ) %>% 
+  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+  left_join(
+    past_deaths_age %>% filter(year != "average") %>% group_by(week_number, place_of_death, age) %>% summarise(
+                                       average_weekly_deaths = sum(number_of_deaths) / 5,
+                                       .groups = "drop"
+                                     ),
+    by = c("week_number", "place_of_death", "age")
+  ) %>% 
+  group_by(place_of_death, age) %>%
+  summarise(
+    period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+    deaths_all_causes = sum(deaths_all_causes),
+    deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+    mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+    .groups = "drop"
+  ) %>%
+  mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
+  kable(caption = "Total deaths at home in data so far for age groups.")
 ```
@@
 ```{r}
-save.image(file = "./workspace.RData", safe = TRUE)
+# save.image(file = "./workspace.RData", safe = TRUE)  # this is calling .GlobalEnv, which is empty
+save(list = ls(all.names = TRUE), file = "./workspace.RData", envir = environment())
 ```

### 2021-03-11 paper_reporting_death_at_home_increase.Rmd
several updates:
* fixed the saving of workspace file in the first .Rmd script;
* added a separate script for the visualisations for the blog, to which I'll move the relevant tables & graphs gradually
* added a table showing percentage increase of home deaths by age group - this should be a function, really, so we can compare across age,x sex, la, and hb
* added some things to the paper .rmd

@@
 
+## Correlation between annual covid cases & annual deaths
+
+The interesting correlation is between the ratio of 2020 home deaths to historical mean, and the covid cases, rather than the total home deaths!
+
+```{r}
+home_deaths_and_cases_2020_hb <-
+  total_deaths_hb %>%
+  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
+  left_join(
+    weekly_covid_cases_hb %>%
+      filter(year == 2020 & week_number == 53),
+    by = c("year", "ref_area")
+  ) %>%
+  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related)
+
+home_deaths_and_cases_2020_la <-
+  total_deaths_la %>%
+  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
+  left_join(
+    weekly_covid_cases_la %>%
+      filter(year == 2020 & week_number == 53),
+    by = c("year", "ref_area")
+  ) %>%
+  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related)
+
+# Non-covid home deaths not included because nearly identical to all deaths
+
+correlations_home_deaths_and_cases <-
+  bind_rows(
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$cumulative_positive, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$cumulative_positive, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$cumulative_positive, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$cumulative_positive, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$cumulative_positive, method = "pearson")) %>% mutate(x="All home deaths", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$cumulative_positive, method = "pearson")) %>% mutate(x="All home deaths", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$cumulative_positive, method = "spearman")) %>% mutate(x="All home deaths", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$cumulative_positive, method = "spearman")) %>% mutate(x="All home deaths", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "LA")
+  ) %>%
+  mutate(year = 2020) %>%
+  relocate(region, year, x, y)
+
+write.xlsx(x = correlations_home_deaths_and_cases, file = paste0(dir_paper, "/correlations_home_deaths_and_cases_by_region.xlsx"))
+```
+
 # Visualisations
@@
 
-## Charts: number of deaths by week & place of death
-
-### Showing historical trends as well as 2020
+## Showing historical trends as well as 2020
 
@@
 
-### Grayscale plot of weekly deaths w/ historical range for 2020
+## Grayscale plot of weekly deaths w/ historical range for 2020
 

### 2021-03-11 visualisations_for_blog.Rmd
several updates:
* fixed the saving of workspace file in the first .Rmd script;
* added a separate script for the visualisations for the blog, to which I'll move the relevant tables & graphs gradually
* added a table showing percentage increase of home deaths by age group - this should be a function, really, so we can compare across age,x sex, la, and hb
* added some things to the paper .rmd

---
title: "Visualisations for blog on SCADR website"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(scico)  # for continuous grayscale palette
library(kable)  # for printing tables nicely
```

# Load workspace

This was previously saved!

```{r}
load(file = "./workspace.RData")
```

# Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
dir_blog <- paste0("./outputs/scadr_blog_",latest_data_date)
if (!dir.exists(dir_blog)) dir.create(dir_blog)
```

# Tables

# Visualisations

## Common elements

```{r}
## modify theme with relevant font
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))
```

## Home deaths by HB: Forth Valley vs Dumfries

```{r}
(home_deaths_by_hb_fv_vs_dag_2week_average <-
  merged_deaths_hb %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid, ma2w_deaths_mean), 
               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
  mutate(
    deaths_type = factor(x = deaths_type, levels = c("ma2w_deaths_all_causes", "ma2w_deaths_non_covid", "ma2w_deaths_mean"), labels = c("All deaths", "Non Covid-19 deaths", "Historic mean"))
    ) %>%
  mutate(fill_text = "Historic range (min/max)") %>%  # added for ease of using aes(fill) later
  ggplot(aes(x = date_w_c)) +
  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max, fill = fill_text), alpha = 0.5) +
  geom_line(aes(y = ma2w_deaths, colour = deaths_type, linetype = deaths_type)) +
  ylim(0, NA) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~ref_area) +
  scale_linetype_manual(values = c("solid","solid","dashed")) +
  scale_fill_manual(values = c("#4477aa")) +  # specify historic range fill
  scale_colour_manual(values = c("red","blue","grey20")) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.title = element_blank()
  ) +
  labs(x = "Date w/c",
       y = "Deaths, 2-week average (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n")
       # colour = "Cause of death" 
       ) +
  # ## annotate year
  geom_text(aes(x=ymd("2021-01-07"), y=2, label = "2021"), hjust = 0, colour = "grey40") +
  NULL
)
save_output_file(filename = paste0(dir_blog,"/home_deaths_by_hb_fv_vs_dag_2week_average"), extensions = c(".pdf",".svg",".png"), plot = home_deaths_by_hb_fv_vs_dag_2week_average, width = 16, height = 12, dpi = 300, units = "cm")
```

```{r}
total_deaths_hb %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
  filter(year == 2020) %>%
  select(year, place_of_death, ref_area, deaths_all_causes, ratio_annual_deaths_to_historical_mean) %>%
  kable(caption = "Total deaths at home in 2020 for health boards.")

merged_deaths_hb %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
  group_by(year, place_of_death, ref_area, week_number) %>%
  summarise(
    deaths_all_causes = sum(deaths_all_causes),
    # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
  left_join(
    past_deaths_hb %>% filter(year != "average") %>% group_by(week_number, place_of_death, ref_area) %>% summarise(
                                       average_weekly_deaths = sum(number_of_deaths) / 5,
                                       .groups = "drop"
                                     ),
    by = c("week_number", "place_of_death", "ref_area")
  ) %>% 
  group_by(place_of_death, ref_area) %>%
  summarise(
    period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
    deaths_all_causes = sum(deaths_all_causes),
    deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
    mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
    .groups = "drop"
  ) %>%
  mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
  kable(caption = "Total deaths at home in data so far for health boards.")
```

### 2021-03-11 download_latest_data_and_render_reports.R

@@ -204,3 +204,10 @@ render(
   envir = new.env()
 )
 
+render(
+  input = "./visualisations_for_blog.Rmd",
+  output_file = paste0("visualisations_for_blog_",today()),
+  output_dir = dir_reports,
+  envir = new.env()
+)
+

### 2021-03-18 covid_nrs_place_of_death.Rmd
added figures for total deaths by year, and another where it's also by place of death
renamed some historical summaries so it's clear they are averages

started toying around with bed occupancy data!

@@
 
+## Load quarterly bed occupancy data
+
+```{r}
+# TODO: download automatically in other script
+
+beds_raw <- 
+  read_csv("C:/Users/janwork/Downloads/beds_by_nhs_board_of_treatment_and_specialty.csv", guess_max = 3e4) %>%
+  mutate(ref_area = match_area(HB)) %>%  # clean health board names
+  clean_names() %>%
+  # select(-percentage_occupancy_qf) %>%
+  extract(col = quarter, into = c("year", "quarter"), regex = "(\\d{4})Q(\\d)", remove = TRUE) %>%
+  # mutate(across(.cols = matches("percent"), .fns = ~{.x / 100}))
+  mutate(percentage_occupancy = percentage_occupancy / 100)
+
+beds_raw %>%
+  filter(ref_area == "Scotland") %>% count(specialty_name)
+
+beds_raw %>%
+  filter(ref_area == "Scotland" & specialty_name == "Palliative Medicine") %>%
+  ggplot() +
+  aes(x = quarter, y = percentage_occupancy, colour = year, size = year) +
+  scale_y_continuous(labels = function(x) scales::percent(x), limits = c(0,NA)) +
+  # geom_point() +
+  geom_line(aes(group = year)) +
+  scale_colour_viridis_d() +
+  scale_size_manual(values = c(rep(0.5, times = 5),1)) +
+  NULL
+
+```
+
 # Calculate case counts: weekly counts to match weekly death reporting
@@
 
-historical_average_overall <-
+historical_weekly_averageoverall <-
   past_deaths_overall %>%
@@
 
-historical_average_hb <-
+historical_weekly_averagehb <-
   past_deaths_hb %>%
@@
 
-historical_average_la <-
+historical_weekly_averagela <-
   past_deaths_la %>%
@@
 
-historical_average_sex_age <-
+historical_weekly_averagesex_age <-
   past_deaths_sex_age %>%
@@
 
-historical_average_age <-
+historical_weekly_averageage <-
   past_deaths_age %>%
@@
 
-historical_average_sex <-
+historical_weekly_averagesex <-
   past_deaths_sex %>%
@@
   filter(sex=="all" & age=="all") %>%
-  left_join(historical_average_overall) %>%
+  left_join(historical_weekly_averageoverall) %>%
   select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean) %>%
@@
 
-historical_average_hb <- compute_moving_average_for_historical_data(historical_average_hb)
-historical_average_la <- compute_moving_average_for_historical_data(historical_average_la)
-historical_average_sex_age <- compute_moving_average_for_historical_data(historical_average_sex_age)
-historical_average_sex <- compute_moving_average_for_historical_data(historical_average_sex)
-historical_average_age <- compute_moving_average_for_historical_data(historical_average_age)
+historical_weekly_averagehb <- compute_moving_average_for_historical_data(historical_weekly_averagehb)
+historical_weekly_averagela <- compute_moving_average_for_historical_data(historical_weekly_averagela)
+historical_weekly_averagesex_age <- compute_moving_average_for_historical_data(historical_weekly_averagesex_age)
+historical_weekly_averagesex <- compute_moving_average_for_historical_data(historical_weekly_averagesex)
+historical_weekly_averageage <- compute_moving_average_for_historical_data(historical_weekly_averageage)
 ```
@@
   filter(sex=="all" & age=="all") %>%
-  inner_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+  inner_join(historical_weekly_averageoverall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
             by = c("place_of_death", "week_number", "sex", "age"))
@@
   rolling_deaths_hb %>%  # use data with rolling movingincluded
-  inner_join(historical_average_hb, by = c("week_number","ref_area","place_of_death"))
+  inner_join(historical_weekly_averagehb, by = c("week_number","ref_area","place_of_death"))
 
@@
   rolling_deaths_la %>%  # use data with rolling movingincluded
-  inner_join(historical_average_la, by = c("week_number","ref_area","place_of_death"))
+  inner_join(historical_weekly_averagela, by = c("week_number","ref_area","place_of_death"))
 
@@
   rolling_deaths_sex_age %>%  # use data with rolling movingincluded
-  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))
+  inner_join(historical_weekly_averagesex_age, by = c("place_of_death", "week_number", "sex", "age"))
 
@@
   rolling_deaths_sex %>%
-  inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))
+  inner_join(historical_weekly_averagesex, by = c("place_of_death", "week_number", "sex"))
 
@@
   rolling_deaths_age %>%
-  inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))
+  inner_join(historical_weekly_averageage, by = c("place_of_death", "week_number", "age"))
 ```
@@
   )
+```
+
+# Deaths over time
+
+## Number of deaths by year
+
+```{r}
+# TODO: finish this here or elsewhere!
+# total_deaths_overall
+# 
+# past_deaths_overall %>%
+#   group_by(year, place_of_death) %>%
+#   summarise(number_of_deaths = sum(number_of_deaths, na.rm = TRUE), .groups="drop")
+# 
+# # TODO: can this be generalised in a function?
+# merged_deaths_age %>%
+#   # filter(place_of_death %in% c("Home & other non-institution")) %>%
+#   group_by(year, place_of_death, age, week_number) %>%
+#   summarise(
+#     deaths_all_causes = sum(deaths_all_causes),
+#     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+#     .groups = "drop"
+#   ) %>% 
+#   mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+#   left_join(
+#     past_deaths_age %>% filter(year != "average") %>% group_by(week_number, place_of_death, age) %>% summarise(
+#                                        average_weekly_deaths = sum(number_of_deaths) / 5,
+#                                        .groups = "drop"
+#                                      ),
+#     by = c("week_number", "place_of_death", "age")
+#   ) %>% 
+#   group_by(place_of_death, age) %>%
+#   summarise(
+#     period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+#     deaths_all_causes = sum(deaths_all_causes),
+#     deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+#     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+#     mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+#     .groups = "drop"
+#   ) %>%
+#   mutate(
+#     percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
+#     percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
+#     ) %>%
+#   (function(x) {
+#     bind_rows(
+#       x,
+#       x %>% group_by(age, period) %>% summarise(
+#         deaths_all_causes = sum(deaths_all_causes),
+#         deaths_non_covid = sum(deaths_non_covid),
+#         mean_annual_deaths_2015_2019 = sum(mean_annual_deaths_2015_2019)
+#       ) %>%
+#         mutate(
+#           percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
+#           percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
+#         )
+#     )
+#   }) %>%
+#   View()
+#   # kable(caption = "Total deaths at home in data so far for age groups.")
 ```
@@
 figure_proportion_home_deaths_over_time
+```
+
+# Number of deaths by place over time
+
+```{r}
+(figure_number_of_deaths_by_place_by_year <-
+  bind_rows(
+    proportions_of_historical_deaths_by_place %>% filter(year==2019) %>% mutate(covid_status = "Non Covid-19 deaths"),  # add extra rows of 2019 data for plotting line between 2019 and 2020
+    proportions_of_historical_deaths_by_place %>% mutate(covid_status = "All deaths"),
+    proportions_of_current_non_covid_deaths_by_place %>% mutate(covid_status = "Non Covid-19 deaths"),
+    proportions_of_current_deaths_by_place %>% mutate(covid_status = "All deaths")
+  ) %>%
+  filter(!place_of_death %in% c("All","Other")) %>%  # don't show Other
+  filter(year != 2021) %>%  # don't include 2021
+  mutate(
+    place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])   # move home to 2nd position to match order in plot
+    ) %>%
+  ggplot(., aes(x=year, y=number_of_deaths, colour = place_of_death)) +
+  geom_line(data = . %>% filter(covid_status == "All deaths"), aes(group = place_of_death, linetype = covid_status)) +
+  geom_line(data = . %>% filter(covid_status == "Non Covid-19 deaths"), aes(group = place_of_death, linetype = covid_status)) +
+  geom_point(data = . %>% filter(!covid_status=="Non Covid-19 deaths" | !year==2019)) +
+  geom_text(data = . %>% filter(year==2016) %>% mutate(y_pos = c(3.1e4, 1.1e4, 1.8e4)), aes(y = y_pos, label = place_of_death), hjust = 0) +
+  scale_colour_viridis_d(option="C", end = 0.8, guide = FALSE) +
+  scale_linetype_manual(values = c("solid", "dashed")) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    legend.box="vertical",
+    legend.margin=margin()
+    ) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ","), limits = c(0, NA)) +
+  labs(
+    x = "Year", y = "Deaths (N)",
+    caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_number_of_deaths_by_place_by_year"), extensions = c(".pdf",".svg",".png"), plot = figure_number_of_deaths_by_place_by_year, width = 15, height = 11, dpi = 300, units = "cm")
+```
+
+## Overall, regardless of place
+
+```{r}
+number_of_covid_deaths_2020 <- total_deaths_overall %>% filter(year==2020 & place_of_death == "All") %>% .$deaths_covid_related
+proportion_of_covid_deaths_2020 <- total_deaths_overall %>% filter(year==2020 & place_of_death == "All") %>% {scales::percent(.$deaths_covid_related / .$deaths_all_causes, accuracy = 0.1)}
+
+(figure_number_of_deaths_by_year <-
+bind_rows(
+  merged_proportions_of_deaths_by_place %>%  # total deaths regardless of place
+    select(year, total_deaths) %>% 
+    distinct %>%
+    mutate(covid_status = "All deaths")
+    ,
+  merged_proportions_of_non_covid_deaths_by_place %>%
+    select(year, total_deaths) %>%
+    distinct %>%
+    # filter(year %in% 2019:2020) %>%
+    filter(year %in% 2020) %>%
+    mutate(covid_status = "Non Covid-19 deaths")
+  ) %>%
+  filter(year!=2021) %>%  # remove 2021
+  ggplot(
+    ., aes(x = year, y = total_deaths, fill = covid_status)
+  ) +
+  scale_fill_manual(values = c("red","blue")) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    legend.box="vertical",
+    legend.margin=margin()
+    ) +
+  geom_col(position = "dodge", colour = "grey20") +
+  # geom_line() +
+  # geom_point() +
+  scale_x_continuous(breaks = 2015:2020) +
+  # annotate(geom = "text", x = 2016, y = 65e3, hjust = 0, label = glue::glue("2020 COVID-related deaths N={format(number_of_covid_deaths_2020, big.mark=',')}\n({proportion_of_covid_deaths_2020} of all deaths)")) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ","), limits = c(0, NA)) +
+  labs(
+    x = "Year", y = "Deaths (N)",
+    subtitle = glue::glue("2020 COVID-related deaths N={format(number_of_covid_deaths_2020, big.mark=',')}\n({proportion_of_covid_deaths_2020} of all deaths)"),
+    caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_number_of_deaths_by_year"), extensions = c(".pdf",".svg",".png"), plot = figure_number_of_deaths_by_year, width = 12, height = 10, dpi = 300, units = "cm")
 ```
@@
 ```{r}
+# TODO: can this be generalised in a function?
 merged_deaths_age %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  # filter(place_of_death %in% c("Home & other non-institution")) %>%
   group_by(year, place_of_death, age, week_number) %>%
@@
   ) %>%
-  mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
-  kable(caption = "Total deaths at home in data so far for age groups.")
+  mutate(
+    percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
+    percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
+    ) %>%
+  (function(x) {
+    bind_rows(
+      x,
+      x %>% group_by(age, period) %>% summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_non_covid = sum(deaths_non_covid),
+        mean_annual_deaths_2015_2019 = sum(mean_annual_deaths_2015_2019)
+      ) %>%
+        mutate(
+          percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
+          percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
+        )
+    )
+  }) %>%
+  View()
+  # kable(caption = "Total deaths at home in data so far for age groups.")
 ```

### 2021-03-31 covid_nrs_place_of_death.Rmd
incorporated "main report" from nrs website so I can do breakdown of weekly deaths by cause of death... hope it works next week!
also started making some visualisations of cause of death

@@
 ```{r}
-latest_data_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
-
-dir_outputs <- paste0("./outputs/",latest_data_date)
+latest_data_modified_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
+
+dir_outputs <- paste0("./outputs/",latest_data_modified_date)
 if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
@@
 
+### Define locations for weekly data fiels
+
 ```{r}
@@
 files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
-
+```
+
+### Load  'main report' data
+
+We are only loading deaths aggregated by cause from the 'main report' data! These are compiled in two sheets:
+
+* Table 3 (2020)
+* Table 3  (2021)
+
+However, the number of spaces might differ! So we will extract the table names automatically.
+
+The ICD-10 codes used for causes of death:
+
+>4) The ICD 10 codes for disease categories are as follows:	
+Cancer: C00-C97	
+Dementia and Alzheimer's: F01, F03 and G30	
+Circulatory: I00:I99	
+Respiratory: J00-J99	
+COVID-19: U07
+
+The format is a bit complicated to read automatically; we will extract the number of weeks available in the data from the filename.
+
+~~These data come chopped up into 4 sheets, 2 by 2 for each of year 2020/2021 and all deaths/covid-related deaths. They are named:
+
+* Table 1 (2020)
+* Table 1 (2021)
+* Table 2 (2020)
+* Table 2 (2021)
+
+~~
+
+```{r}
+weeks_available_in_nrs_weekly_data_by_cause_of_death_2021 <-  # extract the maximum weeks available from the weekly-updated 2021 data
+  files_weekly_deaths$main_report %>%
+  str_extract(string = ., pattern = "week-\\d{2}") %>%  # extract eg "week-12"
+  str_extract(string = ., pattern = "\\d{2}") %>%  # extract "12"
+  as.integer()
+
+## extract the sheet names for cause of death data
+nrs_weekly_data_sheet_2020 <- 
+  excel_sheets(files_weekly_deaths$main_report)[which(str_detect(tolower(excel_sheets(files_weekly_deaths$main_report)), pattern = "table 3.*2020"))]
+nrs_weekly_data_sheet_2021 <- 
+  excel_sheets(files_weekly_deaths$main_report)[which(str_detect(tolower(excel_sheets(files_weekly_deaths$main_report)), pattern = "table 3.*2021"))]
+
+## construct column names for the data
+headers_in_nrs_data_2020 <-
+  c(
+    "place_of_death",
+    "cause_of_death",
+    paste0("week_", 1:53),
+    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
+    "remove_1",
+    "remove_2"
+  )
+headers_in_nrs_data_2021 <-
+  c(
+    "place_of_death",
+    "cause_of_death",
+    paste0(
+      "week_",
+      1:weeks_available_in_nrs_weekly_data_by_cause_of_death_2021
+    ),
+    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
+    "remove_1"
+  )
+
+## extract data... this is dense but I've tried to annotate it clearly!
+weekly_deaths_cause_of_death_before_split_2020 <-
+  readxl::read_excel(files_weekly_deaths$main_report, sheet = nrs_weekly_data_sheet_2020, col_names = headers_in_nrs_data_2020, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
+  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
+  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
+  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
+  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
+  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
+  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
+  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "number_of_deaths", values_drop_na = TRUE) %>%  # convert to long data format
+  mutate(
+    week_number = as.integer(parse_number(week_number)),  # extract week number
+    place_of_death =  # find best match in the reference names used in other datasets
+      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)]  # the reference is in the variable order_of_place_of_death_levels
+    )
+
+weekly_deaths_cause_of_death_before_split_2021 <-
+  readxl::read_excel(files_weekly_deaths$main_report, sheet = nrs_weekly_data_sheet_2021, col_names = headers_in_nrs_data_2021, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
+  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
+  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
+  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
+  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
+  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
+  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
+  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "number_of_deaths", values_drop_na = TRUE) %>%  # convert to long data format
+  mutate(
+    week_number = as.integer(parse_number(week_number)),  # extract week number
+    place_of_death =  # find best match in the reference names used in other datasets
+      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)]  # the reference is in the variable order_of_place_of_death_levels
+    )
+
+## compile the weekly data
+weekly_deaths_cause_of_death <-
+  bind_rows(
+    weekly_deaths_cause_of_death_before_split_2020 %>% filter(!str_detect(tolower(dataset), pattern = "difference|average")),
+    weekly_deaths_cause_of_death_before_split_2021 %>% filter(!str_detect(tolower(dataset), pattern = "difference|average"))
+  ) %>%
+  mutate(
+    year = as.integer(parse_number(dataset)),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+    ) %>%
+  select(-dataset)
+
+## compile historical data - note that it's already summarised as average over the 5 year period, so we don't have the raw data, nor do we have minima/maxima
+historical_average_cause_of_death <-
+  weekly_deaths_cause_of_death_before_split_2020 %>%  # only data from the 2020 sheet are used - the 2021 data is the same entries for historical deaths!
+  filter(str_detect(tolower(dataset), pattern = "average")) %>%
+  select(-dataset) %>%
+  rename(deaths_mean = number_of_deaths) %>%
+  mutate(deaths_min = NA_real_, deaths_max = NA_real_)
+```
+
+### Load 'ad-hoc' query data
+
+These are data aggregated by sex, age, and region (HB/LA)!
+
+```{r}
 # checking out structure
@@
 
-beds_raw <- 
-  read_csv("C:/Users/janwork/Downloads/beds_by_nhs_board_of_treatment_and_specialty.csv", guess_max = 3e4) %>%
-  mutate(ref_area = match_area(HB)) %>%  # clean health board names
-  clean_names() %>%
-  # select(-percentage_occupancy_qf) %>%
-  extract(col = quarter, into = c("year", "quarter"), regex = "(\\d{4})Q(\\d)", remove = TRUE) %>%
-  # mutate(across(.cols = matches("percent"), .fns = ~{.x / 100}))
-  mutate(percentage_occupancy = percentage_occupancy / 100)
-
-beds_raw %>%
-  filter(ref_area == "Scotland") %>% count(specialty_name)
-
-beds_raw %>%
-  filter(ref_area == "Scotland" & specialty_name == "Palliative Medicine") %>%
-  ggplot() +
-  aes(x = quarter, y = percentage_occupancy, colour = year, size = year) +
-  scale_y_continuous(labels = function(x) scales::percent(x), limits = c(0,NA)) +
-  # geom_point() +
-  geom_line(aes(group = year)) +
-  scale_colour_viridis_d() +
-  scale_size_manual(values = c(rep(0.5, times = 5),1)) +
-  NULL
+# beds_raw <- 
+#   read_csv("./beds_by_nhs_board_of_treatment_and_specialty.csv", guess_max = 3e4) %>%
+#   mutate(ref_area = match_area(HB)) %>%  # clean health board names
+#   clean_names() %>%
+#   # select(-percentage_occupancy_qf) %>%
+#   extract(col = quarter, into = c("year", "quarter"), regex = "(\\d{4})Q(\\d)", remove = TRUE) %>%
+#   # mutate(across(.cols = matches("percent"), .fns = ~{.x / 100}))
+#   mutate(percentage_occupancy = percentage_occupancy / 100)
+# 
+# beds_raw %>%
+#   filter(ref_area == "Scotland") %>% count(specialty_name)
+# 
+# beds_raw %>%
+#   filter(ref_area == "Scotland" & specialty_name == "Palliative Medicine") %>%
+#   ggplot() +
+#   aes(x = quarter, y = percentage_occupancy, colour = year, size = year) +
+#   scale_y_continuous(labels = function(x) scales::percent(x), limits = c(0,NA)) +
+#   # geom_point() +
+#   geom_line(aes(group = year)) +
+#   scale_colour_viridis_d() +
+#   scale_size_manual(values = c(rep(0.5, times = 5),1)) +
+#   NULL
 
@@
 
-historical_weekly_averageoverall <-
+historical_weekly_average_overall <-
   past_deaths_overall %>%
@@
 
-historical_weekly_averagehb <-
+historical_weekly_average_hb <-
   past_deaths_hb %>%
@@
 
-historical_weekly_averagela <-
+historical_weekly_average_la <-
   past_deaths_la %>%
@@
 
-historical_weekly_averagesex_age <-
+historical_weekly_average_sex_age <-
   past_deaths_sex_age %>%
@@
 
-historical_weekly_averageage <-
+historical_weekly_average_age <-
   past_deaths_age %>%
@@
 
-historical_weekly_averagesex <-
+historical_weekly_average_sex <-
   past_deaths_sex %>%
@@
   compute_historical_annual_average_deaths(age)
+
+historical_annual_average_cause_of_death <-  # this isn't quite the right figure because we don't have the raw data to work from, but weekly averages across 2015-2019
+  historical_average_cause_of_death %>%
+  group_by(place_of_death, cause_of_death) %>%
+  summarise(
+    annual_deaths_mean = sum(deaths_mean),
+    annual_deaths_min = NA_real_,
+    annual_deaths_max = NA_real_,
+    .groups = "drop"
+  )
 ```
@@
   filter(sex=="all" & age=="all") %>%
-  left_join(historical_weekly_averageoverall) %>%
+  left_join(historical_weekly_average_overall) %>%
   select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean) %>%
@@
 
-historical_weekly_averagehb <- compute_moving_average_for_historical_data(historical_weekly_averagehb)
-historical_weekly_averagela <- compute_moving_average_for_historical_data(historical_weekly_averagela)
-historical_weekly_averagesex_age <- compute_moving_average_for_historical_data(historical_weekly_averagesex_age)
-historical_weekly_averagesex <- compute_moving_average_for_historical_data(historical_weekly_averagesex)
-historical_weekly_averageage <- compute_moving_average_for_historical_data(historical_weekly_averageage)
+historical_weekly_average_hb <- compute_moving_average_for_historical_data(historical_weekly_average_hb)
+historical_weekly_average_la <- compute_moving_average_for_historical_data(historical_weekly_average_la)
+historical_weekly_average_sex_age <- compute_moving_average_for_historical_data(historical_weekly_average_sex_age)
+historical_weekly_average_sex <- compute_moving_average_for_historical_data(historical_weekly_average_sex)
+historical_weekly_average_age <- compute_moving_average_for_historical_data(historical_weekly_average_age)
 ```
@@
   filter(sex=="all" & age=="all") %>%
-  inner_join(historical_weekly_averageoverall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+  inner_join(historical_weekly_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
             by = c("place_of_death", "week_number", "sex", "age"))
@@
   rolling_deaths_hb %>%  # use data with rolling movingincluded
-  inner_join(historical_weekly_averagehb, by = c("week_number","ref_area","place_of_death"))
+  inner_join(historical_weekly_average_hb, by = c("week_number","ref_area","place_of_death"))
 
@@
   rolling_deaths_la %>%  # use data with rolling movingincluded
-  inner_join(historical_weekly_averagela, by = c("week_number","ref_area","place_of_death"))
+  inner_join(historical_weekly_average_la, by = c("week_number","ref_area","place_of_death"))
 
@@
   rolling_deaths_sex_age %>%  # use data with rolling movingincluded
-  inner_join(historical_weekly_averagesex_age, by = c("place_of_death", "week_number", "sex", "age"))
+  inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))
 
@@
   rolling_deaths_sex %>%
-  inner_join(historical_weekly_averagesex, by = c("place_of_death", "week_number", "sex"))
+  inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex"))
 
@@
   rolling_deaths_age %>%
-  inner_join(historical_weekly_averageage, by = c("place_of_death", "week_number", "age"))
+  inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age"))
+
+merged_deaths_cause_of_death <-
+  weekly_deaths_cause_of_death %>%
+  inner_join(
+    historical_average_cause_of_death  , by = c("place_of_death", "week_number", "cause_of_death")
+  )
 ```
@@
   compute_ratio_of_annual_deaths_to_historical
+
+## bespoke computation because format is different... deaths split by cause!
+total_deaths_cause_of_death <-
+  merged_deaths_cause_of_death %>%
+  group_by(year, place_of_death, cause_of_death) %>%
+  summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
+  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death")) %>%
+  mutate(
+    ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
+  )
 ```
@@
 
-date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")
+date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2021, '%d %B %Y')} (inclusive).")
 
@@
 
-## Overall deaths at home 2020 by cause of death
-
-```{r}
-# TODO: add yearly range for all deaths
-# TODO: split data by covid-underlying cause of desath & covid-factor in death
-merged_deaths_overall %>%
-  filter(sex=="all", age=="all") %>%
-  filter(year==2020) %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  group_by(ref_area, year) %>%
-  summarise(
-    deaths_all_causes = sum(deaths_all_causes, na.rm=TRUE),
-    deaths_covid_related = sum(deaths_covid_related, na.rm=TRUE),
-    deaths_non_covid = sum(deaths_non_covid, na.rm=TRUE)
-  )
+# Deaths by cause of death
+
+## Annual deaths by cause
+
+```{r}
+(figure_annual_deaths_by_cause <-
+  total_deaths_cause_of_death %>%
+  filter(year == 2020 & cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
+  # pivot_longer(cols = c(deaths, annual_deaths_mean), names_to = "deaths_type", values_to = "deaths") %>%
+  ggplot() +
+  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), position = "dodge", colour = "grey10") +
+  geom_point(aes(x = cause_of_death, y = annual_deaths_mean), position = "dodge", colour = "grey10") +
+  scale_fill_viridis_d() +
+  facet_wrap(~place_of_death) +
+  NULL
+)
+```
+
+## By cause, at home
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+(figure_deaths_at_home_by_cause <-
+  merged_deaths_cause_of_death %>%
+  filter(str_starts(place_of_death,"Home")) %>%
+  filter(cause_of_death != "All") %>%
+  # filter(!place_of_death %in% c("All","Other")) %>%
+  pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
+  filter(!(deaths_type=="deaths_mean"&cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  ggplot(aes(x = date_w_c)) +
+  # annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_line(aes(y = number_of_deaths, colour = cause_of_death, linetype = deaths_type)) +
+  scale_colour_viridis_d() +
+  scale_linetype_manual(labels = c("Historic mean 2015-2019","Deaths since 2020"), values = c("dashed","solid")) +
+  # geom_line(aes(y = number_of_deaths  , colour = deaths_type)) +
+  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_wrap(cause_of_death~place_of_death, scales = "free_y") +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Weekly Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
+       ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_cause, width = 28, height = 14, dpi = 300, units = "cm")
 ```

### 2021-03-31 download_latest_data_and_render_reports.R

@@ -11,10 +11,24 @@
 
 library(tidyverse)
 library(lubridate)  # dealing with dates
-library(curl)
-library(knitr)
-library(rmarkdown)
-library(extrafont)
+library(curl)  # for downloading & last modified date extraction
+library(knitr)  # for rendering .Rmd
+library(rmarkdown)  # for rendering .Rmd
+library(extrafont)  # for setting fonts
+library(rvest)  # for extracting url from links
+library(xml2)  # for the url_absolute() function
+
+# Find latest NRS weekly deaths spreadsheet URL ---------------------------
+
+url_nrs_weekly_deaths <- "https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland"
+
+all_nrs_links <-
+  read_html(url_nrs_weekly_deaths) %>% 
+  html_nodes("a") %>%
+  html_attr("href")  # find all links on the page
+
+weekly_deaths_url_relative <- all_nrs_links[which(str_detect(all_nrs_links, pattern = "covid-deaths-21-data.*xlsx"))]  # keep only links that are probably weekly deaths
+weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
 
 # Load table keeping track of latest data ---------------------------------
 
@@ -32,8 +46,9 @@ if (file.exists(file_data_sources)) {
   table_of_data_sources <- read_csv(file = file_data_sources)
 } else {
   table_of_data_sources <- tibble(
-    short_name = c("sex_age","la","hb"),
+    short_name = c("main_report","sex_age","la","hb"),
     url = c(
+      weekly_deaths_url_absolute,  # found automatically above!
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx"

### 2021-03-31 paper_reporting_death_at_home_increase.Rmd
addded more exports to blog & paper

--- 2021-03-10 paper_reporting_death_at_home_increase.Rmd
@@
 ```{r}
-dir_paper <- paste0("./outputs/paper_",latest_data_date)
+dir_paper <- paste0("./outputs/paper_",latest_data_modified_date)
 if (!dir.exists(dir_paper)) dir.create(dir_paper)
@@
 
+## Table of annual covid cases & annual deaths
+
+```{r}
+home_deaths_and_cases_2020_hb <-
+  total_deaths_hb %>%
+  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
+  left_join(
+    weekly_covid_cases_hb %>%
+      filter(year == 2020 & week_number == 53),
+    by = c("year", "ref_area")
+  ) %>%
+  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related)
+
+home_deaths_and_cases_2020_la <-
+  total_deaths_la %>%
+  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
+  left_join(
+    weekly_covid_cases_la %>%
+      filter(year == 2020 & week_number == 53),
+    by = c("year", "ref_area")
+  ) %>%
+  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related)
+
+# Non-covid home deaths not included because nearly identical to all deaths
+```
+
+We'll use the LA data, and export a table for the paper!
+
+```{r}
+table_2_paper <-
+  home_deaths_and_cases_2020_la %>%
+  mutate(
+    `Percent increase` = scales::percent(ratio_annual_deaths_to_historical_mean - 1, accuracy = 0.1),
+    `COVID cases per 100,000` = format(round(crude_rate_positive_per100k, digits=1), big.mark = ",")
+    ) %>%
+  select(`Local Authority` = ref_area, `All Deaths (Home)` = deaths_all_causes, `Percent increase`, `COVID cases per 100,000`)
+
+write.xlsx(x = table_2_paper, file = paste0(dir_paper, "/percent_increase_and_covid_cases_by_region.xlsx"))
+```
+
+### Correlation between annual covid cases & annual deaths
+
+The interesting correlation is between the ratio of 2020 home deaths to historical mean, and the covid cases, rather than the total home deaths!
+
+```{r}
+correlations_home_deaths_and_cases <-
+  bind_rows(
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$cumulative_positive, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$cumulative_positive, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$cumulative_positive, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$cumulative_positive, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$cumulative_positive, method = "pearson")) %>% mutate(x="All home deaths", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$cumulative_positive, method = "pearson")) %>% mutate(x="All home deaths", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$cumulative_positive, method = "spearman")) %>% mutate(x="All home deaths", y = "Total cases", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "HB"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$cumulative_positive, method = "spearman")) %>% mutate(x="All home deaths", y = "Total cases", region = "LA"),
+    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "LA")
+  ) %>%
+  mutate(year = 2020) %>%
+  relocate(region, year, x, y)
+
+write.xlsx(x = correlations_home_deaths_and_cases, file = paste0(dir_paper, "/correlations_home_deaths_and_cases_by_region.xlsx"))
+```
+
 # Visualisations

### 2021-03-31 visualisations_for_blog.Rmd
addded more exports to blog & paper

@@
 library(scico)  # for continuous grayscale palette
-library(kable)  # for printing tables nicely
+library(knitr)  # for printing tables nicely
 ```
@@
 ```{r}
-dir_blog <- paste0("./outputs/scadr_blog_",latest_data_date)
+dir_blog <- paste0("./outputs/scadr_blog_",latest_data_modified_date)
 if (!dir.exists(dir_blog)) dir.create(dir_blog)
@@
 
+Note: decided agaisnt doing that, hard to explain why we're singling HBs out!
+
 ```{r}
-(home_deaths_by_hb_fv_vs_dag_2week_average <-
-  merged_deaths_hb %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
-  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid, ma2w_deaths_mean), 
-               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
-  mutate(
-    deaths_type = factor(x = deaths_type, levels = c("ma2w_deaths_all_causes", "ma2w_deaths_non_covid", "ma2w_deaths_mean"), labels = c("All deaths", "Non Covid-19 deaths", "Historic mean"))
-    ) %>%
-  mutate(fill_text = "Historic range (min/max)") %>%  # added for ease of using aes(fill) later
-  ggplot(aes(x = date_w_c)) +
-  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max, fill = fill_text), alpha = 0.5) +
-  geom_line(aes(y = ma2w_deaths, colour = deaths_type, linetype = deaths_type)) +
-  ylim(0, NA) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~ref_area) +
-  scale_linetype_manual(values = c("solid","solid","dashed")) +
-  scale_fill_manual(values = c("#4477aa")) +  # specify historic range fill
-  scale_colour_manual(values = c("red","blue","grey20")) +
-  theme(
-    legend.position = "top",
-    plot.caption = element_text(size = 10, colour = "gray60"),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, hjust = 1),
-    legend.title = element_blank()
-  ) +
-  labs(x = "Date w/c",
-       y = "Deaths, 2-week average (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n")
-       # colour = "Cause of death" 
-       ) +
-  # ## annotate year
-  geom_text(aes(x=ymd("2021-01-07"), y=2, label = "2021"), hjust = 0, colour = "grey40") +
-  NULL
-)
-save_output_file(filename = paste0(dir_blog,"/home_deaths_by_hb_fv_vs_dag_2week_average"), extensions = c(".pdf",".svg",".png"), plot = home_deaths_by_hb_fv_vs_dag_2week_average, width = 16, height = 12, dpi = 300, units = "cm")
+# (home_deaths_by_hb_fv_vs_dag_2week_average <-
+#   merged_deaths_hb %>%
+#   filter(place_of_death %in% c("Home & other non-institution")) %>%
+#   filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
+#   pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid, ma2w_deaths_mean), 
+#                names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
+#   mutate(
+#     deaths_type = factor(x = deaths_type, levels = c("ma2w_deaths_all_causes", "ma2w_deaths_non_covid", "ma2w_deaths_mean"), labels = c("All deaths", "Non Covid-19 deaths", "Historic mean"))
+#     ) %>%
+#   mutate(fill_text = "Historic range (min/max)") %>%  # added for ease of using aes(fill) later
+#   ggplot(aes(x = date_w_c)) +
+#   annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+#   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+#   geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max, fill = fill_text), alpha = 0.5) +
+#   geom_line(aes(y = ma2w_deaths, colour = deaths_type, linetype = deaths_type)) +
+#   ylim(0, NA) +  # start scale at 0 for better understanding the scale of deaths
+#   facet_grid(~ref_area) +
+#   scale_linetype_manual(values = c("solid","solid","dashed")) +
+#   scale_fill_manual(values = c("#4477aa")) +  # specify historic range fill
+#   scale_colour_manual(values = c("red","blue","grey20")) +
+#   theme(
+#     legend.position = "top",
+#     plot.caption = element_text(size = 10, colour = "gray60"),
+#     plot.margin = margin(5, 15, 5, 5),
+#     axis.text.x = element_text(angle = 60, hjust = 1),
+#     legend.title = element_blank()
+#   ) +
+#   labs(x = "Date w/c",
+#        y = "Deaths, 2-week average (N)",
+#        subtitle = date_range_text,
+#        caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n")
+#        # colour = "Cause of death" 
+#        ) +
+#   # ## annotate year
+#   geom_text(aes(x=ymd("2021-01-07"), y=2, label = "2021"), hjust = 0, colour = "grey40") +
+#   NULL
+# )
+# save_output_file(filename = paste0(dir_blog,"/home_deaths_by_hb_fv_vs_dag_2week_average"), extensions = c(".pdf",".svg",".png"), plot = home_deaths_by_hb_fv_vs_dag_2week_average, width = 16, height = 12, dpi = 300, units = "cm")
 ```
@@
 ```{r}
-total_deaths_hb %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
-  filter(year == 2020) %>%
-  select(year, place_of_death, ref_area, deaths_all_causes, ratio_annual_deaths_to_historical_mean) %>%
-  kable(caption = "Total deaths at home in 2020 for health boards.")
+# total_deaths_hb %>%
+#   filter(place_of_death %in% c("Home & other non-institution")) %>%
+#   filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
+#   filter(year == 2020) %>%
+#   select(year, place_of_death, ref_area, deaths_all_causes, ratio_annual_deaths_to_historical_mean) %>%
+#   kable(caption = "Total deaths at home in 2020 for health boards.")
+# 
+# merged_deaths_hb %>%
+#   filter(place_of_death %in% c("Home & other non-institution")) %>%
+#   filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
+#   group_by(year, place_of_death, ref_area, week_number) %>%
+#   summarise(
+#     deaths_all_causes = sum(deaths_all_causes),
+#     # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+#     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+#     .groups = "drop"
+#   ) %>% 
+#   mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+#   left_join(
+#     past_deaths_hb %>% filter(year != "average") %>% group_by(week_number, place_of_death, ref_area) %>% summarise(
+#                                        average_weekly_deaths = sum(number_of_deaths) / 5,
+#                                        .groups = "drop"
+#                                      ),
+#     by = c("week_number", "place_of_death", "ref_area")
+#   ) %>% 
+#   group_by(place_of_death, ref_area) %>%
+#   summarise(
+#     period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+#     deaths_all_causes = sum(deaths_all_causes),
+#     deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+#     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+#     mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+#     .groups = "drop"
+#   ) %>%
+#   mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
+#   kable(caption = "Total deaths at home in data so far for health boards.")
+```
 
+## Home death increase by HB, 2020 to now
+
+For this figure, I can't take annual average deaths to compare against because only for 2020 do we have the complete year, whereas for 2021 only part of the year is available.
+Instead, I add up weekly totals for the period in question, and average over the totals; in other words, if looking at the period 2020 w1 to 2021 w15, weeks 1-15 from 2015-2019 get used twice!
+
+```{r, fig.height=12, fig.width=8}
+(figure_map_ratio_home_deaths_to_historical_hb <-
   merged_deaths_hb %>%
   filter(place_of_death %in% c("Home & other non-institution")) %>%
-  filter(ref_area %in% c("Dumfries and Galloway", "Forth Valley")) %>%
   group_by(year, place_of_death, ref_area, week_number) %>%
@@
     deaths_all_causes = sum(deaths_all_causes),
-    # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
@@
   mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
-  kable(caption = "Total deaths at home in data so far for health boards.")
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase_deaths_over_historic), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B", labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(
+    axis.text = element_blank(), 
+    panel.grid = element_blank(), 
+    aspect.ratio = 1.4,  # needed for Scotland HB plot
+    plot.margin = margin(5,5,5,5),
+    ) +
+  labs(x="", y="",
+       title = str_wrap("Increased home deaths since 2020, by Health Board", width=40),
+       subtitle = str_wrap(date_range_text, width=50),
+       caption = paste0(c(source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease"
+       ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_blog,"/figure_map_ratio_home_deaths_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_to_historical_hb, width = 12, height = 14, dpi = 300, units = "cm")
 ```
+
+```{r}
+table_increase_home_deaths_by_hb <-
+  merged_deaths_hb %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  group_by(year, place_of_death, ref_area, week_number) %>%
+  summarise(
+    deaths_all_causes = sum(deaths_all_causes),
+    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+    .groups = "drop"
+  ) %>% 
+  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+  left_join(
+    past_deaths_hb %>% filter(year != "average") %>% group_by(week_number, place_of_death, ref_area) %>% summarise(
+                                       average_weekly_deaths = sum(number_of_deaths) / 5,
+                                       .groups = "drop"
+                                     ),
+    by = c("week_number", "place_of_death", "ref_area")
+  ) %>% 
+  group_by(place_of_death, ref_area) %>%
+  summarise(
+    period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+    deaths_all_causes = sum(deaths_all_causes),
+    deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+    mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+    .groups = "drop"
+  ) %>%
+  mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
+  arrange(percent_increase_deaths_over_historic) %>%
+  left_join(
+    weekly_covid_cases_hb %>%
+    filter(year == 2021 & week_number == weeks_available_2021) %>%
+    select(ref_area, crude_rate_positive_per100k),
+    by = c("ref_area")
+    ) %>%
+  mutate(
+    `Percent increase` = scales::percent(percent_increase_deaths_over_historic, accuracy = 0.1),
+    `COVID cases per 100,000` = format(round(crude_rate_positive_per100k, digits=0), big.mark = ",")
+    ) %>%
+  select(`Health Board` = ref_area, `Home deaths` = deaths_all_causes, `Percent increase`, `COVID cases per 100,000`)
+
+write_csv(x = table_increase_home_deaths_by_hb, file = paste0(dir_blog,"/table_increase_home_deaths_by_hb.csv"))
+```

### 2021-04-14 covid_nrs_place_of_death.Rmd
Commented out bit of code where I was trying to cluster case rates & home deaths - not used at the minute

@@
 
-```{r}
-cumulative_case_2020_based_grouping_la <-
-  weekly_covid_cases_la %>% 
-  filter(ref_area!="Scotland") %>%
-  filter(year == 2020) %>% 
-  filter(week_number == max(week_number)) %>%  # keep last week
-  select(ref_area, cumulative_positive, crude_rate_positive_per100k)
-
-cumulative_case_2020_based_grouping_hb <-
-  weekly_covid_cases_hb %>% filter(year == 2020) %>% 
-  filter(ref_area!="Scotland") %>%
-  filter(year == 2020) %>% 
-  filter(week_number == max(week_number)) %>%  # keep last week 
-  select(ref_area, cumulative_positive, crude_rate_positive_per100k)
-
-k_means_clustering_cases_and_per100k_2020_hb <- map(
-  .x = 2:4,
-  .f = ~kmeans(x=cumulative_case_2020_based_grouping_hb %>% select(cumulative_positive, crude_rate_positive_per100k), centers = .x, nstart = 1e3)
-) %>% set_names(nm=paste0("k",2:4))
-
-clusters_hb <-
-  bind_cols(
-    cumulative_case_2020_based_grouping_hb %>% select(ref_area),
-    map_dfc(.x = k_means_clustering_cases_and_per100k_2020_hb, .f = function(x) return(x$cluster))
-  )
-
-cumulative_case_2020_based_grouping_hb <-
-  cumulative_case_2020_based_grouping_hb %>%
-  left_join(clusters_hb, by="ref_area") %>%
-  mutate(
-    quartiles_cases = ntile(cumulative_positive, n=4),
-    quartiles_per100k = ntile(crude_rate_positive_per100k, n=4)
-  )
-
-# TODO: apply to LA also!
-
-hb_cases_plots <- list()
-
-hb_cases_plots$cumulative_cases_vs_per100k <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
-  geom_point() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_log10 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
-  geom_point() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases (log)",
-    y="COVID cases per 100,000 population (log)",
-    title="Cases vs. population rate (log scale axes)"
-  ) +
-  scale_x_log10() +
-  scale_y_log10() +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_cases))) +
-  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_per100k))) +
-  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate",
-    colour = "Quartile (Cases per 100,000)"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_k2 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k2))) +
-  geom_point(size=3, aes(colour=factor(k2))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_k3 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k3))) +
-  geom_point(size=3, aes(colour=factor(k3))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_k4 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k4))) +
-  geom_point(size=3, aes(colour=factor(k4))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots_assembled <- patchwork::wrap_plots(hb_cases_plots, ncol = 2)
-
-hb_cases_plots_assembled
+NOTE: I've commented this out - I ended up not using it. TODO: remove in a future iteration!
+
+```{r}
+# cumulative_case_2020_based_grouping_la <-
+#   weekly_covid_cases_la %>% 
+#   filter(ref_area!="Scotland") %>%
+#   filter(year == 2020) %>% 
+#   filter(week_number == max(week_number)) %>%  # keep last week
+#   select(ref_area, cumulative_positive, crude_rate_positive_per100k)
+# 
+# cumulative_case_2020_based_grouping_hb <-
+#   weekly_covid_cases_hb %>% filter(year == 2020) %>% 
+#   filter(ref_area!="Scotland") %>%
+#   filter(year == 2020) %>% 
+#   filter(week_number == max(week_number)) %>%  # keep last week 
+#   select(ref_area, cumulative_positive, crude_rate_positive_per100k)
+# 
+# k_means_clustering_cases_and_per100k_2020_hb <- map(
+#   .x = 2:4,
+#   .f = ~kmeans(x=cumulative_case_2020_based_grouping_hb %>% select(cumulative_positive, crude_rate_positive_per100k), centers = .x, nstart = 1e3)
+# ) %>% set_names(nm=paste0("k",2:4))
+# 
+# clusters_hb <-
+#   bind_cols(
+#     cumulative_case_2020_based_grouping_hb %>% select(ref_area),
+#     map_dfc(.x = k_means_clustering_cases_and_per100k_2020_hb, .f = function(x) return(x$cluster))
+#   )
+# 
+# cumulative_case_2020_based_grouping_hb <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   left_join(clusters_hb, by="ref_area") %>%
+#   mutate(
+#     quartiles_cases = ntile(cumulative_positive, n=4),
+#     quartiles_per100k = ntile(crude_rate_positive_per100k, n=4)
+#   )
+# 
+# # TODO: apply to LA also!
+# 
+# 
+# hb_cases_plots <- list()
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
+#   geom_point() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases",
+#     y="COVID cases per 100,000 population",
+#     title="Cases vs. population rate"
+#   ) +
+#   NULL
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k_log10 <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
+#   geom_point() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases (log)",
+#     y="COVID cases per 100,000 population (log)",
+#     title="Cases vs. population rate (log scale axes)"
+#   ) +
+#   scale_x_log10() +
+#   scale_y_log10() +
+#   NULL
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_cases))) +
+#   geom_point(size=3, aes(colour=factor(quartiles_cases))) +
+#   scale_colour_viridis_d() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases",
+#     y="COVID cases per 100,000 population",
+#     title="Cases vs. population rate"
+#   ) +
+#   NULL
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_per100k))) +
+#   geom_point(size=3, aes(colour=factor(quartiles_cases))) +
+#   scale_colour_viridis_d() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases",
+#     y="COVID cases per 100,000 population",
+#     title="Cases vs. population rate",
+#     colour = "Quartile (Cases per 100,000)"
+#   ) +
+#   NULL
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k_k2 <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k2))) +
+#   geom_point(size=3, aes(colour=factor(k2))) +
+#   scale_colour_viridis_d() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases",
+#     y="COVID cases per 100,000 population",
+#     title="Cases vs. population rate"
+#   ) +
+#   NULL
+# 
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k_k3 <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k3))) +
+#   geom_point(size=3, aes(colour=factor(k3))) +
+#   scale_colour_viridis_d() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases",
+#     y="COVID cases per 100,000 population",
+#     title="Cases vs. population rate"
+#   ) +
+#   NULL
+# 
+# 
+# hb_cases_plots$cumulative_cases_vs_per100k_k4 <-
+#   cumulative_case_2020_based_grouping_hb %>%
+#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k4))) +
+#   geom_point(size=3, aes(colour=factor(k4))) +
+#   scale_colour_viridis_d() +
+#   geom_label_repel() +
+#   # theme_minimal() +
+#   labs(
+#     x="COVID cases",
+#     y="COVID cases per 100,000 population",
+#     title="Cases vs. population rate"
+#   ) +
+#   NULL
+# 
+# 
+# hb_cases_plots_assembled <- patchwork::wrap_plots(hb_cases_plots, ncol = 2)
+# 
+# hb_cases_plots_assembled
 ```
@@
   # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
-  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), position = "dodge", colour = "grey10") +
-  geom_point(aes(x = cause_of_death, y = annual_deaths_mean), position = "dodge", colour = "grey10") +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), colour = "grey10") +
+  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
   scale_fill_viridis_d() +

### 2021-04-14 using_UPRN_to_determine_household_structure.Rmd
Created an Rmd for working out how to use UPRN to compute household members for deceased persons from CHI and death records data. Needs cleaning up a bit, and I should time it!

---
title: "Using UPRN to determine household composition"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Overview

Universal Property Reference Number (UPRN) is a high-resolution identifier for distinguishing properties. This allows research to include household structure and work out from administrative records which individuals live in the same property. In the Deaths at home study, we can use it to determine which people lived with others (as opposed to alone) in the same household at the time of death.

# Problem

Death records span a period of time, and so do address changes recorded in CHI data. Furthermore, we probably want to only include adults living in the same household because we are interested in the availability of informal care (with children we can assume the care relationship is the other way round - e.g. the deceased person was a carer/guardian for the child).

There is computational complexity involved because we will be consulting the entirety of CHI data, e.g. all people residing in Scotland and known to health services. It is therefore useful to plan the algorithm to reducing the data required as much as possible.

Because of the longitudinal nature of the data, we need to work out when address changes happened, to ensure that people lived in the same property at the same time.

Here is a high level overview of what we need to achieve:

**Find CHI records that match the UPRN of the deceased individuals, where**:

* the individuals in the CHI records were resident at that address at the time of the deceased person's death (specified by month & year)
  - CHI records have a date of address change linked to each address
  - from that we can define "episodes" of living at an address - starting from the date of change, and ending on the date of the next address entry (if different)
  - we are looking for records where the date of death falls within the "episode" of living at that UPRN
* they were aged 18+ at the date of the deceased person's death
* they were alive at the date of the deceased person's death
* note that we will likely encounter situations where a deceased person was a household member to another deceased person, so we can't simply separate the data into deceased and alive individuals

# Outline of algorithm

0. Extract from death records the deceased persons' dates of death & UPRN at death, and an id linking the same individual to their CHI records
1. Narrow down CHI data to individuals who ever lived at the same UPRN as the deceased persons, joining on the deceased person's id & death date
  - This is a many-to-many mapping - the same person may have lived with multiple deceased people over time; the same deceased person may have lived with multiple people over time; an alive person at time A becomes a deceased person at time B; various combinations thereof; etc.
2. Remove CHI records where the CHI person id matches the deceased person ID (this is so we don't double-count the deceased person in their own home)
3. Keep records only if age at date of death of deceased person was 18+
4. Compute period of residence at UPRN from change of address & date of death if the person died
5. Keep records if date of death of deceased person falls within period of residence
6. Finally, for every combination of deceased person id and UPRN, compute the number of distinct CHI ids - this is the number of same household members aged 18+

## Join operation

In step 1, we filter CHI records to those that ever included a UPRN matching a deceased person's UPRN & join the deceased person's id and date of death.
This can be implemented as a series of steps:

1. First, an `inner join` operation, keeping only those records that have a matching UPRN - this will massively reduce the amount of data in consideration so it's likely worth doing first; we are left only with records for individuals who ever lived at the relevant UPRN, and only with the record containing that UPRN
2. Add back UPRN records for the remaining individuals - this is needed to compute how long they lived at that UPRN
  - Start by taking the unique person ids of the data from the previous step
  - This followed by a `left_join` on the person ids from the CHI data - we now have all CHI records but only for the individuals who lived at a deceased person's UPRN at any point
3. For the records where the UPRN matches the deceased person's UPRN, compute the period of time at the UPRN:
  - sort address changes by date
  - get the next entry's address change date - this is the end of the period lived at relevant UPRN

Step 3 is the most computationally intensive, because it requires a search through each person's records separately! The smaller the dataset at this point, the faster this step can be.

Note that step 3 could be done first, followed by 1, skipping step 2. The practicality of this depends on the size of the data.

## Same algorithm simplified, using minimal data

1. Extract unique UPRN values from deaths file,
2. Filter CHI records to keep records with UPRN matching unique UPRN value from step 1,
3. Extract person ids from step 2,
4. Get CHI records matching person ids from step 3,
5. Order CHI records from step 4 by date and compute "episode" end date by using the following records' start dates
6. Filter CHI records so only those matching the UPRN from death records (keep only records from step 2, i.e. UPRN "episodes")
7. Join death records to CHI episodes on UPRN
8. Filter to keep CHI episodes where the date of death is between date address change and date "episode" end

This leaves us with one CHI record for each individual who lived with a deceased person at the same UPRN at the time the deceased person died.

# Coded example

```{r}
data_deaths <- tribble(
  ~id_death, ~uprn_death, ~date_death,
  "H", "d", ymd("2020-04-01"),
  "I", "e", ymd("2020-06-01"),
  "J", "f", ymd("2020-08-01")
)

# TODO: add date of birth? and explain what kinds of situations are included in the CHI data below

data_chi <- tribble(
  ~id_chi, ~uprn_chi, ~date_changed,
  "H", "d", ymd("2001-01-01"),
  "I", "e", ymd("2010-04-01"),
  "J", "f", ymd("2012-09-01"),  # the deceased people at their addresses
  "A", "d", ymd("2020-05-01"),  # lived at same address but moved after death
  "A", "n", ymd("2020-07-01"),  # at other address afterwards
  "B", "x", ymd("2020-01-01"),  # not relevant
  "C", "x", ymd("2020-01-01"),  # not relevant
  "D", "x", ymd("2020-01-01"),  # not relevant
  "E", "x", ymd("2020-01-01"),  # not relevant
  "F", "x", ymd("2020-01-01"),  # not relevant
  "G", "e", ymd("2020-01-01")  # moved in with I before I died
)

step1_1 <- data_chi %>% inner_join(data_deaths, by = c("uprn_chi"="uprn_death"))
step1_2 <- step1_1 %>% select(id_chi) %>% distinct %>% left_join(data_chi) %>% left_join(data_deaths, by = c("uprn_chi"="uprn_death"))
step1_3 <- step1_2 %>% group_by(id_chi) %>% arrange(date_changed) %>% mutate(date_moved_out = lead(date_changed, n=1, default = ymd("2022-01-01")))

step2 <- step1_3 %>% filter(id_chi != id_death)
step3 <- step2 #%>% filter(age_in_years_chi_at_death >= 18)  # TODO: include age/dob
step4_and_5 <- step3 %>% filter(date_changed < date_death & date_death < date_moved_out)
step6_final <- 
  step4_and_5 %>% 
  group_by(id_death, uprn_chi) %>% 
  summarise(number_of_other_adult_residents = n_distinct(id_chi), .groups = "drop") %>%
  tidyr::complete(nesting(id_death = data_deaths$id_death, uprn_chi = data_deaths$uprn_death), fill(number_of_other_adult_residents = 0))  # TODO: fix this, doesn't work!

```

# Simulating effect of operation order

Note: I was stuck for a while waiting for the combination of `dplyr::group_by()` and `dplyr::lead` to finish computing. At least on my setup, this is very slow - the "simpler" (less tidy) approach is to sort the dataset by id & date - this way we get continuous runs of records belonging to the same person - then compute the `lead` of the date (i.e. the end date for each episode) and `lead` of the person id. We can then set the "end date" to `NA` where the id doesn't match the next record's id (i.e. when one person's run of records ends and another person's begins).

I thought this was what `group_by()` + `lead` was doing under the hood, but for some reason it's incredibly slow!

```{r}
set.seed(42)
n_individuals = 5e6
n_addresses = 2e6
min_records = 1
max_records = 6

n_deaths_year = 6e5
years = 2015:2020
total_deaths = length(years)*n_deaths_year

# possible_dates_character <- 
#   as.character(
#     seq.Date(
#       from = as.Date("2015-01-01"),  # TODO: complete this, and then sample from the vector instead, possibly character to make it less intense!
#       to = as.Date("2020-12-31"),
#       by = "days"
#     ))

fake_deaths <- tibble(
  id_death = sample(x = 1:n_individuals, size = total_deaths, replace = FALSE), # unique ids of people who died
  uprn_death = sample(x=1:n_addresses, size = total_deaths, replace = TRUE),
  date_death = as.Date(x = rdunif(n = total_deaths,  # random dates
                                 a = as.integer(as.Date("2015-01-01")),
                                 b = as.integer(as.Date("2020-12-31"))
                                 ),
                      origin = "1970-01-01")
)

fake_chi <- 
  tibble(
    id_chi = 1:n_individuals,  # integer id
    n_rows = round(runif(n = n_individuals, min = min_records, max = max_records))  # a random number of records for each id, these are "uncounted" in the next step
    ) %>%
  uncount(weights = n_rows) %>%  # create N entries for each id based on above random N
  mutate(
    uprn_chi = sample(x = 1:n_addresses, size = nrow(.), replace = TRUE),  # sample addresses with replacement
    date_address_change = as.Date(x = rdunif(n = nrow(.),  # random dates
                                 a = as.integer(as.Date("2015-01-01")),
                                 b = as.integer(as.Date("2020-12-31"))),
                      origin = "1970-01-01")
    ) %>%
  arrange(id_chi, date_address_change) %>%
  group_by(id_chi) %>%
  mutate(episode_id = 1:n()) %>%
  ungroup

# 1. Extract unique UPRN values from deaths file,
# 2. Filter CHI records to keep records with UPRN matching unique UPRN value from step 1,
# 3. Extract person ids from step 2,
# 4. Get CHI records matching person ids from step 3,
# 5. Order CHI records from step 4 by date and compute "episode" end date by using the following records' start dates
# 6. Filter CHI records so only those matching the UPRN from death records (keep only records from step 2, i.e. UPRN "episodes")
# 7. Join death records to CHI episodes on UPRN
# 8. Filter to keep CHI episodes where the date of death is between date address change and date "episode" end

chi_ids_with_matching_uprn <-
  fake_chi %>% 
  filter(uprn_chi %in% fake_deaths$uprn_death) %>%  # Step 1 + 2: get unique UPRNs and only keep CHI records where they match the UPRN
  select(id_chi) %>%  # Step 3
  distinct  # we omitted getting unique UPRNs, so we'll get unique IDs here instead - this is faster than running unique() on the uPRNs for some reason!

chi_with_matching_uprn <- 
  chi_ids_with_matching_uprn %>% left_join(fake_chi, by = "id_chi")  # Step 4

sorted_by_date <- chi_with_matching_uprn %>%
  arrange(id_chi, date_address_change)  # Step 5

episodic_chi <-
  sorted_by_date %>%
  mutate(
    date_moved_address = lead(date_address_change, n=1, default = NA_Date_),  # get date from next entry
    id_date_end = lead(id_chi, n=1, default = NA_integer_),  # get person id from next entry - used for checking that the next entry actually belongs to the same person!
    date_moved_address = if_else(id_chi!=id_date_end, NA_Date_, date_moved_address)  # set date to NA if it doesn't belong to same person
  ) %>% 
  select(-id_date_end)  # remove next entry id, no longer needed

household_members <-
  episodic_chi %>%
  inner_join(fake_deaths, by = c("uprn_chi"="uprn_death")) %>%  # step 6 & 7: inner_join ensures only records with matching uprn are retained; step 7 joins the death records
  filter(date_death > date_address_change & date_death < date_moved_address)  # step 8: keep only records where the death occurred during uprn "episode"

## this takes forever
# sorted_by_date %>%
#   group_by(id_chi) %>%
#   mutate(episode_id = 1:n()) %>%
#   # mutate(date_moved_address = lead(date_address_change, n=1, default = ymd("2021-01-01"))) %>%  # takes forever! both with 5e5 and 5e6 people
#   mutate(date_moved_address = lead(date_address_change, n=1, default = NA_Date_)) %>%  # takes forever! both with 5e5 and 5e6 people
#   ungroup

number_of_household_members <-
  household_members %>%
  group_by(id_death) %>%
  summarise(num_household_members = n_distinct(id_chi), .groups = "drop") %>%
  tidyr::complete(id_death = fake_deaths$id_death, fill = list(num_household_members = 0))

# chis_with_matching_uprn_episodic <- 
#   chi_with_matching_uprn %>%
#   select(id_chi, episode_id) %>%
#   mutate(next_episode_id = episode_id + 1) %>%
#   left_join(fake_chi, by = c("id_chi", "next_episode_id"="episode_id")) %>%
#   mutate(date_episode_end = replace_na(date_address_change, replace = ymd("2021-01-01")))
#

```

### 2021-04-16 covid_nrs_place_of_death_data.Rmd
Split the .Rmd into one to deal with data and another to deal with visualisations for easier working

--- 2021-04-14 covid_nrs_place_of_death.Rmd
@@
 ---
-title: "Place of death in NRS COVID-19 deaths"
+title: "Place of death in NRS COVID-19 deaths: data processing for analysis"
 author: "Jan Savinc"
@@
 
-# Visualisations
-
-## Common elements
-
-```{r}
-loadfonts()  # load imported fonts
-# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
-
-theme_set(theme_minimal(base_size = 12) +
-            theme(panel.grid.minor = element_blank(),
-                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
-                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
-                  text=element_text(family="Calibri")
-            ))
-
-date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2021, '%d %B %Y')} (inclusive).")
-
-dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."
-
-only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
-only_2021_text <- glue::glue("Data for 2021 only include weeks 1-{format(weeks_available_2021)} (w/c {format(most_recent_date_available_2021)}).")
-
-source_nrs_text <- "Source: National Records of Scotland"
-source_phs_text <- "Source: Public Health Scotland"
-
-death_registrations_text <- "Note: all figures are based on date of registration."
-
-vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."
-
-```
-
-## Function to save files to multiple formats
-
-Saving to multiple formats is easy enough - the trouble is with saving PDF files with custom fonts, since they need to be embedded in the .pdf file. This requires:
-
-* telling R where to load the font from, using `extrafont::font_import()` - this was done in a previous step
-* registering fonts with R using `extrafont::loadfonts()` - note: this isn't strictly necessary, since once imported, `extrafont` keeps track of fonts to load when the `extrafont` package is loaded
-* ~~telling R where the `gs` executable is, i.e. the installation of ghostscript on the current machine~~
-* ~~embedding the fonts in the `.pdf` files once they've been written, using `extrafont::embed_fonts()`~~
-* instead of manually embedding fonts (which resulted in garbled text on my setup!), just use `device=device_cairo` as an argument to `ggsave()` when saving `.pdf` files and it just works
-
-```{r}
-save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {
-
-walk(
-  .x = extensions,
-  .f = function(ext) {
-  # if (tolower(ext)==".pdf") device <- cairo_pdf
-  ggsave(filename = paste0(filename, ext), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (tolower(ext)==".pdf") cairo_pdf else NULL)
-  }
-)
-
-# TODO: remove below bit, not needed when using device=device_cairo
-# if (any(str_detect(string = tolower(extensions), pattern = "pdf"))) {
-#   pdf_filename <- paste0(filename, extensions[which(any(str_detect(string = tolower(extensions), pattern = "pdf")))])
-#   embed_fonts(file = pdf_filename)  # overwrites file with one with font embedded
-# }
-# invisible(filename)  # return invisibly
-}
-
-```
-
-# Proportion home deaths over time
-
-```{r}
-figure_proportion_home_deaths_over_time <-
-  merged_proportions_of_deaths_by_place %>%
-  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
-  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
-  geom_point() +
-  geom_line(size = 1) +
-  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
-  scale_x_continuous(breaks = 2015:2021) +
-  scale_y_continuous(label = scales::percent) +
-  scale_colour_viridis_d(option="C") +
-  # theme_minimal() +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(
-    x = "Year", y= "Proportion of yearly deaths (%)",
-    colour = "",
-    subtitle = paste0("* ",only_2021_text),
-    caption = paste0(c(source_nrs_text),collapse="\n")
-    ) +
-  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
-
-figure_proportion_home_deaths_over_time
-```
-
-# Number of deaths by place over time
-
-```{r}
-(figure_number_of_deaths_by_place_by_year <-
-  bind_rows(
-    proportions_of_historical_deaths_by_place %>% filter(year==2019) %>% mutate(covid_status = "Non Covid-19 deaths"),  # add extra rows of 2019 data for plotting line between 2019 and 2020
-    proportions_of_historical_deaths_by_place %>% mutate(covid_status = "All deaths"),
-    proportions_of_current_non_covid_deaths_by_place %>% mutate(covid_status = "Non Covid-19 deaths"),
-    proportions_of_current_deaths_by_place %>% mutate(covid_status = "All deaths")
-  ) %>%
-  filter(!place_of_death %in% c("All","Other")) %>%  # don't show Other
-  filter(year != 2021) %>%  # don't include 2021
-  mutate(
-    place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])   # move home to 2nd position to match order in plot
-    ) %>%
-  ggplot(., aes(x=year, y=number_of_deaths, colour = place_of_death)) +
-  geom_line(data = . %>% filter(covid_status == "All deaths"), aes(group = place_of_death, linetype = covid_status)) +
-  geom_line(data = . %>% filter(covid_status == "Non Covid-19 deaths"), aes(group = place_of_death, linetype = covid_status)) +
-  geom_point(data = . %>% filter(!covid_status=="Non Covid-19 deaths" | !year==2019)) +
-  geom_text(data = . %>% filter(year==2016) %>% mutate(y_pos = c(3.1e4, 1.1e4, 1.8e4)), aes(y = y_pos, label = place_of_death), hjust = 0) +
-  scale_colour_viridis_d(option="C", end = 0.8, guide = FALSE) +
-  scale_linetype_manual(values = c("solid", "dashed")) +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    legend.box="vertical",
-    legend.margin=margin()
-    ) +
-  scale_y_continuous(labels = function(x) format(x, big.mark = ","), limits = c(0, NA)) +
-  labs(
-    x = "Year", y = "Deaths (N)",
-    caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
-  ) +
-  NULL
-)
-
-save_output_file(filename = paste0(dir_outputs,"/figure_number_of_deaths_by_place_by_year"), extensions = c(".pdf",".svg",".png"), plot = figure_number_of_deaths_by_place_by_year, width = 15, height = 11, dpi = 300, units = "cm")
-```
-
-## Overall, regardless of place
-
-```{r}
-number_of_covid_deaths_2020 <- total_deaths_overall %>% filter(year==2020 & place_of_death == "All") %>% .$deaths_covid_related
-proportion_of_covid_deaths_2020 <- total_deaths_overall %>% filter(year==2020 & place_of_death == "All") %>% {scales::percent(.$deaths_covid_related / .$deaths_all_causes, accuracy = 0.1)}
-
-(figure_number_of_deaths_by_year <-
-bind_rows(
-  merged_proportions_of_deaths_by_place %>%  # total deaths regardless of place
-    select(year, total_deaths) %>% 
-    distinct %>%
-    mutate(covid_status = "All deaths")
-    ,
-  merged_proportions_of_non_covid_deaths_by_place %>%
-    select(year, total_deaths) %>%
-    distinct %>%
-    # filter(year %in% 2019:2020) %>%
-    filter(year %in% 2020) %>%
-    mutate(covid_status = "Non Covid-19 deaths")
-  ) %>%
-  filter(year!=2021) %>%  # remove 2021
-  ggplot(
-    ., aes(x = year, y = total_deaths, fill = covid_status)
-  ) +
-  scale_fill_manual(values = c("red","blue")) +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    legend.box="vertical",
-    legend.margin=margin()
-    ) +
-  geom_col(position = "dodge", colour = "grey20") +
-  # geom_line() +
-  # geom_point() +
-  scale_x_continuous(breaks = 2015:2020) +
-  # annotate(geom = "text", x = 2016, y = 65e3, hjust = 0, label = glue::glue("2020 COVID-related deaths N={format(number_of_covid_deaths_2020, big.mark=',')}\n({proportion_of_covid_deaths_2020} of all deaths)")) +
-  scale_y_continuous(labels = function(x) format(x, big.mark = ","), limits = c(0, NA)) +
-  labs(
-    x = "Year", y = "Deaths (N)",
-    subtitle = glue::glue("2020 COVID-related deaths N={format(number_of_covid_deaths_2020, big.mark=',')}\n({proportion_of_covid_deaths_2020} of all deaths)"),
-    caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
-  ) +
-  NULL
-)
-
-save_output_file(filename = paste0(dir_outputs,"/figure_number_of_deaths_by_year"), extensions = c(".pdf",".svg",".png"), plot = figure_number_of_deaths_by_year, width = 12, height = 10, dpi = 300, units = "cm")
-```
-
-# Cumulative deaths
-
-Note: I've left this unfinished, it doesn't tell a better story than the weekly figures!
-
-```{r}
-text_non_covid_deaths_equivalent_to_all_cause_before_wk12 <- "Note: all-cause deaths and non-covid deaths were considered equivalent before 2020, week 12"
-
-text_annotation_cunulative_all_cause_deaths <- "Hospital all-cause deaths were within the historical range; home & care home deaths exceeded historical ranges."
-
-# TODO: these figures over- and underestimate yearly cumulative deaths (because they use weekly min/max, as opposed to min/max of cumulative deaths)
-
-figure_cumulative_deaths_by_place <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(!place_of_death %in% c("All","Other")) %>%
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-  mutate(deaths_non_covid = if_else(condition = is.na(deaths_non_covid), true = deaths_all_causes, false = deaths_non_covid)) %>%  # replace NA in non-covid deaths with all cause deaths - before covid, non-covid deaths are equivalent to all deaths
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>%
-  mutate(deaths_type = case_when(
-    deaths_type == "deaths_all_causes" ~ "All-cause deaths",
-    deaths_type == "deaths_non_covid" ~ "Non-COVID deaths",
-    TRUE ~ NA_character_
-  )) %>%
-  group_by(place_of_death, deaths_type) %>%
-  mutate(
-    cumulative_deaths = cumsum(replace_na(deaths, replace = 0)),
-    cumulative_historical_mean = cumsum(replace_na(deaths_mean, replace = 0)),
-    cumulative_historical_min = cumsum(replace_na(deaths_min, replace = 0)),
-    cumulative_historical_max = cumsum(replace_na(deaths_max, replace = 0))
-    ) %>%
-  ungroup %>%
-  ggplot(aes(x = date_w_c)) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
-  geom_ribbon(aes(ymin = cumulative_historical_min, ymax = cumulative_historical_max, fill = place_of_death), alpha = 0.4) +
-  geom_line(aes(y = cumulative_historical_mean, colour = place_of_death), linetype = "dashed", alpha = 0.8, show.legend = FALSE) +
-  geom_line(aes(y = cumulative_deaths, colour = place_of_death)) +
-  facet_wrap(~deaths_type) +
-  scale_colour_viridis_d(option = "C") +
-  scale_fill_viridis_d(option = "C") +
-  theme(
-    legend.position = "top",
-    plot.caption = element_text(size = 10, colour = "gray60"),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = "Date w/c",
-       y = "Cumulative deaths (N)",
-       title = "Cumulative deaths by place of death",
-       # subtitle = paste0(c(only_2020_text,light_shaded_area_shows_difference),collapse="\n"),
-       caption = paste0(c(text_non_covid_deaths_equivalent_to_all_cause_before_wk12,death_registrations_text,source_nrs_text), collapse="\n")
-  ) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_cumulative_deaths_by_place"), extensions = c(".pdf",".svg",".png"), plot = figure_cumulative_deaths_by_place, width = 15, height = 12, dpi = 300, units = "cm")
-
-figure_cumulative_deaths_by_place
-```
-
-# Number of deaths visualisations
-
-## Total deaths in 2020 by age group & place of death
-
-```{r, fig.width=12, fig.height=7, warning=FALSE}
-bar_and_error_bars_2020_text <- "Bars are 2020 all-cause deaths, point and bars shows annual mean and range 2015-2019"
-excluded_other_place_2020 <- glue::glue("Excluding records with 'Other' place of death, N={total_deaths_age %>% filter(year==2020 & place_of_death=='Other') %>% .$deaths_all_causes %>% sum}")
-
-light_shaded_area_shows_difference <- "Lighter area in the top of bar shows difference between 2020 deaths and historical maximum 2015-2019"
-
-# TODO: make the distinction max vs 2020 figures clearer! maybe annotate?
-
-figure_annual_deaths_by_age_group_2020 <-
-  total_deaths_age %>% 
-  filter(year==2020) %>%
-  filter(place_of_death != "Other") %>%
-  ggplot(aes(x = age, y = deaths_all_causes)) +
-  # geom_col(fill = "#4477AA", alpha = 0.5, colour = "grey10") +
-  geom_col(fill = "grey20", colour = "grey10", alpha = 0.5) +
-  geom_col(aes(x=age, y=annual_deaths_max), fill = "#4477AA", colour = "grey10", alpha = 0.5) +  # historical maximum over the actual figure
-  geom_point(aes(x=age, y=annual_deaths_mean)) +
-  geom_errorbar(aes(x = age, ymin = annual_deaths_min, ymax = annual_deaths_max)) +
-  facet_wrap(~place_of_death) +
-  scale_y_continuous(labels = function(x) format(x,big.mark=",")) +
-  labs(x = "Age",
-       y = "Deaths (N)",
-       title = "Deaths in 2020 by age group & place of death",
-       subtitle = paste0(c(only_2020_text,light_shaded_area_shows_difference),collapse="\n"),
-       caption = paste0(c(excluded_other_place_2020,bar_and_error_bars_2020_text,death_registrations_text,source_nrs_text), collapse="\n")
-  )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_annual_deaths_by_age_group_2020"), extensions = c(".pdf",".svg",".png"), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300, units = "in")
-
-figure_annual_deaths_by_age_group_2020
-```
-
-## Recorded deaths in Scotland, by week number & location, with annotations (after DH's plots)
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
-
-annotations_figure_place_of_death <-
-# datapasta::tribble_paste(merged_deaths_overall %>% select(place_of_death))
-  tibble::tribble(
-    ~place_of_death, ~x, ~y, ~label,
-                 "Hospital", 20L, 850, "Shaded area shows range of 2015-2019 deaths; dashed line shows weekly average",
-                 "Hospital", 25L, 250, "Hospital deaths dipped below historical minimum and remained low until Winter.",
-                "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
-                     "Home & other non-institution", 35L, 500, "The gap between historical high and both types of death at home was high throughout."
-                  ) %>% 
-  mutate(
-    place_of_death = factor(place_of_death, levels = levels(merged_deaths_overall$place_of_death)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
-    label = str_wrap(label, width = 40)
-    )
-
-figure_place_of_death_after_dh <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(!place_of_death %in% c("All","Other")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  geom_text(data = annotations_figure_place_of_death, aes(x=x,y=y, label=label)) +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_grid(~place_of_death) +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death_after_dh"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death_after_dh, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_place_of_death_after_dh
-```
-
-### Place of death with historic range, updated version
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
-annotations_figure_place_of_death_explaining_historic_range <- tribble(
-   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
-  "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
-  "curve", "Hospital", NA, 29, 830, 32, 550,
-  "text", "Hospital", "Historic mean", 22, 800, NA, NA,
-  "curve", "Hospital", NA, 22, 780, 26, 510
-) %>% 
-  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-
-figure_place_of_death <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(!place_of_death %in% c("All","Other")) %>%
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(
-    legend.position = "top",
-    plot.caption = element_text(size = 10, colour = "gray60"),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = "Date w/c",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
-  geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death, width = 28, height = 14, dpi = 300, units = "cm")
-
-figure_place_of_death
-```
-
-## All of Scotland, deaths at home, for policy briefing
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-# TODO: add start of lockdowns 1-3?
-# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
-total_deaths_at_home_in_period <- glue::glue("Total deaths at home N={format(merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death %in% c('Home & other non-institution')) %>% .$deaths_all_causes %>% sum, big.mark=',')}")
-
-figure_deaths_at_home_policy_briefing <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 21, y = 575, hjust=0) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  NULL
-
-## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_policy_briefing"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_policy_briefing, width = 14, height = 11, units = "cm", dpi = 300)
-
-figure_deaths_at_home_policy_briefing
-```
-
-## All of Scotland, deaths at home - date axis
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60"), plot.margin = margin(5,15,5,5)) +
-  labs(x = "Date w/c",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 10, year_number = 2020), y = 100, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 10, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 14, year_number = 2020), y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic mean", colour = "grey20") +
-  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  ## annotate year
-  geom_text(aes(x=ymd("2021-01-07"), y=30, label = "2021"), hjust = 0, colour = "grey40") +
-  NULL
-
-## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home, width = 18, height = 15, units = "cm", dpi = 300)
-
-figure_deaths_at_home
-```
-
-## All of Scotland, deaths at home - week number
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home_wwek_number <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_vline(xintercept = 53.5, linetype = "dotted", colour="grey40", show.legend = FALSE) +  # line should be between last week 2020 and 1st week 2021
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  geom_text(aes(x = 10, y = 100, label = "Historic range (min & max)"), colour = "grey20") +
-  geom_curve(aes(x = 10, xend = 14, y = 110, yend = 250), curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  geom_text(aes(x = 20, y = 50, label = "Historic mean"), colour = "grey20") +
-  geom_curve(aes(x = 20, xend = 25, y = 60, yend = 270), curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  ## annotate year
-  geom_text(aes(x=54, y=30, label = "2021"), hjust = 0, colour = "grey40") +
-  NULL
-
-## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_wwek_number"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_wwek_number, width = 18, height = 15, units = "cm", dpi = 300)
-
-figure_deaths_at_home_wwek_number
-```
-
-## Deaths at home in 2020, all of Scotland
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home_2020 <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  filter(year == 2020) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_grid(~place_of_death) +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       # subtitle = date_range_text,
-       subtitle = "Data from 2020 shown, starting with w/c 30 Dec 2019.",
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_2020
-```
-
-## Deaths in hospital, date axis
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_in_hospital <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Hospital")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60"), plot.margin = margin(5,15,5,5)) +
-  labs(x = "Date w/c",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  annotate(geom = "text", x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), y = 850, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom = "curve", x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 32, year_number = 2020), y = 830, yend = 550, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), y = 800, label = "Historic mean", colour = "grey20") +
-  annotate(geom = "curve", x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 26, year_number = 2020), y = 780, yend = 510, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  ## annotate year
-  geom_text(aes(x=ymd("2021-01-07"), y=30, label = "2021"), hjust = 0, colour = "grey40") +
-  NULL
-
-## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_in_hospital"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_in_hospital, width = 18, height = 15, units = "cm", dpi = 300)
-
-figure_deaths_in_hospital
-```
-
-## Deaths at home, by health board
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-island_health_boards_merged <- "Note: Orkney, Shetland, & Western health boards were merged into 'Island HBs'; Borders & Dumfries and Galloway health boards were merged into 'B,D&G HB'"
-vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"
-
-figure_deaths_at_home_by_health_board <-
-  merged_deaths_hb %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~ref_area, scales = "free_y") +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_health_board"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_health_board, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_by_health_board
-```
-
-## Deaths at home with 2-week moving average, by health board
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-moving_average_text <- "All figures represent deaths averaged over past 2 weeks; this includes historical range and historical mean."
-
-figure_deaths_at_home_2week_average_by_health_board <-
-  merged_deaths_hb %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid), 
-               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = ma2w_deaths, colour = deaths_type)) +
-  geom_line(aes(y = ma2w_deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~ref_area, scales = "free_y") +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(moving_average_text, vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_health_board"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_2week_average_by_health_board
-```
-
-## Deaths at home, by local authority
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home_by_local_authority <-
-  merged_deaths_la %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~ref_area, scales = "free_y") +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_local_authority"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_local_authority, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_by_local_authority
-```
-
-## Deaths at home with 2-week moving average, by local authority
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home_2week_average_by_local_authority <-
-  merged_deaths_la %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid), 
-               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = ma2w_deaths, colour = deaths_type)) +
-  geom_line(aes(y = ma2w_deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_wrap(~ref_area, scales = "free_y") +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(moving_average_text, vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2week_average_by_local_authority"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_2week_average_by_local_authority
-```
-
-## Deaths at home by age
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_deaths_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_age %>% filter(year==2021) %>% .$week_number)} in 2021.")
-
-figure_deaths_at_home_by_age <-
-  merged_deaths_age %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  filter(!age %in% c('0','1-14')) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  # facet_wrap(~age, scales = "free_y") +
-  facet_wrap(~age) +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text,removed_under_15_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_age"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_by_age
-```
-
-### Percent increase in home deaths over historic average, by age
-
-```{r}
-# TODO: can this be generalised in a function?
-merged_deaths_age %>%
-  # filter(place_of_death %in% c("Home & other non-institution")) %>%
-  group_by(year, place_of_death, age, week_number) %>%
-  summarise(
-    deaths_all_causes = sum(deaths_all_causes),
-    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-    .groups = "drop"
-  ) %>% 
-  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
-  left_join(
-    past_deaths_age %>% filter(year != "average") %>% group_by(week_number, place_of_death, age) %>% summarise(
-                                       average_weekly_deaths = sum(number_of_deaths) / 5,
-                                       .groups = "drop"
-                                     ),
-    by = c("week_number", "place_of_death", "age")
-  ) %>% 
-  group_by(place_of_death, age) %>%
-  summarise(
-    period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
-    deaths_all_causes = sum(deaths_all_causes),
-    deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-    mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
-    .groups = "drop"
-  ) %>%
-  mutate(
-    percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
-    percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
-    ) %>%
-  (function(x) {
-    bind_rows(
-      x,
-      x %>% group_by(age, period) %>% summarise(
-        deaths_all_causes = sum(deaths_all_causes),
-        deaths_non_covid = sum(deaths_non_covid),
-        mean_annual_deaths_2015_2019 = sum(mean_annual_deaths_2015_2019)
-      ) %>%
-        mutate(
-          percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
-          percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
-        )
-    )
-  }) %>%
-  View()
-  # kable(caption = "Total deaths at home in data so far for age groups.")
-```
-
-## Deaths at home by sex
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home_by_sex <-
-  merged_deaths_sex %>%
-  mutate(sex = if_else(sex=="F", "Female", "Male")) %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  facet_grid(~sex) +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_sex"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_sex, width = 12, height = 9, dpi = 300, units = "in")
-
-figure_deaths_at_home_by_sex
-```
-
-# Deaths by cause of death
-
-## Annual deaths by cause
-
-```{r}
-(figure_annual_deaths_by_cause <-
-  total_deaths_cause_of_death %>%
-  filter(year == 2020 & cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
-  # pivot_longer(cols = c(deaths, annual_deaths_mean), names_to = "deaths_type", values_to = "deaths") %>%
-  ggplot() +
-  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
-  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), colour = "grey10") +
-  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
-  geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
-  scale_fill_viridis_d() +
-  facet_wrap(~place_of_death) +
-  NULL
-)
-```
-
-## By cause, at home
-
-```{r, fig.width=12, fig.height=9, warning=FALSE}
-(figure_deaths_at_home_by_cause <-
-  merged_deaths_cause_of_death %>%
-  filter(str_starts(place_of_death,"Home")) %>%
-  filter(cause_of_death != "All") %>%
-  # filter(!place_of_death %in% c("All","Other")) %>%
-  pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
-  filter(!(deaths_type=="deaths_mean"&cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-  ggplot(aes(x = date_w_c)) +
-  # annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_line(aes(y = number_of_deaths, colour = cause_of_death, linetype = deaths_type)) +
-  scale_colour_viridis_d() +
-  scale_linetype_manual(labels = c("Historic mean 2015-2019","Deaths since 2020"), values = c("dashed","solid")) +
-  # geom_line(aes(y = number_of_deaths  , colour = deaths_type)) +
-  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  facet_wrap(cause_of_death~place_of_death, scales = "free_y") +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    plot.caption = element_text(size = 10, colour = "gray60"),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = "Date w/c",
-       y = "Weekly Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
-       ) +
-  NULL
-)
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_cause, width = 28, height = 14, dpi = 300, units = "cm")
-```
-
-# Case counts & home deaths - comparing plots
-
-## Scotland-wide figure
-
-```{r}
-# TODO: wokr out how to include both... two panels? two axes on same plot?
-
-ratio_max_weekly_deaths_to_cases <- 600 / 16e3
-
-figure_weekly_deaths_to_cases_part_1 <-
-  merged_deaths_overall %>%
-  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
-  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = week_number_run_over)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  # geom_line(aes(y = weekly_positive * ratio_max_weekly_deaths_to_cases), colour="purple") +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  # scale_y_continuous(sec.axis = sec_axis(trans = ~./ratio_max_weekly_deaths_to_cases, name = "Weekly cases")) +
-  # ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  # theme_minimal() +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  # annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       )
-
-figure_weekly_deaths_to_cases_part_2 <-
-  merged_deaths_overall %>%
-  filter(sex=="all" & age=="all" & str_starts(place_of_death, "Home")) %>%
-  left_join(weekly_covid_cases_overall, by=c("year","week_number","date_w_c","week_number_run_over","ref_area")) %>%
-  ggplot(aes(x=week_number_run_over, y=weekly_positive)) +
-  geom_line(size=1) +
-  geom_vline(xintercept = 54, linetype = "dotted") +
-  # theme_minimal() +
-  facet_grid(~ref_area) +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  scale_x_continuous(breaks = seq(0,60,by=10)) +
-  scale_y_continuous(labels = function(x) format(x, big.mark=",")) +
-  labs(
-    x = "Week number", y = "COVID-19 positive cases (N)",
-    subtitle = "Note the different scale for positive cases!",
-    caption = paste0(c(source_phs_text), collapse="\n")
-  )
-
-figure_weekly_deaths_to_cases <- wrap_plots(figure_weekly_deaths_to_cases_part_1,figure_weekly_deaths_to_cases_part_2, ncol = 1)
-
-## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_outputs,"/figure_weekly_deaths_to_cases"), extensions = c(".pdf",".svg",".png"), plot = figure_weekly_deaths_to_cases,width = 15, height = 15, units = "cm", dpi = 300)
-
-figure_weekly_deaths_to_cases
-```
-
-# Map-based visualisations
-
-## Ratio of 2020 home deaths to historical home deaths by LA / HB
-
-```{r}
-areas_relatively_higher_number_home_deaths_text <- "Lighter areas represent areas with relatively higher number of home deaths in 2020 compared to historical mean"
-home_deaths_ratio_text <- "Ratio computed from total home deaths in 2020 divided by mean annual home deaths 2015-2019"
-
-## max/min used for setting scale of fill colour so it's consistent between plots!
-maximum_ratio_between_hb_and_la_home_deaths <-
-  bind_rows(
-    total_deaths_la,
-    total_deaths_hb
-  ) %>%
-  filter(year==2020 & place_of_death == "Home & other non-institution") %>%
-  .$ratio_annual_deaths_to_historical_mean %>%
-  max
-
-minimum_ratio_between_hb_and_la_home_deaths <-
-  bind_rows(
-    total_deaths_la,
-    total_deaths_hb
-  ) %>%
-  filter(year==2020 & place_of_death == "Home & other non-institution") %>%
-  .$ratio_annual_deaths_to_historical_mean %>%
-  min %>%
-  round(digits=1)
-```
-
-### LA
-
-```{r, fig.height=12, fig.width=8}
-# TODO: work out the size of these plots
-figure_map_ratio_home_deaths_2020_to_historical_la <-
-  total_deaths_la %>%
-  filter(
-    year == 2020 & place_of_death == "Home & other non-institution"
-  ) %>%
-  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
-  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
-  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
-  geom_sf(colour="grey10") +
-  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
-  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
-  theme(axis.text = element_blank(), panel.grid = element_blank()) +
-  labs(x="", y="",
-       title = str_wrap("Increased home deaths in 2020 relative to historical mean 2015-2019, by Local Authority", width=50),
-       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
-       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       caption = paste0(c(source_nrs_text), collapse="\n"),
-       fill = "Home death\nincrease"
-       ) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 12, height = 14, dpi = 300, units = "cm")
-
-figure_map_ratio_home_deaths_2020_to_historical_la
-```
-
-#### Grayscale version of the same
-
-```{r}
-figure_map_ratio_home_deaths_2020_to_historical_la_bw <-
-  figure_map_ratio_home_deaths_2020_to_historical_la +
-  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) scales::percent(x, accuracy = 1)) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la_bw, width = 12, height = 14, dpi = 300, units = "cm")
-```
-
-### HB
-
-```{r, fig.height=12, fig.width=8}
-figure_map_ratio_home_deaths_2020_to_historical_hb <-
-  total_deaths_hb %>%
-  filter(
-    year == 2020 & place_of_death == "Home & other non-institution"
-  ) %>%
-  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
-  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
-  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
-  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
-  geom_sf(colour="grey10") +
-  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
-  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
-  theme(
-    axis.text = element_blank(), 
-    panel.grid = element_blank(), 
-    aspect.ratio = 1.4,  # needed for Scotland HB plot
-    plot.margin = margin(5,5,5,5),
-    # legend.position = c(1.2,0.4)  # this was manually tweaked until just right  # problem - edge gets cut
-    ) +
-  labs(x="", y="",
-       title = str_wrap("Increased home deaths in 2020 relative to historical mean 2015-2019, by Health Board", width=50),
-       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
-       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       caption = paste0(c(source_nrs_text), collapse="\n"),
-       # fill = "Ratio"
-       fill = "Home death\nincrease"
-       ) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 12, height = 14, dpi = 300, units = "cm")
-
-figure_map_ratio_home_deaths_2020_to_historical_hb
-```
-
-#### Grayscale version of the same
-
-```{r}
-figure_map_ratio_home_deaths_2020_to_historical_hb_bw <-
-  figure_map_ratio_home_deaths_2020_to_historical_hb +
-  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) scales::percent(x, accuracy = 1)) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb_bw, width = 12, height = 14, dpi = 300, units = "cm")
-```
-
-## Per capita cases of COVID-19 by region
-
-```{r}
-areas_relatively_higher_number_cases <- "Lighter areas represent areas with relatively higher number of COVID-19 cases per 100,000 population in 2020"
-```
-
-### LA
-
-```{r, fig.height=12, fig.width=8}
-figure_map_per_capita_covid_cases_2020_la <-
-  weekly_covid_cases_la %>%
-  filter(year == 2020 & week_number == 53) %>%
-  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
-  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
-  geom_sf(colour="grey10") +
-  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
-  theme(axis.text = element_blank(), panel.grid = element_blank()) +
-  labs(x="", y="",
-       title = str_wrap("Total COVID-19 cases in 2020 per 100,000 population, by Local Authority",width=50),
-       subtitle = "",
-       caption = paste0(c(source_phs_text), collapse="\n"),
-       fill = str_wrap("Cases per 100,000 population", width = 15)
-       ) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la, width = 12, height = 14, dpi = 300, units = "cm")
-
-figure_map_per_capita_covid_cases_2020_la
-```
-
-#### Grayscale version of the same
-
-```{r}
-figure_map_per_capita_covid_cases_2020_la_bw <-
-  figure_map_per_capita_covid_cases_2020_la +
-  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) format(x, big.mark=",")) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la_bw, width = 12, height = 14, dpi = 300, units = "cm")
-```
-
-### HB
-
-```{r, fig.height=12, fig.width=8}
-figure_map_per_capita_covid_cases_2020_hb <-
-  weekly_covid_cases_hb %>%
-  filter(year == 2020 & week_number == 53) %>%
-  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
-  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
-  geom_sf(colour="grey10") +
-  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
-  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
-  labs(x="", y="",
-       title = str_wrap("Total cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
-       subtitle = "",
-       caption = paste0(c(source_phs_text), collapse="\n"),
-       fill = str_wrap("Cases per 100,000 population", width = 15)
-       ) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb, width = 12, height = 14, dpi = 300, units = "cm")
-
-figure_map_per_capita_covid_cases_2020_hb
-```
-
-#### Grayscale version of the same
-
-```{r}
-figure_map_per_capita_covid_cases_2020_hb_bw <-
-  figure_map_per_capita_covid_cases_2020_hb +
-  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) format(x, big.mark=",")) +
-  NULL
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb_bw, width = 12, height = 14, dpi = 300, units = "cm")
-```
-
-## Merged maps of home death ratios & covid cases
-
-### HB
-
-```{r}
-# TODO: work out where all the extra space comes from around plot & legends
-
-figure_map_home_deaths_ratios_2020_and_per_capita_cases <-
-  wrap_plots(
-    figure_map_ratio_home_deaths_2020_to_historical_hb + 
-      labs(
-        caption = source_nrs_text,   # keep source captions only
-        fill = str_wrap("Home deaths increase over historic mean",width = 20),
-        subtitle = NULL,
-        title = str_wrap("Increased home deaths 2020", width = 20),
-        x = NULL,
-        y = NULL
-        ) +
-      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
-    ,
-      # theme(legend.position = "left"),
-    figure_map_per_capita_covid_cases_2020_hb + 
-      labs(
-        caption = source_phs_text,
-        title = str_wrap("COVID cases 2020", width = 20),
-        x = NULL,
-        y = NULL
-        ) + 
-      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
-    ,
-    nrow = 1  # force to same row  
-  )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_map_home_deaths_ratios_2020_and_per_capita_cases"), extensions = c(".pdf",".svg",".png"), plot = figure_map_home_deaths_ratios_2020_and_per_capita_cases, width = 20, height = 10, dpi = 300, units = "cm")
-```
-
 # Save most recent workspace for further analyses

### 2021-04-16 covid_nrs_place_of_death_visualisations.Rmd
Split the .Rmd into one to deal with data and another to deal with visualisations for easier working

--- 2021-03-18 covid_nrs_place_of_death.Rmd
@@
 
-# Set up SPARQL query
-
-Note: not needed if using `opendatascot`
-To access the latest NRS data using their API, we first setup a SPARQL query.
-
-```{r}
-# endpoint <- "http://statistics.gov.scot/sparql.json"
-endpoint <- "http://statistics.gov.scot/sparql"
-```
-
 # Copyright attribution
@@
 
-# Context of lockdowns
-
-A recent timeline of COVID-19 events in Scotland is [available here](https://spice-spotlight.scot/2021/01/29/timeline-of-coronavirus-covid-19-in-scotland/).
-Including these might help in the interpreration fo the figures, though the figures are already pretty busy.
-This is a potential future TODO!
-
-# Load info about latest data
-
-A separate script checks for updated data and downloads it, also updating these files with the locations of the downloaded data:
-
-```{r}
-file_data_sources <- "./table_of_data_sources.csv"
-file_historical_data_sources <- "./table_of_historical_data_sources.csv"
-file_case_trends_data_sources <- "./table_of_case_trends_data_sources.csv"
-
-table_of_data_sources <- read_csv(file = file_data_sources)
-table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
-table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)
-
-table_of_data_sources %>%
-  knitr::kable(caption = "Latest data used in this report.")
-```
-
-## Set up output directory
-
-Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:
-
-```{r}
-latest_data_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
-
-dir_outputs <- paste0("./outputs/",latest_data_date)
-if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
-```
-
-# Import & wrangle data
-
-We're importing two kinds of data:
-
-1. Weekly deaths involving COVID-19 from Scottish gov't open data service, broken down by week & place of death (home or institution)
-2. Historical deaths 2015-2019, grouped by week & place of death
-
-For both of these, we'll also look at crosstabulated figures for:
-
-* Health board
-* Local Authority
-* Sex
-* Age
-* ~~Ethnicity~~ *Note: not sure this is available*
-
-There are two ways of obtaining the data:
-
-1.
-~~Using their SPARQL-based API ("data cube"), which can be accessed using the `opendatascot` package in R; the data can be seen on [the Scottish Gov't open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)~~
-    - Update: it's difficult to keep track of data updates using the API - it's simpler to query the .xlsx/.csv files (next bullet point) for when they were last modified and download the weekly update once it's been released!
-2.
-There are a number of datasets, including the demographic/geographic crosstabulations available on the "Related Statistics" section of the [deaths involving COVID-19 website]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland), under the *Ad-hoc queries section*.
-They are updated weekly, and also contain historical figures.
-
-Note that the [*About* tab of the "data cube" interface](https://statistics.gov.scot/resource?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19) says the following about Health Board & Local Authority figures:
-
->Health Board and Local Authority figures include non-residents.
-Deaths are allocated to areas based on the usual residence of the deceased.
-If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.
-
-This means we'll need to download those figures from the website rather than using the API!
-
-Note that the weekly deaths due to COVID-19 in 2020 & 2021 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the weekly figures for individual years & the appropriate cross-tabulations.
-
-```{r}
-# TODO: remove this section, params not used anymore
-# # "unpack" the data source tables
-# table_of_data_sources <- params$data_sources
-# table_of_historical_data_sources <- params$historical_data_sources
-# table_of_case_trends_data_sources <- params$case_trends_data_sources
-```
-
-## Parsing week numbers
-
-In the `opendatascot`-imported data, `refPeriod` refers to the week number (called `week of occurrence` in the `.xlsx`/`.csv`-imported data has), as defined by [ISO 8601](https://en.wikipedia.org/wiki/ISO_week_date), and can be computed from a date with `lubridate::isoweek()`; the first week of 2020 is the week starting with Monday 2019-12-30, ~~and the corresponding start dates for each week can be computed by adding weeks using `lubridate::dweeks()` to that date.~~ This is non-trivial, because the number of iso weeks in a year isn't constant.
-The `ISOweek::ISOweek2date()` function deals with this issue in a crude manner, and allows us to compute the date (w/c on date) from the year & week number.
-It relies on being provided a valid week number (some years have 52 and some have 53), but it's probably safe to assume that the NRS data has valid dates.
-
-Invalid week numbers will produce a date in the next year: e.g.
-week 53 in 2021 would actually start in 2022, which makes it invalid.
-We can test for that by checking that the year provided is the same year in the date produced.
-
-We'll also compute a 'run-over' week number of 2021 in case we plot both years on a contibuous X-axis - in this case, week 1 of 2021 follows week 53 of 2020, so we start counting from 54 in 2021.
-
-```{r}
-date_first_week_2020 <- ymd("2019-12-30")
-ISOweek::ISOweek2date(weekdate = "2019-W53-1")
-ISOweek::ISOweek2date(weekdate = "2020-W01-1")
-ISOweek::ISOweek2date(weekdate = "2020-W52-1")
-ISOweek::ISOweek2date(weekdate = "2020-W53-1")
-ISOweek::ISOweek2date(weekdate = "2021-W01-1")
-ISOweek::ISOweek2date(weekdate = "2021-W53-1")
-
-compute_start_date_from_week_number <- function(week_number, year_number) {
-  ## this assumes a valid week number! There are 52 or 53 weeks in a year
-  isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
-  # computed_dates <- ISOweek::ISOweek2date(isoweek_string)
-  computed_dates <- rep(NA_Date_,times = length(week_number))
-  computed_dates[which(!is.na(week_number))] <- ISOweek::ISOweek2date(isoweek_string[which(!is.na(week_number))])  # return NA instead of failing with ISOweek!
-  ## invalid weeks will result in a computed date in a year after the requested year...
-  ## a valid week can result in a date that starts in the year before!
-  if (any(year_number<year(ymd(computed_dates)), na.rm = TRUE)) stop("invalid week number provided!")
-  return(computed_dates)
-}
-
-tibble(
-    week_number = c(1,2,3,52,53)
-  ) %>%
-  mutate(
-    date = compute_start_date_from_week_number(week_number = week_number, year_number = 2020),
-    iso_week_from_date = isoweek(date)
-  ) %>% knitr::kable(caption = "Examples of computing start date for each week, and computing isoweek from the dates!")
-```
-
-## Consistent Place of death coding
-
-Place of death was named differently between datasets (location/place) and the different places were also named slightly differently, so we find all of them and recode them to the same:
-
-```{r}
-order_of_place_of_death_levels <- c("Hospital", "Care home", "Home & other non-institution", "Other", "All")
-
-recode_place_of_death <- function(data_tbl) {
-  data_tbl %>%
-    mutate(place_of_death = case_when(
-      str_detect(place_of_death, pattern = regex(pattern = "all", ignore_case = TRUE)) ~ "All",
-      str_detect(place_of_death, pattern = regex(pattern = "care.*home", ignore_case = TRUE)) ~ "Care home",
-      str_detect(place_of_death, pattern = regex(pattern = "(non.*institut)|home", ignore_case = TRUE)) ~ "Home & other non-institution",
-      str_detect(place_of_death, pattern = regex(pattern = "hospital", ignore_case = TRUE)) ~ "Hospital",
-      str_detect(place_of_death, pattern = regex(pattern = "other", ignore_case = TRUE)) ~ "Other",
-      TRUE ~ NA_character_
-      ) %>%
-        factor(x = ., levels = order_of_place_of_death_levels)
-    )
-}
-```
-
-## Consistent Cause of death coding
-
-The `opendatascot`-imported data distinguishes "all causes" deaths from "covid-related" deaths - the non-covid-related deaths are the difference between those.
-In contrast, the other weekly datasets code cause of death as either covid-related or non-covid-related, in which case we need to compute all cause deaths as the sum of the two.
-
-We'll change the wording of these to a consistent format so it's easier to compute consistent columns.
-The below function will also compute non-covid deaths for data where all deaths and covid deaths are provided, and all deaths where covid- and non-covid-related deaths are provided.
-
-This involves pivotting the number of deaths variable into wide format based on the cause of death. Because of this, we need to deal with implicit missing values:
-
-The first COVID-19-related deaths were reported from week 12 onwards.
-In some of the datasets, a missing/omitted value represents 0 cases in that category, when a 0 should have been recorded.
-E.g.
-deaths in "Other institutions" are very low, and if a record is missing for that category, we should assume 0 deaths.
-
-On 3 Februrary 2021 as I was updating the script, the weekly figures started distinguishing cause of death: COVID-19 can be a contributory factor, or an underlying cause of death.
-For the purposes of this study, COVID-19 deaths are in the minority, and we will disregard this distinction for now.
-
-```{r}
-## helper function to make consistently coded cause of death
-recode_cause_of_death <- function(data_tbl) {
-  data_tbl %>%
-    mutate(
-      cause_of_death = case_when(
-        cause_of_death == "all-causes" ~ "deaths_all_causes",
-        cause_of_death == "all-causes-average-of-corresponding-week-over-previous-5-years" ~ "deaths_nrs_past_average_all_causes",
-        cause_of_death == "covid-19-related" ~ "deaths_covid_related",
-        cause_of_death == "COVID-19 mentioned" ~ "deaths_covid_related",
-        cause_of_death == "COVID-19 contributory factor" ~ "deaths_covid_related",
-        cause_of_death == "COVID-19 underlying cause" ~ "deaths_covid_related",
-        cause_of_death == "Non-COVID-19" ~ "deaths_non_covid",
-        TRUE ~ NA_character_
-      )
-    )
-}
-
-# TODO:  still too complicated to safely use as a function...
-## find which combinations of geographic area, cause of death, and place of death are missing and set the number of deaths to 0 for them
-# complete_weekly_deaths_with_implicit_zeros <- function(data_tbl) {
-#   data_tbl %>%
-#   tidyr::complete(
-#       nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
-#       ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
-#       fill = list(number_of_deaths = 0)
-#       )
-# }
-
-## calculate number of deaths after recoding cause of death -
-## recoding results in two entries for covid-related deaths, so we need to sum over them
-## this is achieved by grouping over all variables other than number of deaths
-recalculate_number_of_deaths_after_recoding_cause_of_death <- function(data_tbl) {
-  data_tbl %>%
-  group_by_all %>% 
-  ungroup(number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")  # recompute the sum of deaths
-}
-
-## covid death reporting started in week 12, 2020 - before then it doesn't make sense to distunguish between covid and non-covid deaths, and only the all cause death figure should be kept
-## by using the 'overrun' weeks we ensure that 2021 or later isn't affected
-## note: we're keeping all cause deaths!
-set_covid_and_non_covid_deaths_to_zero_before_reporting_started <- function(data_tbl) {
-  data_tbl %>%
-    mutate(
-      deaths_non_covid = if_else(
-        condition = week_number_run_over < 12,
-        true = NA_real_,
-        false = deaths_non_covid
-      ),
-      deaths_covid_related = if_else(
-        condition = week_number_run_over < 12,
-        true = NA_real_,
-        false = deaths_covid_related
-      )
-    )
-}
-
-# TODO: delete below function, it's too complicated and I've broken it down to the individual functions
-# recode_cause_of_death_and_compute_columns <- function(data_tbl) {
-#   potential_grouping_variables <- c("ref_area","sex","age","ref_period")
-#   grouping_variables <- potential_grouping_variables[which(potential_grouping_variables %in% names(data_tbl))]
-#   
-#   interim_data_tbl <- 
-#     data_tbl %>%
-#     recode_cause_of_death() %>%
-#     tidyr::complete(
-#       nesting(year, week_number, week_number_run_over, date_w_c),
-#       !!!rlang::syms(grouping_variables),
-#       cause_of_death, place_of_death,
-#       fill = list(number_of_deaths = 0)
-#       ) %>%
-#     pivot_wider(names_from = "cause_of_death", values_from = "number_of_deaths")
-#   
-#   ## compute non-covid deaths from all deaths minus covid deaths
-#   if (!"deaths_non_covid" %in% names(interim_data_tbl)) {
-#     interim_data_tbl <-
-#       interim_data_tbl %>%
-#       mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)
-#   }
-#   ## compute all deaths from non-covid deaths + covid deaths
-#   if (!"deaths_all_causes" %in% names(interim_data_tbl)) {
-#     interim_data_tbl <-
-#       interim_data_tbl %>%
-#       mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related)
-#   }
-#   
-#   interim_data_tbl <-
-#     interim_data_tbl %>%  # set as missing covid- & non-covid-related deaths before week 12
-#     mutate(
-#       deaths_non_covid = if_else(
-#         condition = week_number_run_over < 12,
-#         true = NA_real_,
-#         false = deaths_non_covid
-#       ),
-#       deaths_covid_related = if_else(
-#         condition = week_number_run_over < 12,
-#         true = NA_real_,
-#         false = deaths_covid_related
-#       )
-#     )
-#   
-#   return(interim_data_tbl)
-# }
-```
-
-## Consistent use of missing value
-
-```{r}
-replace_na_with_0_from_week_12 <- function(data_tbl) {
-  data_tbl %>%
-    mutate(
-      value = if_else(
-        condition = week_number_roll_over >=12 & is.na(number_of_deaths),
-        true = 0,
-        false = number_of_deaths
-      )
-    )
-}
-```
-
-## Merging health boards
-
-Smaller health boards were merged... TODO!
-
-> "Note: Orkney, Shetland, & Western health boards were merged into 'Island HBs'; Borders & Dumfries and Galloway health boards were merged into 'B,D&G HB'"
-
-## Load weekly COVID-19 related deaths
-
-~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above!
-Even for the Scotland-wide data, place of death is collapsed to 'All' for the granular sex & age categories, actually.
-
-I've also attempted loading the case counts using `ckanr` using the CKAN API from the opendata.nhs.scot website, but gave that up in favour of using their `.csv` links instead to download daily-updated data.
-
-Note about weekly LA/HB data:
-
-> From 14/12/2020 confirmed positive cases are now allocated to an area based on the postcode recorded in the testing system.
-Previously cases were allocated to local areas mainly based on their usual residential address recorded in the Community Health Index (CHI) database.
-This change will impact on newly reported cases on this day.
-
-> Please note that the updated methodology for determining COVID-19 related ICU admissions has been applied to the historic data.
-The updated definition is as follows: A patient who has tested positive for COVID at any time in the 21 days prior to admission to ICU, or who have tested positive from the date of admission up to and including the date of ICU discharge.
-
-> In addition, on 30th October it came to our attention that there is an ongoing issue when linking ICU data to the lab data for COVID-19 test results.
-Any COVID-19 positive patients with a missing CHI that had a first positive test in the community are unable to be linked to the ICU data.
-As a result the COVID-19 positive numbers could be underreported by a small number.
-
-`crude_rate_positive`: Crude rate of total positive cases (cumulative) per 100,000 population.
-
-```{r}
-download_dir <- "./downloaded_data"
-if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist
-
-files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
-
-# checking out structure
-
-# ods_all_datasets() %>% View()
-# ods_structure("deaths")  # can we import all deaths-related data this way?
-# ods_structure("coronavirus-covid-19-management-information")
-
-# ods_dataset("coronavirus-covid-19-management-information", geography = "hb") %>%
-#   filter(variable == "testing-new-cases-reported")
-# 
-# ods_dataset("coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported")
-# ods_dataset("coronavirus-covid-19-management-information", geography = "la", variable = "testing-new-cases-reported")
-# ods_dataset("coronavirus-covid-19-management-information", variable = "testing-new-cases-reported")
-# 
-# ods_print_query(dataset = "coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported") %>% dput
-# ods_print_query(dataset = "coronavirus-covid-19-management-information", variable = "testing-new-cases-reported") %>% dput
-#   
-# modified_query <- "PREFIX qb: <http://purl.org/linked-data/cube#>                 PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>                 PREFIX xsd: <http://www.w3.org/2001/XMLSchema#> select   ?refArea ?refPeriod ?measureType ?variable ?value where { ?data qb:dataSet <http://statistics.gov.scot/data/coronavirus-covid-19-management-information>.
-?data <http://purl.org/linked-data/sdmx/2009/dimension#refArea> ?refArea.
-?data <http://purl.org/linked-data/sdmx/2009/dimension#refPeriod> ?refPeriod.
-?data <http://purl.org/linked-data/cube#measureType> ?measureType.
-?data <http://statistics.gov.scot/def/dimension/variable> ?variable.
-?data ?measureType ?value.
-filter (regex(str(?variable ), 'testing-new-cases-reported'))  } order by ?refPeriod ?refArea"
-# 
-# modified_query %>%
-#   opendatascot:::ods_query_database(endpoint, query = .) %>%
-#   opendatascot:::pre_process_data()
-# 
-# query <- ""
-
-# 
-# library(ckanr)
-# ckanr_setup(url = "https://www.opendata.nhs.scot/")
-# # package_list(as = "table")
-# # tag_list(as = "table")
-# package_search(q = 'daily', rows = 2, as = 'table') %>% str
-# 
-# 
-# download_data_from_opendata_via_ckan_api <- function(resource_id) {
-#   ckan <- ckanr::src_ckan("https://www.opendata.nhs.scot/")
-#   dplyr::tbl(src = ckan$con, from = resource_id_daily_case_trends_la, ) %>% 
-#   as_tibble()
-# }
-# 
-# resource_id_daily_case_trends_la <- "427f9a25-db22-4014-a3bc-893b68243055"
-# 
-# resource_show(id = resource_id_daily_case_trends_la, as = "table")
-
-# ods_print_query("deaths")
-
-# ods_structure("deaths-involving-coronavirus-covid-19")
-# ods_schemes("deaths-involving-coronavirus-covid-19")
-# ods_print_query("deaths-involving-coronavirus-covid-19")
-# ods_print_query("deaths-involving-coronavirus-covid-19", geography = "la")
-# ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003
-# phsmethods::match_area("S92000003")  # "Scotland"
-
-## helper function that links the 9-character geography code to its name
-make_human_readable_ref_area <- function(data_tbl, ref_area = refArea) {
-  enquoted_ref_area <- enquo(ref_area)
-  data_tbl %>%
-    mutate(
-      !!enquoted_ref_area := phsmethods::match_area(!!enquoted_ref_area)
-    )
-}
-
-# raw_covid_deaths <- list(
-#   scotland = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc"),
-#   health_board = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "hb"),
-#   local_authority = ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la")
-# ) %>%
-#   map(.x = ., .f = ~make_human_readable_ref_area(data_tbl = .x))
-
-weekly_deaths_overall <-
-  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
-  make_human_readable_ref_area %>%
-  clean_names() %>%
-  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
-  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
-  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
-  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
-  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
-  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
-  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
-  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
-  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
-  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021, we only want the weekly figures
-  mutate(
-    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
-    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
-  ) %>%
-  select(-c(measure_type)) %>%  #  measure_type is count for all data
-  recode_place_of_death() %>%
-  recode_cause_of_death() %>%
-  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # not needed here probably, but no harm in including it
-  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
-  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing
-
-# weekly_deaths_sex_age_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-
-weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
-  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
-  mutate(
-    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
-    week_number = parse_integer(str_sub(week_number, 4, 5)),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
-  ) %>%
-  recode_cause_of_death() %>%
-  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
-  tidyr::complete(  # complete implicit missing entries and make them 0
-    nesting(week_number,week_number_run_over,date_w_c,year),
-    age,sex,place_of_death,cause_of_death,
-    fill = list(number_of_deaths = 0)
-  ) %>%
-  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
-  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
-  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
-  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
-
-# 
-# weekly_deaths_la_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-
-weekly_deaths_la <- 
-  read_excel(files_weekly_deaths$la, sheet = 2, skip = 2) %>% 
-  clean_names %>%
-  rename(week_number = week_of_occurrence, ref_area = council, place_of_death = location_of_death, number_of_deaths = deaths) %>%
-  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
-  mutate(
-    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
-    week_number = parse_integer(str_sub(week_number, 4, 5)),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
-  ) %>%
-  recode_cause_of_death() %>%
-  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
-  tidyr::complete(
-      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
-      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
-      fill = list(number_of_deaths = 0)
-      ) %>%
-  # complete_weekly_deaths_with_implicit_zeros() %>%
-  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
-  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
-  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
-  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
-
-# weekly_deaths_hb_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-
-weekly_deaths_hb <- 
-  read_excel(files_weekly_deaths$hb, sheet = 2, skip = 2) %>% 
-  clean_names %>%
-  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
-  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
-  mutate(
-    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
-    week_number = parse_integer(str_sub(week_number, 4, 5)),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
-  ) %>%
-  recode_cause_of_death() %>%
-  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
-  tidyr::complete(
-      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
-      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
-      fill = list(number_of_deaths = 0)
-      ) %>%
-  # complete_weekly_deaths_with_implicit_zeros() %>%
-  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
-  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # make separate columns for causes of death figures
-  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
-  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
-
-## compute weekly data for sex and age separately
-## sum over sex or age, respectively
-weekly_deaths_sex <-
-  weekly_deaths_sex_age %>%
-  filter(age!="all") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
-  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
-
-weekly_deaths_age <-
-  weekly_deaths_sex_age %>%
-  filter(sex!="all") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
-  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
-```
-
-### Checking week number coding
-
-```{r}
-weekly_deaths_sex_age %>% 
-  count(year, date_w_c, week_number, week_number_run_over) %>% 
-  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
-  select(-n) %>% 
-  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
-```
-
-## Load historical deaths data
-
-The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data.
-
-```{r, warning=FALSE}
-files_historical_deaths <- set_names(x = as.list(table_of_historical_data_sources$path_latest), nm = table_of_historical_data_sources$short_name)
-
-## overall file - where did I get the link from? it's not obviously available from the website...
-# past_deaths_overall_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-# 
-# past_deaths_hb_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-# 
-# past_deaths_la_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-# 
-# past_deaths_sex_age_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-
-past_deaths_overall <- 
-  read_csv(files_historical_deaths$overall, skip = 2) %>%
-  rename(year = 1) %>%  # rename first column
-  mutate(
-    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
-  ) %>%
-  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
-  drop_na(`1`) %>%
-  pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
-  mutate(week_number=as.integer(week_number)) %>%
-  recode_place_of_death()
-
-past_deaths_hb <- 
-  readxl::read_excel(files_historical_deaths$hb, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
-  clean_names() %>%
-  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
-  rename(  # same format as current data
-    week_number = matches("week"),
-    place_of_death = matches("location"),
-    ref_area = health_board
-  ) %>%
-  mutate(
-    week_number = as.integer(week_number)
-  ) %>%
-  tidyr::complete(
-    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
-    fill = list(number_of_deaths = 0)
-  ) %>% 
-  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
-
-past_deaths_la <-
-  readxl::read_excel(files_historical_deaths$la, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
-  clean_names() %>%
-  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
-  rename(  # same format as current data
-    week_number = matches("week"),
-    place_of_death = matches("location"),
-    ref_area = council_area
-  ) %>%
-  mutate(
-    week_number = as.integer(week_number)
-  ) %>%
-  tidyr::complete(
-    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
-    fill = list(number_of_deaths = 0)
-  ) %>% 
-  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
-
-past_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(files_historical_deaths$sex_age, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
-  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
-  mutate(
-    year = 2000L + as.integer(year),
-    week_number = as.integer(week_number)
-  ) %>%
-  tidyr::complete(
-    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
-    fill = list(number_of_deaths = 0)
-  ) %>% 
-  recode_place_of_death()
-
-## compute weekly data for sex and age separately
-## sum over sex or age, respectively
-past_deaths_sex <-
-  past_deaths_sex_age %>%
-  group_by_all %>%
-  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
-
-past_deaths_age <-
-  past_deaths_sex_age %>%
-  group_by_all %>%
-  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
-```
-
-### Do we get the same number of deaths between datasets?
-
-```{r}
-## check that the results are the same as overall deaths
-map2_dfr(
-  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
-  .y = c("overall","health_board","council_area","sex_age"),
-  .f = function(data_tbl, name) {
-    data_tbl %>%
-      mutate(dataset=name,year=as.numeric(year)) %>%
-      filter(!is.na(year) & place_of_death!="All") %>%  # to remove summary values from the data
-      group_by(dataset,year) %>%
-        summarise(total = sum(number_of_deaths), .groups = "drop")
-  }
-) %>% 
-  pivot_wider(names_from=year,values_from=total) %>%
-  knitr::kable(caption = "Number of historical deaths by year as calculated from different datasets.
-The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other.
-The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other.
-This may be due to the following (verbatim from data description online): Health Board and Local Authority figures include non-residents.
-Deaths are allocated to areas based on the usual residence of the deceased.
-If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
-```
-
-## Load map data with `sf`
-
-I previously downloaded shapefiles for:
-
-* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
-* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
-  - Note: this is a fairly high resolution, don't need this much detail!
-  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)
-
-```{r}
-# shapefile_hb <- st_read(dsn = "./map_data/hb")  # too detailed!
-shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
-shapefile_la <- st_read(dsn = "./map_data/la")
-
-# merged_deaths_la %>%
-#   filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
-#   left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
-#   ggplot(aes(fill = deaths_all_causes, geometry = geometry, group=ref_area)) +
-#   geom_sf()
-# 
-# merged_deaths_hb %>%
-#   filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
-#   left_join(shapefile_hb_uk_wide,by=c("ref_area"="name")) %>%
-#   ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
-#   geom_sf() +
-#   theme(aspect.ratio = 1.3)
-
-# ggplot(data = shapefile_hb_uk_wide, aes(geometry=geometry)) + geom_sf()
-
-## simple UK map
-# ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
-#   geom_polygon() +
-#   coord_map()
-```
-
-## Load COVID case counts by LA/HB
-
-```{r}
-files_case_counts <- set_names(x = as.list(table_of_case_trends_data_sources$path_latest), nm = table_of_case_trends_data_sources$short_name)
-
-cases_hb <- read_csv(
-  files_case_counts$hb, 
-  col_types = 
-    cols(
-    .default = col_double(),
-    `_id` = col_skip(),
-    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
-    Date = col_date(format = "%Y%m%d"),
-    HB = col_character(),
-    HBName = col_character(),
-    HospitalAdmissionsQF = col_character(),
-    ICUAdmissionsQF = col_character()
-    ),
-  ) %>% clean_names() %>%
-  rename(ref_area = hb_name) %>%
-  mutate(
-    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
-    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
-    ref_area = phsmethods::match_area(hb)  # convert to standard HB names
-    )
-
-cases_la <- read_csv(
-  file = files_case_counts$la, 
-  col_types = 
-    cols(
-    .default = col_double(),
-    `_id` = col_skip(),
-    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
-    Date = col_date(format = "%Y%m%d"),
-    CA = col_character(),
-    CAName = col_character()
-  )
-) %>% clean_names() %>%
-  rename(ref_area = ca_name) %>%
-  mutate(
-    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
-    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
-    ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
-    )
-```
-
-## Load quarterly bed occupancy data
-
-```{r}
-# TODO: download automatically in other script
-
-beds_raw <- 
-  read_csv("C:/Users/janwork/Downloads/beds_by_nhs_board_of_treatment_and_specialty.csv", guess_max = 3e4) %>%
-  mutate(ref_area = match_area(HB)) %>%  # clean health board names
-  clean_names() %>%
-  # select(-percentage_occupancy_qf) %>%
-  extract(col = quarter, into = c("year", "quarter"), regex = "(\\d{4})Q(\\d)", remove = TRUE) %>%
-  # mutate(across(.cols = matches("percent"), .fns = ~{.x / 100}))
-  mutate(percentage_occupancy = percentage_occupancy / 100)
-
-beds_raw %>%
-  filter(ref_area == "Scotland") %>% count(specialty_name)
-
-beds_raw %>%
-  filter(ref_area == "Scotland" & specialty_name == "Palliative Medicine") %>%
-  ggplot() +
-  aes(x = quarter, y = percentage_occupancy, colour = year, size = year) +
-  scale_y_continuous(labels = function(x) scales::percent(x), limits = c(0,NA)) +
-  # geom_point() +
-  geom_line(aes(group = year)) +
-  scale_colour_viridis_d() +
-  scale_size_manual(values = c(rep(0.5, times = 5),1)) +
-  NULL
-
-```
-
-# Calculate case counts: weekly counts to match weekly death reporting
-
-Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks.
-This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
-
-However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!
-
-Note also that the `crude_rate_positive_per100k` represents the cumulative rate!
-
-```{r}
-weekly_covid_cases_hb <-
-  cases_hb %>%
-  group_by(year, week_number, ref_area) %>%
-  summarise(
-    weekly_positive = sum(daily_positive),
-    cumulative_positive = max(cumulative_positive),
-    crude_rate_positive_per100k = max(crude_rate_positive),
-    .groups = "drop"
-  ) %>%
-  mutate(
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
-    )
-
-## extract Scotland-wide data and remove from HB split
-weekly_covid_cases_overall <- weekly_covid_cases_hb %>% filter(ref_area=="Scotland")
-weekly_covid_cases_hb <- weekly_covid_cases_hb %>% filter(ref_area!="Scotland")
-
-weekly_covid_cases_la <-
-  cases_la %>%
-  group_by(year, week_number, ref_area) %>%
-  summarise(
-    weekly_positive = sum(daily_positive),
-    cumulative_positive = max(cumulative_positive),
-    crude_rate_positive_per100k = max(crude_rate_positive),
-    .groups = "drop"
-  ) %>%
-  mutate(
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
-    )
-```
-
-# Comparing geographic areas by case counts
-
-We can use the yearly cumulative case count to group geographic areas by how many cases they've had by the end of 2020.
-
-```{r}
-cumulative_case_2020_based_grouping_la <-
-  weekly_covid_cases_la %>% 
-  filter(ref_area!="Scotland") %>%
-  filter(year == 2020) %>% 
-  filter(week_number == max(week_number)) %>%  # keep last week
-  select(ref_area, cumulative_positive, crude_rate_positive_per100k)
-
-cumulative_case_2020_based_grouping_hb <-
-  weekly_covid_cases_hb %>% filter(year == 2020) %>% 
-  filter(ref_area!="Scotland") %>%
-  filter(year == 2020) %>% 
-  filter(week_number == max(week_number)) %>%  # keep last week 
-  select(ref_area, cumulative_positive, crude_rate_positive_per100k)
-
-k_means_clustering_cases_and_per100k_2020_hb <- map(
-  .x = 2:4,
-  .f = ~kmeans(x=cumulative_case_2020_based_grouping_hb %>% select(cumulative_positive, crude_rate_positive_per100k), centers = .x, nstart = 1e3)
-) %>% set_names(nm=paste0("k",2:4))
-
-clusters_hb <-
-  bind_cols(
-    cumulative_case_2020_based_grouping_hb %>% select(ref_area),
-    map_dfc(.x = k_means_clustering_cases_and_per100k_2020_hb, .f = function(x) return(x$cluster))
-  )
-
-cumulative_case_2020_based_grouping_hb <-
-  cumulative_case_2020_based_grouping_hb %>%
-  left_join(clusters_hb, by="ref_area") %>%
-  mutate(
-    quartiles_cases = ntile(cumulative_positive, n=4),
-    quartiles_per100k = ntile(crude_rate_positive_per100k, n=4)
-  )
-
-# TODO: apply to LA also!
-
-hb_cases_plots <- list()
-
-hb_cases_plots$cumulative_cases_vs_per100k <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
-  geom_point() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_log10 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
-  geom_point() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases (log)",
-    y="COVID cases per 100,000 population (log)",
-    title="Cases vs. population rate (log scale axes)"
-  ) +
-  scale_x_log10() +
-  scale_y_log10() +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_cases))) +
-  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_per100k))) +
-  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate",
-    colour = "Quartile (Cases per 100,000)"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_k2 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k2))) +
-  geom_point(size=3, aes(colour=factor(k2))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_k3 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k3))) +
-  geom_point(size=3, aes(colour=factor(k3))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots$cumulative_cases_vs_per100k_k4 <-
-  cumulative_case_2020_based_grouping_hb %>%
-  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k4))) +
-  geom_point(size=3, aes(colour=factor(k4))) +
-  scale_colour_viridis_d() +
-  geom_label_repel() +
-  # theme_minimal() +
-  labs(
-    x="COVID cases",
-    y="COVID cases per 100,000 population",
-    title="Cases vs. population rate"
-  ) +
-  NULL
-
-hb_cases_plots_assembled <- patchwork::wrap_plots(hb_cases_plots, ncol = 2)
-
-hb_cases_plots_assembled
-```
-
-# Calculate historical averages, minima & maxima
-
-Due to the counting of weeks, most years have 52 weeks with some years (2015 & 2020) having 53.
-Therefore we need to decide how to deal with week 53 in 2020 since only 2015 has a corresponding weekly count.
-The solution used here is to calculate a weekly rate for week 53 from week 53 in 2015, and week 52 in 2016-2019 - this represents the "last week of the year", which includes December 31st.
-
-```{r, warning=FALSE}
-map2_dfr(
-  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_la, past_deaths_sex_age),
-  .y = c("overall","health_board","council_area","sex_age"),
-  .f = function(data_tbl, name) {
-    data_tbl %>%
-      mutate(dataset=name,year=as.numeric(year)) %>%
-      filter(year %in% 2016:2019 & week_number==53) %>%
-      count(dataset,year)
-  }
-) %>% 
-  pivot_wider(names_from=year,values_from=n) %>%
-  knitr::kable(caption = "Number of entries by dataset for week number 53 in years 2016-2019 (which only had 52 weeks).
-Only the 'overall' historical deaths dataset has week 53 entries for years with 52 weeks, and all the entries are for 0 deaths.")
-
-## helper function that creates week 53 data by taking week 52 in years with no year 53 and relabels it as 53
-## CAUTION: only use this for computing averages! otherwise it's making up data that wasnae there
-create_week_53_data_for_years_with_52_weeks <- function(data_tbl) {
-  real_data <-
-    data_tbl %>%
-    filter(!(year %in% 2016:2019 & week_number==53))  # remove any entries for week 53 in years with 52 weeks - in some datasets they existed but had 0 listed
-  fake_data <-
-    data_tbl %>%
-    filter(year %in% 2016:2019 & week_number==52) %>%
-    mutate(week_number=53)
-  return(
-    bind_rows(
-      real_data,
-      fake_data
-    )
-  )
-}
-
-historical_weekly_averageoverall <-
-  past_deaths_overall %>%
-  filter(year!="average") %>%  # remove averages
-  group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
-  filter(!all(number_of_deaths==0 & week_number==53)) %>%
-  ungroup %>%
-  create_week_53_data_for_years_with_52_weeks %>%
-  ungroup %>%
-  group_by(week_number,place_of_death) %>% 
-  summarise(
-    across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
-    .groups = "drop"
-  ) %>% rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths"))
-
-week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
-
-# TODO: weekly averages may be too detailed for this level od subcategorising!
-
-historical_weekly_averagehb <-
-  past_deaths_hb %>%
-  create_week_53_data_for_years_with_52_weeks() %>%
-  group_by(
-    week_number, place_of_death, ref_area
-  ) %>%
-  summarise(
-    deaths_mean = mean(number_of_deaths),
-    deaths_min = min(number_of_deaths),
-    deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-
-historical_weekly_averagela <-
-  past_deaths_la %>%
-  create_week_53_data_for_years_with_52_weeks() %>%
-  group_by(
-    week_number, place_of_death, ref_area
-  ) %>%
-  summarise(
-    deaths_mean = mean(number_of_deaths),
-    deaths_min = min(number_of_deaths),
-    deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-
-historical_weekly_averagesex_age <-
-  past_deaths_sex_age %>%
-  create_week_53_data_for_years_with_52_weeks() %>%
-  group_by(
-    week_number, place_of_death, age, sex
-  ) %>%
-  summarise(
-    deaths_mean = mean(number_of_deaths),
-    deaths_min = min(number_of_deaths),
-    deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-
-historical_weekly_averageage <-
-  past_deaths_age %>%
-  create_week_53_data_for_years_with_52_weeks() %>%
-  group_by(
-    week_number, place_of_death, age
-  ) %>%
-  summarise(
-    deaths_mean = mean(number_of_deaths),
-    deaths_min = min(number_of_deaths),
-    deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-
-historical_weekly_averagesex <-
-  past_deaths_sex %>%
-  create_week_53_data_for_years_with_52_weeks() %>%
-  group_by(
-    week_number, place_of_death, sex
-  ) %>%
-  summarise(
-    deaths_mean = mean(number_of_deaths),
-    deaths_min = min(number_of_deaths),
-    deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-
-compute_historical_annual_average_deaths <- function(data_tbl, ...) {
-  grouping_vars <- enquos(...)
-  data_tbl %>%
-    group_by(year, place_of_death, !!!grouping_vars) %>%
-    summarise(  # calculate annual totals first...
-    number_of_deaths = sum(number_of_deaths), 
-    .groups = "keep"
-  ) %>%
-  ungroup(year) %>%  # remove year as grouping to average over year now
-  summarise(
-    annual_deaths_mean = mean(number_of_deaths),
-    annual_deaths_min = min(number_of_deaths),
-    annual_deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-}
-
-historical_annual_average_overall <-
-  past_deaths_overall %>%
-  compute_historical_annual_average_deaths()
-
-historical_annual_average_hb <-
-  past_deaths_hb %>%
-  compute_historical_annual_average_deaths(ref_area)
-
-historical_annual_average_la <-
-  past_deaths_la %>%
-  compute_historical_annual_average_deaths(ref_area)
-
-historical_annual_average_sex_age <-
-  past_deaths_sex_age %>%
-  compute_historical_annual_average_deaths(sex, age)
-
-historical_annual_average_sex <-
-  past_deaths_sex %>%
-  compute_historical_annual_average_deaths(sex)
-
-historical_annual_average_age <-
-  past_deaths_age %>%
-  compute_historical_annual_average_deaths(age)
-```
-
-## Check weekly historic averages against NRS computed data
-
-```{r}
-weekly_deaths_overall %>%
-  filter(sex=="all" & age=="all") %>%
-  left_join(historical_weekly_averageoverall) %>%
-  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean) %>%
-  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
-  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e.
-all other entries had the same figures!).
-Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019.
-For all the other datasets the mean deaths computed agree!")
-```
-
-# Calculate moving averages (rolling mean) over multiple weeks
-
-Once data are broken down by region there are very small figures on a weekly basis, so we can compute a moving average over several weeks instead.
-This is different to merging figures into 2-week chunks, for example, as information from consecutive weeks will overlap.
-
-The average will be computed to the "left" of the current value (a week's moving average is the average of values at this week + n-1 past weeks, or "right" aligned).
-
-Moving averages for historical data will be computed over the previously computed weekly means & ranges.
-Computing moving averages over historical ranges produces wider spread data than computing ranges of moving averages (see `order_of_min_max_rolling_mean.R`).
-
-```{r}
-weeks_to_average <- 2
-
-## compute moving average over weekly data, using average over k weeks
-## using historical data to compute the moving average in the first few weeks of current data
-## Note: historical data needs to be the non-summarised (raw) data!
-## to compute moving average in first few week(s) of year, we add in the previous years' data and code the weeks so that they retain the right order
-## i.e. in 2020 the week_number_run_over starts at 1, and the last week of 2019 is coded as 0, the 2nd last as -1, etc.
-## once the moving average has been computed, we remove the past data;
-## this approach works for 2021 data seamlessly, using the last weeks of 2020 for the moving average in 2021!
-compute_moving_average_for_current_data <- function(current_data, historical_data) {
-  bind_rows(
-    historical_data %>%  # add historical data and format it so that it matches current data
-      filter(year == 2019) %>%  # keep only last year before 2020
-      rename(deaths_all_causes = number_of_deaths) %>%  # in historical deaths there is no split between covid/non-covid so we rename to all cause deaths
-      mutate(week_number_run_over = -(52-week_number)),
-    current_data
-  ) %>%
-    arrange(week_number_run_over) %>%  # just in case the order got messed up, re-order by week_number_run_over for the computation
-    group_by_all() %>% ungroup(matches("^deaths_|week|date|year")) %>%  # clever way of getting grouping by ref_area, sex, age, whatever is available
-    mutate(across(
-        .cols = matches("^deaths_"), 
-        .fns = ~zoo::rollmean(.x, k=weeks_to_average, fill=NA, align="right"), 
-        .names = "ma{weeks_to_average}w_{.col}"
-        )
-    ) %>%
-    ungroup %>%
-    filter(year > 2019)  # keep only current data
-}
-
-rolling_deaths_hb <-
-  compute_moving_average_for_current_data(current_data = weekly_deaths_hb, historical_data = past_deaths_hb)
-
-rolling_deaths_la <-
-  compute_moving_average_for_current_data(current_data = weekly_deaths_la, historical_data = past_deaths_la)
-
-rolling_deaths_sex_age <-
-  compute_moving_average_for_current_data(current_data = weekly_deaths_sex_age, historical_data = past_deaths_sex_age)
-
-rolling_deaths_age <-
-  compute_moving_average_for_current_data(current_data = weekly_deaths_age, historical_data = past_deaths_age)
-
-rolling_deaths_sex <-
-  compute_moving_average_for_current_data(current_data = weekly_deaths_sex, historical_data = past_deaths_sex)
-
-## inspect the calculations  
-rolling_deaths_la %>% filter(week_number_run_over %in% c(1:3,52:54) & str_detect(ref_area,"Glasgow") & place_of_death=="Hospital") %>% select(year, week_number_run_over, ref_area, place_of_death, deaths_all_causes, ma2w_deaths_all_causes) %>% knitr::kable(caption="Showing how rolling average was computed: week 1 is average of week 52 of 2019 (not shown) and week 1 of 2020; week 2 is average of weeks 1 & 2, etc.")
-
-rolling_deaths_sex_age %>% filter(week_number_run_over %in% c(1:3,52:54) & str_detect(sex,"F") & age == "65-74" & place_of_death=="Hospital") %>% select(year, week_number_run_over, sex, age, place_of_death, deaths_all_causes, ma2w_deaths_all_causes) %>% knitr::kable(caption = "Showing how rolling average was computed: week 1 is average of week 52 of 2019 (not shown) and week 1 of 2020; week 2 is average of weeks 1 & 2, etc.")
-
-## helper function that takes historical data and computes a moving average over k weeks
-## with the additional trick of using the last week of all the years as the week prior to the first week
-## strictly speaking we should be using the last week(s) of 2014 to compute the moving average for the first week(s) in 2015, depending on k; and we shouldn't be using the last week of 2019 in the computation either
-## however, the overall effect of this is small since we're averaging over the entire 2015-2019 period
-## the inaccuracy of this approach increases with increasing k
-## the solution for fixing this is straightforward - just download 2014 data as well, and merge the "current" data and the historical data, and then perform the moving average calculation as usual
-compute_moving_average_for_historical_data <- function(historical_data) {
-  last_week_in_previous_year_required <- 53 + 2 - weeks_to_average  # this computes the last week needed from the previous year: if averaging over 2 weeks, we need week 3, if over 3 weeks, the lat week needed is 52, etc.
-  bind_rows(
-    historical_data %>%  # keep last week from all years and convert it to week 0
-      filter(week_number %in% last_week_in_previous_year_required:53) %>% 
-      mutate(week_number = week_number - 53),  # set week number for data in previous to negative numbers but retaining order 
-    historical_data
-  ) %>%
-  arrange(week_number) %>%  # sort in ascending order of week_number
-  group_by_all() %>% ungroup(matches("^deaths_"),week_number) %>%
-  mutate(across(
-      .cols = matches("^deaths_"), 
-      .fns = ~zoo::rollmean(.x, k=weeks_to_average, fill=NA, align="right"), 
-      .names = "ma{weeks_to_average}w_{.col}"
-      )
-  ) %>%
-  ungroup %>%
-  filter(week_number!=0)
-}
-
-historical_weekly_averagehb <- compute_moving_average_for_historical_data(historical_weekly_averagehb)
-historical_weekly_averagela <- compute_moving_average_for_historical_data(historical_weekly_averagela)
-historical_weekly_averagesex_age <- compute_moving_average_for_historical_data(historical_weekly_averagesex_age)
-historical_weekly_averagesex <- compute_moving_average_for_historical_data(historical_weekly_averagesex)
-historical_weekly_averageage <- compute_moving_average_for_historical_data(historical_weekly_averageage)
-```
-
-# Merge weekly COVID/non-COVID-related deaths in 2020/2021 with historic death rates
-
-```{r}
-## compute rolling average
-merged_deaths_overall <-
-  weekly_deaths_overall %>%
-  filter(sex=="all" & age=="all") %>%
-  inner_join(historical_weekly_averageoverall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
-            by = c("place_of_death", "week_number", "sex", "age"))
-
-merged_deaths_hb <-
-  rolling_deaths_hb %>%  # use data with rolling movingincluded
-  inner_join(historical_weekly_averagehb, by = c("week_number","ref_area","place_of_death"))
-
-merged_deaths_la <-
-  rolling_deaths_la %>%  # use data with rolling movingincluded
-  inner_join(historical_weekly_averagela, by = c("week_number","ref_area","place_of_death"))
-
-merged_deaths_sex_age <-
-  rolling_deaths_sex_age %>%  # use data with rolling movingincluded
-  inner_join(historical_weekly_averagesex_age, by = c("place_of_death", "week_number", "sex", "age"))
-
-merged_deaths_sex <-
-  rolling_deaths_sex %>%
-  inner_join(historical_weekly_averagesex, by = c("place_of_death", "week_number", "sex"))
-
-merged_deaths_age <-
-  rolling_deaths_age %>%
-  inner_join(historical_weekly_averageage, by = c("place_of_death", "week_number", "age"))
-```
-
-## Weekly deaths including past weekly deaths
-
-For this dataset, we simply `bind_rows` weekly deaths by year and place of death with past weekly deaths by year and location.
-Note that past data for years with 52 weeks includes an entry of 0 deaths for week 53, so we'll remove them, otherwise the date computation is thrown off!
-
-```{r}
-weekly_deaths_current_and_past <-
-  bind_rows(
-    weekly_deaths_overall %>%
-    filter(sex=="all" & age=="all" & place_of_death!="All") %>%  # keep only aggregated values over sex & age
-    select(year, week_number, week_number_run_over, place_of_death, matches("deaths"), -deaths_nrs_past_average_all_causes)
-    ,
-    past_deaths_overall %>%
-      rename(deaths_all_causes = number_of_deaths) %>%
-      filter(place_of_death!="All" & year!="average") %>%  # remove aggregated entries
-      mutate(year = as.integer(year)) %>%
-      filter(!(year %in% 2016:2019 & week_number==53))  # remove week 53 entries from years 2016-2019
-  ) %>%
-  mutate(date_w_c = compute_start_date_from_week_number(week_number = week_number, year_number = year))
-```
-
-# Annual deaths for 2020 & 2021
-
-For visualising data using maps, we need to compute data for the entire year.
-Note: the non-covid and covid-related deaths in 2020 don't add up to all cause deaths due to the covid status only starting in week 12!
-
-```{r}
-compute_total_deaths_in_year <- function(data_tbl, ...) {
-  grouping_vars <- enquos(...)
-  data_tbl %>%
-    group_by(year, place_of_death, !!!grouping_vars) %>%
-    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%  # compute sums of all types of death
-    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)  # compute a smaller number of all-cause deaths since covid started (over 2020, covid- and non-covid-related deaths don't add up to all-cause deaths because pre-covid deaths are included also!)
-}
-
-compute_ratio_of_annual_deaths_to_historical <- function(data_tbl) {
-  data_tbl %>%
-    mutate(
-      ratio_annual_deaths_to_historical_mean = deaths_all_causes / annual_deaths_mean,
-      ratio_annual_deaths_to_historical_min = deaths_all_causes / annual_deaths_min,
-      ratio_annual_deaths_to_historical_max = deaths_all_causes / annual_deaths_max
-    )
-}
-
-total_deaths_la <-
-  merged_deaths_la %>%
-  compute_total_deaths_in_year(ref_area) %>%
-  left_join(historical_annual_average_la, by = c("place_of_death", "ref_area")) %>%
-  compute_ratio_of_annual_deaths_to_historical
-
-total_deaths_hb <-
-  merged_deaths_hb %>%
-  compute_total_deaths_in_year(ref_area) %>%
-  left_join(historical_annual_average_hb, by = c("place_of_death", "ref_area")) %>%
-  compute_ratio_of_annual_deaths_to_historical
-
-total_deaths_sex <-
-  merged_deaths_sex %>%
-  compute_total_deaths_in_year(sex) %>%
-  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
-  compute_ratio_of_annual_deaths_to_historical
-
-total_deaths_age <-
-  merged_deaths_age %>%
-  compute_total_deaths_in_year(age) %>%
-  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
-  compute_ratio_of_annual_deaths_to_historical
-
-total_deaths_overall <-
-  merged_deaths_overall %>%
-  compute_total_deaths_in_year() %>%
-  left_join(historical_annual_average_overall, by = c("place_of_death")) %>%
-  compute_ratio_of_annual_deaths_to_historical
-```
-
-# Constants about latest data available
-
-```{r}
-## constants for use in graphs & annotations
-weeks_available_2021 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2021])
-most_recent_date_available_2021 <- max(merged_deaths_age$date_w_c)
-earliest_date <- min(merged_deaths_age$date_w_c)
-```
-
-# Comparisons of annual rates with current rates
-
-```{r}
-table_current_deaths_compared_to_past <-
-  bind_rows(
-    ## all available data
-    merged_deaths_overall %>%
-      filter(age == "all" &
-               sex == "all"
-             ) %>% 
-      group_by(year, place_of_death, week_number) %>%
-      summarise(
-        deaths_all_causes = sum(deaths_all_causes),
-        # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-        .groups = "drop"
-      ) %>% 
-      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
-      left_join(
-        past_deaths_overall %>% filter(year != "average") %>% group_by(week_number, place_of_death) %>% summarise(
-                                           average_weekly_deaths = sum(number_of_deaths) / 5,
-                                           .groups = "drop"
-                                         ),
-        by = c("week_number", "place_of_death")
-      ) %>% 
-      group_by(place_of_death) %>%
-      summarise(
-        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
-        deaths_all_causes = sum(deaths_all_causes),
-        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
-        .groups = "drop"
-      )
-    ,
-    ## 2020 only
-    merged_deaths_overall %>%
-      filter(age == "all" &
-               sex == "all" &
-               year == 2020
-             ) %>% 
-      group_by(place_of_death) %>%
-      summarise(
-        period = "2020",
-        deaths_all_causes = sum(deaths_all_causes),
-        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-        .groups = "drop"
-      ) %>% 
-      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
-    left_join(
-      historical_annual_average_overall %>% 
-        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
-      by = "place_of_death"
-    )
-    ,
-    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
-    merged_deaths_overall %>%
-      filter(
-        age == "all" &
-          sex == "all" &
-          year == 2020 & week_number >= 12
-      ) %>%
-      group_by(place_of_death) %>%
-      summarise(
-        period = "2020, week 12 - week 53",
-        deaths_all_causes = sum(deaths_all_causes),
-        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-        .groups = "drop"
-      ) %>% 
-    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
-    left_join(
-      past_deaths_overall %>% 
-        filter(week_number >= 12 & year != "average") %>%
-        group_by(place_of_death) %>%
-        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
-      by = "place_of_death"
-    )
-  ) %>%
-  mutate(
-    ratio_deaths_all_causes_to_historical = deaths_all_causes / mean_annual_deaths_2015_2019,
-    ratio_non_covid_deaths_to_historical = deaths_non_covid / mean_annual_deaths_2015_2019,
-    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
-  )
-
-write_csv(x = table_current_deaths_compared_to_past, file = file.path(dir_outputs, "annual_death_rates_comparison_with_historical_deaths.csv"))
-
-table_current_deaths_compared_to_past %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
-```
-
-# Comparing historical rates of home death to total death to 2020/2021
-
-```{r}
-proportions_of_historical_deaths_by_place <-
-  past_deaths_overall %>% 
-  group_by(year, place_of_death) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%
-  filter(place_of_death != "All" & !is.na(as.integer(year))) %>%
-  group_by(year) %>%
-  mutate(
-    total_deaths = sum(number_of_deaths)
-  ) %>%
-  ungroup %>%
-  mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))
-
-proportions_of_current_deaths_by_place <-
-  total_deaths_overall %>%
-  select(year, place_of_death, number_of_deaths = deaths_all_causes) %>%
-  filter(place_of_death != "All") %>%
-  group_by(year) %>%
-  mutate(
-    total_deaths = sum(number_of_deaths)
-  ) %>%
-  ungroup %>%
-  mutate(proportion_of_total = number_of_deaths / total_deaths)
-
-merged_proportions_of_deaths_by_place <-
-  bind_rows(
-    proportions_of_historical_deaths_by_place,
-    proportions_of_current_deaths_by_place
-  )
-```
-
-## Including non-COVID deaths
-
-Note: non-covid deaths are defined as all-cause deaths minus covid-related deaths.
-Otherwise the number of non-covid deaths only includes non-covid deaths once covid-related deaths started being reported in 2020 (week 12).
-
-```{r}
-proportions_of_current_non_covid_deaths_by_place <-
-  total_deaths_overall %>%
-  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
-  select(year, place_of_death, number_of_deaths = deaths_non_covid) %>%
-  filter(place_of_death != "All") %>%
-  group_by(year) %>%
-  mutate(
-    total_deaths = sum(number_of_deaths)
-  ) %>%
-  ungroup %>%
-  mutate(proportion_of_total = number_of_deaths / total_deaths)
-
-merged_proportions_of_non_covid_deaths_by_place <-
-  bind_rows(
-    proportions_of_historical_deaths_by_place,  # all historical deaths are non-covid!
-    proportions_of_current_non_covid_deaths_by_place
-  )
-```
-
-# Deaths over time
-
-## Number of deaths by year
-
-```{r}
-# TODO: finish this here or elsewhere!
-# total_deaths_overall
-# 
-# past_deaths_overall %>%
-#   group_by(year, place_of_death) %>%
-#   summarise(number_of_deaths = sum(number_of_deaths, na.rm = TRUE), .groups="drop")
-# 
-# # TODO: can this be generalised in a function?
-# merged_deaths_age %>%
-#   # filter(place_of_death %in% c("Home & other non-institution")) %>%
-#   group_by(year, place_of_death, age, week_number) %>%
-#   summarise(
-#     deaths_all_causes = sum(deaths_all_causes),
-#     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-#     .groups = "drop"
-#   ) %>% 
-#   mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
-#   left_join(
-#     past_deaths_age %>% filter(year != "average") %>% group_by(week_number, place_of_death, age) %>% summarise(
-#                                        average_weekly_deaths = sum(number_of_deaths) / 5,
-#                                        .groups = "drop"
-#                                      ),
-#     by = c("week_number", "place_of_death", "age")
-#   ) %>% 
-#   group_by(place_of_death, age) %>%
-#   summarise(
-#     period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
-#     deaths_all_causes = sum(deaths_all_causes),
-#     deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
-#     deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
-#     mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
-#     .groups = "drop"
-#   ) %>%
-#   mutate(
-#     percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
-#     percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
-#     ) %>%
-#   (function(x) {
-#     bind_rows(
-#       x,
-#       x %>% group_by(age, period) %>% summarise(
-#         deaths_all_causes = sum(deaths_all_causes),
-#         deaths_non_covid = sum(deaths_non_covid),
-#         mean_annual_deaths_2015_2019 = sum(mean_annual_deaths_2015_2019)
-#       ) %>%
-#         mutate(
-#           percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1,
-#           percent_increase_non_covid_deaths_over_historic = (deaths_non_covid / mean_annual_deaths_2015_2019) - 1
-#         )
-#     )
-#   }) %>%
-#   View()
-#   # kable(caption = "Total deaths at home in data so far for age groups.")
+# Load most recent data
+
+```{r}
+load(file = "./workspace.RData")
 ```
@@
 # Visualisations
+
+Note: visualisations have been moved to a separate .Rmd file
 
@@
   }) %>%
-  View()
-  # kable(caption = "Total deaths at home in data so far for age groups.")
+  # View()
+  kable(caption = "Total deaths at home in data so far for age groups.")
 ```
@@
 
-## Overall deaths at home 2020 by cause of death
-
-```{r}
-# TODO: add yearly range for all deaths
-# TODO: split data by covid-underlying cause of desath & covid-factor in death
-merged_deaths_overall %>%
-  filter(sex=="all", age=="all") %>%
-  filter(year==2020) %>%
-  filter(place_of_death == "Home & other non-institution") %>%
-  group_by(ref_area, year) %>%
-  summarise(
-    deaths_all_causes = sum(deaths_all_causes, na.rm=TRUE),
-    deaths_covid_related = sum(deaths_covid_related, na.rm=TRUE),
-    deaths_non_covid = sum(deaths_non_covid, na.rm=TRUE)
-  )
+# Deaths by cause of death
+
+## Annual deaths by cause
+
+```{r}
+(figure_annual_deaths_by_cause <-
+  total_deaths_cause_of_death %>%
+  filter(year == 2020 & cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
+  # pivot_longer(cols = c(deaths, annual_deaths_mean), names_to = "deaths_type", values_to = "deaths") %>%
+  ggplot() +
+  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), colour = "grey10") +
+  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  scale_fill_viridis_d() +
+  facet_wrap(~place_of_death) +
+  NULL
+)
+```
+
+## By cause, at home
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+(figure_deaths_at_home_by_cause <-
+  merged_deaths_cause_of_death %>%
+  filter(str_starts(place_of_death,"Home")) %>%
+  filter(cause_of_death != "All") %>%
+  # filter(!place_of_death %in% c("All","Other")) %>%
+  pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
+  filter(!(deaths_type=="deaths_mean"&cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  ggplot(aes(x = date_w_c)) +
+  # annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_line(aes(y = number_of_deaths, colour = cause_of_death, linetype = deaths_type)) +
+  scale_colour_viridis_d() +
+  scale_linetype_manual(labels = c("Historic mean 2015-2019","Deaths since 2020"), values = c("dashed","solid")) +
+  # geom_line(aes(y = number_of_deaths  , colour = deaths_type)) +
+  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_wrap(cause_of_death~place_of_death, scales = "free_y") +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Weekly Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
+       ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_cause, width = 28, height = 14, dpi = 300, units = "cm")
 ```
@@
 ```{r}
-# save.image(file = "./workspace.RData", safe = TRUE)  # this is calling .GlobalEnv, which is empty
-save(list = ls(all.names = TRUE), file = "./workspace.RData", envir = environment())
+save(list = ls(all.names = TRUE), file = "./workspace_with_visualisations.RData", envir = environment())
 ```

### 2021-04-16 download_latest_data_and_render_reports.R
Split the .Rmd into one to deal with data and another to deal with visualisations for easier working

@@ -205,10 +205,16 @@ dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
 
 render(
-  input = "./covid_nrs_place_of_death.Rmd",
-  output_file = paste0("covid_nrs_place_of_death_",today()),
+  input = "./covid_nrs_place_of_death_data.Rmd",
+  output_file = paste0("covid_nrs_place_of_death_data_",today()),
   output_dir = dir_reports,
-  # params = params, 
+  envir = new.env()
+)
+
+render(
+  input = "./covid_nrs_place_of_death_visualisations.Rmd",
+  output_file = paste0("covid_nrs_place_of_death_visualisations",today()),
+  output_dir = dir_reports,
   envir = new.env()
 )
 

### 2021-04-16 using_UPRN_to_determine_household_structure.Rmd
Cleaned up UPRN proof of concept script

@@
 library(tidyverse)
+library(lubridate)  # for working with dates
 ```
@@
 
-# possible_dates_character <- 
-#   as.character(
-#     seq.Date(
-#       from = as.Date("2015-01-01"),  # TODO: complete this, and then sample from the vector instead, possibly character to make it less intense!
-#       to = as.Date("2020-12-31"),
-#       by = "days"
-#     ))
+possible_dates_character <-  # I thought that keeping dates as character vector until I need to compute with them would speed things up, but it's okay actually
+  as.character(
+    seq.Date(
+      from = as.Date("2015-01-01"),  # TODO: complete this, and then sample from the vector instead, possibly character to make it less intense!
+      to = as.Date("2020-12-31"),
+      by = "days"
+    ))
 
@@
 )
+
+# fake_deaths %>% ggplot(aes(x=date_death)) + geom_bar()
 
@@
     ) %>%
-  arrange(id_chi, date_address_change) %>%
-  group_by(id_chi) %>%
-  mutate(episode_id = 1:n()) %>%
-  ungroup
+  arrange(id_chi, date_address_change)
 
@@
   summarise(num_household_members = n_distinct(id_chi), .groups = "drop") %>%
-  tidyr::complete(id_death = fake_deaths$id_death, fill = list(num_household_members = 0))
+  tidyr::complete(id_death = fake_deaths$id_death, fill = list(num_household_members = 0))  # add records of individuals with no household members
 
-# chis_with_matching_uprn_episodic <- 
-#   chi_with_matching_uprn %>%
-#   select(id_chi, episode_id) %>%
-#   mutate(next_episode_id = episode_id + 1) %>%
-#   left_join(fake_chi, by = c("id_chi", "next_episode_id"="episode_id")) %>%
-#   mutate(date_episode_end = replace_na(date_address_change, replace = ymd("2021-01-01")))
-#
-
+number_of_household_members %>% ggplot(aes(x=num_household_members)) + geom_bar() + theme_minimal()
 ```

### 2021-04-16 visualisations_for_blog.Rmd
modified blog & paper scripts to load the workspace with visualsiations included
Added age group to blog visualisations

@@
 ```{r}
-load(file = "./workspace.RData")
+load(file = "./workspace_with_visualisations.RData")
 ```
@@
 
+### Table showing increases by area
+
 ```{r}
@@
 ```
+
+## Home deaths by age group
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_deaths_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_age %>% filter(year==2021) %>% .$week_number)} in 2021.")
+
+## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
+annotations_figure_home_deaths_by_age <- tribble(
+   ~geom, ~place_of_death, ~age, ~label, ~x, ~y, ~xend, ~yend,
+  "text", "Home & other non-institution", "45-64",  "Historic range (min & max)", 17, 20, NA, NA,
+  "curve", "Home & other non-institution", "45-64", NA, 22, 25, 25, 43,
+  "text", "Home & other non-institution", "45-64", "Historic mean", 42, 20, NA, NA,
+  "curve", "Home & other non-institution", "45-64", NA, 42, 25, 44, 64
+) %>% 
+  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020)))
+
+(figure_deaths_at_home_by_age <-
+  merged_deaths_age %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  filter(!age %in% c('0','1-14')) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  # ggplot(aes(x = week_number_run_over)) +
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
+  # geom_vline(xintercept = 54, linetype = "dotted") +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +
+  # facet_wrap(~age, scales = "free_y") +
+  facet_wrap(~age) +
+  # theme_minimal() +
+
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+   theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(removed_under_15_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  geom_text(data = annotations_figure_home_deaths_by_age %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
+  geom_curve(data = annotations_figure_home_deaths_by_age %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), linetype = c("solid","dashed"), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=10, label = "2021", hjust = 0, colour = "grey40") +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_blog,"/figure_deaths_at_home_by_age"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_age, width = 12, height = 9, dpi = 300, units = "in")
+```
+
+### Table showing increases by age
+
+```{r}
+table_increase_home_deaths_by_age <-
+  merged_deaths_age %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  group_by(year, place_of_death, age, week_number) %>%
+  summarise(
+    deaths_all_causes = sum(deaths_all_causes),
+    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+    .groups = "drop"
+  ) %>% 
+  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+  left_join(
+    past_deaths_age %>% filter(year != "average") %>% group_by(week_number, place_of_death, age) %>% summarise(
+                                       average_weekly_deaths = sum(number_of_deaths) / 5,
+                                       .groups = "drop"
+                                     ),
+    by = c("week_number", "place_of_death", "age")
+  ) %>% 
+  group_by(place_of_death, age) %>%
+  summarise(
+    period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+    deaths_all_causes = sum(deaths_all_causes),
+    deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+    deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+    mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+    .groups = "drop"
+  ) %>%
+  mutate(percent_increase_deaths_over_historic = (deaths_all_causes / mean_annual_deaths_2015_2019) - 1) %>%
+  arrange(percent_increase_deaths_over_historic) %>%
+  mutate(
+    `Percent increase` = scales::percent(percent_increase_deaths_over_historic, accuracy = 0.1)
+    ) %>%
+  filter(!age %in% c("0", "1-14")) %>%  # remove the v.small groups
+  select(`Age Group` = age, `Home deaths` = deaths_all_causes, `Percent increase`)
+
+write_csv(x = table_increase_home_deaths_by_age, file = paste0(dir_blog,"/table_increase_home_deaths_by_age.csv"))
+```

### 2021-04-16 covid_nrs_place_of_death_visualisations.Rmd
adding cause of death breakdown

@@
 
+## Table of annual deaths by cause
+
+```{r}
+table_deaths_by_cause_2020 <-
+  total_deaths_cause_of_death %>%
+  select(-matches("min|max")) %>%
+  # filter(year == 2020 & cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
+  filter(year == 2020 & cause_of_death!="All") %>%
+  group_by(place_of_death) %>%
+  mutate(
+    proportion_annual_deaths = deaths / sum(deaths),
+    historical_proportion_annual_deaths = annual_deaths_mean / sum(annual_deaths_mean)
+    ) %>%
+  ungroup
+
+table_deaths_by_cause_2020 %>%
+  mutate(across(.cols = matches("ratio|prop"), .fns = ~scales::percent(.x, accuracy=0.1))) %>%
+  select(Period = year, "Place of death" = place_of_death, "Cause of death" = cause_of_death, "Annual deaths" = deaths, "Proportion of annual deaths" = proportion_annual_deaths, "Historic average" = annual_deaths_mean, "Historic proportion" = historical_proportion_annual_deaths, "Ratio 2020 to historic" = ratio_annual_deaths_to_historical_mean) %>%
+  write_csv(file = paste0(dir_outputs,"/table_causes_of_death_2020.csv"))
+```
+
 ## Annual deaths by cause
@@
   # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
-  geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  # geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
   scale_fill_viridis_d() +
@@
   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_line(aes(y = number_of_deaths, colour = cause_of_death, linetype = deaths_type)) +
+  # geom_line(aes(y = number_of_deaths, colour = cause_of_death, linetype = deaths_type)) +
+  geom_line(aes(y = number_of_deaths, linetype = deaths_type)) +
   scale_colour_viridis_d() +
@@
        subtitle = date_range_text,
-       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
+       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,source_nrs_text), collapse="\n")
        ) +

### 2021-04-16 covid_nrs_place_of_death_visualisations.Rmd
Added graph of deaths by cause

@@
   ggplot() +
-  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
+  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, alpha = deaths_type), position = "dodge", colour = "grey10") +
   geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), colour = "grey10") +
-  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
-  # geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  geom_point(aes(x = cause_of_death, y = annual_deaths_mean)) +
+  coord_flip() +
   scale_fill_viridis_d() +
   facet_wrap(~place_of_death) +
-  NULL
-)
+  # scale_alpha_manual(
+  #   values = c(0.4,1), labels = c("Historic mean", "2020 deaths")
+  # ) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    legend.position = "top", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    subtitle = "Bars are 2020 deaths, dots are (•) mean annual deaths 2015-2019"
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_annual_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_annual_deaths_by_cause, width = 12, height = 6, dpi = 300, units = "in")
 ```

### 2021-05-04 paper_reporting_death_at_home_increase.Rmd
modified blog & paper scripts to load the workspace with visualsiations included
added figure 2 and added TIFF files

@@
 ## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_paper,"/figure1"), extensions = c(".pdf",".svg",".png"), plot = figure1, width = 28, height = 11, units = "cm", dpi = 300)
+save_output_file(filename = paste0(dir_paper,"/figure1"), extensions = c(".pdf",".svg",".png",".tiff"), plot = figure1, width = 28, height = 11, units = "cm", dpi = 300)
+
+save_output_file(filename = paste0(dir_paper,"/figure1_hires"), extensions = c(".png"), plot = figure1, width = 28, height = 11, dpi = 1200, units = "cm")
 ```
@@
 save_output_file(filename = paste0(dir_paper,"/figure2cd_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2cd_bw, width = 20, height = 12, dpi = 300, units = "cm")
+
+figure2 <-
+  wrap_plots(
+    list(
+      figure2c_bw,
+      figure2d_bw
+      ), 
+    nrow = 1)
+
+save_output_file(filename = paste0(dir_paper,"/figure2"), extensions = c(".pdf",".wmf",".png",".tiff"), plot = figure2, width = 20, height = 12, dpi = 300, units = "cm")
+
+save_output_file(filename = paste0(dir_paper,"/figure2_hires"), extensions = c(".png"), plot = figure2, width = 20, height = 12, dpi = 1200, units = "cm")
 ```

### 2021-05-05 paper_reporting_death_at_home_increase.Rmd
added figure and table 3 - age group breakdowns

@@
 
+## Table of deaths by age
+
+```{r}
+table_3_paper <-
+  total_deaths_age %>%
+  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
+  filter(!age %in% c("0","1-14")) %>%
+  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related) %>%
+  mutate(
+    `Percent increase (All home deaths)` = scales::percent(ratio_annual_deaths_to_historical_mean - 1, accuracy = 0.1)
+    ) %>%
+  select(`Age` = age, `All Deaths (Home)` = deaths_all_causes, `Non-covid deaths (Home)` = deaths_non_covid_2020, `Percent increase (All home deaths)`)
+
+write.xlsx(x = table_3_paper, file = paste0(dir_paper, "/percent_increase_by_age.xlsx"))
+```
+
 # Visualisations
@@
 
+## Grayscale plot of weekly deaths by age, 2020
+
+```{r}
+(figure3 <-
+  merged_deaths_age %>%
+  filter(year == 2020) %>%
+  filter(!age %in% c("0","1-14")) %>%  # remove youngest age groups
+  # filter(!place_of_death %in% c("All","Other")) %>%
+  filter(str_starts(place_of_death,"Home")) %>%
+  # mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(fill_text = "Historical range") %>%  # added for ease of using aes(fill) later
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid, deaths_mean), 
+               names_to = "deaths_type", values_to = "deaths") %>%
+  mutate(deaths_type = factor(x = deaths_type, levels = c("deaths_all_causes", "deaths_non_covid", "deaths_mean"), labels = c("All deaths", "Non Covid-19 deaths", "Historical mean"))) %>%
+  ggplot(aes(x = week_number_run_over)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = fill_text), alpha = 0.5) +
+  geom_line(aes(y = deaths, linetype = deaths_type, colour = deaths_type)) +
+  scale_colour_manual(values = c("black", "black", "gray30")) +
+  scale_x_continuous(breaks = seq(0,60,by=10)) +
+  ylim(0, NA) +  # start scale at 0 for better understanding the scale of deaths
+  facet_wrap(~age) +
+  scale_fill_manual(values = c("grey70")) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    legend.box.spacing = unit(0, "cm"),
+    plot.caption = element_text(size = 10, colour = "gray60")
+    ) +
+  labs(x = "Week number",
+       y = "Deaths (N)",
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n")
+       ) +
+  NULL
+)
+## figure for a bit less than half A4 page is approx 14*11cm
+save_output_file(filename = paste0(dir_paper,"/figure3"), extensions = c(".pdf",".svg",".png",".tiff"), plot = figure3, width = 28, height = 11, units = "cm", dpi = 300)
+
+save_output_file(filename = paste0(dir_paper,"/figure3_hires"), extensions = c(".png"), plot = figure3, width = 28, height = 11, dpi = 1200, units = "cm")
+```
+
 ## Maps: home death increase & case counts

### 2021-05-05 download_latest_data_and_render_reports.R
modified download function so that it works with dynamically changing filenames/urls for the weekly report of covid-related deaths

@@ -72,7 +72,7 @@ if (file.exists(file_historical_data_sources)) {
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx"
     )
   ) %>% mutate(
-    filename = basename(url),
+    # filename = basename(url),
     last_modified = ymd_hms("2020-01-01_01:01:01"),  # start with a date in the past
     path_latest = NA_character_  
   )
@@ -98,20 +98,25 @@ if (file.exists(file_case_trends_data_sources)) {
 
 # Function to download latest data if available ---------------------------
 
-download_file_if_newer_available <- function(url, filename, last_modified, directory, record_tbl) {
+# download_file_if_newer_available <- function(url, filename, last_modified, directory, record_tbl) {
+download_file_if_newer_available <- function(url, short_name, filename=NULL, last_modified, directory, record_tbl) {  # note:removed filename argument, I can get it from the url!
   todays_file <- curl_fetch_disk(url, path = tempfile())
-  message(paste0("Checking if we have latest file for: ",filename))
+  message(paste0("Checking if we have latest file for: ",short_name))
+  message(paste0("The latest local file is: ", basename(record_tbl$path_latest[which(record_tbl$short_name == short_name)])))
   message(paste0("Latest file was last modified at: ",todays_file$modified))
   if (last_modified < todays_file$modified) {
     message("Newer file available, downloading...")
     date_string <- strftime(todays_file$modified, format = "%F")  # prefix filename with last modified date
     ## copy file to folder
-    filename_prefixed <- paste0(date_string,"_",filename)
-    file_path <- file.path(directory, filename_prefixed)
+    if (is.null(filename)) {  # if a filename is not provided infer name from url
+      filename <- basename(url)
+      }
+    file_path <-  file.path(directory, paste0(date_string,"_",filename))
     file.copy(from = todays_file$content, to = file_path, overwrite = TRUE)
     ## set last_modified date & save
-    record_tbl$last_modified[which(record_tbl$filename == filename)] <- todays_file$modified
-    record_tbl$path_latest[which(record_tbl$filename == filename)] <- file_path
+    record_tbl$last_modified[which(record_tbl$short_name == short_name)] <- todays_file$modified
+    record_tbl$filename[which(record_tbl$short_name == short_name)] <- filename
+    record_tbl$path_latest[which(record_tbl$short_name == short_name)] <- file_path
   } else {
[…]

### 2021-05-06 covid_nrs_place_of_death_visualisations.Rmd
updated cause of death figure

--- 2021-04-16 covid_nrs_place_of_death_visualisations.Rmd
@@
 
+## get "canonical" order for causes of death in 2020
+order_of_causes_of_death <- table_deaths_by_cause_2020 %>% filter(place_of_death == "All") %>% arrange(desc(deaths)) %>% pull(cause_of_death)
+
+## sort the table by "canonical" order
+table_deaths_by_cause_2020 <-
+  table_deaths_by_cause_2020 %>%
+  group_by(year, place_of_death) %>%
+  arrange(cause_of_death = factor(cause_of_death, levels = order_of_causes_of_death), .by_group = TRUE) %>%
+  ungroup
+
 table_deaths_by_cause_2020 %>%
@@
   # pivot_longer(cols = c(deaths, annual_deaths_mean), names_to = "deaths_type", values_to = "deaths") %>%
+   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+   mutate(
+     cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+     cause_of_death = fct_rev(cause_of_death)
+     ) %>%
   ggplot() +
-  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
-  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), colour = "grey10") +
-  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
-  # geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, alpha = deaths_type), position = "dodge", colour = "grey10") +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
+  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean)) +
+  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = annual_deaths_mean, yend = annual_deaths_mean), size = 1) +
+  coord_flip() +
   scale_fill_viridis_d() +
   facet_wrap(~place_of_death) +
-  NULL
-)
+  # scale_alpha_manual(
+  #   values = c(0.4,1), labels = c("Historic mean", "2020 deaths")
+  # ) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    # legend.position = "top", 
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    # subtitle = "Bars are 2020 deaths, dots are (•) mean annual deaths 2015-2019"
+    subtitle = "Bars are 2020 deaths, line segments are mean annual deaths 2015-2019"
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_annual_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_annual_deaths_by_cause, width = 12, height = 6, dpi = 300, units = "in")
 ```

### 2021-05-06 covid_nrs_place_of_death_visualisations.Rmd
updated cause of death graph and table & made figure 4 from it for the paper

--- 2021-04-16 covid_nrs_place_of_death_visualisations.Rmd
@@
 
+## get "canonical" order for causes of death in 2020
+order_of_causes_of_death <- table_deaths_by_cause_2020 %>% filter(place_of_death == "All") %>% arrange(desc(deaths)) %>% pull(cause_of_death)
+
+## sort the table by "canonical" order
+table_deaths_by_cause_2020 <-
+  table_deaths_by_cause_2020 %>%
+  group_by(year, place_of_death) %>%
+  arrange(cause_of_death = factor(cause_of_death, levels = order_of_causes_of_death), .by_group = TRUE) %>%
+  ungroup
+
 table_deaths_by_cause_2020 %>%
   mutate(across(.cols = matches("ratio|prop"), .fns = ~scales::percent(.x, accuracy=0.1))) %>%
+  mutate(across(.cols = matches("ratio|^historical"), .fns = ~if_else(condition = cause_of_death=="COVID-19", true = "", false = .x))) %>%
   select(Period = year, "Place of death" = place_of_death, "Cause of death" = cause_of_death, "Annual deaths" = deaths, "Proportion of annual deaths" = proportion_annual_deaths, "Historic average" = annual_deaths_mean, "Historic proportion" = historical_proportion_annual_deaths, "Ratio 2020 to historic" = ratio_annual_deaths_to_historical_mean) %>%
@@
   filter(year == 2020 & cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
-  # pivot_longer(cols = c(deaths, annual_deaths_mean), names_to = "deaths_type", values_to = "deaths") %>%
+   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+   mutate(
+     cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+     cause_of_death = fct_rev(cause_of_death)
+     ) %>%
   ggplot() +
-  # geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths, group = deaths_type, linetype = deaths_type), position = "dodge", colour = "grey10") +
-  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), colour = "grey10") +
-  # geom_point(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
-  # geom_line(aes(x = cause_of_death, y = annual_deaths_mean), colour = "grey10") +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
+  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = annual_deaths_mean, yend = annual_deaths_mean), size = 0.8, colour = "grey1") +
+  coord_flip() +
   scale_fill_viridis_d() +
   facet_wrap(~place_of_death) +
-  NULL
-)
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    subtitle = "Bars are 2020 deaths, line segments are mean annual deaths 2015-2019"
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_annual_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_annual_deaths_by_cause, width = 10, height = 3, dpi = 300, units = "in")
 ```

### 2021-05-06 paper_reporting_death_at_home_increase.Rmd
updated cause of death graph and table & made figure 4 from it for the paper

@@
 ```
+
+## Causes of death
+
+```{r}
+(figure4 <-
+  figure_annual_deaths_by_cause +
+  scale_fill_grey() +
+  labs(x = NULL, y = NULL, title = NULL, subtitle = NULL)
+)
+
+save_output_file(filename = paste0(dir_paper,"/figure4"), extensions = c(".pdf",".svg",".png"), plot = figure4, width = 10, height = 3, dpi = 300, units = "in")
+save_output_file(filename = paste0(dir_paper,"/figure4_hires"), extensions = c(".png"), plot = figure4, width = 10, height = 3, dpi = 1200, units = "in")
+```

### 2021-05-14 download_latest_data_and_render_reports.R
hacked up a solution for downloading the latest main report from the NRS data. it's a mess now, but it works (I think so anyway!)

@@ -100,6 +100,9 @@ if (file.exists(file_case_trends_data_sources)) {
 
 # download_file_if_newer_available <- function(url, filename, last_modified, directory, record_tbl) {
 download_file_if_newer_available <- function(url, short_name, filename=NULL, last_modified, directory, record_tbl) {  # note:removed filename argument, I can get it from the url!
+  if (short_name == "main_report") {
+    url <- weekly_deaths_url_absolute  # replace with latest url of main report 
+  }
   todays_file <- curl_fetch_disk(url, path = tempfile())
   message(paste0("Checking if we have latest file for: ",short_name))
   message(paste0("The latest local file is: ", basename(record_tbl$path_latest[which(record_tbl$short_name == short_name)])))
@@ -114,6 +117,9 @@ download_file_if_newer_available <- function(url, short_name, filename=NULL, las
     file_path <-  file.path(directory, paste0(date_string,"_",filename))
     file.copy(from = todays_file$content, to = file_path, overwrite = TRUE)
     ## set last_modified date & save
+    if (short_name == "main_report") {
+      record_tbl$url <- url  # replace with latest url of main report 
+    }
     record_tbl$last_modified[which(record_tbl$short_name == short_name)] <- todays_file$modified
     record_tbl$filename[which(record_tbl$short_name == short_name)] <- filename
     record_tbl$path_latest[which(record_tbl$short_name == short_name)] <- file_path

### 2021-05-14 covid_nrs_place_of_death_visualisations.Rmd
added a plot of deaths over time by place, "normalised" so that the historic mean is set to 0 and only deviations are plotted.
added a plot of causes of death from 2020 until the latest available data.

@@
 
+### Place of death, deviation from mean
+
+```{r}
+(figure_place_of_death_deviation_from_mean <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>%
+  mutate(
+    deaths = deaths - deaths_mean,
+    deaths_min = deaths_min - deaths_mean,
+    deaths_max = deaths_max - deaths_mean
+  ) %>%
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  annotate(geom = "text", x=ymd("2021-01-07"), y=-320, label = "2021", hjust = 0, colour = "grey40") +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = 0), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Weekly Deaths: difference from 2015-2019 mean (N)",
+       subtitle = paste0(c(date_range_text), collapse="\n"),
+       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death_deviation_from_mean"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death_deviation_from_mean, width = 28, height = 14, dpi = 300, units = "cm")
+```
+
 ## All of Scotland, deaths at home, for policy briefing
@@
 
-## Table of annual deaths by cause
+## Table of 2020 annual deaths by cause
 
@@
 save_output_file(filename = paste0(dir_outputs,"/figure_annual_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_annual_deaths_by_cause, width = 10, height = 3, dpi = 300, units = "in")
+```
+
+## Deaths by cause, from 2020 until current day
+
+```{r}
+(figure_deaths_by_cause <-
+  merged_deaths_cause_of_death %>%
+  filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+   cause_of_death = fct_rev(cause_of_death)
+ ) %>%
+  group_by(place_of_death, cause_of_death) %>%
+  summarise(
+    deaths = sum(number_of_deaths),
+    deaths_mean = sum(deaths_mean),
+    .groups="drop"
+  ) %>%
+  ggplot() +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
+  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
+  coord_flip() +
+  scale_fill_viridis_d() +
+  facet_wrap(~place_of_death) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean annual deaths 2015-2019"),collapse="\n")
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause, width = 10, height = 3, dpi = 300, units = "in")
 ```

### 2021-05-18 covid_nrs_place_of_death_data.Rmd
added a table of excess deaths

@@
   )
+```
+
+# Excess deaths
+
+## Excess deaths by place of death
+
+```{r}
+total_deaths_overall %>%
+  mutate(excess_deaths = deaths_all_causes - annual_deaths_min)
+
+excess_deaths_2020_table <-
+  past_deaths_overall %>%
+  group_by(year, place_of_death) %>%
+  summarise(annual_deaths = sum(number_of_deaths), .groups = "drop") %>%
+  filter(year==2019) %>%
+  pivot_wider(names_from = year, names_prefix = "annual_deaths_", values_from = annual_deaths) %>%
+  left_join(
+    total_deaths_overall %>% filter(year==2020), by = "place_of_death"
+    ) %>%
+  mutate(
+    excess_deaths_vs_2019 = deaths_all_causes - annual_deaths_2019,
+    # excess_deaths_vs_2019_prop = excess_deaths_vs_2019 / sum(excess_deaths_vs_2019),
+    excess_deaths_vs_mean = deaths_all_causes - annual_deaths_mean,
+    excess_deaths_vs_min = deaths_all_causes - annual_deaths_min,
+    excess_deaths_vs_max = deaths_all_causes - annual_deaths_max
+  ) %>%
+  select(year, place_of_death, annual_deaths_2019, annual_deaths_mean, deaths_all_causes, matches("excess")) %>%
+  mutate(disaggregated_place_of_death = place_of_death != "All") %>%
+  group_by(disaggregated_place_of_death) %>%
+  mutate(
+    proportion_excess_deaths_vs_2019 = excess_deaths_vs_2019 / sum(excess_deaths_vs_2019),
+    proportion_excess_deaths_vs_mean = excess_deaths_vs_mean / sum(excess_deaths_vs_mean)
+  ) %>%
+  ungroup %>%
+  select(-disaggregated_place_of_death)
+
+excess_deaths_table_2019 <-
+  excess_deaths_2020_table %>%
+  select(place_of_death, annual_deaths_2019, deaths_all_causes, excess_deaths_vs_2019, proportion_excess_deaths_vs_2019) %>%
+  mutate(
+    proportion_excess_deaths_vs_2019 = scales::percent(proportion_excess_deaths_vs_2019, accuracy = 0.1)
+    ) %>%
+  mutate_if(is.numeric, ~format(.x, big.mark=",")) %>%
+  rename(
+    "Place of death" = place_of_death,
+    "2019 deaths" = annual_deaths_2019,
+    "2020 deaths (all causes)" = deaths_all_causes,
+    "Excess deaths" = excess_deaths_vs_2019,
+    "Proportion of excess deaths by place" = proportion_excess_deaths_vs_2019
+  )
+
+write.xlsx(x = list("Excess w.r.t. 2019" = excess_deaths_table_2019), file = paste0(dir_outputs,"/excess_deaths_Scotland_compared_to_2019.xlsx"))
 ```

### 2021-06-09 covid_nrs_place_of_death_visualisations.Rmd
- added deviation from mean graph showing deviations in absolute terms - added graph with both 2020 and 2021 trends on same week number scale, for easier comparison of same period both years; made 2020 transparent, but become more opaque later in the year when there isn't 2021 data yet, which I think is pretty neat!

@@
 
-## All of Scotland, deaths at home, for policy briefing
+## Deaths at home, for policy briefing
 
@@
 
-## All of Scotland, deaths at home - date axis
+## Deaths at home - date axis
 
@@
 
-## All of Scotland, deaths at home - week number
+### Home, deviation from mean
+
+```{r}
+(figure_deaths_at_home_deviation_from_mean <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>%
+  mutate(
+    deaths = deaths - deaths_mean,
+    deaths_min = deaths_min - deaths_mean,
+    deaths_max = deaths_max - deaths_mean
+  ) %>%
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = 0), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Weekly Deaths: difference from 2015-2019 average (N)",
+       subtitle = paste0(c(date_range_text), collapse="\n"),
+       caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+   ## annotate historic range
+  annotate(geom = "text", x=ymd("2021-01-07"), y=-120, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 16, year_number = 2020), y = -90, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 14, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 18, year_number = 2020), y = -80, yend = -50, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 37, year_number = 2020), y = -120, label = "Historic weekly average, set to 0", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 37, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 42, year_number = 2020), y = -110, yend = -5, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  ## annotate year
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_deviation_from_mean"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_deviation_from_mean, width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+## Deaths at home - week number
 
 ```{r, fig.width=12, fig.height=9, warning=FALSE}
-figure_deaths_at_home_wwek_number <-
+figure_deaths_at_home_week_number <-
   merged_deaths_overall %>%
@@
 ## figure for a bit less than half A4 page is approx 14*11cm
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_wwek_number"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_wwek_number, width = 18, height = 15, units = "cm", dpi = 300)
-
-figure_deaths_at_home_wwek_number
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_week_number"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home_week_number, width = 18, height = 15, units = "cm", dpi = 300)
+
+figure_deaths_at_home_week_number
 ```
@@
 
+## 2020 and 2021 on same graph
+
+For these graphs, we use the week number, and superimpose lines for 2020 and 2021!
+
+### Home deaths, both years
+
+```{r}
+(figure_deaths_at_home_2020_and_2021 <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
+  mutate(year = factor(year)) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
+               names_to = "deaths_type", values_to = "deaths") %>%
+  ggplot(aes(x = week_number)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  # geom_line(aes(y = deaths, colour = deaths_type, alpha = year)) +
+  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number), show.legend = FALSE) +
+  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
+  scale_x_continuous(breaks = seq(0,60,by=12)) +
+  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths",
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  scale_alpha_continuous(range = c(0.2, 0.8), guide = NULL) +
+  annotate(geom = "text", x = 18, y = 550, label = "2020", alpha = 0.6) +
+  annotate(geom = "text", x = 4, y = 530, label = "2021") +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Weekly deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  annotate(geom = "text", x = 10, y = 100, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom = "curve", x = 10, xend = 14, y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x = 20, y = 50, label = "Historic mean", colour = "grey20") +
+  annotate(geom = "curve", x = 20, xend = 25, y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020_and_2021, width = 18, height = 14, dpi = 300, units = "cm")
+```
+
 # Deaths by cause of death
@@
   filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
+  # filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
@@
 save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause, width = 10, height = 3, dpi = 300, units = "in")
+
+(figure_deaths_by_cause_across_places <-
+  merged_deaths_cause_of_death %>%
+  filter(cause_of_death!="All" & place_of_death %in% c("All")) %>%
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+   cause_of_death = fct_rev(cause_of_death)
+ ) %>%
+  group_by(place_of_death, cause_of_death) %>%
+  summarise(
+    deaths = sum(number_of_deaths),
+    deaths_mean = sum(deaths_mean),
+    .groups="drop"
+  ) %>%
+  ggplot() +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
+  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
+  coord_flip() +
+  scale_fill_viridis_d() +
+  facet_wrap(~place_of_death) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean annual deaths 2015-2019"),collapse="\n")
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_cause_all_places"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause_across_places, width = 7, height = 3, dpi = 300, units = "in")
 ```

### 2021-06-25 covid_nrs_place_of_death_visualisations.Rmd
added some more detail to visualisations of causes of death over time; added two plots for blog based on cause fo death visualisations

@@
     y = "Deaths",
-    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean annual deaths 2015-2019"),collapse="\n")
+    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean deaths for same period 2015-2019"),collapse="\n")
   ) +
@@
   filter(cause_of_death != "All") %>%
-  # filter(!place_of_death %in% c("All","Other")) %>%
   pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
-  filter(!(deaths_type=="deaths_mean"&cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
+  mutate(deaths_type = if_else(deaths_type == "deaths_mean", true = "Historic mean 2015-2019", false = "Deaths in 2020 & 2021")) %>%
+  filter(!(deaths_type=="deaths_mean" & cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
   ggplot(aes(x = date_w_c)) +
-  # annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "skyblue", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  annotation_custom(grid::textGrob(label = "2021",  # solution to place annotation in same place in each facet
+                                 x = unit(0.75, "npc"), 
+                                 y = unit(0.95, "npc"),
+                                 gp = grid::gpar(col = "black"))) +
   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  # geom_line(aes(y = number_of_deaths, colour = cause_of_death, linetype = deaths_type)) +
-  geom_line(aes(y = number_of_deaths, linetype = deaths_type)) +
-  scale_colour_viridis_d() +
-  scale_linetype_manual(labels = c("Historic mean 2015-2019","Deaths since 2020"), values = c("dashed","solid")) +
-  # geom_line(aes(y = number_of_deaths  , colour = deaths_type)) +
-  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  facet_wrap(cause_of_death~place_of_death, scales = "free_y") +
+  geom_line(aes(y = number_of_deaths, colour = deaths_type, linetype = deaths_type)) +
+  scale_linetype_manual(values = c("solid", "dashed")) +
+  scale_colour_manual(values = c("grey10","grey70")) +
+  facet_wrap(cause_of_death~., scales = "free_y") +
   theme(
@@
        y = "Weekly Deaths (N)",
-       subtitle = date_range_text,
+       subtitle = paste0(c("Deaths registered at home", date_range_text), collapse = "\n"),
        caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,source_nrs_text), collapse="\n")

### 2021-06-25 visualisations_for_blog.Rmd
added some more detail to visualisations of causes of death over time; added two plots for blog based on cause fo death visualisations

@@
 ```
+
+## Deaths at home - noting point when 1st blog published
+
+I'll reuse `figure_deaths_at_home` here, but make a note of the data being available up until 15th February when the first blog was published, to comment on the number of deaths since.
+
+```{r}
+weeks_available_for_first_blog <- isoweek(ymd("2021-02-15"))
+
+(figure_place_of_death_noting_first_blog_data <-
+  figure_deaths_at_home +
+  geom_vline(xintercept = ymd("2021-02-15"), linetype = "dotted", colour = "orange", size = 1.5) +
+  annotate(geom = "label", x = ymd("2021-02-15"), y = 550, fill = "orange", label = "Data that was available in 1st blog", hjust = 1)
+)
+
+save_output_file(filename = paste0(dir_blog,"/figure_place_of_death_noting_first_blog_data"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death_noting_first_blog_data, width = 12, height = 9, dpi = 300, units = "in")
+```
+
+## Proportions covid deaths by week:
+
+```{r}
+merged_deaths_overall %>%
+    filter(sex == "all" & age == "all") %>%
+    filter(place_of_death %in% c("Home & other non-institution")) %>%
+    select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>% filter(week_number > 12) %>% select(year, week_number, deaths_all_causes, deaths_covid_related) %>% mutate(percent_covid = scales::percent(deaths_covid_related/deaths_all_causes, accuracy = 0.1)) %>% print(n=nrow(.))
+
+## maximum covid deaths
+merged_deaths_overall %>%
+    filter(sex == "all" & age == "all") %>%
+    filter(place_of_death %in% c("Home & other non-institution")) %>%
+    slice_max(order_by = deaths_covid_related, n = 8) %>%
+    select(year, week_number, deaths_all_causes, deaths_covid_related) %>% mutate(percent_covid = scales::percent(deaths_covid_related/deaths_all_causes, accuracy = 0.1)) %>% print(n=nrow(.))
+```
+
+## Causes of death up until now
+
+```{r}
+(figure_deaths_by_cause_combined <-
+  (figure_deaths_by_cause_across_places + labs(y = NULL)) /  # coordinates were flipped for these graphs!
+  figure_deaths_by_cause + labs(subtitle = NULL))
+```
+
+## Causes of death at home up until now
+
+```{r}
+(figure_deaths_by_cause_at_home <-
+  merged_deaths_cause_of_death %>%
+  # filter(cause_of_death!="All" & !place_of_death %in% c("Other","All")) %>%
+  filter(cause_of_death!="All" & str_starts(place_of_death,"Home")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 20), levels = order_of_causes_of_death %>% str_wrap(., width = 20)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+   cause_of_death = fct_rev(cause_of_death)
+ ) %>%
+  group_by(place_of_death, cause_of_death) %>%
+  summarise(
+    deaths = sum(number_of_deaths),
+    deaths_mean = sum(deaths_mean),
+    .groups="drop"
+  ) %>%
+  ggplot() +
+  geom_col(aes(x = cause_of_death, fill = cause_of_death, y = deaths), alpha = 0.8, colour = "grey10") +
+  geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
+  coord_flip() +
+  scale_fill_viridis_d() +
+  facet_wrap(~place_of_death) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are average deaths for same period 2015-2019"),collapse="\n")
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_blog,"/figure_deaths_by_cause_at_home"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_cause_at_home, width = 7, height = 4, dpi = 300, units = "in")
+```
+
+## By cause, at home
+
+```{r, fig.width=12, fig.height=9, warning=FALSE}
+(figure_deaths_at_home_by_cause_over_time <-
+  merged_deaths_cause_of_death %>%
+  filter(str_starts(place_of_death,"Home")) %>%
+  filter(cause_of_death != "All") %>%
+  pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
+  mutate(deaths_type = if_else(deaths_type == "deaths_mean", true = "Historic average 2015-2019", false = "Deaths in 2020 & 2021")) %>%
+  filter(!(deaths_type=="deaths_mean" & cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  ggplot(aes(x = date_w_c)) +
+  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "skyblue", alpha = 0.2) +  # shade area - geom_rect doesn't work here
+  annotation_custom(grid::textGrob(label = "2021",  # solution to place annotation in same place in each facet
+                                 x = unit(0.9, "npc"), 
+                                 y = unit(0.95, "npc"),
+                                 gp = grid::gpar(col = "black"))) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
+  geom_line(aes(y = number_of_deaths, colour = deaths_type, linetype = deaths_type)) +
+  scale_linetype_manual(values = c("solid", "dashed")) +
+  scale_colour_manual(values = c("grey10","grey70")) +
+  facet_wrap(cause_of_death~., scales = "free_y") +
+  ylim(c(0,NA)) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Weekly Deaths (N)",
+       subtitle = paste0(c("Deaths registered at home", date_range_text), collapse = "\n"),
+       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,source_nrs_text), collapse="\n")
+       ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_blog,"/figure_deaths_at_home_by_cause_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_by_cause_over_time, width = 28, height = 14, dpi = 300, units = "cm")
+```

### 2021-07-29 covid_nrs_place_of_death_data.Rmd
there is an error in the openly available data that ODS pulls from, so I recompute the weekly_deaths_overall object from another source, namely from the sex-disaggregated dataset weekly_deaths_sex;
I've messaged the relevant person at ScotGov but for he time being this fix works

@@
 
-# weekly_deaths_sex_age_file <-
-#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
-#                 destfile = tempfile(),
-#                 quiet = FALSE)
-
 weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
@@
   summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
+```
+
+#### Fixing overall weekly deaths
+
+As of 28 July 2021, the [ScotGov open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19) reports the non-covid deaths as if they were all-cause deaths, and this messed up my entire pipeline!
+Below is a quick fix, where instead of using the open data API, I construct the 'overall' weekly deaths data, `weekly_deaths_overall`, from the NRS-published ad-hoc query data on deaths by sex.
+
+```{r}
+## quick fix whilst the open data for overall deaths doesn't work...
+# TODO: remove this chunk once no longer needed
+
+weekly_deaths_overall_averages_and_other_bits <-  # keep the nrs-computed past averages from the broken data, as well as other columns needed
+  weekly_deaths_overall %>%
+  select(-c(deaths_covid_related, deaths_non_covid, deaths_all_causes))
+  # select(age, sex, place_of_death, week_number_run_over, deaths_nrs_past_average_all_causes, ref_area, ref_period)
+
+weekly_deaths_overall <-  # this rewrites weekly_deaths_overall!
+  weekly_deaths_sex %>%
+  group_by_all() %>%
+  ungroup(sex, starts_with("deaths_")) %>%
+  summarise(across(starts_with("deaths_"), .fns = sum), .groups = "drop") %>%  # adds together figures between F & M
+  ({function (x) bind_rows(  # this takes the data at this point and a copy of it, and then summarises the copy across place of death, creating an 'All' category, the figures for which are sums across place of death
+    x,
+    x %>% group_by_all() %>% ungroup(place_of_death, starts_with("deaths_")) %>% summarise(across(starts_with("deaths_"), .fns = sum), .groups = "drop") %>% mutate(place_of_death = "All")
+  )}) %>%
+  mutate(sex = "all", age = "all") %>%  # add these so that I don't end up breaking any further analysis
+  arrange(week_number_run_over, place_of_death) %>%
+  left_join(
+    weekly_deaths_overall_averages_and_other_bits,
+    by = c("week_number", "week_number_run_over", "year", "date_w_c", "place_of_death", "sex", "age")
+    )  # add the average numbers in also so I don't break any further analysis
 ```

### 2021-07-29 exploring_isds1_data.Rmd
Started exploring ISD(S)1 data to work out what we will request for the linked data study

---
title: "Exploring ISD(S)1 Data"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
library(curl)  # for downloading stuff
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
```

# Introduction

ISD(S)1 contains annual & quarterly statistics of hospital bed usage amongst other things. It is submitted on a monthly basis so monthly data should also be available!

The structure of the file is complicated - it contains entries grouped by health board, location, and specialty, so it allocates beds to specialties within each hospital.

From the notes accompanying ISD(S)1:

>Please note this release includes Scotland going into emergency measures due to COVID-19. During this pandemic, NHS Boards, hospitals, and healthcare providers have been required to change their normal way of working to manage their COVID-19 response. As such, this is directly impacting on the volume of hospital activity and trends observed. For example, the inpatient and outpatient activity has decreased by just over a fifth when comparing October to December 2020 to the same quarter last year.

>For this publication, it was decided not to contact the NHS Boards directly to ask whether they were aware of data quality issues which may affect this publication, as we would normally do. This decision was taken in order to allow the NHS Boards to focus their efforts locally on their COVID-19 response.

>Many trends observed will be influenced by and attributable to the data completeness levels, small numbers, and the impact of COVID-19 highlighted above. Therefore, caution should be taken when comparing quarterly information.

>From 2020 onwards, there has been a marked increase in the use of the XSU (Unspecified) specialty across many of the NHS Boards. This is most likely due to their COVID response and not allocating beds to specific specialties.

>Records have been identified with potentially duplicate information coming from more than one NHS Board for the same location. The numbers concerned are very low and the impact is not significant. Additionally, there are issues with the allocation of NHS hospitals to NHS Boards and private hospital activity to the ‘non-NHS Provider code’. Data Management have investigated and found that in many of these cases, this can be attributed to visiting consultant activity and NHS Boards sending their patients to private locations.

## Availability within ODS

```{r}
all_datasets <- ods_all_datasets()

all_datasets %>% filter(str_detect(Name, pattern = regex("isd|bed|occup", ignore_case = TRUE)))
## none of these are exactly what we're looking for
```

# Download data & load

The latest ISD(S)1 dataset is available from the [OpenData NHS website]() - this includes a data dictionary also!

ISD(S)1 is also included in the [PHS release here](https://publichealthscotland.scot/publications/acute-hospital-activity-and-nhs-beds-information-quarterly/acute-hospital-activity-and-nhs-beds-information-quarterly-quarter-ending-31-december-2020/data-summary/)

the list of locations in use in the NHS is available from the [National Reference Files website of ISD](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/), with the layout described [at this address](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#2)

The list of specialty codes is available [from the National Reference Files](http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/specialt?63379634), with the [layout described here](http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#12)

```{r}
isds1_download_dir <- "./downloaded_data/"
isds1_url <- "https://www.opendata.nhs.scot/dataset/554b0e7a-ccac-4bb6-82db-1a8b306fcb36/resource/f272bb7d-5320-4491-84c1-614a2c064007/download/beds_by_nhs_board_of_treatment_and_specialty.csv"
isd1_local_path <- file.path(isds1_download_dir, "beds_by_nhs_board_of_treatment_and_specialty.csv")
if (!file.exists(isd1_local_path)) curl_download(url = isds1_url, destfile = isds1_local_path, quiet = FALSE)

location_codes_url <- "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/location?63379634"
location_codes_local_path <- file.path(isds1_download_dir, "location_codes")
if (!file.exists(location_codes_local_path)) curl_download(url = location_codes_url, destfile = location_codes_local_path, quiet = FALSE)

location_codes <- read_fwf(file = location_codes_local_path, fwf_cols(
  location_code = c(1,5),
  location_name = c(6,75)
))

specialty_codes_url <- "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/specialt?63379634"
specialty_codes_local_path <- file.path(isds1_download_dir, "specialty_codes")
if (!file.exists(specialty_codes_local_path)) curl_download(url = specialty_codes_url, destfile = specialty_codes_local_path, quiet = FALSE)

specialty_codes <- read_fwf(file = specialty_codes_local_path, fwf_cols(
  specialty_code = c(1,4),
  specialty_description = c(5,54)#,
  # lower_age = c(55,57),
  # upper_age = c(58,60)
))
```

# Human-readable fields

```{r}
isds1 <- read_csv(isd1_local_path, guess_max = 1e5) %>%
  mutate(
    year = as.integer(str_sub(Quarter, 1, 4)),
    quarter = as.integer(str_sub(Quarter, 6, 6)),
    health_board = match_area(HB),
    AvailableBeds = AllStaffedBeds - TotalOccupiedBeds
  ) %>%
  left_join(location_codes, by = c("Location" = "location_code")) %>%
  left_join(specialty_codes, by = c("Specialty" = "specialty_code"))

```

# Printing some of the ISD(S)1 data

```{r}
isds1 %>% count(health_board)

isds1 %>%
  filter(year == 2020 & quarter == 4) %>%
  filter(health_board == "Orkney") %>%
  count(location_name)

## palliative medicine data
isds1 %>%
  filter(year == 2020) %>%
  filter(!is.na(location_name)) %>%
  filter(specialty_description == "Palliative Medicine") %>%
  select(year, quarter, health_board, location_name, SpecialtyName, PercentageOccupancy, PercentageOccupancyQF) %>%
  print(n = nrow(.))

isds1 %>% 
  filter(!is.na(location_name)) %>%
  filter(year %in% 2015:2021) %>%
  filter(specialty_description == "Palliative Medicine") %>%
  select(Quarter, health_board, location_name, AllStaffedBeds, TotalOccupiedBeds, PercentageOccupancy) # %>%
  # mutate(testp = round(TotalOccupiedBeds/AllStaffedBeds*100, digits=1))  # this is exactly it

## occupancy plot
isds1 %>% 
  filter(!is.na(location_name)) %>%
  filter(year %in% 2015:2021) %>%
  filter(specialty_description == "Palliative Medicine") %>%
  ggplot(data = .) +
  geom_point(aes(x = Quarter, y = PercentageOccupancy, group = location_name, colour = location_name), position = position_dodge()) +
  geom_line(aes(x = Quarter, y = PercentageOccupancy, group = location_name, colour = location_name)) +
  facet_wrap(health_board ~ .) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45))

## total beds occupied
isds1 %>% 
  filter(!is.na(location_name)) %>%
  filter(year %in% 2015:2021) %>%
  filter(specialty_description == "Palliative Medicine") %>%
  ggplot(data = .) +
  geom_point(aes(x = Quarter, y = TotalOccupiedBeds, group = location_name, colour = location_name), position = position_dodge()) +
  geom_line(aes(x = Quarter, y = TotalOccupiedBeds, group = location_name, colour = location_name)) +
  facet_wrap(health_board ~ .) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45))

## available beds (all beds - total occ.)
isds1 %>% 
  filter(!is.na(location_name)) %>%
  filter(year %in% 2015:2021) %>%
  filter(specialty_description == "Palliative Medicine") %>%
  ggplot(data = .) +
  geom_point(aes(x = Quarter, y = AvailableBeds, group = location_name, colour = location_name), position = position_dodge()) +
  geom_line(aes(x = Quarter, y = AvailableBeds, group = location_name, colour = location_name)) +
  facet_wrap(health_board ~ .) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45))
```

There seems to be quite a bit of year to year variation in hospital occupancy in palliative medicine specialty entries, so it's not clear that it changed with Covid (i.e. Q2 of 2020 and later)
It might still help explain the variation of home deaths during covid, when looked at regionally

# Print session info

```{r}
sessionInfo()
```

### 2021-08-02 covid_nrs_place_of_death_data.Rmd
scotgov open data has been fixed, so I've commented out the fix I was using to correct their erroneous data

--- 2021-05-18 covid_nrs_place_of_death_data.Rmd
@@
   summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
+```
+
+#### Fixing overall weekly deaths
+
+As of 28 July 2021, the [ScotGov open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19) reports the non-covid deaths as if they were all-cause deaths, and this messed up my entire pipeline!
+Below is a quick fix, where instead of using the open data API, I construct the 'overall' weekly deaths data, `weekly_deaths_overall`, from the NRS-published ad-hoc query data on deaths by sex.
+
+```{r}
+## quick fix whilst the open data for overall deaths doesn't work...
+# TODO: remove this chunk once no longer needed
+
+# weekly_deaths_overall_averages_and_other_bits <-  # keep the nrs-computed past averages from the broken data, as well as other columns needed
+#   weekly_deaths_overall %>%
+#   select(-c(deaths_covid_related, deaths_non_covid, deaths_all_causes))
+#   # select(age, sex, place_of_death, week_number_run_over, deaths_nrs_past_average_all_causes, ref_area, ref_period)
+# 
+# weekly_deaths_overall <-  # this rewrites weekly_deaths_overall!
+#   weekly_deaths_sex %>%
+#   group_by_all() %>%
+#   ungroup(sex, starts_with("deaths_")) %>%
+#   summarise(across(starts_with("deaths_"), .fns = sum), .groups = "drop") %>%  # adds together figures between F & M
+#   ({function (x) bind_rows(  # this takes the data at this point and a copy of it, and then summarises the copy across place of death, creating an 'All' category, the figures for which are sums across place of death
+#     x,
+#     x %>% group_by_all() %>% ungroup(place_of_death, starts_with("deaths_")) %>% summarise(across(starts_with("deaths_"), .fns = sum), .groups = "drop") %>% mutate(place_of_death = "All")
+#   )}) %>%
+#   mutate(sex = "all", age = "all") %>%  # add these so that I don't end up breaking any further analysis
+#   arrange(week_number_run_over, place_of_death) %>%
+#   left_join(
+#     weekly_deaths_overall_averages_and_other_bits,
+#     by = c("week_number", "week_number_run_over", "year", "date_w_c", "place_of_death", "sex", "age")
+#     )  # add the average numbers in also so I don't break any further analysis
 ```

### 2021-08-02 exploring_isds1_data.Rmd
saved one of the graphs for hospital beds data

@@
 
+An alternative location for downloading [hospital bed data is here](https://publichealthscotland.scot/media/7634/beds-by-nhs-board-of-treatment-and-specialty-to-december-2020.csv).
+This appears to be somewhat more neatly formatted and ready for analysis.
+
 ```{r}
@@
 ```{r}
+if (!dir.exists("./isd(s)1_exploration")) dir.create("./isd(s)1_exploration")
+
 isds1 %>% count(health_board)
@@
 ## occupancy plot
+(occupancy_2020_palliative_specialty_hospitals <-
   isds1 %>% 
@@
   ggplot(data = .) +
-  geom_point(aes(x = Quarter, y = PercentageOccupancy, group = location_name, colour = location_name), position = position_dodge()) +
-  geom_line(aes(x = Quarter, y = PercentageOccupancy, group = location_name, colour = location_name)) +
+  geom_point(aes(x = Quarter, y = PercentageOccupancy, group = location_name, colour = location_name), position = position_dodge(width = 0.5)) +
+  geom_line(aes(x = Quarter, y = PercentageOccupancy, group = location_name, colour = location_name), position = position_dodge(width = 0.5)) +
+  annotate(geom = "rect", xmin = "2020Q2", xmax = "2020Q4", ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.4) +
   facet_wrap(health_board ~ .) +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45))
+  theme(axis.text.x = element_text(angle = 80, vjust = 0))
+  )
+occupancy_2020_palliative_specialty_hospitals %>%
+  ggsave(plot = ., filename = "./isd(s)1_exploration/occupancy_2020_palliative_specialty_hospitals.png", width = 15, height = 6, dpi = 300)
 
@@
   ggplot(data = .) +
-  geom_point(aes(x = Quarter, y = TotalOccupiedBeds, group = location_name, colour = location_name), position = position_dodge()) +
-  geom_line(aes(x = Quarter, y = TotalOccupiedBeds, group = location_name, colour = location_name)) +
+  geom_point(aes(x = Quarter, y = TotalOccupiedBeds, group = location_name, colour = location_name), position = position_dodge(width = 0.5)) +
+  geom_line(aes(x = Quarter, y = TotalOccupiedBeds, group = location_name, colour = location_name), position = position_dodge(width = 0.5)) +
+  annotate(geom = "rect", xmin = "2020Q2", xmax = "2020Q4", ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.4) +
   facet_wrap(health_board ~ .) +
@@
   ggplot(data = .) +
-  geom_point(aes(x = Quarter, y = AvailableBeds, group = location_name, colour = location_name), position = position_dodge()) +
-  geom_line(aes(x = Quarter, y = AvailableBeds, group = location_name, colour = location_name)) +
+  geom_point(aes(x = Quarter, y = AvailableBeds, group = location_name, colour = location_name), position = position_dodge(width = 0.5)) +
+  geom_line(aes(x = Quarter, y = AvailableBeds, group = location_name, colour = location_name), position = position_dodge(width = 0.5)) +
+  annotate(geom = "rect", xmin = "2020Q2", xmax = "2020Q4", ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.4) +
   facet_wrap(health_board ~ .) +

### 2021-08-04 mocking_up_distance_calculations.Rmd
Started computing distance to hospital computations, using data zones as geographical unit (and computing euclidean distance between OS grid reference coordinates)

---
title: "Mocking up distance/time to travel calculations for hospitals in Scotland"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
library(curl)  # for downloading stuff
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(tictoc)  # for measuring time taken for calculations
```

# Introduction

# Download files

The hospitals data needs to be downloaded, whereas the centroids file was provided by Dr Laurie Berry!

```{r}
url_hospitals <- "https://www.opendata.nhs.scot/dataset/cbd1802e-0e04-4282-88eb-d7bdcfb120f0/resource/c698f450-eeed-41a0-88f7-c1e40a568acc/download/current-hospital_flagged20210506.csv"

if (!file.exists("./distance_to_hospital/current-hospital_flagged20210506.csv")) download.file(url = url_hospitals, destfile = "./distance_to_hospital/current-hospital_flagged20210506.csv")

# 
# url_datazones <- "https://maps.gov.scot/ATOM/shapefiles/SG_DataZoneCent_2011.zip"
# 
# if (!file.exists("./distance_to_hospital/SG_DataZoneCent_2011.zip")) download.file(url = url_datazones, destfile = "./distance_to_hospital/SG_DataZoneCent_2011.zip")
```

# Load data

```{r}
hospitals <- read_csv(file = "./distance_to_hospital/current-hospital_flagged20210506.csv")

datazones <- read_excel("./distance_to_hospital/datazone2011 centroids.xlsx", sheet = 1)

pythagorean_distance <- function(x1, y1, x2, y2) {
  distance <- sqrt((x1 - x2)^2 + (y1 - y2)^2)
}

## check that the XCoordinate and YCoordinate and Easting and Northing are the same thing, by checking the DataZone centroid location against hospital location
hospitals %>%
  select(LocationName, DataZone, XCoordinate, YCoordinate, BasedOnPostcode) %>%
  left_join(datazones, by = c("DataZone")) %>%
  mutate(distance = pythagorean_distance(XCoordinate, YCoordinate, Easting, Northing)) %>%
  arrange(desc(distance))

isds1 <- read_csv("./distance_to_hospital/beds_by_nhs_board_of_treatment_and_specialty.csv", guess=1e5)
```

## Defining acute hospitals

From the [Hospital care website of ISD Scotland](https://www.isdscotland.org/health-topics/hospital-care/):

> This section of our website brings together information on different aspects of acute hospital care, sourced from hospital administrative systems across Scotland. Note that 'Acute' hospital care excludes obstetric, psychiatric and long stay care services that are covered elsewhere on the ISD website. For reference, a listing of all Scottish hospitals (acute and non-acute) can be found by following the Hospitals link on the menu to the left of this page.

Therefore, we can use an ISD(S)1 publication, in this case on Bed Occupancy (https://www.isdscotland.org/Health-Topics/Hospital-Care/Publications/2019-11-26/Acute-Hospital-Publication/data-files/docs/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv), to define Acute hospitals - all hospitals included in this publication are considered acute!

```{r}
acute_hospitals <- hospitals %>% semi_join(isds1, by = "Location")  # keep only hospitals present in the isd(s)1 dataset
```

This reduces the number of hospitals from `r n_distinct(hospitals$LocationName)` to `r n_distinct(acute_hospitals$LocationName)`!

# Compute straight-line distance for each datazone to nearest hospital

For this purpose, I've removed some specialist units that operate within a larger hospital department and are listed at the same geographical grid reference! Otherwise we get two units that are equidistant for some data zones.

```{r}
tic()
smallest_distance_straight_line_datazones <-
  crossing(
    datazones %>% select(DataZone), 
    hospitals %>% select(LocationName) %>% filter(!LocationName %in% c("Orchard View","William Fraser Centre", "Gartnavel General Hospital (Admin Purposes)"))  # remove units located in same place as another hospital
    ) %>%
  left_join(datazones %>% select(DataZone, Easting, Northing), by = "DataZone") %>%
  left_join(hospitals %>% select(LocationName, XCoordinate, YCoordinate), by = "LocationName") %>%
  mutate(distance = pythagorean_distance(x1 = Easting, x2 = XCoordinate, y1 = Northing, y2 = YCoordinate)) %>%
  group_by(DataZone) %>%
  slice_min(order_by = distance, n = 1) %>%
  ungroup
toc(log = TRUE)
```

## Acute hospital distances

```{r}
tic()
smallest_distance_straight_line_datazones_acute_hospitals <-
  crossing(
    datazones %>% select(DataZone), 
    acute_hospitals %>% select(LocationName)
    ) %>%
  left_join(datazones %>% select(DataZone, Easting, Northing), by = "DataZone") %>%
  left_join(hospitals %>% select(LocationName, XCoordinate, YCoordinate), by = "LocationName") %>%
  mutate(distance = pythagorean_distance(x1 = Easting, x2 = XCoordinate, y1 = Northing, y2 = YCoordinate)) %>%
  group_by(DataZone) %>%
  slice_min(order_by = distance, n = 1) %>%
  ungroup
toc(log = TRUE)
```

# Print session info

```{r}
sessionInfo()
```

### 2021-08-04 distance_to_nearest_hospital.Rmd
Expanded the distance to nearest hospital calculations & renamed file

---
title: "Distance/time to travel calculations for hospitals in Scotland"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
library(curl)  # for downloading stuff
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(tictoc)  # for measuring time taken for calculations
```

# Introduction

# Download files & load

The following data needs downloading:

* Hospitals with geographic coordinates
* Datazone centroids with geographic coordinates (provided by Dr Laurie Berrie!)
* Postcodes with geographic coordinates

The coordinates are given in Easting & Northing coordinates in the OS National Grid basis. We can compute the straight-line distance between them using the Pythagorean identity function:

```{r}
pythagorean_distance <- function(x1, y1, x2, y2) {
  distance <- sqrt((x1 - x2)^2 + (y1 - y2)^2)
}
```

## Hospitals

```{r}
url_hospitals <- "https://www.opendata.nhs.scot/dataset/cbd1802e-0e04-4282-88eb-d7bdcfb120f0/resource/c698f450-eeed-41a0-88f7-c1e40a568acc/download/current-hospital_flagged20210506.csv"

if (!file.exists("./distance_to_hospital/current-hospital_flagged20210506.csv")) download.file(url = url_hospitals, destfile = "./distance_to_hospital/current-hospital_flagged20210506.csv")
hospitals <- read_csv(file = "./distance_to_hospital/current-hospital_flagged20210506.csv")

## check that the XCoordinate and YCoordinate and Easting and Northing are the same thing, by checking the DataZone centroid location against hospital location
hospitals %>%
  select(LocationName, DataZone, XCoordinate, YCoordinate, BasedOnPostcode) %>%
  left_join(datazones, by = c("DataZone")) %>%
  mutate(distance = pythagorean_distance(XCoordinate, YCoordinate, Easting, Northing)) %>%
  arrange(desc(distance))
```

### Defining acute hospitals

From the [Hospital care website of ISD Scotland](https://www.isdscotland.org/health-topics/hospital-care/):

> This section of our website brings together information on different aspects of acute hospital care, sourced from hospital administrative systems across Scotland. Note that 'Acute' hospital care excludes obstetric, psychiatric and long stay care services that are covered elsewhere on the ISD website. For reference, a listing of all Scottish hospitals (acute and non-acute) can be found by following the Hospitals link on the menu to the left of this page.

Therefore, we can use an ISD(S)1 publication, in this case on Bed Occupancy (https://www.isdscotland.org/Health-Topics/Hospital-Care/Publications/2019-11-26/Acute-Hospital-Publication/data-files/docs/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv), to define Acute hospitals - all hospitals included in this publication are considered acute!

```{r}
isds1 <- read_csv("./distance_to_hospital/beds_by_nhs_board_of_treatment_and_specialty.csv", guess=1e5)

acute_hospitals <- hospitals %>% semi_join(isds1, by = "Location")  # keep only hospitals present in the isd(s)1 dataset
```

This reduces the number of hospitals from `r n_distinct(hospitals$LocationName)` to `r n_distinct(acute_hospitals$LocationName)`!

## DataZones

```{r}
# url_datazones <- "https://maps.gov.scot/ATOM/shapefiles/SG_DataZoneCent_2011.zip"
# 
# if (!file.exists("./distance_to_hospital/SG_DataZoneCent_2011.zip")) download.file(url = url_datazones, destfile = "./distance_to_hospital/SG_DataZoneCent_2011.zip")
datazones <- read_excel("./distance_to_hospital/datazone2011 centroids.xlsx", sheet = 1)
```

## Postcode directory

~~I'm using the August 2020 postcode directory [available here](https://geoportal.statistics.gov.uk/datasets/ons::ons-postcode-directory-august-2020/about)

I've opted for the `.csv` version and limited the columns to read because it is over 1GB in size!~~

The 2021-1 version of the Scottish postcode directory is [available here](https://www.nrscotland.gov.uk/statistics-and-data/geography/our-products/scottish-postcode-directory/2021-1)

There are a few postcodes with duplicate entries for the same postcode where the geographical location is slightly different between them. I've opted to keep the last entry in a series of duplicates, as it probably reflects the most recent information available.

```{r}
# postcodes <- read_csv(file = "./distance_to_hospital/ONSPD_AUG_2020_UK.csv", col_names = c("postcode_PCDS","easting","northing"), col_types = "__c________ii_____________________________________", skip = 1)  # this format ensures we only read the postcode and the easting & northing data

# paste0(rep("_", times = 50), collapse="")  # used for precisely specifying number of skipped columns!

postcodes <- read_csv(file = "./distance_to_hospital/SPD_PostcodeIndex_Cut_21_1_CSV/SmallUser.csv", col_names = c("postcode","easting","northing"), col_types = "c____ii__________________________________________________", skip = 1)  # this format ensures we only read the postcode and the easting & northing data

postcodes <-  # keep last entry where there is more than one record per postcode
  postcodes %>%
  group_by(postcode) %>%
  slice(n()) %>%
  ungroup
```

# Compute straight-line distance for each datazone to nearest hospital

For this purpose, I've removed some specialist units that operate within a larger hospital department and are listed at the same geographical grid reference! Otherwise we get two units that are equidistant for some data zones.

```{r}
tic()
smallest_distance_straight_line_datazones <-
  crossing(
    datazones %>% select(DataZone), 
    hospitals %>% select(LocationName) %>% filter(!LocationName %in% c("Orchard View","William Fraser Centre", "Gartnavel General Hospital (Admin Purposes)"))  # remove units located in same place as another hospital
    ) %>%
  left_join(datazones %>% select(DataZone, Easting, Northing), by = "DataZone") %>%
  left_join(hospitals %>% select(LocationName, XCoordinate, YCoordinate), by = "LocationName") %>%
  mutate(distance = pythagorean_distance(x1 = Easting, x2 = XCoordinate, y1 = Northing, y2 = YCoordinate)) %>%
  group_by(DataZone) %>%
  slice_min(order_by = distance, n = 1) %>%
  ungroup %>%
  select(DataZone, LocationName, distance)
toc(log = TRUE)
```

## Acute hospital distances

```{r}
tic()
smallest_distance_straight_line_datazones_acute_hospitals <-
  crossing(
    datazones %>% select(DataZone), 
    acute_hospitals %>% select(LocationName)
    ) %>%
  left_join(datazones %>% select(DataZone, Easting, Northing), by = "DataZone") %>%
  left_join(hospitals %>% select(LocationName, XCoordinate, YCoordinate), by = "LocationName") %>%
  mutate(distance = pythagorean_distance(x1 = Easting, x2 = XCoordinate, y1 = Northing, y2 = YCoordinate)) %>%
  group_by(DataZone) %>%
  slice_min(order_by = distance, n = 1) %>%
  ungroup %>%
  select(DataZone, LocationName, distance)
toc(log = TRUE)
```

# Postcode straight-line distance to acute hospitals

```{r}
tic()
smallest_distance_straight_line_postcodes_acute_hospitals <-
  crossing(
    postcodes %>% select(postcode), 
    acute_hospitals %>% select(LocationName)
    ) %>%
  left_join(postcodes %>% select(postcode, easting, northing), by = "postcode") %>%
  left_join(hospitals %>% select(LocationName, XCoordinate, YCoordinate), by = "LocationName") %>%
  mutate(distance = pythagorean_distance(x1 = easting, x2 = XCoordinate, y1 = northing, y2 = YCoordinate)) %>%
  group_by(postcode) %>%
  slice_min(order_by = distance, n = 1) %>%
  ungroup %>%
  select(postcode, LocationName, distance) %>%
  distinct
toc(log = TRUE)

smallest_distance_straight_line_postcodes_acute_hospitals %>%
  write_csv(file = "./distance_to_hospital/postcode_distance_to_nearest_acute_hospital_straightline.csv")
```

# Print session info

```{r}
sessionInfo()
```

### 2021-08-04 distance_to_nearest_hospital.Rmd
reorganise the file and wrote up an introduction

@@
 library(tidyverse)  # for tidy workflow
-library(opendatascot)  # importing data from ScotGov open data website
-library(phsmethods)  # methods for working with PHS data
 library(readxl)  # for reading excel files
-library(curl)  # for downloading stuff
-library(lubridate)  # dealing with dates
-library(janitor)  # for cleaning column names
-library(ISOweek)  # for computing date from ISO week number + year
-library(sf)  # for mapping
-library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
-library(patchwork)  # for assembling plots
-library(extrafont)  # for working with fonts
-library(openxlsx)  # for creating xlsx files
-library(knitr)  # for displaying tables in rmd
 library(tictoc)  # for measuring time taken for calculations
@@
 # Introduction
+
+For investigating increased numbers of deaths at home & other non-institutional settings in Scotland, we'll take into account the distance/travel time to the nearest hospital.
+For this purpose, we need to compute the distance to nearest hospital.
+
+There are two basic approaches: straight-line distance, and road distance / road travel time:
+
+## Straight-line distance
+
+This can be computed using OS National Grid Reference values (Easting & Northing), using the locations of hospitals, and either:
+
+* data zones
+* postcodes
+
+Postcodes are more precise, but also more privacy-invasive when working with administrative records.
+For this reason, we would ideally provide the computed distances at data linkage rather than after data delivery.
+
+Straight-line distances don't necessarily represent a realistic measure of distance or travel time to hospital, however! The main benefit is its ease of calculation.
+
+## Road distance / travel time
+
+This is more realistic & representative than straight-line distance, but more difficult to obtain, since road distances and road spee (used in calculating travel time) aren't openly available like the OS grid reference data are.
 
@@
 if (!file.exists("./distance_to_hospital/current-hospital_flagged20210506.csv")) download.file(url = url_hospitals, destfile = "./distance_to_hospital/current-hospital_flagged20210506.csv")
+
 hospitals <- read_csv(file = "./distance_to_hospital/current-hospital_flagged20210506.csv")
@@
 
-# Compute straight-line distance for each datazone to nearest hospital
+# DataZone straight-line distance to nearest hospital
 
-For this purpose, I've removed some specialist units that operate within a larger hospital department and are listed at the same geographical grid reference!
+For this purpose, I've removed some duplicate entries in the list of hospitals - I think they are specialist units that operate within a larger hospital and are listed at the same geographical grid reference!
 Otherwise we get two units that are equidistant for some data zones.

### 2021-08-16 covid_nrs_place_of_death_data.Rmd
added proportion of yearly deaths broken down by geography & started plotting visualisation;
also added some spaces to the save_output_file function to make it easier to read

@@
     proportions_of_current_non_covid_deaths_by_place
+  )
+```
+
+## By geography
+
+```{r}
+proportions_of_deaths_by_place_by_geography <-
+  bind_rows(
+    ## historic deaths
+    bind_rows(  # merge HB & LA past deaths, and tag them appropriately
+      past_deaths_hb %>% mutate(geography = "hb"),
+      past_deaths_la %>% mutate(geography = "la")
+    ) %>%
+    group_by(year, place_of_death, ref_area, geography) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%  # compute the total number of yearly deaths by place of death & geography
+    filter(place_of_death != "All" & !is.na(as.integer(year))) %>%  # remove blank/irrelevant entries
+    group_by(year, ref_area, geography) %>%
+    mutate(
+      total_deaths = sum(number_of_deaths)  # compute total annual deaths across place of death, but within each geographical area & year
+    ) %>%
+    ungroup %>%
+    mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))  # compute proportion of deaths in any palce relative to total annual deaths in geographical area
+    ,
+    ## current deaths
+    bind_rows(  # merge HB & LA past deaths, and tag them appropriately
+      total_deaths_hb %>% mutate(geography = "hb"),
+      total_deaths_la %>% mutate(geography = "la")
+    ) %>%
+    select(year, place_of_death, number_of_deaths = deaths_all_causes, geography, ref_area) %>%  # keep only relevant columns & rename deaths due to all causes
+    group_by(year, place_of_death, ref_area, geography) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%  # compute the total number of yearly deaths by place of death & geography
+    filter(place_of_death != "All" & !is.na(as.integer(year))) %>%  # remove blank/irrelevant entries
+    group_by(year, ref_area, geography) %>%
+    mutate(
+      total_deaths = sum(number_of_deaths)  # compute total annual deaths across place of death, but within each geographical area & year
+    ) %>%
+    ungroup %>%
+    mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))  # compute proportion of deaths in any palce relative to total annual deaths in geographical area
   )

### 2021-08-16 covid_nrs_place_of_death_visualisations.Rmd
added proportion of yearly deaths broken down by geography & started plotting visualisation;
also added some spaces to the save_output_file function to make it easier to read

@@
 library(knitr)  # for displaying tables in rmd
+library(lemon)  # for showing axis & axis labels in every facet in ggplot
 ```
@@
 ```{r}
-figure_proportion_home_deaths_over_time <-
+(figure_proportion_home_deaths_over_time <-
   merged_proportions_of_deaths_by_place %>%
@@
   NULL
+)
 
 save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
-
-figure_proportion_home_deaths_over_time
 ```
@@
 
+# Proportions of home deaths by geography
+
+## HB
+
+```{r}
+# (figure_proportion_home_deaths_over_time <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "hb") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point() +
+  geom_line(size = 1) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  # facet_wrap(~ ref_area, scales = "free_x") +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_x_continuous(breaks = 2015:2021) +
+  # scale_y_continuous(breaks = seq(0.1, 0.6, by = 0.1), label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C") +
+  # theme_minimal() +
+  theme(
+    legend.position = "top", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = "Year", y= "Proportion of yearly deaths (%)",
+    colour = "",
+    subtitle = paste0("* ",only_2021_text),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+# )
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
+```
+
 # Save most recent workspace for further analyses

### 2021-08-16 covid_nrs_place_of_death_data.Rmd
added script to download & import population estimates for LA/HB based on 2019 data

@@
 earliest_date <- min(merged_deaths_age$date_w_c)
+```
+
+# Ordering geographical data by population size
+
+The latest population estimates by HB/LA are available [from NRS Scotland here](https://www.opendata.nhs.scot/dataset/population-estimates).
+
+I'm using the estimates for 2020 (based on 2019 data) to compute the relative sizes of health boards!
+
+```{r}
+population_estimate_la_2019 <- read_csv(file = "./population_estimates/ca2019_pop_est_29062021.csv") %>%
+  select(Year, CA, Sex, AllAges) %>%
+  clean_names() %>%
+  make_human_readable_ref_area(ref_area = ca) %>%
+  filter(year==2020) %>%
+  filter(ca != "Scotland") %>%
+  filter(sex == "All") %>%
+  rename(la = ca)
+
+population_estimate_hb_2019 <- read_csv(file = "./population_estimates/hb2019_pop_est_29062021.csv") %>%
+  select(Year, HB, Sex, AllAges) %>%
+  clean_names() %>%
+  make_human_readable_ref_area(ref_area = hb) %>%
+  filter(year==2020) %>%
+  filter(hb != "Scotland") %>%
+  filter(sex == "All")
+```
+
+## Compute order
+
+```{r}
+order_la_by_population <-
+  population_estimate_la_2019 %>%
+  arrange(desc(all_ages)) %>%
+  .$la
+
+order_hb_by_population <-
+  population_estimate_hb_2019 %>%
+  arrange(desc(all_ages)) %>%
+  .$hb
 ```

### 2021-08-16 download_latest_data_and_render_reports.R

@@ -208,6 +208,24 @@ la_shapefile_url <- "https://opendata.arcgis.com/datasets/910f48f3c4b3400aa9eb0a
 # download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")
 # download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
 
+# Download population estimates data --------------------------------------
+
+dir_population_estimates <- "./population_estimates"
+if (!dir.exists(dir_population_estimates)) dir.create(dir_population_estimates)
+
+population_estimates_2019_la_url <- "https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/09ebfefb-33f4-4f6a-8312-2d14e2b02ace/download/ca2019_pop_est_29062021.csv"
+population_estimates_2019_hb_url <- "https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1/download/hb2019_pop_est_29062021.csv"
+
+population_estimates_2019_la_file <- "ca2019_pop_est_29062021.csv"
+population_estimates_2019_hb_file <- "hb2019_pop_est_29062021.csv"
+
+if (!file.exists(file.path(dir_population_estimates, population_estimates_2019_la_file))) {
+  curl_download(url = population_estimates_2019_la_url, destfile = file.path(dir_population_estimates, population_estimates_2019_la_file), quiet = FALSE)
+}
+if (!file.exists(file.path(dir_population_estimates, population_estimates_2019_hb_file))) {
+  curl_download(url = population_estimates_2019_hb_url, destfile = file.path(dir_population_estimates, population_estimates_2019_hb_file), quiet = FALSE)
+}
+
 # Import Calibri font -----------------------------------------------------
 
 if (!any(fonttable()$FullName=="Calibri")) font_import(pattern = "calibri", prompt = FALSE)  # only run first time if Calibri not yet imported

### 2021-08-16 covid_nrs_place_of_death_visualisations.Rmd
added plots of proportion of deaths by place & geography

@@
 ```{r}
-# (figure_proportion_home_deaths_over_time <-
+(figure_proportion_of_deaths_by_place_by_hb <-
   proportions_of_deaths_by_place_by_geography %>%
@@
   mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
   ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
@@
   geom_line(size = 1) +
-  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
-  # facet_wrap(~ ref_area, scales = "free_x") +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
   facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
   scale_x_continuous(breaks = 2015:2021) +
-  # scale_y_continuous(breaks = seq(0.1, 0.6, by = 0.1), label = function(x) scales::percent(x, accuracy = 1)) +
   scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
   scale_colour_viridis_d(option="C") +
-  # theme_minimal() +
   theme(
     legend.position = "top",
+    legend.text = element_text(size = 12),
     plot.caption = element_text(size = 10, colour = "gray60"),
@@
   labs(
-    x = "Year", y= "Proportion of yearly deaths (%)",
-    colour = "",
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    colour = NULL,
     subtitle = paste0("* ",only_2021_text),
-    caption = paste0(c(source_nrs_text),collapse="\n")
+    caption = paste0(c("Health boards sorted by 2020 population",source_nrs_text),collapse="\n")
     ) +
@@
   NULL
-# )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_by_place_by_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_by_place_by_hb, width = 12, height = 10, units = "in", dpi = 300)
+```
+
+## LA
+
+```{r}
+(figure_proportion_of_deaths_by_place_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange hb/la by population size
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point() +
+  geom_line(size = 1) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE, ncol = 4) +
+  scale_x_continuous(breaks = 2015:2021) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C") +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    colour = NULL,
+    subtitle = paste0("* ",only_2021_text),
+    caption = paste0(c("Local Authorities sorted by 2020 population",source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_by_place_by_la"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_by_place_by_la, width = 12, height = 14, units = "in", dpi = 300)
 ```

### 2021-08-17 covid_nrs_place_of_death_visualisations.Rmd
updated 2020 & 2021 on same graph plot, and added plots showing proportion of home deaths by geography compared to scottish average

--- 2021-08-16 covid_nrs_place_of_death_visualisations.Rmd
@@
 
+## 2020 and 2021 on same graph
+
+For these graphs, we use the week number, and superimpose lines for 2020 and 2021!
+
+### Home deaths, both years
+
+```{r}
+(figure_deaths_at_home_2020_and_2021 <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%
+  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
+  mutate(year = factor(year)) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
+               names_to = "deaths_type", values_to = "deaths") %>%
+  ggplot(aes(x = week_number)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
+  geom_line(data = .
+%>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
+  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
+  scale_x_continuous(breaks = seq(0,60,by=12)) +
+  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths",
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
+  annotate(geom = "text", x = 18, y = 550, label = "2020", alpha = 0.4) +
+  annotate(geom = "text", x = 4, y = 530, label = "2021") +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Weekly deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  annotate(geom = "text", x = 10, y = 100, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom = "curve", x = 10, xend = 14, y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x = 20, y = 50, label = "Historic mean", colour = "grey20") +
+  annotate(geom = "curve", x = 20, xend = 25, y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020_and_2021, width = 18, height = 14, dpi = 300, units = "cm")
+```
+
 ## Deaths at home, for policy briefing
@@
 figure_deaths_at_home_by_sex
-```
-
-## 2020 and 2021 on same graph
-
-For these graphs, we use the week number, and superimpose lines for 2020 and 2021!
-
-### Home deaths, both years
-
-```{r}
-(figure_deaths_at_home_2020_and_2021 <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
-  mutate(year = factor(year)) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
-               names_to = "deaths_type", values_to = "deaths") %>%
-  ggplot(aes(x = week_number)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  # geom_line(aes(y = deaths, colour = deaths_type, alpha = year)) +
-  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number), show.legend = FALSE) +
-  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
-  scale_x_continuous(breaks = seq(0,60,by=12)) +
-  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths",
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  scale_alpha_continuous(range = c(0.2, 0.8), guide = NULL) +
-  annotate(geom = "text", x = 18, y = 550, label = "2020", alpha = 0.6) +
-  annotate(geom = "text", x = 4, y = 530, label = "2021") +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Weekly deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  annotate(geom = "text", x = 10, y = 100, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom = "curve", x = 10, xend = 14, y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x = 20, y = 50, label = "Historic mean", colour = "grey20") +
-  annotate(geom = "curve", x = 20, xend = 25, y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  NULL
-)
-
-save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020_and_2021, width = 18, height = 14, dpi = 300, units = "cm")
 ```
@@
 
-## HB
-
-```{r}
-# (figure_proportion_home_deaths_over_time <-
+## all places of death, HB
+
+```{r}
+(figure_proportion_of_deaths_by_place_by_hb <-
   proportions_of_deaths_by_place_by_geography %>%
@@
   mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
   ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
-  geom_point() +
-  geom_line(size = 1) +
-  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
-  # facet_wrap(~ ref_area, scales = "free_x") +
+  geom_point(aes(size = place_of_death)) +
+  geom_line(aes(size = place_of_death)) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
   facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
   scale_x_continuous(breaks = 2015:2021) +
-  # scale_y_continuous(breaks = seq(0.1, 0.6, by = 0.1), label = function(x) scales::percent(x, accuracy = 1)) +
   scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
-  scale_colour_viridis_d(option="C") +
-  # theme_minimal() +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
   theme(
     legend.position = "top",
+    legend.text = element_text(size = 12),
     plot.caption = element_text(size = 10, colour = "gray60"),
@@
   labs(
-    x = "Year", y= "Proportion of yearly deaths (%)",
-    colour = "",
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    subtitle = paste0(c(paste0("* ",only_2021_text),"Health boards sorted by 2020 population estimate."),collapse="\n"),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2, title = NULL), 
+         size = guide_legend(nrow=2, title = NULL)) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_by_place_by_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_by_place_by_hb, width = 12, height = 10, units = "in", dpi = 300)
+```
+
+## all places of death, LA
+
+```{r}
+(figure_proportion_of_deaths_by_place_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange hb/la by population size
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point(aes(size = place_of_death)) +
+  geom_line(aes(size = place_of_death)) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE, ncol = 4) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = 2015:2021) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    subtitle = paste0(c(paste0("* ",only_2021_text),"Local authorities sorted by 2020 population estimate."),collapse="\n"),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2, title = NULL), 
+         size = guide_legend(nrow=2, title = NULL)) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_by_place_by_la"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_by_place_by_la, width = 12, height = 14, units = "in", dpi = 300)
+```
+
+## Home deaths + Scottish average, HB
+
+```{r}
+(figure_proportion_of_deaths_at_home_by_hb <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "hb") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
+  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
+  geom_point() +
+  geom_line() +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = 2015:2021) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
+    linetype = NULL,
+    subtitle = paste0(c(paste0("* ",only_2021_text),"Health boards sorted by 2020 population estimate."),collapse="\n"),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_at_home_by_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_at_home_by_hb, width = 12, height = 10, units = "in", dpi = 300)
+```
+
+## Home deaths + Scottish average, LA
+
+```{r}
+(figure_proportion_of_deaths_at_home_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
+  geom_point() +
+  geom_line() +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE, ncol = 4) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = 2015:2021) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
+    linetype = NULL,
+    subtitle = paste0(c(paste0("* ",only_2021_text),"Local authorities sorted by 2020 population estimate."),collapse="\n"),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_at_home_by_la"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_at_home_by_la, width = 12, height = 14, units = "in", dpi = 300)
+```
+
+## HB: compact plot of home deaths only
+
+```{r}
+
+# (figure_proportion_of_deaths_by_place_by_hb <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "hb") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
+  mutate(year = factor(x = year, levels = 2015:2021)) %>%  # arrange hb/la by population size
+  ggplot(., aes(x=proportion_of_total, y=ref_area, colour=year, shape=year)) +
+  geom_point(size = 2) +
+  scale_shape_manual(values = c(rep(4, times = 5), rep(16, times = 2))) +
+  scale_x_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="D", direction = -1) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    colour = NULL,
     subtitle = paste0("* ",only_2021_text),
-    caption = paste0(c(source_nrs_text),collapse="\n")
-    ) +
-  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+    caption = paste0(c("Health boards sorted by 2020 population estimate",source_nrs_text),collapse="\n")
+    ) +
+  # guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
   NULL
 # )
-
-save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
 ```

### 2021-08-17 covid_nrs_place_of_death_visualisations.Rmd
modified figure_number_of_deaths_by_place_by_year to include 2021 also!

@@
   filter(!place_of_death %in% c("All","Other")) %>%  # don't show Other
-  filter(year != 2021) %>%  # don't include 2021
+  # filter(year != 2021) %>%  # don't include 2021
   mutate(
@@
   geom_text(data = . %>% filter(year==2016) %>% mutate(y_pos = c(3.1e4, 1.1e4, 1.8e4)), aes(y = y_pos, label = place_of_death), hjust = 0) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=y, label="*"), inherit.aes = FALSE) +
   scale_colour_viridis_d(option="C", end = 0.8, guide = FALSE) +
   scale_linetype_manual(values = c("solid", "dashed")) +
+  scale_x_continuous(breaks = 2015:2021) +
   theme(
@@
     x = "Year", y = "Deaths (N)",
+    subtitle = paste0("* ",only_2021_text),
     caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n")

### 2021-08-18 covid_nrs_place_of_death_visualisations.Rmd
added a graph showing absolute number of deaths by place over time ending in 2020

@@
   geom_text(data = . %>% filter(year==2016) %>% mutate(y_pos = c(3.1e4, 1.1e4, 1.8e4)), aes(y = y_pos, label = place_of_death), hjust = 0) +
-  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=y, label="*"), inherit.aes = FALSE) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=number_of_deaths, label="*"), inherit.aes = FALSE) +
   scale_colour_viridis_d(option="C", end = 0.8, guide = FALSE) +
@@
 save_output_file(filename = paste0(dir_outputs,"/figure_number_of_deaths_by_place_by_year"), extensions = c(".pdf",".svg",".png"), plot = figure_number_of_deaths_by_place_by_year, width = 15, height = 11, dpi = 300, units = "cm")
+```
+
+### Only up to incl. 2020
+
+```{r}
+(figure_number_of_deaths_by_place_by_year_2020 <-
+  bind_rows(
+    proportions_of_historical_deaths_by_place %>% filter(year==2019) %>% mutate(covid_status = "Non Covid-19 deaths"),  # add extra rows of 2019 data for plotting line between 2019 and 2020
+    proportions_of_historical_deaths_by_place %>% mutate(covid_status = "All deaths"),
+    proportions_of_current_non_covid_deaths_by_place %>% mutate(covid_status = "Non Covid-19 deaths"),
+    proportions_of_current_deaths_by_place %>% mutate(covid_status = "All deaths")
+  ) %>%
+  filter(!place_of_death %in% c("All","Other")) %>%  # don't show Other
+  filter(year != 2021) %>%  # don't include 2021
+  mutate(
+    place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])   # move home to 2nd position to match order in plot
+    ) %>%
+  ggplot(., aes(x=year, y=number_of_deaths, colour = place_of_death)) +
+  geom_line(data = . %>% filter(covid_status == "All deaths"), aes(group = place_of_death, linetype = covid_status)) +
+  geom_line(data = . %>% filter(covid_status == "Non Covid-19 deaths"), aes(group = place_of_death, linetype = covid_status)) +
+  geom_point(data = . %>% filter(!covid_status=="Non Covid-19 deaths" | !year==2019)) +
+  geom_text(data = . %>% filter(year==2016) %>% mutate(y_pos = c(3.1e4, 1.1e4, 1.8e4)), aes(y = y_pos, label = place_of_death), hjust = 0) +
+  scale_colour_viridis_d(option="C", end = 0.8, guide = FALSE) +
+  scale_linetype_manual(values = c("solid", "dashed")) +
+  scale_x_continuous(breaks = 2015:2021) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    legend.box="vertical",
+    legend.margin=margin()
+    ) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ","), limits = c(0, NA)) +
+  labs(
+    x = "Year", y = "Deaths (N)",
+    caption = paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
+  ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_number_of_deaths_by_place_by_year_2020"), extensions = c(".pdf",".svg",".png"), plot = figure_number_of_deaths_by_place_by_year_2020, width = 15, height = 11, dpi = 300, units = "cm")
 ```

### 2021-08-19 covid_nrs_place_of_death_data.Rmd
slight format change in the weekly deaths spreadsheet caused an error due to mismatch in number of expected and actual columns in 2021 data, so I hacked up a solution

@@
   )
+## in week 32 of 2021, the format changed somewhat and the blank column in penultimate location in Table 3 (2021) was removed
+## so this is a hack that gets around that!
+expected_ncol_2021 <- length(headers_in_nrs_data_2021)
+actual_ncol_2021 <- ncol(readxl::read_excel(files_weekly_deaths$main_report, sheet = nrs_weekly_data_sheet_2021))
+headers_in_nrs_data_2021 <- headers_in_nrs_data_2021[1:actual_ncol_2021]  # subset the first N of the defined headers, where N is actual number of columns in file
 

### 2021-08-23 distance_to_nearest_hospital.Rmd
Tried to test if a different list of acute hospitals from another PHS report on quarterly hospital activity would produce different list  of hospitals... but the data ren't given on hospital level, so I can't produce a list of hospitals from those! Also recorded a caveat about the appropriateness of the 'acute hospital' definition

@@
 library(tictoc)  # for measuring time taken for calculations
+
+# TODO: work out if OSRM would work...
+# library(osrm)  # for OSM routing data
 ```
@@
 
-## Hospitals
-
-```{r}
-url_hospitals <- "https://www.opendata.nhs.scot/dataset/cbd1802e-0e04-4282-88eb-d7bdcfb120f0/resource/c698f450-eeed-41a0-88f7-c1e40a568acc/download/current-hospital_flagged20210506.csv"
-
-if (!file.exists("./distance_to_hospital/current-hospital_flagged20210506.csv")) download.file(url = url_hospitals, destfile = "./distance_to_hospital/current-hospital_flagged20210506.csv")
-
-hospitals <- read_csv(file = "./distance_to_hospital/current-hospital_flagged20210506.csv")
-
-## check that the XCoordinate and YCoordinate and Easting and Northing are the same thing, by checking the DataZone centroid location against hospital location
-hospitals %>%
-  select(LocationName, DataZone, XCoordinate, YCoordinate, BasedOnPostcode) %>%
-  left_join(datazones, by = c("DataZone")) %>%
-  mutate(distance = pythagorean_distance(XCoordinate, YCoordinate, Easting, Northing)) %>%
-  arrange(desc(distance))
-```
-
-### Defining acute hospitals
-
-From the [Hospital care website of ISD Scotland](https://www.isdscotland.org/health-topics/hospital-care/):
-
-> This section of our website brings together information on different aspects of acute hospital care, sourced from hospital administrative systems across Scotland.
-Note that 'Acute' hospital care excludes obstetric, psychiatric and long stay care services that are covered elsewhere on the ISD website.
-For reference, a listing of all Scottish hospitals (acute and non-acute) can be found by following the Hospitals link on the menu to the left of this page.
-
-Therefore, we can use an ISD(S)1 publication, in this case on Bed Occupancy (https://www.isdscotland.org/Health-Topics/Hospital-Care/Publications/2019-11-26/Acute-Hospital-Publication/data-files/docs/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv), to define Acute hospitals - all hospitals included in this publication are considered acute!
-
-```{r}
-isds1 <- read_csv("./distance_to_hospital/beds_by_nhs_board_of_treatment_and_specialty.csv", guess=1e5)
-
-acute_hospitals <- hospitals %>% semi_join(isds1, by = "Location")  # keep only hospitals present in the isd(s)1 dataset
-```
-
-This reduces the number of hospitals from `r n_distinct(hospitals$LocationName)` to `r n_distinct(acute_hospitals$LocationName)`!
-
 ## DataZones
@@
 ```
+
+## Hospitals
+
+```{r}
+url_hospitals <- "https://www.opendata.nhs.scot/dataset/cbd1802e-0e04-4282-88eb-d7bdcfb120f0/resource/c698f450-eeed-41a0-88f7-c1e40a568acc/download/current-hospital_flagged20210506.csv"
+
+if (!file.exists("./distance_to_hospital/current-hospital_flagged20210506.csv")) download.file(url = url_hospitals, destfile = "./distance_to_hospital/current-hospital_flagged20210506.csv")
+
+hospitals <- read_csv(file = "./distance_to_hospital/current-hospital_flagged20210506.csv")
+
+## check that the XCoordinate and YCoordinate and Easting and Northing are the same thing, by checking the DataZone centroid location against hospital location
+hospitals %>%
+  select(LocationName, DataZone, XCoordinate, YCoordinate, BasedOnPostcode) %>%
+  left_join(datazones, by = c("DataZone")) %>%
+  mutate(distance = pythagorean_distance(XCoordinate, YCoordinate, Easting, Northing)) %>%
+  arrange(desc(distance))
+```
+
+### Defining acute hospitals
+
+From the [Hospital care website of ISD Scotland](https://www.isdscotland.org/health-topics/hospital-care/):
+
+> This section of our website brings together information on different aspects of acute hospital care, sourced from hospital administrative systems across Scotland.
+Note that 'Acute' hospital care excludes obstetric, psychiatric and long stay care services that are covered elsewhere on the ISD website.
+For reference, a listing of all Scottish hospitals (acute and non-acute) can be found by following the Hospitals link on the menu to the left of this page.
+
+Therefore, we can use an ISD(S)1 publication, in this case on Bed Occupancy (https://www.isdscotland.org/Health-Topics/Hospital-Care/Publications/2019-11-26/Acute-Hospital-Publication/data-files/docs/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv), to define Acute hospitals - all hospitals included in this publication are considered acute!
+
+Note: when I spoke to Kirsty Anderson (PHS analyst) on 19 August 2021, she suggestd I extract the list of acute hospitals from the Inpatient activity file of the report, so I will check that also.
+[The report is available here](https://publichealthscotland.scot/media/7504/inpatient-and-daycase-stays-by-nhs-board-of-residence-to-december-2019.csv)
+However, these data aren't provided on hospital-level, but on health-board level instead, so a list of hospitals canot be obtained this way!
+
+**Important: Kirsty Anderson of PHS said that the definition of 'acute' is not a singular definition used across the NHS.
+'Acute' was defined for the purposes of their reporting hospital activity, and may not be appropriate in other contexts!**
+
+```{r}
+isds1_beds <- read_csv("./distance_to_hospital/beds_by_nhs_board_of_treatment_and_specialty.csv", guess=1e5)
+
+acute_hospitals <- hospitals %>% semi_join(isds1_beds, by = "Location")  # keep only hospitals present in the isd(s)1 dataset
+
+# isds1_inpatients <- read_csv("./distance_to_hospital/inpatient-and-daycase-stays-by-nhs-board-of-residence-to-december-2019.csv", guess=1e5)  # note: this doesn't include per-hospital data
+```
+
+This reduces the number of hospitals from `r n_distinct(hospitals$LocationName)` to `r n_distinct(acute_hospitals$LocationName)`!
 

### 2021-08-24 covid_nrs_place_of_death_visualisations.Rmd
fixed graph size so the subtitle doesn't get cut off!

@@
 
-save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 12, height = 10, units = "cm", dpi = 300)
+save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 13, height = 10, units = "cm", dpi = 300)
 ```

### 2021-08-24 covid_nrs_place_of_death_visualisations.Rmd
added deaths by place, with both yers on same plot. pretty pleased with how it looks!

@@
 save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020_and_2021, width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+### Deaths by place, both years
+
+```{r}
+annotations_deaths_by_place_2020_and_2021 <- tribble(
+  ~place_of_death, ~x, ~y, ~label,
+  "Hospital", 4, 400, 2021,
+  "Hospital", 18, 800, 2020,
+  "Home & other non-institution", 4, 540, 2021,
+  "Home & other non-institution", 18, 610, 2020,
+  "Care home", 4, 160, 2021,
+  "Care home", 11, 600, 2020
+) %>% mutate(alpha = if_else(label==2021, 1, 0.4)) %>%   # change alpha for 2020
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+
+arrows_deaths_by_place_2020_and_2021 <- tribble(
+   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
+  "text", "Home & other non-institution", "Historic range (min & max)", 15, 90, NA, NA,
+  "curve", "Home & other non-institution", NA, 10, 110, 14, 250,
+  "text", "Home & other non-institution", "Historic mean", 34, 40, NA, NA,
+  "curve", "Home & other non-institution", NA, 34, 60, 39, 275
+) %>% 
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+
+(figure_deaths_by_place_2020_and_2021 <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
+  mutate(year = factor(year)) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
+               names_to = "deaths_type", values_to = "deaths") %>%
+  ggplot(aes(x = week_number)) +
+  # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "grey", alpha = 0.3) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  geom_line(data = .
+%>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.5) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
+  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type), size = 1.1) +
+  scale_x_continuous(breaks = seq(0,60,by=12)) +
+  ylim(0, 850) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths",
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  geom_text(data = annotations_deaths_by_place_2020_and_2021 %>% filter(label==2021), aes(x=x, y=y, label=label), fontface = "bold") +
+  geom_text(data = annotations_deaths_by_place_2020_and_2021 %>% filter(label==2020), aes(x=x, y=y, label=label), alpha = 0.6) +
+  geom_text(data = arrows_deaths_by_place_2020_and_2021 %>% filter(geom=="text"), aes(x=x,y=y,label=label)) +
+  geom_curve(data = arrows_deaths_by_place_2020_and_2021 %>% filter(geom=="curve") %>% slice(1), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), alpha = 0.5) +
+  geom_curve(data = arrows_deaths_by_place_2020_and_2021 %>% filter(geom=="curve") %>% slice(2), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), linetype = "dashed") +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Weekly deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_place_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_place_2020_and_2021, width = 28, height = 14, dpi = 300, units = "cm")
 ```

### 2021-08-24 covid_nrs_place_of_death_data.Rmd
added a distinction between underlying and contributory causes and starting making plot of those

@@
 #     )  # add the average numbers in also so I don't break any further analysis
+```
+
+#### Weekly Covid as contributory vs underlying cause of death
+
+```{r}
+weekly_covid_contributory_vs_underlying_cause <-
+  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
+  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
+  mutate(
+    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
+    week_number = parse_integer(str_sub(week_number, 4, 5)),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+  ) %>%
+  filter(
+    cause_of_death %in% c("COVID-19 contributory factor", "COVID-19 underlying cause")
+  ) %>%
+  tidyr::complete(  # complete implicit missing entries and make them 0
+    nesting(week_number,week_number_run_over,date_w_c,year),
+    age,sex,place_of_death,cause_of_death,
+    fill = list(number_of_deaths = 0)
+  ) %>%
+  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
+  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
+  clean_names() %>%
+  filter(age!="all") %>%
+  filter(sex!="all") %>%
+  filter(place_of_death != "All") %>%
+  group_by(week_number, week_number_run_over, year, date_w_c, place_of_death) %>%  # sum over age & sex
+  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
+  ({function(x)
+    bind_rows(
+      x,
+      x %>% 
+        group_by(week_number, week_number_run_over, year, date_w_c) %>%  # sum across place of death
+        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
+        mutate(place_of_death = "All")
+    )
+  })
+
+## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure
+stopifnot(
+  (weekly_deaths_overall %>%
+  filter(sex=="all" & age=="all") %>%
+  inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
+  filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
+  nrow) == 0
+)
+
+annual_covid_contributory_vs_underlying_cause <-
+  weekly_covid_contributory_vs_underlying_cause %>%
+  group_by(year, place_of_death) %>%
+  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
+  ({function(x)
+    bind_rows(
+      x %>% mutate(year = as.character(year)),
+      x %>% 
+        group_by(place_of_death) %>%  # sum across eyars
+        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
+        mutate(year = "2020-2021")
+    )
+  }) %>%
+  mutate(deaths_covid_related = covid_19_contributory_factor + covid_19_underlying_cause) %>%
+  mutate(
+    proportion_contributory = covid_19_contributory_factor / deaths_covid_related,
+    proportion_underlying = covid_19_underlying_cause / deaths_covid_related
+    )
 ```

### 2021-08-24 covid_nrs_place_of_death_visualisations.Rmd
added a distinction between underlying and contributory causes and starting making plot of those

--- 2021-08-24 covid_nrs_place_of_death_visualisations.Rmd
@@
 save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_at_home_2020_and_2021, width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+### Deaths by place, both years
+
+```{r}
+annotations_deaths_by_place_2020_and_2021 <- tribble(
+  ~place_of_death, ~x, ~y, ~label,
+  "Hospital", 4, 400, 2021,
+  "Hospital", 18, 800, 2020,
+  "Home & other non-institution", 4, 540, 2021,
+  "Home & other non-institution", 18, 610, 2020,
+  "Care home", 4, 160, 2021,
+  "Care home", 11, 600, 2020
+) %>% mutate(alpha = if_else(label==2021, 1, 0.4)) %>%   # change alpha for 2020
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+
+arrows_deaths_by_place_2020_and_2021 <- tribble(
+   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
+  "text", "Home & other non-institution", "Historic range (min & max)", 15, 90, NA, NA,
+  "curve", "Home & other non-institution", NA, 10, 110, 14, 250,
+  "text", "Home & other non-institution", "Historic mean", 34, 40, NA, NA,
+  "curve", "Home & other non-institution", NA, 34, 60, 39, 275
+) %>% 
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+
+(figure_deaths_by_place_2020_and_2021 <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
+  mutate(year = factor(year)) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
+               names_to = "deaths_type", values_to = "deaths") %>%
+  ggplot(aes(x = week_number)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "grey", alpha = 0.3) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  geom_line(data = .
+%>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.5) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
+  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type), size = 1.1) +
+  scale_x_continuous(breaks = seq(0,60,by=12)) +
+  ylim(0, 850) +  # start scale at 0 for better understanding the scale of deaths
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths",
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  geom_text(data = annotations_deaths_by_place_2020_and_2021 %>% filter(label==2021), aes(x=x, y=y, label=label), fontface = "bold") +
+  geom_text(data = annotations_deaths_by_place_2020_and_2021 %>% filter(label==2020), aes(x=x, y=y, label=label), alpha = 0.6) +
+  geom_text(data = arrows_deaths_by_place_2020_and_2021 %>% filter(geom=="text"), aes(x=x,y=y,label=label)) +
+  geom_curve(data = arrows_deaths_by_place_2020_and_2021 %>% filter(geom=="curve") %>% slice(1), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), alpha = 0.5) +
+  geom_curve(data = arrows_deaths_by_place_2020_and_2021 %>% filter(geom=="curve") %>% slice(2), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), linetype = "dashed") +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(x = "Week number",
+       y = "Weekly deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_deaths_by_place_2020_and_2021"), extensions = c(".pdf",".svg",".png"), plot = figure_deaths_by_place_2020_and_2021, width = 28, height = 14, dpi = 300, units = "cm")
 ```
@@
 
+# Covid as underlying vs contributory cause
+
+```{r}
+(figure_covid_contributory_vs_underlying_cause <-
+annual_covid_contributory_vs_underlying_cause %>%
+  pivot_longer(cols = -c(year, place_of_death, deaths_covid_related), names_to = "covid_category", values_to = "value") %>%
+  extract(col = covid_category, into = c("measure", "covid_category"), regex = "(covid_19|proportion)_(.*)", remove = TRUE) %>%
+  mutate(
+    measure = if_else(measure == "covid_19", "n", measure),
+    covid_category = str_remove(covid_category, pattern = "\\_.*$"),
+    covid_category = paste0(covid_category, " cause")
+  ) %>%
+  ggplot(aes(x = place_of_death, y = value, group = covid_category, fill = covid_category)) +
+    facet_wrap(year~measure,scales = "free", nrow = 3) +
+    geom_col(position = position_dodge(width = 1))
+)
+```
+
 # Save most recent workspace for further analyses

### 2021-08-24 covid_nrs_place_of_death_visualisations.Rmd
made a plot of proportions of covid by whether underlying or contributory

@@
     covid_category = str_remove(covid_category, pattern = "\\_.*$"),
-    covid_category = paste0(covid_category, " cause")
-  ) %>%
+    covid_category = paste0(covid_category, " cause"),
+    year = str_replace(year, pattern = "2021", replacement = "2021*"),
+    year = factor(year, levels = c("2020", "2021*", "2020-2021*")),
+    place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))
+  ) %>%
+  filter(measure =="proportion") %>%
+  filter(place_of_death != "Other") %>%
   ggplot(aes(x = place_of_death, y = value, group = covid_category, fill = covid_category)) +
-    facet_wrap(year~measure,scales = "free", nrow = 3) +
-    geom_col(position = position_dodge(width = 1))
-)
+    facet_grid(~year) +
+    geom_col(position = position_stack()) +
+    # geom_col(position = position_dodge(width = 1)) +
+    scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
+    scale_fill_viridis_d(labels = c("Contributory", "Underlying")) +
+    # geom_text(data = .
+%>% filter(covid_category == "contributory cause"), inherit.aes = TRUE, aes(label = paste0("Contributory cause: ", scales::percent(value, accuracy = 0.1))), x = 0.7, nudge_y = -0.2, hjust = 1) +
+    # geom_text(data = .
+%>% filter(covid_category == "underlying cause"), inherit.aes = TRUE, aes(label = paste0("Underlying cause: ", scales::percent(value, accuracy = 0.1))), x = 0.5, nudge_y = 0.2, hjust = 1) +
+    geom_label(data = . %>% filter(covid_category == "contributory cause"), inherit.aes = TRUE, aes(label = scales::percent(value, accuracy = 0.1), y = 1.02-value), fill = "white",  hjust = 0.5) +
+    geom_label(data = . %>% filter(covid_category == "underlying cause"), inherit.aes = TRUE, aes(label = scales::percent(value, accuracy = 0.1), y = value-0.2), fill = "white", hjust = 0.5) +
+    labs(
+      x = NULL, y = NULL,
+      subtitle = paste0(c(paste0("* ",only_2021_text)),collapse="\n"),
+      caption = paste0(c(source_nrs_text),collapse="\n")
+      ) +
+    guides(fill = guide_legend(nrow=1, title = NULL)) +
+    theme(
+      legend.position = "top",
+      legend.text = element_text(size = 10),
+      plot.caption = element_text(size = 10, colour = "gray60"),
+      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
+      ) +
+    NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_covid_contributory_vs_underlying_cause"), extensions = c(".pdf",".svg",".png"), plot = figure_covid_contributory_vs_underlying_cause, width = 6, height = 6, units = "in", dpi = 300)
 ```

### 2021-09-15 covid_nrs_place_of_death_data.Rmd
cleaning up commented out bits of code

@@
 
-## Load quarterly bed occupancy data
-
-```{r}
-# TODO: download automatically in other script
-
-# beds_raw <- 
-#   read_csv("./beds_by_nhs_board_of_treatment_and_specialty.csv", guess_max = 3e4) %>%
-#   mutate(ref_area = match_area(HB)) %>%  # clean health board names
-#   clean_names() %>%
-#   # select(-percentage_occupancy_qf) %>%
-#   extract(col = quarter, into = c("year", "quarter"), regex = "(\\d{4})Q(\\d)", remove = TRUE) %>%
-#   # mutate(across(.cols = matches("percent"), .fns = ~{.x / 100}))
-#   mutate(percentage_occupancy = percentage_occupancy / 100)
-# 
-# beds_raw %>%
-#   filter(ref_area == "Scotland") %>% count(specialty_name)
-# 
-# beds_raw %>%
-#   filter(ref_area == "Scotland" & specialty_name == "Palliative Medicine") %>%
-#   ggplot() +
-#   aes(x = quarter, y = percentage_occupancy, colour = year, size = year) +
-#   scale_y_continuous(labels = function(x) scales::percent(x), limits = c(0,NA)) +
-#   # geom_point() +
-#   geom_line(aes(group = year)) +
-#   scale_colour_viridis_d() +
-#   scale_size_manual(values = c(rep(0.5, times = 5),1)) +
-#   NULL
-
-```
-
 # Calculate case counts: weekly counts to match weekly death reporting
@@
     )
-```
-
-# Comparing geographic areas by case counts
-
-We can use the yearly cumulative case count to group geographic areas by how many cases they've had by the end of 2020.
-
-NOTE: I've commented this out - I ended up not using it. TODO: remove in a future iteration!
-
-```{r}
-# cumulative_case_2020_based_grouping_la <-
-#   weekly_covid_cases_la %>% 
-#   filter(ref_area!="Scotland") %>%
-#   filter(year == 2020) %>% 
-#   filter(week_number == max(week_number)) %>%  # keep last week
-#   select(ref_area, cumulative_positive, crude_rate_positive_per100k)
-# 
-# cumulative_case_2020_based_grouping_hb <-
-#   weekly_covid_cases_hb %>% filter(year == 2020) %>% 
-#   filter(ref_area!="Scotland") %>%
-#   filter(year == 2020) %>% 
-#   filter(week_number == max(week_number)) %>%  # keep last week 
-#   select(ref_area, cumulative_positive, crude_rate_positive_per100k)
-# 
-# k_means_clustering_cases_and_per100k_2020_hb <- map(
-#   .x = 2:4,
-#   .f = ~kmeans(x=cumulative_case_2020_based_grouping_hb %>% select(cumulative_positive, crude_rate_positive_per100k), centers = .x, nstart = 1e3)
-# ) %>% set_names(nm=paste0("k",2:4))
-# 
-# clusters_hb <-
-#   bind_cols(
-#     cumulative_case_2020_based_grouping_hb %>% select(ref_area),
-#     map_dfc(.x = k_means_clustering_cases_and_per100k_2020_hb, .f = function(x) return(x$cluster))
-#   )
-# 
-# cumulative_case_2020_based_grouping_hb <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   left_join(clusters_hb, by="ref_area") %>%
-#   mutate(
-#     quartiles_cases = ntile(cumulative_positive, n=4),
-#     quartiles_per100k = ntile(crude_rate_positive_per100k, n=4)
-#   )
-# 
-# # TODO: apply to LA also!
-# 
-# 
-# hb_cases_plots <- list()
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
-#   geom_point() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases",
-#     y="COVID cases per 100,000 population",
-#     title="Cases vs. population rate"
-#   ) +
-#   NULL
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k_log10 <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
-#   geom_point() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases (log)",
-#     y="COVID cases per 100,000 population (log)",
-#     title="Cases vs. population rate (log scale axes)"
-#   ) +
-#   scale_x_log10() +
-#   scale_y_log10() +
-#   NULL
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_cases))) +
-#   geom_point(size=3, aes(colour=factor(quartiles_cases))) +
-#   scale_colour_viridis_d() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases",
-#     y="COVID cases per 100,000 population",
-#     title="Cases vs. population rate"
-#   ) +
-#   NULL
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_per100k))) +
-#   geom_point(size=3, aes(colour=factor(quartiles_cases))) +
-#   scale_colour_viridis_d() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases",
-#     y="COVID cases per 100,000 population",
-#     title="Cases vs. population rate",
-#     colour = "Quartile (Cases per 100,000)"
-#   ) +
-#   NULL
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k_k2 <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k2))) +
-#   geom_point(size=3, aes(colour=factor(k2))) +
-#   scale_colour_viridis_d() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases",
-#     y="COVID cases per 100,000 population",
-#     title="Cases vs. population rate"
-#   ) +
-#   NULL
-# 
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k_k3 <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k3))) +
-#   geom_point(size=3, aes(colour=factor(k3))) +
-#   scale_colour_viridis_d() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases",
-#     y="COVID cases per 100,000 population",
-#     title="Cases vs. population rate"
-#   ) +
-#   NULL
-# 
-# 
-# hb_cases_plots$cumulative_cases_vs_per100k_k4 <-
-#   cumulative_case_2020_based_grouping_hb %>%
-#   ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k4))) +
-#   geom_point(size=3, aes(colour=factor(k4))) +
-#   scale_colour_viridis_d() +
-#   geom_label_repel() +
-#   # theme_minimal() +
-#   labs(
-#     x="COVID cases",
-#     y="COVID cases per 100,000 population",
-#     title="Cases vs. population rate"
-#   ) +
-#   NULL
-# 
-# 
-# hb_cases_plots_assembled <- patchwork::wrap_plots(hb_cases_plots, ncol = 2)
-# 
-# hb_cases_plots_assembled
 ```

### 2021-09-15 compile_avoidable_death_codes.Rmd
started a script to compile 'avoidable deaths' codes from NRS website

---
title: "Compiling avoidable deaths codes (as coded by NRS)"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

NRS publish an annual report on the number of avoidable/treatable deaths. The latest report is from December 2020 and reports on data from 2019, available on [the NRS website](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/avoidable-mortality).

The data used in the report can be downloaded in .csv and .xlsx format - I'll extract the code list for avoidable deaths from the excel file (.xlsx): [https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2019/avoid-mortality-19-all-tabs.xlsx](https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2019/avoid-mortality-19-all-tabs.xlsx)

# Load packages

```{r}
library(tidyverse)
library(readxl)  # for loading excel files
library(janitor)  # for cleaning data
library(knitr)  # for pretty tables in the report
```

# Download file

```{r}
url_avoidable_deaths_2019 <- "https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2019/avoid-mortality-19-all-tabs.xlsx"
path_avoidable_deaths_2019 <- "./downloaded_data/avoid-mortality-19-all-tabs.xlsx"

if (!file.exists(path_avoidable_deaths_2019)) download.file(url = url_avoidable_deaths_2019, destfile = path_avoidable_deaths_2019, mode = "wb")
```

# Load definitions

The excel sheet containing definitions is helpfully named 'Definitions':

```{r}
avoidable_deaths_raw <- read_excel(path = path_avoidable_deaths_2019, sheet = "Definitions", range = "A4:E225") %>%
  clean_names
```

## Inclusions and exclusions

The column `ICD-10 codes` lists codes in the following formats:

* most codes are listed as comma-separated values (e.g. `A35, A36, A80`) - these can be easily 'split' into several rows with one value per row
* some codes are listed as ranges (using dash/hyphen, e.g. `A00-A09`) - these need to be parsed. This can be done in a 'dumb' manner without reference to the specifics of ICD-10 where not all codes in such a range exist, because invalid codes simply won't match actual data
* some codes are listed with exclusions, which are always listed in parentheses (e.g. `A40 (excl. A40.3), A41 (excl. A41.3)`). This could be somewhat complex because we might need to track which exclusion applies to which code. However, it looks like there are only 2 entries with excepetions, and there is only ever one exception per listed code or code range, which means the exclusion can simply go in a new column.
    - Note that it is possible that the same death record may contain both codes that are included under avoidable deaths and excluded. This is something to check in the actual data!

```{r}
avoidable_deaths_raw %>% filter(str_detect(string = icd_10_codes, pattern = regex("excl", ignore_case = TRUE))) %>% kable(caption = "Entries with exclusions listed")

avoidable_deaths_raw %>% filter(str_detect(string = icd_10_codes, pattern = regex("\\-", ignore_case = TRUE))) %>% kable(caption = "Entries specifying code ranges")

# TODO: add code to check length so we can handle decimal codes also!
parse_icd_10_range <- function(code_range) {
  ## helper function for generating code ranges - based on the icd package, function .icd10cm_get_majors_possible()
  if (is.na(code_range) | !str_detect(string = code_range, pattern="\\-")) return(code_range)

  letter_1 <- str_replace(string = code_range, pattern = "^([A-Za-z]).*$", replacement = "\\1")
  letter_2 <- str_replace(string = code_range, pattern = "^.+\\-([A-Za-z]).*$", replacement = "\\1")
  letters <- LETTERS[which(LETTERS==letter_1) : which(LETTERS==letter_2)]

  decimal <- str_detect(string = code_range, pattern = "\\.")
  number_of_digits <- 2
  if (decimal) number_of_digits <- 3

  number_1 <- str_replace(string = str_remove_all(code_range, pattern="\\."), pattern = "^[A-Za-z](\\d{1,3})\\-.*$", replacement = "\\1")
  number_2 <- str_replace(string = str_remove_all(code_range, pattern="\\."), pattern = "^.+\\-[A-Za-z](\\d{1,3})$", replacement = "\\1")

  code_1 <- paste0(letter_1,number_1)
  code_2 <- paste0(letter_2,number_2)

  if (decimal) {
    candidates <- sort(apply(
        expand.grid(letters, 0:9, 0:9, 0:9),
        MARGIN = 1,
        FUN = paste0,
        collapse = ""
      ))
  } else {
    candidates <- sort(apply(
        expand.grid(letters, 0:9, 0:9),
        MARGIN = 1,
        FUN = paste0,
        collapse = ""
      ))
  }

  parsed_codes <- candidates[seq.int(
    from = which(candidates == code_1),
    to = which(candidates == code_2)
  )]
  if (decimal) parsed_codes <- paste0(str_sub(parsed_codes, 1, 3), ".", str_sub(parsed_codes, 4, 4))
  parsed_codes <- paste0(parsed_codes, collapse=",")
  return(parsed_codes)
}
```

```{r}
# TODO: show how the ranges have been parsed just to be sure
avoidable_deaths <-
  avoidable_deaths_raw %>%
  drop_na(condition_group_and_cause) %>%  # remove blank rows used for blank space in the excel file
  separate_rows(icd_10_codes, sep=",\\s*") %>%  # separate comma-separated values into rows so we have one entry per row
  extract(col = icd_10_codes, into = "exclusions", regex = "\\(excl\\.\\s*(.+)\\)", remove = FALSE) %>%  # extract the exclusions and put them in a separate column
  mutate(
    icd_10_codes = str_remove_all(string = icd_10_codes, pattern = "\\s*\\(excl.+\\)"),  # remove the exclusions from the code ranges
    icd_10_codes = map_chr(.x = icd_10_codes, .f = ~parse_icd_10_range(.x))
  )
```

# Notes on usage

When using this with real-world data, make sure to check whether the same record contains both an inclusion and an exclusion code according to this list! What to do in this case will differ from case to case.

### 2021-09-17 compile_avoidable_death_codes.Rmd
Completed compiling avoidable death codes; also added a table showing how code ranges were parsed

@@
 
-# TODO: add code to check length so we can handle decimal codes also!
 parse_icd_10_range <- function(code_range) {
@@
 ```{r}
-# TODO: show how the ranges have been parsed just to be sure
 avoidable_deaths <-
+  avoidable_deaths_raw %>%
+  drop_na(condition_group_and_cause) %>%  # remove blank rows used for blank space in the excel file
+  mutate(
+    group = if_else(condition = is.na(icd_10_codes), true = condition_group_and_cause, false = NA_character_),
+    original_codes = icd_10_codes
+    ) %>%
+  fill(group, .direction = "down") %>%  # forward-fill group
+  drop_na(icd_10_codes) %>%
+  separate_rows(icd_10_codes, sep=",\\s*", ) %>%  # separate comma-separated values into rows so we have one entry per row
+  extract(col = icd_10_codes, into = "exclusions", regex = "\\(excl\\.\\s*(.+)\\)", remove = FALSE) %>%  # extract the exclusions and put them in a separate column
+  mutate(
+    icd_10_codes = str_remove_all(string = icd_10_codes, pattern = "\\s*\\(excl.+\\)"),  # remove the exclusions from the code ranges
+    icd_10_codes = map_chr(.x = icd_10_codes, .f = ~parse_icd_10_range(.x))
+  ) %>%
+  separate_rows(icd_10_codes, sep=",\\s*") %>%  # separate comma-separated values again now that we've converted them from the ranges
+  separate(col = age, into = c("age_min", "age_max") ,sep = "\\-", remove = TRUE) %>%
+  mutate(
+    treatable = case_when(  # make more explicit the treatable/preventable distinction
+      str_detect(treatable, pattern = "50") ~ 0.5,
+      str_detect(treatable, pattern = "\\*") ~ 1,
+      TRUE ~ 0
+    ),
+    preventable = case_when(
+      str_detect(preventable, pattern = "50") ~ 0.5,
+      str_detect(preventable, pattern = "\\•") ~ 1,
+      TRUE ~ 0
+    ),
+  ) %>%
+  relocate(group, condition_group_and_cause, original_codes)
+```
+
+## Checking parsing of ranges
+
+```{r}
 avoidable_deaths_raw %>%
@@
   extract(col = icd_10_codes, into = "exclusions", regex = "\\(excl\\.\\s*(.+)\\)", remove = FALSE) %>%  # extract the exclusions and put them in a separate column
+  filter(str_detect(string = icd_10_codes, pattern = "\\-")) %>%
   mutate(
     icd_10_codes = str_remove_all(string = icd_10_codes, pattern = "\\s*\\(excl.+\\)"),  # remove the exclusions from the code ranges
-    icd_10_codes = map_chr(.x = icd_10_codes, .f = ~parse_icd_10_range(.x))
-  )
+    icd_10_codes_parsed = map_chr(.x = icd_10_codes, .f = ~parse_icd_10_range(.x))
+  ) %>%
+  select(icd_10_codes, icd_10_codes_parsed) %>% 
+  kable(caption = "Parsing of codes defined as ranges.")
 ```
@@
 What to do in this case will differ from case to case.
+
+# Save as `.csv`
+
+```{r}
+if (!dir.exists("./avoidable_deaths")) dir.create("./avoidable_deaths")
+write_csv(avoidable_deaths, file = "./avoidable_deaths/avoidable_deaths_2019_inclusions.csv")
+```

### 2021-09-23 hospital_classification.Rmd
added hospital classification .Rmd

---
title: "Hospital classification in Scotland"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(readxl)  # for reading excel files
library(knitr)  # for printing tables
```

# Classifications overview

A simple (with no classifications) list of Scotland's NHS hospitals is available from the [Scottish Open Data website]("https://www.opendata.nhs.scot/dataset/cbd1802e-0e04-4282-88eb-d7bdcfb120f0/resource/c698f450-eeed-41a0-88f7-c1e40a568acc/download/current-hospital_flagged20210506.csv")

From my research so far it appears there is not a single NHS-wide classification of hospitals. Different departments/units use classifications for their own purposes, it seems. For our purposes, we need hospitals that 'a person nearing their end of life might go to/be taken to' - in other words, we are looking for hospitals providing acute treatment, but probably also hospitals with particular specialties. However, there are probably specifics to which hospitals, if any, people migh use in any geographical area for a particular situation, which is probably not possible to infer from administrative data alone.

The two classifications I have found so far:

* `Acute hospitals` as defined within ISD(S)1 reports on hospital activity, for example on [Bed Occupancy]("https://www.isdscotland.org/Health-Topics/Hospital-Care/Publications/2019-11-26/Acute-Hospital-Publication/data-files/docs/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv"). Whilst this does not include an explicit list of hospitals, we can infer the acute hospitals as the unique locations reported in the data.
* `Hospital classifications` as reported in the `Costs` section of ISD Scotland's website: there are two files, one listing the [classification code for each hospital]("https://beta.isdscotland.org/media/7790/hospital_profile_2020.xlsx"), and another listing the [descriptions of the codes]("https://beta.isdscotland.org/media/7786/costs_hospital_classification_2020.xlsx).

The above can be combined and cross-checked (e.g. are all teaching hospitals also acute hospitals?) to form one or more listings of relevant hospitals.

# Downloading classifications

```{r}
dir_classifications <- "./hospital_classification/"
if (!dir.exists(dir_classifications)) dir.create(dir_classifications)

url_isds1 <- "https://www.isdscotland.org/Health-Topics/Hospital-Care/Publications/2019-11-26/Acute-Hospital-Publication/data-files/docs/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv"
url_hospital_classifications <- "https://beta.isdscotland.org/media/7790/hospital_profile_2020.xlsx"
url_hospital_classifications_descriptions <- "https://beta.isdscotland.org/media/7786/costs_hospital_classification_2020.xlsx"

walk(  # download files if they aren't already there
  .x = list(url_isds1,url_hospital_classifications,url_hospital_classifications_descriptions),
  .f = function(x) {
    url <- x
    filename <- basename(x)
    path <- file.path(dir_classifications,filename)
    if (!file.exists(path)) {
      download.file(url = url, destfile = path, quiet = TRUE)
      }
  }
)
```

# Parsing classifications

```{r}
hospital_classification <- 
  read_excel(path  = "./hospital_classification/hospital_profile_2020.xlsx", sheet = 1) %>%
  rename(sub_classification = HospGroup)  # rename to match below

classification_descriptions <- 
  read_excel(path  = "./hospital_classification/costs_hospital_classification_2020.xlsx", sheet = 1, skip = 2) %>% 
  slice(-c(1:2)) %>%
  fill(Main, .direction = "down") %>%
  mutate(
    main_classification = if_else(condition = is.na(Sub), true = Title, false = NA_character_),
    sub_classification = paste0(Main, Sub)
    ) %>%
  fill(main_classification, .direction = "down") %>%
  drop_na(Sub) %>%
  select(-Main, -Sub) %>%
  rename(description = Title)

acute_hospitals <- read_csv("./hospital_classification/20191126_Beds_by_Health_Board_of_Treatment_and_Specialty.csv") %>%
  select(hb_code, loc_code) %>% distinct

hospital_classification <-
  hospital_classification %>%
  left_join(classification_descriptions, by = "sub_classification") %>%
  mutate(acute = location %in% acute_hospitals$loc_code)
```

# Checking classifications

## Are all acute hospitals general/teaching hospitals?

```{r}
hospital_classification %>%
  count(main_classification, acute) %>%
  kable(caption = "Table of Acute hospital status by main classification.")

hospital_classification %>%
  filter((main_classification == "General Hospitals (mainly acute)" & !acute) | (main_classification == "Other" & acute)) %>%
  select(org_loc:acute) %>%
  kable(caption = "Unexpected combinations - acute non-general hospitals, and general non-acute hospitals.")
```

## Which hospital sub-classifications are in `Acute hospitals`?

```{r}
hospital_classification %>%
  filter(acute) %>%
  count(sub_classification, description) %>%
  kable(caption = "Acute hospitals by sub-classification.")
```

# Conclusion

It looks like for our purposes, `Acute` hospitals as per above are the best fit - they capture all general hospitals. Just using the `General hospital` classification would omit a few hospitals in `Other` that are also classed as `Acute`, and would also include `Hospitals for sick children` unless  sub-classifications were used.

# Save classification as `.csv`

```{r}
write_csv(hospital_classification, file = "./hospital_classification/hospital_classification.csv")
```

### 2021-09-30 covid_nrs_place_of_death_visualisations.Rmd
changed headings a bit so they make more sense
cleaning up commented out bits of code
fixed decimal place in proportion of deaths by place over time plot

@@
   scale_x_continuous(breaks = 2015:2021) +
-  scale_y_continuous(label = scales::percent) +
+  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
   scale_colour_viridis_d(option="C") +
@@
 
-## Annual deaths by cause
+## Annual deaths by cause, 2020 only
 
@@
 
-## Deaths by cause, from 2020 until current day
+## Deaths by cause, 2020 to now
+
+It would be great to have a separate 2019 bar to show changes from last year only, however, cause of death data are only provided as an average 2015-2019!
 

### 2021-09-30 highlands_data_and_visualisations.Rmd
started working on highlands data excerpt

---
title: "Deaths at home in NHS Highlands / North Highland LA"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
```

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

# Load most recent data

```{r}
load(file = "./workspace_with_visualisations.RData")
```

# Visualisations

Note: visualisations have been moved to a separate .Rmd file

## Common elements

```{r}
loadfonts()  # load imported fonts
# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))

date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2021, '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
only_2021_text <- glue::glue("Data for 2021 only include weeks 1-{format(weeks_available_2021)} (w/c {format(most_recent_date_available_2021)}).")

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

death_registrations_text <- "Note: all figures are based on date of registration."

vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."

```

## Function to save files to multiple formats

Saving to multiple formats is easy enough - the trouble is with saving PDF files with custom fonts, since they need to be embedded in the .pdf file. This requires:

* telling R where to load the font from, using `extrafont::font_import()` - this was done in a previous step
* registering fonts with R using `extrafont::loadfonts()` - note: this isn't strictly necessary, since once imported, `extrafont` keeps track of fonts to load when the `extrafont` package is loaded
* ~~telling R where the `gs` executable is, i.e. the installation of ghostscript on the current machine~~
* ~~embedding the fonts in the `.pdf` files once they've been written, using `extrafont::embed_fonts()`~~
* instead of manually embedding fonts (which resulted in garbled text on my setup!), just use `device=device_cairo` as an argument to `ggsave()` when saving `.pdf` files and it just works

```{r}
save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {

  walk(
    .x = extensions,
    .f = function(ext) {
    # if (tolower(ext)==".pdf") device <- cairo_pdf
    ggsave(filename = paste0(filename, ext), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (tolower(ext)==".pdf") cairo_pdf else NULL)
    }
  )

# TODO: remove below bit, not needed when using device=device_cairo
# if (any(str_detect(string = tolower(extensions), pattern = "pdf"))) {
#   pdf_filename <- paste0(filename, extensions[which(any(str_detect(string = tolower(extensions), pattern = "pdf")))])
#   embed_fonts(file = pdf_filename)  # overwrites file with one with font embedded
# }
# invisible(filename)  # return invisibly
}

```

# NHS Highlands / North Highland LA data

## Create directory

```{r}
dir_outputs_highlands <- gsub(x = dir_outputs, pattern = "outputs", replacement = "outputs/highlands")
if (!dir.exists(dir_outputs_highlands)) dir.create(dir_outputs_highlands, recursive = TRUE)
```

## Table of deaths

## Deaths by place

```{r}
## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
annotations_figure_place_of_death_explaining_historic_range_nhs_highland <- tribble(
   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
  "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
  "curve", "Hospital", NA, 29, 830, 32, 550,
  "text", "Hospital", "Historic mean", 22, 800, NA, NA,
  "curve", "Hospital", NA, 22, 780, 26, 510
) %>% 
  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)

(figure_place_of_death_nhs_highland <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = "Date w/c",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
  geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
  NULL
)

# save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death, width = 28, height = 14, dpi = 300, units = "cm")
```

## Home deaths

### NHS Highland

```{r}
(figure_deaths_at_home_2020_and_2021_nhs_highland <-
  merged_deaths_hb %>%
  filter(ref_area == "Highland") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  annotate(geom = "text", x = 15, y = 36, label = "2020", alpha = 0.4) +
  annotate(geom = "text", x = 8, y = 43, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "NHS Highland",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 11, xend = 15, y = 7, yend = 11, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 20, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_nhs_highlands"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_nhs_highland, width = 18, height = 14, dpi = 300, units = "cm")
```

### Highland LA

```{r}
(figure_deaths_at_home_2020_and_2021_highland_la <-
  merged_deaths_la %>%
  filter(ref_area == "Highland") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  annotate(geom = "text", x = 21, y = 31, label = "2020", alpha = 0.4) +
  annotate(geom = "text", x = 10, y = 28, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "Highland (Local authority)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 9, xend = 10.5, y = 6, yend = 8, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 15, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_highland_la"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_highland_la, width = 18, height = 14, dpi = 300, units = "cm")
```

### Combined

```{r}
# TODO: complete combined graph
(bind_rows(  # combine the two geographical types
  merged_deaths_hb %>%
    filter(ref_area == "Highland") %>%
    filter(place_of_death %in% c("Home & other non-institution")) %>%
    mutate(geography = "Health board"),
  merged_deaths_la %>%
    filter(ref_area == "Highland") %>%
    filter(place_of_death %in% c("Home & other non-institution")) %>%
    mutate(geography = "Local authority")
) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(year~geography) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "Highland (Local authority)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 9, xend = 10.5, y = 6, yend = 8, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 15, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)
```

## Proportion of deaths by place by year

```{r}
# TODO: add NHS highland & LA!
(figure_proportion_deaths_by_place_by_year_highland <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>% mutate(geography = "scotland"),
    proportions_of_deaths_by_place_by_geography
    ) %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(geography = case_when(
    geography == "hb" ~ "NHS Highland",
    geography == "la" ~ "Highland (Local Authority)",
    tolower(geography) == "scotland" ~ "Scotland",
    TRUE ~ NA_character_
  )) %>%
  filter(
    geography == "Scotland" | (!is.na(ref_area) & ref_area == "Highland")
  ) %>%
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(size = 1) +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  facet_wrap(~geography) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(
    x = "Year", y= "Proportion of yearly deaths (%)",
    colour = "",
    subtitle = paste0("* ",only_2021_text),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 13, height = 10, units = "cm", dpi = 300)
```

# Print session info

```{r}
sessionInfo()
```

### 2021-09-30 highlands_data_and_visualisations.Rmd
completed 2 graphs so far: deaths at home ocer time, and proportion of deaths by place over time - next up is breakdown of causes of death!

@@
 library(lemon)  # for showing axis & axis labels in every facet in ggplot
+library(patchwork)  # for assembling multiple plots
 ```
@@
 
-### NHS Highland
+```{r}
+annotations_deaths_at_home_highland <- tribble(
+  ~geography, ~place_of_death, ~x, ~y, ~label,
+  "NHS Highland", "Home & other non-institution", 15, 36, 2020,
+  "NHS Highland", "Home & other non-institution", 8, 43, 2021,
+  "Highland (Local authority)", "Home & other non-institution", 21, 31, 2020,
+  "Highland (Local authority)", "Home & other non-institution", 10, 28, 2021,
+) %>% mutate(alpha = if_else(label==2021, 1, 0.4)) %>%   # change alpha for 2020
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
 
-```{r}
-(figure_deaths_at_home_2020_and_2021_nhs_highland <-
+arrows_deaths_at_home_highland <- tribble(
+  ~geography, ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
+  "NHS Highland", "text", "Home & other non-institution", "Historic range (min & max)", 11, 5, NA, NA,
+  "NHS Highland", "curve", "Home & other non-institution", NA, 11, 7, 15, 11,
+  "NHS Highland", "text", "Home & other non-institution", "Historic mean", 20, 1, NA, NA,
+  "NHS Highland", "curve", "Home & other non-institution", NA, 23, 3, 29, 18
+) %>% 
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+
+(figure_deaths_at_home_2020_and_2021_highland <-
+  bind_rows(  # combine the two geographical types
     merged_deaths_hb %>%
@@
       filter(place_of_death %in% c("Home & other non-institution")) %>%
+      mutate(geography = "NHS Highland"),
+    merged_deaths_la %>%
+      filter(ref_area == "Highland") %>%
+      filter(place_of_death %in% c("Home & other non-institution")) %>%
+      mutate(geography = "Highland (Local authority)")
+  ) %>%
   mutate(year = factor(year)) %>%
@@
   ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths",
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
-  annotate(geom = "text", x = 15, y = 36, label = "2020", alpha = 0.4) +
-  annotate(geom = "text", x = 8, y = 43, label = "2021") +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Weekly deaths (N)",
-       title = "NHS Highland",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom = "curve", x = 11, xend = 15, y = 7, yend = 11, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
-  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 20, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  NULL
-)
-
-save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_nhs_highlands"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_nhs_highland, width = 18, height = 14, dpi = 300, units = "cm")
-```
-
-### Highland LA
-
-```{r}
-(figure_deaths_at_home_2020_and_2021_highland_la <-
-  merged_deaths_la %>%
-  filter(ref_area == "Highland") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  mutate(year = factor(year)) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
-               names_to = "deaths_type", values_to = "deaths") %>%
-  ggplot(aes(x = week_number)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
-  geom_line(data = .
-%>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
-  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
-  scale_x_continuous(breaks = seq(0,60,by=12)) +
-  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths",
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
-  annotate(geom = "text", x = 21, y = 31, label = "2020", alpha = 0.4) +
-  annotate(geom = "text", x = 10, y = 28, label = "2021") +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
-  labs(x = "Week number",
-       y = "Weekly deaths (N)",
-       title = "Highland (Local authority)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom = "curve", x = 9, xend = 10.5, y = 6, yend = 8, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
-  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 15, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  NULL
-)
-
-save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_highland_la"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_highland_la, width = 18, height = 14, dpi = 300, units = "cm")
-```
-
-### Combined
-
-```{r}
-# TODO: complete combined graph
-(bind_rows(  # combine the two geographical types
-  merged_deaths_hb %>%
-    filter(ref_area == "Highland") %>%
-    filter(place_of_death %in% c("Home & other non-institution")) %>%
-    mutate(geography = "Health board"),
-  merged_deaths_la %>%
-    filter(ref_area == "Highland") %>%
-    filter(place_of_death %in% c("Home & other non-institution")) %>%
-    mutate(geography = "Local authority")
-) %>%
-  mutate(year = factor(year)) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
-               names_to = "deaths_type", values_to = "deaths") %>%
-  ggplot(aes(x = week_number)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  scale_x_continuous(breaks = seq(0,60,by=12)) +
-  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
-  facet_grid(year~geography) +
+  facet_grid(~geography) +
   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
@@
        y = "Weekly deaths (N)",
-       title = "Highland (Local authority)",
+       title = "Highland: Health Board & Local Authority",
        subtitle = date_range_text,
@@
   ## annotate historic range
-  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom = "curve", x = 9, xend = 10.5, y = 6, yend = 8, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
-  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 15, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  geom_text(data = annotations_deaths_at_home_highland %>% filter(label==2021), aes(x=x, y=y, label=label), fontface = "bold") +
+  geom_text(data = annotations_deaths_at_home_highland %>% filter(label==2020), aes(x=x, y=y, label=label), alpha = 0.6) +
+  geom_text(data = arrows_deaths_at_home_highland %>% filter(geom=="text"), aes(x=x,y=y,label=label)) +
+  geom_curve(data = arrows_deaths_at_home_highland %>% filter(geom=="curve") %>% slice(1), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), alpha = 0.5) +
+  geom_curve(data = arrows_deaths_at_home_highland %>% filter(geom=="curve") %>% slice(2), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), linetype = "dashed") +
   NULL
 )
+
+save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_highland"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_highland, width = 24, height = 14, dpi = 300, units = "cm")
 ```
@@
 ```{r}
-# TODO: add NHS highland & LA!
 (figure_proportion_deaths_by_place_by_year_highland <-
@@
   geom_line(size = 1) +
-  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
   scale_x_continuous(breaks = 2015:2021) +
@@
     colour = "",
+    title = "Highland Health Board & Local Authority",
     subtitle = paste0("* ",only_2021_text),
@@
 
-save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 13, height = 10, units = "cm", dpi = 300)
+save_output_file(filename = paste0(dir_outputs_highlands,"/figure_proportion_home_deaths_over_time_highland"), extensions = c(".pdf",".png"), plot = figure_proportion_deaths_by_place_by_year_highland, width = 22, height = 10, units = "cm", dpi = 300)
 ```

### 2021-09-30 highlands_data_and_visualisations.Rmd
cleaned up script a bit and added a table summarising number of deaths in Highland; deprivation and cause of death aren't available on HB/LA level so they can't be included.

@@
 ---
-title: "Deaths at home in NHS Highlands / North Highland LA"
+title: "Deaths at home in NHS Highland / Highland LA"
 author: "Jan Savinc"
@@
   )
-
-# TODO: remove below bit, not needed when using device=device_cairo
-# if (any(str_detect(string = tolower(extensions), pattern = "pdf"))) {
-#   pdf_filename <- paste0(filename, extensions[which(any(str_detect(string = tolower(extensions), pattern = "pdf")))])
-#   embed_fonts(file = pdf_filename)  # overwrites file with one with font embedded
-# }
-# invisible(filename)  # return invisibly
 }
-
 ```
 
-# NHS Highlands / North Highland LA data
+# NHS Highlands / Highland LA data
+
+On 28 September, Michael Loynd of NHS Highland asked for a data excerpt for NHS Highland and North Highland (sic) local authority.
 
@@
 
-## Deaths by place
+```{r}
+table_current_deaths_compared_to_past_highland <-
+  bind_rows(
+    ## all available data
+    merged_deaths_hb %>%
+      mutate(geography = "HB") %>%
+      filter(ref_area == "Highland") %>%
+      group_by(year, place_of_death, week_number, geography) %>%
+      summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+      left_join(
+        past_deaths_hb %>% filter(year != "average" & ref_area == "Highland") %>% group_by(week_number, place_of_death) %>% summarise(
+                                           average_weekly_deaths = sum(number_of_deaths) / 5,
+                                           .groups = "drop"
+                                         ),
+        by = c("week_number", "place_of_death")
+      ) %>% 
+      group_by(place_of_death, geography) %>%
+      summarise(
+        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+        .groups = "drop"
+      )
+    ,
+    ## 2020 only
+    merged_deaths_hb %>%
+      mutate(geography = "HB") %>%
+      filter(ref_area == "Highland" & year == 2020) %>%
+      group_by(place_of_death, geography) %>%
+      summarise(
+        period = "2020",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_hb %>% 
+        filter(ref_area == "Highland") %>%
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+      by = "place_of_death"
+    )
+    ,
+    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
+    merged_deaths_hb %>%
+      mutate(geography = "HB") %>%
+      filter(ref_area == "Highland" &
+          year == 2020 & week_number >= 12
+      ) %>%
+      group_by(place_of_death, geography) %>%
+      summarise(
+        period = "2020, week 12 - week 53",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      past_deaths_hb %>% 
+        filter(ref_area == "Highland", week_number >= 12 & year != "average") %>%
+        group_by(place_of_death) %>%
+        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
+      by = "place_of_death"
+    ),
+    ## all available data
+    merged_deaths_la %>%
+      mutate(geography = "LA") %>%
+      filter(ref_area == "Highland") %>%
+      group_by(year, place_of_death, week_number, geography) %>%
+      summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+      left_join(
+        past_deaths_la %>% filter(year != "average" & ref_area == "Highland") %>% group_by(week_number, place_of_death) %>% summarise(
+                                           average_weekly_deaths = sum(number_of_deaths) / 5,
+                                           .groups = "drop"
+                                         ),
+        by = c("week_number", "place_of_death")
+      ) %>% 
+      group_by(place_of_death, geography) %>%
+      summarise(
+        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2021}, 2021"),
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+        .groups = "drop"
+      )
+    ,
+    ## 2020 only
+    merged_deaths_la %>%
+      mutate(geography = "LA") %>%
+      filter(ref_area == "Highland" & year == 2020) %>%
+      group_by(place_of_death, geography) %>%
+      summarise(
+        period = "2020",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_la %>% 
+        filter(ref_area == "Highland") %>%
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+      by = "place_of_death"
+    )
+    ,
+    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
+    merged_deaths_la %>%
+      mutate(geography = "LA") %>%
+      filter(ref_area == "Highland" &
+          year == 2020 & week_number >= 12
+      ) %>%
+      group_by(place_of_death, geography) %>%
+      summarise(
+        period = "2020, week 12 - week 53",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      past_deaths_la %>% 
+        filter(ref_area == "Highland", week_number >= 12 & year != "average") %>%
+        group_by(place_of_death) %>%
+        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
+      by = "place_of_death"
+    )
+  ) %>%
+  mutate(
+    ratio_deaths_all_causes_to_historical = deaths_all_causes / mean_annual_deaths_2015_2019,
+    ratio_non_covid_deaths_to_historical = deaths_non_covid / mean_annual_deaths_2015_2019,
+    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
+  ) %>% relocate(geography, place_of_death)
 
-```{r}
-## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
-annotations_figure_place_of_death_explaining_historic_range_nhs_highland <- tribble(
-   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
-  "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
-  "curve", "Hospital", NA, 29, 830, 32, 550,
-  "text", "Hospital", "Historic mean", 22, 800, NA, NA,
-  "curve", "Hospital", NA, 22, 780, 26, 510
-) %>% 
-  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-
-(figure_place_of_death_nhs_highland <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(!place_of_death %in% c("All","Other")) %>%
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(
-    legend.position = "top",
-    plot.caption = element_text(size = 10, colour = "gray60"),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = "Date w/c",
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
-       colour = "Cause of death"
-       ) +
-  ## annotate historic range
-  geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
-  geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
-  NULL
-)
-
-# save_output_file(filename = paste0(dir_outputs,"/figure_place_of_death"), extensions = c(".pdf",".svg",".png"), plot = figure_place_of_death, width = 28, height = 14, dpi = 300, units = "cm")
+write_csv(x = table_current_deaths_compared_to_past_highland, file = file.path(dir_outputs_highlands, "annual_death_rates_comparison_with_historical_deaths_highland.csv"))
 ```
@@
 
+## Cause of death & deprivation
+
+Unfortunately, as of 30 September 2021 these aren't available in combination with HB/LA, so they can't be reported for Highland individually.
+
 # Print session info

### 2021-09-30 download_latest_data_and_render_reports.R
Added new Highlands rmd to the rendering pipeline

@@ -263,3 +263,10 @@ render(
   envir = new.env()
 )
 
+render(
+  input = "./highlands_data_and_visualisations.Rmd",
+  output_file = paste0("highlands_data_and_visualisations_",today()),
+  output_dir = dir_reports,
+  envir = new.env()
+)
+

### 2021-10-01 download_latest_data_and_render_reports.R
commented out a line where a url for a shapefile no longer works

@@ -205,7 +205,7 @@ la_shapefile_url <- "https://opendata.arcgis.com/datasets/910f48f3c4b3400aa9eb0a
 
 # TODO: these have been downloaded so don't need to be downloaded again... change to check if they exist already first?
 # download_shapefile(url = health_boards_shapefile_url, directory = dir_map_data, subdirectory = "hb")
-# download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")
+# # download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")  # returning 404
 # download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
 
 # Download population estimates data --------------------------------------

### 2021-10-01 covid_nrs_place_of_death_data.Rmd
updated LA boundaries shapefile since link changed; changed output directory creation to be recursive, otherwise it fails!

@@
 dir_outputs <- paste0("./outputs/",latest_data_modified_date)
-if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
+if (!dir.exists(dir_outputs)) dir.create(dir_outputs, recursive = TRUE)
 ```
@@
 
-* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
+* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-december-2020-uk-buc/
 * Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards

### 2021-10-01 download_latest_data_and_render_reports.R
updated LA boundaries shapefile to latest since the previous one was removed; updated map data logic to only download if subdirectory doesn't exist yet (i.e.e hasn't yet been downloaded)

@@ -192,6 +192,11 @@ dir_map_data <- "./map_data"
 if (!dir.exists(dir_map_data)) dir.create(dir_map_data)
 
 download_shapefile <- function(url, directory, subdirectory) {
+  if (dir.exists(file.path(directory, subdirectory))) {
+    message(glue::glue("Map data for:{subdirectory} has already been downloaded, skipping."))
+    return(FALSE)
+  }
+  message(glue::glue("Downloading map data for:{subdirectory}...."))
   zip_file <- tempfile()
   downloaded_zip <- curl_download(url = url, destfile = zip_file, quiet = FALSE) 
   unzip(zipfile = zip_file, exdir = file.path(directory, subdirectory), overwrite = TRUE)
@@ -201,12 +206,13 @@ download_shapefile <- function(url, directory, subdirectory) {
 health_boards_shapefile_url <- "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip"  # massive resolution
 health_boards_uk_wide_url <- "https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip"
 # https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
-la_shapefile_url <- "https://opendata.arcgis.com/datasets/910f48f3c4b3400aa9eb0af9f8989bbe_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"
+# la_shapefile_url <- "https://opendata.arcgis.com/datasets/910f48f3c4b3400aa9eb0af9f8989bbe_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"  # this is the May 2020 link that no longer works
+la_shapefile_url <- "https://opendata.arcgis.com/api/v3/datasets/69dc11c7386943b4ad8893c45648b1e1_0/downloads/data?format=shp&spatialRefId=27700"
 
 # TODO: these have been downloaded so don't need to be downloaded again... change to check if they exist already first?
-# download_shapefile(url = health_boards_shapefile_url, directory = dir_map_data, subdirectory = "hb")
-# # download_shapefile(url = la_shapefile_url, directory = dir_map_data, subdirectory = "la")  # returning 404
-# download_shapefile(url = health_boards_uk_wide_url, directory = dir_map_data, subdirectory = "hb_uk_wide")
[…]

### 2021-10-15 location_code_exploration.Rmd
Started exploring location/institution codes to figure out what the Other death location in NHS Borders data represents...

---
title: "Exploring location codes to work out what the Other location for deaths in NHS Borders represent"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(tidyverse)
library(phsmethods)
```

# Question: what actual places does the `Other` category for place of death represent?

The `Other` category for place of death represents a very small fraction of annual deaths in Scotland. On looking into geographical variations in home deaths, I discovered that it's only the health board NHS Borders where `Other` place of deaths represent a sizable proportion of deaths. The question is what institutions these deaths represent - the first approximation is that these are likely to be prison & school deaths, since most other institutions are already accounted for in the Hospital/Home & other non-institutional setting/Care Home distinction (hospices are included partly in Hospital and partly in Care Home, the reason being explained below).

NHS Borders also has a relatively small population being 4th smallest of the 14 HBs, so perhaps the `Other` category appears sizable when it is in fact very small and would appear insignificant in the context of a larger HB.

# Location/institutiion codes in reporting NRS deaths

The following excerpt is from the [following NRS emthodology in reporting Covid-19 deaths document](https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-21-methodology.pdf):

> Public Health Scotland maintains a list of institutions across Scotland. These are mainly health-related institutions but the list also includes prisons and schools etc. NRS use this list to assign an institution code to the place of death on each death record. The institution codes are set up in such a way that the last letter of the code designates the type of institution. For the purposes of NRS death statistics, hospitals are counted as codes ending in H, care homes as codes ending in J, K, R, S, T, U or V, home / non-institution codes ending in N and other institution is everything else. The codes used for care homes are not an exact match. V contains some non-NHS hospitals, J and K contain some hospitals but mainly care homes. This grouping of codes is the closest we can get with the institution code structure which PHS has and, importantly, it’s consistent with previous practice. It should also be noted that there is no separate category for hospices. In some cases hospices have a code which means they are counted in the care home category, but in other cases hospices have a code which is counted in the hospital category. It is not possible to separate hospices out into a category of their own under the current coding structure used by PHS.

As will become clear, there are lots of codes in the `Other` category; from manual review they mainly comprise health centres, clinics, GP surgeries (A,B,C), residential living institutions,  primary schools (L), nurseries &  colleges (W), high schools (Q).

## Data availability

NRS publish aggregated weekly deaths by HB & place of death, which is the most granular data we can investigate. This means we can get figures for people who died in `Other` settings, but we still can't tell what these settings are.
NRS also provide a list of all locations with their associated code, and from their postcodes we can also infer which Health Board they belong to,

The list of location codes can be found on the [National Reference Files website](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/), with the relevant data dictionary at the [SMR file reference](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#Location%20Codes)

More information about the Scottish Postcode Directory (SPD) can be found on the [ISD website](https://www.isdscotland.org/Products-and-Services/GPD-Support/Geography/), and the actual directory is available to [download from a NRS website](https://www.nrscotland.gov.uk/statistics-and-data/geography/nrs-postcode-extract). I'm using the `Small users` file, which has more granular-level data than the `Large users` file.

# Downloading data

```{r}
## download location codes
if (!file.exists("./downloaded_data/location_codes")) download.file(url = "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/location?52573360", destfile = "./downloaded_data/location_codes", method = "curl")
```

# Loading locations, institutions, postcodes

```{r}
locations <- read_fwf(
  file = "./downloaded_data/location_codes", 
  col_positions = 
    fwf_cols(location_code = c(1,5), location_name = c(6,75), postcode = c(251,258))  # from the data dictionary in the SMR reference above
  ) %>%
  mutate(institution_code_nrs = str_sub(location_code, 5, 5))

## I'd downloaded the postcode data previously
postcodes <- 
  read_csv(
    file = "./distance_to_hospital/SPD_PostcodeIndex_Cut_21_1_CSV/SmallUser.csv", 
    col_names = c("postcode","hb_code"), 
    col_types = paste0(c("c",rep("_", times = 14), "c", rep("_", times = 57 - 16)), collapse=""),  # there are 57 columns, and we only need number 1 and number 16, both as character vector
    skip = 1
    ) %>%
  mutate(hb = phsmethods::match_area(hb_code))
```

# Coding place of death

```{r}
codes_hospital <- c("H")
codes_home <- c("N")
codes_care_home <- c("J", "K", "R", "S", "T", "U","V")
```

## what locations were counted as Home?

```{r}
locations %>% filter(institution_code_nrs %in% codes_home) %>% print(n=nrow(.))
```

## what are the `Other` locations?

Note: not all of these will actually appear as places of death in the death registrations, these are *potential* places of death coded as `Other`

```{r}
locations %>% filter(!institution_code_nrs %in% c(codes_hospital, codes_home, codes_care_home))
```

Note that there are too many to parse... and a lot of them are units within a hospital, which probably gets coded as a Hospital death regardless of which unit the death happeed in.

## maybe narrow it down to TD postcode - Borders LA
locations %>% filter(!institution_code_nrs %in% c(codes_hospital, codes_home, codes_care_home) & str_starts(string = postcode, pattern = "TD")) %>% print(n=nrow(.))

locations %>% filter(str_detect(string = location_name, pattern = regex("prison", ignore_case = TRUE)))

## narrow it down to Borders HB & Other locations
locations %>% 
  filter(!institution_code_nrs %in% c(codes_hospital, codes_home, codes_care_home)) %>%
  left_join(postcodes, by = "postcode") %>%
  filter(hb == "Borders") %>%
  arrange(word(location_name ,-1)) %>%
  print(n=nrow(.))

### 2021-10-19 covid_nrs_place_of_death_data.Rmd
exported proportions by place and geography over time file

@@
   )
+
+## write to csv
+proportions_of_deaths_by_place_by_geography %>%
+  write_csv(file = paste0(dir_outputs,"/proportions_of_deaths_by_place_by_geography.csv"))
 ```

### 2021-10-19 location_code_exploration.Rmd
getting closer with the location codes now, added a data dictionary and a few more bits

@@
 
+## From data dictionary
+
+These were provided by Evan Williams by email on 19 March 2021:
+
+>A - Admin Office
+B - Health Centre, GP Surgery
+C - Clinic Premises
+E - Extra-mural clinic
+H - NHS Hospital
+J - Joint User Hospital
+K - Contractual Hospital
+L - School
+M - Non NHS Maternity
+N - Non-Institutional, e.g. Domiciliary
+P - Prison
+Q - School
+R - Home for the Elderly
+S - Other Home
+T - Miscellaneous Premises
+V - Private Nursing Home, Private Hospital etc
+W - School
+
+```{r}
+# datapasta::vector_paste_vertical()  # use with above list
+institution_codes <-
+  c("A - Admin Office",
+     "B - Health Centre, GP Surgery",
+     "C - Clinic Premises",
+     "E - Extra-mural clinic",
+     "H - NHS Hospital",
+     "J - Joint User Hospital",
+     "K - Contractual Hospital",
+     "L - School",
+     "M - Non NHS Maternity",
+     "N - Non-Institutional, e.g. Domiciliary",
+     "P - Prison",
+     "Q - School",
+     "R - Home for the Elderly",
+     "S - Other Home",
+     "T - Miscellaneous Premises",
+     "V - Private Nursing Home, Private Hospital etc",
+     "W - School") %>%
+  tibble(institution_code_nrs = .) %>%
+  separate(col = institution_code_nrs, into = c("institution_code_nrs","institution_type"), sep = " - ", remove = TRUE)
+
+locations <-
+  locations %>%
+  left_join(institution_codes, by = "institution_code_nrs")
+
+## any institutions without a type?
+locations %>% filter(is.na(institution_type)) %>% count(institution_code_nrs)
+## YES: G & Z codes don't have a type associated
+```
+
+Using the above data dictionary, we are still left with G & Z-type institutions that don't appear in the dictionary...
+
+* `G` seems to cover clinics & medical units
+* `Z` seems to cover community centres, games halls, town halls
+
+## From NRS Covid-19 methodology paper:
+
 ```{r}
@@
 
-Note that there are too many to parse... and a lot of them are units within a hospital, which probably gets coded as a Hospital death regardless of which unit the death happeed in.
+Note that there are too many to parse... and a lot of them are units within a hospital, which probably gets coded as a Hospital death regardless of which unit the death happened in.
 
-## maybe narrow it down to TD postcode - Borders LA
-locations %>% filter(!institution_code_nrs %in% c(codes_hospital, codes_home, codes_care_home) & str_starts(string = postcode, pattern = "TD")) %>% print(n=nrow(.))
-
-locations %>% filter(str_detect(string = location_name, pattern = regex("prison", ignore_case = TRUE)))
-
-## narrow it down to Borders HB & Other locations
+```{r}
+locations_hb_borders_other <-
   locations %>% 
@@
   left_join(postcodes, by = "postcode") %>%
-  filter(hb == "Borders") %>%
-  arrange(word(location_name ,-1)) %>%
-  print(n=nrow(.))
+  filter(hb == "Borders")
+
+locations_hb_borders_other %>% count(institution_code_nrs)
+locations_hb_borders_other %>% filter(institution_code_nrs=="Q")
+
+## any prisons?
+locations_hb_borders_other %>% filter(str_detect(string = location_name, pattern = regex("prison|hmp", ignore_case = TRUE))) %>% print(n=nrow(.))
+
+locations_hb_borders_other %>% filter(!str_detect(string = location_name, pattern = regex("school", ignore_case = TRUE))) %>% print(n=nrow(.))
+```
+
+## What does the first letter in the `location code` mean?
+
+I suspect `B` is for Borders HB...
+
+```{r}
+locations_hb_borders_other %>% count(str_sub(location_code,1,1))
+
+locations %>% anti_join(locations_hb_borders_other) %>% count(str_sub(location_code, 1, 1))
+
+locations %>% anti_join(locations_hb_borders_other) %>% filter(str_sub(location_code, 1, 1) == "B") %>% count(is.na(postcode))
+
+locations %>% anti_join(locations_hb_borders_other) %>% filter(str_sub(location_code, 1, 1) == "B")
+```

### 2021-11-08 extract_elixhauser_icd_10_codes_quan_paper_from_icd_package.Rmd
Wrote up extraction of Elixhauser comorbidity codes

---
title: "Extract Elixhauser ICD-10 comorbidity codes from `icd` package"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
bibliography: references_extract_elixhauser.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

To use Elixhauser comorbidity codes [@elixhauser] in our study, we needed to provide the criteria for each comorbidity to eDRIS - these will be used to derive binary variables for each person based on their historical hospitalisation records. The original codes were provided for ICD-9-CM, but have since been translated to ICD-10 [@quan]. Unfortunately, the paper is from 2005 and only available as a `.pdf` file, and the codes are listed in a table broken over two pages, so they are not very easy to extract from the paper. The OCR done on the paper isn't great, either, misreading codes, for example, `I12` was read as `112` instead.

Fortunately, the `icd` package [@icd] contains a list of Elixhauser code from the Quan paper, so they can be extracted from the package instead of having to be input by hand.

# Install & load `icd` package

As of November 2021, the `icd` package isn't available on CRAN, so I've used the github version instead!

```{r, warning=FALSE, message=FALSE}
## uncomment if installation needed - note it will take a while because it needs compilation
# remotes::install_github("jackwasey/icd")

library(tidyverse)  # for tidy workflow
library(openxlsx)  # for writing excel files
```

# Get the codes out of `icd`

The object `icd::icd10_map_quan_elix` contains the Quan et al. (2005) provided codes, but its default printing method only shows the first 7 comorbidities and the first 7 codes associated with each. We need to `unclass()` it to get the underlying `list` object!

Note that the convention here is that ICD-10 stands for ICD-10-CM (Clinical Modification, the version of ICD-10 used in the USA) - some of these codes don't exist in the UK (WHO) version of ICD-10. It is also clear that rather than having been specified generally, e.g. F18 for drug abuse, all valid ICD-10-CM codes within F18 were enumerated... for our purposes, this is too detailed and some of the very detailed codes don't exist in the UK code list.

For this reason, I actually saved the codes from the `icd` package as a `.csv` and went through them to remove most codes specified at 5th character detail or more, to produce a more compact code that relates to ICD-10 codes rather than ICD-10-CM.

```{r}
list_of_elixhauser_codes_icd10cm <- unclass(icd::icd10_map_quan_elix)

tbl_of_elixhauser_codes_icd10cm <- 
  enframe(list_of_elixhauser_codes_icd10cm, name = "comorbidity", value = "code") %>%
  unnest(code) %>%
  mutate(
    code = as.character(code),
    code_decimal = paste0(str_sub(code, start = 1, end = 3), ".", str_sub(code, start = 4))
    )

write.csv(x = tbl_of_elixhauser_codes_icd10cm, file = "./elixhauser_comorbidities_quan_2005_icd10cm.csv")
```

After manual review, these codes can be loaded and cleaned up...

```{r, warning=FALSE, message=FALSE}
tbl_of_elixhauser_codes_icd10 <- 
  read_csv(file = "./elixhauser_comorbidities_quan_2005_icd10_manual.csv") %>%
  rename(icd10_prefix = code) %>%
  mutate(icd10_prefix_decimal = if_else(
    condition = nchar(icd10_prefix) > 3,
    true = paste0(str_sub(icd10_prefix, start = 1, end = 3), ".", str_sub(icd10_prefix, start = 4)),
    false = icd10_prefix
  ))

# datapasta::tribble_paste(tbl_of_elixhauser_codes_icd10 %>% select(comorbidity) %>% distinct %>% mutate(comorbidity_long = " "))
long_comorbidity_names <- tibble::tribble(
                              ~comorbidity, ~comorbidity_long,
                                     "CHF",               "Congestive heart failure",
                              "Arrhythmia",               "cardiac arrhythmias",
                                "Valvular",               "Valvular disease",
                                    "PHTN",               "Pulmonary circulation disorders",
                                     "PVD",               "Peripheral vascular disorders",
                                     "HTN",               "Hypertension, uncomplicated",
                                   "HTNcx",               "Hypertension, complicated",
                               "Paralysis",               "Paralysis",
                              "NeuroOther",               "Other neurological disorders",
                               "Pulmonary",               "Chronic pulmonary disease",
                                      "DM",               "Diabetes, uncomplicated",
                                    "DMcx",               "Diabetes, complicated",
                             "Hypothyroid",               "Hypothyroidism",
                                   "Renal",               "Renal failure",
                                   "Liver",               "Liver disease",
                                     "PUD",               "Peptic ulcer disease",
                                     "HIV",               "AIDS/HIV",
                                "Lymphoma",               "Lymphoma",
                                    "Mets",               "Metastatic cancer",
                                   "Tumor",               "Solid tumour without metastasis",
                               "Rheumatic",               "Rheumatoid arthritis/collagen vascular diseases",
                            "Coagulopathy",               "Coagulopathy",
                                 "Obesity",               "Obesity",
                              "WeightLoss",               "Weight loss",
                             "FluidsLytes",               "Fluid and electrolyte disorders",
                               "BloodLoss",               "Blood loss anaemia",
                                  "Anemia",               "Deficiency anaemia",
                                 "Alcohol",               "Alcohol abuse",
                                   "Drugs",               "Drug abuse",
                               "Psychoses",               "Psychoses",
                              "Depression",               "Depression"
                            )

tbl_of_elixhauser_codes_icd10 <-
  tbl_of_elixhauser_codes_icd10 %>%
  left_join(long_comorbidity_names) %>%
  relocate(comorbidity, comorbidity_long)

## condensed version of the table
tbl_of_elixhauser_codes_icd10_condensed <-
  tbl_of_elixhauser_codes_icd10 %>%
  group_by(comorbidity, comorbidity_long) %>%
  summarise(across(.cols = matches("icd10"), .fns = ~paste0(.x, collapse = ", ")), .groups = "drop")
```

# Save tables for use/publishing

```{r}
if (!dir.exists("./comorbidity/")) dir.create("./comorbidity/")

## write the table & condensed table separately as .csv files
tbl_of_elixhauser_codes_icd10 %>% write_csv(file = "./comorbidity/elixhauser_comorbidities_quan_2005_icd10.csv")
tbl_of_elixhauser_codes_icd10_condensed %>% write_csv(file = "./comorbidity/elixhauser_comorbidities_quan_2005_icd10_condensed.csv")

descriptive_note_for_excel <- tibble(
  Notes = c("ICD-10 codes are taken from the Quan et al. 2005 paper 'Coding Algorithms for Defining Comorbidities in ICD-9-CM and ICD-10 Administrative Data', https://www.jstor.org/stable/3768193, and are based on the Elixhauser et al. 1998 paper 'Comorbidity measures for use with administrative data'.", "The codes should be treated as prefixes - any ICD-10 code matching the first 3 or 4 characters provided in this list should be included as a comorbidity of the appropriate type even if the actual diagnosis code is longer; depending on the data format, this can be implemented using the regex caret operator (^) to match the start of a string, e.g. ^E52 to capture codes starting with E52. Codes are provided both with and without a decimal comma.", "The 'icd' package for R (https://github.com/jackwasey/icd) was also used in preparing the files; the package provides a list of ICD-10-CM codes which were truncated to ICD-10 codes.")
  )

write.xlsx(x = list(
  "Description" = descriptive_note_for_excel,
  "Prefixes (long)" = tbl_of_elixhauser_codes_icd10,
  "Prefixes (condensed)" = tbl_of_elixhauser_codes_icd10_condensed
), file = "./comorbidity/elixhauser_comorbidities_quan_2005_icd10.xlsx")
```

# REFERENCES

### 2021-11-08 covid_nrs_place_of_death_data.Rmd
added a .csv of locations coded as 'other'

@@
   write_csv(file = paste0(dir_outputs,"/proportions_of_deaths_by_place_by_geography.csv"))
+```
+
+## Number & proportion of deaths in `Other` location
+
+```{r}
+annual_deaths_coded_as_other_place_hb <-
+  proportions_of_deaths_by_place_by_geography %>% 
+  filter(geography=="hb" & place_of_death == "Other") %>% 
+  mutate(num_prop = paste0(format(number_of_deaths, big.mark = ","), " (", scales::percent(proportion_of_total, accuracy = 0.01),")")) %>%
+  select(year, place_of_death, ref_area, num_prop) %>% 
+  pivot_wider(names_from = year, values_from = num_prop)
+
+annual_deaths_coded_as_other_place_hb %>% write_csv(file = paste0(dir_outputs,"/annual_deaths_coded_as_other_place_hb.csv"))
 ```

### 2021-11-08 location_code_exploration.Rmd
expanded location exploration

@@
 library(phsmethods)
+library(readxl)  # for reading excel files
+library(lubridate)  # for working with dates & times
+library(janitor)
 ```
@@
 
-As will become clear, there are lots of codes in the `Other` category; from manual review they mainly comprise health centres, clinics, GP surgeries (A,B,C), residential living institutions,  primary schools (L), nurseries &  colleges (W), high schools (Q).
+As will become clear, there are lots of codes in the `Other` category.
+
+A list of all locations and their associated codes are available on the [NRS website](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-background-information/code-lists-used-in-vital-events-statistics), specifically at [the following link: Institutions](https://www.nrscotland.gov.uk/files//statistics/vital-events/institution-codes-october-2021.xlsx).
 
@@
 
-The list of location codes can be found on the [National Reference Files website](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/), with the relevant data dictionary at the [SMR file reference](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#Location%20Codes)
+~~The list of location codes can be found on the [National Reference Files website](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/), with the relevant data dictionary at the [SMR file reference](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#Location%20Codes)~~
+Note: see above for the NRS-provided list of institutions which also explains the type of institution and health board coding.
 
@@
 ## download location codes
-if (!file.exists("./downloaded_data/location_codes")) download.file(url = "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/location?52573360", destfile = "./downloaded_data/location_codes", method = "curl")
+## this is for the National Reference File, but it's better to use the NRS-provided file instead
+# if (!file.exists("./downloaded_data/location_codes")) download.file(url = "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/location?52573360", destfile = "./downloaded_data/location_codes", method = "curl")
+
+institutions_file_url <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/institution-codes-october-2021.xlsx"
+institutions_file_path <- "./downloaded_data/institution-codes-october-2021.xlsx"
+
+if (!file.exists(institutions_file_path)) download.file(url = institutions_file_url, destfile = institutions_file_path, method = "curl")
 ```
@@
   mutate(institution_code_nrs = str_sub(location_code, 5, 5))
+
+institution_codes <-
+  c("A - Admin Office",
+     "B - Health Centre, GP Surgery",
+     "C - Clinic Premises",
+     "E - Extra-mural clinic",
+     "H - NHS Hospital",
+     "J - Joint User Hospital",
+     "K - Contractual Hospital",
+     "L - School",
+     "M - Non NHS Maternity",
+     "N - Non-Institutional, e.g. Domiciliary",
+     "P - Prison",
+     "Q - School",
+     "R - Home for the Elderly",
+     "S - Other Home",
+     "T - Miscellaneous Premises",
+     "V - Private Nursing Home, Private Hospital etc",
+     "W - School") %>%
+  tibble(institution_code_nrs = .) %>%
+  separate(col = institution_code_nrs, into = c("institution_code_nrs","institution_type"), sep = " - ", remove = TRUE)
+
+institution_hb_codes <- read_excel(path = institutions_file_path, sheet = 1, range = "A7:B21", col_names = c("hb_code","hb"))
+institution_types <- read_excel(path = institutions_file_path, sheet = 1, range = "E7:G21", col_names = c("institution_type_code","institution_type","institution_type_detail"))
+
+institution_types <-
+  bind_rows(
+    institution_types,
+    institution_codes %>% rename(institution_type_code = institution_code_nrs) %>% anti_join(institution_types, by = "institution_type_code")
+  )
+
+institutions <- 
+  read_excel(path = institutions_file_path, sheet = 1, skip = 28, col_types	= rep(c("text","date"), each = 3)) %>%
+  clean_names %>%
+  filter(nchar(location_code)==5) %>%  # extract only actual entries, ignoring all the headers & formatting rows
+  mutate(
+    institution_type_code = str_sub(location_code, 5, 5),
+    hb_code = str_sub(location_code, 1, 1)
+  ) %>% 
+  left_join(institution_hb_codes, by = "hb_code") %>%
+  left_join(institution_types %>% select(institution_type, institution_type_code), by = "institution_type_code")
 
@@
 codes_care_home <- c("J", "K", "R", "S", "T", "U","V")
+
+institutions <-
+  institutions %>%
+  mutate(place_of_death = case_when(
+    institution_type_code %in% codes_hospital ~ "Hospital",
+    institution_type_code %in% codes_home ~ "Home & other non-institutional setting",
+    institution_type_code %in% codes_care_home ~ "Care Home",
+    TRUE ~ "Other"
+  ))
 ```
@@
 
-## What does the first letter in the `location code` mean?
+```{r}
+## relatively new (2016 or later institutions)
+institutions %>%
+  filter(hb=="Borders" & place_of_death == "Other") %>%
+  filter(year(location_closure_date) > 2016 | is.na(location_closure_date)) %>%
+  filter(year(location_start_date) >= 2016) %>%
+  View()
 
-I suspect `B` is for Borders HB...
+institutions %>%
+  filter(hb=="Borders" & place_of_death == "Other") %>%
+  filter(str_detect(location_name, pattern = regex("base|royal|raf|army|milit", ignore_case = TRUE))) %>%
+  .$location_name
 
-```{r}
-locations_hb_borders_other %>% count(str_sub(location_code,1,1))
-
-locations %>% anti_join(locations_hb_borders_other) %>% count(str_sub(location_code, 1, 1))
-
-locations %>% anti_join(locations_hb_borders_other) %>% filter(str_sub(location_code, 1, 1) == "B") %>% count(is.na(postcode))
-
-locations %>% anti_join(locations_hb_borders_other) %>% filter(str_sub(location_code, 1, 1) == "B")
+institutions %>%
+  filter(hb=="Borders" & place_of_death == "Other") %>%
+  filter(year(location_closure_date) > 2016 | is.na(location_closure_date)) %>%
+  View()
 ```

### 2021-11-15 visualisations_for_blog.Rmd
added plotly plot for geographical variation of place of death, needs a bit more work

@@
 library(knitr)  # for printing tables nicely
+library(lemon)  # for adding all axis labels to facet_wrap plots
 ```
@@
 ```
+
+# Plotly version of proportions of deaths by place by year by geography
+
+```{r}
+library(plotly)  # for interactive plots!
+library(htmlwidgets)  # for saving standalone html file
+
+plotly_proportion_of_deaths_by_place_by_hb <-
+  (proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "hb") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  # geom_point(aes(size = place_of_death)) +
+  geom_line(
+    aes(
+      size = place_of_death, 
+      group = place_of_death, 
+      text = paste0(  # using dummy text aesthetic, construct pairs of variable name: value, separated with line break <br|
+        "HB: ", ref_area,
+        "<br>Place of death: ", place_of_death,
+        "<br>Year: ", year,
+        "<br>Proportion: ", scales::percent(proportion_of_total, accuracy = 0.1)
+        )
+      )
+    ) +
+  # geom_line(aes(size = place_of_death)) +
+  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = 2015:2021) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    subtitle = paste0(c(paste0("* ",only_2021_text),"Health boards sorted by 2020 population estimate."),collapse="\n"),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2, title = NULL), 
+         size = guide_legend(nrow=2, title = NULL)) +
+  NULL) %>%
+  ggplotly(p = ., tooltip = c("text")) %>%
+  layout(legend = list(y = 1.1, orientation = 'h'))  # y=1.1 puts legend above plot
+
+# TODO: add average home death increase in Scotland
+
+plotly_proportion_of_deaths_by_place_by_hb <- plotly_build(plotly_proportion_of_deaths_by_place_by_hb)
+for (i in c(1:14, 29:56)) {
+  plotly_proportion_of_deaths_by_place_by_hb$x$data[[i]]$visible <- "legendonly"
+}
+
+plotly_proportion_of_deaths_by_place_by_hb %>%
+  partial_bundle() %>%
+  saveWidget(file = paste0(dir_blog,"/plotly_proportion_of_deaths_by_place_by_hb.html"))
+
+```

### 2021-11-19 plotly_graph_deaths_by_place_by_year.R
moved the plotly code to separate script so I can run it as I need to

@@ -1,0 +1,104 @@
+## Graph: Proportion of deaths in Scotland by place, 2015-2021
+## plotly version
+
+
+# Load packages -----------------------------------------------------------
+
+library(tidyverse)  # for tidy workflow
+library(lemon)  # for adding all axis labels to facet_wrap plots
+library(plotly)  # for interactive plots!
+library(htmlwidgets)  # for saving standalone html file
+
+
+# Theme setup to match blog style -----------------------------------------
+
+## set theme to match the blog style so far
+theme_set(theme_minimal(base_size = 12) +
+            theme(panel.grid.minor = element_blank(),
+                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
+                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
+                  text=element_text(family="Calibri")
+            ))
+
+
+# Create graph ------------------------------------------------------------
+
+plotly_proportion_of_deaths_by_place_by_hb <-
+  (proportions_of_deaths_by_place_by_geography %>%
+     filter(geography == "hb") %>%
+     mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+     mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
+     ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+     # geom_point(aes(size = place_of_death)) +
+     geom_line(
+       aes(
+         size = place_of_death, 
+         group = place_of_death, 
+         text = paste0(  # using dummy text aesthetic, construct pairs of variable name: value, separated with line break <br|
+           "HB: ", ref_area,
+           "<br>Place of death: ", place_of_death,
+           "<br>Year: ", year,
+           "<br>Proportion: ", scales::percent(proportion_of_total, accuracy = 0.1)
+         )
+       )
+     ) +
+     # geom_line(aes(size = place_of_death)) +
+     geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+     facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
[…]

### 2021-11-19 plotly_graph_deaths_by_place_by_year.R
oops, forgot to load the latest data!

@@ -10,6 +10,12 @@ library(plotly)  # for interactive plots!
 library(htmlwidgets)  # for saving standalone html file
 
 
+
+# Load data ---------------------------------------------------------------
+
+load(file = "./workspace_with_visualisations.RData")
+
+
 # Theme setup to match blog style -----------------------------------------
 
 ## set theme to match the blog style so far

### 2022-01-13 download_latest_data_and_render_reports.R
amended download script to find 2022 data also

@@ -27,7 +27,9 @@ all_nrs_links <-
   html_nodes("a") %>%
   html_attr("href")  # find all links on the page
 
-weekly_deaths_url_relative <- all_nrs_links[which(str_detect(all_nrs_links, pattern = "covid-deaths-21-data.*xlsx"))]  # keep only links that are probably weekly deaths
+nrs_filename_pattern <- "covid-deaths-(21|22)-data.*xlsx"  # to accommodate 2022 data also!
+
+weekly_deaths_url_relative <- all_nrs_links[which(str_detect(all_nrs_links, pattern = nrs_filename_pattern))]  # keep only links that are probably weekly deaths
 weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
 
 # Load table keeping track of latest data ---------------------------------

### 2022-01-13 covid_nrs_place_of_death_data.Rmd
added table output for SPPC data request - summary of 2020 and 2021 data

@@
   )
+
+current_and_historical_deaths_by_cause_and_location <-
+  bind_rows(
+    ## 2020 & 2021 separately
+    total_deaths_cause_of_death %>%
+      select(year, place_of_death, cause_of_death, deaths) %>%
+      mutate(year = as.character(year)),
+    ## 2020 & 2021 averaged
+    total_deaths_cause_of_death %>%
+      select(year, place_of_death, cause_of_death, deaths) %>%
+      group_by(place_of_death,  cause_of_death) %>%
+      summarise(deaths = sum(deaths), year = "2020-2021", .groups = "drop"),
+    ## historical data
+    total_deaths_cause_of_death %>%
+      select(year, place_of_death, cause_of_death, deaths = annual_deaths_mean) %>%
+      mutate(year = "2015-2019") %>%
+      distinct
+  ) %>%
+  filter(place_of_death != "All") %>%
+  group_by(year, cause_of_death) %>%
+  mutate(proportion = deaths / sum(deaths)) %>%
+  ungroup %>%
+  arrange(year, cause_of_death)
+
 ```
@@
     )
+    ,
+    ## 2021 only
+    merged_deaths_overall %>%
+      filter(age == "all" &
+               sex == "all" &
+               year == 2021
+             ) %>% 
+      group_by(place_of_death) %>%
+      summarise(
+        period = "2021",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_overall %>% 
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+      by = "place_of_death"
+    )
   ) %>%
@@
 table_current_deaths_compared_to_past %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
+```
+
+# Yearly death rates
+
+```{r}
+table_annual_death_rates <-
+  merged_deaths_overall %>%
+    filter(age == "all" &
+             sex == "all" &
+             year %in% c(2020:2021)
+           ) %>% 
+    group_by(year, place_of_death) %>%
+    summarise(
+      deaths_all_causes = sum(deaths_all_causes),
+      deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+      .groups = "drop"
+    ) %>% 
+    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+  left_join(
+    historical_annual_average_overall %>% 
+      select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+    by = "place_of_death"
+  ) %>%
+  mutate(
+    proportion_covid_deaths = scales::percent(deaths_covid_related / deaths_all_causes, accuracy = 0.1),
+    change_deaths_all_causes_from_historical = scales::percent((deaths_all_causes / mean_annual_deaths_2015_2019) - 1, accuracy = 0.1),
+    change_non_covid_deaths_from_historical = scales::percent((deaths_non_covid / mean_annual_deaths_2015_2019) - 1, accuracy = 0.1)
+  )
+
+table_annual_death_rates_by_cause <-
+  total_deaths_cause_of_death %>%
+  select(-matches("min|max")) %>%
+  filter(year %in% c(2020:2021)) %>%
+  mutate(
+    change_from_historical_annual_deaths_2015_2019 = scales::percent(ratio_annual_deaths_to_historical_mean - 1, accuracy = 0.1) %>% if_else(. == "Inf", NA_character_, .)
+    ) %>%
+  mutate(temporary_group_for_all_causes = cause_of_death=="All") %>%
+  group_by(year,place_of_death, temporary_group_for_all_causes) %>%
+  mutate(
+    proportion_annual_deaths_in_place = scales::percent(deaths / sum(deaths), accuracy = 0.1) %>% if_else(. == "100.0%", NA_character_, .),
+    historical_proportion_annual_deaths_in_place = scales::percent(annual_deaths_mean / sum(annual_deaths_mean), accuracy = 0.1) %>% if_else(. == "100.0%", NA_character_, .)
+    ) %>%
+  ungroup %>%
+  rename(mean_annual_deaths_2015_2019 = annual_deaths_mean) %>%
+  select(-c(ratio_annual_deaths_to_historical_mean, temporary_group_for_all_causes))
+
+notes_annual_death_rates <-
+  tibble( notes = c(
+    glue::glue("Note: these data are based on weekly figures reported by NRS; years don't align perfectly with calendar years, for example: week 1 of 2020 started on {ISOweek2date('2020-W01-1')}, and week 53 of 2020 ended on {ISOweek2date('2020-W53-7')}; week 1 of 2021 started on {ISOweek2date('2021-W01-1')}, and week 52 of 2021 ended on {ISOweek2date('2021-W52-7')}.")
+    )
+  )
+
+write.xlsx(
+    x = list(
+      "2020 + 2021" = table_annual_death_rates,
+      "Causes of death (COD)" = table_annual_death_rates_by_cause,
+      "notes" = notes_annual_death_rates
+      ), 
+    file = file.path("./outputs", "annual_death_rates_and_causes_of_death_2020_2021.xlsx")
+      )
 ```
@@
 ```{r}
-total_deaths_overall %>%
-  mutate(excess_deaths = deaths_all_causes - annual_deaths_min)
-
 excess_deaths_2020_table <-

### 2022-01-13 covid_nrs_place_of_death_visualisations.Rmd
work in progress - distribution of causes of death across places - e.g. cancer deaths distribution between years

@@
 
+## Proportion of cause of death by location
+
+This graph shows each of the causes of death divided by place of death - e.g. what proportion of cancer deaths happened at home vs hospital over time?
+
+```{r}
+# TODO: proportion of each cause of death that happens in a place
+
+# total_deaths_cause_of_death
+(
+  current_and_historical_deaths_by_cause_and_location %>%
+  ggplot() +
+  geom_col(aes(x = place_of_death, fill = place_of_death, y = proportion), alpha = 0.8, colour = "grey10") +
+  # geom_segment(aes(x = as.numeric(cause_of_death)-0.45, xend = as.numeric(cause_of_death)+0.45, y = deaths_mean, yend = deaths_mean), size = 0.8, colour = "grey1") +
+  # coord_flip() +
+  scale_fill_viridis_d() +
+  facet_grid(cause_of_death~year) +
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL,
+    y = "Deaths",
+    subtitle = paste0(c(paste0("Bars are deaths ",str_remove_all(date_range_text,"Data ")),"Line segments are mean deaths for same period 2015-2019"),collapse="\n")
+  ) +
+  NULL
+)
+
+(
+current_and_historical_deaths_by_cause_and_location %>%
+  # filter(year!="2020-2021") %>%
+  filter(!year %in% c("2020","2021")) %>%
+  filter(place_of_death!="Other") %>%
+  filter(!cause_of_death %in% c("COVID-19","All")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+   cause_of_death = fct_rev(cause_of_death)
+  ) %>%
+  ggplot(aes(x = place_of_death, group = year, y = proportion, fill = year)) +
+  geom_col(position = position_dodge(width = 0.9)) +
+  facet_grid(cause_of_death~.) +
+  scale_fill_viridis_d() +
+  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
+  coord_flip() +
+  theme(
+    panel.spacing = unit(1, units = "lines"),
+    legend.position = "top", legend.title = element_blank()
+    ) +
+  NULL
+)
+```
+
 # Case counts & home deaths - comparing plots

### 2022-01-13 covid_nrs_place_of_death_data.Rmd
tidied some commented out stuff

@@
 
-# TODO:  still too complicated to safely use as a function...
-## find which combinations of geographic area, cause of death, and place of death are missing and set the number of deaths to 0 for them
-# complete_weekly_deaths_with_implicit_zeros <- function(data_tbl) {
-#   data_tbl %>%
-#   tidyr::complete(
-#       nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
-#       ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
-#       fill = list(number_of_deaths = 0)
-#       )
-# }
-
 ## calculate number of deaths after recoding cause of death -
@@
 }
-
-# TODO: delete below function, it's too complicated and I've broken it down to the individual functions
-# recode_cause_of_death_and_compute_columns <- function(data_tbl) {
-#   potential_grouping_variables <- c("ref_area","sex","age","ref_period")
-#   grouping_variables <- potential_grouping_variables[which(potential_grouping_variables %in% names(data_tbl))]
-#   
-#   interim_data_tbl <- 
-#     data_tbl %>%
-#     recode_cause_of_death() %>%
-#     tidyr::complete(
-#       nesting(year, week_number, week_number_run_over, date_w_c),
-#       !!!rlang::syms(grouping_variables),
-#       cause_of_death, place_of_death,
-#       fill = list(number_of_deaths = 0)
-#       ) %>%
-#     pivot_wider(names_from = "cause_of_death", values_from = "number_of_deaths")
-#   
-#   ## compute non-covid deaths from all deaths minus covid deaths
-#   if (!"deaths_non_covid" %in% names(interim_data_tbl)) {
-#     interim_data_tbl <-
-#       interim_data_tbl %>%
-#       mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)
-#   }
-#   ## compute all deaths from non-covid deaths + covid deaths
-#   if (!"deaths_all_causes" %in% names(interim_data_tbl)) {
-#     interim_data_tbl <-
-#       interim_data_tbl %>%
-#       mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related)
-#   }
-#   
-#   interim_data_tbl <-
-#     interim_data_tbl %>%  # set as missing covid- & non-covid-related deaths before week 12
-#     mutate(
-#       deaths_non_covid = if_else(
-#         condition = week_number_run_over < 12,
-#         true = NA_real_,
-#         false = deaths_non_covid
-#       ),
-#       deaths_covid_related = if_else(
-#         condition = week_number_run_over < 12,
-#         true = NA_real_,
-#         false = deaths_covid_related
-#       )
-#     )
-#   
-#   return(interim_data_tbl)
-# }
 ```

### 2022-01-13 covid_nrs_place_of_death_visualisations.Rmd
made a nice graph showing cause of death distribution by place, but ggtext doesn't seem to work

--- 2021-09-30 covid_nrs_place_of_death_visualisations.Rmd
@@
 library(lemon)  # for showing axis & axis labels in every facet in ggplot
+library(ggtext)  # for showing coloured title text
 ```
@@
 
+# Proportion of cause of death by location
+
+This graph shows each of the causes of death divided by place of death - e.g. what proportion of cancer deaths happened at home vs hospital over time?
+
+```{r}
+(
+figure_cause_of_death_by_location <-
+  current_and_historical_deaths_by_cause_and_location %>%
+  filter(year!="2020-2021") %>%
+  filter(!cause_of_death %in% c("All")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+  ) %>%
+  ggplot(aes(x = year, y = proportion, colour = place_of_death, group = place_of_death)) +
+  geom_line() +
+  geom_point() +
+  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
+  scale_colour_viridis_d() +
+  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
+  labs(
+    x = "Year",
+    y = "Percentage yearly deaths (%)",
+    # title = "More cancer deaths in 2020 and 2021 happened at <span style='color:#31688E;'>Home</span> than in <span style='color:#440154;'>Hospital</span>",
+    title = "In 2020 and 2021 more cancer deaths happened at Home than in Hospital",
+    caption = paste0(c(source_nrs_text), collapse = "\n")
+  ) +
+  theme(
+    panel.spacing = unit(1, units = "lines"),
+    legend.position = "top", legend.title = element_blank()
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+)
+
+save_output_file(filename = paste0(dir_outputs,"/figure_cause_of_death_by_location"), extensions = c(".pdf",".png"), plot = figure_cause_of_death_by_location, width = 26, height = 16, units = "cm", dpi = 300)
+```
+
 # Case counts & home deaths - comparing plots

### 2022-01-13 covid_nrs_place_of_death_visualisations.Rmd
ah, found the issue with ggtext - needed to add a theme(plot.title = element_textbox()) to render the html in the title text

@@
     # title = "More cancer deaths in 2020 and 2021 happened at <span style='color:#31688E;'>Home</span> than in <span style='color:#440154;'>Hospital</span>",
-    title = "In 2020 and 2021 more cancer deaths happened at Home than in Hospital",
+    title = "In 2020 and 2021 more cancer deaths happened at <span style='color:#31688E;'>Home</span> than in <span style='color:#440154;'>Hospital</span>",
     caption = paste0(c(source_nrs_text), collapse = "\n")
@@
     panel.spacing = unit(1, units = "lines"),
-    legend.position = "top", legend.title = element_blank()
+    legend.position = "top", legend.title = element_blank(),
+    plot.title = element_textbox_simple()  # to render the html in the title
     ) +

### 2022-02-02 covid_nrs_place_of_death_data.Rmd
minor update: deleted unnecessary line about download dir

@@
 ```{r}
-download_dir <- "./downloaded_data"
-if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist
-
 files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
@@
 ```{r}
+# TODO: change this to also keep track of latest available year
+# TODO: modify above notes to include 2022 data also
+# TODO: 2020 data aren't included in the latest files, so I might need to keep them separate
+
 weeks_available_in_nrs_weekly_data_by_cause_of_death_2021 <-  # extract the maximum weeks available from the weekly-updated 2021 data

### 2022-02-02 covid_nrs_place_of_death_data.Rmd
started making changes to support 2022 data

@@
 
+ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")
+
 weekly_deaths_overall <-
-  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
+  # ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
+  ods_raw %>%
   make_human_readable_ref_area %>%
@@
   rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
-  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021, we only want the weekly figures
-  mutate(
-    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
-    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
+  filter(!ref_period %in% as.character(2020:2022)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
+  mutate(
+    year = parse_integer(str_extract(ref_period, pattern="^(202[0-3])")),
+    week_number = parse_integer(str_remove(ref_period, pattern = "^(202[0-3])\\-")),
     date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
     number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
   ) %>%

### 2022-02-02 covid_nrs_place_of_death_visualisations.Rmd
started making changes to support 2022 data

--- 2022-01-13 covid_nrs_place_of_death_visualisations.Rmd
@@
 ```{r, fig.width=12, fig.height=9, warning=FALSE}
+(
 figure_deaths_at_home <-
@@
   NULL
+)
 
@@
 save_output_file(filename = paste0(dir_outputs,"/figure_deaths_at_home"), extensions = c(".pdf",".svg",".png",".jpg"), plot = figure_deaths_at_home, width = 18, height = 15, units = "cm", dpi = 300)
-
-figure_deaths_at_home
 ```
@@
     y = "Percentage yearly deaths (%)",
-    # title = "More cancer deaths in 2020 and 2021 happened at <span style='color:#31688E;'>Home</span> than in <span style='color:#440154;'>Hospital</span>",
-    title = "In 2020 and 2021 more cancer deaths happened at Home than in Hospital",
-    caption = paste0(c(source_nrs_text), collapse = "\n")
+    title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
+    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",source_nrs_text), collapse = "\n")
   ) +

### 2022-02-02 download_latest_data_and_render_reports.R

@@ -27,10 +27,13 @@ all_nrs_links <-
   html_nodes("a") %>%
   html_attr("href")  # find all links on the page
 
-nrs_filename_pattern <- "covid-deaths-(21|22)-data.*xlsx"  # to accommodate 2022 data also!
+nrs_filename_pattern <- "covid-deaths-(21|22)-data.*xlsx"  # to accommodate 2022 data also once the filename changes!
 
 weekly_deaths_url_relative <- all_nrs_links[which(str_detect(all_nrs_links, pattern = nrs_filename_pattern))]  # keep only links that are probably weekly deaths
 weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
+
+## the format changed between 2021 and 2022 for weekly deaths data, and the 2022 data no longer include 2020 figures, so we need to load the 2020 data separately from the latest available data
+weekly_deaths_2020_and_2021_last_file <- "https://www.nrscotland.gov.uk/files/statistics/covid19/covid-deaths-21-data-week-52.xlsx"
 
 # Load table keeping track of latest data ---------------------------------
 
@@ -66,12 +69,13 @@ if (file.exists(file_historical_data_sources)) {
   table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
 } else {
   table_of_historical_data_sources <- tibble(
-    short_name = c("overall","sex_age","la","hb"),
+    short_name = c("overall","sex_age","la","hb","2020-2021"),
     url = c(
       "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
-      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx"
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-21-data-week-52.xlsx"  # last 2021 data which included 2020 also
     )
   ) %>% mutate(
     # filename = basename(url),
[…]

### 2022-02-03 covid_nrs_place_of_death_data_2022.Rmd
Started compiling updated script for parsing latest NRS data; luckily msot ofthe script pulls from the ad-hoc query files which haven't changed format, and there's only a small amount of change needed from hte previous script

---
title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2022 data"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd

## load functions
source("./functions.R")
```

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

# Dates represent event registration, not occurrence!

From NRS website:

> All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.

# Load info about latest data

A separate script checks for updated data and downloads it, also updating these files with the locations of the downloaded data:

```{r}
file_data_sources <- "./table_of_data_sources.csv"
file_historical_data_sources <- "./table_of_historical_data_sources.csv"
file_case_trends_data_sources <- "./table_of_case_trends_data_sources.csv"

table_of_data_sources <- read_csv(file = file_data_sources)
table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)

table_of_data_sources %>%
  knitr::kable(caption = "Latest data used in this report.")

files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
```

## Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
latest_data_modified_date <- strftime(max(table_of_data_sources$last_modified), format="%F")

dir_outputs <- paste0("./outputs/",latest_data_modified_date)
if (!dir.exists(dir_outputs)) dir.create(dir_outputs, recursive = TRUE)
```

# Import & wrangle data

There are several files that the data are imported from.

## Overall data

```{r}
ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")

weekly_deaths_overall <-
  # ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
  ods_raw %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
  filter(!ref_period %in% as.character(2020:2022)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(202[0-3])")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(202[0-3])\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-c(measure_type)) %>%  #  measure_type is count for all data
  recode_place_of_death() %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # not needed here probably, but no harm in including it
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing
```

## Sex & age

```{r}
weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
```

### Sex & age separate splits

```{r}
## compute weekly data for sex and age separately
## sum over sex or age, respectively
weekly_deaths_sex <-
  weekly_deaths_sex_age %>%
  filter(age!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")

weekly_deaths_age <-
  weekly_deaths_sex_age %>%
  filter(sex!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
```

## LA & HB

```{r}
weekly_deaths_la <- 
  read_excel(files_weekly_deaths$la, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = council, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  parse_year_and_week_number_from_ad_hoc_data %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(
      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
      fill = list(number_of_deaths = 0)
      ) %>%
  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
              quiet = FALSE)

weekly_deaths_hb <- 
  read_excel(files_weekly_deaths$hb, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  parse_year_and_week_number_from_ad_hoc_data %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(
      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
      fill = list(number_of_deaths = 0)
      ) %>%
  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
```

## Weekly Covid as contributory vs underlying cause of death

```{r}
weekly_covid_contributory_vs_underlying_cause <-
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  filter(
    cause_of_death %in% c("COVID-19 contributory factor", "COVID-19 underlying cause")
  ) %>%
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  clean_names() %>%
  filter(age!="all") %>%
  filter(sex!="all") %>%
  filter(place_of_death != "All") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, place_of_death) %>%  # sum over age & sex
  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
  ({function(x)
    bind_rows(
      x,
      x %>% 
        group_by(week_number, week_number_run_over, year, date_w_c) %>%  # sum across place of death
        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
        mutate(place_of_death = "All")
    )
  })

## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure
stopifnot(
  (weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
  filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
  nrow) == 0
)

annual_covid_contributory_vs_underlying_cause <-
  weekly_covid_contributory_vs_underlying_cause %>%
  group_by(year, place_of_death) %>%
  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
  ({function(x)
    bind_rows(
      x %>% mutate(year = as.character(year)),
      x %>% 
        group_by(place_of_death) %>%  # sum across eyars
        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
        mutate(year = "2020-2021")
    )
  }) %>%
  mutate(deaths_covid_related = covid_19_contributory_factor + covid_19_underlying_cause) %>%
  mutate(
    proportion_contributory = covid_19_contributory_factor / deaths_covid_related,
    proportion_underlying = covid_19_underlying_cause / deaths_covid_related
    )
```

## 'Main report' file

The format of the main report file has changed from what was used in 2021.

However, the main report file is still only being used for cause of death data, which simplifies parsing the file!

Causes of death are reported in `sheet 8` of the report, and include figure from 2021 and 2022. The worksheet contains data split by location of death, which are contained within different tables arranged vertically, so their boundaries need to be written out; the tables have been pre-filled for all of 2021 and 2022, so some proportion of entries will be blank for 2022.

Note that excess deaths and 5-year average deaths are computed differently between 2020-2021 and 2022:

> For data registered in 2021, any comparison to a five year average is being made with the 2015-2019 average. For 2022, the comparison is being made to the average of 2016, 2017, 2018, 2019, 2021.

> Excess deaths are calculated by comparing the current year to the five year average from previous years. NRS, along with ONS and NISRA, agreed to leave 2020 out of five year average calculations so the five-year average which is used to compare 2022 figures against covers the years 2016, 2017, 2018, 2019 and 2021. More information about this can be found in a paper published on our website (see link on the right). Moveable public holidays, when registration offices are closed, affect the number of registrations made in the published weeks and in the corresponding weeks in previous years.

### Weekly causes of death

```{r}
sheet_cause_of_death_data <- "8"

cell_ranges_for_causes_of_death_data <- list(
  "All" = "A6:V110",
  "Care home" = "A113:V217",
  "Home & other non-institution" = "A220:V324",
  "Hospital" = "A327:V431",
  "Other" = "A434:V538"
)

# TODO: complete cleaning of this bit
# TODO: canonical causes of death

# weekly_deaths_cause_of_death <-
weekly_cause_of_death_data <- 
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

### Historical causes of death

Historical here refers to the 2015-2019 period for 2020 & 2021, and 2016-2021 (skipping 2020) for 2022. There is another 'historical' dataset, which includes cause of death figures for 2020.

```{r}

```

### 2022-02-04 covid_nrs_place_of_death_data_2022.Rmd
added hsitorical figure calculations. phew!

@@
 files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
+files_historical_deaths <- set_names(x = as.list(table_of_historical_data_sources$path_latest), nm = table_of_historical_data_sources$short_name)
+files_case_counts <- set_names(x = as.list(table_of_case_trends_data_sources$path_latest), nm = table_of_case_trends_data_sources$short_name)
 ```
@@
 ```{r}
+## a separate raw file makes it easier to debug, and also reduces the number of requests I make while debugging!
 ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")
@@
 weekly_deaths_overall <-
-  # ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
   ods_raw %>%
@@
 
-# TODO: complete cleaning of this bit
-# TODO: canonical causes of death
-
-# weekly_deaths_cause_of_death <-
-weekly_cause_of_death_data <- 
+weekly_cause_of_death_2021_onwards <-
   map_dfr(
@@
 ```{r}
+historical_average_cause_of_death_2021_onwards <-
+  map_dfr(
+    .x = cell_ranges_for_causes_of_death_data,
+    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
+    .id = "place_of_death"
+  ) %>%
+  janitor::clean_names() %>%
+  rename(year = registration_year, date_w_c = week_beginning) %>%
+  select(place_of_death, year, week_number, date_w_c, matches("five_year_average")) %>%
+  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "deaths_mean") %>%
+  mutate(
+    year = as.integer(year),
+    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
+    cause_of_death = return_pretty_cause_of_death(cause_of_death),
+    cause_of_death = if_else(cause_of_death == "All causes", "All", cause_of_death),
+    date_w_c = as.Date(date_w_c),
+    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+    historic_period = historic_period_depending_on_year(year)
+  ) %>%
+  select(-name) %>%
+  drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future
+```
 
+### Data from 2020 & 2021
+
+The reports available from NRS in 2022 include data from 2021-2022, but we can also include data from 2020 for context. For this I'll load the last 'main report' file from 2021.
+
+The weekly causes of death data for 2020 is included in the sheet `Table 3 (2020)`.
+
+#### 'Main report' file
+
+```{r}
+nrs_weekly_data_sheet_2020 <- 
+  excel_sheets(files_historical_deaths$`2020-2021`)[which(str_detect(tolower(excel_sheets(files_historical_deaths$`2020-2021`)), pattern = "table 3.*2020"))]
+
+## construct column names for the data
+headers_in_nrs_data_2020 <-
+  c(
+    "place_of_death",
+    "cause_of_death",
+    paste0("week_", 1:53),
+    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
+    "remove_1",
+    "remove_2"
+  )
+
+## extract data... this is dense but I've tried to annotate it clearly!
+weekly_deaths_cause_of_death_2020 <-
+  readxl::read_excel(files_historical_deaths$`2020-2021`, sheet = nrs_weekly_data_sheet_2020, col_names = headers_in_nrs_data_2020, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
+  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
+  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
+  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
+  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
+  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
+  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
+  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "number_of_deaths", values_drop_na = TRUE) %>%  # convert to long data format
+  mutate(
+    week_number = as.integer(parse_number(week_number)),  # extract week number
+    place_of_death =  # find best match in the reference names used in other datasets
+      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)],  # the reference is in the variable order_of_place_of_death_levels
+    year = 2020,
+    dataset = str_remove_all(dataset , "Registered deaths\\: "),
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+    ) %>%
+  filter(dataset!="Difference") %>%
+  pivot_wider(names_from = dataset, values_from = number_of_deaths) %>%
+  rename(deaths_mean = `five year average`, number_of_deaths = `2020`)
 ```
+
+#### Merge 2020 data with current data
+
+Note: the 5-year average for 2022 (spanning 2016-2019 & 2021) is only published along with the running weekly counts of deaths by cause of death; it's just that NRS has decided to only publish them at the same time as new data comes out!
+This is a bit of a bummer - it means we can't populate the whole year of 'historical context' data for 2022.
+
+They could be calculated by hand if the 2015-2019 weekly data were available broken down by cause of death - as it stands, only the average 2015-2019 figure is available.
+
+```{r}
+weekly_cause_of_death_data <-
+  bind_rows(
+    weekly_cause_of_death_2021_onwards,
+    weekly_deaths_cause_of_death_2020 %>% 
+      select(-deaths_mean) %>%
+      filter(cause_of_death != "All")  # don't need all cause data in causes breadkown!
+  )
+
+historical_average_cause_of_death <-
+  bind_rows(
+    historical_average_cause_of_death_2021_onwards,
+    weekly_deaths_cause_of_death_2020 %>% 
+      select(-number_of_deaths) %>% 
+      mutate(historic_period = historic_period_depending_on_year(year))
+  )
+```
+
+## Historical weekly deaths
+
+The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data. These are downloaded from the ad-hoc section also.
+
+```{r, warning=FALSE}
+weekly_deaths_2015_2019_overall <- 
+  read_csv(files_historical_deaths$overall, skip = 2) %>%
+  rename(year = 1) %>%  # rename first column
+  mutate(
+    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
+  ) %>%
+  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
+  drop_na(`1`) %>%
+  pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
+  mutate(week_number=as.integer(week_number)) %>%
+  recode_place_of_death()
+
+weekly_deaths_2015_2019_hb <- 
+  readxl::read_excel(files_historical_deaths$hb, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  clean_names() %>%
+  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
+  rename(  # same format as current data
+    week_number = matches("week"),
+    place_of_death = matches("location"),
+    ref_area = health_board
+  ) %>%
+  mutate(
+    week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
+
+weekly_deaths_2015_2019_la <-
+  readxl::read_excel(files_historical_deaths$la, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  clean_names() %>%
+  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
+  rename(  # same format as current data
+    week_number = matches("week"),
+    place_of_death = matches("location"),
+    ref_area = council_area
+  ) %>%
+  mutate(
+    week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
+
+weekly_deaths_2015_2019_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
+  readxl::read_excel(files_historical_deaths$sex_age, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
+  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
+  mutate(
+    year = 2000L + as.integer(year),
+    week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()
+
+## compute weekly data for sex and age separately
+## sum over sex or age, respectively
+weekly_deaths_2015_2019_sex <-
+  weekly_deaths_2015_2019_sex_age %>%
+  group_by_all %>%
+  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
+
+weekly_deaths_2015_2019_age <-
+  weekly_deaths_2015_2019_sex_age %>%
+  group_by_all %>%
+  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
+```
+
+## Map data
+
+I'm using `sf` to read these and to include them invisualisations.
+
+* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-december-2020-uk-buc/
+* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
+  - Note: this is a fairly high resolution, don't need this much detail!
+  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)
+
+```{r}
+shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
+shapefile_la <- st_read(dsn = "./map_data/la")
+```
+
+## COVID case counts by LA/HB
+
+```{r}
+cases_hb <- read_csv(
+  files_case_counts$hb, 
+  col_types = 
+    cols(
+    .default = col_double(),
+    `_id` = col_skip(),
+    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
+    Date = col_date(format = "%Y%m%d"),
+    HB = col_character(),
+    HBName = col_character(),
+    HospitalAdmissionsQF = col_character(),
+    ICUAdmissionsQF = col_character()
+    ),
+  ) %>% clean_names() %>%
+  rename(ref_area = hb_name) %>%
+  mutate(
+    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
+    ref_area = phsmethods::match_area(hb)  # convert to standard HB names
+    )
+
+cases_la <- read_csv(
+  file = files_case_counts$la, 
+  col_types = 
+    cols(
+    .default = col_double(),
+    `_id` = col_skip(),
+    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
+    Date = col_date(format = "%Y%m%d"),
+    CA = col_character(),
+    CAName = col_character()
+  )
+) %>% clean_names() %>%
+  rename(ref_area = ca_name) %>%
+  mutate(
+    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
+    ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
+    )
+```
+
+# Calculate case counts: weekly counts to match weekly death reporting
+
+Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks.
+This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
+
+However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!
+
+Note also that the `crude_rate_positive_per100k` represents the cumulative rate!
+
+```{r}
+weekly_covid_cases_hb <-
+  cases_hb %>%
+  group_by(year, week_number, ref_area) %>%
+  summarise(
+    weekly_positive = sum(daily_positive),
+    cumulative_positive = max(cumulative_positive),
+    crude_rate_positive_per100k = max(crude_rate_positive),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
+    )
+
+## extract Scotland-wide data and remove from HB split
+weekly_covid_cases_overall <- weekly_covid_cases_hb %>% filter(ref_area=="Scotland")
+weekly_covid_cases_hb <- weekly_covid_cases_hb %>% filter(ref_area!="Scotland")
+
+weekly_covid_cases_la <-
+  cases_la %>%
+  group_by(year, week_number, ref_area) %>%
+  summarise(
+    weekly_positive = sum(daily_positive),
+    cumulative_positive = max(cumulative_positive),
+    crude_rate_positive_per100k = max(crude_rate_positive),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
+    )
+```
+
+# Historical averages, minima & maxima
+
+Due to the counting of weeks, most years have 52 weeks with some years (2015 & 2020) having 53.
+Therefore we need to decide how to deal with week 53 in 2020 since only 2015 has a corresponding weekly count.
+The solution used here is to calculate a weekly rate for week 53 from week 53 in 2015, and week 52 in 2016-2019 - this represents the "last week of the year", which includes December 31st.
+
+Note also that since 2022, the 5-year historical period has been defined as 2016-2021, with 2020 excluded, so we need to calculate two weekly averages!
+
+```{r, warning=FALSE}
+historical_weekly_average_overall <-
+  bind_rows(
+    ## 2015-2019 period
+    weekly_deaths_2015_2019_overall %>%
+    filter(year!="average") %>%  # remove averages
+    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
+    filter(!all(number_of_deaths==0 & week_number==53)) %>%
+    ungroup %>%
+    create_week_53_data_for_years_with_52_weeks %>%
+    group_by(week_number,place_of_death) %>% 
+    summarise(
+      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
+      .groups = "drop"
+    ) %>% 
+    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
+    mutate(historic_period = historic_period_depending_on_year(2020))
+    ,
+    ## 2016-2021, excluding 2020
+    bind_rows(
+      ## 2016-2019
+      weekly_deaths_2015_2019_overall %>%
+        mutate(year = as.integer(year)) %>%
+        filter(year %in% 2016:2019),
+      ## 2021
+      weekly_deaths_overall %>% 
+        rename(number_of_deaths = deaths_all_causes) %>%
+        filter(year == 2021 & age == "all" & sex == "all")
+      ) %>%
+    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
+    filter(!all(number_of_deaths==0 & week_number==53)) %>%
+    ungroup %>%
+    create_week_53_data_for_years_with_52_weeks %>%
+    group_by(week_number,place_of_death) %>% 
+    summarise(
+      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
+      .groups = "drop"
+    ) %>% 
+    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
+    mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+
+week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
+
+# TODO: weekly averages may be too detailed for this level od subcategorising!
+
+deaths_mean_min_max <- function(data_tbl) {
+  data_tbl %>%
+    summarise(
+      deaths_mean = mean(number_of_deaths),
+      deaths_min = min(number_of_deaths),
+      deaths_max = max(number_of_deaths),
+      .groups = "drop"
+    )
+}
+
+make_2016_to_2019_and_2021_data <- function(historical_data, current_data) {
+  bind_rows(
+      historical_data %>% filter(year %in% 2016:2019),
+      current_data %>% filter(year == 2021) %>% rename(number_of_deaths = deaths_all_causes)
+    )
+}
+
+historical_weekly_average_hb <-
+  bind_rows(
+    ## 2015-2019
+    weekly_deaths_2015_2019_hb %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, ref_area
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2020))
+    ,
+    ## 2016-2019, 2021
+    make_2016_to_2019_and_2021_data(
+      historical_data = weekly_deaths_2015_2019_hb, 
+      current_data = weekly_deaths_hb
+      ) %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, ref_area
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+
+historical_weekly_average_la <-
+  bind_rows(
+    ## 2015-2019
+    weekly_deaths_2015_2019_la %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, ref_area
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2020))
+    ,
+    ## 2016-2019, 2021
+    make_2016_to_2019_and_2021_data(
+      historical_data = weekly_deaths_2015_2019_la, 
+      current_data = weekly_deaths_la
+      ) %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, ref_area
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+
+historical_weekly_average_sex_age <-
+  bind_rows(
+    ## 2015-2019
+    weekly_deaths_2015_2019_sex_age %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, age, sex
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2020))
+    ,
+    ## 2016-2019, 2021
+    make_2016_to_2019_and_2021_data(
+      historical_data = weekly_deaths_2015_2019_sex_age, 
+      current_data = weekly_deaths_sex_age
+      ) %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, age, sex
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+
+historical_weekly_average_age <-
+  bind_rows(
+    ## 2015-2019
+    weekly_deaths_2015_2019_age %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, age
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2020))
+    ,
+    ## 2016-2019, 2021
+    make_2016_to_2019_and_2021_data(
+      historical_data = weekly_deaths_2015_2019_age, 
+      current_data = weekly_deaths_age
+      ) %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, age
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+
+historical_weekly_average_sex <-
+  bind_rows(
+    ## 2015-2019
+    weekly_deaths_2015_2019_sex %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, sex
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2020))
+    ,
+    ## 2016-2019, 2021
+    make_2016_to_2019_and_2021_data(
+      historical_data = weekly_deaths_2015_2019_sex, 
+      current_data = weekly_deaths_sex
+      ) %>%
+    create_week_53_data_for_years_with_52_weeks() %>%
+    group_by(
+      week_number, place_of_death, sex
+    ) %>%
+    deaths_mean_min_max %>%
+    mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+
+# TODO: include the split between historical periods here also
+compute_historical_annual_average_deaths <- function(data_tbl, ...) {
+  grouping_vars <- enquos(...)
+  data_tbl %>%
+    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
+    summarise(  # calculate annual totals first...
+    number_of_deaths = sum(number_of_deaths), 
+    .groups = "keep"
+  ) %>%
+  ungroup(year) %>%  # remove year as grouping to average over year now
+  summarise(
+    annual_deaths_mean = mean(number_of_deaths),
+    annual_deaths_min = min(number_of_deaths),
+    annual_deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+}
+
+historical_annual_average_overall <-
+  bind_rows(
+      ## 2016-2019
+      weekly_deaths_2015_2019_overall %>%
+        mutate(year = as.integer(year)) %>%
+        filter(year %in% 2016:2019) %>%
+        mutate(historic_period = historic_period_depending_on_year(2020)),
+      ## 2021
+      weekly_deaths_overall %>% 
+        rename(number_of_deaths = deaths_all_causes) %>%
+        filter(year == 2021 & age == "all" & sex == "all") %>%
+        mutate(historic_period = historic_period_depending_on_year(2022))
+      ) %>%
+  compute_historical_annual_average_deaths()
+
+make_2016_to_2019_and_2021_data_with_historic_periods <- function (historical_data, current_data) {
+  bind_rows(
+    historical_data %>% 
+      filter(year %in% 2016:2019) %>%
+      mutate(historic_period = historic_period_depending_on_year(2020)),
+    current_data %>% 
+      filter(year == 2021) %>% 
+      rename(number_of_deaths = deaths_all_causes) %>%
+      mutate(historic_period = historic_period_depending_on_year(2022))
+  )
+}
+
+historical_annual_average_hb <-
+  make_2016_to_2019_and_2021_data_with_historic_periods(
+    historical_data = weekly_deaths_2015_2019_hb, 
+    current_data = weekly_deaths_hb
+      ) %>%
+  compute_historical_annual_average_deaths(ref_area)
+
+historical_annual_average_la <-
+  make_2016_to_2019_and_2021_data_with_historic_periods(
+    historical_data = weekly_deaths_2015_2019_la, 
+    current_data = weekly_deaths_la
+      ) %>%
+  compute_historical_annual_average_deaths(ref_area)
+
+historical_annual_average_sex_age <-
+  make_2016_to_2019_and_2021_data_with_historic_periods(
+    historical_data = weekly_deaths_2015_2019_sex_age, 
+    current_data = weekly_deaths_sex_age
+      ) %>%
+  compute_historical_annual_average_deaths(sex, age)
+
+historical_annual_average_sex <-
+  make_2016_to_2019_and_2021_data_with_historic_periods(
+    historical_data = weekly_deaths_2015_2019_sex, 
+    current_data = weekly_deaths_sex
+      ) %>%
+  compute_historical_annual_average_deaths(sex)
+
+historical_annual_average_age <-
+  make_2016_to_2019_and_2021_data_with_historic_periods(
+    historical_data = weekly_deaths_2015_2019_age, 
+    current_data = weekly_deaths_age
+      ) %>%
+  compute_historical_annual_average_deaths(age)
+```
+
+## Historical annual average by cause
+
+I can approximate this by taking the sum of the average weekly deaths for `2015-2019`; for the `2016-2019 + 2021` period this isn't readily available because NRS only publish the weekly averages alongside the weekly updated data, but not for the entire year.
+Therefore I'll just calculate the `2015-2019` period average for now, and a compromise `2015-2019 + 2021` average:
+
+$$\frac{5}{6}average\_deaths_{2015-2019} + \frac{1}{6}deaths_{2021}$$
+
+```{r}
+# TODO: incorporate the 2016-2019 + 2021 period
+historical_annual_average_cause_of_death_2015_2019 <-
+  historical_average_cause_of_death %>%
+  filter(historic_period == "2015-2019" & year == 2021) %>%  # this is slightly strange but some of the averages disagree, probably because the 2015-2019 data were updated in 2020, and the 2021 data were published with modified figures.
+NRS deaths data are truly a living document!
+  group_by(place_of_death, cause_of_death, historic_period) %>%
+  summarise(
+    annual_deaths_mean = sum(deaths_mean),
+    .groups = "drop"
+  )
+
+deaths_by_cause_2021 <-
+  weekly_cause_of_death_2021_onwards %>% 
+  filter(year == 2021) %>% 
+  group_by(place_of_death, cause_of_death) %>% 
+  summarise(deaths_2021 = sum(number_of_deaths), .groups = "drop")
+
+historical_annual_average_cause_of_death <-
+  bind_rows(
+    historical_annual_average_cause_of_death_2015_2019
+    ,
+    ## compromise figures for 2015-2019 + 2021
+    historical_annual_average_cause_of_death_2015_2019 %>%
+    filter(cause_of_death != "All") %>%
+    left_join(
+      deaths_by_cause_2021, by = c("place_of_death", "cause_of_death")
+    ) %>%
+    mutate(
+      annual_deaths_mean = (5/6 * annual_deaths_mean) + (1/6 * deaths_2021),
+      deaths_2021 = NULL,
+      historic_period = "2015-2019, 2021"
+    )
+  )
+```
+
+## Check weekly historic averages against NRS computed data
+
+```{r}
+weekly_deaths_overall %>%
+  filter(sex=="all" & age=="all") %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%
+  left_join(historical_weekly_average_overall) %>%
+  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean, historic_period) %>%
+  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
+  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e.
+all other entries had the same figures!).
+Week 53 was computed as the average of deaths in week 53 in 2015 and weeks 52 in 2016-2019.
+For all the other datasets the mean deaths computed agree!")
+```

### 2022-02-04 covid_nrs_place_of_death_data_2022.Rmd
nearly there with the data munging! can start visualising stuff soon I hope!

@@
 ```{r}
-weekly_cause_of_death_data <-
+weekly_deaths_cause_of_death <-
   bind_rows(
@@
 ```
+
+# Merge weekly deaths with historic death rates
+
+```{r}
+## compute rolling average
+merged_deaths_overall <-
+  weekly_deaths_overall %>%
+  filter(sex=="all" & age=="all") %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(historical_weekly_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
+  by = c("place_of_death", "week_number", "sex", "age", "historic_period"))
+
+merged_deaths_hb <-
+  weekly_deaths_hb %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(historical_weekly_average_hb, by = c("week_number","ref_area","place_of_death", "historic_period"))
+
+merged_deaths_la <-
+  weekly_deaths_la %>%  # use data with weekly movingincluded
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(historical_weekly_average_la, by = c("week_number","ref_area","place_of_death", "historic_period"))
+
+merged_deaths_sex_age <-
+  weekly_deaths_sex_age %>%  # use data with weekly movingincluded
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age", "historic_period"))
+
+merged_deaths_sex <-
+  weekly_deaths_sex %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex", "historic_period"))
+
+merged_deaths_age <-
+  weekly_deaths_age %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age", "historic_period"))
+
+merged_deaths_cause_of_death <-
+  weekly_deaths_cause_of_death %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+  inner_join(
+    historical_average_cause_of_death, by = c("place_of_death", "week_number", "cause_of_death", "historic_period", "year", "date_w_c")
+  )
+```
+
+## Weekly deaths including past weekly deaths
+
+For this dataset, we simply `bind_rows` weekly deaths by year and place of death with past weekly deaths by year and location.
+Note that past data for years with 52 weeks includes an entry of 0 deaths for week 53, so we'll remove them, otherwise the date computation is thrown off!
+
+```{r}
+weekly_deaths_current_and_past <-
+  bind_rows(
+    weekly_deaths_overall %>%
+    filter(sex=="all" & age=="all" & place_of_death!="All") %>%  # keep only aggregated values over sex & age
+    select(year, week_number, week_number_run_over, place_of_death, matches("deaths"), -deaths_nrs_past_average_all_causes)
+    ,
+    past_deaths_overall %>%
+      rename(deaths_all_causes = number_of_deaths) %>%
+      filter(place_of_death!="All" & year!="average") %>%  # remove aggregated entries
+      mutate(year = as.integer(year)) %>%
+      filter(!(year %in% 2016:2019 & week_number==53))  # remove week 53 entries from years 2016-2019
+  ) %>%
+  mutate(date_w_c = compute_start_date_from_week_number(week_number = week_number, year_number = year))
+```
+
+# Annual deaths for 2020 & 2021
+
+For visualising data using maps, we need to compute data for the entire year.
+Note: the non-covid and covid-related deaths in 2020 don't add up to all cause deaths due to the covid status only starting in week 12!
+
+```{r}
+# TODO: check through these to see if they compute correctly!
+
+compute_total_deaths_in_year <- function(data_tbl, ...) {
+  grouping_vars <- enquos(...)
+  data_tbl %>%
+    group_by(year, place_of_death, !!!grouping_vars) %>%
+    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%  # compute sums of all types of death
+    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)  # compute a smaller number of all-cause deaths since covid started (over 2020, covid- and non-covid-related deaths don't add up to all-cause deaths because pre-covid deaths are included also!)
+}
+
+compute_ratio_of_annual_deaths_to_historical <- function(data_tbl) {
+  data_tbl %>%
+    mutate(
+      ratio_annual_deaths_to_historical_mean = deaths_all_causes / annual_deaths_mean,
+      ratio_annual_deaths_to_historical_min = deaths_all_causes / annual_deaths_min,
+      ratio_annual_deaths_to_historical_max = deaths_all_causes / annual_deaths_max
+    )
+}
+
+total_deaths_la <-
+  merged_deaths_la %>%
+  compute_total_deaths_in_year(ref_area) %>%
+  left_join(historical_annual_average_la, by = c("place_of_death", "ref_area")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_hb <-
+  merged_deaths_hb %>%
+  compute_total_deaths_in_year(ref_area) %>%
+  left_join(historical_annual_average_hb, by = c("place_of_death", "ref_area")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_sex <-
+  merged_deaths_sex %>%
+  compute_total_deaths_in_year(sex) %>%
+  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_age <-
+  merged_deaths_age %>%
+  compute_total_deaths_in_year(age) %>%
+  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+total_deaths_overall <-
+  merged_deaths_overall %>%
+  compute_total_deaths_in_year() %>%
+  left_join(historical_annual_average_overall, by = c("place_of_death")) %>%
+  compute_ratio_of_annual_deaths_to_historical
+
+## bespoke computation because format is different... deaths split by cause!
+total_deaths_cause_of_death <-
+  merged_deaths_cause_of_death %>%
+  group_by(year, place_of_death, cause_of_death) %>%
+  summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
+  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death")) %>%
+  mutate(
+    ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
+  )
+
+current_and_historical_deaths_by_cause_and_location <-
+  bind_rows(
+    ## 2020 & 2021 separately
+    total_deaths_cause_of_death %>%
+      select(year, place_of_death, cause_of_death, deaths) %>%
+      mutate(year = as.character(year)),
+    ## 2020 & 2021 averaged
+    total_deaths_cause_of_death %>%
+      select(year, place_of_death, cause_of_death, deaths) %>%
+      group_by(place_of_death,  cause_of_death) %>%
+      summarise(deaths = sum(deaths), year = "2020-2021", .groups = "drop"),
+    ## historical data
+    total_deaths_cause_of_death %>%
+      select(year, place_of_death, cause_of_death, deaths = annual_deaths_mean) %>%
+      mutate(year = "2015-2019") %>%
+      distinct
+  ) %>%
+  filter(place_of_death != "All") %>%
+  group_by(year, cause_of_death) %>%
+  mutate(proportion = deaths / sum(deaths)) %>%
+  ungroup %>%
+  arrange(year, cause_of_death)
+
+```

### 2022-02-11 covid_nrs_place_of_death_data_2022.Rmd
I have a strange feeling I've pushed this file before...

@@
 
+## helper function to remove rows of data where the historic period is inappropriate
+remove_invalid_historic_period_for_year <- function(data_tbl) {
+  data_tbl %>%
+    filter(!(historic_period == "2016-2019, 2021" & year < 2022))
+}
+
+total_deaths_overall <-
+  merged_deaths_overall %>%
+  compute_total_deaths_in_year() %>%
+  left_join(historical_annual_average_overall, by = c("place_of_death")) %>%
+  compute_ratio_of_annual_deaths_to_historical %>%
+  remove_invalid_historic_period_for_year()
+
 total_deaths_la <-
@@
   left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
-  compute_ratio_of_annual_deaths_to_historical
-
-total_deaths_overall <-
-  merged_deaths_overall %>%
-  compute_total_deaths_in_year() %>%
-  left_join(historical_annual_average_overall, by = c("place_of_death")) %>%
   compute_ratio_of_annual_deaths_to_historical

### 2022-02-11 covid_nrs_place_of_death_data_2022.Rmd
Started making visualisations using new data

@@
 latest_data_modified_date <- strftime(max(table_of_data_sources$last_modified), format="%F")
-
-dir_outputs <- paste0("./outputs/",latest_data_modified_date)
+latest_year <- year(latest_data_modified_date)
+
+dir_outputs <- file.path("./outputs", latest_year, latest_data_modified_date)
 if (!dir.exists(dir_outputs)) dir.create(dir_outputs, recursive = TRUE)
@@
     ,
-    past_deaths_overall %>%
+    weekly_deaths_2015_2019_overall %>%
       rename(deaths_all_causes = number_of_deaths) %>%
@@
 ```{r}
-# TODO: check through these to see if they compute correctly!
-
 compute_total_deaths_in_year <- function(data_tbl, ...) {
@@
   left_join(historical_annual_average_la, by = c("place_of_death", "ref_area")) %>%
-  compute_ratio_of_annual_deaths_to_historical
+  compute_ratio_of_annual_deaths_to_historical %>%
+  remove_invalid_historic_period_for_year()
 
@@
   left_join(historical_annual_average_hb, by = c("place_of_death", "ref_area")) %>%
-  compute_ratio_of_annual_deaths_to_historical
+  compute_ratio_of_annual_deaths_to_historical %>%
+  remove_invalid_historic_period_for_year()
 
@@
   left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
-  compute_ratio_of_annual_deaths_to_historical
+  compute_ratio_of_annual_deaths_to_historical %>%
+  remove_invalid_historic_period_for_year()
 
@@
   left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
-  compute_ratio_of_annual_deaths_to_historical
+  compute_ratio_of_annual_deaths_to_historical %>%
+  remove_invalid_historic_period_for_year()
+```
+
+## Annual deaths by cause of death
+
+```{r}
+# TODO: work out what to do about this - is it even used?
 
 ## bespoke computation because format is different... deaths split by cause!
+## Note: this isn't appropriate for 2022 yet because we don't have the annual death figure yet
 total_deaths_cause_of_death <-
   merged_deaths_cause_of_death %>%
-  group_by(year, place_of_death, cause_of_death) %>%
+  group_by(year, place_of_death, cause_of_death, historic_period) %>%
   summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
-  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death")) %>%
+  left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death", "historic_period")) %>%
   mutate(
     ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
-  )
+  ) %>%
+  filter(year != 2022)
 
@@
   arrange(year, cause_of_death)
-
-```
+```
+
+## Historical average weekly deaths, including for incomplete year 2022
+
+```{r}
+merged_deaths_hb %>%
+  group_by(year, ref_area, place_of_death, historic_period) %>%
+  summarise(
+    deaths_all_causes = sum(deaths_all_causes, na.rm = TRUE),
+    deaths_mean = sum(deaths_mean, na.rm = TRUE),
+    .groups = "drop"
+  ) %>%
+  mutate(proportion_home_deaths_increase = (deaths_all_causes/deaths_mean) - 1)
+
+# TODO: fix below to include historic_period
+# proportions_of_historical_deaths_by_place <-
+#   historical_weekly_average_overall %>% 
+#   group_by(year, place_of_death) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%
+#   filter(place_of_death != "All" & !is.na(as.integer(year))) %>%
+#   group_by(year) %>%
+#   mutate(
+#     total_deaths = sum(number_of_deaths)
+#   ) %>%
+#   ungroup %>%
+#   mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))
+# 
+# proportions_of_current_deaths_by_place <-
+#   total_deaths_overall %>%
+#   select(year, place_of_death, number_of_deaths = deaths_all_causes) %>%
+#   filter(place_of_death != "All") %>%
+#   group_by(year) %>%
+#   mutate(
+#     total_deaths = sum(number_of_deaths)
+#   ) %>%
+#   ungroup %>%
+#   mutate(proportion_of_total = number_of_deaths / total_deaths)
+# 
+# 
+# merged_proportions_of_deaths_by_place <-
+#   bind_rows(
+#     proportions_of_historical_deaths_by_place,
+#     proportions_of_current_deaths_by_place
+#   )
+```
+
+# Constants about latest data available
+
+These are used for saying in plots what data were included e.g. weeks 1-12 in 2022.
+
+```{r}
+## constants for use in graphs & annotations
+weeks_available_2022 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2022])
+most_recent_date_available_2022 <- max(merged_deaths_age$date_w_c)
+
+earliest_date <- min(merged_deaths_age$date_w_c)
+```
+
+# Ordering geographical data by population size
+
+The latest population estimates by HB/LA are available [from NRS Scotland here](https://www.opendata.nhs.scot/dataset/population-estimates).
+
+I'm using the estimates for 2020 (based on 2019 data) to compute the relative sizes of health boards!
+
+```{r}
+population_estimate_la_2019 <- read_csv(file = "./population_estimates/ca2019_pop_est_29062021.csv") %>%
+  select(Year, CA, Sex, AllAges) %>%
+  clean_names() %>%
+  make_human_readable_ref_area(ref_area = ca) %>%
+  filter(year==2020) %>%
+  filter(ca != "Scotland") %>%
+  filter(sex == "All") %>%
+  rename(la = ca)
+
+population_estimate_hb_2019 <- read_csv(file = "./population_estimates/hb2019_pop_est_29062021.csv") %>%
+  select(Year, HB, Sex, AllAges) %>%
+  clean_names() %>%
+  make_human_readable_ref_area(ref_area = hb) %>%
+  filter(year==2020) %>%
+  filter(hb != "Scotland") %>%
+  filter(sex == "All")
+```
+
+## Compute order
+
+```{r}
+order_la_by_population <-
+  population_estimate_la_2019 %>%
+  arrange(desc(all_ages)) %>%
+  .$la
+
+order_hb_by_population <-
+  population_estimate_hb_2019 %>%
+  arrange(desc(all_ages)) %>%
+  .$hb
+```
+
+# Comparisons of annual rates with current rates
+
+```{r}
+table_current_deaths_compared_to_past <-
+  bind_rows(
+    ## all available data
+    merged_deaths_overall %>%
+      filter(age == "all" &
+               sex == "all"
+             ) %>% 
+      group_by(year, place_of_death, week_number) %>%
+      summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        # deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+      left_join(
+        weekly_deaths_2015_2019_overall %>% filter(year != "average") %>% group_by(week_number, place_of_death) %>% summarise(
+                                           average_weekly_deaths = sum(number_of_deaths) / 5,
+                                           .groups = "drop"
+                                         ),
+        by = c("week_number", "place_of_death")
+      ) %>% 
+      group_by(place_of_death) %>%
+      summarise(
+        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2022}, 2022"),
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+        .groups = "drop"
+      )
+    ,
+    ## 2020 only
+    merged_deaths_overall %>%
+      filter(age == "all" &
+               sex == "all" &
+               year == 2020
+             ) %>% 
+      group_by(place_of_death) %>%
+      summarise(
+        period = "2020",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_overall %>% 
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+      by = "place_of_death"
+    )
+    ,
+    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
+    merged_deaths_overall %>%
+      filter(
+        age == "all" &
+          sex == "all" &
+          year == 2020 & week_number >= 12
+      ) %>%
+      group_by(place_of_death) %>%
+      summarise(
+        period = "2020, week 12 - week 53",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      weekly_deaths_2015_2019_overall %>% 
+        filter(week_number >= 12 & year != "average") %>%
+        group_by(place_of_death) %>%
+        summarise(mean_annual_deaths_2015_2019 = sum(number_of_deaths) / 5),
+      by = "place_of_death"
+    )
+    ,
+    ## 2021 only
+    merged_deaths_overall %>%
+      filter(age == "all" &
+               sex == "all" &
+               year == 2021
+             ) %>% 
+      group_by(place_of_death) %>%
+      summarise(
+        period = "2021",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_overall %>% 
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
+      by = "place_of_death"
+    )
+  ) %>%
+  mutate(
+    ratio_deaths_all_causes_to_historical = deaths_all_causes / mean_annual_deaths_2015_2019,
+    ratio_non_covid_deaths_to_historical = deaths_non_covid / mean_annual_deaths_2015_2019,
+    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
+  )
+
+write_csv(x = table_current_deaths_compared_to_past, file = file.path(dir_outputs, "annual_death_rates_comparison_with_historical_deaths.csv"))
+
+table_current_deaths_compared_to_past %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
+```
+
+# Moving averages: 4 weeks
+
+```{r}
+merged_deaths_overall <-
+  compute_moving_average_for_current_data(
+    current_data = merged_deaths_overall, 
+    historical_data = 
+      weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
+    num_weeks_to_average = 4)
+
+merged_deaths_hb <-
+  compute_moving_average_for_current_data(
+    current_data = merged_deaths_hb, 
+    historical_data = 
+      weekly_deaths_2015_2019_hb, 
+    num_weeks_to_average = 4)
+
+merged_deaths_la <-
+  compute_moving_average_for_current_data(
+    current_data = merged_deaths_la, 
+    historical_data = 
+      weekly_deaths_2015_2019_la, 
+    num_weeks_to_average = 4)
+
+merged_deaths_sex <-
+  compute_moving_average_for_current_data(
+    current_data = merged_deaths_sex, 
+    historical_data = 
+      weekly_deaths_2015_2019_sex, 
+    num_weeks_to_average = 4)
+
+merged_deaths_age <-
+  compute_moving_average_for_current_data(
+    current_data = merged_deaths_age, 
+    historical_data = 
+      weekly_deaths_2015_2019_age, 
+    num_weeks_to_average = 4)
+
+merged_deaths_sex_age <-
+  compute_moving_average_for_current_data(
+    current_data = merged_deaths_sex_age, 
+    historical_data = 
+      weekly_deaths_2015_2019_sex_age, 
+    num_weeks_to_average = 4)
+```
+
+# TODO: other thigns like excess death & historical rates by geography
+
+```{r}
+# TODO: copy some of the old calculations over
+```
+
+# Save most recent workspace for further analyses
+
+```{r}
+# save.image(file = "./workspace.RData", safe = TRUE)  # this is calling .GlobalEnv, which is empty
+save(list = ls(all.names = TRUE), file = "./workspace.RData", envir = environment())
+```
+
+# Print session info
+
+```{r}
+sessionInfo()
+```

### 2022-02-11 covid_nrs_place_of_death_visualisations_2022.Rmd
Started making visualisations using new data

---
title: "Place of death in NRS COVID-19 deaths: visualisation for 2022 data"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text
```

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

# Dates represent event registration, not occurrence!

From NRS website:

  > All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.

# Load most recent data

```{r}
load(file = "./workspace.RData")
```

# Visualisations

Note: visualisations have been moved to a separate .Rmd file

## Common elements

```{r}
loadfonts()  # load imported fonts
# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))

date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2021, '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
only_2022_text <- glue::glue("Data for 2022 only include weeks 1-{format(weeks_available_2022)} (w/c {format(most_recent_date_available_2022)}).")

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

death_registrations_text <- "Note: all figures are based on date of registration."

vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."

five_year_historical_mean <- "For 2020-2021 data, the 5-year historical period is 2015-2019; for 2022, the period is 2016-2019 + 2021, skipping 2020."

vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"

# removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death.")
```

## Function to save files to multiple formats

Saving to multiple formats is easy enough - the trouble is with saving PDF files with custom fonts, since they need to be embedded in the .pdf file. This requires:

  * telling R where to load the font from, using `extrafont::font_import()` - this was done in a previous step
* registering fonts with R using `extrafont::loadfonts()` - note: this isn't strictly necessary, since once imported, `extrafont` keeps track of fonts to load when the `extrafont` package is loaded
* ~~telling R where the `gs` executable is, i.e. the installation of ghostscript on the current machine~~
* ~~embedding the fonts in the `.pdf` files once they've been written, using `extrafont::embed_fonts()`~~
  * instead of manually embedding fonts (which resulted in garbled text on my setup!), just use `device=device_cairo` as an argument to `ggsave()` when saving `.pdf` files and it just works

```{r}
save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {

  walk(
    .x = extensions,
    .f = function(ext) {
      # if (tolower(ext)==".pdf") device <- cairo_pdf
      ggsave(filename = paste0(filename, ext), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (tolower(ext)==".pdf") cairo_pdf else NULL)
    }
  )
}
```

# Place of death over time

```{r}
## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
annotations_figure_place_of_death_explaining_historic_range <- tribble(
   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
  "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
  "curve", "Hospital", NA, 29, 830, 32, 550,
  "text", "Hospital", "Historic mean", 22, 800, NA, NA,
  "curve", "Hospital", NA, 22, 780, 26, 510
) %>% 
  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)

figure_place_of_death <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
  geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
  NULL

save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = figure_place_of_death, width = 28, height = 14, dpi = 300, units = "cm")
```

# Home deaths by HB over time: 2020-2022, 4-week average

```{r}
figure_home_deaths_over_time_by_hb_4_week_average <-
  merged_deaths_hb %>%
  filter(place_of_death == "Home & other non-institution") %>%
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  pivot_longer(cols = c(ma4w_deaths_all_causes, ma4w_deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max), fill = "#4477aa", alpha = 0.5) +
  # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = ma4w_deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  # scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~ref_area, scales = "free_y") +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  labs(x = NULL,
       y = "Deaths (average of past 4 weeks)",
       title = "Deaths at Home*: 4-week average by NHS Health Board",
       subtitle = date_range_text,
       caption = paste0(c(
         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text
         ), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x=ymd("2021-01-07"), y=0, label = "2021", hjust = 0, vjust = -1, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=0, label = "'22", hjust = 0, vjust = -1, colour = "grey40") +
  NULL

save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = figure_home_deaths_over_time_by_hb_4_week_average, width = 30, height = 20, dpi = 300, units = "cm")
```

# Geographical variation over time: cumulative deaths at home over time

For this visualisation, we'll compare home deaths over time (e.g. years 2020, 2021, and 2022) with their historic period annual average, at HB-level. For this, we need to calculate the 2022 rate to be comparable to the complete data for years 2020 & 2021 - this is achieved by taking the sum of deaths in 2022 and dividing by the sum of the weekly average deaths in its historic period

```{r}
merged_deaths_hb %>%
  # filter(week_number %in% 1:4) %>%
  group_by(year, ref_area, place_of_death, historic_period) %>%
  summarise(
    deaths_all_causes = sum(deaths_all_causes, na.rm = TRUE),
    deaths_mean = sum(deaths_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ratio_annual_deaths_to_historical_mean = deaths_all_causes/deaths_mean,
    percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1
    ) %>%
  filter(
    place_of_death == "Home & other non-institution"
  ) %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  # scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_binned(
    labels = function(x) scales::percent(x, accuracy = 1), 
    type = "viridis",
    n.breaks = 7
    ) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(), 
    aspect.ratio = 1.4,  # needed for Scotland HB plot
    plot.margin = margin(5,5,5,5),
    # legend.position = c(1.2,0.4)  # this was manually tweaked until just right  # problem - edge gets cut
    ) +
  facet_grid(~year) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths in 2020 relative to 5-year historical mean, by Health Board", width=50),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(source_nrs_text,five_year_historical_mean), collapse="\n"),
       # fill = "Ratio"
       fill = "Home death\nincrease"
       ) +
  NULL
```

### 2022-02-18 covid_nrs_place_of_death_data_2022.Rmd
changed author name & email

@@
   set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
-              quiet = FALSE)
 
@@
 
-```{r}
+In the `2015-2019` period there were no Covid-19 deaths, so the historical average should be `NA`, which can possibly be interpreted as 0 for the purposes of comparing a year to the historical average.
+For the `2016-2019, 2021` period the average number of covid deaths should be equivalent to the deaths in 2021 - we wouldn't want to make an average between 4 years of 0 deaths and one year with any deaths!
+Similarly if the period were `2017-2019, 2021-2022`, we would compute the average for 2021-2022, and so on.
+
+```{r}
+## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
+## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
+## however, for the latter, we set the dates to 2022
+average_weekly_covid_deaths_2021_onwards <-
+  bind_rows(
+    ## set 2021 averages to NA
+    weekly_cause_of_death_2021_onwards %>%
+    filter(cause_of_death == "COVID-19" & year == 2021) %>%
+    mutate(number_of_deaths = NA_real_) %>%
+    rename(deaths_mean = number_of_deaths),
+    ## use 2021 weekly deaths as 2022 historic average
+    weekly_cause_of_death_2021_onwards %>%
+    filter(cause_of_death == "COVID-19" & year == 2021) %>%
+    mutate(year = 2022) %>%
+    select(-c(week_number_run_over, date_w_c)) %>%
+    rename(deaths_mean = number_of_deaths) %>%
+    inner_join(weekly_cause_of_death_2021_onwards %>% 
+                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
+              )
+  ) %>%
+  mutate(historic_period = historic_period_depending_on_year(year))
+
 historical_average_cause_of_death_2021_onwards <-
+  bind_rows(
+    ## the hacked averages for 2021 and 2022 above
+    average_weekly_covid_deaths_2021_onwards,
+    ## extract averages from most recent file for 2021 onwards
     map_dfr(
@@
     drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future
+  )
 ```
@@
   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
-  inner_join(
-    historical_average_cause_of_death, by = c("place_of_death", "week_number", "cause_of_death", "historic_period", "year", "date_w_c")
+  left_join(  # left_join so that the historical Covid-19 rates are blank!
+    historical_average_cause_of_death %>% select(-week_number_run_over), 
+    by = c("place_of_death", "week_number", "cause_of_death", "historic_period", "year", "date_w_c")
   )
@@
 ```{r}
-# TODO: work out what to do about this - is it even used?
-
 ## bespoke computation because format is different... deaths split by cause!
-## Note: this isn't appropriate for 2022 yet because we don't have the annual death figure yet
 total_deaths_cause_of_death <-
@@
   group_by(year, place_of_death, cause_of_death, historic_period) %>%
-  summarise(deaths = sum(number_of_deaths), .groups="drop") %>%
+  summarise(deaths = sum(number_of_deaths, na.rm = TRUE), .groups="drop") %>%
   left_join(historical_annual_average_cause_of_death, by = c("place_of_death", "cause_of_death", "historic_period")) %>%
   mutate(
-    ratio_annual_deaths_to_historical_mean = deaths / annual_deaths_mean
-  ) %>%
-  filter(year != 2022)
+    ratio_annual_deaths_to_historical_mean = if_else(  # only compute this for complete years, we can't compare 2022 to annual average!
+      condition = year == 2022,
+      true = NA_real_,
+      false = deaths / annual_deaths_mean
+    )
+  )
 
@@
   bind_rows(
-    ## 2020 & 2021 separately
+    ## 2020-2022 separately
     total_deaths_cause_of_death %>%
@@
       mutate(year = as.character(year)),
-    ## 2020 & 2021 averaged
-    total_deaths_cause_of_death %>%
-      select(year, place_of_death, cause_of_death, deaths) %>%
-      group_by(place_of_death,  cause_of_death) %>%
-      summarise(deaths = sum(deaths), year = "2020-2021", .groups = "drop"),
     ## historical data
-    total_deaths_cause_of_death %>%
+    total_deaths_cause_of_death %>%  # these are for 2020-2021, but we use the historical mean
+      filter(year == 2020) %>%
       select(year, place_of_death, cause_of_death, deaths = annual_deaths_mean) %>%
-      mutate(year = "2015-2019") %>%
-      distinct
+      mutate(year = "2015-2019")
   ) %>%
@@
 
+# Annual deaths by place, 2015-2022
+
+```{r}
+merged_proportions_of_deaths_by_place <-
+  bind_rows(
+    ## 2015-2019
+    weekly_deaths_2015_2019_overall %>% 
+    group_by(year, place_of_death) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%
+    filter(place_of_death != "All" & !is.na(as.integer(year))) %>%
+    group_by(year) %>%
+    mutate(
+      total_deaths = sum(number_of_deaths)
+    ) %>%
+    ungroup %>%
+    mutate(
+      proportion_of_total = number_of_deaths / total_deaths, 
+      year=as.integer(year)
+      ),
+    ## 2020 onwards
+    total_deaths_overall %>%
+    select(year, place_of_death, number_of_deaths = deaths_all_causes) %>%
+    distinct %>%  # remove duplicate entries for 2022 due to having multiple possible historic periods
+    filter(place_of_death != "All") %>%
+    group_by(year) %>%
+    mutate(
+      total_deaths = sum(number_of_deaths)
+    ) %>%
+    ungroup %>%
+    mutate(proportion_of_total = number_of_deaths / total_deaths)
+  )
+```
+
 # TODO: other thigns like excess death & historical rates by geography

### 2022-02-18 covid_nrs_place_of_death_visualisations_2022.Rmd
changed author name & email

@@
 
-Note: visualisations have been moved to a separate .Rmd file
-
 ## Common elements
@@
 
-date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2021, '%d %B %Y')} (inclusive).")
+date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2022, '%d %B %Y')} (inclusive).")
 
@@
 
-save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = figure_place_of_death, width = 28, height = 14, dpi = 300, units = "cm")
+figure_place_of_death %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
+```
+
+## Home deaths only
+
+```{r}
+figure_place_of_death_home_only <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
+       ) +
+  ## annotate historic range
+  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  NULL
+
+figure_place_of_death_home_only %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+# Proportion annual deaths by place
+
+Here we'll show 4 lines, one for each place of death, for each year so far.
+
+```{r}
+figure_proportion_annual_deaths_by_place <-
+  merged_proportions_of_deaths_by_place %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point() +
+  geom_line(size = 1) +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C") +
+  # theme_minimal() +
+  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  labs(
+    x = "Year", y= "Proportion of yearly deaths (%)",
+    colour = "",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c(source_nrs_text),collapse="\n")
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+
+figure_proportion_annual_deaths_by_place %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
+```
+
+# Causes of death over time
+
+## Proportion of causes by place, yearly
+
+This graph shows each of the causes of death divided by place of death - e.g. what proportion of cancer deaths happened at home vs hospital over time?
+
+```{r}
+figure_cause_of_death_annual_proportion_by_place <-
+  current_and_historical_deaths_by_cause_and_location %>%
+  filter(!cause_of_death %in% c("All")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  # mutate(
+  #  cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+  # ) %>%
+  ggplot(aes(x = year, y = proportion, colour = place_of_death, group = place_of_death)) +
+  geom_line() +
+  geom_point() +
+  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
+  scale_colour_viridis_d() +
+  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
+  labs(
+    x = "Year",
+    y = "Percentage yearly deaths (%)",
+    title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
+    subtitle = date_range_text,
+    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",source_nrs_text), collapse = "\n")
+  ) +
+  theme(
+    panel.spacing = unit(1, units = "lines"),
+    legend.position = "top", legend.title = element_blank(),
+    plot.title = element_textbox_simple()  # to render the html in the title
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+
+figure_cause_of_death_annual_proportion_by_place %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_annual_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 26, height = 16, dpi = 300, units = "cm")
+```
+
+## Deviation from mean
+
+```{r}
+figure_place_of_death_deviation_from_mean <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>%
+  mutate(  # subtract mean from all figures to flatten graph
+    deaths = deaths - deaths_mean,
+    deaths_min = deaths_min - deaths_mean,
+    deaths_max = deaths_max - deaths_mean
+  ) %>%
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "#4477aa"), alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = 0, linetype = "dashed"), colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  scale_linetype_manual(values = c("dashed"), labels = "Weekly average (scaled to 0)") +
+  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (difference from weekly average)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
+       ) +
+  annotate(geom = "text", x=ymd("2021-01-28"), y=-250, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=-250, label = "'22", hjust = 0, colour = "grey40") +
+  NULL
+
+figure_place_of_death_deviation_from_mean %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_deviation_from_mean", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
 ```
@@
 ```{r}
-figure_home_deaths_over_time_by_hb_4_week_average <-
+(figure_home_deaths_over_time_by_hb_4_week_average <-
   merged_deaths_hb %>%
@@
   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
-  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max, fill = "#4477aa"), alpha = 0.5) +
   # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
   geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = ma4w_deaths_mean), linetype = "dashed", size = 0.2) +
+  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), size = 0.2) +
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
@@
                       ) +
+  scale_linetype_manual(values = c("dashed"), labels = "Weekly average") +
+  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
   theme(
     legend.position = "top",
+    legend.title = element_blank(),
     plot.caption = element_text(size = 10, colour = "gray60"),
@@
          "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text
-         ), collapse="\n"),
-       colour = "Cause of death"
+         ), collapse="\n")
        ) +
@@
   NULL
+ )
 
-save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = figure_home_deaths_over_time_by_hb_4_week_average, width = 30, height = 20, dpi = 300, units = "cm")
+figure_home_deaths_over_time_by_hb_4_week_average %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 20, dpi = 300, units = "cm")
 ```

### 2022-02-18 download_latest_data_and_render_reports.R
updated downloading & rendering script to use the 2022 scripts for data & visualisations, but there's an error with the covid trends files and they don't work at the minute

@@ -29,7 +29,12 @@ all_nrs_links <-
 
 nrs_filename_pattern <- "covid-deaths-(21|22)-data.*xlsx"  # to accommodate 2022 data also once the filename changes!
 
-weekly_deaths_url_relative <- all_nrs_links[which(str_detect(all_nrs_links, pattern = nrs_filename_pattern))]  # keep only links that are probably weekly deaths
+# TODO: incorporate both weekly and monthly data?
+weekly_deaths_url_relative <- 
+  all_nrs_links[which(
+    str_detect(all_nrs_links, pattern = nrs_filename_pattern) &
+    !str_detect(all_nrs_links, pattern = "monthly")  # remove "monthly" file
+    )]  # keep only links that are probably weekly deaths
 weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
 
 ## the format changed between 2021 and 2022 for weekly deaths data, and the 2022 data no longer include 2020 figures, so we need to load the 2020 data separately from the latest available data
@@ -249,37 +254,37 @@ dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
 
 render(
-  input = "./covid_nrs_place_of_death_data.Rmd",
+  input = "./covid_nrs_place_of_death_data_2022.Rmd",
   output_file = paste0("covid_nrs_place_of_death_data_",today()),
   output_dir = dir_reports,
   envir = new.env()
 )
 
 render(
-  input = "./covid_nrs_place_of_death_visualisations.Rmd",
+  input = "./covid_nrs_place_of_death_visualisations_2022.Rmd",
   output_file = paste0("covid_nrs_place_of_death_visualisations",today()),
   output_dir = dir_reports,
   envir = new.env()
 )
-
-render(
-  input = "./paper_reporting_death_at_home_increase.Rmd",
-  output_file = paste0("paper_reporting_death_at_home_increase_",today()),
-  output_dir = dir_reports,
-  envir = new.env()
-)
-
-render(
-  input = "./visualisations_for_blog.Rmd",
-  output_file = paste0("visualisations_for_blog_",today()),
-  output_dir = dir_reports,
-  envir = new.env()
-)
-
-render(
-  input = "./highlands_data_and_visualisations.Rmd",
-  output_file = paste0("highlands_data_and_visualisations_",today()),
-  output_dir = dir_reports,
[…]

### 2022-02-18 covid_nrs_place_of_death_visualisations_2022.Rmd
added reminder to myself

@@
 ```{r}
+# TODO: fix causes of deaths and reorder by number of deaths!
 figure_cause_of_death_annual_proportion_by_place <-

### 2022-02-21 covid_nrs_place_of_death_data_2022.Rmd
added more visualisations and cleaned up existing

@@
 
+##
+
+## By HB/LA
+
+```{r}
+proportions_of_deaths_by_place_by_geography <-
+  bind_rows(
+    ## historic deaths
+    bind_rows(  # merge HB & LA past deaths, and tag them appropriately
+      weekly_deaths_2015_2019_hb %>% mutate(geography = "hb"),
+      weekly_deaths_2015_2019_la %>% mutate(geography = "la")
+    ) %>%
+    group_by(year, place_of_death, ref_area, geography) %>% summarise(number_of_deaths = sum(number_of_deaths, na.rm = TRUE), .groups="drop") %>%  # compute the total number of yearly deaths by place of death & geography
+    filter(place_of_death != "All" & !is.na(as.integer(year))) %>%  # remove blank/irrelevant entries
+    group_by(year, ref_area, geography) %>%
+    mutate(
+      total_deaths = sum(number_of_deaths)  # compute total annual deaths across place of death, but within each geographical area & year
+    ) %>%
+    ungroup %>%
+    mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))  # compute proportion of deaths in any palce relative to total annual deaths in geographical area
+    ,
+    ## current deaths
+    bind_rows(  # merge HB & LA past deaths, and tag them appropriately
+      total_deaths_hb %>% mutate(geography = "hb"),
+      total_deaths_la %>% mutate(geography = "la")
+    ) %>%
+    select(year, place_of_death, number_of_deaths = deaths_all_causes, geography, ref_area) %>%  # keep only relevant columns & rename deaths due to all causes
+    group_by(year, place_of_death, ref_area, geography) %>% summarise(number_of_deaths = sum(number_of_deaths), .groups="drop") %>%  # compute the total number of yearly deaths by place of death & geography
+    filter(place_of_death != "All" & !is.na(as.integer(year))) %>%  # remove blank/irrelevant entries
+    group_by(year, ref_area, geography) %>%
+    mutate(
+      total_deaths = sum(number_of_deaths)  # compute total annual deaths across place of death, but within each geographical area & year
+    ) %>%
+    ungroup %>%
+    mutate(proportion_of_total = number_of_deaths / total_deaths, year=as.integer(year))  # compute proportion of deaths in any palce relative to total annual deaths in geographical area
+  )
+
+## write to csv
+proportions_of_deaths_by_place_by_geography %>%
+  write_csv(file = paste0(dir_outputs,"/proportions_of_deaths_by_place_by_geography.csv"))
+```
+
+# Deaths by cause, ordered by number of deaths in 2020
+
+```{r}
+## get "canonical" order for causes of death in 2020
+order_of_causes_of_death <- 
+  total_deaths_cause_of_death %>% 
+  filter(year == 2020 & place_of_death == "All") %>% 
+  arrange(desc(deaths)) %>% 
+  pull(cause_of_death)
+```
+
 # TODO: other thigns like excess death & historical rates by geography

### 2022-02-21 covid_nrs_place_of_death_visualisations_2022.Rmd
added more visualisations and cleaned up existing

@@
   labs(
-    x = "Year", y= "Proportion of yearly deaths (%)",
-    colour = "",
+    x = "Year", y = "Proportion of yearly deaths (%)",
+    colour = NULL,
     subtitle = paste0("* ",only_2022_text),
-    caption = paste0(c(source_nrs_text),collapse="\n")
+    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
     ) +
@@
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
+```
+
+# Proportion of cumulative deaths by place, over the course of a year
+
+First we compute the cumulative proportion of place of death within each year. As the weeks go by, the proportions of deaths in each place change, and they change at different rates.
+
+```{r}
+cumulative_deaths_by_place_by_year <-
+  bind_rows(
+    ## current data
+    merged_deaths_overall %>%
+      filter(age == "all" & sex == "all"),
+    ## 2015-2019
+    weekly_deaths_2015_2019_overall %>%
+      mutate(year = as.integer(year)) %>%  # remove the 'average' figures
+      drop_na(year) %>%
+      rename(deaths_all_causes = number_of_deaths)
+  ) %>%
+  arrange(year, week_number) %>%
+  group_by(year, place_of_death) %>%
+  mutate(
+    cumulative_deaths = cumsum(deaths_all_causes)
+  ) %>%
+  group_by(year, week_number, place_of_death == "All"
+  ) %>%
+  mutate(
+    proportion_cumulative_deaths = cumulative_deaths / sum(cumulative_deaths)
+  ) %>%
+  ungroup %>%
+  select(year, place_of_death, week_number, matches("cumulative_deaths"))
+
+cumulative_deaths_by_place_by_year_min_max_2015_2019 <-
+  cumulative_deaths_by_place_by_year %>%
+  filter(year %in% 2015:2019) %>%
+  group_by(place_of_death, week_number) %>%
+  summarise(
+    proportion_cumulative_deaths_min = min(proportion_cumulative_deaths),
+    proportion_cumulative_deaths_max = max(proportion_cumulative_deaths),
+    .groups = "drop"
+  )
+```
+
+## Weekly proportion of deaths by place
+
+These are non-cumulative proportions of place of death for each week.
+You can see how the proportions of place of death are fairly noisy over time, and in 2015-2019 they are relatively stable; it's not until 2020 that a big shift occurs.
+
+```{r}
+figure_weekly_proportion_of_death_by_place <-
+  bind_rows(
+    ## current data
+    merged_deaths_overall %>%
+      filter(age == "all" & sex == "all"),
+    ## 2015-2019
+    weekly_deaths_2015_2019_overall %>%
+      mutate(year = as.integer(year)) %>%  # remove the 'average' figures
+      drop_na(year) %>%
+      rename(deaths_all_causes = number_of_deaths)
+  ) %>%
+  group_by(year, week_number, place_of_death == "All") %>%
+  mutate(proportion_of_deaths = deaths_all_causes / sum(deaths_all_causes)) %>%
+  ungroup %>%
+  filter(place_of_death != "All") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  ggplot(
+    data = ., aes(x = week_number, y = proportion_of_deaths, colour = place_of_death)
+  ) +
+  geom_line() +
+  facet_wrap(~year) +
+  scale_colour_viridis_d() +
+  scale_x_continuous(breaks = scales::breaks_extended(6)) +
+  scale_y_continuous(breaks = scales::breaks_extended(6), label = function(x) scales::percent(x = x, accuracy = 1)) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+  ) +
+  labs(
+    x = "Week number", y = "Weekly proportion of deaths (%)",
+    colour = NULL,
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
+  ) +
+  NULL
+
+figure_weekly_proportion_of_death_by_place %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_proportion_of_death_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
+```
+
+## Yearly graphs, 2015-2022
+
+This is a graph to see whether the relative proportion of deaths by place changes over time - we know the annual proportions for the 4 places of death, for example, but their relative proportion changes over the year.
+This gives us context to be able to decide whether a decline in home deaths early in the year represents a genuine decline or just the general trend of home deaths being lower early in a year.
+
+```{r}
+figure_cumulative_weekly_proportion_of_deaths_by_place <-
+  cumulative_deaths_by_place_by_year %>%
+  filter(place_of_death != "All") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  ggplot(data = ., aes(x = week_number, y = proportion_cumulative_deaths, group = place_of_death, colour = place_of_death)) +
+  geom_line() +
+  facet_wrap(~year) +
+  scale_colour_viridis_d() +
+  scale_fill_viridis_d() +
+  scale_x_continuous(breaks = scales::breaks_extended(6)) +
+  scale_y_continuous(breaks = scales::breaks_extended(6), label = function(x) scales::percent(x = x, accuracy = 1)) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+  ) +
+  labs(
+    x = "Week number", y = "Cumulative proportion of place of death (%)",
+    colour = NULL,
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
+  ) +
+  NULL
+
+figure_cumulative_weekly_proportion_of_deaths_by_place %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cumulative_weekly_proportion_of_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
+```
+
+## Cumulative proportion of place of death, showing 2015-2019 as range
+
+This is a similar graph to the above, but we include the weekly range (min/max) of the weekly proportions of place of death, as historic context for the data from 2020 onwards.
+
+```{r}
+figure_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range <-
+  cumulative_deaths_by_place_by_year %>%
+  filter(place_of_death != "All") %>%
+  filter(!year %in% 2015:2019) %>%
+  left_join(cumulative_deaths_by_place_by_year_min_max_2015_2019, by = c("week_number", "place_of_death")) %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  ggplot(data = ., aes(x = week_number, y = proportion_cumulative_deaths, group = place_of_death, colour = place_of_death)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = proportion_cumulative_deaths_min, ymax = proportion_cumulative_deaths_max, fill = place_of_death, colour = NULL), alpha = 0.4) +
+  facet_wrap(~year) +
+  scale_colour_viridis_d() +
+  scale_fill_viridis_d() +
+  scale_x_continuous(breaks = scales::breaks_extended(6)) +
+  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+  ) +
+  labs(
+    x = "Week number", y = "Cumulative proportion of place of death (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Shaded areas represent the range of cumulative weekly proportion of deaths by place for the 2015-2019 period",death_registrations_text, source_nrs_text),collapse="\n")
+  ) +
+  NULL
+
+figure_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
 ```
@@
   filter(!cause_of_death %in% c("All")) %>%
+  filter(!year %in% c("2020-2021")) %>%
   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-  # mutate(
-  #  cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
-  # ) %>%
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+  ) %>%
   ggplot(aes(x = year, y = proportion, colour = place_of_death, group = place_of_death)) +
@@
   geom_point() +
+  geom_text(data = ~filter(.x, year==2022), aes(x=4.2, y=proportion, label="*"), inherit.aes = FALSE) +  # 4.2 here is 4th (final) place on the scale!
   facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
@@
     x = "Year",
-    y = "Percentage yearly deaths (%)",
+    y = "Percentage yearly deaths by place (%)",
     title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
-    subtitle = date_range_text,
-    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",source_nrs_text), collapse = "\n")
+    subtitle = paste0("\n* ",only_2022_text),
+    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",death_registrations_text,source_nrs_text), collapse = "\n")
   ) +
@@
 figure_cause_of_death_annual_proportion_by_place %>%
-save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_annual_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 26, height = 16, dpi = 300, units = "cm")
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_annual_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
 ```
 
-## Deviation from mean
+## Weekly deaths by cause, at home
+
+```{r}
+annotations_for_2021_and_2022_in_home_deaths_by_cause <-
+  merged_deaths_cause_of_death %>%
+  filter(str_starts(place_of_death,"Home")) %>%
+  filter(cause_of_death != "All") %>%
+  group_by(cause_of_death) %>%
+  summarise(max = max(number_of_deaths), .groups = "drop")
+
+figure_weekly_deaths_by_cause_at_home <-
+  merged_deaths_cause_of_death %>%
+  filter(str_starts(place_of_death,"Home")) %>%
+  filter(cause_of_death != "All") %>%
+  mutate(deaths_mean = if_else(historic_period=="2015-2019" & cause_of_death=="COVID-19", NA_real_, deaths_mean)) %>%  # set historic covid deaths to NA for 2020-2021
+  pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
+  mutate(deaths_type = if_else(deaths_type == "deaths_mean", true = "Historic mean 2015-2019", false = "Deaths in 2020 & 2021")) %>%
+  filter(!(deaths_type=="deaths_mean" & cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  # geom_line(aes(y = number_of_deaths, colour = deaths_type, linetype = deaths_type)) +
+  geom_line(aes(y = number_of_deaths, colour = deaths_type)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dotted", colour="grey70") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dotted", colour="grey70") +  # start of 2022
+  # scale_linetype_manual(values = c("solid", "dashed")) +
+  scale_colour_manual(values = c("grey10","grey50")) +
+  geom_text(data = annotations_for_2021_and_2022_in_home_deaths_by_cause, aes(y = max), x = ymd("2021-01-20"), label = "2021", hjust = 0) +
+  geom_text(data = annotations_for_2021_and_2022_in_home_deaths_by_cause, aes(y = max), x = ymd("2022-01-20"), label = "'22", hjust = 0) +
+  facet_wrap(cause_of_death~., scales = "free_y") +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = "Date w/c",
+       y = "Weekly Deaths (N)",
+       subtitle = paste0(c("Deaths registered at home", date_range_text), collapse = "\n"),
+       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,source_nrs_text), collapse="\n")
+       ) +
+  NULL
+
+figure_weekly_deaths_by_cause_at_home %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_deaths_by_cause_at_home", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 18, dpi = 300, units = "cm")
+```
+
+# Deviation from mean
 
@@
        caption = paste0(c(
-         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text
+         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text,death_registrations_text,source_nrs_text
          ), collapse="\n")
@@
 
-# Geographical variation over time: cumulative deaths at home over time
+# Geographical variations
+
+## Proportion of deaths by place, by HB
+
+```{r}
+figure_proportion_of_deaths_by_place_by_hb <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "hb") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point(aes(size = place_of_death)) +
+  geom_line(aes(size = place_of_death)) +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  guides(colour = guide_legend(nrow=2, title = NULL), 
+         size = guide_legend(nrow=2, title = NULL)) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Health boards sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL
+
+figure_proportion_of_deaths_by_place_by_hb %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_by_place_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+### Home deaths only, HB
+
+```{r}
+figure_proportion_of_deaths_at_home_by_hb <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "hb") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
+  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
+  geom_point() +
+  geom_line() +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80")
+    ) +
+  labs(
+    linetype = NULL,
+    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Health boards sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL
+
+figure_proportion_of_deaths_at_home_by_hb %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## cumulative deaths at home over time
 
@@
 ```{r}
+#TODO: continue working on this visualisation
+#TODO: figure out some interesting time points to visualise
+#TODO: maybe this could be done as a shiny widget? with a date slider and a selection of visualisations: map of covid cases, 
 merged_deaths_hb %>%
@@
        # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       caption = paste0(c(source_nrs_text,five_year_historical_mean), collapse="\n"),
+       caption = paste0(c(five_year_historical_mean,death_registrations_text,source_nrs_text), collapse="\n"),
        # fill = "Ratio"

### 2022-02-21 hospital_classification.Rmd
Added a little bit to hospital classifications following AG's comments about hospitals without A&E units

@@
 write_csv(hospital_classification, file = "./hospital_classification/hospital_classification.csv")
+
+## acute hospitals only
+write_csv(
+  hospital_classification %>% filter(acute), 
+  file = "./hospital_classification/hospital_classification_acute_only.csv")
 ```
+
+# Further considerations
+
+## Hospitals without A&E units
+
+On 24 January 2022, Andrew Gasiorowski contacted me to check if we should include hospitals the following hospitals which no longer have A&E units and only MIU (minor injuries units).
+We could remove these from the 'acute hospitals' list since they don't :
+
+* Golden Jubilee
+* New Stobhill 
+* New Victoria
+
+All of the above hospitals are classed as `Other - Miscellaneous hospitals in which function and/or costs are not comparable with other hospitals`.
+
+MIU is likely not the place a person might have gone to seek help for COVID-related symptoms, however, it's possible that some hospitalisations staretd with a MIU visit, e.g. for falls
+
+This may not be a big issue if the MIUs are close to A&Es, in which case the A&Es are a good approximation in terms of location.
+
+```{r}
+hospital_classification %>% 
+  filter(acute) %>%
+  filter(str_detect(org_loc, regex("stobhill|new victoria|jubilee", ignore_case = TRUE)))
+```
+
+# Session info
+
+```{r}
+sessionInfo()
+```

### 2022-02-23 covid_nrs_place_of_death_visualisations_2022.Rmd
added weekly proportions of causes of death by place - interseting results w.r.t. cancer deaths in 2020 and 2021,with home deaths overtaking hospital deaths

@@
 library(ggtext)  # for showing coloured title text
+
+## load functions
+source("./functions.R")
 ```
@@
   # theme_minimal() +
-  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
+  theme(
+    legend.position = "top", 
+    plot.caption = element_text(size = 10, colour = "gray60")
+    ) +
   labs(
@@
 ```{r}
-# TODO: fix causes of deaths and reorder by number of deaths!
 figure_cause_of_death_annual_proportion_by_place <-
@@
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_annual_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
+```
+
+## Weekly proportion of deaths by place
+
+This is the same idea as above, but looking at weekly proportions of cause of death by place - where did deaths due to different causes happen, and did their proportion vary over the year (e.g.
+seasonal effects)
+
+```{r}
+figure_cause_of_death_weekly_proportion_by_place <-
+  merged_deaths_cause_of_death %>%
+  filter(!cause_of_death %in% c("All")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other", "All"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+  ) %>%
+  group_by(year, week_number, cause_of_death, place_of_death != "All") %>%
+  mutate(proportion_weekly_total = number_of_deaths / sum(number_of_deaths)) %>%
+  ungroup %>%
+  filter(place_of_death != "All") %>%
+  ggplot(aes(x = date_w_c, y = proportion_weekly_total, colour = place_of_death, group = place_of_death)) +
+  geom_line() +
+  annotate(geom = "text", x=ymd("2021-01-14"), y=0.95, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-14"), y=0.95, label = "'22", hjust = 0, colour = "grey40") +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE, ncol = 2) +
+  scale_colour_viridis_d() +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
+  labs(
+    x = "Year",
+    y = "Percentage weekly deaths by place (%)",
+    # title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
+    subtitle = paste0("\n* ",only_2022_text),
+    caption = paste0(c("The 2015-2019 figures are average weekly deaths.",death_registrations_text,source_nrs_text), collapse = "\n")
+  ) +
+  theme(
+    panel.spacing = unit(1, units = "lines"),
+    legend.position = "top", legend.title = element_blank(),
+    plot.title = element_textbox_simple(),  # to render the html in the title
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+
+figure_cause_of_death_weekly_proportion_by_place %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
+```
+
+### With 2015-2019 context also
+
+```{r}
+figure_weekly_proportion_of_death_by_place_with_2015_2019 <-
+  bind_rows(
+    ## 2020 onwards
+    merged_deaths_cause_of_death %>%
+      mutate(
+        year = as.character(year),
+        week_number_run_over = week_number_run_over + 53  # push 2020+ forward by a year, to 'squeeze' in the historic period year
+        )
+    ,
+    ## 2015-2019 data, weekly averages make up one "year"
+    merged_deaths_cause_of_death %>%
+      filter(year==2020) %>%
+      mutate(
+        year = "2015-2019",
+        number_of_deaths = deaths_mean
+        )
+  ) %>%
+  mutate(year = factor(year, levels = c("2015-2019",as.character(2020:2022)))) %>%
+  filter(!cause_of_death %in% c("All")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other", "All"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  mutate(
+   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
+  ) %>%
+  group_by(year, week_number, cause_of_death, place_of_death != "All") %>%
+  mutate(proportion_weekly_total = number_of_deaths / sum(number_of_deaths)) %>%
+  ungroup %>%
+  filter(place_of_death != "All") %>%
+  ggplot(aes(x = week_number_run_over, y = proportion_weekly_total, colour = place_of_death, group = place_of_death)) +
+  geom_line() +
+  # annotate(geom = "text", x=1+1, y=0.95, label = "2015-19", hjust = 0, colour = "grey40") +
+  # annotate(geom = "text", x=54+1, y=0.95, label = "2020", hjust = 0, colour = "grey40") +
+  # annotate(geom = "text", x=107+1, y=0.95, label = "2021", hjust = 0, colour = "grey40") +
+  # annotate(geom = "text", x=159+1, y=0.95, label = "'22", hjust = 0, colour = "grey40") +
+  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE, ncol = 2) +
+  scale_colour_viridis_d() +
+  scale_x_continuous(breaks = c(1,54,107,159), labels = c("2015-2019",as.numeric(2020:2022))) +
+  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
+  labs(
+    x = "Year",
+    y = "Percentage weekly deaths by place (%)",
+    subtitle = paste0("\n* ",only_2022_text),
+    caption = paste0(c("The 2015-2019 figures are average weekly deaths",death_registrations_text,source_nrs_text), collapse = "\n")
+  ) +
+  theme(
+    panel.spacing = unit(1, units = "lines"),
+    legend.position = "top", legend.title = element_blank(),
+    plot.title = element_textbox_simple(),  # to render the html in the title
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+    ) +
+  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+  NULL
+
+figure_weekly_proportion_of_death_by_place_with_2015_2019 %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place_2015_2019", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
 ```

### 2022-03-07 covid_nrs_place_of_death_data_2022.Rmd
Modified code to account for missing "crude_positive_rate" in new case trends data - the rate is now computed from population estimate and cumulative cases; also included the "functions.R" file - forgot to commit earlier!

@@
     ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
-    )
-```
-
-# Calculate case counts: weekly counts to match weekly death reporting
-
-Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks.
-This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
-
-However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!
-
-Note also that the `crude_rate_positive_per100k` represents the cumulative rate!
-
-```{r}
-weekly_covid_cases_hb <-
-  cases_hb %>%
-  group_by(year, week_number, ref_area) %>%
-  summarise(
-    weekly_positive = sum(daily_positive),
-    cumulative_positive = max(cumulative_positive),
-    crude_rate_positive_per100k = max(crude_rate_positive),
-    .groups = "drop"
-  ) %>%
-  mutate(
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
-    )
-
-## extract Scotland-wide data and remove from HB split
-weekly_covid_cases_overall <- weekly_covid_cases_hb %>% filter(ref_area=="Scotland")
-weekly_covid_cases_hb <- weekly_covid_cases_hb %>% filter(ref_area!="Scotland")
-
-weekly_covid_cases_la <-
-  cases_la %>%
-  group_by(year, week_number, ref_area) %>%
-  summarise(
-    weekly_positive = sum(daily_positive),
-    cumulative_positive = max(cumulative_positive),
-    crude_rate_positive_per100k = max(crude_rate_positive),
-    .groups = "drop"
-  ) %>%
-  mutate(
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
-    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
     )
@@
 
+# Calculate case counts: weekly counts to match weekly death reporting
+
+Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks.
+This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
+
+However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!
+
+Note also that the `crude_rate_positive_per100k` represents the cumulative rate!
+
+## Calculating population size of areas
+
+Update 7 Mr 2022: at some point the file structure changed, and the crude positive rate per 100k was removed...
+however, the crude mortality rate remains, so we can infer the population from the `crude_rate_deaths` and `cumulative_deaths` figures, and compute the `crude_rate_positive` from those.
+
+The trouble is that early on in the data, when `cumulative_deaths` and `crude_rate_deaths` are both 0, we can't infer the population size.
+If the population size is constant in the data, we can calculate it for each `ref_area` (HB or LA) and use that as a lookup table.
+
+It looks like the same population estimate was used for data 2020-2022, and it is the **mid-year 2019 estimate**.
+
+## Importing & calculating case trends
+
+```{r}
+weekly_covid_cases_hb <-
+  cases_hb %>%
+  left_join(
+    population_estimate_hb_2019 %>%
+    add_scotland_total_to_population_estimate %>%
+    select(ref_area = hb, population_size = all_ages)
+    ) %>%
+  group_by(year, week_number, ref_area) %>%
+  summarise(
+    weekly_positive = sum(daily_positive, na.rm = TRUE),
+    cumulative_positive = max(cumulative_positive, na.rm = TRUE),
+    # crude_rate_positive_per100k = max(crude_rate_positive),  # this column is no longer in the data, so we compute it below...
+    crude_rate_positive_per100k = max(cumulative_positive / population_size * 1e5, na.rm = TRUE),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
+    )
+
+## extract Scotland-wide data and remove from HB split
+weekly_covid_cases_overall <- weekly_covid_cases_hb %>% filter(ref_area=="Scotland")
+weekly_covid_cases_hb <- weekly_covid_cases_hb %>% filter(ref_area!="Scotland")
+
+weekly_covid_cases_la <-
+  cases_la %>%
+  left_join(
+    population_estimate_la_2019 %>%
+    add_scotland_total_to_population_estimate %>%
+    select(ref_area = la, population_size = all_ages)
+    ) %>%
+  group_by(year, week_number, ref_area) %>%
+  summarise(
+    weekly_positive = sum(daily_positive, na.rm = TRUE),
+    cumulative_positive = max(cumulative_positive, na.rm = TRUE),
+    # crude_rate_positive_per100k = max(crude_rate_positive),  # this column is no longer in the data, so we compute it below...
+    crude_rate_positive_per100k = max(cumulative_positive / population_size * 1e5, na.rm = TRUE),
+    .groups = "drop"
+  ) %>%
+  mutate(
+    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
+    )
+```
+
 # Comparisons of annual rates with current rates

### 2022-03-07 download_latest_data_and_render_reports.R

@@ -89,6 +89,10 @@ if (file.exists(file_historical_data_sources)) {
   )
 }
 
+## Case trend data from the following pages
+## Daily COVID-19 Cases in Scotland, subsection Daily Case Trends By...:
+## LA: https://www.opendata.nhs.scot/dataset/covid-19-in-scotland/resource/427f9a25-db22-4014-a3bc-893b68243055
+## HB: https://www.opendata.nhs.scot/dataset/covid-19-in-scotland/resource/2dd8534b-0a6f-4744-9253-9565d62f96c2
 
 if (file.exists(file_case_trends_data_sources)) {
   table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)

### 2022-03-07 functions.R

@@ -1,0 +1,239 @@
+# Functions for use in the analysis
+
+
+compute_start_date_from_week_number <- function(week_number, year_number) {
+  ## this assumes a valid week number! There are 52 or 53 weeks in a year
+  isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
+  # computed_dates <- ISOweek::ISOweek2date(isoweek_string)
+  computed_dates <- rep(NA_Date_,times = length(week_number))
+  computed_dates[which(!is.na(week_number))] <- ISOweek::ISOweek2date(isoweek_string[which(!is.na(week_number))])  # return NA instead of failing with ISOweek!
+  ## invalid weeks will result in a computed date in a year after the requested year...
+  ## a valid week can result in a date that starts in the year before!
+  if (any(year_number<year(ymd(computed_dates)), na.rm = TRUE)) stop("invalid week number provided!")
+  return(computed_dates)
+}
+
+# Place of death was named differently between datasets (location/place) and the different places were also named slightly differently, so we find all of them and recode them to the same:
+order_of_place_of_death_levels <- c("Hospital", "Care home", "Home & other non-institution", "Other", "All")
+
+recode_place_of_death <- function(data_tbl) {
+  data_tbl %>%
+    mutate(place_of_death = case_when(
+      str_detect(place_of_death, pattern = regex(pattern = "all", ignore_case = TRUE)) ~ "All",
+      str_detect(place_of_death, pattern = regex(pattern = "care.*home", ignore_case = TRUE)) ~ "Care home",
+      str_detect(place_of_death, pattern = regex(pattern = "(non.*institut)|home", ignore_case = TRUE)) ~ "Home & other non-institution",
+      str_detect(place_of_death, pattern = regex(pattern = "hospital", ignore_case = TRUE)) ~ "Hospital",
+      str_detect(place_of_death, pattern = regex(pattern = "other", ignore_case = TRUE)) ~ "Other",
+      TRUE ~ NA_character_
+    ) %>%
+      factor(x = ., levels = order_of_place_of_death_levels)
+    )
+}
+
+
+## Consistent Cause of death coding
[…]

### 2022-03-28 covid_nrs_place_of_death_visualisations_2022.Rmd
added bits of code to keep 2015-2019 historic period in 2022, and a place of death figure using the same

@@
 removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death.")
+
+historic_period_2015_2019_text <- "Historic mean & range for 2015-2019 throughout."
+historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
 ```
@@
        subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
+       caption = paste0(c(death_registrations_text,historic_period_2022_text,source_nrs_text), collapse="\n")
        ) +
@@
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+### Home deaths with 2015-2019 historic period means & range only
+
+```{r}
+merged_deaths_overall_2015_2019_historic_period <-
+  bind_rows(
+    ## pre-2022 data
+    merged_deaths_overall %>% filter(year < 2022),
+    ## 2022 data
+    merged_deaths_overall %>%
+      filter(year >= 2022) %>%  # 2022 or later!?
+      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
+      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
+      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
+      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
+        current_data = ., 
+        historical_data = 
+          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
+        num_weeks_to_average = 4)
+  )
+
+figure_place_of_death_home_2015_2019_historic_period <-
+  merged_deaths_overall_2015_2019_historic_period %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n")
+       ) +
+  ## annotate historic range
+  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "2022", hjust = 0, colour = "grey40") +
+  NULL
+
+figure_place_of_death_home_2015_2019_historic_period %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_2015-2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
 ```

### 2022-05-18 download_latest_data_and_render_reports.R
renamed avoidable deaths to 2019

@@ -42,7 +42,6 @@ weekly_deaths_url_relative <- # keep only links that are probably weekly deaths
     n = 1
   )
 
-
 weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
 
 ## the format changed between 2021 and 2022 for weekly deaths data, and the 2022 data no longer include 2020 figures, so we need to load the 2020 data separately from the latest available data

### 2022-05-18 compile_avoidable_death_codes_2019.Rmd
Fixed issue with some of the diseases that were treatable not being labelled appropriately because I used an asterisk rather than • character

--- 2021-09-17 compile_avoidable_death_codes.Rmd
@@
       str_detect(treatable, pattern = "50") ~ 0.5,
-      str_detect(treatable, pattern = "\\*") ~ 1,
+      str_detect(treatable, pattern = "\\•") ~ 1,
       TRUE ~ 0

### 2022-05-18 compile_avoidable_death_codes_2019.Rmd
added a little check that I should have done originally

@@
   relocate(group, condition_group_and_cause, original_codes)
+
+## all codes should be treatable, preventable, or 50-50
+avoidable_deaths %>% count(treatable, preventable)
 ```

### 2022-05-18 compile_avoidable_death_codes_2020.Rmd
Updated avoidable mortality extraction script for latest data (2020)

--- 2022-05-18 compile_avoidable_death_codes_2019.Rmd
@@
 ---
-title: "Compiling avoidable deaths codes (as coded by NRS)"
+title: "Compiling avoidable deaths codes (as coded by NRS), 2020"
 author: "Jan Savinc"
@@
 NRS publish an annual report on the number of avoidable/treatable deaths.
-The latest report is from December 2020 and reports on data from 2019, available on [the NRS website](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/avoidable-mortality).
+The latest report is from March 2022 and reports on data from 2020, available on [the NRS website](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/avoidable-mortality).
 
-The data used in the report can be downloaded in .csv and .xlsx format - I'll extract the code list for avoidable deaths from the excel file (.xlsx): [https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2019/avoid-mortality-19-all-tabs.xlsx](https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2019/avoid-mortality-19-all-tabs.xlsx)
+The data used in the report can be downloaded in .csv and .xlsx format - I'll extract the code list for avoidable deaths from the excel file (.xlsx), sheet `Table A`: [https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2020/avoid-mortality-20-all-tabs.xlsx](https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2020/avoid-mortality-20-all-tabs.xlsx)
+
+The main change is that the NRS has joined efforts with ONS and adopted the 'Joint OECD/Eurostat' definition of avoidable mortality, including codelists. These include Covid-19, also!
 
@@
 ```{r}
-url_avoidable_deaths_2019 <- "https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2019/avoid-mortality-19-all-tabs.xlsx"
-path_avoidable_deaths_2019 <- "./downloaded_data/avoid-mortality-19-all-tabs.xlsx"
+url_avoidable_deaths_2020 <- "https://www.nrscotland.gov.uk/files//statistics/avoidable-mortality/2020/avoid-mortality-20-all-tabs.xlsx"
+path_avoidable_deaths_2020 <- file.path("./downloaded_data", basename(url_avoidable_deaths_2020))
 
-if (!file.exists(path_avoidable_deaths_2019)) download.file(url = url_avoidable_deaths_2019, destfile = path_avoidable_deaths_2019, mode = "wb")
+if (!file.exists(path_avoidable_deaths_2020)) download.file(url = url_avoidable_deaths_2020, destfile = path_avoidable_deaths_2020, mode = "wb")
 ```
@@
 ```{r}
-avoidable_deaths_raw <- read_excel(path = path_avoidable_deaths_2019, sheet = "Definitions", range = "A4:E225") %>%
+avoidable_deaths_raw <- read_excel(path = path_avoidable_deaths_2020, sheet = "Table A", range = "A4:E115") %>%
   clean_names
@@
 if (!dir.exists("./avoidable_deaths")) dir.create("./avoidable_deaths")
-write_csv(avoidable_deaths, file = "./avoidable_deaths/avoidable_deaths_2019_inclusions.csv")
+write_csv(avoidable_deaths, file = "./avoidable_deaths/avoidable_deaths_2020_inclusions.csv")
 ```

### 2022-05-24 compile_avoidable_death_codes_2020.Rmd
Fixed error in parsing age variable where spaces and some kind of invisible trailing character were erroneously included

@@
 
+## Parsing clean data
+
 ```{r}
@@
   separate_rows(icd_10_codes, sep=",\\s*") %>%  # separate comma-separated values again now that we've converted them from the ranges
-  separate(col = age, into = c("age_min", "age_max") ,sep = "\\-", remove = TRUE) %>%
+  extract(col = age, into = c("age_min", "age_max"), regex = "(\\d+)[^0-9]+(\\d+)", remove = TRUE) %>%
   mutate(

### 2022-05-24 compile_avoidable_death_codes_2020.Rmd
Fixed error in parsing age variable where spaces and some kind of invisible trailing character were erroneously included

@@
   separate_rows(icd_10_codes, sep=",\\s*") %>%  # separate comma-separated values again now that we've converted them from the ranges
-  extract(col = age, into = c("age_min", "age_max"), regex = "(\\d+)[^0-9]+(\\d+)", remove = TRUE) %>%
+  extract(col = age, into = c("age_min", "age_max"), regex = "(\\d+)[^0-9]+(\\d+)", remove = TRUE, convert = TRUE) %>%
   mutate(
@@
 
+## Checking parsing of age
+
+There was previously an error with this, with some ages having a leading space, and a trailing invisible character.
+This is slightly overkill to check since the minimum and maximum ages are the same for all conditions, but it may change in the future!
+I fixed this by using `extract()` to parse the digits from the entries rather than just splitting them on the hyphen symbol.
+
+```{r}
+avoidable_deaths_raw %>% 
+  count(age) %>%
+  kable(caption = "The different possible entries for 'age' in the raw spreadsheet. Note how there are two identical-looking entries, probably due to an invisible character.")
+
+avoidable_deaths%>% 
+  count(age_min, age_max) %>%
+  kable(caption = "The different possible entries for 'age' in the clean data after parsing.")
+```
+
 # Notes on usage

### 2022-05-25 compile_avoidable_death_codes_2020.Rmd
fixed error in parsing ranges; I needed to replace all cases of regexes picking up a hyphen to picking up the \\p{Pd} pattern for hyphen-like characters

@@
 ```{r}
+## note: "\\p{Pd}" captures all types of hyphens/dashes
+
 avoidable_deaths_raw %>% filter(str_detect(string = icd_10_codes, pattern = regex("excl", ignore_case = TRUE))) %>% kable(caption = "Entries with exclusions listed")
 
-avoidable_deaths_raw %>% filter(str_detect(string = icd_10_codes, pattern = regex("\\-", ignore_case = TRUE))) %>% kable(caption = "Entries specifying code ranges")
+avoidable_deaths_raw %>% filter(str_detect(string = icd_10_codes, pattern = regex("\\p{Pd}", ignore_case = TRUE))) %>% kable(caption = "Entries specifying code ranges")
 
@@
   ## helper function for generating code ranges - based on the icd package, function .icd10cm_get_majors_possible()
-  if (is.na(code_range) | !str_detect(string = code_range, pattern="\\-")) return(code_range)
+  if (is.na(code_range) | !str_detect(string = code_range, pattern="\\p{Pd}")) return(code_range)
 
   letter_1 <- str_replace(string = code_range, pattern = "^([A-Za-z]).*$", replacement = "\\1")
-  letter_2 <- str_replace(string = code_range, pattern = "^.+\\-([A-Za-z]).*$", replacement = "\\1")
+  letter_2 <- str_replace(string = code_range, pattern = "^.+\\p{Pd}([A-Za-z]).*$", replacement = "\\1")
   letters <- LETTERS[which(LETTERS==letter_1) : which(LETTERS==letter_2)]
@@
 
-  number_1 <- str_replace(string = str_remove_all(code_range, pattern="\\."), pattern = "^[A-Za-z](\\d{1,3})\\-.*$", replacement = "\\1")
-  number_2 <- str_replace(string = str_remove_all(code_range, pattern="\\."), pattern = "^.+\\-[A-Za-z](\\d{1,3})$", replacement = "\\1")
+  number_1 <- str_replace(string = str_remove_all(code_range, pattern="\\."), pattern = "^[A-Za-z](\\d{1,3})\\p{Pd}.*$", replacement = "\\1")
+  number_2 <- str_replace(string = str_remove_all(code_range, pattern="\\."), pattern = "^.+\\p{Pd}[A-Za-z](\\d{1,3})$", replacement = "\\1")
 
@@
   drop_na(condition_group_and_cause) %>%  # remove blank rows used for blank space in the excel file
+  mutate(across(everything(), ~str_replace_all(string = .x, pattern = "\\p{Pd}", replacement = "-"))) %>%  # replace all hyphen-likes with normal hyphen
   mutate(
@@
   drop_na(icd_10_codes) %>%
-  separate_rows(icd_10_codes, sep=",\\s*", ) %>%  # separate comma-separated values into rows so we have one entry per row
+  separate_rows(icd_10_codes, sep=",\\s*") %>%  # separate comma-separated values into rows so we have one entry per row
   extract(col = icd_10_codes, into = "exclusions", regex = "\\(excl\\.\\s*(.+)\\)", remove = FALSE) %>%  # extract the exclusions and put them in a separate column
@@
   extract(col = icd_10_codes, into = "exclusions", regex = "\\(excl\\.\\s*(.+)\\)", remove = FALSE) %>%  # extract the exclusions and put them in a separate column
-  filter(str_detect(string = icd_10_codes, pattern = "\\-")) %>%
+  filter(str_detect(string = icd_10_codes, pattern = "\\p{Pd}")) %>%
   mutate(
@@
 avoidable_deaths_raw %>%
+  filter(age!="0-74")
+
+avoidable_deaths_raw %>% 
   count(age) %>%
@@
 
-avoidable_deaths%>% 
+avoidable_deaths %>% 
   count(age_min, age_max) %>%

### 2022-06-08 covid_nrs_place_of_death_visualisations_2022.Rmd
Added a few more LA-related plots

@@
                   axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
-                  text=element_text(family="Calibri")
+                  text=element_text(family="Calibri"),
+                  axis.text = element_text(size = 9)
             ))
@@
 ```{r}
-figure_proportion_annual_deaths_by_place <-
+(figure_proportion_annual_deaths_by_place <-
   merged_proportions_of_deaths_by_place %>%
@@
   NULL
-
+)
 figure_proportion_annual_deaths_by_place %>%
@@
 
-# Home deaths by HB over time: 2020-2022, 4-week average
+# Home deaths by geography over time: 2020-2022, 4-week average
+
+## Home deaths by HB, 4-week average
 
@@
 
+## Home deaths by LA, 4-week average
+
+```{r}
+(figure_home_deaths_over_time_by_la_4_week_average <-
+  merged_deaths_la %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  pivot_longer(cols = c(ma4w_deaths_all_causes, ma4w_deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max, fill = "#4477aa"), alpha = 0.5) +
+  # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), size = 0.2) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  # scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~ref_area, scales = "free_y") +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  scale_linetype_manual(values = c("dashed"), labels = "Weekly average") +
+  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+    ) +
+  labs(x = NULL,
+       y = "Deaths (average of past 4 weeks)",
+       title = "Deaths at Home*: 4-week average by Scottish Local Authority",
+       subtitle = date_range_text,
+       caption = paste0(c(
+         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text,death_registrations_text,source_nrs_text
+         ), collapse="\n")
+       ) +
+  ## annotate historic range
+  annotate(geom = "text", x=ymd("2021-01-07"), y=0, label = "2021", hjust = 0, vjust = -1, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=0, label = "'22", hjust = 0, vjust = -1, colour = "grey40") +
+  NULL
+ )
+
+figure_home_deaths_over_time_by_la_4_week_average %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 40, height = 20, dpi = 300, units = "cm")
+```
+
 # Geographical variations
@@
 ```{r}
-figure_proportion_of_deaths_by_place_by_hb <-
+(figure_proportion_of_deaths_by_place_by_hb <-
   proportions_of_deaths_by_place_by_geography %>%
@@
     panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
-    strip.background = element_rect(fill = NA, colour = "gray80")
+    strip.background = element_rect(fill = NA, colour = "gray80"),
+    axis.text.x = element_text(angle = 45, hjust = 1)
     ) +
@@
     ) +
-  NULL
+  NULL)
 
@@
 
-### Home deaths only, HB
-
-```{r}
-figure_proportion_of_deaths_at_home_by_hb <-
+## Proportion of deaths by place, by LA
+
+```{r}
+(figure_proportion_of_deaths_by_place_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point(aes(size = place_of_death)) +
+  geom_line(aes(size = place_of_death)) +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80"),
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    ) +
+  guides(colour = guide_legend(nrow=2, title = NULL), 
+         size = guide_legend(nrow=2, title = NULL)) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL)
+
+figure_proportion_of_deaths_by_place_by_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_by_place_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Home deaths only, HB
+
+```{r}
+(figure_proportion_of_deaths_at_home_by_hb <-
   proportions_of_deaths_by_place_by_geography %>%
@@
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Home deaths only, LA
+
+```{r}
+(figure_proportion_of_deaths_at_home_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
+  geom_point() +
+  geom_line() +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80"),
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    ) +
+  labs(
+    linetype = NULL,
+    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL)
+
+figure_proportion_of_deaths_at_home_by_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Map-based visualisations
+
+```{r}
+## merge the 2015-2019 period historical weekly averages to 2022 data
+merged_deaths_la_historic_period_2015_2019 <-
+  bind_rows(
+    merged_deaths_la %>% filter(year %in% 2020:2021),
+    merged_deaths_la %>% filter(year == 2022) %>%
+      select(-c(deaths_mean, deaths_min, deaths_max)) %>%
+      mutate(historic_period = "2015-2019") %>%
+      left_join(historical_weekly_average_la, by = c("week_number", "place_of_death", "ref_area", "historic_period"))
+  ) %>%
+  left_join(weekly_covid_cases_la, by = c("year","week_number","date_w_c","week_number_run_over","ref_area"))
+```
+
+### Home deaths increase since 2019 by LA
+
+For this we can take the total annual deaths in each LA and compare to historical figures, showing the ratio for each year so far; alternatively we can also average over 2020-2022, and compare that to the average!
+Because we only have partial data for 2022, we need to employ the weekly comparison method -
+
+```{r}
+(figure_map_home_deaths_increase_from_2020_la <-
+  merged_deaths_la_historic_period_2015_2019 %>%
+  filter(historic_period == "2015-2019") %>%
+  group_by(ref_area, place_of_death) %>%
+  summarise(
+    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
+    percent_increase = ratio_deaths_to_historic - 1,
+    .groups = "drop"
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  labs(x="", y="",
+       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority", width=40),
+       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease relative\nto 2015-2019"
+       ) +
+  NULL)
+
+figure_map_home_deaths_increase_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
+```
+
+### Home deaths increase since 2019 by LA, by year
+
+For this we can take the total annual deaths in each LA and compare to historical figures, showing the ratio for each year so far; alternatively we can also average over 2020-2022, and compare that to the average!
+Because we only have partial data for 2022, we need to employ the weekly comparison method -
+
+```{r}
+(figure_map_home_deaths_increase_by_year_la <-
+  merged_deaths_la_historic_period_2015_2019 %>%
+  filter(historic_period == "2015-2019") %>%
+  group_by(year, ref_area, place_of_death) %>%
+  summarise(
+    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
+    percent_increase = ratio_deaths_to_historic - 1,
+    .groups = "drop"
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(
+    axis.text = element_blank(), 
+    panel.grid = element_blank(),
+    # legend.position = "bottom"
+    ) +
+  facet_wrap(~year, nrow = 1) +
+  labs(x="", y="",
+       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
+       caption = paste0(c(source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease relative\nto 2015-2019"
+       ) +
+  NULL)
+
+figure_map_home_deaths_increase_by_year_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_by_year_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 32, height = 18, dpi = 300, units = "cm")
+```
+
+### Home deaths increase vs. covid cases by LA, all deaths since 2019
+
+```{r}
+# TODO: continue working on this!
+(figure_map_covid_cases_from_2020_la <-
+  weekly_covid_cases_la %>%
+  group_by(ref_area) %>%
+  summarise(
+    ratio_covid_cases = max(crude_rate_positive_per100k),
+    .groups = "drop"
+  ) %>% 
+  # left_join(population_estimate_la_2019, by = c("ref_area"="la")) %>%
+  # dplyr::group_by(ref_area) %>%
+  # dplyr::summarise(
+  #   ratio_covid_cases = sum(weekly_positive, na.rm=TRUE) / all_ages *1e5, 
+  #   .groups = "drop"
+  # ) %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  labs(x="", y="",
+       title = str_wrap("Total COVID-19 cases since the beginning of 2020 per 100,000 population, by Local Authority", width=50),
+       subtitle = "",
+       caption = paste0(c(source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
+       ) +
+  NULL)
+
+figure_map_covid_cases_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("figure_map_covid_cases_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
 ```

### 2022-06-09 covid_nrs_place_of_death_visualisations_2022.Rmd
finished the home deaths increase vs cumulative covid rate by LA plot

@@
 source_phs_text <- "Source: Public Health Scotland"
+
+crude_rate_text <- "Total cases = crude cumulative positive cases over time."
 
@@
 
+## Re-merge weekly deaths with 2015-2019 data for 2022 for LA & HB
+
 ```{r}
@@
        # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       caption = paste0(c(source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
        title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
-       caption = paste0(c(source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
 ```{r}
-# TODO: continue working on this!
 (figure_map_covid_cases_from_2020_la <-
@@
   ) %>% 
-  # left_join(population_estimate_la_2019, by = c("ref_area"="la")) %>%
-  # dplyr::group_by(ref_area) %>%
-  # dplyr::summarise(
-  #   ratio_covid_cases = sum(weekly_positive, na.rm=TRUE) / all_ages *1e5, 
-  #   .groups = "drop"
-  # ) %>%
   left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
@@
   labs(x="", y="",
-       title = str_wrap("Total COVID-19 cases since the beginning of 2020 per 100,000 population, by Local Authority", width=50),
+       title = str_wrap("Total COVID-19 cases since the start of 2020, by Local Authority", width=40),
+       subtitle = "",
+       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
+       ) +
+  NULL)
+
+figure_map_covid_cases_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_covid_cases_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
+
+## combine maps
+(figure_map_home_deaths_increase_vs_covid_case_from_2020_la <-
+  wrap_plots(
+    figure_map_home_deaths_increase_from_2020_la +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+      ,
+    figure_map_covid_cases_from_2020_la +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+  )
+)
+
+figure_map_home_deaths_increase_vs_covid_case_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_vs_covid_case_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 16, dpi = 300, units = "cm")
+```
+
+#### LA covid rates over time
+
+I didn't finish this plot because the scale goes from 500 per 100k to 25k per 100k, which is a pain to depict in a meaningful way. Mayeb with a separate scale for each plot?
+
+```{r}
+## the problem with this plot is that there were between 500 per 100k cases to 25k per 100k cases over time and over LAs - how do we visualise that?
+(figure_map_covid_cases_by_year_la <-
+  weekly_covid_cases_la %>%
+  left_join(population_estimate_la_2019 %>% rename(population_estimate = all_ages, year_estimate = year), by = c("ref_area"="la")) %>%
+  group_by(ref_area, year, population_estimate) %>%
+  summarise(
+    cumulative_cases = sum(weekly_positive, na.rm=TRUE),
+    ratio_covid_cases = cumulative_cases / population_estimate * 1e5,
+    .groups = "drop"
+  ) %>%
+  distinct %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  # ggplot(aes(geometry=geometry, fill=log(ratio_covid_cases)), colour="grey10") +
+  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(option = "B", labels = function(x) format(exp(x), big.mark=",")) +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  facet_wrap(~year, nrow = 1) +
+  labs(x="", y="",
+       title = str_wrap("Total COVID-19 cases since the start of 2020 per 100,000 population, by Local Authority", width=50),
        subtitle = "",
@@
   NULL)
-
-figure_map_covid_cases_from_2020_la %>%
-save_output_file(filename = file.path(dir_outputs, paste0(c("figure_map_covid_cases_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
 ```

### 2022-06-10 covid_nrs_place_of_death_visualisations_2022.Rmd
added cause of death plots for data sicne 2020 and by year; also added a line that hopefulyl stops the annyoing messages about missing fonts

--- 2022-03-28 covid_nrs_place_of_death_visualisations_2022.Rmd
@@
 ```{r}
-loadfonts()  # load imported fonts
+loadfonts(device="win")  # load imported fonts
 # Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
+windowsFonts("Calibri" = windowsFont("Calibri"))  # to suppress the multiple error messages
 
@@
 ## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
-annotations_figure_place_of_death_explaining_historic_range <- tribble(
+annotations_fig_place_of_death_explaining_historic_range <- tribble(
    ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
@@
 
-figure_place_of_death <-
+fig_place_of_death <-
   merged_deaths_overall %>%
@@
   ## annotate historic range
-  geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
-  geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
+  geom_text(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
+  geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
   annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
@@
 
-figure_place_of_death %>%
+fig_place_of_death %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
@@
 ```{r}
-figure_place_of_death_home_only <-
+fig_place_of_death_home_only <-
   merged_deaths_overall %>%
@@
 
-figure_place_of_death_home_only %>%
+fig_place_of_death_home_only %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
@@
 
-figure_place_of_death_home_2015_2019_historic_period <-
+fig_place_of_death_home_2015_2019_historic_period <-
   merged_deaths_overall_2015_2019_historic_period %>%
@@
 
-figure_place_of_death_home_2015_2019_historic_period %>%
+fig_place_of_death_home_2015_2019_historic_period %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_2015-2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
@@
 ```{r}
-figure_proportion_annual_deaths_by_place <-
+(fig_proportion_annual_deaths_by_place <-
   merged_proportions_of_deaths_by_place %>%
@@
   NULL
-
-figure_proportion_annual_deaths_by_place %>%
+)
+fig_proportion_annual_deaths_by_place %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
@@
 ```{r}
-figure_weekly_proportion_of_death_by_place <-
+fig_weekly_proportion_of_death_by_place <-
   bind_rows(
@@
 
-figure_weekly_proportion_of_death_by_place %>%
+fig_weekly_proportion_of_death_by_place %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_proportion_of_death_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
@@
 ```{r}
-figure_cumulative_weekly_proportion_of_deaths_by_place <-
+fig_cumulative_weekly_proportion_of_deaths_by_place <-
   cumulative_deaths_by_place_by_year %>%
@@
 
-figure_cumulative_weekly_proportion_of_deaths_by_place %>%
+fig_cumulative_weekly_proportion_of_deaths_by_place %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cumulative_weekly_proportion_of_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
@@
 ```{r}
-figure_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range <-
+fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range <-
   cumulative_deaths_by_place_by_year %>%
@@
 
-figure_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range %>%
+fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
@@
 
+# Causes of death compared to historic context
+
+Here I first merge the 2022 data series with the historic period 2015-2019 averages for weekly figures - this way we can compare the total in 2022 so far with the historical figures up to the same week.
+
+```{r}
+merged_deaths_cause_of_death_with_2015_2019_historic_period <-
+  bind_rows(
+    merged_deaths_cause_of_death %>%
+      filter(year %in% 2020:2021)
+    ,
+    merged_deaths_cause_of_death %>%
+      filter(year == 2022) %>%
+      select(-c(historic_period,deaths_mean)) %>%
+      left_join(
+        merged_deaths_cause_of_death %>%
+          filter(year==2021) %>%
+          select(place_of_death, week_number, cause_of_death, historic_period, deaths_mean),
+        by = c("place_of_death", "week_number", "cause_of_death")
+      )
+  )
+
+total_deaths_cause_of_death_so_far <-
+  merged_deaths_cause_of_death_with_2015_2019_historic_period %>%
+  group_by(year, cause_of_death, place_of_death, historic_period) %>%
+  summarise(
+    number_of_deaths = sum(number_of_deaths, na.rm = TRUE),
+    deaths_mean = sum(deaths_mean, na.rm = TRUE),
+    .groups = "drop"
+  ) %>%
+  group_by(year, place_of_death, historic_period) %>%
+  mutate(
+    proportion_all_deaths = number_of_deaths / sum(number_of_deaths, na.rm = TRUE),
+    proportion_all_deaths_historic = deaths_mean / sum(deaths_mean, na.rm = TRUE)
+  ) %>%
+  ungroup
+```
+
+## Cuases so far by place
+
+```{r}
+(fig_deaths_so_far_by_cause <-
+  total_deaths_cause_of_death_so_far %>%
+  filter(!place_of_death %in% c("Other","All")) %>%
+  group_by(cause_of_death, place_of_death, historic_period) %>%
+  summarise(
+    number_of_deaths = sum(number_of_deaths, na.rm = TRUE),
+    deaths_mean = sum(deaths_mean, na.rm = TRUE),
+    .groups = "drop"
+  ) %>%
+  group_by(place_of_death, historic_period) %>%
+  mutate(
+    proportion_all_deaths = number_of_deaths / sum(number_of_deaths, na.rm = TRUE),
+    proportion_all_deaths_historic = deaths_mean / sum(deaths_mean, na.rm = TRUE)
+  ) %>%
+  ungroup %>%
+  mutate(
+     cause_of_death = fct_rev(factor(cause_of_death)),
+     place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))
+     ) %>%
+  ggplot() +
+  geom_col(aes(y = cause_of_death, fill = cause_of_death, x = number_of_deaths), alpha = 0.8, colour = "grey10") +
+  geom_segment(aes(y = as.numeric(cause_of_death)-0.45, yend = as.numeric(cause_of_death)+0.45, x = deaths_mean, xend = deaths_mean), size = 0.8, colour = "grey1") +
+  scale_fill_viridis_d() +
+  facet_grid(~place_of_death) +
+  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
+  scale_y_discrete(labels = ~str_wrap(.x, width = 20)) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL, y = NULL,
+    title = str_wrap("Total deaths since start of 2020 by cause and place of death, compared to 2015-2019", width = 60),
+    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2022_text)), collapse="\n"),
+    caption = paste0(c(death_registrations_text, source_nrs_text),collapse="\n")
+  ) +
+  NULL)
+
+fig_deaths_so_far_by_cause %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_deaths_so_far_by_cause", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 12, dpi = 300, units = "cm")
+```
+
+## Causes by place & year
+
+```{r}
+(fig_annual_deaths_by_cause <-
+  total_deaths_cause_of_death_so_far %>%
+  filter(!place_of_death %in% c("Other","All")) %>%
+  mutate(
+     cause_of_death = fct_rev(factor(cause_of_death)),
+     place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))
+     ) %>%
+  ggplot() +
+  geom_col(aes(y = cause_of_death, fill = cause_of_death, x = number_of_deaths), alpha = 0.8, colour = "grey10") +
+  geom_segment(aes(y = as.numeric(cause_of_death)-0.45, yend = as.numeric(cause_of_death)+0.45, x = deaths_mean, xend = deaths_mean), size = 0.8, colour = "grey1") +
+  scale_fill_viridis_d() +
+  facet_grid(year~place_of_death) +
+  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
+  scale_y_discrete(labels = ~str_wrap(.x, width = 20)) +
+  theme(
+    legend.position = "none", 
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    legend.title = element_blank(), 
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
+    ) +
+  labs(
+    x = NULL, y = NULL,
+    title = "Annual deaths by cause and place of death, compared to 2015-2019",
+    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2022_text)), collapse="\n"),
+    caption = paste0(c(death_registrations_text, source_nrs_text),collapse="\n")
+  ) +
+  NULL)
+
+fig_annual_deaths_by_cause %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_annual_deaths_by_cause", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
+```
+
 # Causes of death over time
@@
 ```{r}
-figure_cause_of_death_annual_proportion_by_place <-
+fig_cause_of_death_annual_proportion_by_place <-
   current_and_historical_deaths_by_cause_and_location %>%
@@
 
-figure_cause_of_death_annual_proportion_by_place %>%
+fig_cause_of_death_annual_proportion_by_place %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_annual_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
@@
 ```{r}
-figure_cause_of_death_weekly_proportion_by_place <-
+fig_cause_of_death_weekly_proportion_by_place <-
   merged_deaths_cause_of_death %>%
@@
 
-figure_cause_of_death_weekly_proportion_by_place %>%
+fig_cause_of_death_weekly_proportion_by_place %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
@@
 ```{r}
-figure_weekly_proportion_of_death_by_place_with_2015_2019 <-
+(fig_weekly_proportion_of_death_by_place_with_2015_2019 <-
   bind_rows(
@@
   NULL
-
-figure_weekly_proportion_of_death_by_place_with_2015_2019 %>%
+)
+fig_weekly_proportion_of_death_by_place_with_2015_2019 %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place_2015_2019", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
@@
 
-figure_weekly_deaths_by_cause_at_home <-
+fig_weekly_deaths_by_cause_at_home <-
   merged_deaths_cause_of_death %>%
@@
 
-figure_weekly_deaths_by_cause_at_home %>%
+fig_weekly_deaths_by_cause_at_home %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_deaths_by_cause_at_home", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 18, dpi = 300, units = "cm")
@@
 ```{r}
-figure_place_of_death_deviation_from_mean <-
+fig_place_of_death_deviation_from_mean <-
   merged_deaths_overall %>%
@@
 
-figure_place_of_death_deviation_from_mean %>%
+fig_place_of_death_deviation_from_mean %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_deviation_from_mean", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
@@
 
-# Home deaths by HB over time: 2020-2022, 4-week average
-
-```{r}
-(figure_home_deaths_over_time_by_hb_4_week_average <-
+# Home deaths by geography over time: 2020-2022, 4-week average
+
+## Home deaths by HB, 4-week average
+
+```{r}
+(fig_home_deaths_over_time_by_hb_4_week_average <-
   merged_deaths_hb %>%
@@
 
-figure_home_deaths_over_time_by_hb_4_week_average %>%
+fig_home_deaths_over_time_by_hb_4_week_average %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 20, dpi = 300, units = "cm")
@@
 
+## Home deaths by LA, 4-week average
+
+```{r}
+(fig_home_deaths_over_time_by_la_4_week_average <-
+  merged_deaths_la %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  pivot_longer(cols = c(ma4w_deaths_all_causes, ma4w_deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max, fill = "#4477aa"), alpha = 0.5) +
+  # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), size = 0.2) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  # scale_x_continuous(breaks = seq(0,60,by=10)) +
+  facet_wrap(~ref_area, scales = "free_y") +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  scale_linetype_manual(values = c("dashed"), labels = "Weekly average") +
+  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+    ) +
+  labs(x = NULL,
+       y = "Deaths (average of past 4 weeks)",
+       title = "Deaths at Home*: 4-week average by Scottish Local Authority",
+       subtitle = date_range_text,
+       caption = paste0(c(
+         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text,death_registrations_text,source_nrs_text
+         ), collapse="\n")
+       ) +
+  ## annotate historic range
+  annotate(geom = "text", x=ymd("2021-01-07"), y=0, label = "2021", hjust = 0, vjust = -1, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=0, label = "'22", hjust = 0, vjust = -1, colour = "grey40") +
+  NULL
+ )
+
+fig_home_deaths_over_time_by_la_4_week_average %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 40, height = 20, dpi = 300, units = "cm")
+```
+
 # Geographical variations
@@
 ```{r}
-figure_proportion_of_deaths_by_place_by_hb <-
+(fig_proportion_of_deaths_by_place_by_hb <-
   proportions_of_deaths_by_place_by_geography %>%
@@
     ) +
-  NULL
-
-figure_proportion_of_deaths_by_place_by_hb %>%
+  NULL)
+
+fig_proportion_of_deaths_by_place_by_hb %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_by_place_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
@@
 
-### Home deaths only, HB
-
-```{r}
-figure_proportion_of_deaths_at_home_by_hb <-
+## Proportion of deaths by place, by LA
+
+```{r}
+(fig_proportion_of_deaths_by_place_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
+  geom_point(aes(size = place_of_death)) +
+  geom_line(aes(size = place_of_death)) +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80"),
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    ) +
+  guides(colour = guide_legend(nrow=2, title = NULL), 
+         size = guide_legend(nrow=2, title = NULL)) +
+  labs(
+    x = NULL, y= "Proportion of yearly deaths (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL)
+
+fig_proportion_of_deaths_by_place_by_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_by_place_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Home deaths only, HB
+
+```{r}
+(fig_proportion_of_deaths_at_home_by_hb <-
   proportions_of_deaths_by_place_by_geography %>%
@@
     ) +
-  NULL
-
-figure_proportion_of_deaths_at_home_by_hb %>%
+  NULL)
+
+fig_proportion_of_deaths_at_home_by_hb %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Home deaths only, LA
+
+```{r}
+(fig_proportion_of_deaths_at_home_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
+  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
+  geom_point() +
+  geom_line() +
+  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80"),
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    ) +
+  labs(
+    linetype = NULL,
+    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
+    subtitle = paste0("* ",only_2022_text),
+    caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL)
+
+fig_proportion_of_deaths_at_home_by_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Map-based visualisations
+
+## Re-merge weekly deaths with 2015-2019 data for 2022 for LA & HB
+
+```{r}
+## merge the 2015-2019 period historical weekly averages to 2022 data
+merged_deaths_la_historic_period_2015_2019 <-
+  bind_rows(
+    merged_deaths_la %>% filter(year %in% 2020:2021),
+    merged_deaths_la %>% filter(year == 2022) %>%
+      select(-c(deaths_mean, deaths_min, deaths_max)) %>%
+      mutate(historic_period = "2015-2019") %>%
+      left_join(historical_weekly_average_la, by = c("week_number", "place_of_death", "ref_area", "historic_period"))
+  ) %>%
+  left_join(weekly_covid_cases_la, by = c("year","week_number","date_w_c","week_number_run_over","ref_area"))
+
+merged_deaths_hb_historic_period_2015_2019 <-
+  bind_rows(
+    merged_deaths_hb %>% filter(year %in% 2020:2021),
+    merged_deaths_hb %>% filter(year == 2022) %>%
+      select(-c(deaths_mean, deaths_min, deaths_max)) %>%
+      mutate(historic_period = "2015-2019") %>%
+      left_join(historical_weekly_average_hb, by = c("week_number", "place_of_death", "ref_area", "historic_period"))
+  ) %>%
+  left_join(weekly_covid_cases_hb, by = c("year","week_number","date_w_c","week_number_run_over","ref_area"))
+```
+
+### Home deaths increase since 2019 by LA
+
+For this we can take the total annual deaths in each LA and compare to historical figures, showing the ratio for each year so far; alternatively we can also average over 2020-2022, and compare that to the average!
+Because we only have partial data for 2022, we need to employ the weekly comparison method -
+
+```{r}
+(fig_map_home_deaths_increase_from_2020_la <-
+  merged_deaths_la_historic_period_2015_2019 %>%
+  filter(historic_period == "2015-2019") %>%
+  group_by(ref_area, place_of_death) %>%
+  summarise(
+    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
+    percent_increase = ratio_deaths_to_historic - 1,
+    .groups = "drop"
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  labs(x="", y="",
+       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority", width=40),
+       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease relative\nto 2015-2019"
+       ) +
+  NULL)
+
+fig_map_home_deaths_increase_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
+```
+
+### Home deaths increase since 2019 by HB
+
+```{r}
+(fig_map_home_deaths_increase_from_2020_hb <-
+  merged_deaths_hb_historic_period_2015_2019 %>%
+  filter(historic_period == "2015-2019") %>%
+  group_by(ref_area, place_of_death) %>%
+  summarise(
+    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
+    percent_increase = ratio_deaths_to_historic - 1,
+    .groups = "drop"
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
+  labs(x="", y="",
+       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board", width=40),
+       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
+       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease relative\nto 2015-2019"
+       ) +
+  NULL)
+
+fig_map_home_deaths_increase_from_2020_hb %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_from_2020_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
+```
+
+### Home deaths increase since 2019 by LA, by year
+
+For this we can take the total annual deaths in each LA and compare to historical figures, showing the ratio for each year so far; alternatively we can also average over 2020-2022, and compare that to the average!
+Because we only have partial data for 2022, we need to employ the weekly comparison method -
+
+```{r}
+(fig_map_home_deaths_increase_by_year_la <-
+  merged_deaths_la_historic_period_2015_2019 %>%
+  filter(historic_period == "2015-2019") %>%
+  group_by(year, ref_area, place_of_death) %>%
+  summarise(
+    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
+    percent_increase = ratio_deaths_to_historic - 1,
+    .groups = "drop"
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(
+    axis.text = element_blank(), 
+    panel.grid = element_blank(),
+    # legend.position = "bottom"
+    ) +
+  facet_wrap(~year, nrow = 1) +
+  labs(x="", y="",
+       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
+       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease relative\nto 2015-2019"
+       ) +
+  NULL)
+
+fig_map_home_deaths_increase_by_year_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_by_year_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 32, height = 18, dpi = 300, units = "cm")
+```
+
+### Home deaths increase vs. covid cases by LA, all deaths since 2019
+
+```{r}
+(fig_map_covid_cases_from_2020_la <-
+  weekly_covid_cases_la %>%
+  group_by(ref_area) %>%
+  summarise(
+    ratio_covid_cases = max(crude_rate_positive_per100k),
+    .groups = "drop"
+  ) %>% 
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  labs(x="", y="",
+       title = str_wrap("Total COVID-19 cases since the start of 2020, by Local Authority", width=40),
+       subtitle = "",
+       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
+       ) +
+  NULL)
+
+fig_map_covid_cases_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_covid_cases_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
+
+## combine maps
+(fig_map_home_deaths_increase_vs_covid_case_from_2020_la <-
+  wrap_plots(
+    fig_map_home_deaths_increase_from_2020_la +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+      ,
+    fig_map_covid_cases_from_2020_la +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+  )
+)
+
+fig_map_home_deaths_increase_vs_covid_case_from_2020_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_vs_covid_case_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 16, dpi = 300, units = "cm")
+```
+
+### Home deaths increase since 2019 by HB, by year
+
+See above for how this was done for LA!
+
+```{r}
+(fig_map_home_deaths_increase_by_year_hb <-
+  merged_deaths_hb_historic_period_2015_2019 %>%
+  filter(historic_period == "2015-2019") %>%
+  group_by(year, ref_area, place_of_death) %>%
+  summarise(
+    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
+    percent_increase = ratio_deaths_to_historic - 1,
+    .groups = "drop"
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  theme(
+    axis.text = element_blank(), 
+    panel.grid = element_blank(),
+    aspect.ratio = 1.4
+    ) +
+  facet_wrap(~year, nrow = 1) +
+  labs(x="", y="",
+       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board",
+       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       fill = "Home death\nincrease relative\nto 2015-2019"
+       ) +
+  NULL)
+
+fig_map_home_deaths_increase_by_year_hb %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_by_year_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 32, height = 18, dpi = 300, units = "cm")
+```
+
+### Home deaths increase vs. covid cases by HB, all deaths since 2019
+
+```{r}
+(fig_map_covid_cases_from_2020_hb <-
+  weekly_covid_cases_hb %>%
+  group_by(ref_area) %>%
+  summarise(
+    ratio_covid_cases = max(crude_rate_positive_per100k),
+    .groups = "drop"
+  ) %>% 
+  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
+  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
+  geom_sf(colour="grey10") +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  theme(
+    axis.text = element_blank(), 
+    panel.grid = element_blank(),
+    aspect.ratio = 1.4
+    ) +
+  labs(x="", y="",
+       title = str_wrap("Total COVID-19 cases since the start of 2020, by NHS Health Board", width=40),
+       subtitle = "",
+       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
+       ) +
+  NULL)
+
+fig_map_covid_cases_from_2020_hb %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_covid_cases_from_2020_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
+
+## combine maps
+(fig_map_home_deaths_increase_vs_covid_case_from_2020_hb <-
+  wrap_plots(
+    fig_map_home_deaths_increase_from_2020_hb +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+      ,
+    fig_map_covid_cases_from_2020_hb +
+      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
+  )
+)
+
+fig_map_home_deaths_increase_vs_covid_case_from_2020_hb %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_vs_covid_case_from_2020_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 16, dpi = 300, units = "cm")
+```
+
+#### LA covid rates over time
+
+I didn't finish this plot because the scale goes from 500 per 100k to 25k per 100k, which is a pain to depict in a meaningful way. Mayeb with a separate scale for each plot?
+
+```{r}
+## the problem with this plot is that there were between 500 per 100k cases to 25k per 100k cases over time and over LAs - how do we visualise that?
+(fig_map_covid_cases_by_year_la <-
+  weekly_covid_cases_la %>%
+  left_join(population_estimate_la_2019 %>% rename(population_estimate = all_ages, year_estimate = year), by = c("ref_area"="la")) %>%
+  group_by(ref_area, year, population_estimate) %>%
+  summarise(
+    cumulative_cases = sum(weekly_positive, na.rm=TRUE),
+    ratio_covid_cases = cumulative_cases / population_estimate * 1e5,
+    .groups = "drop"
+  ) %>%
+  distinct %>%
+  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
+  # ggplot(aes(geometry=geometry, fill=log(ratio_covid_cases)), colour="grey10") +
+  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
+  geom_sf(colour="grey10") +
+  # scale_fill_viridis_b(option = "B", labels = function(x) format(exp(x), big.mark=",")) +
+  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  theme(axis.text = element_blank(), panel.grid = element_blank()) +
+  facet_wrap(~year, nrow = 1) +
+  labs(x="", y="",
+       title = str_wrap("Total COVID-19 cases since the start of 2020 per 100,000 population, by Local Authority", width=50),
+       subtitle = "",
+       caption = paste0(c(source_phs_text), collapse="\n"),
+       fill = str_wrap("Cases per 100,000 population", width = 15)
+       ) +
+  NULL)
 ```

### 2022-06-10 covid_nrs_place_of_death_data_2022.Rmd
fixed the table of ratios of current deaths to historic - didn't account for histoirc periods and resulted in double entries in some cases

@@
         period = glue::glue("Data from week 1, 2020 to week {weeks_available_2022}, 2022"),
+        historic_period = "2015-2019",
         deaths_all_causes = sum(deaths_all_causes),
@@
              ) %>% 
-      group_by(place_of_death) %>%
+      group_by(place_of_death, historic_period) %>%
       summarise(
@@
       historical_annual_average_overall %>% 
-        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
-      by = "place_of_death"
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean, historic_period),
+      by = c("place_of_death","historic_period")
     )
@@
         period = "2020, week 12 - week 53",
+        historic_period = "2015-2019",
         deaths_all_causes = sum(deaths_all_causes),
@@
         period = "2021",
+        historic_period = "2015-2019",
         deaths_all_causes = sum(deaths_all_causes),
@@
       historical_annual_average_overall %>% 
-        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean),
-      by = "place_of_death"
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean, historic_period),
+      by = c("place_of_death","historic_period")
+    )
+    ,
+    ## 2022 data so far
+    merged_deaths_overall %>%
+      filter(
+        age == "all" &
+          sex == "all" &
+          year == 2022
+      ) %>%
+      group_by(place_of_death, week_number) %>%
+      summarise(
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+    mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      weekly_deaths_2015_2019_overall %>% filter(year != "average") %>% group_by(week_number, place_of_death) %>% summarise(
+                                         average_weekly_deaths = sum(number_of_deaths) / 5,
+                                         .groups = "drop"
+                                       ),
+      by = c("week_number", "place_of_death")
+    ) %>% 
+    group_by(place_of_death) %>%
+    summarise(
+      period = glue::glue("2022 up to week {weeks_available_2022}"),
+      historic_period = "2015-2019",
+      deaths_all_causes = sum(deaths_all_causes),
+      deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
+      deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+      mean_annual_deaths_2015_2019 = sum(average_weekly_deaths),
+      .groups = "drop"
     )
@@
 
-##
-
 ## By HB/LA

### 2022-06-10 download_latest_data_and_render_reports.R
added quiet = TRUE flag to the render() commands in the starting script

@@ -264,25 +264,27 @@ if (!any(fonttable()$FullName=="Calibri")) font_import(pattern = "calibri", prom
 dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
 
-render(
+rmarkdown::render(
   input = "./covid_nrs_place_of_death_data_2022.Rmd",
   output_file = paste0("covid_nrs_place_of_death_data_",today()),
   output_dir = dir_reports,
-  envir = new.env()
+  envir = new.env(),
+  quiet = TRUE
 )
 
-render(
+rmarkdown::render(
   input = "./covid_nrs_place_of_death_visualisations_2022.Rmd",
   output_file = paste0("covid_nrs_place_of_death_visualisations",today()),
   output_dir = dir_reports,
-  envir = new.env()
+  envir = new.env(),
+  quiet = TRUE
 )
 # 
 # render(
 #   input = "./paper_reporting_death_at_home_increase.Rmd",
 #   output_file = paste0("paper_reporting_death_at_home_increase_",today()),
 #   output_dir = dir_reports,
-#   envir = new.env()
+#   envir = new.env()https
 # )
 # 
 # render(

### 2022-06-27 covid_nrs_place_of_death_visualisations_2022.Rmd
tidied up visualisations a bit, and changed direction of colour scale so that high cases/home deaths = dark; also changed the parallel maps to have same colour coding

@@
 
-## Map-based visualisations
+# Map-based visualisations
 
@@
 
-### Home deaths increase since 2019 by LA
+## Home deaths increase since 2019 by LA
 
@@
   # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
-  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
   theme(axis.text = element_blank(), panel.grid = element_blank()) +
@@
 
-### Home deaths increase since 2019 by HB
+## Home deaths increase since 2019 by HB
 
@@
   # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
-  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
   theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
@@
 
-### Home deaths increase since 2019 by LA, by year
+## Home deaths increase since 2019 by LA, by year
 
@@
   geom_sf(colour="grey10") +
-  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
-  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
   theme(
@@
     panel.grid = element_blank(),
-    # legend.position = "bottom"
     ) +
@@
 
-### Home deaths increase vs. covid cases by LA, all deaths since 2019
+## Home deaths increase vs. covid cases by LA, all deaths since 2019
 
@@
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
   theme(axis.text = element_blank(), panel.grid = element_blank()) +
@@
 
-### Home deaths increase since 2019 by HB, by year
+## Home deaths increase since 2019 by HB, by year
 
@@
   geom_sf(colour="grey10") +
-  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
-  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1)) +
+  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
   theme(
@@
 
-### Home deaths increase vs. covid cases by HB, all deaths since 2019
+## Home deaths increase vs. covid cases by HB, all deaths since 2019
 
@@
   geom_sf(colour="grey10") +
-  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
   theme(
@@
 
-#### LA covid rates over time
+### LA covid rates over time
 
@@
   left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
-  # ggplot(aes(geometry=geometry, fill=log(ratio_covid_cases)), colour="grey10") +
   ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
   geom_sf(colour="grey10") +
-  # scale_fill_viridis_b(option = "B", labels = function(x) format(exp(x), big.mark=",")) +
-  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
+  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
   theme(axis.text = element_blank(), panel.grid = element_blank()) +
@@
 
-## cumulative deaths at home over time
+# Cumulative deaths at home over time
 

### 2022-07-22 covid_nrs_place_of_death_data_2022.Rmd
modified script to account for change in data - from mid-July 2022 the report no longer includes cause of death figures for 2021

@@
 
-Causes of death are reported in `sheet 8` of the report, and include figure from 2021 and 2022.
-The worksheet contains data split by location of death, which are contained within different tables arranged vertically, so their boundaries need to be written out; the tables have been pre-filled for all of 2021 and 2022, so some proportion of entries will be blank for 2022.
+Causes of death are reported in `sheet 8` of the report, ~~and include figure from 2021 and 2022.
+The worksheet contains data split by location of death, which are contained within different tables arranged vertically, so their boundaries need to be written out; the tables have been pre-filled for all of 2021 and 2022, so some proportion of entries will be blank for 2022.~~
+
+Note: the above no longer holds as of 20 July 2022 - only 2022 data re included in `sheet 8` now, which means the 2021 data need to be imported from an older file.
 
@@
 
+# TODO: delete below, it's for the previous releases that included 2021 data also
+# cell_ranges_for_causes_of_death_data <- list(
+#   "All" = "A6:V110",
+#   "Care home" = "A113:V217",
+#   "Home & other non-institution" = "A220:V324",
+#   "Hospital" = "A327:V431",
+#   "Other" = "A434:V538"
+# )
+
 cell_ranges_for_causes_of_death_data <- list(
-  "All" = "A6:V110",
-  "Care home" = "A113:V217",
-  "Home & other non-institution" = "A220:V324",
-  "Hospital" = "A327:V431",
-  "Other" = "A434:V538"
-)
-
-weekly_cause_of_death_2021_onwards <-
+  "All" = "A6:V58",
+  "Care home" = "A61:V113",
+  "Home & other non-institution" = "A116:V168",
+  "Hospital" = "A171:V223",
+  "Other" = "A226:V278"
+)
+
+weekly_deaths_cause_of_death_2022_onwards <-
   map_dfr(
@@
     cause_of_death = return_pretty_cause_of_death(cause_of_death),
-    date_w_c = as.Date(date_w_c),
+    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
     week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
@@
 
-### Historical causes of death
-
-Historical here refers to the 2015-2019 period for 2020 & 2021, and 2016-2021 (skipping 2020) for 2022. There is another 'historical' dataset, which includes cause of death figures for 2020.
-
-In the `2015-2019` period there were no Covid-19 deaths, so the historical average should be `NA`, which can possibly be interpreted as 0 for the purposes of comparing a year to the historical average.
-For the `2016-2019, 2021` period the average number of covid deaths should be equivalent to the deaths in 2021 - we wouldn't want to make an average between 4 years of 0 deaths and one year with any deaths!
-Similarly if the period were `2017-2019, 2021-2022`, we would compute the average for 2021-2022, and so on.
-
-```{r}
-## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
-## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
-## however, for the latter, we set the dates to 2022
-average_weekly_covid_deaths_2021_onwards <-
-  bind_rows(
-    ## set 2021 averages to NA
-    weekly_cause_of_death_2021_onwards %>%
-    filter(cause_of_death == "COVID-19" & year == 2021) %>%
-    mutate(number_of_deaths = NA_real_) %>%
-    rename(deaths_mean = number_of_deaths),
-    ## use 2021 weekly deaths as 2022 historic average
-    weekly_cause_of_death_2021_onwards %>%
-    filter(cause_of_death == "COVID-19" & year == 2021) %>%
-    mutate(year = 2022) %>%
-    select(-c(week_number_run_over, date_w_c)) %>%
-    rename(deaths_mean = number_of_deaths) %>%
-    inner_join(weekly_cause_of_death_2021_onwards %>% 
-                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
-              )
-  ) %>%
-  mutate(historic_period = historic_period_depending_on_year(year))
-
-historical_average_cause_of_death_2021_onwards <-
-  bind_rows(
-    ## the hacked averages for 2021 and 2022 above
-    average_weekly_covid_deaths_2021_onwards,
-    ## extract averages from most recent file for 2021 onwards
-    map_dfr(
-      .x = cell_ranges_for_causes_of_death_data,
-      .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
-      .id = "place_of_death"
-    ) %>%
-    janitor::clean_names() %>%
-    rename(year = registration_year, date_w_c = week_beginning) %>%
-    select(place_of_death, year, week_number, date_w_c, matches("five_year_average")) %>%
-    pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "deaths_mean") %>%
-    mutate(
-      year = as.integer(year),
-      cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
-      cause_of_death = return_pretty_cause_of_death(cause_of_death),
-      cause_of_death = if_else(cause_of_death == "All causes", "All", cause_of_death),
-      date_w_c = as.Date(date_w_c),
-      week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
-      historic_period = historic_period_depending_on_year(year)
-    ) %>%
-    select(-name) %>%
-    drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future
-  )
-```
-
 ### Data from 2020 & 2021
 
-The reports available from NRS in 2022 include data from 2021-2022, but we can also include data from 2020 for context. For this I'll load the last 'main report' file from 2021.
+The reports available from NRS in 2022 include data from 2021-2022 up until week 25, and partly in weeks 26 and 27 where there is data from 2021 for specific places of death but not for all places of death.
+I also include data from 2020 for context.
+For this I'll load the last 'main report' file from 2021.
 
@@
 
-#### 'Main report' file
-
-```{r}
+The weekly causes of death data for 2021 is included in the week 25 'main report' file, table `8`.
+
+#### 2020
+
+```{r}
+# this uses the combined 2020-2021 file
+# nrs_weekly_data_sheet_2020 <- 
+#   excel_sheets(files_historical_deaths$`2020-2021`)[which(str_detect(tolower(excel_sheets(files_historical_deaths$`2020-2021`)), pattern = "table 3.*2020"))]
+
 nrs_weekly_data_sheet_2020 <- 
-  excel_sheets(files_historical_deaths$`2020-2021`)[which(str_detect(tolower(excel_sheets(files_historical_deaths$`2020-2021`)), pattern = "table 3.*2020"))]
+  excel_sheets(files_historical_deaths$`2020`)[which(str_detect(tolower(excel_sheets(files_historical_deaths$`2020`)), pattern = "table 3.*2020"))]
 
@@
 weekly_deaths_cause_of_death_2020 <-
-  readxl::read_excel(files_historical_deaths$`2020-2021`, sheet = nrs_weekly_data_sheet_2020, col_names = headers_in_nrs_data_2020, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
+  readxl::read_excel(files_historical_deaths$`2020`, sheet = nrs_weekly_data_sheet_2020, col_names = headers_in_nrs_data_2020, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
   select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
@@
 
-#### Merge 2020 data with current data
+#### 2021
+
+```{r}
+cell_ranges_for_causes_of_death_data_2021 <- list(
+  "All" = "A6:V58",
+  "Care home" = "A113:V165",
+  "Home & other non-institution" = "A220:V272",
+  "Hospital" = "A327:V379",
+  "Other" = "A434:V486"
+)
+
+weekly_deaths_cause_of_death_2021 <-
+  map_dfr(
+    .x = cell_ranges_for_causes_of_death_data_2021,
+    .f = ~readxl::read_excel(files_historical_deaths$`2021`, sheet = sheet_cause_of_death_data, range = .x), 
+    .id = "place_of_death"
+  ) %>%
+  janitor::clean_names() %>%
+  rename(year = registration_year, date_w_c = week_beginning) %>%
+  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
+  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
+  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
+  mutate(
+    year = as.integer(year),
+    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
+    cause_of_death = return_pretty_cause_of_death(cause_of_death),
+    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
+    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+  ) %>%
+  select(-name) %>%
+  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
+```
+
+#### Historical causes of death
+
+Historical here refers to the 2015-2019 period for 2020 & 2021, and 2016-2021 (skipping 2020) for 2022, useful for a comparison with the historical context of the pre-pandemic years.
+(There is another 'historical' dataset, which includes cause of death figures for 2020 and for 2021.)
+
+In the `2015-2019` period there were no Covid-19 deaths, so the historical average should be `NA`, which can possibly be interpreted as 0 for the purposes of comparing a year to the historical average.
+For the `2016-2019, 2021` period the average number of covid deaths should be equivalent to the deaths in 2021 - we wouldn't want to make an average between 4 years of 0 deaths and one year with any deaths!
+Similarly if the period were `2017-2019, 2021-2022`, we would compute the average for 2021-2022, and so on.
+The year 2020 is considered exceptional and so is not used for computing the historical averages in 2021 and onwards.
+
+Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the
+
+```{r}
+## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
+## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
+## however, for the latter, we set the dates to 2022
+# average_weekly_covid_deaths_2021_onwards <-
+average_weekly_covid_deaths_2021_onwards <-
+  bind_rows(
+    ## set 2021 averages to NA
+    weekly_deaths_cause_of_death_2021 %>%
+    filter(cause_of_death == "COVID-19") %>%
+    mutate(number_of_deaths = NA_real_) %>%
+    rename(deaths_mean = number_of_deaths),
+    ## use 2021 weekly deaths as 2022 historic average
+    weekly_deaths_cause_of_death_2021 %>%
+    filter(cause_of_death == "COVID-19") %>%
+    mutate(year = 2022) %>%
+    select(-c(week_number_run_over, date_w_c)) %>%
+    rename(deaths_mean = number_of_deaths) %>%
+    inner_join(weekly_deaths_cause_of_death_2022_onwards %>% 
+                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
+              )
+  ) %>%
+  mutate(historic_period = historic_period_depending_on_year(year))
+
+average_weekly_noncovid_deaths_2021_onwards <-
+    ## extract 5-year averages from most recent file for 2021 onwards
+    ## this doesn't include covid, which we've manually extracted above!
+  bind_rows(
+    ## 2022+ data
+    map_dfr(
+      .x = cell_ranges_for_causes_of_death_data,
+      .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
+      .id = "place_of_death"
+    ),
+    ## 2021 data
+    map_dfr(
+      .x = cell_ranges_for_causes_of_death_data_2021,
+      .f = ~readxl::read_excel(files_historical_deaths$`2021`, sheet = sheet_cause_of_death_data, range = .x), 
+      .id = "place_of_death"
+    )
+  ) %>%
+    janitor::clean_names() %>%
+    rename(year = registration_year, date_w_c = week_beginning) %>%
+    select(place_of_death, year, week_number, date_w_c, matches("five_year_average")) %>%
+    pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "deaths_mean") %>%
+    mutate(
+      year = as.integer(year),
+      cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
+      cause_of_death = return_pretty_cause_of_death(cause_of_death),
+      cause_of_death = if_else(cause_of_death == "All causes", "All", cause_of_death),
+      date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
+      week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+      historic_period = historic_period_depending_on_year(year)
+    ) %>%
+    select(-name) %>%
+    drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future
+
+historical_average_cause_of_death_2021_onwards <-
+  bind_rows(
+    ## the hacked averages for 2021 and 2022 above
+    average_weekly_covid_deaths_2021_onwards,
+    ## the non-covid deaths in the 'main report' files
+    average_weekly_noncovid_deaths_2021_onwards
+  )
+```
+
+#### Merge 2020 & 2021 data with current data
 
@@
   bind_rows(
-    weekly_cause_of_death_2021_onwards,
+    weekly_deaths_cause_of_death_2022_onwards,
+    weekly_deaths_cause_of_death_2021,
     weekly_deaths_cause_of_death_2020 %>% 
@@
 deaths_by_cause_2021 <-
-  weekly_cause_of_death_2021_onwards %>% 
-  filter(year == 2021) %>% 
+  weekly_deaths_cause_of_death_2021 %>% 
   group_by(place_of_death, cause_of_death) %>% 

### 2022-07-22 download_latest_data_and_render_reports.R

@@ -81,13 +81,15 @@ if (file.exists(file_historical_data_sources)) {
   table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
 } else {
   table_of_historical_data_sources <- tibble(
-    short_name = c("overall","sex_age","la","hb","2020-2021"),
+    # short_name = c("overall","sex_age","la","hb","2020-2021"),
+    short_name = c("overall","sex_age","la","hb","2020","2021"),
     url = c(
       "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
-      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-21-data-week-52.xlsx"  # last 2021 data which included 2020 also
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-21-data-week-52.xlsx",  # last 2021 data which included 2020 also
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-22-data-week-25.xlsx"  # last 2022 data which included 2021 also
     )
   ) %>% mutate(
     # filename = basename(url),
@@ -301,3 +303,4 @@ rmarkdown::render(
 #   envir = new.env()
 # )
 
+message("Finished successfully!")

### 2022-10-24 covid_nrs_place_of_death_visualisations_2022.Rmd
started working on 'years skipped ahead' plot to illustrate the increase in home deaths

@@
 ```
+
+# Years 'skipped ahead' due to Covid
+
+There was a pre-pandemic trend of increasing deaths at home - by plotting a regression line we can compute the 'years skipped ahead' equivalent to the increase in deaths at home due to covid
+
+```{r}
+regression_data <- 
+  merged_proportions_of_deaths_by_place %>% 
+  filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>% 
+  mutate(years_since_2015 = 0:(n()-1))
+
+regression_model <- lm(data = regression_data, formula = number_of_deaths ~ years_since_2015)
+
+predicted_data <- tibble(years_since_2015 = seq(from=0, to=30, by = 0.1)) %>%
+  mutate(year = 2015 + years_since_2015,
+         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015))
+         )
+
+# TODO: make the plot nicer, and print out the predicted year somewhere
+# TODO: figure out how to adjust the incomplete 2022 data
+actual_data <- tibble(year = c(2020:2022)) %>% 
+  left_join(
+    merged_proportions_of_deaths_by_place %>%
+      select(year, number_of_deaths, place_of_death) %>%
+      filter(place_of_death == "Home & other non-institution"),
+    by = "year"
+  ) %>%
+  mutate(
+    predicted_year =  if_else(
+      condition = year != 2022,
+      true = 2015 + (number_of_deaths - coef(regression_model)[1]) / coef(regression_model)[2],
+      ## for 2022, we use slightly different computation
+      false = NA_real_
+    ),
+    label = paste0("Actual home deaths in ", year)
+  )
+
+ggplot(data = predicted_data, aes(x=year, y=predicted)) + 
+  geom_point(data = regression_data, aes(x = year, y = number_of_deaths)) +
+  geom_line(linetype = "dashed") +
+  ggrepel::geom_label_repel(data = actual_data, aes(x = predicted_year, y = number_of_deaths, label = label)) +
+  geom_vline(data = actual_data, aes(xintercept = predicted_year)) +
+  xlim(c(2015,2035)) +
+  theme_minimal() +
+  scale_y_continuous(labels = scales::comma)
+```

### 2022-10-27 functions.R
helper functions for computing new data from regression model

@@ -237,3 +237,12 @@ add_scotland_total_to_population_estimate <- function(data_tbl) {
   new_row$all_ages <- scotland_total
   bind_rows(data_tbl, new_row)
 }
+
+## compute y var from x, given a linear regression
+compute_y_from_x_using_regression <- function(x, regression_model) {
+  coef(regression_model)[1] + (x * coef(regression_model)[2])
+}
+## compute x var from y, given a linear regression
+compute_x_from_y_using_regression <- function(y, regression_model) {
+  (y - coef(regression_model)[1]) / coef(regression_model)[2]
+}

### 2022-10-27 covid_nrs_place_of_death_visualisations_2022.Rmd
completed a first draft version of the 'years skipped ahead' plot.

--- 2022-06-27 covid_nrs_place_of_death_visualisations_2022.Rmd
@@
 ```
+
+# Years 'skipped ahead' due to Covid
+
+There was a pre-pandemic trend of increasing deaths at home - by plotting a regression line we can compute the 'years skipped ahead' equivalent to the increase in deaths at home due to covid
+
+```{r}
+regression_data <- 
+  merged_proportions_of_deaths_by_place %>% 
+  filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>% 
+  mutate(years_since_2015 = 0:(n()-1))
+
+regression_model <- lm(data = regression_data, formula = number_of_deaths ~ years_since_2015)
+
+predicted_data <- tibble(years_since_2015 = seq(from=0, to=30, by = 0.1)) %>%
+  mutate(year = 2015 + years_since_2015,
+         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015))
+         )
+
+## 2022 only has X weeks' worth of data available; we can extrapolate the annual
+## total by computing the percentage of annual data represented by previous
+## years at X weeks, and then averaging across those, or taking the min & max
+extrapolated_2020_data <-
+  historical_weekly_average_overall %>%
+  filter(place_of_death == "Home & other non-institution" & historic_period == "2015-2019") %>%
+  summarise(
+    deaths_mean_so_far = sum(deaths_mean[week_number <= weeks_available_2022], na.rm= TRUE),
+    deaths_mean_total = sum(deaths_mean, na.rm= TRUE),
+    proportion_of_deaths_so_far = deaths_mean_so_far / deaths_mean_total
+    )
+
+predicted_data_annotations <- tribble(
+  ~x, ~y, ~label,
+  2021, 20e3, "Predicted home deaths 2020-2022 using 2015-2019 data",
+  2033, 23e3, "Actual home deaths 2020-2022*",
+  2033, 17.5e3, "If the 2015-2019 trend continued we wouldn't observe this number of home deaths until 2031",
+)
+
+actual_and_predicted_data <-
+  bind_rows(
+    merged_proportions_of_deaths_by_place %>%
+    select(year, number_of_deaths, place_of_death) %>%
+    filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
+      mutate(data_type = "Actual")
+    ,
+    merged_proportions_of_deaths_by_place %>%
+    select(year, number_of_deaths, place_of_death) %>%
+    filter(place_of_death == "Home & other non-institution" & year %in% 2020:2022) %>%
+    mutate(
+      data_type = "Predicted",
+      number_of_deaths = if_else(year %in% 2020:2022, compute_y_from_x_using_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
+      )
+    ,
+    merged_proportions_of_deaths_by_place %>%
+      select(year, number_of_deaths, place_of_death) %>%
+      filter(place_of_death == "Home & other non-institution" & year %in% 2020:2022) %>%
+      mutate(
+        number_of_deaths = if_else(year == 2022, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
+        year = 2015 + compute_x_from_y_using_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
+        data_type = "Actual"  # note: this is actual figures but
+      )
+  )
+
+fig_expected_home_deaths_regression <-
+  ggplot(data = actual_and_predicted_data, aes(x=year, y=number_of_deaths)) +
+  geom_rect(aes(xmin = 2019.5, xmax = 2022.5, ymin = 15.5e3, ymax = 18.5e3), fill = "yellow", alpha = 0.3) +
+  geom_rect(aes(xmin = 2030, xmax = 2035, ymin = 19.5e3, ymax = 22e3), fill = "yellow", alpha = 0.3) +
+  geom_text(data = predicted_data_annotations, aes(x=x,y=y,label=str_wrap(label, width = 40))) +
+  geom_point(aes(shape = data_type)) +
+  geom_line(data = predicted_data, aes(x = year, y = predicted), linetype = "dotted") +
+  theme_minimal() +
+  theme(legend.position = "top", legend.title = element_blank()) +
+  xlim(c(2015,2037)) +
+  scale_y_continuous(labels = scales::comma) +
+  labs(y = NULL, x = "Year (hypothetical)", title = "Increasing deaths at home", subtitle = str_wrap("Home deaths were on the increase before the pandemic.\nIf we assumed the same annual increases from 2015-2019 we wouldn't expect the pandemic levels of home deaths until 2031!", width = 90), caption = paste0(c(glue::glue("*Annual home deaths for 2022 extrapolated from data for weeks 1-{weeks_available_2022}"),source_nrs_text), collapse="\n"))
+
+fig_expected_home_deaths_regression %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_expected_home_deaths_regression", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 12, dpi = 300, units = "cm")
+```

### 2022-11-28 covid_nrs_place_of_death_visualisations_2022.Rmd
added place of death graph for all 3 PoD, using 2015-2019 historic range throughout, incl. 2022

@@
 
+## With 2015-2019 historic range only
+
+```{r}
+merged_deaths_overall_2015_2019_historic_period <-
+  bind_rows(
+    ## pre-2022 data
+    merged_deaths_overall %>% filter(year < 2022),
+    ## 2022 data
+    merged_deaths_overall %>%
+      filter(year >= 2022) %>%  # 2022 or later!?
+      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
+      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
+      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
+      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
+        current_data = ., 
+        historical_data = 
+          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
+        num_weeks_to_average = 4)
+  )
+
+fig_place_of_death_2015_2019_range <-
+  merged_deaths_overall_2015_2019_historic_period %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  geom_text(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
+  geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  NULL
+
+fig_place_of_death_2015_2019_range %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
+```
+
 ## Home deaths only
@@
 
-### Home deaths with 2015-2019 historic period means & range only
-
-```{r}
-merged_deaths_overall_2015_2019_historic_period <-
-  bind_rows(
-    ## pre-2022 data
-    merged_deaths_overall %>% filter(year < 2022),
-    ## 2022 data
-    merged_deaths_overall %>%
-      filter(year >= 2022) %>%  # 2022 or later!?
-      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
-      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
-      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
-      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
-        current_data = ., 
-        historical_data = 
-          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
-        num_weeks_to_average = 4)
-  )
-
+## Home deaths with 2015-2019 historic period means & range only
+
+```{r}
 fig_place_of_death_home_2015_2019_historic_period <-

### 2023-01-05 covid_nrs_place_of_death_data_2022.Rmd
fixed change in ref_period variable

@@
 
+## Note: as of 12 Dec 2022, the format of the ref_period variable changed from
+## e.g. "2020-W12" to "w/c 2019-12-31"
+## this means the w/c value is already provided
+## and also the week number is not provided
+
 weekly_deaths_overall <-
@@
   rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
-  filter(!ref_period %in% as.character(2020:2022)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
-  mutate(
-    year = parse_integer(str_extract(ref_period, pattern="^(202[0-3])")),
-    week_number = parse_integer(str_remove(ref_period, pattern = "^(202[0-3])\\-")),
-    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+  filter(!ref_period %in% as.character(2020:2023)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
+  parse_ref_period() %>%
+  mutate(
     number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
-    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
   ) %>%

### 2023-01-05 compile_medications_from_bnf.Rmd
added script to compile medications for dementia & pall.care from BNF

---
title: "Compiling list of medications based on BNF codes for dementia/AD & palliative care drugs"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

For the analysis of prescription/dispensing data, lists of specific medicines are needed to make the analysis tractable and applicable to dementia/Alzheimer's disease and for the analysis of anticipatory medicines used in palliative care.

## Source of BNF codes

One source for a list of all BNF codes is the Information Services Portal of the NHS Business Services Authority, https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126. This is available after putting in a captcha-like password, so I've downloaded the file manually.

File used: `01-01-2023 : 84`

# Load packages

```{r}
library(tidyverse)
library(readxl)  # for loading excel files
library(janitor)  # for cleaning data
library(knitr)  # for pretty tables in the report
```

# Download file

Note: this was done manually from the above website!

# Load BNF data

```{r}
## the .csv file is inside a .zip file!
bnf <- read_csv(file = unzip( "./bnf/20230104_1672833147341_BNF_Code_Information.zip", "20230104_1672833147341_BNF_Code_Information.csv")) %>%
  clean_names()
```

# Defining medicines

## Helper functions for searching BNF

```{r}
find_substance <- function(keyword) {
  bnf %>%
    filter(str_detect(bnf_chemical_substance, pattern = regex(keyword, ignore_case = TRUE))) %>%
    group_by(bnf_chemical_substance, bnf_paragraph) %>%
    summarise(
      n = n(),
      products = paste(bnf_presentation, collapse = ";")
    )
}
```

## Dementia / Alzheimer's disease

The medicines of interest are:

- donepezil
- galantamine
- rivastigmine
- memantine

~~I can manually search for the above keywords in the `chemical substance` field in the BNF and include all relevant medicines.~~

Note that the above drugs correspond to the entirety of the BNF section `4.11, Drugs for dementia`

```{r}
bnf_dementia <- 
  bnf %>%
  filter(bnf_section == "Drugs for dementia") %>%
  mutate(medication = word(bnf_chemical_substance, 1)) %>%
  select(medication, bnf_section, bnf_section_code, bnf_chemical_substance_code) %>%
  distinct
```

## Palliative care / anticipatory medicines

Marie Curie lists [the following anticipatory medicines](https://www.mariecurie.org.uk/professionals/palliative-care-knowledge-zone/symptom-control/anticipatory-medicines):

- Medicine for pain in palliative care – an appropriate opioid, for example, morphine, diamorphine, oxycodone or alfentanil.
- Medicine for breathlessness – midazolam or an opioid.
- Medicine for anxiety – midazolam.
- Medicine for delirium or agitation – haloperidol, levomepromazine, midazolam or phenobarbital.
- Medicine for nausea and vomiting – cyclizine, metoclopramide, haloperidol or levomepromazine.
- Medicine for noisy chest secretions – hyoscine hydrobromide or glycopyrronium.

Some of the above are specific chemicals, and some are classes of medicines (e.g. opioids). For opioids, some examples are listed, but I will include the entire class, i.e. the BNF section `4.7.2, Opioid analgesics`. For the other medicines, I will only include the specific substances listed.

Note that alfentanil is not listed as an opioid in the BNF but as a general anaesthetic!

```{r}
bnf_palliative <-
  ## pain: opioids + alfentanil
  list(
    bind_rows(
      bnf %>% filter(bnf_paragraph_code == "040702"),  # opioids
      bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("alfentanil", ignore_case = TRUE)))
    ) %>% mutate(medicine = "Pain")
    ,
    ## breathlessness: midazolam or opioid
    bind_rows(
      bnf %>% filter(bnf_paragraph_code == "040702"),  # opioids
      bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("midazolam", ignore_case = TRUE)))
    ) %>% mutate(medicine = "Breathlessness")
    ,
    ## anxiety
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("midazolam", ignore_case = TRUE))) %>% 
      mutate(medicine = "Anxiety")
    ,
    ## delirium/agitation
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("haloperidol|levomepromazine|midazolam|phenobarbital", ignore_case = TRUE))) %>% 
      mutate(medicine = "Delirium or agitation")
    ,
    ## nausea or vomiting
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("cyclizine|metoclopramide|haloperidol|levomepromazine", ignore_case = TRUE))) %>% 
      mutate(medicine = "Nausea or vomiting")
    ,
    ## noisy chest secretions
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("hyoscine hydrobromide|glycopyrronium", ignore_case = TRUE))) %>% 
      mutate(medicine = "Noisy chest secretions")
  ) %>%
  bind_rows() %>%
  select(medicine, bnf_chemical_substance, bnf_chemical_substance_code) %>%
  distinct %>%
  pivot_wider(names_from = medicine, values_from = medicine, names_prefix = "bin_", values_fill = FALSE, values_fn = ~TRUE) %>%
  clean_names()
```

# Conclusion

We are now left with two files listing medicines used in dementia/AD and palliative/anticipatory care; the palliative list is slightly more complex and it includes, for each substance, a number of binary variables indicating its use (e.g. pain, breathlessness, agitation, etc.)

# Save .csv files

```{r}
write_csv(bnf_dementia, file = "./bnf/bnf_codes_dementia_alzheimers.csv")

write_csv(bnf_palliative, file = "./bnf/bnf_codes_palliative_anticipatory.csv")
```

### 2023-01-05 compile_medications_from_bnf.Rmd
updated script a bit & results

@@
   distinct %>%
-  pivot_wider(names_from = medicine, values_from = medicine, names_prefix = "bin_", values_fill = FALSE, values_fn = ~TRUE) %>%
+  pivot_wider(names_from = medicine, values_from = medicine, names_prefix = "bin_", values_fill = 0L, values_fn = ~1L) %>%
   clean_names()

### 2023-01-05 functions.R
oops, forgot to include the date parsing script

@@ -246,3 +246,35 @@ compute_y_from_x_using_regression <- function(x, regression_model) {
 compute_x_from_y_using_regression <- function(y, regression_model) {
   (y - coef(regression_model)[1]) / coef(regression_model)[2]
 }
+
+## parse the date given in ref_period depending on format
+## this assumes there is a ref_period variable in the data!
+parse_ref_period = function(data_tbl) {
+  if (!"ref_period" %in% names(data_tbl)) 
+    stop("A 'ref_period' column not found in the data!")
+  ## this assumes the entirety of ref_period has the same format
+  ## I think it's unlikely for the data to come as a mix of the two formats
+  if (str_starts(data_tbl[["ref_period"]][1], pattern = "w\\/c")) {
+    ## continue with assumption that ref_period is given as w/c
+    data_tbl %>%
+      mutate(
+        date_w_c = 
+          ymd(str_extract(ref_period, pattern = "\\d{4}\\-\\d{1,2}\\-\\d{1,2}")),
+        year = as.integer(str_sub(ISOweek::ISOweek(date_w_c), 1, 4)),
+        week_number = lubridate::isoweek(date_w_c),
+        week_number_run_over = compute_week_run_over(week_number = week_number, year = year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+      )
+  }
+  else {
+    ## otherwise, ref_period is given in isoweek format: e.g. 2020-W01
+    data_tbl %>%
+      mutate(
+        year = parse_integer(str_extract(ref_period, pattern="^(202[0-3])")),
+        week_number = parse_integer(str_remove(ref_period, pattern = "^(202[0-3])\\-")),
+        date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
+        number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
+        week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+      )
+  }
+    
+}

### 2023-01-09 covid_nrs_place_of_death_data_2022.Rmd
There was an error reading the HB data due to that week's xlsx missing the first sheet, 'contents'. I've changed so I'm reading from the sheet named 'Data' instead of a hard-coded 2nd sheet in the file, so this kind of error won't break future analyses!

@@
 weekly_deaths_la <- 
-  read_excel(files_weekly_deaths$la, sheet = 2, skip = 2) %>% 
+  read_excel(files_weekly_deaths$la, sheet = "Data", skip = 2) %>% 
   clean_names %>%
@@
 weekly_deaths_hb <- 
-  read_excel(files_weekly_deaths$hb, sheet = 2, skip = 2) %>% 
+  read_excel(files_weekly_deaths$hb, sheet = "Data", skip = 2) %>% 
   clean_names %>%

### 2023-01-11 parsing_toc/example.Rmd
toy example for parsing .md files created alongside rendering rmarkdown files, generating a listing of 3rd level headings (###) and captions of tables using DT::datatable()

---
title: "Example of report with tables"
output: 
  html_document:
    keep_md: true
    number_sections: true
date: "2022-11-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(DT)
```

# Top level header

## Subheading L2 without much content

### What

### is this

## Subheading

### L3 subheading without a table

Just some lorem ipsum

### Table 1: SW characters shorter than 100

Some lorem ipsum that doesn't need to be in a listing of tables.

```{r}
starwars %>% filter(height<100) %>%
datatable(caption = "SW characters shorter than 100 units, presumably centimeters.")
```

### Table 2: SW characters taller than 200

Some more lorem ipsum that doesn't need to be in a listing of tables.

```{r}
starwars %>% filter(height>200) %>%
  datatable(caption = "SW characters taller than 200 units, presumably centimeters.")
```

## Subheading 2

### Table 2 contains `glue`

Some more lorem ipsum that doesn't need to be in a listing of tables.

```{r}
starwars %>% filter(height>200) %>%
  datatable(caption = glue::glue("SW characters taller than 200 units, presumably centimeters."))
```

### 2023-01-11 parsing_toc/parsing_markdown_toc.R

@@ -1,0 +1,61 @@
+# TODO: incorporate parsing statements containing glue!
+
+## for .Rmd - this can ignore the glue::glue text but won't incorporate the text
+## that glue would insert
+library(tidyverse)
+tibble(text = read_lines("./parsing_toc/example.Rmd")) %>%
+  # find your h3 tags to tie the output to
+  mutate(h3_cnt = cumsum(str_detect(text, "^### ")),
+         # find the titles
+         h3 = str_extract(text, "(?<=### ).*") %>% replace_na(""),
+         # find the captions
+         # caption = str_extract(text, "(?<=caption = \\\").+(?=\\\")") %>% replace_na("")
+         caption = 
+           str_extract(text, "(?<=caption).+(?=\\\")") %>% 
+           str_remove(pattern = "\\s*\\=\\s*(\\w+\\:{0,2}\\w*\\()*\\\"") %>% 
+           replace_na("")
+  ) %>%
+  group_by(h3_cnt) %>%
+  summarise(title = paste(h3, collapse = ""),
+            caption = paste(caption, collapse = ""))
+
+## using .md file
+tibble(text = read_lines("./parsing_toc/example.md")) %>%
+  ## preprocess: remove code blocks containing comments -
+  ## otherwise they are indistinguishable from markdown headings!
+  mutate(
+    open_block = cumsum(str_detect(text, "^\\`{3}\\.+")),
+    close_block = cumsum(str_detect(text, "^\\`{3}$"))
+  ) %>%
+  ## quick & dirty: remove lines that are code blocks and start with a '#'
+  ## because these indicate comments rather than headings!
+  filter(!(open_block>close_block & str_detect(text,"^#+"))) %>%
+  mutate(
+    h1 = cumsum(str_detect(text, "^# "))
+  ) %>% 
+  group_by(h1) %>%
+  mutate(
+    h2 = cumsum(str_detect(text, "^## "))
+  ) %>%
+  group_by(h1,h2) %>%  # this ensure counts are nested within h1 also
+  mutate(
+    h3 = cumsum(str_detect(text, "^### "))
+  ) %>%
+  ungroup %>%
+  mutate(
+    heading_number = glue::glue("{h1}.{h2}.{h3}"),
+    heading_text = str_extract(text, "(?<=### ).*") %>% replace_na(""),
+    # find the captions
+    caption = 
+     str_extract(text, "(?<=\\<caption\\>).+(?=\\<\\\\\\/caption\\>)") %>% 
[…]

### 2023-01-11 parsing_toc/parsing_markdown_toc.R
fleshed TOC parsing script out a bit

@@ -45,9 +45,10 @@ tibble(text = read_lines("./parsing_toc/example.md")) %>%
   mutate(
     heading_number = glue::glue("{h1}.{h2}.{h3}"),
     heading_text = str_extract(text, "(?<=### ).*") %>% replace_na(""),
-    # find the captions
+    # find the captions inside tags <caption> BLABLA <\/caption>
     caption = 
      str_extract(text, "(?<=\\<caption\\>).+(?=\\<\\\\\\/caption\\>)") %>% 
+      textutils::HTMLdecode() %>%  # decodes things like "&amp" to "&"
      replace_na("")
   ) %>%
   distinct %>%

### 2023-01-17 covid_nrs_place_of_death_data_2022.Rmd
started modifying 2022 code to work with 2023; I'll just make a new file instead

@@
   rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
-  filter(!ref_period %in% as.character(2020:2023)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
+  filter(!ref_period %in% as.character(2020:2029)) %>% # remove totals for 2020, 2021, 2022, etc., we only want the weekly figures
   parse_ref_period() %>%

### 2023-01-17 download_latest_data_and_render_reports.R
modified to download the newly released 2023 data; TODO: do all the rest of the scripts that assume data up to 2022...

@@ -27,7 +27,7 @@ all_nrs_links <-
   html_nodes("a") %>%
   html_attr("href")  # find all links on the page
 
-nrs_filename_pattern <- "covid-deaths-(21|22)-data.*xlsx"  # to accommodate 2022 data also once the filename changes!
+nrs_filename_pattern <- "covid-deaths-(2\\d)-data.*xlsx"  # this will last for the decade... 202X
 
 # TODO: incorporate both weekly and monthly data?
 weekly_deaths_url_relative <- # keep only links that are probably weekly deaths

### 2023-01-30 compile_medications_from_bnf.Rmd
I've included a list of medicines for polypharmacy definitions, reading it using docxtractor::read_docx(); this doesn't work with a downloaded file but can read straight from the URL! needed to specify path to LibreOffice binary for it to work

@@
 
+## List of discrete medications
+
+To count the number of different prescriptions or measure polypharmacy, we need to define discrete medications that can be prescribed/dispensed, whilst removing things like dressings, vaccines, etc.
+
+[Guthrie et al., 2015 ("The rising tide of polypharmacy and drug-drug interactions: population database analysis 1995–2010")](https://doi.org/10.1186/s12916-015-0322-7) define this in their Additional file, table S1, which I'll copy-paste and export as .csv below.
+
 # Load packages
@@
 library(knitr)  # for pretty tables in the report
+library(docxtractr)  # for reading data from table in .docx file
 ```
@@
 
+## BNF codes
+
 Note: this was done manually from the above website!
+
+## List of medicines
+
+Note: there appears to be an error with using `docxtractor::read_docx()` with a file downloaded to OneDrive; reading directly from the URL works fine, however!
+Therefore I've skipped downloading the file below.
+
+```{r}
+url_medicines <- "https://static-content.springer.com/esm/art%3A10.1186%2Fs12916-015-0322-7/MediaObjects/12916_2015_322_MOESM1_ESM.doc"
+
+# path_medicines <- file.path("./bnf", basename(url_medicines))
+# download.file(url = url_medicines, destfile = path_medicines)
+```
 
@@
   clean_names()
+```
+
+## List of medicines
+
+```{r}
+# doc <- read_docx(path_medicines)
+## error with reading from OneDrive folder?
+doc <- read_docx(url_medicines)
+extracted_table <- 
+  docx_extract_tbl(doc, 1)
 ```
@@
 
+## Discrete medicines from Guthrie et al., 2015 paper
+
+```{r}
+# TODO: process this if needed?
+medicines_list <- 
+  extracted_table %>%
+  clean_names() %>%
+  mutate(
+    chapter = as.integer(str_extract(drug_class_name, pattern = "(?<=Chapter )\\d+"))
+  ) %>%
+  filter(bnf_subsection!="") %>%
+  fill(chapter, .direction = "down")
+```
+
 # Conclusion

### 2023-01-30 download_latest_data_and_render_reports.R
changed script to use the 2023 data processing and visualisation scripts; those still need updating, however!

@@ -267,7 +267,7 @@ dir_reports <- "./reports"
 if (!dir.exists(dir_reports)) dir.create(dir_reports)
 
 rmarkdown::render(
-  input = "./covid_nrs_place_of_death_data_2022.Rmd",
+  input = "./covid_nrs_place_of_death_data_2023.Rmd",
   output_file = paste0("covid_nrs_place_of_death_data_",today()),
   output_dir = dir_reports,
   envir = new.env(),
@@ -275,7 +275,7 @@ rmarkdown::render(
 )
 
 rmarkdown::render(
-  input = "./covid_nrs_place_of_death_visualisations_2022.Rmd",
+  input = "./covid_nrs_place_of_death_visualisations_2023.Rmd",
   output_file = paste0("covid_nrs_place_of_death_visualisations",today()),
   output_dir = dir_reports,
   envir = new.env(),

### 2023-02-13 compile_medications_from_bnf.Rmd
added more testing code for BNF medicine compiling

@@
 library(docxtractr)  # for reading data from table in .docx file
+library(fuzzyjoin)  # for regex merges
 ```
@@
   ) %>%
-  filter(bnf_subsection!="") %>%
-  fill(chapter, .direction = "down")
+  fill(chapter, .direction = "down") %>%  # complete blanks between chapters with the last non-blank chapter
+  filter(bnf_subsection!="")  # remove chapters from the drug class column
+
+zero_pad <- function(x) str_pad(string = x, width = 2, pad = "0")
+
+test_list <-
+  medicines_list %>%
+  separate(col = bnf_subsection, into = paste0("c",1:4), sep = "\\.", remove = FALSE) %>%
+  mutate(across(c1:c3, zero_pad)) %>%  # zero-pad digits except for 4th character, c4
+  unite(col = "code", c1:c4, sep = "", na.rm = TRUE, remove = TRUE)
+```
+
+## Test against BNF to see if I get the same substances
+
+Note: these need to be done in sections by how long the code given is - does it pick up a chapter, a section, a subsection, or specific medicines only?
+
+```{r}
+test_merge <-
+  # test_list %>% 
+  # select(drug_class_name, bnf_subsection, code) %>%
+  # mutate(code = paste0("^",code)) %>%
+  # regex_right_join(bnf, by = c("code"="bnf_chemical_substance_code"))
+  bnf %>%
+  regex_right_join(
+    test_list %>%
+      select(drug_class_name, bnf_subsection, code) %>%
+      mutate(code = paste0("^",code)),
+    by = c(bnf_chemical_substance_code = "code")
+  )
+
+# View(test_merge)
+
+## what's left after matching all of the bnf to the provided codes?
+test_merge2 <-
+  bnf %>%
+  regex_anti_join(
+    test_list %>%
+      select(drug_class_name, bnf_subsection, code) %>%
+      mutate(code = paste0("^",code)),
+    by = c(bnf_chemical_substance_code = "code")
+  )
+test_merge2 %>% count(bnf_section, bnf_section_code)
 ```

### 2023-02-13 covid_nrs_place_of_death_data_2023.Rmd
editing historic periods for 2023 data (& beyond - fianlly writing abstract enough code!)

--- 2023-01-17 covid_nrs_place_of_death_data_2022.Rmd
@@
 ---
-title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2022 data"
+title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data"
 author: "Jan Savinc"
@@
 
-# TODO: delete below, it's for the previous releases that included 2021 data also
+## updated for 2023 which contains both 2022 and 2023
+cell_ranges_for_causes_of_death_data <- list(
+  "All" = "A6:V110",
+  "Care home" = "A113:V217",
+  "Home & other non-institution" = "A220:V324",
+  "Hospital" = "A327:V431",
+  "Other" = "A434:V538"
+)
+
+## this was for only 2022 data
 # cell_ranges_for_causes_of_death_data <- list(
-#   "All" = "A6:V110",
-#   "Care home" = "A113:V217",
-#   "Home & other non-institution" = "A220:V324",
-#   "Hospital" = "A327:V431",
-#   "Other" = "A434:V538"
+#   "All" = "A6:V58",
+#   "Care home" = "A61:V113",
+#   "Home & other non-institution" = "A116:V168",
+#   "Hospital" = "A171:V223",
+#   "Other" = "A226:V278"
 # )
 
-cell_ranges_for_causes_of_death_data <- list(
-  "All" = "A6:V58",
-  "Care home" = "A61:V113",
-  "Home & other non-institution" = "A116:V168",
-  "Hospital" = "A171:V223",
-  "Other" = "A226:V278"
-)
-
+# TODO: this will eventually only have 2023 data, change when that happens
 weekly_deaths_cause_of_death_2022_onwards <-
@@
 
+#### 2022
+
+```{r}
+# TODO: deal with 2022 data here, now that it's 'historical'
+cell_ranges_for_causes_of_death_data_2022 <- list(
+  "All" = "A6:V58",
+  "Care home" = "A61:V113",
+  "Home & other non-institution" = "A116:V168",
+  "Hospital" = "A171:V223",
+  "Other" = "A226:V278"
+)
+
+```
+
 #### Historical causes of death
 
-Historical here refers to the 2015-2019 period for 2020 & 2021, and 2016-2021 (skipping 2020) for 2022, useful for a comparison with the historical context of the pre-pandemic years.
-(There is another 'historical' dataset, which includes cause of death figures for 2020 and for 2021.)
+From the [methodology for computing excess deaths during the covid pandemic](https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-methodology-excess-deaths.pdf):
+
+> For 2022, the excess deaths calculation would usually incorporate the mortality data for the most recent 5 year period (2017 – 2021).
+However, given the unusual nature of the 2020 data and the decision previously taken for 2021, it has been agreed with ONS that the excess deaths calculation will compare 2022 mortality data with the average of 2016, 2017, 2018, 2019 and 2021, and therefore continue to exclude 2020.
+Similarly for 2023, the five year average will again leave out 2020 and will be built from the data from 2017, 2018, 2019, 2021 and 2022.
+
+'Historical' here refers to the 5-year period preceding each year, useful for a historic comparison. The historic period depedning on year has been defined as:
+
+| year|historic_period      |
+|----:|:--------------------|
+| 2020|2015-2019            |
+| 2021|2015-2019            |
+| 2022|2016-2019, 2021      |
+| 2023|2017-2019, 2021-2022 |
+
+There is a second sense of 'historical' here, which denotes cause of death figures for 2020 to 2022; these were only provided as 5-year averages for 2015-2019, whereas we have a per-cause breakdown from 2020 onwards - to combine these, a 5-year historical average needs to be recomputed using a combination of the 2015-2019 data and the 2020+ data (using weighted averages).
 
@@
 
-Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the
+Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the **Note: I don't remember what came after this...**
 
@@
       select(-deaths_mean) %>%
-      filter(cause_of_death != "All")  # don't need all cause data in causes breadkown!
+      filter(cause_of_death != "All")  # don't need all cause data in causes breakdown!
   )
@@
 
-Note also that since 2022, the 5-year historical period has been defined as 2016-2021, with 2020 excluded, so we need to calculate two weekly averages!
+Since 2021, the 5-year historic context has been amended to exclude 2020 (see [above section on historical cause of death data](#historical-causes-of-death)).
+Because of this, the weekly average needs to be recalculated depending on which historic period is of interest.
+
+## Historic periods by year
+
+This is a reference table for the year of data, its historic period, and a nested list of vectors of years of historical data.
+
+```{r}
+historic_period_by_year <-
+  tibble(year = 2020:latest_year) %>%
+  mutate(
+    historic_period = historic_period_depending_on_year(year),
+    years_in_historic_period = map(.x = as.integer(str_sub(historic_period, 1, 4)), .f = generate_5_years_historic_period_excl_2020)
+    )
+```
+
+## Historic averages, overall
 
 ```{r, warning=FALSE}
+## helper functions to filter historic data for different 
+get_historic_and_current_data_from_year_range <- function(historic_data, current_data, year_range) {
+  year_range <- year_range[year_range!=2020]  # remove 2020 from range
+  historic_range <- year_range[year_range < 2020]
+  current_range <- year_range[year_range >= 2020]
+  bind_rows(
+    historic_data %>% 
+      filter(year %in% historic_range),
+    current_data %>% 
+      filter(year %in% current_range) %>% 
+      rename(number_of_deaths = deaths_all_causes)  # also rename to historic data format
+  )
+}
+
 historical_weekly_average_overall <-
-  bind_rows(
-    ## 2015-2019 period
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = 
           weekly_deaths_2015_2019_overall %>%
-    filter(year!="average") %>%  # remove averages
+          filter(year!="average") %>%
+          mutate(year = as.integer(year))
+          ,
+        current_data = 
+          weekly_deaths_overall %>%
+          filter(age == "all" & sex == "all"),
+        year_range = years_in_historic_period
+      ) %>%
       group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
@@
       rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
-    mutate(historic_period = historic_period_depending_on_year(2020))
-    ,
-    ## 2016-2021, excluding 2020
-    bind_rows(
-      ## 2016-2019
-      weekly_deaths_2015_2019_overall %>%
-        mutate(year = as.integer(year)) %>%
-        filter(year %in% 2016:2019),
-      ## 2021
-      weekly_deaths_overall %>% 
-        rename(number_of_deaths = deaths_all_causes) %>%
-        filter(year == 2021 & age == "all" & sex == "all")
-      ) %>%
-    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
-    filter(!all(number_of_deaths==0 & week_number==53)) %>%
-    ungroup %>%
-    create_week_53_data_for_years_with_52_weeks %>%
-    group_by(week_number,place_of_death) %>% 
-    summarise(
-      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
-      .groups = "drop"
-    ) %>% 
-    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
-    mutate(historic_period = historic_period_depending_on_year(2022))
-  )
+      mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct()
 
@@
 }
-
-make_2016_to_2019_and_2021_data <- function(historical_data, current_data) {
-  bind_rows(
-      historical_data %>% filter(year %in% 2016:2019),
-      current_data %>% filter(year == 2021) %>% rename(number_of_deaths = deaths_all_causes)
-    )
-}
-
+```
+
+## Historic averages, HB, LA, sex, age
+
+```{r}
 historical_weekly_average_hb <-
-  bind_rows(
-    ## 2015-2019
-    weekly_deaths_2015_2019_hb %>%
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_hb, 
+        current_data = weekly_deaths_hb,
+        year_range = years_in_historic_period
+      ) %>%
       create_week_53_data_for_years_with_52_weeks() %>%
@@
       deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2020))
-    ,
-    ## 2016-2019, 2021
-    make_2016_to_2019_and_2021_data(
-      historical_data = weekly_deaths_2015_2019_hb, 
-      current_data = weekly_deaths_hb
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+
+historical_weekly_average_la <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_la, 
+        current_data = weekly_deaths_la,
+        year_range = years_in_historic_period
       ) %>%
@@
       deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2022))
-  )
-
-historical_weekly_average_la <-
-  bind_rows(
-    ## 2015-2019
-    weekly_deaths_2015_2019_la %>%
-    create_week_53_data_for_years_with_52_weeks() %>%
-    group_by(
-      week_number, place_of_death, ref_area
-    ) %>%
-    deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2020))
-    ,
-    ## 2016-2019, 2021
-    make_2016_to_2019_and_2021_data(
-      historical_data = weekly_deaths_2015_2019_la, 
-      current_data = weekly_deaths_la
-      ) %>%
-    create_week_53_data_for_years_with_52_weeks() %>%
-    group_by(
-      week_number, place_of_death, ref_area
-    ) %>%
-    deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2022))
-  )
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
 
 historical_weekly_average_sex_age <-
-  bind_rows(
-    ## 2015-2019
-    weekly_deaths_2015_2019_sex_age %>%
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_sex_age, 
+        current_data = weekly_deaths_sex_age,
+        year_range = years_in_historic_period
+      ) %>%
       create_week_53_data_for_years_with_52_weeks() %>%
@@
       deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2020))
-    ,
-    ## 2016-2019, 2021
-    make_2016_to_2019_and_2021_data(
-      historical_data = weekly_deaths_2015_2019_sex_age, 
-      current_data = weekly_deaths_sex_age
-      ) %>%
-    create_week_53_data_for_years_with_52_weeks() %>%
-    group_by(
-      week_number, place_of_death, age, sex
-    ) %>%
-    deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2022))
-  )
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
 
 historical_weekly_average_age <-
-  bind_rows(
-    ## 2015-2019
-    weekly_deaths_2015_2019_age %>%
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_age, 
+        current_data = weekly_deaths_age,
+        year_range = years_in_historic_period
+      ) %>%
       create_week_53_data_for_years_with_52_weeks() %>%
@@
       deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2020))
-    ,
-    ## 2016-2019, 2021
-    make_2016_to_2019_and_2021_data(
-      historical_data = weekly_deaths_2015_2019_age, 
-      current_data = weekly_deaths_age
-      ) %>%
-    create_week_53_data_for_years_with_52_weeks() %>%
-    group_by(
-      week_number, place_of_death, age
-    ) %>%
-    deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2022))
-  )
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
 
 historical_weekly_average_sex <-
-  bind_rows(
-    ## 2015-2019
-    weekly_deaths_2015_2019_sex %>%
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_sex, 
+        current_data = weekly_deaths_sex,
+        year_range = years_in_historic_period
+      ) %>%
       create_week_53_data_for_years_with_52_weeks() %>%
@@
       deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2020))
-    ,
-    ## 2016-2019, 2021
-    make_2016_to_2019_and_2021_data(
-      historical_data = weekly_deaths_2015_2019_sex, 
-      current_data = weekly_deaths_sex
-      ) %>%
-    create_week_53_data_for_years_with_52_weeks() %>%
-    group_by(
-      week_number, place_of_death, sex
-    ) %>%
-    deaths_mean_min_max %>%
-    mutate(historic_period = historic_period_depending_on_year(2022))
-  )
-
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+```
+
+## Historic averages, annual average
+
+```{r}
 # TODO: include the split between historical periods here also

### 2023-02-13 functions.R

@@ -166,13 +166,28 @@ return_pretty_cause_of_death <- function(raw_cause_of_death) {
   causes_of_death_nrs$pretty[match(x = raw_cause_of_death, table = causes_of_death_nrs$raw)]
 }
 
+
 ## helper function to return the historic period for average number of death calculations
+# TODO: at some point this will need updating!
 historic_period_depending_on_year <- function(year) {
   case_when(
     year %in% 2020:2021 ~ "2015-2019",
     year == 2022 ~ "2016-2019, 2021",
+    year == 2023 ~ "2017-2019, 2021-2022",
+    year == 2024 ~ "2018-2019, 2021-2023",
     TRUE ~ NA_character_
   )
+}
+
+## helper function for generating sequences of years excluding 2020
+## this is for use with computign historic periods
+generate_5_years_historic_period_excl_2020 <- function(starting_year) {
+  seq_years <- seq.int(from = starting_year, by = 1, length.out = 5)
+  if (2020 %in% seq_years) {
+    seq_years <- c(seq_years, tail(seq_years, 1) + 1L)
+  }
+  seq_years <- seq_years[seq_years!=2020]
+  return(seq_years)
 }
 
 ## helper function that creates week 53 data by taking week 52 in years with no year 53 and relabels it as 53

### 2023-02-13 covid_nrs_place_of_death_data_2023.Rmd
moved section for historic cause of death data around for rclarity; finished processing historic period data for different years

@@
 ```{r}
-# TODO: deal with 2022 data here, now that it's 'historical'
+# TODO: deal with 2022 data here, now that it's 'historical'; this will involve loading the last released data for 2022 released in the weekly releases for 2023, I assume somewhere mid-2023, like it was for 2021 data in 2022 releases
 cell_ranges_for_causes_of_death_data_2022 <- list(
@@
 
-#### Historical causes of death
+## Historical weekly deaths
+
+The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data. These are downloaded from the ad-hoc section also.
+
+```{r, warning=FALSE}
+weekly_deaths_2015_2019_overall <- 
+  read_csv(files_historical_deaths$overall, skip = 2) %>%
+  rename(year = 1) %>%  # rename first column
+  mutate(
+    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
+  ) %>%
+  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
+  drop_na(`1`) %>%
+  pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
+  mutate(week_number=as.integer(week_number)) %>%
+  recode_place_of_death()
+
+weekly_deaths_2015_2019_hb <- 
+  readxl::read_excel(files_historical_deaths$hb, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  clean_names() %>%
+  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
+  rename(  # same format as current data
+    week_number = matches("week"),
+    place_of_death = matches("location"),
+    ref_area = health_board
+  ) %>%
+  mutate(
+    week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
+
+weekly_deaths_2015_2019_la <-
+  readxl::read_excel(files_historical_deaths$la, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
+  clean_names() %>%
+  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
+  rename(  # same format as current data
+    week_number = matches("week"),
+    place_of_death = matches("location"),
+    ref_area = council_area
+  ) %>%
+  mutate(
+    week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
+
+weekly_deaths_2015_2019_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
+  readxl::read_excel(files_historical_deaths$sex_age, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
+  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
+  mutate(
+    year = 2000L + as.integer(year),
+    week_number = as.integer(week_number)
+  ) %>%
+  tidyr::complete(
+    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
+    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
+    fill = list(number_of_deaths = 0)
+  ) %>% 
+  recode_place_of_death()
+
+## compute weekly data for sex and age separately
+## sum over sex or age, respectively
+weekly_deaths_2015_2019_sex <-
+  weekly_deaths_2015_2019_sex_age %>%
+  group_by_all %>%
+  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
+
+weekly_deaths_2015_2019_age <-
+  weekly_deaths_2015_2019_sex_age %>%
+  group_by_all %>%
+  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
+  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
+```
+
+## Map data
+
+I'm using `sf` to read these and to include them invisualisations.
+
+* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-december-2020-uk-buc/
+* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
+  - Note: this is a fairly high resolution, don't need this much detail!
+  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)
+
+```{r}
+shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
+shapefile_la <- st_read(dsn = "./map_data/la")
+```
+
+## COVID case counts by LA/HB
+
+```{r}
+cases_hb <- read_csv(
+  files_case_counts$hb, 
+  col_types = 
+    cols(
+    .default = col_double(),
+    `_id` = col_skip(),
+    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
+    Date = col_date(format = "%Y%m%d"),
+    HB = col_character(),
+    HBName = col_character(),
+    HospitalAdmissionsQF = col_character(),
+    ICUAdmissionsQF = col_character()
+    ),
+  ) %>% clean_names() %>%
+  rename(ref_area = hb_name) %>%
+  mutate(
+    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
+    ref_area = phsmethods::match_area(hb)  # convert to standard HB names
+    )
+
+cases_la <- read_csv(
+  file = files_case_counts$la, 
+  col_types = 
+    cols(
+    .default = col_double(),
+    `_id` = col_skip(),
+    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
+    Date = col_date(format = "%Y%m%d"),
+    CA = col_character(),
+    CAName = col_character()
+  )
+) %>% clean_names() %>%
+  rename(ref_area = ca_name) %>%
+  mutate(
+    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
+    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
+    ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
+    )
+```
+
+# Historic data for deaths
 
@@
 There is a second sense of 'historical' here, which denotes cause of death figures for 2020 to 2022; these were only provided as 5-year averages for 2015-2019, whereas we have a per-cause breakdown from 2020 onwards - to combine these, a 5-year historical average needs to be recomputed using a combination of the 2015-2019 data and the 2020+ data (using weighted averages).
+This is done in [the later section recomputing historical averages for causes of death](#historical-annual-average-by-cause).
+
+# Historical averages, minima & maxima
+
+Due to the counting of weeks, most years have 52 weeks with some years (2015 & 2020) having 53.
+Therefore we need to decide how to deal with week 53 in 2020 since only 2015 has a corresponding weekly count.
+The solution used here is to calculate a weekly rate for week 53 from week 53 in 2015, and week 52 in 2016-2019 - this represents the "last week of the year", which includes December 31st.
+
+Since 2021, the 5-year historic context has been amended to exclude 2020 (see [above section on historical cause of death data](#historical-causes-of-death)).
+Because of this, the weekly average needs to be recalculated depending on which historic period is of interest.
+
+## Historic periods by year
+
+This is a reference table for the year of data, its historic period, and a nested list of vectors of years of historical data.
+
+```{r}
+historic_period_by_year <-
+  tibble(year = 2020:latest_year) %>%
+  mutate(
+    historic_period = historic_period_depending_on_year(year),
+    years_in_historic_period = map(.x = as.integer(str_sub(historic_period, 1, 4)), .f = generate_5_years_historic_period_excl_2020)
+    )
+```
+
+## Historic averages, overall
+
+```{r, warning=FALSE}
+## helper functions to filter historic data for different 
+get_historic_and_current_data_from_year_range <- function(historic_data, current_data, year_range) {
+  year_range <- year_range[year_range!=2020]  # remove 2020 from range
+  historic_range <- year_range[year_range < 2020]
+  current_range <- year_range[year_range >= 2020]
+  bind_rows(
+    historic_data %>% 
+      filter(year %in% historic_range),
+    current_data %>% 
+      filter(year %in% current_range) %>% 
+      rename(number_of_deaths = deaths_all_causes)  # also rename to historic data format
+  )
+}
+
+historical_weekly_average_overall <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = 
+          weekly_deaths_2015_2019_overall %>%
+          filter(year!="average") %>%
+          mutate(year = as.integer(year))
+          ,
+        current_data = 
+          weekly_deaths_overall %>%
+          filter(age == "all" & sex == "all"),
+        year_range = years_in_historic_period
+      ) %>%
+      group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
+      filter(!all(number_of_deaths==0 & week_number==53)) %>%
+      ungroup %>%
+      create_week_53_data_for_years_with_52_weeks %>%
+      group_by(week_number,place_of_death) %>% 
+      summarise(
+        across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
+        .groups = "drop"
+      ) %>% 
+      rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
+      mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct()
+
+week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
+```
+
+## Historic averages, HB, LA, sex, age
+
+```{r}
+## helper function to summarise with mean, min, and max
+deaths_mean_min_max <- function(data_tbl) {
+  data_tbl %>%
+    summarise(
+      deaths_mean = mean(number_of_deaths),
+      deaths_min = min(number_of_deaths),
+      deaths_max = max(number_of_deaths),
+      .groups = "drop"
+    )
+}
+
+historical_weekly_average_hb <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_hb, 
+        current_data = weekly_deaths_hb,
+        year_range = years_in_historic_period
+      ) %>%
+      create_week_53_data_for_years_with_52_weeks() %>%
+      group_by(
+        week_number, place_of_death, ref_area
+      ) %>%
+      deaths_mean_min_max %>%
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+
+historical_weekly_average_la <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_la, 
+        current_data = weekly_deaths_la,
+        year_range = years_in_historic_period
+      ) %>%
+      create_week_53_data_for_years_with_52_weeks() %>%
+      group_by(
+        week_number, place_of_death, ref_area
+      ) %>%
+      deaths_mean_min_max %>%
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+
+historical_weekly_average_sex_age <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_sex_age, 
+        current_data = weekly_deaths_sex_age,
+        year_range = years_in_historic_period
+      ) %>%
+      create_week_53_data_for_years_with_52_weeks() %>%
+      group_by(
+        week_number, place_of_death, age, sex
+      ) %>%
+      deaths_mean_min_max %>%
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+
+historical_weekly_average_age <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_age, 
+        current_data = weekly_deaths_age,
+        year_range = years_in_historic_period
+      ) %>%
+      create_week_53_data_for_years_with_52_weeks() %>%
+      group_by(
+        week_number, place_of_death, age
+      ) %>%
+      deaths_mean_min_max %>%
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+
+historical_weekly_average_sex <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_sex, 
+        current_data = weekly_deaths_sex,
+        year_range = years_in_historic_period
+      ) %>%
+      create_week_53_data_for_years_with_52_weeks() %>%
+      group_by(
+        week_number, place_of_death, sex
+      ) %>%
+      deaths_mean_min_max %>%
+      mutate(historic_period = historic_period)
+    }
+  ) %>% 
+  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+```
+
+## Historic averages, annual average
+
+Here I'm computing the 5-year annual average for various groupings of variables, and across the various possible historic periods.
+Instead of taking the weekly averages for a historic period computed in the above section, I recompute this separately from the raw data for historic periods and current data.
+
+```{r}
+## helper function to compute annual average per historic period, per place of
+## death, and per various grouping variables as needed
+compute_historical_annual_average_deaths <- function(data_tbl, ...) {
+  grouping_vars <- enquos(...)
+  data_tbl %>%
+    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
+    summarise(  # calculate annual totals first...
+    number_of_deaths = sum(number_of_deaths), 
+    .groups = "keep"
+  ) %>%
+  ungroup(year) %>%  # remove year as grouping to average over year now
+  summarise(
+    annual_deaths_mean = mean(number_of_deaths),
+    annual_deaths_min = min(number_of_deaths),
+    annual_deaths_max = max(number_of_deaths),
+    .groups = "drop"
+  )
+}
+
+## Note, there was an error in these before, where I erroneously omitted 2015
+## figures from the annual averages computation for 2015-2019; this had minimal
+## effect on the mean/min/max values because the algorithm used averaged over
+## the available years (as opposed to e.g. dividing 4 years of data by 5)
+
+# the below commented out section is no longer used, but I've left it to
+# demonstrate the error in previous data - I think I started 'splitting' the
+# 2015-2019 afrom the 2016-2021 historic periods but didn't end up implementing
+# both, and instead left a confusing entry that compute the 2016-2019 period
+# only.
+# make_2016_to_2019_and_2021_data_with_historic_periods <- function(historical_data, current_data) {
+#   bind_rows(
+#     historical_data %>% 
+#       filter(year %in% 2016:2019) %>%
+#       mutate(historic_period = historic_period_depending_on_year(2020)),
+#     current_data %>% 
+#       filter(year == 2021) %>% 
+#       rename(number_of_deaths = deaths_all_causes) %>%
+#       mutate(historic_period = historic_period_depending_on_year(2022))
+#   )
+# }
+
+historical_annual_average_overall <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = 
+          weekly_deaths_2015_2019_overall %>%  # a quirk of the weekly deaths 'overall' data
+          filter(year != "average") %>%
+          mutate(year = as.integer(year)), 
+        current_data = 
+          weekly_deaths_overall %>%
+          filter(sex == "all" & age == "all"),
+        year_range = years_in_historic_period
+      ) %>%
+        mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+  compute_historical_annual_average_deaths()
+
+historical_annual_average_hb <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_hb, 
+        current_data = weekly_deaths_hb, 
+        year_range = years_in_historic_period
+      ) %>%
+        mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+  compute_historical_annual_average_deaths(ref_area)
+
+historical_annual_average_la <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_la, 
+        current_data = weekly_deaths_la, 
+        year_range = years_in_historic_period
+      ) %>%
+        mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+  compute_historical_annual_average_deaths(ref_area)
+
+historical_annual_average_sex_age <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_sex_age, 
+        current_data = weekly_deaths_sex_age, 
+        year_range = years_in_historic_period
+      ) %>%
+        mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+  compute_historical_annual_average_deaths(sex, age)
+
+historical_annual_average_sex <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_sex, 
+        current_data = weekly_deaths_sex, 
+        year_range = years_in_historic_period
+      ) %>%
+        mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+  compute_historical_annual_average_deaths(sex)
+
+historical_annual_average_age <-
+  pmap_dfr(
+    .l = historic_period_by_year,
+    .f = function(year, historic_period, years_in_historic_period) {
+      get_historic_and_current_data_from_year_range(
+        historic_data = weekly_deaths_2015_2019_age, 
+        current_data = weekly_deaths_age, 
+        year_range = years_in_historic_period
+      ) %>%
+        mutate(historic_period = historic_period)
+    }
+  ) %>%
+  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+  compute_historical_annual_average_deaths(age)
+```
+
+## Check weekly historic averages against NRS computed data
+
+```{r}
+weekly_deaths_overall %>%
+  filter(sex=="all" & age=="all") %>%
+  mutate(historic_period = historic_period_depending_on_year(year)) %>%
+  left_join(historical_weekly_average_overall) %>%
+  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean, historic_period) %>%
+  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
+  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e.
+all other entries had the same figures!).
+Week 53 was computed as the average of deaths in week 53 in 2015 and weeks 52 in 2016-2019.
+For all the other datasets the mean deaths computed agree!")
+```
+
+# Historic average by cause of death
 
@@
 
-Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the **Note: I don't remember what came after this...**
-
-```{r}
+```{r}
+# TODO: change this so it works for 2023 and onwards
+
 ## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
@@
 ## however, for the latter, we set the dates to 2022
-# average_weekly_covid_deaths_2021_onwards <-
 average_weekly_covid_deaths_2021_onwards <-
@@
 
-#### Merge 2020 & 2021 data with current data
+## Merge 2020 & 2021 data with current data
 
@@
 
-## Historical weekly deaths
-
-The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data. These are downloaded from the ad-hoc section also.
-
-```{r, warning=FALSE}
-weekly_deaths_2015_2019_overall <- 
-  read_csv(files_historical_deaths$overall, skip = 2) %>%
-  rename(year = 1) %>%  # rename first column
-  mutate(
-    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
-  ) %>%
-  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
-  drop_na(`1`) %>%
-  pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
-  mutate(week_number=as.integer(week_number)) %>%
-  recode_place_of_death()
-
-weekly_deaths_2015_2019_hb <- 
-  readxl::read_excel(files_historical_deaths$hb, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
-  clean_names() %>%
-  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
-  rename(  # same format as current data
-    week_number = matches("week"),
-    place_of_death = matches("location"),
-    ref_area = health_board
-  ) %>%
-  mutate(
-    week_number = as.integer(week_number)
-  ) %>%
-  tidyr::complete(
-    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
-    fill = list(number_of_deaths = 0)
-  ) %>% 
-  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
-
-weekly_deaths_2015_2019_la <-
-  readxl::read_excel(files_historical_deaths$la, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
-  clean_names() %>%
-  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
-  rename(  # same format as current data
-    week_number = matches("week"),
-    place_of_death = matches("location"),
-    ref_area = council_area
-  ) %>%
-  mutate(
-    week_number = as.integer(week_number)
-  ) %>%
-  tidyr::complete(
-    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
-    fill = list(number_of_deaths = 0)
-  ) %>% 
-  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also
-
-weekly_deaths_2015_2019_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(files_historical_deaths$sex_age, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
-  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
-  mutate(
-    year = 2000L + as.integer(year),
-    week_number = as.integer(week_number)
-  ) %>%
-  tidyr::complete(
-    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
-    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
-    fill = list(number_of_deaths = 0)
-  ) %>% 
-  recode_place_of_death()
-
-## compute weekly data for sex and age separately
-## sum over sex or age, respectively
-weekly_deaths_2015_2019_sex <-
-  weekly_deaths_2015_2019_sex_age %>%
-  group_by_all %>%
-  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
-
-weekly_deaths_2015_2019_age <-
-  weekly_deaths_2015_2019_sex_age %>%
-  group_by_all %>%
-  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
-  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
-```
-
-## Map data
-
-I'm using `sf` to read these and to include them invisualisations.
-
-* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-december-2020-uk-buc/
-* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
-  - Note: this is a fairly high resolution, don't need this much detail!
-  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)
-
-```{r}
-shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
-shapefile_la <- st_read(dsn = "./map_data/la")
-```
-
-## COVID case counts by LA/HB
-
-```{r}
-cases_hb <- read_csv(
-  files_case_counts$hb, 
-  col_types = 
-    cols(
-    .default = col_double(),
-    `_id` = col_skip(),
-    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
-    Date = col_date(format = "%Y%m%d"),
-    HB = col_character(),
-    HBName = col_character(),
-    HospitalAdmissionsQF = col_character(),
-    ICUAdmissionsQF = col_character()
-    ),
-  ) %>% clean_names() %>%
-  rename(ref_area = hb_name) %>%
-  mutate(
-    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
-    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
-    ref_area = phsmethods::match_area(hb)  # convert to standard HB names
-    )
-
-cases_la <- read_csv(
-  file = files_case_counts$la, 
-  col_types = 
-    cols(
-    .default = col_double(),
-    `_id` = col_skip(),
-    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
-    Date = col_date(format = "%Y%m%d"),
-    CA = col_character(),
-    CAName = col_character()
-  )
-) %>% clean_names() %>%
-  rename(ref_area = ca_name) %>%
-  mutate(
-    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
-    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
-    ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
-    )
-```
-
-# Historical averages, minima & maxima
-
-Due to the counting of weeks, most years have 52 weeks with some years (2015 & 2020) having 53.
-Therefore we need to decide how to deal with week 53 in 2020 since only 2015 has a corresponding weekly count.
-The solution used here is to calculate a weekly rate for week 53 from week 53 in 2015, and week 52 in 2016-2019 - this represents the "last week of the year", which includes December 31st.
-
-Since 2021, the 5-year historic context has been amended to exclude 2020 (see [above section on historical cause of death data](#historical-causes-of-death)).
-Because of this, the weekly average needs to be recalculated depending on which historic period is of interest.
-
-## Historic periods by year
-
-This is a reference table for the year of data, its historic period, and a nested list of vectors of years of historical data.
-
-```{r}
-historic_period_by_year <-
-  tibble(year = 2020:latest_year) %>%
-  mutate(
-    historic_period = historic_period_depending_on_year(year),
-    years_in_historic_period = map(.x = as.integer(str_sub(historic_period, 1, 4)), .f = generate_5_years_historic_period_excl_2020)
-    )
-```
-
-## Historic averages, overall
-
-```{r, warning=FALSE}
-## helper functions to filter historic data for different 
-get_historic_and_current_data_from_year_range <- function(historic_data, current_data, year_range) {
-  year_range <- year_range[year_range!=2020]  # remove 2020 from range
-  historic_range <- year_range[year_range < 2020]
-  current_range <- year_range[year_range >= 2020]
-  bind_rows(
-    historic_data %>% 
-      filter(year %in% historic_range),
-    current_data %>% 
-      filter(year %in% current_range) %>% 
-      rename(number_of_deaths = deaths_all_causes)  # also rename to historic data format
-  )
-}
-
-historical_weekly_average_overall <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = 
-          weekly_deaths_2015_2019_overall %>%
-          filter(year!="average") %>%
-          mutate(year = as.integer(year))
-          ,
-        current_data = 
-          weekly_deaths_overall %>%
-          filter(age == "all" & sex == "all"),
-        year_range = years_in_historic_period
-      ) %>%
-      group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
-      filter(!all(number_of_deaths==0 & week_number==53)) %>%
-      ungroup %>%
-      create_week_53_data_for_years_with_52_weeks %>%
-      group_by(week_number,place_of_death) %>% 
-      summarise(
-        across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
-        .groups = "drop"
-      ) %>% 
-      rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>%
-  distinct()
-
-week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
-
-# TODO: weekly averages may be too detailed for this level od subcategorising!
-
-deaths_mean_min_max <- function(data_tbl) {
-  data_tbl %>%
-    summarise(
-      deaths_mean = mean(number_of_deaths),
-      deaths_min = min(number_of_deaths),
-      deaths_max = max(number_of_deaths),
-      .groups = "drop"
-    )
-}
-```
-
-## Historic averages, HB, LA, sex, age
-
-```{r}
-historical_weekly_average_hb <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_hb, 
-        current_data = weekly_deaths_hb,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, ref_area
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_la <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_la, 
-        current_data = weekly_deaths_la,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, ref_area
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_sex_age <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_sex_age, 
-        current_data = weekly_deaths_sex_age,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, age, sex
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_age <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_age, 
-        current_data = weekly_deaths_age,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, age
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_sex <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_sex, 
-        current_data = weekly_deaths_sex,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, sex
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-```
-
-## Historic averages, annual average
-
-```{r}
-# TODO: include the split between historical periods here also
-compute_historical_annual_average_deaths <- function(data_tbl, ...) {
-  grouping_vars <- enquos(...)
-  data_tbl %>%
-    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
-    summarise(  # calculate annual totals first...
-    number_of_deaths = sum(number_of_deaths), 
-    .groups = "keep"
-  ) %>%
-  ungroup(year) %>%  # remove year as grouping to average over year now
-  summarise(
-    annual_deaths_mean = mean(number_of_deaths),
-    annual_deaths_min = min(number_of_deaths),
-    annual_deaths_max = max(number_of_deaths),
-    .groups = "drop"
-  )
-}
-
-historical_annual_average_overall <-
-  bind_rows(
-      ## 2016-2019
-      weekly_deaths_2015_2019_overall %>%
-        mutate(year = as.integer(year)) %>%
-        filter(year %in% 2016:2019) %>%
-        mutate(historic_period = historic_period_depending_on_year(2020)),
-      ## 2021
-      weekly_deaths_overall %>% 
-        rename(number_of_deaths = deaths_all_causes) %>%
-        filter(year == 2021 & age == "all" & sex == "all") %>%
-        mutate(historic_period = historic_period_depending_on_year(2022))
-      ) %>%
-  compute_historical_annual_average_deaths()
-
-make_2016_to_2019_and_2021_data_with_historic_periods <- function (historical_data, current_data) {
-  bind_rows(
-    historical_data %>% 
-      filter(year %in% 2016:2019) %>%
-      mutate(historic_period = historic_period_depending_on_year(2020)),
-    current_data %>% 
-      filter(year == 2021) %>% 
-      rename(number_of_deaths = deaths_all_causes) %>%
-      mutate(historic_period = historic_period_depending_on_year(2022))
-  )
-}
-
-historical_annual_average_hb <-
-  make_2016_to_2019_and_2021_data_with_historic_periods(
-    historical_data = weekly_deaths_2015_2019_hb, 
-    current_data = weekly_deaths_hb
-      ) %>%
-  compute_historical_annual_average_deaths(ref_area)
-
-historical_annual_average_la <-
-  make_2016_to_2019_and_2021_data_with_historic_periods(
-    historical_data = weekly_deaths_2015_2019_la, 
-    current_data = weekly_deaths_la
-      ) %>%
-  compute_historical_annual_average_deaths(ref_area)
-
-historical_annual_average_sex_age <-
-  make_2016_to_2019_and_2021_data_with_historic_periods(
-    historical_data = weekly_deaths_2015_2019_sex_age, 
-    current_data = weekly_deaths_sex_age
-      ) %>%
-  compute_historical_annual_average_deaths(sex, age)
-
-historical_annual_average_sex <-
-  make_2016_to_2019_and_2021_data_with_historic_periods(
-    historical_data = weekly_deaths_2015_2019_sex, 
-    current_data = weekly_deaths_sex
-      ) %>%
-  compute_historical_annual_average_deaths(sex)
-
-historical_annual_average_age <-
-  make_2016_to_2019_and_2021_data_with_historic_periods(
-    historical_data = weekly_deaths_2015_2019_age, 
-    current_data = weekly_deaths_age
-      ) %>%
-  compute_historical_annual_average_deaths(age)
-```
-
 ## Historical annual average by cause
+
+Cause of death data for 2015-2019 was only given in terms of weekly averages over 5 years.
+With the later historic periods (2016-2019 + 2021 and so on) I compute a weighted average between the 5-year average and the current weekly data.
+
+The below compromise figure is no longer used, but the reasoning for the weighted average is the same (except for the quirk of the denominator of 6 in the below case, where I'm including the 5 years of historic data plus the one year of current data, hence 6 years in the denominator):
 
@@
   )
-```
-
-## Check weekly historic averages against NRS computed data
-
-```{r}
-weekly_deaths_overall %>%
-  filter(sex=="all" & age=="all") %>%
-  mutate(historic_period = historic_period_depending_on_year(year)) %>%
-  left_join(historical_weekly_average_overall) %>%
-  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean, historic_period) %>%
-  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
-  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e.
-all other entries had the same figures!).
-Week 53 was computed as the average of deaths in week 53 in 2015 and weeks 52 in 2016-2019.
-For all the other datasets the mean deaths computed agree!")
 ```

### 2023-02-13 covid_nrs_place_of_death_visualisations_2023.Rmd
modified place of death graphs to use 2023 data and albel start of 2023

--- 2022-10-27 covid_nrs_place_of_death_visualisations_2022.Rmd
@@
 ---
-title: "Place of death in NRS COVID-19 deaths: visualisation for 2022 data"
+title: "Place of death in NRS COVID-19 deaths: visualisation for 2023 data"
 author: "Jan Savinc"
@@
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
@@
   geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
   annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
   NULL
@@
 
-## Home deaths only
-
-```{r}
-fig_place_of_death_home_only <-
-  merged_deaths_overall %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
-  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non Covid-19 deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    plot.caption = element_text(size = 10, colour = "gray60"),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = NULL,
-       y = "Deaths (N)",
-       subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text,historic_period_2022_text,source_nrs_text), collapse="\n")
-       ) +
-  ## annotate historic range
-  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
-  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
-  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
-  NULL
-
-fig_place_of_death_home_only %>%
-save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
-```
-
-### Home deaths with 2015-2019 historic period means & range only
+## With 2015-2019 historic range only
 
@@
 
+fig_place_of_death_2015_2019_range <-
+  merged_deaths_overall_2015_2019_historic_period %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  geom_text(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
+  geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
+  NULL
+
+fig_place_of_death_2015_2019_range %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
+```
+
+## Home deaths only
+
+```{r}
+fig_place_of_death_home_only <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,historic_period_2022_text,source_nrs_text), collapse="\n")
+       ) +
+  ## annotate historic range
+  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
+  NULL
+
+fig_place_of_death_home_only %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+## Home deaths with 2015-2019 historic period means & range only
+
+```{r}
 fig_place_of_death_home_2015_2019_historic_period <-
@@
   annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "2022", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
   NULL

### 2023-02-14 covid_nrs_place_of_death_data_2023.Rmd
cleaned up the processing of cause of death data; found discrepancy between 5-year weekly averages for 2020 and 2021 CoD data, which supposedly should be the same?

@@
   rename(year = registration_year, date_w_c = week_beginning) %>%
-  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
+  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
+  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
   pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
@@
     year = as.integer(year),
+    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
     cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
     cause_of_death = return_pretty_cause_of_death(cause_of_death),
-    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
+    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
     week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
@@
   select(-name) %>%
-  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
+  pivot_wider(names_from = measure_type, values_from = number_of_deaths)
+  # filter(!is.na(number_of_deaths))   # this will also remove the historic mean values, so I've opted not to remove blank entries!
 ```
@@
 The weekly causes of death data for 2021 is included in the week 25 'main report' file, table `8`.
+
+~~Note that due to how I originally processed these files, the weekly CoD data for 2020 *includes* the 5-year historic weekly average for 2015-2019, whereas I calculate that manually later for the 2021 data and later.~~ This is now changed so that the weekly CoD data is now
 
@@
     year = as.integer(year),
+    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
     cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
@@
   select(-name) %>%
+  pivot_wider(names_from = measure_type, values_from = number_of_deaths) %>%
   filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
@@
 
+Cause of death data comes on a per-year basis, with the current year's data including the preceding year; at some point mid-year, this changes, and only the current year's data are included, and the preceding year's data need to be loaded from the last release that included them.
+(e.g.
+mid-2022, causes of death for 2021 stopped being released alongside 2022 data, so I load them from the last 2022 release that included 2021 causes of death also).
+They therefore come as one dataset per year, with the current year as a separate dataset.
+The weekly figures for cause of death are included alongside the 5-year historic average for that week except for Covid, which has been set to NA for all data up and including 2023 data as of February 2023.
+
 In the `2015-2019` period there were no Covid-19 deaths, so the historical average should be `NA`, which can possibly be interpreted as 0 for the purposes of comparing a year to the historical average.
+
 For the `2016-2019, 2021` period the average number of covid deaths should be equivalent to the deaths in 2021 - we wouldn't want to make an average between 4 years of 0 deaths and one year with any deaths!
@@
 ```{r}
-# TODO: change this so it works for 2023 and onwards
-
-## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
-## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
-## however, for the latter, we set the dates to 2022
-average_weekly_covid_deaths_2021_onwards <-
+# TODO: include 2023 data once it's reported separately
+
+# TODO: I still need to separately process the most recent year because the average values aren't provided
+
+## here I separately merge the datasets for cause of death data from different years
+## I treat covid deaths separately to recalculate a realistic esstimate of average deaths in prior years
+weekly_deaths_cause_of_death_noncovid <-
   bind_rows(
-    ## set 2021 averages to NA
-    weekly_deaths_cause_of_death_2021 %>%
+    weekly_deaths_cause_of_death_2020,
+    weekly_deaths_cause_of_death_2021,
+    weekly_deaths_cause_of_death_2022_onwards
+  ) %>%
+  filter(cause_of_death != "COVID-19")
+
+weekly_deaths_cause_of_death_covid <-
+  bind_rows(
+    weekly_deaths_cause_of_death_2020,
+    weekly_deaths_cause_of_death_2021,
+    weekly_deaths_cause_of_death_2022_onwards
+  ) %>%
   filter(cause_of_death == "COVID-19") %>%
-    mutate(number_of_deaths = NA_real_) %>%
-    rename(deaths_mean = number_of_deaths),
-    ## use 2021 weekly deaths as 2022 historic average
-    weekly_deaths_cause_of_death_2021 %>%
-    filter(cause_of_death == "COVID-19") %>%
-    mutate(year = 2022) %>%
-    select(-c(week_number_run_over, date_w_c)) %>%
-    rename(deaths_mean = number_of_deaths) %>%
-    inner_join(weekly_deaths_cause_of_death_2022_onwards %>% 
-                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
-              )
-  ) %>%
-  mutate(historic_period = historic_period_depending_on_year(year))
-
-average_weekly_noncovid_deaths_2021_onwards <-
-    ## extract 5-year averages from most recent file for 2021 onwards
-    ## this doesn't include covid, which we've manually extracted above!
-  bind_rows(
-    ## 2022+ data
-    map_dfr(
-      .x = cell_ranges_for_causes_of_death_data,
-      .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
-      .id = "place_of_death"
-    ),
-    ## 2021 data
-    map_dfr(
-      .x = cell_ranges_for_causes_of_death_data_2021,
-      .f = ~readxl::read_excel(files_historical_deaths$`2021`, sheet = sheet_cause_of_death_data, range = .x), 
-      .id = "place_of_death"
-    )
-  ) %>%
-    janitor::clean_names() %>%
-    rename(year = registration_year, date_w_c = week_beginning) %>%
-    select(place_of_death, year, week_number, date_w_c, matches("five_year_average")) %>%
-    pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "deaths_mean") %>%
-    mutate(
-      year = as.integer(year),
-      cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
-      cause_of_death = return_pretty_cause_of_death(cause_of_death),
-      cause_of_death = if_else(cause_of_death == "All causes", "All", cause_of_death),
-      date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
-      week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
-      historic_period = historic_period_depending_on_year(year)
-    ) %>%
-    select(-name) %>%
-    drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future
-
-historical_average_cause_of_death_2021_onwards <-
-  bind_rows(
-    ## the hacked averages for 2021 and 2022 above
-    average_weekly_covid_deaths_2021_onwards,
-    ## the non-covid deaths in the 'main report' files
-    average_weekly_noncovid_deaths_2021_onwards
-  )
-```
-
-## Merge 2020 & 2021 data with current data
-
-Note: the 5-year average for 2022 (spanning 2016-2019 & 2021) is only published along with the running weekly counts of deaths by cause of death; it's just that NRS has decided to only publish them at the same time as new data comes out!
-This is a bit of a bummer - it means we can't populate the whole year of 'historical context' data for 2022.
-
-They could be calculated by hand if the 2015-2019 weekly data were available broken down by cause of death - as it stands, only the average 2015-2019 figure is available.
-
-```{r}
+  group_by(year) %>%
+  group_map(
+    .f = function(data_tbl, year_tbl) {
+      current_year <- year_tbl$year  # get year from the function arguments
+      year_range <- infer_year_range_for_calculating_covid_deaths(current_year)  # work out which years of data to use
+      weekly_means <-  
+        weekly_deaths_cause_of_death_covid %>%
+        filter(year %in% year_range) %>%  # filter the covid data to the relevant years to compute from
+        create_week_53_data_for_years_with_52_weeks() %>%  # create 53rd week if needed
+        group_by(place_of_death, cause_of_death, week_number) %>%
+        summarise(
+          deaths_mean = mean(number_of_deaths, na.rm=TRUE),
+          # year_range = paste0(year_range,collapse=","),  # for testing
+          .groups = "drop"
+        )
+      data_tbl %>%
+        mutate(year = current_year) %>%  # this gets removed in the grouping, so I re-add it here
+        select(-deaths_mean) %>%  # remove the NA/0 mean deaths
+        create_week_53_data_for_years_with_52_weeks() %>%  # create 53rd week if needed
+        left_join(weekly_means, by = c("place_of_death", "cause_of_death", "week_number"))  # merge the mean deaths
+    }
+  ) %>%
+  bind_rows()  # merge into single tibble from list
+
+## re-merge non-covid and covid deaths again
+# TODO: make CoD, PoD factors
 weekly_deaths_cause_of_death <-
   bind_rows(
-    weekly_deaths_cause_of_death_2022_onwards,
-    weekly_deaths_cause_of_death_2021,
-    weekly_deaths_cause_of_death_2020 %>% 
-      select(-deaths_mean) %>%
-      filter(cause_of_death != "All")  # don't need all cause data in causes breakdown!
-  )
-
-historical_average_cause_of_death <-
-  bind_rows(
-    historical_average_cause_of_death_2021_onwards,
-    weekly_deaths_cause_of_death_2020 %>% 
-      select(-number_of_deaths) %>% 
+    weekly_deaths_cause_of_death_noncovid,
+    weekly_deaths_cause_of_death_covid
+  ) %>%
   mutate(historic_period = historic_period_depending_on_year(year))
-  )
-```
-
-## Historical annual average by cause
-
-Cause of death data for 2015-2019 was only given in terms of weekly averages over 5 years.
-With the later historic periods (2016-2019 + 2021 and so on) I compute a weighted average between the 5-year average and the current weekly data.
-
-The below compromise figure is no longer used, but the reasoning for the weighted average is the same (except for the quirk of the denominator of 6 in the below case, where I'm including the 5 years of historic data plus the one year of current data, hence 6 years in the denominator):
-
-I can approximate this by taking the sum of the average weekly deaths for `2015-2019`; for the `2016-2019 + 2021` period this isn't readily available because NRS only publish the weekly averages alongside the weekly updated data, but not for the entire year.
-Therefore I'll just calculate the `2015-2019` period average for now, and a compromise `2015-2019 + 2021` average:
-
-$$\frac{5}{6}average\_deaths_{2015-2019} + \frac{1}{6}deaths_{2021}$$
-
-```{r}
-# TODO: incorporate the 2016-2019 + 2021 period
-historical_annual_average_cause_of_death_2015_2019 <-
-  historical_average_cause_of_death %>%
-  filter(historic_period == "2015-2019" & year == 2021) %>%  # this is slightly strange but some of the averages disagree, probably because the 2015-2019 data were updated in 2020, and the 2021 data were published with modified figures.
-NRS deaths data are truly a living document!
-  group_by(place_of_death, cause_of_death, historic_period) %>%
+
+## note that these might differ from actual historic figures a bit because of
+## my hack to create a fake 53rd week in years with only 52 weeks for easier 
+## comparison
+
+# TODO: I get different numbers of 5-year annual average deaths between 2020 and 2021, how come?
+historical_annual_average_cause_of_death <-
+  weekly_deaths_cause_of_death %>%
+  group_by(year, historic_period, place_of_death, cause_of_death) %>%
   summarise(
@@
     .groups = "drop"
-  )
-
-deaths_by_cause_2021 <-
-  weekly_deaths_cause_of_death_2021 %>% 
-  group_by(place_of_death, cause_of_death) %>% 
-  summarise(deaths_2021 = sum(number_of_deaths), .groups = "drop")
-
-historical_annual_average_cause_of_death <-
-  bind_rows(
-    historical_annual_average_cause_of_death_2015_2019
-    ,
-    ## compromise figures for 2015-2019 + 2021
-    historical_annual_average_cause_of_death_2015_2019 %>%
-    filter(cause_of_death != "All") %>%
-    left_join(
-      deaths_by_cause_2021, by = c("place_of_death", "cause_of_death")
-    ) %>%
-    mutate(
-      annual_deaths_mean = (5/6 * annual_deaths_mean) + (1/6 * deaths_2021),
-      deaths_2021 = NULL,
-      historic_period = "2015-2019, 2021"
-    )
-  )
+  ) %>% 
+  filter(year!=2020)  # 2020 and 2021 should have identical annual means so I remove one %>%
+  select(-year)
+
+## checking discrepancy:
+## I get different numbers of 5-year annual average deaths between 2020 and 2021, how come?
+weekly_deaths_cause_of_death %>% filter(year %in% 2020:2021) %>% select(place_of_death, cause_of_death, year, week_number, deaths_mean) %>%
+  pivot_wider(names_from = year, values_from = deaths_mean) %>%
+  filter(`2020`!=`2021`) %>%
+  arrange(desc(week_number)) %>%
+  summarise(across(matches("\\d{4}"), ~sum(.x, na.rm=TRUE)))
+## looks like 2020 and 2021 don't actually use the same 5-year weekly average deaths by cause!!!
+# TODO: act on this knowledge...
 ```

### 2023-02-14 functions.R

@@ -293,3 +293,13 @@ parse_ref_period = function(data_tbl) {
   }
     
 }
+
+## helper function to infer how many years of Covid data to average from for
+## historic figures, given current year covid data started in 2020: 2020 has no
+## historic figures, in 2021 we take the weekly average from 2020, in 2022 we
+## take from 2020-2021, and so on
+infer_year_range_for_calculating_covid_deaths <- function(year) {
+  year_range <- seq.int(from = year-5L, to = year-1L)
+  year_range <- year_range[year_range>=2020]
+  return(year_range)
+}

### 2023-02-21 covid_nrs_place_of_death_data_2023.Rmd
modified data to avoid hard-coding current year - lots more to change I'm sure!

@@
 ## a separate raw file makes it easier to debug, and also reduces the number of requests I make while debugging!
+## Note: as of 16 Feb 2023, this returns a strange error
+# TODO: if error persists, load data from the main report file instead?
 ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")
@@
 
-weekly_deaths_cause_of_death_covid <-
+weekly_deaths_cause_of_death_covid_before_processing <-
   bind_rows(
@@
     weekly_deaths_cause_of_death_2022_onwards
-  ) %>%
+  )
+
+weekly_deaths_cause_of_death_covid <-
+  weekly_deaths_cause_of_death_covid_before_processing %>%
   filter(cause_of_death == "COVID-19") %>%
@@
       weekly_means <-  
-        weekly_deaths_cause_of_death_covid %>%
+        weekly_deaths_cause_of_death_covid_before_processing %>%  # bind again weekly deaths
         filter(year %in% year_range) %>%  # filter the covid data to the relevant years to compute from
@@
   ) %>% 
-  filter(year!=2020)  # 2020 and 2021 should have identical annual means so I remove one %>%
+  filter(year!=2020) %>%  # 2020 and 2021 should have identical annual means so I remove one %>%
   select(-year)
@@
 
+## Note: previously I computed the historical average separately fro mthe weekly cause of death data
+## Now it's already computed in the same dataset, so I just assign it here, no joining needed
 merged_deaths_cause_of_death <-
-  weekly_deaths_cause_of_death %>%
-  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
-  left_join(  # left_join so that the historical Covid-19 rates are blank!
-    historical_average_cause_of_death %>% select(-week_number_run_over), 
-    by = c("place_of_death", "week_number", "cause_of_death", "historic_period", "year", "date_w_c")
-  )
+  weekly_deaths_cause_of_death
+# merged_deaths_cause_of_death <-
+#   weekly_deaths_cause_of_death %>%
+#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+#   left_join(  # left_join so that the historical Covid-19 rates are blank!
+#     historical_average_cause_of_death %>% select(-week_number_run_over), 
+#     by = c("place_of_death", "week_number", "cause_of_death", "historic_period", "year", "date_w_c")
+#   )
 ```
@@
 
-These are used for saying in plots what data were included e.g. weeks 1-12 in 2022.
+These are used for saying in plots what data were included e.g. weeks 1-12 in 2023.
 
@@
 ## constants for use in graphs & annotations
-weeks_available_2022 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2022])
-most_recent_date_available_2022 <- max(merged_deaths_age$date_w_c)
+weeks_available_2023 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2023])
+most_recent_date_available_2023 <- max(merged_deaths_age$date_w_c)
 
 earliest_date <- min(merged_deaths_age$date_w_c)
+current_year <- isoyear(most_recent_date_available_2023)
 ```
@@
       summarise(
-        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2022}, 2022"),
+        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2023}, 2022"),
         historic_period = "2015-2019",
@@
     summarise(
-      period = glue::glue("2022 up to week {weeks_available_2022}"),
+      period = glue::glue("2022 up to week {weeks_available_2023}"),
       historic_period = "2015-2019",

### 2023-02-21 covid_nrs_place_of_death_visualisations_2023.Rmd
modified data to avoid hard-coding current year - lots more to change I'm sure!

--- 2022-10-27 covid_nrs_place_of_death_visualisations_2022.Rmd
@@
 
-date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2022, '%d %B %Y')} (inclusive).")
+date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023, '%d %B %Y')} (inclusive).")
 
@@
 only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
-only_2022_text <- glue::glue("Data for 2022 only include weeks 1-{format(weeks_available_2022)} (w/c {format(most_recent_date_available_2022)}).")
+only_2023_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023)} (w/c {format(most_recent_date_available_2023)}).")
 
@@
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
@@
 
+fig_place_of_death_2015_2019_range <-
+  merged_deaths_overall_2015_2019_historic_period %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n"),
+       colour = "Cause of death"
+       ) +
+  ## annotate historic range
+  geom_text(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
+  geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
+  NULL
+
+fig_place_of_death_2015_2019_range %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
+```
+
+## Home deaths only
+
+```{r}
+fig_place_of_death_home_only <-
+  merged_deaths_overall %>%
+  filter(sex == "all" & age == "all") %>%
+  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
+  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
+  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
+               names_to = "deaths_type", values_to = "deaths") %>% 
+  ggplot(aes(x = date_w_c)) +
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
+  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
+  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
+  geom_line(aes(y = deaths, colour = deaths_type)) +
+  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
+  facet_grid(~place_of_death) +
+  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
+                      labels = c("All deaths", 
+                                 "Non Covid-19 deaths"),
+                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
+                      ) +
+  theme(
+    legend.position = "top",
+    legend.title = element_blank(),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    plot.margin = margin(5, 15, 5, 5),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
+  ) +
+  labs(x = NULL,
+       y = "Deaths (N)",
+       subtitle = date_range_text,
+       caption = paste0(c(death_registrations_text,historic_period_2022_text,source_nrs_text), collapse="\n")
+       ) +
+  ## annotate historic range
+  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
+  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
+  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
+  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
+  NULL
+
+fig_place_of_death_home_only %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
+```
+
+## Home deaths with 2015-2019 historic period means & range only
+
+```{r}
 fig_place_of_death_home_2015_2019_historic_period <-
@@
     colour = NULL,
-    subtitle = paste0("* ",only_2022_text),
+    subtitle = paste0("* ",only_2023_text),
     caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
@@
     x = "Week number", y = "Cumulative proportion of place of death (%)",
-    subtitle = paste0("* ",only_2022_text),
+    subtitle = paste0("* ",only_2023_text),
     caption = paste0(c("Shaded areas represent the range of cumulative weekly proportion of deaths by place for the 2015-2019 period",death_registrations_text, source_nrs_text),collapse="\n")
@@
     title = str_wrap("Total deaths since start of 2020 by cause and place of death, compared to 2015-2019", width = 60),
-    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2022_text)), collapse="\n"),
+    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2023_text)), collapse="\n"),
     caption = paste0(c(death_registrations_text, source_nrs_text),collapse="\n")
@@
     title = "Annual deaths by cause and place of death, compared to 2015-2019",
-    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2022_text)), collapse="\n"),
+    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2023_text)), collapse="\n"),
     caption = paste0(c(death_registrations_text, source_nrs_text),collapse="\n")
@@
     title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
-    subtitle = paste0("\n* ",only_2022_text),
+    subtitle = paste0("\n* ",only_2023_text),
     caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",death_registrations_text,source_nrs_text), collapse = "\n")
@@
   scale_colour_viridis_d() +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
   scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
@@
     # title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
-    subtitle = paste0("\n* ",only_2022_text),
+    subtitle = paste0("\n* ",only_2023_text),
     caption = paste0(c("The 2015-2019 figures are average weekly deaths.",death_registrations_text,source_nrs_text), collapse = "\n")
@@
     y = "Percentage weekly deaths by place (%)",
-    subtitle = paste0("\n* ",only_2022_text),
+    subtitle = paste0("\n* ",only_2023_text),
     caption = paste0(c("The 2015-2019 figures are average weekly deaths",death_registrations_text,source_nrs_text), collapse = "\n")
@@
   ggplot(aes(x = date_w_c)) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
   # geom_line(aes(y = number_of_deaths, colour = deaths_type, linetype = deaths_type)) +
@@
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "#4477aa"), alpha = 0.5) +
@@
   ggplot(aes(x = date_w_c)) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
+  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
   geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max, fill = "#4477aa"), alpha = 0.5) +
@@
     x = NULL, y= "Proportion of yearly deaths (%)",
-    subtitle = paste0("* ",only_2022_text),
+    subtitle = paste0("* ",only_2023_text),
     caption = paste0(c("Health boards sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
@@
     x = NULL, y= "Proportion of yearly deaths (%)",
-    subtitle = paste0("* ",only_2022_text),
+    subtitle = paste0("* ",only_2023_text),
     caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
@@
     x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
-    subtitle = paste0("* ",only_2022_text),
+    subtitle = paste0("* ",only_2023_text),
     caption = paste0(c("Health boards sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
@@
     x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
-    subtitle = paste0("* ",only_2022_text),
+    subtitle = paste0("* ",only_2023_text),
     caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
@@
        # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2023_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
        title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
-       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2023_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
        subtitle = "",
-       caption = paste0(c(only_2022_text,crude_rate_text,source_phs_text), collapse="\n"),
+       caption = paste0(c(only_2023_text,crude_rate_text,source_phs_text), collapse="\n"),
        fill = str_wrap("Cases per 100,000 population", width = 15)
@@
        title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board",
-       caption = paste0(c(only_2022_text,source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2023_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
   summarise(
-    deaths_mean_so_far = sum(deaths_mean[week_number <= weeks_available_2022], na.rm= TRUE),
+    deaths_mean_so_far = sum(deaths_mean[week_number <= weeks_available_2023], na.rm= TRUE),
     deaths_mean_total = sum(deaths_mean, na.rm= TRUE),
     proportion_of_deaths_so_far = deaths_mean_so_far / deaths_mean_total
-    )
-
-predicted_data_annotations <- tribble(
-  ~x, ~y, ~label,
-  2021, 20e3, "Predicted home deaths 2020-2022 using 2015-2019 data",
-  2033, 23e3, "Actual home deaths 2020-2022*",
-  2033, 17.5e3, "If the 2015-2019 trend continued we wouldn't observe this number of home deaths until 2031",
     )
@@
     filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
-      mutate(data_type = "Actual")
+      mutate(data_type = "Historic")
     ,
@@
     select(year, number_of_deaths, place_of_death) %>%
-    filter(place_of_death == "Home & other non-institution" & year %in% 2020:2022) %>%
+    filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
     mutate(
       data_type = "Predicted",
-      number_of_deaths = if_else(year %in% 2020:2022, compute_y_from_x_using_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
+      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
       )
@@
       select(year, number_of_deaths, place_of_death) %>%
-      filter(place_of_death == "Home & other non-institution" & year %in% 2020:2022) %>%
+      filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
       mutate(
-        number_of_deaths = if_else(year == 2022, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
+        number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
         year = 2015 + compute_x_from_y_using_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
-        data_type = "Actual"  # note: this is actual figures but
-      )
+        data_type = "Actual",  # note: this is actual figures, but at the extrapolated year
+        actual_year = 2020:current_year
+      )
+  ) %>%
+  mutate(data_type = factor(data_type, levels = unique(data_type)))
+
+## define the limites of the rectangles used to highlight areas of the graph
+regression_rectangles <-
+  actual_and_predicted_data %>% 
+  filter(data_type %in% c("Predicted","Actual")) %>%
+  group_by(data_type) %>%
+  summarise(year_min = min(year), year_max = max(year), deaths_min = min(number_of_deaths), deaths_max = max(number_of_deaths))
+
+## annotations to go above the rectangles
+predicted_data_annotations <- tribble(
+  ~x, ~y, ~label,
+  rowMeans(regression_rectangles[1,c("year_min","year_max")]), regression_rectangles[[1,"deaths_max"]]+2000, glue::glue("Predicted home deaths 2020-{current_year} using 2015-2019 data"),
+  rowMeans(regression_rectangles[2,c("year_min","year_max")]), regression_rectangles[[2,"deaths_max"]]+2000, glue::glue("Actual home deaths 2020-{current_year}*"),
+  rowMeans(regression_rectangles[2,c("year_min","year_max")]), regression_rectangles[[2,"deaths_min"]]-3000, "If the 2015-2019 trend continued we wouldn't observe this number of home deaths until 2031",
 )
@@
   ggplot(data = actual_and_predicted_data, aes(x=year, y=number_of_deaths)) +
-  geom_rect(aes(xmin = 2019.5, xmax = 2022.5, ymin = 15.5e3, ymax = 18.5e3), fill = "yellow", alpha = 0.3) +
-  geom_rect(aes(xmin = 2030, xmax = 2035, ymin = 19.5e3, ymax = 22e3), fill = "yellow", alpha = 0.3) +
-  geom_text(data = predicted_data_annotations, aes(x=x,y=y,label=str_wrap(label, width = 40))) +
+  # highlighting predicted data for 2020 onwards
+  statebins:::geom_rrect(aes(xmin = regression_rectangles[[1,"year_min"]]-0.5, xmax = regression_rectangles[[1,"year_max"]]+0.5, ymin = regression_rectangles[[1,"deaths_min"]]-1000, ymax = regression_rectangles[[1,"deaths_max"]]+1000), fill = alpha("grey",0), colour = "red") +
+  # highlighting actual data
+  statebins:::geom_rrect(aes(xmin = regression_rectangles[[2,"year_min"]]-0.5, xmax = regression_rectangles[[2,"year_max"]]+0.5, ymin = regression_rectangles[[2,"deaths_min"]]-1000, ymax = regression_rectangles[[2,"deaths_max"]]+1000), fill = alpha("grey",0), colour = "orange") +
+  geom_text(data = predicted_data_annotations, aes(x=x,y=y,label=str_wrap(label, width = 60))) +
+  geom_text(data = actual_and_predicted_data %>% filter(actual_year %in% 2020:current_year), aes(x = year, y = number_of_deaths + 500, label = if_else(actual_year==current_year, paste0(actual_year,"*"), as.character(actual_year))), angle = 30) +
   geom_point(aes(shape = data_type)) +
-  geom_line(data = predicted_data, aes(x = year, y = predicted), linetype = "dotted") +
+  geom_line(data = predicted_data %>% filter(year <= ceiling(max(regression_rectangles$year_max))), aes(x = year, y = predicted), linetype = "dotted") +
   theme_minimal() +
   theme(legend.position = "top", legend.title = element_blank()) +
-  xlim(c(2015,2037)) +
   scale_y_continuous(labels = scales::comma) +
-  labs(y = NULL, x = "Year (hypothetical)", title = "Increasing deaths at home", subtitle = str_wrap("Home deaths were on the increase before the pandemic.\nIf we assumed the same annual increases from 2015-2019 we wouldn't expect the pandemic levels of home deaths until 2031!", width = 90), caption = paste0(c(glue::glue("*Annual home deaths for 2022 extrapolated from data for weeks 1-{weeks_available_2022}"),source_nrs_text), collapse="\n"))
+  scale_x_continuous(breaks = 2015:2045, minor_breaks = NULL, limits = c(2015,ceiling(max(regression_rectangles$year_max)))) +
+  theme(axis.text.x = element_text(size = rel(0.8), angle = 45, hjust = 1)) +
+  labs(y = NULL, x = "Year (hypothetical)", title = "Increasing deaths at home", subtitle = str_wrap("Home deaths were on the increase before the pandemic.\nIf we assumed the same annual increases from 2015-2019 we wouldn't expect the pandemic levels of home deaths until 2031!", width = 90), caption = paste0(c(glue::glue("*Annual home deaths for 2023 extrapolated from data for weeks 1-{weeks_available_2023}"),source_nrs_text), collapse="\n"))
 

### 2023-02-22 covid_nrs_place_of_death_data_2023.Rmd
modified output tables to account for 2023 data

@@
     ,
-    ## 2022 data so far
+    ## 2022
+    merged_deaths_overall %>%
+      filter(age == "all" &
+               sex == "all" &
+               year == 2022
+             ) %>% 
+      group_by(place_of_death) %>%
+      summarise(
+        period = "2022",
+        historic_period = "2015-2019",
+        deaths_all_causes = sum(deaths_all_causes),
+        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
+        .groups = "drop"
+      ) %>% 
+      mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
+    left_join(
+      historical_annual_average_overall %>% 
+        select(place_of_death, mean_annual_deaths_2015_2019 = annual_deaths_mean, historic_period),
+      by = c("place_of_death","historic_period")
+    ),
+    ## 2023 data so far
     merged_deaths_overall %>%
@@
           sex == "all" &
-          year == 2022
+          year == 2023
       ) %>%
@@
     summarise(
-      period = glue::glue("2022 up to week {weeks_available_2023}"),
+      period = glue::glue("2023, up to week {weeks_available_2023}"),
       historic_period = "2015-2019",
@@
 
+# Causes of death over time, proportion inside and between places
+
+Table of number of deaths by cause, place of death and year, showing proportion of deaths by cause within each place `prop_cod_within_pod` (e.g.
+Covid deaths were 2% of home deaths) and proportion of deaths by place within each cause `prop_pod_within_cod` (e.g.
+70% of all Covid deaths happened in hospital).
+Note that because historic data was given averaged over the 5 year period, the historic figures are given as a mean over the period rather than an annual total.
+
+```{r}
+table_proportions_of_cause_of_death <-
+  bind_rows(
+    ## historic periods
+    historical_annual_average_cause_of_death %>%
+      rename(year = historic_period, deaths = annual_deaths_mean),
+    ## current data
+    total_deaths_cause_of_death %>%
+      select(-c(historic_period,annual_deaths_mean:ncol(.))) %>%
+      mutate(year = as.character(year))
+  ) %>%
+  filter(place_of_death!="All" & cause_of_death!="All") %>%
+  group_by(year,place_of_death) %>%
+  mutate(prop_cod_within_pod = deaths / sum(deaths)) %>%
+  group_by(year,cause_of_death) %>%
+  mutate(prop_pod_within_cod = deaths / sum(deaths)) %>%
+  ungroup %>%
+  (function(data_tbl){
+    bind_rows(
+      data_tbl,
+      ## all cause
+      data_tbl %>%
+        group_by(year, place_of_death) %>%
+        summarise(
+          cause_of_death = "All",
+          deaths = sum(deaths),
+          prop_cod_within_pod = 1,
+          .groups = "drop"
+        ) %>%
+        group_by(year, cause_of_death) %>%
+        mutate(prop_pod_within_cod = deaths/sum(deaths)) %>%
+        ungroup,
+      ## all place of death
+      data_tbl %>%
+        group_by(year, cause_of_death) %>%
+        summarise(
+          place_of_death = "All",
+          deaths = sum(deaths),
+          prop_pod_within_cod = 1,
+          .groups = "drop"
+        ) %>%
+        group_by(year, place_of_death) %>%
+        mutate(prop_cod_within_pod = deaths/sum(deaths)) %>%
+        ungroup
+    )
+  }) %>%
+  mutate(
+    year = factor(year, levels = unique(year)),
+    cause_of_death = factor(cause_of_death, levels = order_of_causes_of_death),
+    place_of_death = factor(place_of_death, levels = order_of_place_of_death_levels[c(5,1,3,2,4)])
+    ) %>%
+  arrange(year, place_of_death, cause_of_death)
+
+table_proportions_of_cause_of_death %>%
+  write_csv(file = paste0(dir_outputs,"/proportions_of_deaths_by_place_&_cause.csv"))
+```
+
 # TODO: other thigns like excess death & historical rates by geography

### 2023-03-03 visualisations_for_short_paper_feb2023.qmd
added script for visualisations for short paper

---
title: "Visualisations for short paper, Feb 2023"
author: "Jan savinc"
format: html
editor_options: 
  chunk_output_type: console
---

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text

## load functions
source("./functions.R")
```

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

# Dates represent event registration, not occurrence!

From NRS website:

  > All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.

# Load most recent data

```{r}
load(file = "./workspace.RData")
```

# Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
dir_paper <- file.path(dir_outputs, "short_paper_feb2023")
if (!dir.exists(dir_paper)) dir.create(dir_paper)
```

# Visualisations

## Common elements

```{r}
loadfonts(device="win")  # load imported fonts
# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
windowsFonts("Calibri" = windowsFont("Calibri"))  # to suppress the multiple error messages

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri"),
                  axis.text = element_text(size = 9)
            ))

date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023, '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
only_2023_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023)} (w/c {format(most_recent_date_available_2023)}).")

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

crude_rate_text <- "Total cases = crude cumulative positive cases over time."

death_registrations_text <- "Note: all figures are based on date of registration."

vertical_line_text <- "Vertical dashed line at week number 54 represents start of 2021."

five_year_historical_mean <- "For 2020-2021 data, the 5-year historical period is 2015-2019; for 2022, the period is 2016-2019 + 2021, skipping 2020."

vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"

# removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death.")

historic_period_2015_2019_text <- "Historic mean & range for 2015-2019 throughout."
historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
```

# Place of death image

## Wrangle data

```{r}
merged_deaths_overall_2015_2019_historic_period <-
  bind_rows(
    ## pre-2022 data
    merged_deaths_overall %>% filter(year < 2022),
    ## 2022 data
    merged_deaths_overall %>%
      filter(year >= 2022) %>%  # 2022 or later!?
      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
        current_data = ., 
        historical_data = 
          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
        num_weeks_to_average = 4)
  )
```

## With 2015-2019 historic range only

Note that the footnote to the image should be separate to the file so I'll remove the title and captions and export them as a text file instead.

```{r}
fig_paper_pod_2015_2019_range <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

fig_paper_pod_2015_2019_range %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")

fig_paper_pod_2015_2019_range_footnote <- c(
  date_range_text,
  paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n")
)

fig_paper_pod_2015_2019_range_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_", latest_data_modified_date, ".txt")))
```

### As above, but split into 1a,b,c

```{r}
p1 <- 
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

p2 <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Hospital")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = NULL
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

p3 <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Care home")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = NULL
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

fig_paper_pod_2015_2019_range_1abc <-
  p1 + (p2 / p3) +
    plot_layout( 
      guides = "collect",
      widths = c(2,1)
    ) +
    plot_annotation(tag_levels = "a", tag_suffix = ")") &
    # if (interactive()) theme_minimal() else NULL &  # for testing
    theme(
      legend.position = "bottom", 
      plot.tag = element_text(size = rel(1.1))
      )

fig_paper_pod_2015_2019_range_1abc %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range_abc_layout", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png",".tiff"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")

fig_paper_pod_2015_2019_range_1abc_footnote <- c(
  "a) Home & other non-insitution, b) Hospital, c) Care home; 'Other' place not shown due to small numbers.",
  fig_paper_pod_2015_2019_range_footnote
)

fig_paper_pod_2015_2019_range_1abc_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_1abc_footnote_", latest_data_modified_date, ".txt")))
```

### 2023-03-03 download_latest_data_and_render_reports.R

@@ -281,6 +281,13 @@ rmarkdown::render(
   envir = new.env(),
   quiet = TRUE
 )
+rmarkdown::render(
+  input = "./visualisations_for_short_paper_feb2023.qmd",
+  output_file = paste0("visualisations_for_short_paper_feb2023",today()),
+  output_dir = dir_reports,
+  envir = new.env(),
+  quiet = TRUE
+)
 # 
 # render(
 #   input = "./paper_reporting_death_at_home_increase.Rmd",

### 2023-03-03 covid_nrs_place_of_death_visualisations_2023.Rmd
moved function for saving plots to functions.r

@@
 historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
-```
-
-## Function to save files to multiple formats
-
-Saving to multiple formats is easy enough - the trouble is with saving PDF files with custom fonts, since they need to be embedded in the .pdf file. This requires:
-
-  * telling R where to load the font from, using `extrafont::font_import()` - this was done in a previous step
-* registering fonts with R using `extrafont::loadfonts()` - note: this isn't strictly necessary, since once imported, `extrafont` keeps track of fonts to load when the `extrafont` package is loaded
-* ~~telling R where the `gs` executable is, i.e. the installation of ghostscript on the current machine~~
-* ~~embedding the fonts in the `.pdf` files once they've been written, using `extrafont::embed_fonts()`~~
-  * instead of manually embedding fonts (which resulted in garbled text on my setup!), just use `device=device_cairo` as an argument to `ggsave()` when saving `.pdf` files and it just works
-
-```{r}
-save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {
-
-  walk(
-    .x = extensions,
-    .f = function(ext) {
-      # if (tolower(ext)==".pdf") device <- cairo_pdf
-      ggsave(filename = paste0(filename, ext), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (tolower(ext)==".pdf") cairo_pdf else NULL)
-    }
-  )
-}
 ```

### 2023-03-03 functions.R

@@ -303,3 +303,67 @@ infer_year_range_for_calculating_covid_deaths <- function(year) {
   year_range <- year_range[year_range>=2020]
   return(year_range)
 }
+
+
+## Function to save files to multiple formats Saving to multiple formats is easy
+## enough - the trouble is with saving PDF files with custom fonts, since they
+## need to be embedded in the .pdf file. This requires:
+## 1) telling R where to load the font from, using `extrafont::font_import()` - 
+## this was done in a previous step
+## 2) registering fonts with R using `extrafont::loadfonts()` - 
+## note: this isn't strictly necessary, since once imported, `extrafont` keeps
+## track of fonts to load when the `extrafont` package is loaded
+## 3) instead of manually embedding fonts (which resulted in garbled text on my
+## setup!), just use `device=device_cairo` as an argument to `ggsave()` when
+## saving `.pdf` files and it just works
+
+save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {
+  walk(
+    .x = extensions,
+    .f = function(ext) {
+      ggsave(
+        filename = paste0(filename, ext),
+        plot = plot,
+        width = width,
+        height = height,
+        units = units,
+        dpi = dpi,
+        device = 
+          if (tolower(ext) == ".pdf") cairo_pdf
+          else NULL
+      )
+    }
+  )
+}
+
+
+## helper function for plotting - adds a vertical dashed line and writes the
+## year at the start of the year, for the requested years
+## this saves me having to change graphs to add the year line to each
+add_vertical_lines_and_annotations_for_years <- function(years, y_annotation, shorten_year = TRUE) {
+  c(
+    map(
+      .x = years,
+      .f = function(year) {
+        annotate(
+          geom = "text",
+          x = compute_start_date_from_week_number(week_number = 1, year_number = year),
+          y = y_annotation,
+          label = if (shorten_year) paste0("'",str_sub(year,3,4)) else year,
+          hjust = 0,
+          colour = "grey40"
+        )
+      }
+    ),
+    map(
+      .x = years,
+      .f = function(year) {
+        geom_vline(
+          xintercept = compute_start_date_from_week_number(week_number = 1, year_number = year),
+          linetype = "dashed",
+          colour = "grey50"
+        )
+      }
+    )
+  )
+}

### 2023-04-20 covid_nrs_place_of_death_data_2023.Rmd
made changes to accommodate the disaggregatd data by age,sex,LA/HB ne longer being available alongside place of death added figure 2 for the paper

@@
 
+TODO: now that the ad-hoc files are no longer available, some of the data will need to be compiled from the tabs in the weekly data.
+
 ## Overall data
 
+Note: this contains data split by sex and age, but only for ALL places of death; individual PoD data are available for ALL sex and ALL age only
+
 ```{r}
 ## a separate raw file makes it easier to debug, and also reduces the number of requests I make while debugging!
-## Note: as of 16 Feb 2023, this returns a strange error
-# TODO: if error persists, load data from the main report file instead?
 ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")
@@
 
-```{r}
+Note: the last sex & age split of the data provided by NRS was Week 9 of 2023, though the data can still be split using the 'overall' dataset from Open Data Scotland above!
+
+```{r}
+# weekly_deaths_sex_age <-
+#   weekly_deaths_overall %>%
+#   filter(age!="all" & place_of_death != "All")
+
+# TODO: delete the below line, used for testing
+# files_weekly_deaths$main_report <- "./downloaded_data/2023-03-22_weekly-deaths-23.xlsx"
+
+sheet_covid_deaths_sex_and_age_data <- "1"
+sheet_all_cause_deaths_sex_and_age_data <- "4"
+
+cell_ranges_for_sex_and_age_data <- list(
+  "All" = "A6:K110",
+  "Female" = "A113:K217",
+  "Male" = "A220:K324"
+)
+
+# TODO: this still doesn't split data by PoD...
+weekly_deaths_sex_age <-
+  full_join(
+    x =
+      map_dfr(
+        .x = cell_ranges_for_sex_and_age_data,
+        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_covid_deaths_sex_and_age_data, range = .x), 
+        .id = "sex"
+      ) %>%
+      pivot_longer(
+        cols = `All ages`:`85+`,
+        names_to = "age",
+        values_to = "deaths_covid_related"
+      ) %>%
+      janitor::clean_names(),
+    y =
+      map_dfr(
+        .x = cell_ranges_for_sex_and_age_data,
+        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_all_cause_deaths_sex_and_age_data, range = .x), 
+        .id = "sex"
+      ) %>%
+      pivot_longer(
+        cols = `All ages`:`85+`,
+        names_to = "age",
+        values_to = "deaths_all_causes"
+      ) %>%
+      janitor::clean_names(),
+    by = c("sex", "registration_year", "week_number", "week_beginning", "age")
+  ) %>%
+  rename(year = registration_year, date_w_c = week_beginning) %>%
+  mutate(
+    year = as.integer(year),
+    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
+    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+    deaths_non_covid = deaths_all_causes - deaths_covid_related  # compute non-covid from all-cause minus covid
+  ) %>%
+  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
+
+# TODO: this is obsolete now, remove!
 weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
@@
 ```{r}
+# TODO: need to compute that from another source for the time being
 weekly_covid_contributory_vs_underlying_cause <-
@@
 
+# TODO: this is currently commented out as it doesn't compute correctly
 ## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure
-stopifnot(
-  (weekly_deaths_overall %>%
-  filter(sex=="all" & age=="all") %>%
-  inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
-  filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
-  nrow) == 0
-)
+
+# stopifnot(
+#   (weekly_deaths_overall %>%
+#   filter(sex=="all" & age=="all") %>%
+#   inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
+#   filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
+#   nrow) == 0
+# )
 
@@
 
+There is as of April 2023 a split in the availability of data:
+
+- Data are being continually updated for the all age all sex all HB/LA dataset
+- Data split by age, sex, HB or LA stopped being updated (technically they are still available but not without place of death data)
+
+Therefore there are two separate most recent available data, I'll call them 'basic' and 'disaggregated'/
+
 ```{r}
 ## constants for use in graphs & annotations
-weeks_available_2023 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2023])
-most_recent_date_available_2023 <- max(merged_deaths_age$date_w_c)
-
-earliest_date <- min(merged_deaths_age$date_w_c)
+weeks_available_2023 <- max(merged_deaths_overall$week_number[merged_deaths_overall$year==2023])
+most_recent_date_available_2023 <- max(merged_deaths_overall$date_w_c)
+
+earliest_date <- min(merged_deaths_overall$date_w_c)
 current_year <- isoyear(most_recent_date_available_2023)
+
+weeks_available_2023_disaggregated <- max(merged_deaths_age$week_number[merged_deaths_age$year==2023])
+most_recent_date_available_2023_disaggregated <- max(merged_deaths_age$date_w_c)
 ```

### 2023-04-20 covid_nrs_place_of_death_visualisations_2023.Rmd
made changes to accommodate the disaggregatd data by age,sex,LA/HB ne longer being available alongside place of death added figure 2 for the paper

@@
 date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023, '%d %B %Y')} (inclusive).")
+date_range_text_disaggregated <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023_disaggregated, '%d %B %Y')} (inclusive).")
 
@@
 only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
+
 only_2023_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023)} (w/c {format(most_recent_date_available_2023)}).")
+only_2023_disaggregated_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023_disaggregated)} (w/c {format(most_recent_date_available_2023_disaggregated)}).")
 
@@
   geom_point() +
-  geom_line(size = 1) +
-  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  geom_line(linewidth = 1) +
+  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
   scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
@@
   geom_point() +
-  geom_text(data = ~filter(.x, year==2022), aes(x=4.2, y=proportion, label="*"), inherit.aes = FALSE) +  # 4.2 here is 4th (final) place on the scale!
+  geom_text(data = ~filter(.x, year==as.character(latest_year)), aes(x=n_distinct(current_and_historical_deaths_by_cause_and_location$year)+0.2, y=proportion, label="*"), inherit.aes = FALSE) +  # cobble together the final year position from the data!
   facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
@@
   geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), size = 0.2) +
+  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), linewidth = 0.2) +
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
@@
   geom_point(aes(size = place_of_death)) +
-  geom_line(aes(size = place_of_death)) +
-  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  geom_line(aes(linewidth = place_of_death)) +
+  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
   facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
-  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
   scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
@@
     x = NULL, y= "Proportion of yearly deaths (%)",
-    subtitle = paste0("* ",only_2023_text),
-    caption = paste0(c("Health boards sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    subtitle = paste0("* ",only_2023_disaggregated_text),
+    caption = paste0(c("Health boards ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
     ) +
@@
     x = NULL, y= "Proportion of yearly deaths (%)",
-    subtitle = paste0("* ",only_2023_text),
-    caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    subtitle = paste0("* ",only_2023_disaggregated_text),
+    caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
     ) +
@@
   left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
-  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
   mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
@@
   geom_line() +
-  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
   facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
-  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
   scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
@@
     x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
-    subtitle = paste0("* ",only_2023_text),
-    caption = paste0(c("Health boards sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    subtitle = paste0("* ",only_2023_disaggregated_text),
+    caption = paste0(c("Health boards ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
     ) +
@@
   left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
-  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
   mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
@@
     x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
-    subtitle = paste0("* ",only_2023_text),
-    caption = paste0(c("Local Authorities sorted by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    subtitle = paste0("* ",only_2023_disaggregated_text),
+    caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
     ) +
@@
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+```
+
+## Absolute num. home deaths by LA
+
+```{r}
+# TODO: remove theme_minimal when done testing
+# TODO: add Scotland graph also
+# TODO: check  if all PoD would work?
+
+la_group <- tibble(  # group LAs into 4 groups of 8
+  ref_area = factor(order_la_by_population),
+  # la_group = rep(1:4, each = 8))
+  la_group = rep(1:6, times = c(6,5,5,5,5,6))
+  ) %>%
+  group_by(la_group) %>%
+  mutate(la_group_label = paste0(ref_area[1], " - ", ref_area[n()])) %>%
+  ungroup %>%
+  mutate(la_group_label = factor(la_group_label, levels = unique(la_group_label)))
+
+(fig_number_of_deaths_by_place_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(la_group, by = "ref_area") %>%  # group LAs
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  ggplot(. , aes(x=year, y=number_of_deaths, colour=ref_area, shape = ref_area)) +
+    theme_minimal() +
+    geom_point(size = rel(2)) +
+    geom_line() +
+    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=number_of_deaths, label="*"), inherit.aes = FALSE) +
+    facet_rep_wrap(~ la_group_label, repeat.tick.labels = TRUE, scales = "free_y") +
+    scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+    scale_y_continuous(label = scales::comma) +
+    scale_shape_manual(values = map(c(6,5,5,5,5,6), scales::shape_pal()) %>% unlist) +
+    scale_colour_manual(values = map(c(6,5,5,5,5,6), viridis::viridis_pal()) %>% unlist) +
+    theme(
+      legend.position = "top",
+      legend.title = element_blank(),
+      legend.text = element_text(size = 12),
+      plot.caption = element_text(size = 10, colour = "gray60"),
+      panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+      strip.background = element_rect(fill = NA, colour = "gray80"),
+      axis.text.x = element_text(angle = 45, hjust = 1)
+      ) +
+    guides(
+      colour = guide_legend(ncol = 6, title = NULL), 
+      shape = guide_legend(ncol = 6, title = NULL)
+      ) +
+    labs(
+      x = NULL, y = NULL,
+      title = "Deaths at home by LA",
+  subtitle = paste0("* ",only_2023_disaggregated_text),
+  caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+      ) +
+    NULL
+)
+
+fig_number_of_deaths_by_place_by_la %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_number_of_deaths_by_place_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
 ```
@@
        # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
-       caption = paste0(c(only_2023_text,source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2023_disaggregated_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
        title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
-       caption = paste0(c(only_2023_text,source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2023_disaggregated_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
        subtitle = "",
-       caption = paste0(c(only_2023_text,crude_rate_text,source_phs_text), collapse="\n"),
+       caption = paste0(c(only_2023_disaggregated_text,crude_rate_text,source_phs_text), collapse="\n"),
        fill = str_wrap("Cases per 100,000 population", width = 15)
@@
        title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board",
-       caption = paste0(c(only_2023_text,source_nrs_text), collapse="\n"),
+       caption = paste0(c(only_2023_disaggregated_text,source_nrs_text), collapse="\n"),
        fill = "Home death\nincrease relative\nto 2015-2019"
@@
   theme(axis.text.x = element_text(size = rel(0.8), angle = 45, hjust = 1)) +
-  labs(y = NULL, x = "Year (hypothetical)", title = "Increasing deaths at home", subtitle = str_wrap("Home deaths were on the increase before the pandemic.\nIf we assumed the same annual increases from 2015-2019 we wouldn't expect the pandemic levels of home deaths until 2031!", width = 90), caption = paste0(c(glue::glue("*Annual home deaths for 2023 extrapolated from data for weeks 1-{weeks_available_2023}"),source_nrs_text), collapse="\n"))
+  labs(y = NULL, x = "Year (hypothetical)", title = "Increasing deaths at home", subtitle = str_wrap("Home deaths were on the increase before the pandemic.\nIf we assumed the same annual increases from 2015-2019 we wouldn't expect the pandemic levels of home deaths until approx.
+2031!", width = 90), caption = paste0(c(glue::glue("*Annual home deaths for 2023 extrapolated from data for weeks 1-{weeks_available_2023}"),source_nrs_text), collapse="\n"))
 

### 2023-04-20 visualisations_for_short_paper_feb2023.qmd
made changes to accommodate the disaggregatd data by age,sex,LA/HB ne longer being available alongside place of death added figure 2 for the paper

@@
   filter(!place_of_death %in% c("All","Other")) %>%
+  mutate(place_of_death = factor(place_of_death, levels = order_of_place_of_death_levels[c(1,3,2)])) %>%
   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
@@
 ```
+
+# Place of death by LA
+
+```{r}
+fig_proportion_of_deaths_at_home_by_la <-
+  proportions_of_deaths_by_place_by_geography %>%
+  filter(geography == "la") %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
+  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
+  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
+  geom_point() +
+  geom_line() +
+  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
+  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
+  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
+  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
+  theme(
+    legend.position = "top",
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(size = 10, colour = "gray60"),
+    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
+    strip.background = element_rect(fill = NA, colour = "gray80"),
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    ) +
+  labs(
+    linetype = NULL,
+    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
+    subtitle = paste0("* ",only_2023_disaggregated_text),
+    # caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
+    ) +
+  NULL
+
+fig_proportion_of_deaths_at_home_by_la %>%
+save_output_file(filename = file.path(dir_paper, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png", ".tiff"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
+
+fig_proportion_of_deaths_at_home_by_la_footnote <- c(
+  "Local Authorities are shown in order of 2020 popoulation estimate.",
+  date_range_text_disaggregated,
+  paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
+)
+fig_proportion_of_deaths_at_home_by_la_footnote %>%
+  write_lines(file = file.path(dir_paper, paste0("fig_proportion_of_deaths_at_home_by_la_footnote_", latest_data_modified_date, ".txt")))
+```

### 2023-04-20 download_latest_data_and_render_reports.R

@@ -20,29 +20,44 @@ library(xml2)  # for the url_absolute() function
 
 # Find latest NRS weekly deaths spreadsheet URL ---------------------------
 
-url_nrs_weekly_deaths <- "https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland"
+## fixed url now and filename no longer updated weekly, so I need to rename the file myself after it's downloaded
+url_nrs_weekly_deaths_2023 <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/weekly-deaths/weekly-deaths-23.xlsx"
 
-all_nrs_links <-
-  read_html(url_nrs_weekly_deaths) %>% 
-  html_nodes("a") %>%
-  html_attr("href")  # find all links on the page
+## Note: previously the filename for the most recent spreadsheet reporting
+## weekly deaths changed weekly and reflected the date of publication (week
+## number & year); additionally, every 4 weeks or so the publication included
+## more detailed statistics, which was also reflected in the filename.
+## Therefore, I was inferring the latest filename by scraping the NRS website
+## and finding the most likely link. As of late March 2023, the website has been
+## renamed to no longer be Covid-specific (though the format of the report has
+## not changed and still contains Covid-related deaths). 
 
-nrs_filename_pattern <- "covid-deaths-(2\\d)-data.*xlsx"  # this will last for the decade... 202X
+## Below is the previously used code to infer the clatest URL for the report
 
-# TODO: incorporate both weekly and monthly data?
-weekly_deaths_url_relative <- # keep only links that are probably weekly deaths
-  tail(  # take last entry (n=1)
-    sort(  # sort so that last entry has the highest week number
-      unique(  # on some occasions, duplicate urls are extracted
-        all_nrs_links[which(
-          str_detect(all_nrs_links, pattern = nrs_filename_pattern) &
-          !str_detect(all_nrs_links, pattern = "monthly")  # remove "monthly" file
-          )])
-      ),
-    n = 1
-  )
[…]

### 2023-04-20 functions.R

@@ -330,7 +330,8 @@ save_output_file <- function(filename, extensions, plot, width, height, units, d
         dpi = dpi,
         device = 
           if (tolower(ext) == ".pdf") cairo_pdf
-          else NULL
+          else NULL,
+        bg = "white"
       )
     }
   )

### 2023-07-13 functions.R
renamed functions for clarity, made saving images more flexible

@@ -254,12 +254,22 @@ add_scotland_total_to_population_estimate <- function(data_tbl) {
 }
 
 ## compute y var from x, given a linear regression
-compute_y_from_x_using_regression <- function(x, regression_model) {
+compute_y_from_x_using_linear_regression <- function(x, regression_model) {
   coef(regression_model)[1] + (x * coef(regression_model)[2])
 }
 ## compute x var from y, given a linear regression
-compute_x_from_y_using_regression <- function(y, regression_model) {
+compute_x_from_y_using_linear_regression <- function(y, regression_model) {
   (y - coef(regression_model)[1]) / coef(regression_model)[2]
+}
+
+## compute y from x, given logistic regression
+compute_y_from_x_using_logistic_regression <- function(x, regression_model) {
+  predictor_var <- names(coef(regression_model)[2])
+  predict.glm(object = regression_model, newdata = tibble({{predictor_var}} := x), type = "response")
+}
+## compute x var from y, given a logistic regression
+compute_x_from_y_using_logistic_regression <- function(y, regression_model) {
+  (1/coef(model_proportions)[2])*(model_proportions$family$linkfun(y)-coef(model_proportions)[1])
 }
 
 ## parse the date given in ref_period depending on format
@@ -317,21 +327,18 @@ infer_year_range_for_calculating_covid_deaths <- function(year) {
 ## setup!), just use `device=device_cairo` as an argument to `ggsave()` when
 ## saving `.pdf` files and it just works
 
-save_output_file <- function(filename, extensions, plot, width, height, units, dpi) {
+save_output_file <- function(filename, extensions, plot, ...) {
   walk(
     .x = extensions,
     .f = function(ext) {
       ggsave(
         filename = paste0(filename, ext),
         plot = plot,
-        width = width,
-        height = height,
-        units = units,
-        dpi = dpi,
         device = 
           if (tolower(ext) == ".pdf") cairo_pdf
           else NULL,
-        bg = "white"
+        bg = "white",
+        ...
       )
     }
   )

### 2023-07-13 covid_nrs_place_of_death_visualisations_2023.Rmd
reorganised visualisations

@@
 
-# Proportion annual deaths by place
+# Number & Proportion annual deaths by place
+
+## Proportion only
 
@@
   merged_proportions_of_deaths_by_place %>%
-  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
+  filter(place_of_death!="All") %>%
   ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
@@
 )
+
 fig_proportion_annual_deaths_by_place %>%
 save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
+```
+
+## Number & proportion
+
+```{r}
+(fig_annual_deaths_by_place <-
+  wrap_plots(
+    # Number
+    merged_proportions_of_deaths_by_place %>%
+    ggplot(data = ., 
+           aes(x = year, y = number_of_deaths, colour = place_of_death, shape = place_of_death, linetype = place_of_death)
+           ) +
+    geom_line(linewidth = 1) +
+    geom_point(size = rel(2)) +
+    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=number_of_deaths, label="*"), inherit.aes = FALSE) +
+    theme_minimal() +
+    theme(legend.position = "top", legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
+    scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year)), minor_breaks = NULL) +
+    scale_y_continuous(labels = scales::comma, breaks = seq.int(from=0, to=7e4, by=1e4), minor_breaks = seq.int(from=0,t=7e4,by=5e3)) +
+    scale_colour_manual(values = c(scales::viridis_pal(option = "C")(4),"grey")) +
+    scale_shape_manual(values = c(scales::shape_pal()(4),NA)) +
+    scale_linetype_manual(values = c(rep("solid", 4),"dashed")) +
+    guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
+    labs(
+      title = "Number of deaths",
+      x = NULL, y = NULL
+    ),
+    # Proportion
+    merged_proportions_of_deaths_by_place %>%
+    filter(place_of_death!="All") %>%
+    ggplot(data = ., 
+           aes(x = year, y = proportion_of_total, colour = place_of_death, shape = place_of_death)
+           ) +
+    geom_line(linewidth = 1) +
+    geom_point(size = rel(2)) +
+    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
+    theme_minimal() +
+    theme(legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
+    scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year)), minor_breaks = NULL) +
+    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
+    scale_colour_viridis_d(option = "C") +
+    labs(
+      title = "Proportion of deaths",
+      x = NULL, y = NULL
+    ), 
+    guides = "collect"
+  ) +
+   plot_annotation(theme = theme(legend.position = "top"), subtitle = paste0("* ",only_2023_text),
+    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n"))
+ )
+
+fig_annual_deaths_by_place %>%
+save_output_file(filename = file.path(dir_outputs, paste0(c("fig_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
 ```
@@
   mutate(year = 2015 + years_since_2015,
-         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015))
+         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015), type = "response")
          )
@@
       data_type = "Predicted",
-      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
+      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_linear_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
       )
@@
         number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
-        year = 2015 + compute_x_from_y_using_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
+        year = 2015 + compute_x_from_y_using_linear_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
         data_type = "Actual",  # note: this is actual figures, but at the extrapolated year

### 2023-07-13 visualisations_for_short_paper_feb2023.qmd
reorganised short paper sript

@@
 only_2023_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023)} (w/c {format(most_recent_date_available_2023)}).")
+only_2023_disaggregated_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023_disaggregated)} (w/c {format(most_recent_date_available_2023_disaggregated)}).")
 
@@
 historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
+
 ```

### 2023-07-13 testing_gganimate_in_two_steps.R
i don't remember exactly but I think I made this more readable

@@ -1,30 +1,31 @@
 # script to render gganimate file as multiple images that will be rendered as a .gif or .mp4 at a later step
 
 library(tidyverse)
+library(ggplot2)
 library(gganimate)
 
 dir_animation <- "./animation_test/"
 if (!dir.exists(dir_animation)) dir.create(dir_animation)
 
 
-iris$group <- seq_len(nrow(iris))
-anim1 <- ggplot(iris, aes(Sepal.Width, Petal.Width, group = group)) +
+animated_plot <- 
+  ggplot(mtcars, aes(x = wt, y = hp, colour = as.factor(cyl))) +
   geom_point() +
-  labs(title = "{closest_state}") +
-  transition_states(Species, transition_length = 3, state_length = 1) +
+  transition_states(cyl, transition_length = 3, state_length = 1) +
   enter_fade() +
-  exit_fade()
+  exit_fade() +
+  labs(title = "Cyl: {closest_state}")
 
 ## save directly as gif
 anim_save(
   filename = file.path(dir_animation, "anim_test_1stage.gif"), 
-  animation = animate(anim1)
+  animation = animate(animated_plot)
   )
 
 ## save directly as mp4
 anim_save(
   filename = file.path(dir_animation, "anim_test_1stage.mp4"), 
-  animation = animate(anim1, 
+  animation = animate(animated_plot, 
                       renderer = av_renderer())
 )
 
@@ -38,7 +39,7 @@ duration = 10
 nframes = fps * duration
 
 animate(
-  anim1,
+  animated_plot,
   fps = fps,
   nframes = nframes,
   renderer = file_renderer(

### 2023-07-13 covid_nrs_place_of_death_data_2023.Rmd
small improvements, i guess?

@@
     mutate(proportion_of_total = number_of_deaths / total_deaths)
-  )
+  ) %>%
+  (function(x) {
+    bind_rows(
+      x,
+      x %>%
+        group_by(year) %>%
+        summarise(
+          place_of_death = "All",
+          number_of_deaths = sum(number_of_deaths, na.rm = TRUE),
+          total_deaths = unique(total_deaths),
+          proportion_of_total = 1
+          )
+    )
+  }) %>%
+  mutate(place_of_death = factor(place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4,5)])) %>%
+  arrange(year, place_of_death)
+
+write_csv(x = merged_proportions_of_deaths_by_place, file = file.path(dir_outputs, "annual_deaths_by_place.csv"))
 ```

### 2023-07-13 covid_nrs_place_of_death_visualisations_2022.Rmd
small improvements, i guess?

@@
       data_type = "Predicted",
-      number_of_deaths = if_else(year %in% 2020:2022, compute_y_from_x_using_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
+      number_of_deaths = if_else(year %in% 2020:2022, compute_y_from_x_using_linear_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
       )
@@
         number_of_deaths = if_else(year == 2022, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
-        year = 2015 + compute_x_from_y_using_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
+        year = 2015 + compute_x_from_y_using_linear_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
         data_type = "Actual"  # note: this is actual figures but

### 2023-07-13 visualisations_for_rse_curious_event.qmd
made new script for the RSE curious event, to maek a simple visualisation

--- 2023-03-03 visualisations_for_short_paper_feb2023.qmd
@@
 ---
-title: "Visualisations for short paper, Feb 2023"
-author: "Jan savinc"
+title: "Visualisations for RSE Curious event"
+author: "Jan Savinc"
 format: html
@@
 ```{r}
-dir_paper <- file.path(dir_outputs, "short_paper_feb2023")
-if (!dir.exists(dir_paper)) dir.create(dir_paper)
+dir_event <- file.path(dir_outputs, "rse_curious_2023")
+if (!dir.exists(dir_event)) dir.create(dir_event)
+```
+
+# Download & process monthly data
+
+This section maybe belongs in the earlier pipeline?
+
+NRS publish monthly mortality reports broken down by various variables, including Place of death over years, available since 2005.
+
+These are updated monthly, and reported for the previous month, so as of 12 July 2023, the latest report was published in June, and contains data for up to May 2023.]
+
+Note also that these exhibit small differences from the weekly data - the weekly data are reported to the ISO 8061 standard, meaning that week 1 of 2020 starts in 2019, for example.
+
+```{r}
+monthly_report_url <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/monthly-mortality/monthly-mortality-may-23-tab-5.xlsx"
+
+temporary_report_path <- tempfile()
+download.file(url = monthly_report_url, destfile = temporary_report_path, mode = "wb")
+
+# excel_sheets(temporary_report_path)
+# deaths_monthly <- read_excel(temporary_report_path, sheet = 5)  # more reliable to use name, though that is liable to change
+deaths_monthly <- 
+  read_excel(temporary_report_path, sheet = "Table M5", skip = 4) %>%
+  clean_names() %>%
+  rename(
+    year = year_of_occurrence,
+    month = month_of_occurrence,
+    place_of_death = location
+  ) %>%
+  select(-provisional_data_marker) %>%
+  mutate(
+    place_of_death = order_of_place_of_death_levels[match(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels, 1, 4))] %>%
+      factor(., levels = order_of_place_of_death_levels)
+  )
 ```
@@
 historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
+
 ```
 
-# Place of death image
+# Monthly place of death image
 
-## Wrangle data
+Monthly deaths might be too detailed...
 
 ```{r}
-merged_deaths_overall_2015_2019_historic_period <-
-  bind_rows(
-    ## pre-2022 data
-    merged_deaths_overall %>% filter(year < 2022),
-    ## 2022 data
-    merged_deaths_overall %>%
-      filter(year >= 2022) %>%  # 2022 or later!?
-      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
-      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
-      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
-      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
-        current_data = ., 
-        historical_data = 
-          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
-        num_weeks_to_average = 4)
-  )
-```
-
-## With 2015-2019 historic range only
-
-Note that the footnote to the image should be separate to the file so I'll remove the title and captions and export them as a text file instead.
-
-```{r}
-fig_paper_pod_2015_2019_range <-
-  merged_deaths_overall_2015_2019_historic_period %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(!place_of_death %in% c("All","Other")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non-Covid deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
+deaths_monthly %>%
+  mutate(
+    date = ymd(glue::glue("{year}-{month}-01"))
+  ) %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  filter(year >= 2015) %>%
+  # ggplot(aes(x = paste(year,month,sep="-"), number_of_deaths)) +
+  ggplot(aes(x = date, y = number_of_deaths)) +
+  add_vertical_lines_and_annotations_for_years(years = 2015:current_year, y_annotation = 30) +
+  geom_line() +
+  # facet_grid(place_of_death~.) +
   theme(
@@
     legend.title = element_blank(),
-    plot.margin = margin(5, 15, 5, 5),
     axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
@@
        ) +
-  ## legend to include historic range & mean
-  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
-  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
   NULL
+```
 
-fig_paper_pod_2015_2019_range %>%
-save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
+# Annual home deaths
 
-fig_paper_pod_2015_2019_range_footnote <- c(
-  date_range_text,
-  paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n")
+```{r}
+deaths_annual <-
+  deaths_monthly %>%
+  group_by(year, place_of_death) %>%
+  summarise(
+    number_of_deaths = sum(number_of_deaths),
+    .groups = "drop"
   )
 
-fig_paper_pod_2015_2019_range_footnote %>%
-  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_", latest_data_modified_date, ".txt")))
-```
+annual_deaths_by_place_incl_covid <-
+  bind_rows(
+    ## 2015-2019
+    merged_proportions_of_deaths_by_place %>%
+    select(year, place_of_death, number_of_deaths) %>%
+    filter(year %in% 2015:2019) %>%
+    mutate(deaths_non_covid = number_of_deaths, deaths_covid_related = 0)
+    ,
+    ## 2020 onwards
+    total_deaths_overall %>%
+      select(year, place_of_death, number_of_deaths = deaths_all_causes, deaths_covid_related) %>%
+      distinct %>%
+      mutate(deaths_non_covid = number_of_deaths - deaths_covid_related)
+  )
 
-### As above, but split into 1a,b,c
+# merged_proportions_of_deaths_by_place %>%
+#   select(year, place_of_death, number_of_deaths_from_weekly = number_of_deaths) %>%
+#   left_join(deaths_annual, by = c("year","place_of_death")) %>%
+#   filter(place_of_death!="All") %>%
+#   ggplot(aes(x=year, colour = place_of_death)) +
+#   geom_line(aes(y = number_of_deaths), linetype = "solid") +
+#   geom_point(aes(y = number_of_deaths), shape = 10) +
+#   geom_line(aes(y = number_of_deaths_from_weekly), linetype = "dashed") +
+#   geom_point(aes(y = number_of_deaths_from_weekly), shape = 15) +
+#   NULL
 
-```{r}
-p1 <- 
-  merged_deaths_overall_2015_2019_historic_period %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Home & other non-institution")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non-Covid deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = NULL,
-       y = "Deaths (N)"
-       ) +
-  ## legend to include historic range & mean
-  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
-  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
-  NULL
-
-p2 <-
-  merged_deaths_overall_2015_2019_historic_period %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Hospital")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non-Covid deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = NULL,
-       y = NULL
-       ) +
-  ## legend to include historic range & mean
-  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
-  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
-  NULL
-
-p3 <-
-  merged_deaths_overall_2015_2019_historic_period %>%
-  filter(sex == "all" & age == "all") %>%
-  filter(place_of_death %in% c("Care home")) %>%
-  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
-               names_to = "deaths_type", values_to = "deaths") %>% 
-  ggplot(aes(x = date_w_c)) +
-  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
-  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
-  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
-  geom_line(aes(y = deaths, colour = deaths_type)) +
-  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
-  facet_grid(~place_of_death) +
-  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
-                      labels = c("All deaths", 
-                                 "Non-Covid deaths"),
-                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
-                      ) +
-  theme(
-    legend.position = "top",
-    legend.title = element_blank(),
-    plot.margin = margin(5, 15, 5, 5),
-    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
-  ) +
-  labs(x = NULL,
-       y = NULL
-       ) +
-  ## legend to include historic range & mean
-  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
-  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
-  NULL
-
-fig_paper_pod_2015_2019_range_1abc <-
-  p1 + (p2 / p3) +
-    plot_layout( 
-      guides = "collect",
-      widths = c(2,1)
-    ) +
-    plot_annotation(tag_levels = "a", tag_suffix = ")") &
-    # if (interactive()) theme_minimal() else NULL &  # for testing
+(fig_annual_home_deaths <- 
+  annual_deaths_by_place_incl_covid %>%
+  filter(place_of_death == "Home & other non-institution") %>%
+  filter(year >= 2015 & year != 2023) %>%
+  pivot_longer(cols = c(deaths_covid_related, deaths_non_covid), names_to = "death_type", values_to = "deaths") %>%
+  mutate(bin_pandemic = year >= 2020) %>%
+  ggplot(aes(x = year, y = deaths, fill = death_type)) +
+  geom_col() +
   theme(
     legend.position = "bottom",
-      plot.tag = element_text(size = rel(1.1))
+    legend.title = element_blank(),
+    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1), 
+    aspect.ratio = 0.5
+  ) +
+  scale_x_continuous(breaks = 2015:2022) +
+  scale_y_continuous(labels = scales::comma) +
+  # scale_fill_manual(values = scales::viridis_pal(option = "D")(3)[c(3,1)],labels = c("Covid-related", "Non-covid")) +
+  scale_fill_manual(values = c("red","grey50"),labels = c("Covid-related", "Non-covid")) +
+  labs(x = NULL,
+       y = NULL,
+       title = "The new normal:\nHome deaths increased by a third during the pandemic...",
+       subtitle = "...and remain increased in 2023!"
+       ) +
+  NULL
 )
 
-fig_paper_pod_2015_2019_range_1abc %>%
-save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range_abc_layout", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png",".tiff"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
-
-fig_paper_pod_2015_2019_range_1abc_footnote <- c(
-  "a) Home & other non-insitution, b) Hospital, c) Care home; 'Other' place not shown due to small numbers.",
-  fig_paper_pod_2015_2019_range_footnote
-)
-
-fig_paper_pod_2015_2019_range_1abc_footnote %>%
-  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_1abc_footnote_", latest_data_modified_date, ".txt")))
+fig_annual_home_deaths %>%
+save_output_file(filename = file.path(dir_event, paste0(c("fig_annual_home_deaths", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
 ```

### 2023-07-31 visualisations_for_short_paper_feb2023.qmd
fixed Health Board to say LA instead; changed output so I generate both 300 and 600 dpi outputs

@@
   pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
-  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
+  mutate(measure = if_else(measure == "proportion_of_total", true = "Local Authority", false = "Scotland")) %>%
   ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
@@
 
+fig_proportion_of_deaths_at_home_by_la %>%
+save_output_file(filename = file.path(dir_paper, paste0(c("fig_proportion_of_deaths_at_home_by_la_600dpi", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png", ".tiff"), plot = ., width = 30, height = 25, dpi = 600, units = "cm")
+
 fig_proportion_of_deaths_at_home_by_la_footnote <- c(

### 2023-08-08 covid_nrs_place_of_death_data_2023.Rmd
changed cell ranges in the spreadsheet - they've been updated to only contain 2023 data instead of mix of 2022 & 2023

@@
 
+# TODO: the data have changed again, now the most recent file no longer includes 2022 and 2023, but only 2023, so the cell ranges below are inaccurate and result in errors!
+# 
+# cell_ranges_for_sex_and_age_data <- list(
+#   "All" = "A6:K110",
+#   "Female" = "A113:K217",
+#   "Male" = "A220:K324"
+# )
 cell_ranges_for_sex_and_age_data <- list(
-  "All" = "A6:K110",
-  "Female" = "A113:K217",
-  "Male" = "A220:K324"
+  "All" = "A6:K58",
+  "Female" = "A61:K113",
+  "Male" = "A116:K168"
 )
@@
 ## updated for 2023 which contains both 2022 and 2023
+# cell_ranges_for_causes_of_death_data <- list(
+#   "All" = "A6:V110",
+#   "Care home" = "A113:V217",
+#   "Home & other non-institution" = "A220:V324",
+#   "Hospital" = "A327:V431",
+#   "Other" = "A434:V538"
+# )
+
+## this was for only 2022 data
+## ...and is correct again for 2023
 cell_ranges_for_causes_of_death_data <- list(
-  "All" = "A6:V110",
-  "Care home" = "A113:V217",
-  "Home & other non-institution" = "A220:V324",
-  "Hospital" = "A327:V431",
-  "Other" = "A434:V538"
-)
-
-## this was for only 2022 data
-# cell_ranges_for_causes_of_death_data <- list(
-#   "All" = "A6:V58",
-#   "Care home" = "A61:V113",
-#   "Home & other non-institution" = "A116:V168",
-#   "Hospital" = "A171:V223",
-#   "Other" = "A226:V278"
-# )
+  "All" = "A6:V58",
+  "Care home" = "A61:V113",
+  "Home & other non-institution" = "A116:V168",
+  "Hospital" = "A171:V223",
+  "Other" = "A226:V278"
+)
 

### 2023-08-08 regressions_for_short_paper_jun2023.qmd
reporting regressions for the short paper, using a linear and a logistic model

---
title: "Increased home deaths regressions"
author: "Jan Savinc"
format: html
editor_options: 
  chunk_output_type: console
---

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text

## load functions
source("./functions.R")
```

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

# Dates represent event registration, not occurrence!

From NRS website:

  > All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.

# Load most recent data

```{r}
load(file = "./workspace.RData")
```

# Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
dir_paper <- file.path(dir_outputs, "short_paper_feb2023")
if (!dir.exists(dir_paper)) dir.create(dir_paper)
```

# Visualisations

## Common elements

```{r}
loadfonts(device="win")  # load imported fonts
# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
windowsFonts("Calibri" = windowsFont("Calibri"))  # to suppress the multiple error messages

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri"),
                  axis.text = element_text(size = 9)
            ))

date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023, '%d %B %Y')} (inclusive).")
date_range_text_disaggregated <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023_disaggregated, '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
only_2023_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023)} (w/c {format(most_recent_date_available_2023)}).")
only_2023_disaggregated_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023_disaggregated)} (w/c {format(most_recent_date_available_2023_disaggregated)}).")

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

crude_rate_text <- "Total cases = crude cumulative positive cases over time."

death_registrations_text <- "Note: all figures are based on date of registration."

vertical_line_text <- "Vertical dashed line at week number 54 represents start of 2021."

five_year_historical_mean <- "For 2020-2021 data, the 5-year historical period is 2015-2019; for 2022, the period is 2016-2019 + 2021, skipping 2020."

vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"

# removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death.")

historic_period_2015_2019_text <- "Historic mean & range for 2015-2019 throughout."
historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
```

# Simple regressions

In this section I calculate a simple linear regression over home deaths 2015-2019 and predict the expected number of home deaths in 2020-2023, to contrast with the actual number of home deaths.
I also calculate the expected proportion of home deaths from the number of home deaths and total home deaths in 2015-2019 and predict the expected proportion of home deaths in 2020-2023, also to contrast with actual numbers.

The regression is simply number of deaths regressed over year, specifically, number of years since 2015. This allows for the easy interpretation of the intercept as the estimated home deaths in 2015, and the beta coefficient as the annual growth in the number of deaths. Seasonality is not explicitly modeled but is captured by the annual data.

In the logistic regression, the dependent variable is the proportion of home deaths of all deaths, or the odds of home death to non-home death in each year, regressed on the year (number of years since 2015)

## Defining regression data

```{r}
regression_data <- 
  merged_proportions_of_deaths_by_place %>% 
  filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>% 
  mutate(years_since_2015 = 0:(n()-1))
```

## Preliminary: extrapolating 2023 data

There are complete data for 2015-2022, but not 2023, for which there are data for weeks 1-`r weeks_available_2023`. For this reason, I estimate the total number of deaths in 2023 by estimating the proportion of average annual deaths represented by weeks 1-`r weeks_available_2023` in 2015-2019, and then dividing deaths so far in 2023 by the proportion, thereby obtaining an estimated total for 2023.

```{r}
## 2023 only has X weeks' worth of data available; we can extrapolate the annual
## total by computing the percentage of annual data represented by previous
## years at X weeks, and then averaging across those, or taking the min & max
extrapolated_2023_data <-
  historical_weekly_average_overall %>%
  filter(place_of_death %in% c("Home & other non-institution","All") & historic_period == "2015-2019") %>%
  group_by(place_of_death) %>%
  summarise(
    deaths_mean_so_far = sum(deaths_mean[week_number <= weeks_available_2023], na.rm= TRUE),
    deaths_mean_total = sum(deaths_mean, na.rm= TRUE),
    proportion_of_deaths_so_far = deaths_mean_so_far / deaths_mean_total
    )
```

## Predicting absolute number of home deaths

```{r}
model_home_deaths <- lm(data = regression_data, formula = number_of_deaths ~ years_since_2015)

home_deaths_actual_and_predicted_data <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
      mutate(data_type = "Historic")
    ,
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
    mutate(
      data_type = "Predicted",
      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_linear_regression(x = year-2015, regression_model = model_home_deaths), number_of_deaths)
      )
    ,
    merged_proportions_of_deaths_by_place %>%
      select(year, number_of_deaths, place_of_death) %>%
      filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
      mutate(
        number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2023_data$proportion_of_deaths_so_far, number_of_deaths),
        year = 2015 + compute_x_from_y_using_linear_regression(y = number_of_deaths, regression_model = model_home_deaths),  # compute predicted X
        data_type = "Actual",  # note: this is actual figures, but at the extrapolated year
        actual_year = 2020:current_year
      )
  ) %>%
  mutate(data_type = factor(data_type, levels = unique(data_type)))
```

## Predicting proportion of home deaths

This takes into account both the number of deaths at home and total number of deaths.

A logistic regression is used via R's `glm(family = binomial)` function, with 'successes' representing home deaths and 'failures' representing non-home deaths.

```{r}
## response variable is formatted as a 2-column matrix of successes and failures
model_proportions <-
  glm(data = regression_data, formula = cbind(number_of_deaths, total_deaths-number_of_deaths) ~ years_since_2015, family = "binomial")

## note this is equivalent to glm(data = regression_data, formula = proportion_of_total ~ years_since_2015, weights = total_deaths, family = "binomial")

model_proportions_predicted <- tibble(years_since_2015 = seq(from=0, to=30, by = 0.1)) %>%
  mutate(year = 2015 + years_since_2015,
         predicted = predict.glm(object = model_proportions, newdata = tibble(years_since_2015), type = "response")
         )

proportions_actual_and_predicted_data <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
      mutate(data_type = "Historic")
    ,
    merged_proportions_of_deaths_by_place %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
    mutate(
      data_type = "Predicted",
      proportion_of_total = if_else(year %in% 2020:current_year, compute_y_from_x_using_logistic_regression(x = year-2015, regression_model = model_proportions), number_of_deaths)
      )
    ,
    merged_proportions_of_deaths_by_place %>%
      filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
      mutate(
        number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2023_data$proportion_of_deaths_so_far, number_of_deaths),
        # year = as.numeric(2015 + MASS::dose.p(model_proportions, p=proportion_of_total)),  # compute predicted X
        year = as.numeric(2015 + compute_x_from_y_using_logistic_regression(y = proportion_of_total, regression_model = model_proportions)),  # compute predicted X
        data_type = "Actual",  # note: this is actual figures, but at the extrapolated year
        actual_year = 2020:current_year
      )
  ) %>%
  mutate(data_type = factor(data_type, levels = unique(data_type)))

```

## Report estimates from models

```{r}
table_estimates <-
  bind_rows(
    ## actual data
    merged_proportions_of_deaths_by_place %>%
      filter(place_of_death == "Home & other non-institution" & year %in% 2015:2023) %>%
      select(-place_of_death) %>%
      rename(deaths_home = number_of_deaths, deaths_all_places = total_deaths, proportion_home_deaths = proportion_of_total) %>%
      mutate(data = "Actual")
    ,
    ## extrapolated 2023 data
    merged_proportions_of_deaths_by_place %>%
      filter(year %in% 2023) %>%
      inner_join(extrapolated_2023_data, by = "place_of_death") %>%
      group_by(year, place_of_death) %>%
      summarise(deaths_extrapolated = number_of_deaths / proportion_of_deaths_so_far, .groups = "drop") %>%
      pivot_wider(names_from = place_of_death, values_from = deaths_extrapolated) %>%
      rename(deaths_home = `Home & other non-institution`, deaths_all_places = All) %>%
        mutate(proportion_home_deaths = merged_proportions_of_deaths_by_place$proportion_of_total[merged_proportions_of_deaths_by_place$place_of_death=="Home & other non-institution" & merged_proportions_of_deaths_by_place$year == 2023]) %>%  # infer proportion of home deaths
      mutate(data = "Extrapolated")
    ) %>%
    mutate(
      predicted_year_linear = if_else(
        year %in% c(2015:2019, 2023) & data == "Actual",
        NA_real_,
        2015 + compute_x_from_y_using_linear_regression(y = deaths_home, regression_model = model_home_deaths)
      ),
      predicted_year_logistic = if_else(
        year %in% c(2015:2019, 2023) & data == "Actual",
        NA_real_,
        2015 + compute_x_from_y_using_logistic_regression(y = proportion_home_deaths, regression_model = model_proportions)
      )
    ) %>%
    relocate(data)

write_csv(table_estimates, file = file.path(dir_paper, "table_actual_vs_predicted_data.csv"))
```

## Report model summaries

Note: regression coefficients for logistic model were exponentiated.

```{r}
table_model_summaries <-
  bind_rows(
    broom::tidy(model_home_deaths) %>%
      mutate(model = "Linear model, absolute number of home deaths"),
    broom::tidy(model_proportions, exponentiate = TRUE) %>%
      mutate(model = "Logistic model, proportion of home deaths")
  ) %>%
  relocate(model)

write_csv(table_model_summaries, file = file.path(dir_paper, "table_regression_models.csv"))
```

# Regression with covariates

Note: decided not to compute those - too much effort for too little gain!

# Previous estimates

These are inferred from the paper:

> Finucane, A.M., Bone, A.E., Evans, C.J. et al. The impact of population ageing on end-of-life care in Scotland: projections of place of death and recommendations for future service provision. BMC Palliat Care 18, 112 (2019). https://doi.org/10.1186/s12904-019-0490-x

Four different scenarios are provided with different assumptions of relative population mortality growth. They calculated the proportion of people in each age and gender group in each place of death for 2004-2016; the estimated age and gender-specific proportions of deaths by place of death were then applied to the ONS projected deaths. Scenario 4 is unrealistic so it won't be considered here (it assumed that home & care home deaths do not increase above their absolute numbers in 2016).

Note that even the highest estimate computes around 19k home deaths in 2040, whereas there were more than 20k home deaths in 2020.

### 2023-08-08 get_scotland_health_boards.R
script to download health board codes

@@ -1,0 +1,10 @@
+## Script to read health board codes and descriptive names and save them as .csv
+
+library(tidyverse)
+raw_hbs <- read_csv(file ="https://www.opendata.nhs.scot/datastore/dump/652ff726-e676-4a20-abda-435b98dd7bdc?bom=True")
+
+clean_hbs <-
+  raw_hbs %>%
+  select(hb_code = HB, health_board_name = HBName)
+
+write_csv(x = clean_hbs, file = "./scotland_health_board_names.csv")

### 2024-02-02 covid_nrs_place_of_death_visualisations_2024.Rmd
quick run through visualisations file, updating for 2024

---
title: "Place of death in NRS COVID-19 deaths: visualisation for 2024 data"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text

## load functions
source("./functions.R")
```

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

# Dates represent event registration, not occurrence!

From NRS website:

  > All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.

# Load most recent data

```{r}
load(file = "./workspace.RData")
```

# Visualisations

## Common elements

```{r}
loadfonts(device="win")  # load imported fonts
# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
windowsFonts("Calibri" = windowsFont("Calibri"))  # to suppress the multiple error messages

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri"),
                  axis.text = element_text(size = 9)
            ))

date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2024, '%d %B %Y')} (inclusive).")
date_range_text_disaggregated <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2024_disaggregated, '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")

only_2024_text <- glue::glue("Data for 2024 only include weeks 1-{format(weeks_available_2024)} (w/c {format(most_recent_date_available_2024)}).")
only_2024_disaggregated_text <- glue::glue("Data for 2024 only include weeks 1-{format(weeks_available_2024_disaggregated)} (w/c {format(most_recent_date_available_2024_disaggregated)}).")

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

crude_rate_text <- "Total cases = crude cumulative positive cases over time."

death_registrations_text <- "Note: all figures are based on date of registration."

vertical_line_text <- "Vertical dotted line at week number 54 represents start of 2021."

five_year_historical_mean <- "For 2020-2021 data, the 5-year historical period is 2015-2019; for 2022, the period is 2016-2019 + 2021, skipping 2020."

vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"

# removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death.")

historic_period_2015_2019_text <- "Historic mean & range for 2015-2019 throughout."
historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
```

# Place of death over time

```{r}
## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
annotations_fig_place_of_death_explaining_historic_range <- tribble(
   ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
  "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
  "curve", "Hospital", NA, 29, 830, 32, 550,
  "text", "Hospital", "Historic mean", 22, 800, NA, NA,
  "curve", "Hospital", NA, 22, 780, 26, 510
) %>% 
  mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)

fig_place_of_death <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2024), linetype = "dashed", colour="grey50") +  # start of 2024
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
  geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2024-01-01"), y=30, label = "'24", hjust = 0, colour = "grey40") +
  NULL

fig_place_of_death %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
```

## With 2015-2019 historic range only

```{r}
merged_deaths_overall_2015_2019_historic_period <-
  bind_rows(
    ## pre-2022 data
    merged_deaths_overall %>% filter(year < 2022),
    ## 2022 data
    merged_deaths_overall %>%
      filter(year >= 2022) %>%  # 2022 or later!?
      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
        current_data = ., 
        historical_data = 
          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
        num_weeks_to_average = 4)
  )

fig_place_of_death_2015_2019_range <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2024), linetype = "dashed", colour="grey50") +  # start of 2024
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
  geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2024-01-01"), y=30, label = "'24", hjust = 0, colour = "grey40") +
  NULL

fig_place_of_death_2015_2019_range %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
```

## Home deaths only

```{r}
fig_place_of_death_home_only <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2024), linetype = "dashed", colour="grey50") +  # start of 2024
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,historic_period_2022_text,source_nrs_text), collapse="\n")
       ) +
  ## annotate historic range
  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2024-01-01"), y=30, label = "'24", hjust = 0, colour = "grey40") +
  NULL

fig_place_of_death_home_only %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_only", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
```

## Home deaths with 2015-2019 historic period means & range only

```{r}
fig_place_of_death_home_2015_2019_historic_period <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%  # home death only
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2024), linetype = "dashed", colour="grey50") +  # start of 2024
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n")
       ) +
  ## annotate historic range
  annotate(geom="text",x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 50, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 29, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 33, year_number = 2020), y = 65, yend = 210, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom="text", x = compute_start_date_from_week_number(week_number = 20, year_number = 2020), y = 100, label = "Historic mean", colour = "grey20") +
  annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2024-01-01"), y=30, label = "'24", hjust = 0, colour = "grey40") +
  NULL

fig_place_of_death_home_2015_2019_historic_period %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_home_2015-2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 14, dpi = 300, units = "cm")
```

# Number & Proportion annual deaths by place

## Proportion only

Here we'll show 4 lines, one for each place of death, for each year so far.

```{r}
(fig_proportion_annual_deaths_by_place <-
  merged_proportions_of_deaths_by_place %>%
  filter(place_of_death!="All") %>%
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(linewidth = 1) +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  # theme_minimal() +
  theme(
    legend.position = "top", 
    plot.caption = element_text(size = 10, colour = "gray60")
    ) +
  labs(
    x = "Year", y = "Proportion of yearly deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2024_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)

fig_proportion_annual_deaths_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
```

## Number & proportion

```{r}
(fig_annual_deaths_by_place <-
  wrap_plots(
    # Number
    merged_proportions_of_deaths_by_place %>%
    ggplot(data = ., 
           aes(x = year, y = number_of_deaths, colour = place_of_death, shape = place_of_death, linetype = place_of_death)
           ) +
    geom_line(linewidth = 1) +
    geom_point(size = rel(2)) +
    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=number_of_deaths, label="*"), inherit.aes = FALSE) +
    theme_minimal() +
    theme(legend.position = "top", legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year)), minor_breaks = NULL) +
    scale_y_continuous(labels = scales::comma, breaks = seq.int(from=0, to=7e4, by=1e4), minor_breaks = seq.int(from=0,t=7e4,by=5e3)) +
    scale_colour_manual(values = c(scales::viridis_pal(option = "C")(4),"grey")) +
    scale_shape_manual(values = c(scales::shape_pal()(4),NA)) +
    scale_linetype_manual(values = c(rep("solid", 4),"dashed")) +
    guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
    labs(
      title = "Number of deaths",
      x = NULL, y = NULL
    ),
    # Proportion
    merged_proportions_of_deaths_by_place %>%
    filter(place_of_death!="All") %>%
    ggplot(data = ., 
           aes(x = year, y = proportion_of_total, colour = place_of_death, shape = place_of_death)
           ) +
    geom_line(linewidth = 1) +
    geom_point(size = rel(2)) +
    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
    theme_minimal() +
    theme(legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year)), minor_breaks = NULL) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_colour_viridis_d(option = "C") +
    labs(
      title = "Proportion of deaths",
      x = NULL, y = NULL
    ), 
    guides = "collect"
  ) +
   plot_annotation(theme = theme(legend.position = "top"), subtitle = paste0("* ",only_2024_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n"))
 )

fig_annual_deaths_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
```

# Proportion of cumulative deaths by place, over the course of a year

First we compute the cumulative proportion of place of death within each year. As the weeks go by, the proportions of deaths in each place change, and they change at different rates.

```{r}
cumulative_deaths_by_place_by_year <-
  bind_rows(
    ## current data
    merged_deaths_overall %>%
      filter(age == "all" & sex == "all"),
    ## 2015-2019
    weekly_deaths_2015_2019_overall %>%
      mutate(year = as.integer(year)) %>%  # remove the 'average' figures
      drop_na(year) %>%
      rename(deaths_all_causes = number_of_deaths)
  ) %>%
  arrange(year, week_number) %>%
  group_by(year, place_of_death) %>%
  mutate(
    cumulative_deaths = cumsum(deaths_all_causes)
  ) %>%
  group_by(year, week_number, place_of_death == "All"
  ) %>%
  mutate(
    proportion_cumulative_deaths = cumulative_deaths / sum(cumulative_deaths)
  ) %>%
  ungroup %>%
  select(year, place_of_death, week_number, matches("cumulative_deaths"))

cumulative_deaths_by_place_by_year_min_max_2015_2019 <-
  cumulative_deaths_by_place_by_year %>%
  filter(year %in% 2015:2019) %>%
  group_by(place_of_death, week_number) %>%
  summarise(
    proportion_cumulative_deaths_min = min(proportion_cumulative_deaths),
    proportion_cumulative_deaths_max = max(proportion_cumulative_deaths),
    .groups = "drop"
  )
```

## Weekly proportion of deaths by place

These are non-cumulative proportions of place of death for each week. You can see how the proportions of place of death are fairly noisy over time, and in 2015-2019 they are relatively stable; it's not until 2020 that a big shift occurs.

```{r}
fig_weekly_proportion_of_death_by_place <-
  bind_rows(
    ## current data
    merged_deaths_overall %>%
      filter(age == "all" & sex == "all"),
    ## 2015-2019
    weekly_deaths_2015_2019_overall %>%
      mutate(year = as.integer(year)) %>%  # remove the 'average' figures
      drop_na(year) %>%
      rename(deaths_all_causes = number_of_deaths)
  ) %>%
  group_by(year, week_number, place_of_death == "All") %>%
  mutate(proportion_of_deaths = deaths_all_causes / sum(deaths_all_causes)) %>%
  ungroup %>%
  filter(place_of_death != "All") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(
    data = ., aes(x = week_number, y = proportion_of_deaths, colour = place_of_death)
  ) +
  geom_line() +
  facet_wrap(~year) +
  scale_colour_viridis_d() +
  scale_x_continuous(breaks = scales::breaks_extended(6)) +
  scale_y_continuous(breaks = scales::breaks_extended(6), label = function(x) scales::percent(x = x, accuracy = 1)) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
  ) +
  labs(
    x = "Week number", y = "Weekly proportion of deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2024_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
  ) +
  NULL

fig_weekly_proportion_of_death_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_proportion_of_death_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
```

## Yearly graphs, 2015-2022

This is a graph to see whether the relative proportion of deaths by place changes over time - we know the annual proportions for the 4 places of death, for example, but their relative proportion changes over the year.
This gives us context to be able to decide whether a decline in home deaths early in the year represents a genuine decline or just the general trend of home deaths being lower early in a year.

```{r}
fig_cumulative_weekly_proportion_of_deaths_by_place <-
  cumulative_deaths_by_place_by_year %>%
  filter(place_of_death != "All") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(data = ., aes(x = week_number, y = proportion_cumulative_deaths, group = place_of_death, colour = place_of_death)) +
  geom_line() +
  facet_wrap(~year) +
  scale_colour_viridis_d() +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks = scales::breaks_extended(6)) +
  scale_y_continuous(breaks = scales::breaks_extended(6), label = function(x) scales::percent(x = x, accuracy = 1)) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
  ) +
  labs(
    x = "Week number", y = "Cumulative proportion of place of death (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2024_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
  ) +
  NULL

fig_cumulative_weekly_proportion_of_deaths_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cumulative_weekly_proportion_of_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
```

## Cumulative proportion of place of death, showing 2015-2019 as range

This is a similar graph to the above, but we include the weekly range (min/max) of the weekly proportions of place of death, as historic context for the data from 2020 onwards.

```{r}
fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range <-
  cumulative_deaths_by_place_by_year %>%
  filter(place_of_death != "All") %>%
  filter(!year %in% 2015:2019) %>%
  left_join(cumulative_deaths_by_place_by_year_min_max_2015_2019, by = c("week_number", "place_of_death")) %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(data = ., aes(x = week_number, y = proportion_cumulative_deaths, group = place_of_death, colour = place_of_death)) +
  geom_line() +
  geom_ribbon(aes(ymin = proportion_cumulative_deaths_min, ymax = proportion_cumulative_deaths_max, fill = place_of_death, colour = NULL), alpha = 0.4) +
  facet_wrap(~year) +
  scale_colour_viridis_d() +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks = scales::breaks_extended(6)) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
  ) +
  labs(
    x = "Week number", y = "Cumulative proportion of place of death (%)",
    subtitle = paste0("* ",only_2024_text),
    caption = paste0(c("Shaded areas represent the range of cumulative weekly proportion of deaths by place for the 2015-2019 period",death_registrations_text, source_nrs_text),collapse="\n")
  ) +
  NULL

fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cumulative_weekly_proportion_of_deaths_by_place_with_historic_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
```

# Causes of death compared to historic context

Here I first merge the 2022 data series with the historic period 2015-2019 averages for weekly figures - this way we can compare the total in 2022 so far with the historical figures up to the same week.

```{r}
merged_deaths_cause_of_death_with_2015_2019_historic_period <-
  bind_rows(
    merged_deaths_cause_of_death %>%
      filter(year %in% 2020:2021)
    ,
    merged_deaths_cause_of_death %>%
      filter(year == 2022) %>%
      select(-c(historic_period,deaths_mean)) %>%
      left_join(
        merged_deaths_cause_of_death %>%
          filter(year==2021) %>%
          select(place_of_death, week_number, cause_of_death, historic_period, deaths_mean),
        by = c("place_of_death", "week_number", "cause_of_death")
      )
  )

total_deaths_cause_of_death_so_far <-
  merged_deaths_cause_of_death_with_2015_2019_historic_period %>%
  group_by(year, cause_of_death, place_of_death, historic_period) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths, na.rm = TRUE),
    deaths_mean = sum(deaths_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(year, place_of_death, historic_period) %>%
  mutate(
    proportion_all_deaths = number_of_deaths / sum(number_of_deaths, na.rm = TRUE),
    proportion_all_deaths_historic = deaths_mean / sum(deaths_mean, na.rm = TRUE)
  ) %>%
  ungroup
```

## Cuases so far by place

```{r}
(fig_deaths_so_far_by_cause <-
  total_deaths_cause_of_death_so_far %>%
  filter(!place_of_death %in% c("Other","All")) %>%
  group_by(cause_of_death, place_of_death, historic_period) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths, na.rm = TRUE),
    deaths_mean = sum(deaths_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(place_of_death, historic_period) %>%
  mutate(
    proportion_all_deaths = number_of_deaths / sum(number_of_deaths, na.rm = TRUE),
    proportion_all_deaths_historic = deaths_mean / sum(deaths_mean, na.rm = TRUE)
  ) %>%
  ungroup %>%
  mutate(
     cause_of_death = fct_rev(factor(cause_of_death)),
     place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))
     ) %>%
  ggplot() +
  geom_col(aes(y = cause_of_death, fill = cause_of_death, x = number_of_deaths), alpha = 0.8, colour = "grey10") +
  geom_segment(aes(y = as.numeric(cause_of_death)-0.45, yend = as.numeric(cause_of_death)+0.45, x = deaths_mean, xend = deaths_mean), size = 0.8, colour = "grey1") +
  scale_fill_viridis_d() +
  facet_grid(~place_of_death) +
  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
  scale_y_discrete(labels = ~str_wrap(.x, width = 20)) +
  theme(
    legend.position = "none", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    legend.title = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  labs(
    x = NULL, y = NULL,
    title = str_wrap("Total deaths since start of 2020 by cause and place of death, compared to 2015-2019", width = 60),
    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2024_text)), collapse="\n"),
    caption = paste0(c(death_registrations_text, source_nrs_text),collapse="\n")
  ) +
  NULL)

fig_deaths_so_far_by_cause %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_deaths_so_far_by_cause", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 12, dpi = 300, units = "cm")
```

## Causes by place & year

```{r}
(fig_annual_deaths_by_cause <-
  total_deaths_cause_of_death_so_far %>%
  filter(!place_of_death %in% c("Other","All")) %>%
  mutate(
     cause_of_death = fct_rev(factor(cause_of_death)),
     place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))
     ) %>%
  ggplot() +
  geom_col(aes(y = cause_of_death, fill = cause_of_death, x = number_of_deaths), alpha = 0.8, colour = "grey10") +
  geom_segment(aes(y = as.numeric(cause_of_death)-0.45, yend = as.numeric(cause_of_death)+0.45, x = deaths_mean, xend = deaths_mean), size = 0.8, colour = "grey1") +
  scale_fill_viridis_d() +
  facet_grid(year~place_of_death) +
  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
  scale_y_discrete(labels = ~str_wrap(.x, width = 20)) +
  theme(
    legend.position = "none", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    legend.title = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  labs(
    x = NULL, y = NULL,
    title = "Annual deaths by cause and place of death, compared to 2015-2019",
    subtitle = paste0(c("Bars are no. of deaths, line segments are mean annual deaths 2015-2019", paste0("* ",only_2024_text)), collapse="\n"),
    caption = paste0(c(death_registrations_text, source_nrs_text),collapse="\n")
  ) +
  NULL)

fig_annual_deaths_by_cause %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_annual_deaths_by_cause", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
```

# Causes of death over time

## Proportion of causes by place, yearly

This graph shows each of the causes of death divided by place of death - e.g. what proportion of cancer deaths happened at home vs hospital over time?

```{r}
fig_cause_of_death_annual_proportion_by_place <-
  current_and_historical_deaths_by_cause_and_location %>%
  filter(!cause_of_death %in% c("All")) %>%
  filter(!year %in% c("2020-2021")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  ggplot(aes(x = year, y = proportion, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  geom_point() +
  geom_text(data = ~filter(.x, year==as.character(latest_year)), aes(x=n_distinct(current_and_historical_deaths_by_cause_and_location$year)+0.2, y=proportion, label="*"), inherit.aes = FALSE) +  # cobble together the final year position from the data!
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
  scale_colour_viridis_d() +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage yearly deaths by place (%)",
    title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
    subtitle = paste0("\n* ",only_2024_text),
    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple()  # to render the html in the title
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

fig_cause_of_death_annual_proportion_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_annual_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 15, dpi = 300, units = "cm")
```

## Weekly proportion of deaths by place

This is the same idea as above, but looking at weekly proportions of cause of death by place - where did deaths due to different causes happen, and did their proportion vary over the year (e.g. seasonal effects)

```{r}
fig_cause_of_death_weekly_proportion_by_place <-
  merged_deaths_cause_of_death %>%
  filter(!cause_of_death %in% c("All")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other", "All"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  group_by(year, week_number, cause_of_death, place_of_death != "All") %>%
  mutate(proportion_weekly_total = number_of_deaths / sum(number_of_deaths)) %>%
  ungroup %>%
  filter(place_of_death != "All") %>%
  ggplot(aes(x = date_w_c, y = proportion_weekly_total, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  annotate(geom = "text", x=ymd("2021-01-14"), y=0.95, label = "2021", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-14"), y=0.95, label = "'22", hjust = 0, colour = "grey40") +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE, ncol = 2) +
  scale_colour_viridis_d() +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage weekly deaths by place (%)",
    # title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
    subtitle = paste0("\n* ",only_2024_text),
    caption = paste0(c("The 2015-2019 figures are average weekly deaths.",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple(),  # to render the html in the title
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

fig_cause_of_death_weekly_proportion_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
```

### With 2015-2019 context also

```{r}
(fig_weekly_proportion_of_death_by_place_with_2015_2019 <-
  bind_rows(
    ## 2020 onwards
    merged_deaths_cause_of_death %>%
      mutate(
        year = as.character(year),
        week_number_run_over = week_number_run_over + 53  # push 2020+ forward by a year, to 'squeeze' in the historic period year
        )
    ,
    ## 2015-2019 data, weekly averages make up one "year"
    merged_deaths_cause_of_death %>%
      filter(year==2020) %>%
      mutate(
        year = "2015-2019",
        number_of_deaths = deaths_mean
        )
  ) %>%
  mutate(year = factor(year, levels = c("2015-2019",as.character(2020:2022)))) %>%
  filter(!cause_of_death %in% c("All")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other", "All"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  group_by(year, week_number, cause_of_death, place_of_death != "All") %>%
  mutate(proportion_weekly_total = number_of_deaths / sum(number_of_deaths)) %>%
  ungroup %>%
  filter(place_of_death != "All") %>%
  ggplot(aes(x = week_number_run_over, y = proportion_weekly_total, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  # annotate(geom = "text", x=1+1, y=0.95, label = "2015-19", hjust = 0, colour = "grey40") +
  # annotate(geom = "text", x=54+1, y=0.95, label = "2020", hjust = 0, colour = "grey40") +
  # annotate(geom = "text", x=107+1, y=0.95, label = "2021", hjust = 0, colour = "grey40") +
  # annotate(geom = "text", x=159+1, y=0.95, label = "'22", hjust = 0, colour = "grey40") +
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE, ncol = 2) +
  scale_colour_viridis_d() +
  scale_x_continuous(breaks = c(1,54,107,159), labels = c("2015-2019",as.numeric(2020:2022))) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage weekly deaths by place (%)",
    subtitle = paste0("\n* ",only_2024_text),
    caption = paste0(c("The 2015-2019 figures are average weekly deaths",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple(),  # to render the html in the title
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)
fig_weekly_proportion_of_death_by_place_with_2015_2019 %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place_2015_2019", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
```

## Weekly deaths by cause, at home

```{r}
annotations_for_2021_and_2022_in_home_deaths_by_cause <-
  merged_deaths_cause_of_death %>%
  filter(str_starts(place_of_death,"Home")) %>%
  filter(cause_of_death != "All") %>%
  group_by(cause_of_death) %>%
  summarise(max = max(number_of_deaths), .groups = "drop")

fig_weekly_deaths_by_cause_at_home <-
  merged_deaths_cause_of_death %>%
  filter(str_starts(place_of_death,"Home")) %>%
  filter(cause_of_death != "All") %>%
  mutate(deaths_mean = if_else(historic_period=="2015-2019" & cause_of_death=="COVID-19", NA_real_, deaths_mean)) %>%  # set historic covid deaths to NA for 2020-2021
  pivot_longer(cols = c(deaths_mean, number_of_deaths), names_to = "deaths_type", values_to = "number_of_deaths") %>%
  mutate(deaths_type = if_else(deaths_type == "deaths_mean", true = "Historic mean 2015-2019", false = "Deaths in 2020 & 2021")) %>%
  filter(!(deaths_type=="deaths_mean" & cause_of_death=="COVID-19")) %>%  # remove historical average for covid, it's 0 anyway
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  ggplot(aes(x = date_w_c)) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  # geom_line(aes(y = number_of_deaths, colour = deaths_type, linetype = deaths_type)) +
  geom_line(aes(y = number_of_deaths, colour = deaths_type)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dotted", colour="grey70") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dotted", colour="grey70") +  # start of 2022
  # scale_linetype_manual(values = c("solid", "dashed")) +
  scale_colour_manual(values = c("grey10","grey50")) +
  geom_text(data = annotations_for_2021_and_2022_in_home_deaths_by_cause, aes(y = max), x = ymd("2021-01-20"), label = "2021", hjust = 0) +
  geom_text(data = annotations_for_2021_and_2022_in_home_deaths_by_cause, aes(y = max), x = ymd("2022-01-20"), label = "'22", hjust = 0) +
  facet_wrap(cause_of_death~., scales = "free_y") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = "Date w/c",
       y = "Weekly Deaths (N)",
       subtitle = paste0(c("Deaths registered at home", date_range_text), collapse = "\n"),
       caption = paste0(c("Note that y-axis scale varies across panels!",death_registrations_text,source_nrs_text), collapse="\n")
       ) +
  NULL

fig_weekly_deaths_by_cause_at_home %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_deaths_by_cause_at_home", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 18, dpi = 300, units = "cm")
```

# Deviation from mean

```{r}
fig_place_of_death_deviation_from_mean <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>%
  mutate(  # subtract mean from all figures to flatten graph
    deaths = deaths - deaths_mean,
    deaths_min = deaths_min - deaths_mean,
    deaths_max = deaths_max - deaths_mean
  ) %>%
  ggplot(aes(x = date_w_c)) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "#4477aa"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = 0, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_linetype_manual(values = c("dashed"), labels = "Weekly average (scaled to 0)") +
  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (difference from weekly average)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n")
       ) +
  annotate(geom = "text", x=ymd("2021-01-28"), y=-250, label = "2021", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=-250, label = "'22", hjust = 0, colour = "grey40") +
  NULL

fig_place_of_death_deviation_from_mean %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_place_of_death_deviation_from_mean", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
```

# Home deaths by geography over time: 2020-2022, 4-week average

## Home deaths by HB, 4-week average

```{r}
(fig_home_deaths_over_time_by_hb_4_week_average <-
  merged_deaths_hb %>%
  filter(place_of_death == "Home & other non-institution") %>%
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  pivot_longer(cols = c(ma4w_deaths_all_causes, ma4w_deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max, fill = "#4477aa"), alpha = 0.5) +
  # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), linewidth = 0.2) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  # scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~ref_area, scales = "free_y") +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_linetype_manual(values = c("dashed"), labels = "Weekly average") +
  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  labs(x = NULL,
       y = "Deaths (average of past 4 weeks)",
       title = "Deaths at Home*: 4-week average by NHS Health Board",
       subtitle = date_range_text,
       caption = paste0(c(
         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text,death_registrations_text,source_nrs_text
         ), collapse="\n")
       ) +
  ## annotate historic range
  annotate(geom = "text", x=ymd("2021-01-07"), y=0, label = "2021", hjust = 0, vjust = -1, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=0, label = "'22", hjust = 0, vjust = -1, colour = "grey40") +
  NULL
 )

fig_home_deaths_over_time_by_hb_4_week_average %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 20, dpi = 300, units = "cm")
```

## Home deaths by LA, 4-week average

```{r}
(fig_home_deaths_over_time_by_la_4_week_average <-
  merged_deaths_la %>%
  filter(place_of_death == "Home & other non-institution") %>%
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  pivot_longer(cols = c(ma4w_deaths_all_causes, ma4w_deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
  geom_ribbon(aes(ymin = ma4w_deaths_min, ymax = ma4w_deaths_max, fill = "#4477aa"), alpha = 0.5) +
  # geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = ma4w_deaths_mean, linetype = "dashed"), linewidth = 0.2) +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  # scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~ref_area, scales = "free_y") +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_linetype_manual(values = c("dashed"), labels = "Weekly average") +
  scale_fill_manual(values = c("#4477aa"), labels = "Historic weekly minima/maxima") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  labs(x = NULL,
       y = "Deaths (average of past 4 weeks)",
       title = "Deaths at Home*: 4-week average by Scottish Local Authority",
       subtitle = date_range_text,
       caption = paste0(c(
         "*Home or other non-institutional setting", vertical_line_text, vertical_scale_differs_between_panels_text,death_registrations_text,source_nrs_text
         ), collapse="\n")
       ) +
  ## annotate historic range
  annotate(geom = "text", x=ymd("2021-01-07"), y=0, label = "2021", hjust = 0, vjust = -1, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-07"), y=0, label = "'22", hjust = 0, vjust = -1, colour = "grey40") +
  NULL
 )

fig_home_deaths_over_time_by_la_4_week_average %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_home_deaths_4week_average_over_time_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 40, height = 20, dpi = 300, units = "cm")
```

# Geographical variations

## Proportion of deaths by place, by HB

```{r}
(fig_proportion_of_deaths_by_place_by_hb <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "hb") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point(aes(size = place_of_death)) +
  geom_line(aes(linewidth = place_of_death)) +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80"),
    axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
  guides(colour = guide_legend(nrow=2, title = NULL), 
         size = guide_legend(nrow=2, title = NULL)) +
  labs(
    x = NULL, y= "Proportion of yearly deaths (%)",
    subtitle = paste0("* ",only_2024_disaggregated_text),
    caption = paste0(c("Health boards ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  NULL)

fig_proportion_of_deaths_by_place_by_hb %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_by_place_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
```

## Proportion of deaths by place, by LA

```{r}
(fig_proportion_of_deaths_by_place_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point(aes(size = place_of_death)) +
  geom_line(aes(linewidth = place_of_death)) +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80"),
    axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
  guides(colour = guide_legend(nrow=2, title = NULL), 
         size = guide_legend(nrow=2, title = NULL)) +
  labs(
    x = NULL, y= "Proportion of yearly deaths (%)",
    subtitle = paste0("* ",only_2024_disaggregated_text),
    caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  NULL)

fig_proportion_of_deaths_by_place_by_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_by_place_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
```

## Home deaths only, HB

```{r}
(fig_proportion_of_deaths_at_home_by_hb <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "hb") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
  geom_point() +
  geom_line() +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80"),
    axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
  labs(
    linetype = NULL,
    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
    subtitle = paste0("* ",only_2024_disaggregated_text),
    caption = paste0(c("Health boards ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  NULL)

fig_proportion_of_deaths_at_home_by_hb %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
```

## Home deaths only, LA

```{r}
(fig_proportion_of_deaths_at_home_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
  geom_point() +
  geom_line() +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80"),
    axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
  labs(
    linetype = NULL,
    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
    subtitle = paste0("* ",only_2024_disaggregated_text),
    caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  NULL)

fig_proportion_of_deaths_at_home_by_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
```

## Absolute num. home deaths by LA

```{r}
# TODO: remove theme_minimal when done testing
# TODO: add Scotland graph also
# TODO: check  if all PoD would work?

la_group <- tibble(  # group LAs into 4 groups of 8
  ref_area = factor(order_la_by_population),
  # la_group = rep(1:4, each = 8))
  la_group = rep(1:6, times = c(6,5,5,5,5,6))
  ) %>%
  group_by(la_group) %>%
  mutate(la_group_label = paste0(ref_area[1], " - ", ref_area[n()])) %>%
  ungroup %>%
  mutate(la_group_label = factor(la_group_label, levels = unique(la_group_label)))

(fig_number_of_deaths_by_place_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(la_group, by = "ref_area") %>%  # group LAs
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  ggplot(. , aes(x=year, y=number_of_deaths, colour=ref_area, shape = ref_area)) +
    theme_minimal() +
    geom_point(size = rel(2)) +
    geom_line() +
    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=number_of_deaths, label="*"), inherit.aes = FALSE) +
    facet_rep_wrap(~ la_group_label, repeat.tick.labels = TRUE, scales = "free_y") +
    scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
    scale_y_continuous(label = scales::comma) +
    scale_shape_manual(values = map(c(6,5,5,5,5,6), scales::shape_pal()) %>% unlist) +
    scale_colour_manual(values = map(c(6,5,5,5,5,6), viridis::viridis_pal()) %>% unlist) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 12),
      plot.caption = element_text(size = 10, colour = "gray60"),
      panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
      strip.background = element_rect(fill = NA, colour = "gray80"),
      axis.text.x = element_text(angle = 45, hjust = 1)
      ) +
    guides(
      colour = guide_legend(ncol = 6, title = NULL), 
      shape = guide_legend(ncol = 6, title = NULL)
      ) +
    labs(
      x = NULL, y = NULL,
      title = "Deaths at home by LA",
  subtitle = paste0("* ",only_2024_disaggregated_text),
  caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
      ) +
    NULL
)

fig_number_of_deaths_by_place_by_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_number_of_deaths_by_place_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")
```

# Map-based visualisations

## Re-merge weekly deaths with 2015-2019 data for 2022 for LA & HB

```{r}
## merge the 2015-2019 period historical weekly averages to 2022 data
merged_deaths_la_historic_period_2015_2019 <-
  bind_rows(
    merged_deaths_la %>% filter(year %in% 2020:2021),
    merged_deaths_la %>% filter(year == 2022) %>%
      select(-c(deaths_mean, deaths_min, deaths_max)) %>%
      mutate(historic_period = "2015-2019") %>%
      left_join(historical_weekly_average_la, by = c("week_number", "place_of_death", "ref_area", "historic_period"))
  ) %>%
  left_join(weekly_covid_cases_la, by = c("year","week_number","date_w_c","week_number_run_over","ref_area"))

merged_deaths_hb_historic_period_2015_2019 <-
  bind_rows(
    merged_deaths_hb %>% filter(year %in% 2020:2021),
    merged_deaths_hb %>% filter(year == 2022) %>%
      select(-c(deaths_mean, deaths_min, deaths_max)) %>%
      mutate(historic_period = "2015-2019") %>%
      left_join(historical_weekly_average_hb, by = c("week_number", "place_of_death", "ref_area", "historic_period"))
  ) %>%
  left_join(weekly_covid_cases_hb, by = c("year","week_number","date_w_c","week_number_run_over","ref_area"))
```

## Home deaths increase since 2019 by LA

For this we can take the total annual deaths in each LA and compare to historical figures, showing the ratio for each year so far; alternatively we can also average over 2020-2022, and compare that to the average!
Because we only have partial data for 2022, we need to employ the weekly comparison method -

```{r}
(fig_map_home_deaths_increase_from_2020_la <-
  merged_deaths_la_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority", width=40),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(only_2024_disaggregated_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

fig_map_home_deaths_increase_from_2020_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
```

## Home deaths increase since 2019 by HB

```{r}
(fig_map_home_deaths_increase_from_2020_hb <-
  merged_deaths_hb_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_hb_home_deaths-1, maximum_ratio_between_hb_and_hb_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board", width=40),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(only_2024_disaggregated_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

fig_map_home_deaths_increase_from_2020_hb %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_from_2020_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")
```

## Home deaths increase since 2019 by LA, by year

For this we can take the total annual deaths in each LA and compare to historical figures, showing the ratio for each year so far; alternatively we can also average over 2020-2022, and compare that to the average!
Because we only have partial data for 2022, we need to employ the weekly comparison method -

```{r}
(fig_map_home_deaths_increase_by_year_la <-
  merged_deaths_la_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(year, ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    ) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by Local Authority",
       caption = paste0(c(only_2024_disaggregated_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

fig_map_home_deaths_increase_by_year_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_by_year_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 32, height = 18, dpi = 300, units = "cm")
```

## Home deaths increase vs. covid cases by LA, all deaths since 2019

```{r}
(fig_map_covid_cases_from_2020_la <-
  weekly_covid_cases_la %>%
  group_by(ref_area) %>%
  summarise(
    ratio_covid_cases = max(crude_rate_positive_per100k),
    .groups = "drop"
  ) %>% 
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020, by Local Authority", width=40),
       subtitle = "",
       caption = paste0(c(only_2024_disaggregated_text,crude_rate_text,source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)

fig_map_covid_cases_from_2020_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_covid_cases_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")

## combine maps
(fig_map_home_deaths_increase_vs_covid_case_from_2020_la <-
  wrap_plots(
    fig_map_home_deaths_increase_from_2020_la +
      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
      ,
    fig_map_covid_cases_from_2020_la +
      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
  )
)

fig_map_home_deaths_increase_vs_covid_case_from_2020_la %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_vs_covid_case_from_2020_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 16, dpi = 300, units = "cm")
```

## Home deaths increase since 2019 by HB, by year

See above for how this was done for LA!

```{r}
(fig_map_home_deaths_increase_by_year_hb <-
  merged_deaths_hb_historic_period_2015_2019 %>%
  filter(historic_period == "2015-2019") %>%
  group_by(year, ref_area, place_of_death) %>%
  summarise(
    ratio_deaths_to_historic = sum(deaths_all_causes, na.rm=TRUE) / sum(deaths_mean, na.rm=TRUE),
    percent_increase = ratio_deaths_to_historic - 1,
    .groups = "drop"
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(limits = c(0, 0.6), labels = function(x) scales::percent(x, accuracy = 1), direction = -1) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    aspect.ratio = 1.4
    ) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = "Increased home deaths since the start of 2020 relative to historical mean 2015-2019, by NHS Health Board",
       caption = paste0(c(only_2024_disaggregated_text,source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease relative\nto 2015-2019"
       ) +
  NULL)

fig_map_home_deaths_increase_by_year_hb %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_by_year_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 32, height = 18, dpi = 300, units = "cm")
```

## Home deaths increase vs. covid cases by HB, all deaths since 2019

```{r}
(fig_map_covid_cases_from_2020_hb <-
  weekly_covid_cases_hb %>%
  group_by(ref_area) %>%
  summarise(
    ratio_covid_cases = max(crude_rate_positive_per100k),
    .groups = "drop"
  ) %>% 
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(),
    aspect.ratio = 1.4
    ) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020, by NHS Health Board", width=40),
       subtitle = "",
       caption = paste0(c(only_2024_disaggregated_text,crude_rate_text,source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)

fig_map_covid_cases_from_2020_hb %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_covid_cases_from_2020_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 14, dpi = 300, units = "cm")

## combine maps
(fig_map_home_deaths_increase_vs_covid_case_from_2020_hb <-
  wrap_plots(
    fig_map_home_deaths_increase_from_2020_hb +
      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
      ,
    fig_map_covid_cases_from_2020_hb +
      theme(plot.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,0,0))
  )
)

fig_map_home_deaths_increase_vs_covid_case_from_2020_hb %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_map_home_deaths_increase_vs_covid_case_from_2020_hb", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 24, height = 16, dpi = 300, units = "cm")
```

### LA covid rates over time

I didn't finish this plot because the scale goes from 500 per 100k to 25k per 100k, which is a pain to depict in a meaningful way. Mayeb with a separate scale for each plot?

```{r}
## the problem with this plot is that there were between 500 per 100k cases to 25k per 100k cases over time and over LAs - how do we visualise that?
(fig_map_covid_cases_by_year_la <-
  weekly_covid_cases_la %>%
  left_join(population_estimate_la_2019 %>% rename(population_estimate = all_ages, year_estimate = year), by = c("ref_area"="la")) %>%
  group_by(ref_area, year, population_estimate) %>%
  summarise(
    cumulative_cases = sum(weekly_positive, na.rm=TRUE),
    ratio_covid_cases = cumulative_cases / population_estimate * 1e5,
    .groups = "drop"
  ) %>%
  distinct %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=ratio_covid_cases), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(labels = function(x) format(x, big.mark=","), direction = -1) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~year, nrow = 1) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases since the start of 2020 per 100,000 population, by Local Authority", width=50),
       subtitle = "",
       caption = paste0(c(source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL)
```

# Cumulative deaths at home over time

For this visualisation, we'll compare home deaths over time (e.g. years 2020, 2021, and 2022) with their historic period annual average, at HB-level. For this, we need to calculate the 2022 rate to be comparable to the complete data for years 2020 & 2021 - this is achieved by taking the sum of deaths in 2022 and dividing by the sum of the weekly average deaths in its historic period

```{r}
#TODO: continue working on this visualisation
#TODO: figure out some interesting time points to visualise
#TODO: maybe this could be done as a shiny widget? with a date slider and a selection of visualisations: map of covid cases, 
merged_deaths_hb %>%
  # filter(week_number %in% 1:4) %>%
  group_by(year, ref_area, place_of_death, historic_period) %>%
  summarise(
    deaths_all_causes = sum(deaths_all_causes, na.rm = TRUE),
    deaths_mean = sum(deaths_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ratio_annual_deaths_to_historical_mean = deaths_all_causes/deaths_mean,
    percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1
    ) %>%
  filter(
    place_of_death == "Home & other non-institution"
  ) %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  # scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_binned(
    labels = function(x) scales::percent(x, accuracy = 1), 
    type = "viridis",
    n.breaks = 7
    ) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(), 
    aspect.ratio = 1.4,  # needed for Scotland HB plot
    plot.margin = margin(5,5,5,5),
    # legend.position = c(1.2,0.4)  # this was manually tweaked until just right  # problem - edge gets cut
    ) +
  facet_grid(~year) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths in 2020 relative to 5-year historical mean, by Health Board", width=50),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(five_year_historical_mean,death_registrations_text,source_nrs_text), collapse="\n"),
       # fill = "Ratio"
       fill = "Home death\nincrease"
       ) +
  NULL
```

# Years 'skipped ahead' due to Covid

There was a pre-pandemic trend of increasing deaths at home - by plotting a regression line we can compute the 'years skipped ahead' equivalent to the increase in deaths at home due to covid

```{r}
regression_data <- 
  merged_proportions_of_deaths_by_place %>% 
  filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>% 
  mutate(years_since_2015 = 0:(n()-1))

regression_model <- lm(data = regression_data, formula = number_of_deaths ~ years_since_2015)

predicted_data <- tibble(years_since_2015 = seq(from=0, to=30, by = 0.1)) %>%
  mutate(year = 2015 + years_since_2015,
         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015), type = "response")
         )

## 2022 only has X weeks' worth of data available; we can extrapolate the annual
## total by computing the percentage of annual data represented by previous
## years at X weeks, and then averaging across those, or taking the min & max
extrapolated_2020_data <-
  historical_weekly_average_overall %>%
  filter(place_of_death == "Home & other non-institution" & historic_period == "2015-2019") %>%
  summarise(
    deaths_mean_so_far = sum(deaths_mean[week_number <= weeks_available_2024], na.rm= TRUE),
    deaths_mean_total = sum(deaths_mean, na.rm= TRUE),
    proportion_of_deaths_so_far = deaths_mean_so_far / deaths_mean_total
    )

actual_and_predicted_data <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
      mutate(data_type = "Historic")
    ,
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
    mutate(
      data_type = "Predicted",
      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_linear_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
      )
    ,
    merged_proportions_of_deaths_by_place %>%
      select(year, number_of_deaths, place_of_death) %>%
      filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
      mutate(
        number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
        year = 2015 + compute_x_from_y_using_linear_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
        data_type = "Actual",  # note: this is actual figures, but at the extrapolated year
        actual_year = 2020:current_year
      )
  ) %>%
  mutate(data_type = factor(data_type, levels = unique(data_type)))

## define the limites of the rectangles used to highlight areas of the graph
regression_rectangles <-
  actual_and_predicted_data %>% 
  filter(data_type %in% c("Predicted","Actual")) %>%
  group_by(data_type) %>%
  summarise(year_min = min(year), year_max = max(year), deaths_min = min(number_of_deaths), deaths_max = max(number_of_deaths))

## annotations to go above the rectangles
predicted_data_annotations <- tribble(
  ~x, ~y, ~label,
  rowMeans(regression_rectangles[1,c("year_min","year_max")]), regression_rectangles[[1,"deaths_max"]]+2000, glue::glue("Predicted home deaths 2020-{current_year} using 2015-2019 data"),
  rowMeans(regression_rectangles[2,c("year_min","year_max")]), regression_rectangles[[2,"deaths_max"]]+2000, glue::glue("Actual home deaths 2020-{current_year}*"),
  rowMeans(regression_rectangles[2,c("year_min","year_max")]), regression_rectangles[[2,"deaths_min"]]-3000, "If the 2015-2019 trend continued we wouldn't observe this number of home deaths until 2031",
)

fig_expected_home_deaths_regression <-
  ggplot(data = actual_and_predicted_data, aes(x=year, y=number_of_deaths)) +
  # highlighting predicted data for 2020 onwards
  statebins:::geom_rrect(aes(xmin = regression_rectangles[[1,"year_min"]]-0.5, xmax = regression_rectangles[[1,"year_max"]]+0.5, ymin = regression_rectangles[[1,"deaths_min"]]-1000, ymax = regression_rectangles[[1,"deaths_max"]]+1000), fill = alpha("grey",0), colour = "red") +
  # highlighting actual data
  statebins:::geom_rrect(aes(xmin = regression_rectangles[[2,"year_min"]]-0.5, xmax = regression_rectangles[[2,"year_max"]]+0.5, ymin = regression_rectangles[[2,"deaths_min"]]-1000, ymax = regression_rectangles[[2,"deaths_max"]]+1000), fill = alpha("grey",0), colour = "orange") +
  geom_text(data = predicted_data_annotations, aes(x=x,y=y,label=str_wrap(label, width = 60))) +
  geom_text(data = actual_and_predicted_data %>% filter(actual_year %in% 2020:current_year), aes(x = year, y = number_of_deaths + 500, label = if_else(actual_year==current_year, paste0(actual_year,"*"), as.character(actual_year))), angle = 30) +
  geom_point(aes(shape = data_type)) +
  geom_line(data = predicted_data %>% filter(year <= ceiling(max(regression_rectangles$year_max))), aes(x = year, y = predicted), linetype = "dotted") +
  theme_minimal() +
  theme(legend.position = "top", legend.title = element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = 2015:2045, minor_breaks = NULL, limits = c(2015,ceiling(max(regression_rectangles$year_max)))) +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 45, hjust = 1)) +
  labs(y = NULL, x = "Year (hypothetical)", title = "Increasing deaths at home", subtitle = str_wrap("Home deaths were on the increase before the pandemic.\nIf we assumed the same annual increases from 2015-2019 we wouldn't expect the pandemic levels of home deaths until approx. 2031!", width = 90), caption = paste0(c(glue::glue("*Annual home deaths for 2024 extrapolated from data for weeks 1-{weeks_available_2024}"),source_nrs_text), collapse="\n"))

fig_expected_home_deaths_regression %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_expected_home_deaths_regression", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 18, height = 12, dpi = 300, units = "cm")
```

### 2024-02-02 covid_nrs_place_of_death_data_2024.Rmd
cursory update to change 2023 to 2024

--- 2023-08-08 covid_nrs_place_of_death_data_2023.Rmd
@@
 ---
-title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data"
+title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2024 data"
 author: "Jan Savinc"
@@
 ## constants for use in graphs & annotations
-weeks_available_2023 <- max(merged_deaths_overall$week_number[merged_deaths_overall$year==2023])
-most_recent_date_available_2023 <- max(merged_deaths_overall$date_w_c)
+weeks_available_2024 <- max(merged_deaths_overall$week_number[merged_deaths_overall$year==2024])
+most_recent_date_available_2024 <- max(merged_deaths_overall$date_w_c)
 
 earliest_date <- min(merged_deaths_overall$date_w_c)
-current_year <- isoyear(most_recent_date_available_2023)
-
-weeks_available_2023_disaggregated <- max(merged_deaths_age$week_number[merged_deaths_age$year==2023])
-most_recent_date_available_2023_disaggregated <- max(merged_deaths_age$date_w_c)
+current_year <- isoyear(most_recent_date_available_2024)
+
+weeks_available_2024_disaggregated <- max(merged_deaths_age$week_number[merged_deaths_age$year==2024])
+most_recent_date_available_2024_disaggregated <- max(merged_deaths_age$date_w_c)
 ```
@@
       summarise(
-        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2023}, 2022"),
+        period = glue::glue("Data from week 1, 2020 to week {weeks_available_2024}, 2022"),
         historic_period = "2015-2019",
@@
     ),
-    ## 2023 data so far
+    ## 2024 data so far
     merged_deaths_overall %>%
@@
           sex == "all" &
-          year == 2023
+          year == 2024
       ) %>%
@@
     summarise(
-      period = glue::glue("2023, up to week {weeks_available_2023}"),
+      period = glue::glue("2024, up to week {weeks_available_2024}"),
       historic_period = "2015-2019",

### 2024-02-02 download_latest_data_and_render_reports.R
updated pipeline to use 2024 data, currently results in error running the data processing script

@@ -20,8 +20,35 @@ library(xml2)  # for the url_absolute() function
 
 # Find latest NRS weekly deaths spreadsheet URL ---------------------------
 
+# Note: as of 2024-02-02, the filename appears to change weekly again, so I'm
+# back to using a script to find the latest filename
+
+url_nrs_weekly_deaths <- "https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-deaths-registered-in-scotland"
+
+all_nrs_links <-
+  read_html(url_nrs_weekly_deaths) %>%
+  html_nodes("a") %>%
+  html_attr("href") %>%  # find all links on the page
+  tibble(links = .)
+
+nrs_filename_pattern <- "weekly-deaths-.*\\.xlsx"
+
+weekly_deaths_url_relative <-
+  all_nrs_links %>%
+  filter(str_detect(links, pattern = nrs_filename_pattern)) %>%
+  filter(!str_detect(links, pattern = "monthly")) %>%  # remove monthly data
+  pull(links)
+
+stopifnot(length(weekly_deaths_url_relative)==1)  # stop if more than one URL found
+
+## get absolute URL
+weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
+
+# 2023 notes below:
+
 ## fixed url now and filename no longer updated weekly, so I need to rename the file myself after it's downloaded
-url_nrs_weekly_deaths_2023 <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/weekly-deaths/weekly-deaths-23.xlsx"
+
+# url_nrs_weekly_deaths_2023 <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/weekly-deaths/weekly-deaths-23.xlsx"
 
 ## Note: previously the filename for the most recent spreadsheet reporting
 ## weekly deaths changed weekly and reflected the date of publication (week
@@ -81,7 +108,7 @@ if (file.exists(file_data_sources)) {
     short_name = c("main_report","sex_age","la","hb"),
     url = c(
       # weekly_deaths_url_absolute,  # found automatically above!
-      url_nrs_weekly_deaths_2023,  # fixed URL from March 2023
+      weekly_deaths_url_absolute,  # inferred latest URL
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
[…]

### 2024-02-07 covid_nrs_place_of_death_visualisations_2024.Rmd
refactored adding vline and annotate to graphs over time to automatically infer from current year of data how many to add

@@
   ggplot(aes(x = date_w_c)) +
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2024), linetype = "dashed", colour="grey50") +  # start of 2024
+  add_vertical_lines_for_start_of_year(years = 2021:current_year) +
   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
@@
        subtitle = date_range_text,
-       caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
+       # caption = paste0(c(death_registrations_text,removed_other_category_text,source_nrs_text), collapse="\n"),
        colour = "Cause of death"
@@
   geom_curve(data = annotations_fig_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2024-01-01"), y=30, label = "'24", hjust = 0, colour = "grey40") +
+  add_annotations_for_start_of_year(years = 2021:current_year) +
   NULL
@@
   ggplot(aes(x = date_w_c)) +
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
-  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
-   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2023), linetype = "dashed", colour="grey50") +  # start of 2023
-   geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2024), linetype = "dashed", colour="grey50") +  # start of 2024
+  add_vertical_lines_for_start_of_year(years = 2021:current_year) +
   scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2024)) +
@@
   annotate(geom="curve",x = compute_start_date_from_week_number(week_number = 22, year_number = 2020), xend = compute_start_date_from_week_number(week_number = 25, year_number = 2020), y = 115, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
-  annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "'21", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2022-01-07"), y=30, label = "'22", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2023-01-02"), y=30, label = "'23", hjust = 0, colour = "grey40") +
-  annotate(geom = "text", x=ymd("2024-01-01"), y=30, label = "'24", hjust = 0, colour = "grey40") +
+  add_annotations_for_start_of_year(years = 2021:current_year) +
   NULL

### 2024-05-08 compile_medications_from_bnf.Rmd
added codes for injectable buscopan, accidentally included hyoscine hydrobromide (not buscopan) - but I dealt with this in the safe haven

@@
 
-One source for a list of all BNF codes is the Information Services Portal of the NHS Business Services Authority, https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126.
+One source for a list of all BNF codes is the [Information Services Portal of the NHS Business Services Authority] (https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126).
 This is available after putting in a captcha-like password, so I've downloaded the file manually.
@@
 
+## Buscopan/Hyoscine Butylbromide injections
+
+In conversation with Jo Bowden, Sarah Mills, Sarah Bowers, we discussed using injectable buscopan/hyoscine butylbromide as a marker of palliative care.
+
+```{r}
+bnf_buscopan <-
+  bnf %>%
+  filter(str_detect(
+    string = bnf_chemical_substance,
+    pattern = regex("hyoscine", ignore_case = TRUE)
+  )) %>% filter(str_detect(bnf_presentation, pattern = regex("inj", ignore_case = TRUE))) %>%
+  select(bnf_section, bnf_section_code, bnf_product, bnf_presentation, bnf_presentation_code)
+```
+
 ## Test against BNF to see if I get the same substances
@@
 write_csv(bnf_palliative, file = "./bnf/bnf_codes_palliative_anticipatory.csv")
+
+write_csv(bnf_buscopan, file = "./bnf/bnf_codes_buscopan.csv")
 ```

### 2024-05-08 covid_nrs_place_of_death_data_2024.Rmd
just a note for myself to update script for 2024 data
modified these to include finalised 2022 data as historical, and to download the latest 2024 dataset; I've deleted & commented out the bits for the disaggregated age & sex data (currently only the split by sex/age or PoD, but not both, is available)

@@
 ## install packages from github if not yet available!
-# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
+### remotes::install_github("datasciencescotland/opendatascot", force = TRUE)  # this is the old repo
+# devtools::install_github("jsphdms/opendatascot")
 # remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)
@@
 
-## Sex & age
-
-Note: the last sex & age split of the data provided by NRS was Week 9 of 2023, though the data can still be split using the 'overall' dataset from Open Data Scotland above!
-
-```{r}
-# weekly_deaths_sex_age <-
-#   weekly_deaths_overall %>%
-#   filter(age!="all" & place_of_death != "All")
-
-# TODO: delete the below line, used for testing
-# files_weekly_deaths$main_report <- "./downloaded_data/2023-03-22_weekly-deaths-23.xlsx"
-
-sheet_covid_deaths_sex_and_age_data <- "1"
-sheet_all_cause_deaths_sex_and_age_data <- "4"
-
-# TODO: the data have changed again, now the most recent file no longer includes 2022 and 2023, but only 2023, so the cell ranges below are inaccurate and result in errors!
-# 
-# cell_ranges_for_sex_and_age_data <- list(
-#   "All" = "A6:K110",
-#   "Female" = "A113:K217",
-#   "Male" = "A220:K324"
-# )
-cell_ranges_for_sex_and_age_data <- list(
-  "All" = "A6:K58",
-  "Female" = "A61:K113",
-  "Male" = "A116:K168"
-)
-
-# TODO: this still doesn't split data by PoD...
-weekly_deaths_sex_age <-
-  full_join(
-    x =
-      map_dfr(
-        .x = cell_ranges_for_sex_and_age_data,
-        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_covid_deaths_sex_and_age_data, range = .x), 
-        .id = "sex"
-      ) %>%
-      pivot_longer(
-        cols = `All ages`:`85+`,
-        names_to = "age",
-        values_to = "deaths_covid_related"
-      ) %>%
-      janitor::clean_names(),
-    y =
-      map_dfr(
-        .x = cell_ranges_for_sex_and_age_data,
-        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_all_cause_deaths_sex_and_age_data, range = .x), 
-        .id = "sex"
-      ) %>%
-      pivot_longer(
-        cols = `All ages`:`85+`,
-        names_to = "age",
-        values_to = "deaths_all_causes"
-      ) %>%
-      janitor::clean_names(),
-    by = c("sex", "registration_year", "week_number", "week_beginning", "age")
-  ) %>%
-  rename(year = registration_year, date_w_c = week_beginning) %>%
-  mutate(
-    year = as.integer(year),
-    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
-    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
-    deaths_non_covid = deaths_all_causes - deaths_covid_related  # compute non-covid from all-cause minus covid
-  ) %>%
-  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
-
-# TODO: this is obsolete now, remove!
-weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
-  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
-  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
-  parse_year_and_week_number_from_ad_hoc_data %>%
-  recode_cause_of_death() %>%
-  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
-  tidyr::complete(  # complete implicit missing entries and make them 0
-    nesting(week_number,week_number_run_over,date_w_c,year),
-    age,sex,place_of_death,cause_of_death,
-    fill = list(number_of_deaths = 0)
-  ) %>%
-  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
-  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
-  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
-  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
-```
-
-### Sex & age separate splits
-
-```{r}
-## compute weekly data for sex and age separately
-## sum over sex or age, respectively
-weekly_deaths_sex <-
-  weekly_deaths_sex_age %>%
-  filter(age!="all") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
-  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
-
-weekly_deaths_age <-
-  weekly_deaths_sex_age %>%
-  filter(sex!="all") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
-  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
-```
-
 ## LA & HB
@@
 
-## Weekly Covid as contributory vs underlying cause of death
-
-```{r}
-# TODO: need to compute that from another source for the time being
-weekly_covid_contributory_vs_underlying_cause <-
-  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
-  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
-  parse_year_and_week_number_from_ad_hoc_data %>%
-  filter(
-    cause_of_death %in% c("COVID-19 contributory factor", "COVID-19 underlying cause")
-  ) %>%
-  tidyr::complete(  # complete implicit missing entries and make them 0
-    nesting(week_number,week_number_run_over,date_w_c,year),
-    age,sex,place_of_death,cause_of_death,
-    fill = list(number_of_deaths = 0)
-  ) %>%
-  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
-  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
-  clean_names() %>%
-  filter(age!="all") %>%
-  filter(sex!="all") %>%
-  filter(place_of_death != "All") %>%
-  group_by(week_number, week_number_run_over, year, date_w_c, place_of_death) %>%  # sum over age & sex
-  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
-  ({function(x)
-    bind_rows(
-      x,
-      x %>% 
-        group_by(week_number, week_number_run_over, year, date_w_c) %>%  # sum across place of death
-        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
-        mutate(place_of_death = "All")
-    )
-  })
-
-# TODO: this is currently commented out as it doesn't compute correctly
-## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure
-
-# stopifnot(
-#   (weekly_deaths_overall %>%
-#   filter(sex=="all" & age=="all") %>%
-#   inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
-#   filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
-#   nrow) == 0
-# )
-
-annual_covid_contributory_vs_underlying_cause <-
-  weekly_covid_contributory_vs_underlying_cause %>%
-  group_by(year, place_of_death) %>%
-  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
-  ({function(x)
-    bind_rows(
-      x %>% mutate(year = as.character(year)),
-      x %>% 
-        group_by(place_of_death) %>%  # sum across eyars
-        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
-        mutate(year = "2020-2021")
-    )
-  }) %>%
-  mutate(deaths_covid_related = covid_19_contributory_factor + covid_19_underlying_cause) %>%
-  mutate(
-    proportion_contributory = covid_19_contributory_factor / deaths_covid_related,
-    proportion_underlying = covid_19_underlying_cause / deaths_covid_related
-    )
-```
-
 ## 'Main report' file
@@
 ```{r}
+# TODO: delete the below line, used for testing
+files_weekly_deaths$main_report <- "../../../Downloads/weekly-deaths-24-week-17.xlsx"
+
 sheet_cause_of_death_data <- "8"
+
+## updated for 2024, this has all of 2023 plus incomplete 2024 data, with blanks up to week 52
+## note that there are fewer columns now because excess deaths have been moved to a separate spreadsheet
+# TODO: 2024 has 52 ISOweeks, but 2024-12-30 is W01 of 2025, or the 53rd week of 2024, so this may change
+cell_ranges_for_causes_of_death_data <- list(
+  "All" = "A6:J110",
+  "Care home" = "A113:J217",
+  "Home & other non-institution" = "A220:J324",
+  "Hospital" = "A327:J431",
+  "Other" = "A434:J538"
+)
 
@@
 ## ...and is correct again for 2023
-cell_ranges_for_causes_of_death_data <- list(
-  "All" = "A6:V58",
-  "Care home" = "A61:V113",
-  "Home & other non-institution" = "A116:V168",
-  "Hospital" = "A171:V223",
-  "Other" = "A226:V278"
-)
-
-# TODO: this will eventually only have 2023 data, change when that happens
-weekly_deaths_cause_of_death_2022_onwards <-
+# cell_ranges_for_causes_of_death_data <- list(
+#   "All" = "A6:V58",
+#   "Care home" = "A61:V113",
+#   "Home & other non-institution" = "A116:V168",
+#   "Hospital" = "A171:V223",
+#   "Other" = "A226:V278"
+# )
+
+weekly_deaths_cause_of_death_2023_onwards <-
   map_dfr(
@@
   janitor::clean_names() %>%
+  janitor::remove_empty(which = "cols") %>%
   rename(year = registration_year, date_w_c = week_beginning) %>%
@@
 
-### Data from 2020 & 2021
+### Data from previous years, 2020-2022
 
@@
 ```{r}
-# TODO: deal with 2022 data here, now that it's 'historical'; this will involve loading the last released data for 2022 released in the weekly releases for 2023, I assume somewhere mid-2023, like it was for 2021 data in 2022 releases
+# TODO: deal with 2022 data here, now that it's 'historical', I load the last known 2023 file that included 2022 data, 2023-07-19_weekly-deaths-23.xlsx
 cell_ranges_for_causes_of_death_data_2022 <- list(
@@
 
+weekly_deaths_cause_of_death_2022 <-
+  map_dfr(
+    .x = cell_ranges_for_causes_of_death_data_2022,
+    .f = ~readxl::read_excel(files_historical_deaths$`2022`, sheet = sheet_cause_of_death_data, range = .x), 
+    .id = "place_of_death"
+  ) %>%
+  janitor::clean_names() %>%
+  rename(year = registration_year, date_w_c = week_beginning) %>%
+  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
+  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
+  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
+  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
+  mutate(
+    year = as.integer(year),
+    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
+    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
+    cause_of_death = return_pretty_cause_of_death(cause_of_death),
+    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
+    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
+  ) %>%
+  select(-name) %>%
+  pivot_wider(names_from = measure_type, values_from = number_of_deaths) %>%
+  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
 ```
@@
   distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_sex_age <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_sex_age, 
-        current_data = weekly_deaths_sex_age,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, age, sex
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_age <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_age, 
-        current_data = weekly_deaths_age,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, age
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
-
-historical_weekly_average_sex <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_sex, 
-        current_data = weekly_deaths_sex,
-        year_range = years_in_historic_period
-      ) %>%
-      create_week_53_data_for_years_with_52_weeks() %>%
-      group_by(
-        week_number, place_of_death, sex
-      ) %>%
-      deaths_mean_min_max %>%
-      mutate(historic_period = historic_period)
-    }
-  ) %>% 
-  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+# 
+# historical_weekly_average_sex_age <-
+#   pmap_dfr(
+#     .l = historic_period_by_year,
+#     .f = function(year, historic_period, years_in_historic_period) {
+#       get_historic_and_current_data_from_year_range(
+#         historic_data = weekly_deaths_2015_2019_sex_age,
+#         current_data = weekly_deaths_sex_age,
+#         year_range = years_in_historic_period
+#       ) %>%
+#       create_week_53_data_for_years_with_52_weeks() %>%
+#       group_by(
+#         week_number, place_of_death, age, sex
+#       ) %>%
+#       deaths_mean_min_max %>%
+#       mutate(historic_period = historic_period)
+#     }
+#   ) %>%
+#   distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+# 
+# historical_weekly_average_age <-
+#   pmap_dfr(
+#     .l = historic_period_by_year,
+#     .f = function(year, historic_period, years_in_historic_period) {
+#       get_historic_and_current_data_from_year_range(
+#         historic_data = weekly_deaths_2015_2019_age, 
+#         current_data = weekly_deaths_age,
+#         year_range = years_in_historic_period
+#       ) %>%
+#       create_week_53_data_for_years_with_52_weeks() %>%
+#       group_by(
+#         week_number, place_of_death, age
+#       ) %>%
+#       deaths_mean_min_max %>%
+#       mutate(historic_period = historic_period)
+#     }
+#   ) %>% 
+#   distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
+# 
+# historical_weekly_average_sex <-
+#   pmap_dfr(
+#     .l = historic_period_by_year,
+#     .f = function(year, historic_period, years_in_historic_period) {
+#       get_historic_and_current_data_from_year_range(
+#         historic_data = weekly_deaths_2015_2019_sex, 
+#         current_data = weekly_deaths_sex,
+#         year_range = years_in_historic_period
+#       ) %>%
+#       create_week_53_data_for_years_with_52_weeks() %>%
+#       group_by(
+#         week_number, place_of_death, sex
+#       ) %>%
+#       deaths_mean_min_max %>%
+#       mutate(historic_period = historic_period)
+#     }
+#   ) %>% 
+#   distinct  # the 2020 and 2021 data get identical historical periodfs & values, so I remove duplicates
 ```
@@
 
-historical_annual_average_sex_age <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_sex_age, 
-        current_data = weekly_deaths_sex_age, 
-        year_range = years_in_historic_period
-      ) %>%
-        mutate(historic_period = historic_period)
-    }
-  ) %>%
-  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
-  compute_historical_annual_average_deaths(sex, age)
-
-historical_annual_average_sex <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_sex, 
-        current_data = weekly_deaths_sex, 
-        year_range = years_in_historic_period
-      ) %>%
-        mutate(historic_period = historic_period)
-    }
-  ) %>%
-  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
-  compute_historical_annual_average_deaths(sex)
-
-historical_annual_average_age <-
-  pmap_dfr(
-    .l = historic_period_by_year,
-    .f = function(year, historic_period, years_in_historic_period) {
-      get_historic_and_current_data_from_year_range(
-        historic_data = weekly_deaths_2015_2019_age, 
-        current_data = weekly_deaths_age, 
-        year_range = years_in_historic_period
-      ) %>%
-        mutate(historic_period = historic_period)
-    }
-  ) %>%
-  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
-  compute_historical_annual_average_deaths(age)
+# historical_annual_average_sex_age <-
+#   pmap_dfr(
+#     .l = historic_period_by_year,
+#     .f = function(year, historic_period, years_in_historic_period) {
+#       get_historic_and_current_data_from_year_range(
+#         historic_data = weekly_deaths_2015_2019_sex_age, 
+#         current_data = weekly_deaths_sex_age, 
+#         year_range = years_in_historic_period
+#       ) %>%
+#         mutate(historic_period = historic_period)
+#     }
+#   ) %>%
+#   distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+#   compute_historical_annual_average_deaths(sex, age)
+# 
+# historical_annual_average_sex <-
+#   pmap_dfr(
+#     .l = historic_period_by_year,
+#     .f = function(year, historic_period, years_in_historic_period) {
+#       get_historic_and_current_data_from_year_range(
+#         historic_data = weekly_deaths_2015_2019_sex, 
+#         current_data = weekly_deaths_sex, 
+#         year_range = years_in_historic_period
+#       ) %>%
+#         mutate(historic_period = historic_period)
+#     }
+#   ) %>%
+#   distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+#   compute_historical_annual_average_deaths(sex)
+# 
+# historical_annual_average_age <-
+#   pmap_dfr(
+#     .l = historic_period_by_year,
+#     .f = function(year, historic_period, years_in_historic_period) {
+#       get_historic_and_current_data_from_year_range(
+#         historic_data = weekly_deaths_2015_2019_age, 
+#         current_data = weekly_deaths_age, 
+#         year_range = years_in_historic_period
+#       ) %>%
+#         mutate(historic_period = historic_period)
+#     }
+#   ) %>%
+#   distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
+#   compute_historical_annual_average_deaths(age)
 ```
@@
 ## here I separately merge the datasets for cause of death data from different years
-## I treat covid deaths separately to recalculate a realistic esstimate of average deaths in prior years
+## I treat covid deaths separately to recalculate a realistic estimate of average deaths in prior years
 weekly_deaths_cause_of_death_noncovid <-
@@
     weekly_deaths_cause_of_death_2021,
-    weekly_deaths_cause_of_death_2022_onwards
+    weekly_deaths_cause_of_death_2022,
+    weekly_deaths_cause_of_death_2023_onwards
   ) %>%
@@
     weekly_deaths_cause_of_death_2021,
-    weekly_deaths_cause_of_death_2022_onwards
+    weekly_deaths_cause_of_death_2022,
+    weekly_deaths_cause_of_death_2023_onwards
   )
@@
   inner_join(historical_weekly_average_la, by = c("week_number","ref_area","place_of_death", "historic_period"))
-
-merged_deaths_sex_age <-
-  weekly_deaths_sex_age %>%  # use data with weekly movingincluded
-  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
-  inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age", "historic_period"))
-
-merged_deaths_sex <-
-  weekly_deaths_sex %>%
-  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
-  inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex", "historic_period"))
-
-merged_deaths_age <-
-  weekly_deaths_age %>%
-  mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
-  inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age", "historic_period"))
+# 
+# merged_deaths_sex_age <-
+#   weekly_deaths_sex_age %>%  # use data with weekly movingincluded
+#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+#   inner_join(historical_weekly_average_sex_age, by = c("place_of_death", "week_number", "sex", "age", "historic_period"))
+# 
+# merged_deaths_sex <-
+#   weekly_deaths_sex %>%
+#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+#   inner_join(historical_weekly_average_sex, by = c("place_of_death", "week_number", "sex", "historic_period"))
+# 
+# merged_deaths_age <-
+#   weekly_deaths_age %>%
+#   mutate(historic_period = historic_period_depending_on_year(year)) %>%  # add historic period so we match with the correct historic period data
+#   inner_join(historical_weekly_average_age, by = c("place_of_death", "week_number", "age", "historic_period"))
 
@@
 
-total_deaths_sex <-
-  merged_deaths_sex %>%
-  compute_total_deaths_in_year(sex) %>%
-  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
-  compute_ratio_of_annual_deaths_to_historical %>%
-  remove_invalid_historic_period_for_year()
-
-total_deaths_age <-
-  merged_deaths_age %>%
-  compute_total_deaths_in_year(age) %>%
-  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
-  compute_ratio_of_annual_deaths_to_historical %>%
-  remove_invalid_historic_period_for_year()
+# total_deaths_sex <-
+#   merged_deaths_sex %>%
+#   compute_total_deaths_in_year(sex) %>%
+#   left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
+#   compute_ratio_of_annual_deaths_to_historical %>%
+#   remove_invalid_historic_period_for_year()
+# 
+# total_deaths_age <-
+#   merged_deaths_age %>%
+#   compute_total_deaths_in_year(age) %>%
+#   left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
+#   compute_ratio_of_annual_deaths_to_historical %>%
+#   remove_invalid_historic_period_for_year()
 ```
@@
 
-weeks_available_2024_disaggregated <- max(merged_deaths_age$week_number[merged_deaths_age$year==2024])
-most_recent_date_available_2024_disaggregated <- max(merged_deaths_age$date_w_c)
+weeks_available_2024_disaggregated <- max(merged_deaths_cause_of_death$week_number[merged_deaths_cause_of_death$year==2024])
+most_recent_date_available_2024_disaggregated <- max(merged_deaths_cause_of_death$date_w_c)
 ```
@@
     num_weeks_to_average = 4)
-
-merged_deaths_sex <-
-  compute_moving_average_for_current_data(
-    current_data = merged_deaths_sex, 
-    historical_data = 
-      weekly_deaths_2015_2019_sex, 
-    num_weeks_to_average = 4)
-
-merged_deaths_age <-
-  compute_moving_average_for_current_data(
-    current_data = merged_deaths_age, 
-    historical_data = 
-      weekly_deaths_2015_2019_age, 
-    num_weeks_to_average = 4)
-
-merged_deaths_sex_age <-
-  compute_moving_average_for_current_data(
-    current_data = merged_deaths_sex_age, 
-    historical_data = 
-      weekly_deaths_2015_2019_sex_age, 
-    num_weeks_to_average = 4)
+# 
+# merged_deaths_sex <-
+#   compute_moving_average_for_current_data(
+#     current_data = merged_deaths_sex, 
+#     historical_data = 
+#       weekly_deaths_2015_2019_sex, 
+#     num_weeks_to_average = 4)
+
+# merged_deaths_age <-
+#   compute_moving_average_for_current_data(
+#     current_data = merged_deaths_age, 
+#     historical_data = 
+#       weekly_deaths_2015_2019_age, 
+#     num_weeks_to_average = 4)
+
+# merged_deaths_sex_age <-
+#   compute_moving_average_for_current_data(
+#     current_data = merged_deaths_sex_age, 
+#     historical_data = 
+#       weekly_deaths_2015_2019_sex_age, 
+#     num_weeks_to_average = 4)
 ```
@@
 
-# TODO: other thigns like excess death & historical rates by geography
-
-```{r}
-# TODO: copy some of the old calculations over
-```
-
 # Save most recent workspace for further analyses

### 2024-05-08 download_latest_data_and_render_reports.R

@@ -37,6 +37,7 @@ weekly_deaths_url_relative <-
   all_nrs_links %>%
   filter(str_detect(links, pattern = nrs_filename_pattern)) %>%
   filter(!str_detect(links, pattern = "monthly")) %>%  # remove monthly data
+  filter(!str_detect(links, pattern = "excess")) %>%  # remove excess deaths data
   pull(links)
 
 stopifnot(length(weekly_deaths_url_relative)==1)  # stop if more than one URL found
@@ -124,15 +125,15 @@ if (file.exists(file_historical_data_sources)) {
   table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
 } else {
   table_of_historical_data_sources <- tibble(
-    # short_name = c("overall","sex_age","la","hb","2020-2021"),
-    short_name = c("overall","sex_age","la","hb","2020","2021"),
+    short_name = c("overall","sex_age","la","hb","2020","2021","2022"),
     url = c(
       "https://www.nrscotland.gov.uk/files/statistics/covid19/weekly-deaths-by-location-2015-2019.csv",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-council-areas.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location-15-19.xlsx",
       "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-21-data-week-52.xlsx",  # last 2021 data which included 2020 also
-      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-22-data-week-25.xlsx"  # last 2022 data which included 2021 also
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-22-data-week-25.xlsx",  # last 2022 data which included 2021 also
+      "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-22-data-final.xlsx"  # finalised 2022 data
     )
   ) %>% mutate(
     # filename = basename(url),

### 2024-05-08 functions.R
helper functions for common plotting tasks

@@ -4,6 +4,7 @@
 compute_start_date_from_week_number <- function(week_number, year_number) {
   ## this assumes a valid week number! There are 52 or 53 weeks in a year
   isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
+  # isoweek_strings <- paste0(year_number,"-W",str_pad(week_number, width=2, pad='0'),"-1")
   # computed_dates <- ISOweek::ISOweek2date(isoweek_string)
   computed_dates <- rep(NA_Date_,times = length(week_number))
   computed_dates[which(!is.na(week_number))] <- ISOweek::ISOweek2date(isoweek_string[which(!is.na(week_number))])  # return NA instead of failing with ISOweek!
@@ -312,6 +313,33 @@ infer_year_range_for_calculating_covid_deaths <- function(year) {
   year_range <- seq.int(from = year-5L, to = year-1L)
   year_range <- year_range[year_range>=2020]
   return(year_range)
+}
+
+## helper function to create vertical lines in graph for start of each year
+## included
+add_vertical_lines_for_start_of_year <- function(years, linetype = "dashed", colour = "grey50") {
+  map(.x = years, .f = function(year) geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = year), linetype = linetype, colour = colour))
+}
+
+## helper function, same as above, but for annotations
+add_annotations_for_start_of_year <- function(years, y=30, colour="grey40", shorten_labels = TRUE) {
+  if (shorten_labels)
+    labels = paste0("'", str_sub(years, 3, 4))
+  else
+    labels = years
+  map2(
+    .x = years,
+    .y = labels,
+    .f = function(year, label)
+      annotate(
+        geom = "text",
+        x = compute_start_date_from_week_number(week_number = 1, year_number = year),
+        y = y,
+        label = label,
+        hjust = 0,
+        colour = colour
+      )
+  )
 }
 
 

### 2024-05-08 example_script_multistate_modelling.r
a few more things added to MSM example

@@ -250,6 +250,49 @@ dummy <- expand.grid(policy = c(TRUE,FALSE), sex = c("F","M"))
 predicted1 <- survfit(cfit1, newdata = dummy)  # from here on you can plot both the original fit and the predicted fit and compare!
 
 
+# example of same-day discharges ------------------------------------------
+
+# There is an issue with the admissions data: some people are discharged on the
+# same day as they are admitted. This would be less of an issue if we knew the
+# time as well as the date, as we could compute actual durations in that case.
+# In the absence of actual times, there are several solutions:
+
+# 3-day study period example
+dt_start <- ymd("2020-01-01")
+dt_end <- ymd("2020-01-03")
+
+transitions_3 <- 
+  bind_rows(
+    tibble(
+      id = 1L,
+      start = c(ymd("2020-01-01"),ymd("2020-01-02"),ymd("2020-01-03")),
+      end = c(c(ymd("2020-01-01"),ymd("2020-01-02"),ymd("2020-01-03")))
+    ),
+    tibble(
+      id = 2L,
+      start = c(ymd("2020-01-01"),ymd("2020-01-03")),
+      end = c(ymd("2020-01-03"),ymd("2020-01-03")),
+    )
+  ) %>%
+  ## using fake start and end times to generate 12-hour intervals
+  mutate(
+    start = ymd_h(paste0(start, " 06")),
+    end = ymd_h(paste0(end, " 18")),
+    interval = iv(start, end)
+  ) %>%
+  group_by(id) %>%
+  reframe(
+    interval = iv_set_difference(iv(ymd_hm("2020-01-01 06:00"), ymd_hm("2020-01-04 18:00")), iv(start, end)),
+    start_home = iv_start(interval),
+    end_home = iv_end(interval)
+  ) %>%
+  ## convert to time since start of study
+  mutate(across(c(start_home,end_home), ~(.x - ymd_hm("2020-01-01 06:00"))/ddays(1)))
+
+
+
+
+
 # working through mstate example ------------------------------------------
 
 

### 2024-07-10 just_in_case_bnf_medications.qmd
started compiling Sarah Mills's codes for palliative care meds, a few more left

---
title: "Just-in-case medications list"
author: "Sarah Mills & Jan Savinc"
format: html
---

# Introduction

Here I'll compile the BNF codes corresponding to medicines Sarah Mills compiled as being "Just in case" for use in palliative care/end of life care.

•       Strong opioids: Alfentanil, Buprenorphine, Diamorphine hydrochloride, Dihydrocodeine tartrate, Fentanyl, Hydromorphone hydrochloride, Meptazinol, Methadone hydrochloride, Morphine sulfate, Oxycodone hydrochloride, Tramadol hydrochloride
•       Anti-muscarinics: Aclidinium bromide, Alverine citrate, Bupropion hydrochloride, Minims atropine sulfate, Atropine sulfate, Minims cyclopentolate hydrochloride, Cyclopentolate hydrochloride, Dicycloverine hydrochloride, Erdosteine, Fesoterodine fumarate, Flavoxate hydrochloride, Glycopyrrolate bromide, Hyoscine, Procyclidine hydrochloride, Propantheline bromide, Propiverine hydrochloride
•       Anti-emetics: Betahistine hydrochloride, Cyclizine, Domperidone, Granisetron hydrochloride, Levomepromazine hydrochloride, Metoclopramide hydrochloride, Ondansetron hydrochloride, Prochlorperazine maleate
•       Anxiolytics: Buspirone hydrochloride, Chlordiazepoxide hydrochloride, Clobazam, Clomethiazole, Clonazepam, Diazepam, Loprazolam, Lorazepam, Lormetazepam, Midazolam, Nitrazepam, Temazepam, Trazodone hydrochloride, Zopiclone, Zolpidem tartrate

The strategy for these is to look at each medication, look at its formulations, and decide which to include.

# BNF source

One source for a list of all BNF codes is the [Information Services Portal of the NHS Business Services Authority] (https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126). This is available after putting in a captcha-like password, so I've downloaded the file manually.

File used: `01-01-2024 : 86`

# Setup

```{r}
library(tidyverse)
library(janitor)
```

# Compile codes for JIC medicines

## Read BNF codes

```{r}
bnf_zip_file <- "./bnf/20240701_1719843096151_BNF_Code_Information.zip"  # manually downloaded to this location
csv_file <- as.character(unzip(bnf_zip_file, list = TRUE)[1])  # get the filename from the zip
temp_dir <- tempdir()  # create temporary directory for unzipping to
unzip(zipfile = bnf_zip_file, exdir = temp_dir)  # extract csv file from zip

bnf <- read_csv(file.path(temp_dir, csv_file)) %>% clean_names()  # read csv file
```

## Encode the list of medicines

```{r}
strong_opioids <- "Alfentanil, Buprenorphine, Diamorphine hydrochloride, Dihydrocodeine tartrate, Fentanyl, Hydromorphone hydrochloride, Meptazinol, Methadone hydrochloride, Morphine sulfate, Oxycodone hydrochloride, Tramadol hydrochloride"

anti_muscarinics <- "Aclidinium bromide, Alverine citrate, Bupropion hydrochloride, Minims atropine sulfate, Atropine sulfate, Minims cyclopentolate hydrochloride, Cyclopentolate hydrochloride, Dicycloverine hydrochloride, Erdosteine, Fesoterodine fumarate, Flavoxate hydrochloride, Glycopyrrolate bromide, Hyoscine, Procyclidine hydrochloride, Propantheline bromide, Propiverine hydrochloride"

anti_emetics <- "Betahistine hydrochloride, Cyclizine, Domperidone, Granisetron hydrochloride, Levomepromazine hydrochloride, Metoclopramide hydrochloride, Ondansetron hydrochloride, Prochlorperazine maleate"

anxiolytics <- "Buspirone hydrochloride, Chlordiazepoxide hydrochloride, Clobazam, Clomethiazole, Clonazepam, Diazepam, Loprazolam, Lorazepam, Lormetazepam, Midazolam, Nitrazepam, Temazepam, Trazodone hydrochloride, Zopiclone, Zolpidem tartrate"

jic <- map2(
  .x = c(strong_opioids, anti_muscarinics, anti_emetics, anxiolytics),
  .y = c("strong opioids", "anti-muscarinics", "anti-emetics", "anxiolytics"),
  .f = function(meds, name) {
    tibble(
      group = name,
      meds = 
        str_split(string = meds, pattern = ",") %>%
        unlist() %>%
        str_trim() %>%
        tolower()
    )
  }
) %>% list_rbind()
```

## Check for non-matches in BNF

A small number of medications don't have exact matches in the BNF when searching by substance:

```{r}
nonmatches <- 
  jic %>%
  anti_join(
    bnf %>% mutate(bnf_chemical_substance = tolower(bnf_chemical_substance)), 
    by = join_by(meds == bnf_chemical_substance)
  )

nonmatches %>% knitr::kable(caption = "Medications that don't match the BNF entries exactly")
```

Note that even exact matches need to be carefully checked as not all formualtions are relevant (specifically combination medicines, for example).

## Adding medicines to the list

```{r}
## initialise the blank list of included medicines
jic_medicines <- tibble()

## helper function to subset the bnf list, finding a keyword in the product name
## or chemical substance
search_bnf <- function(keyword) {
  bnf %>%
    filter(
      str_detect(paste0(bnf_product, bnf_chemical_substance), regex(keyword, ignore_case = TRUE))
    )
}

## helper function to add medicines as I go along
add_entries <- function(meds_to_add, bnf_to_add) {
  bind_rows(
    jic_medicines,
    bind_cols(jic %>% filter(meds==meds_to_add), bnf_to_add)
    ) %>%
    distinct  # to precvent accidental duplication along the way
}
```

## strong opioids

### alfentanil

All formulations make sense to me. Query `Alfentanil 0.1% nasal spray`?

```{r}
# search_bnf(keyword="alfentanil") %>% View()

jic_medicines <- add_entries(meds = "alfentanil", bnf_to_add = search_bnf(keyword="alfentanil"))
```

### buprenorphine

Keep only the analgesics section (0407), not the opioid dependence section (0410).

```{r}
# search_bnf(keyword="buprenorphine") %>% View()

jic_medicines <- add_entries(
  meds = "buprenorphine", 
  bnf_to_add = search_bnf(keyword="buprenorphine") %>%
    filter(bnf_section_code == "0407")
  )
```

### diamorphine hydrochloride

Included both section cough suppressants and analgesics.

```{r}
# search_bnf(keyword="diamorphine hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "diamorphine hydrochloride", 
  bnf_to_add = search_bnf(keyword="diamorphine hydrochloride")
  )
```

### dihydrocodeine tartrate

```{r}
# search_bnf(keyword="dihydrocodeine tartrate") %>% View()

jic_medicines <- add_entries(
  meds = "dihydrocodeine tartrate", 
  bnf_to_add = search_bnf(keyword="dihydrocodeine tartrate")
  )
```

### fentanyl

Keep only the analgesics subsection (0407), not the general anaesthesia and local anaesthesia sections.

```{r}
# search_bnf(keyword="fentanyl") %>% View()

jic_medicines <- add_entries(
  meds = "fentanyl", 
  bnf_to_add = search_bnf(keyword="fentanyl") %>%
    filter(bnf_section_code == "0407")
  )
```

### hydromorphone hydrochloride

```{r}
# search_bnf(keyword="hydromorphone hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "hydromorphone hydrochloride", 
  bnf_to_add = search_bnf(keyword="hydromorphone hydrochloride")
  )
```

### meptazinol

```{r}
# search_bnf(keyword="meptazinol") %>% View()

jic_medicines <- add_entries(
  meds = "meptazinol", 
  bnf_to_add = search_bnf(keyword="meptazinol")
  )
```

### methadone hydrochloride

Kept the cough suppressants section (0309) and analgesics (0407), not drug dependence (0410).

```{r}
# search_bnf(keyword="methadone hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "methadone hydrochloride", 
  bnf_to_add = search_bnf(keyword="methadone hydrochloride") %>%
    filter(bnf_section_code %in% c("0309","0407"))
  )
```

### morphine sulfate

Query gel, intrasite gel?

```{r}
# search_bnf(keyword="morphine sulfate") %>% View()

jic_medicines <- add_entries(
  meds = "morphine sulfate", 
  bnf_to_add = search_bnf(keyword="morphine sulfate")
  )
```

### oxycodone hydrochloride

Query combination: Oxycodone hydrochloride/naloxone hydrochloride?

```{r}
# search_bnf(keyword="oxycodone hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "oxycodone hydrochloride", 
  bnf_to_add = search_bnf(keyword="oxycodone hydrochloride")
  )
```

### tramadol hydrochloride

```{r}
# search_bnf(keyword="tramadol hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "tramadol hydrochloride", 
  bnf_to_add = search_bnf(keyword="tramadol hydrochloride")
  )
```

## anti-muscarinics

### aclidinium bromide

Keep only paragraph code 030102 for antimuscarinic bronchodilators, not compound bronchodilator preparations 030104 (Aclidinium bromide/formoterol)

```{r}
# search_bnf(keyword="aclidinium bromide") %>% View()

jic_medicines <- add_entries(
  meds = "aclidinium bromide", 
  bnf_to_add = search_bnf(keyword="aclidinium bromide") %>%
    filter(bnf_paragraph_code == "030102")
  )
```

### alverine citrate

included combination Alverine citrate/simethicone, as well as Alverine citrate compound preparations (Sterculia 62% / Alverine 0.5% granules and Spasmonal Fibre granules)

```{r}
# search_bnf(keyword="alverine citrate") %>% View()

jic_medicines <- add_entries(
  meds = "alverine citrate", 
  bnf_to_add = search_bnf(keyword="alverine citrate")
  )
```

### bupropion hydrochloride

```{r}
# search_bnf(keyword="bupropion hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "bupropion hydrochloride", 
  bnf_to_add = search_bnf(keyword="bupropion hydrochloride")
  )
```

### minims atropine sulfate

```{r}
# search_bnf(keyword="minims atropine sulfate") %>% View()

jic_medicines <- add_entries(
  meds = "minims atropine sulfate", 
  bnf_to_add = search_bnf(keyword="minims atropine sulfate")
  )
```

### atropine sulfate

Query whether only eye drops, or oral/parenteral is also relevant here? Also "Co-phenotrope (Diphenox hydrochloride/atropine sulfate)" for acute diarrhea?

```{r}
# search_bnf(keyword="atropine sulfate") %>% View()

jic_medicines <- add_entries(
  meds = "atropine sulfate", 
  bnf_to_add = search_bnf(keyword="atropine sulfate") %>%
    # minims already included by an above entry
    filter(!str_detect(bnf_product, regex("minims", ignore_case=TRUE)))
  )
```

### minims cyclopentolate hydrochloride

```{r}
# search_bnf(keyword="minims cyclopentolate hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "minims cyclopentolate hydrochloride", 
  bnf_to_add = search_bnf(keyword="minims cyclopentolate hydrochloride")
  )
```

### cyclopentolate hydrochloride

```{r}
# search_bnf(keyword="cyclopentolate hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "cyclopentolate hydrochloride", 
  bnf_to_add = search_bnf(keyword="cyclopentolate hydrochloride") %>%
    # minims already included by an above entry
    filter(!str_detect(bnf_product, regex("minims", ignore_case=TRUE)))
  )
```

### dicycloverine hydrochloride

```{r}
# search_bnf(keyword="dicycloverine hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "dicycloverine hydrochloride", 
  bnf_to_add = search_bnf(keyword="dicycloverine hydrochloride")
  )
```

### erdosteine

```{r}
# search_bnf(keyword="erdosteine") %>% View()

jic_medicines <- add_entries(
  meds = "erdosteine", 
  bnf_to_add = search_bnf(keyword="erdosteine")
  )
```

### fesoterodine fumarate

```{r}
# search_bnf(keyword="fesoterodine fumarate") %>% View()

jic_medicines <- add_entries(
  meds = "fesoterodine fumarate", 
  bnf_to_add = search_bnf(keyword="fesoterodine fumarate")
  )
```

### flavoxate hydrochloride

```{r}
# search_bnf(keyword="flavoxate hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "flavoxate hydrochloride", 
  bnf_to_add = search_bnf(keyword="flavoxate hydrochloride")
  )
```

### glycopyrrolate bromide

Note: "glycopyrrolate" doesn't appear in the BNF, but "glycopyrronium" does.

Query: this appears under chapters: gastro-intestinal, skin, respiratory, as well as anaesthesia - which ones are relevant?

```{r}
# search_bnf(keyword="glycopyrrolate bromide") %>% View()
# search_bnf(keyword="glycopyrrolate") %>% View()
# search_bnf(keyword="glycopyrr.* bromide") %>% View()

jic_medicines <- add_entries(
  meds = "glycopyrrolate bromide", 
  bnf_to_add = search_bnf(keyword="glycopyrr.* bromide")
  )
```

### hyoscine

Query:this occurs in 2 forms, butylbromide (antispasmodic) and hydrobromide (nausea and vertigo) – are both relevant? And in all formualtions (toral, injectable, transdermal patch)

```{r}
# search_bnf(keyword="hyoscine") %>% View()

jic_medicines <- add_entries(
  meds = "hyoscine", 
  bnf_to_add = search_bnf(keyword="hyoscine")
  )
```

### procyclidine hydrochloride

```{r}
# search_bnf(keyword="procyclidine hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "procyclidine hydrochloride", 
  bnf_to_add = search_bnf(keyword="procyclidine hydrochloride")
  )
```

### propantheline bromide

```{r}
# search_bnf(keyword="propantheline bromide") %>% View()

jic_medicines <- add_entries(
  meds = "propantheline bromide", 
  bnf_to_add = search_bnf(keyword="propantheline bromide")
  )
```

### propiverine hydrochloride

```{r}
# search_bnf(keyword="propiverine hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "propiverine hydrochloride", 
  bnf_to_add = search_bnf(keyword="propiverine hydrochloride")
  )
```

## anti-emetics

### betahistine hydrochloride

```{r}
# search_bnf(keyword="betahistine hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "betahistine hydrochloride", 
  bnf_to_add = search_bnf(keyword="betahistine hydrochloride")
  )
```

### cyclizine

Keeping cyclizine hydrochloride and lactate forms, but not the combination "Morphine tartrate and cyclizine tartrate" in the Analgesics section (0407)

```{r}
# search_bnf(keyword="cyclizine") %>% View()

jic_medicines <- add_entries(
  meds = "cyclizine", 
  bnf_to_add = search_bnf(keyword="cyclizine") %>%
    filter(bnf_section_code != "0407")
  )
```

### domperidone

```{r}
# search_bnf(keyword="domperidone") %>% View()

jic_medicines <- add_entries(
  meds = "domperidone", 
  bnf_to_add = search_bnf(keyword="domperidone")
  )
```

### granisetron hydrochloride

```{r}
# search_bnf(keyword="granisetron hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "granisetron hydrochloride", 
  bnf_to_add = search_bnf(keyword="granisetron hydrochloride")
  )
```

### levomepromazine hydrochloride

```{r}
# search_bnf(keyword="levomepromazine hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "levomepromazine hydrochloride", 
  bnf_to_add = search_bnf(keyword="levomepromazine hydrochloride")
  )
```

### metoclopramide hydrochloride

```{r}
# search_bnf(keyword="metoclopramide hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "metoclopramide hydrochloride", 
  bnf_to_add = search_bnf(keyword="metoclopramide hydrochloride")
  )
```

### ondansetron hydrochloride

```{r}
# search_bnf(keyword="ondansetron hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "ondansetron hydrochloride", 
  bnf_to_add = search_bnf(keyword="ondansetron hydrochloride")
  )
```

### prochlorperazine maleate

```{r}
# search_bnf(keyword="prochlorperazine maleate") %>% View()

jic_medicines <- add_entries(
  meds = "prochlorperazine maleate", 
  bnf_to_add = search_bnf(keyword="prochlorperazine maleate")
  )
```

## anxiolytics

### buspirone hydrochloride

```{r}
# search_bnf(keyword="buspirone hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "buspirone hydrochloride", 
  bnf_to_add = search_bnf(keyword="buspirone hydrochloride")
  )
```

### chlordiazepoxide hydrochloride

```{r}
# search_bnf(keyword="chlordiazepoxide hydrochloride") %>% View()

jic_medicines <- add_entries(
  meds = "chlordiazepoxide hydrochloride", 
  bnf_to_add = search_bnf(keyword="chlordiazepoxide hydrochloride")
  )
```

### clobazam

```{r}
# TODO: continue adding these, only the anxiolyitics left!
```

# delete below, scratch area

```{r}

search_bnf(keyword="alfentanil") %>% View()
search_bnf(keyword="meptazinol") %>% View()
search_bnf(keyword="minims atropine sulfate") %>% View()
search_bnf(keyword="minims cyclopentolate hydrochloride") %>% View()
# search_bnf(keyword="glycopyrrolate") %>% View()  # glycopyrronium rather than glycopyrrolate
search_bnf(keyword="glycopyrronium") %>% View()  # lots, uprobably not all relevant
search_bnf(keyword="cyclizine") %>% View()  # nto all relevant
search_bnf(keyword="midazolam") %>% View()  # nto all relevant
search_bnf(keyword="loprazolam") %>% View()  # nto all relevant

## checking the ones that had matches...
search_bnf(keyword="meptazinol") %>% View()  # nto all relevant

search_bnf(keyword="alfentanil") %>% View()

jic_medicines <-
  bind_rows(
    jic_medicines,
    search_bnf(keyword="alfentanil")
  )
```

### 2024-07-11 just_in_case_bnf_medications.qmd
completed vetting the JIC medications list (v1); need to have it checked by a clinician next (v2)

@@
 
-# Introduction
+# Introduction {#sec-introduction}
 
@@
 The strategy for these is to look at each medication, look at its formulations, and decide which to include.
+
+Versions:
+
+- v1 is my first attempt at compiling the codes, without a clinician checking over the codes
 
@@
 
-Note that even exact matches need to be carefully checked as not all formualtions are relevant (specifically combination medicines, for example).
+Note that even exact matches need to be carefully checked as not all formulations are relevant (specifically combination medicines, for example).
 
@@
 
-Keep only the analgesics section (0407), not the opioid dependence section (0410).
+Keep only the analgesics section (0407), not the opioid dependence section (0410) - the latter uses buprenorphine hydrochloride and a buprenorphine hcl/naloxone combination.
 
@@
 
-Query:this occurs in 2 forms, butylbromide (antispasmodic) and hydrobromide (nausea and vertigo) – are both relevant? And in all formualtions (toral, injectable, transdermal patch)
+Query:this occurs in 2 forms, butylbromide (antispasmodic) and hydrobromide (nausea and vertigo) – are both relevant? And in all formulations (oral, injectable, transdermal patch)
 
@@
 ```{r}
-# TODO: continue adding these, only the anxiolyitics left!
+# search_bnf(keyword="clobazam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "clobazam", 
+  bnf_to_add = search_bnf(keyword="clobazam")
+  )
 ```
 
-# delete below, scratch area
+### clomethiazole
 
 ```{r}
+# search_bnf(keyword="clomethiazole") %>% View()
 
-search_bnf(keyword="alfentanil") %>% View()
-search_bnf(keyword="meptazinol") %>% View()
-search_bnf(keyword="minims atropine sulfate") %>% View()
-search_bnf(keyword="minims cyclopentolate hydrochloride") %>% View()
-# search_bnf(keyword="glycopyrrolate") %>% View()  # glycopyrronium rather than glycopyrrolate
-search_bnf(keyword="glycopyrronium") %>% View()  # lots, uprobably not all relevant
-search_bnf(keyword="cyclizine") %>% View()  # nto all relevant
-search_bnf(keyword="midazolam") %>% View()  # nto all relevant
-search_bnf(keyword="loprazolam") %>% View()  # nto all relevant
-
-## checking the ones that had matches...
-search_bnf(keyword="meptazinol") %>% View()  # nto all relevant
-
-search_bnf(keyword="alfentanil") %>% View()
-
-jic_medicines <-
-  bind_rows(
-    jic_medicines,
-    search_bnf(keyword="alfentanil")
+jic_medicines <- add_entries(
+  meds = "clomethiazole", 
+  bnf_to_add = search_bnf(keyword="clomethiazole")
   )
 ```
+
+### clonazepam
+
+Keeping only systemic clonazepam (BNF subparagraph “control of epilepsy”) and not parenteral (Clonazepam 1mg/1ml inj ampoules and diluent; subparagraph “drugs used in status epilepticus”)
+
+```{r}
+# search_bnf(keyword="clonazepam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "clonazepam", 
+  bnf_to_add = search_bnf(keyword="clonazepam") %>%
+    filter(bnf_paragraph_code == "040801")
+  )
+```
+
+### diazepam
+
+Keeping all formulations, oral solution/suspension, rectal solution tube, suppositories, injection ampoules
+
+```{r}
+# search_bnf(keyword="diazepam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "diazepam", 
+  bnf_to_add = search_bnf(keyword="diazepam")
+  )
+```
+
+### loprazolam
+
+```{r}
+# search_bnf(keyword="loprazolam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "loprazolam", 
+  bnf_to_add = search_bnf(keyword="loprazolam")
+  )
+```
+
+### lorazepam
+
+Keeping all formulations, tablets, oral solution/suspension, injection ampoules
+
+```{r}
+# search_bnf(keyword="lorazepam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "lorazepam", 
+  bnf_to_add = search_bnf(keyword="lorazepam")
+  )
+```
+
+### lormetazepam
+
+```{r}
+# search_bnf(keyword="lormetazepam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "lormetazepam", 
+  bnf_to_add = search_bnf(keyword="lormetazepam")
+  )
+```
+
+### midazolam
+
+keeping only BNF section 0408 “antiepileptic drugs” and not 1501 ”General anaesthesia”; of 0408, keeping both midazolam maleate (tablets, oral liquid, oral syringe) and midazolam hydrochloride (oral syringe, oral solution); (note: 1501 has injection ampoules and infusion vials, as well as oral solution)
+
+```{r}
+# search_bnf(keyword="midazolam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "midazolam", 
+  bnf_to_add = search_bnf(keyword="midazolam") %>%
+    filter(bnf_section_code == "0408")
+  )
+```
+
+### nitrazepam
+
+This picks up flunitrazepam (Rohypnol) also, but I didn't include that.
+
+```{r}
+# search_bnf(keyword="nitrazepam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "nitrazepam", 
+  bnf_to_add = search_bnf(keyword="nitrazepam") %>%
+    filter(bnf_chemical_substance_code == "0401010R0")  # keep only nitrazepam, not flunitrazepam
+  )
+```
+
+### temazepam
+
+```{r}
+# search_bnf(keyword="temazepam") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "temazepam", 
+  bnf_to_add = search_bnf(keyword="temazepam")
+  )
+```
+
+### trazodone hydrochloride
+
+```{r}
+# search_bnf(keyword="trazodone hydrochloride") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "trazodone hydrochloride", 
+  bnf_to_add = search_bnf(keyword="trazodone hydrochloride")
+  )
+```
+
+### zopiclone
+
+```{r}
+# search_bnf(keyword="zopiclone") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "zopiclone", 
+  bnf_to_add = search_bnf(keyword="zopiclone")
+  )
+```
+
+### zolpidem tartrate
+
+```{r}
+# search_bnf(keyword="zolpidem tartrate") %>% View()
+
+jic_medicines <- add_entries(
+  meds = "zolpidem tartrate", 
+  bnf_to_add = search_bnf(keyword="zolpidem tartrate")
+  )
+```
+
+## Check I've included all of them
+
+```{r}
+stopifnot(
+  nrow(
+    ## this should be an empty table - findings meds that were listed in jic but not in jic_medicines
+    jic %>% anti_join(jic_medicines, by = "meds")
+  )==0
+)
+```
+
+# Save resulting file
+
+See @sec-introduction for version listing.
+
+```{r}
+write_csv(jic_medicines, file = "./bnf/bnf_codes_just_in_case_v1.csv")  # v1 before SM checked
+```

### 2024-07-25 just_in_case_bnf_medications.qmd
finished incorporating comments from Sarah Mills into the JIC medications list

@@
 
-•       Strong opioids: Alfentanil, Buprenorphine, Diamorphine hydrochloride, Dihydrocodeine tartrate, Fentanyl, Hydromorphone hydrochloride, Meptazinol, Methadone hydrochloride, Morphine sulfate, Oxycodone hydrochloride, Tramadol hydrochloride
-•       Anti-muscarinics: Aclidinium bromide, Alverine citrate, Bupropion hydrochloride, Minims atropine sulfate, Atropine sulfate, Minims cyclopentolate hydrochloride, Cyclopentolate hydrochloride, Dicycloverine hydrochloride, Erdosteine, Fesoterodine fumarate, Flavoxate hydrochloride, Glycopyrrolate bromide, Hyoscine, Procyclidine hydrochloride, Propantheline bromide, Propiverine hydrochloride
-•       Anti-emetics: Betahistine hydrochloride, Cyclizine, Domperidone, Granisetron hydrochloride, Levomepromazine hydrochloride, Metoclopramide hydrochloride, Ondansetron hydrochloride, Prochlorperazine maleate
-•       Anxiolytics: Buspirone hydrochloride, Chlordiazepoxide hydrochloride, Clobazam, Clomethiazole, Clonazepam, Diazepam, Loprazolam, Lorazepam, Lormetazepam, Midazolam, Nitrazepam, Temazepam, Trazodone hydrochloride, Zopiclone, Zolpidem tartrate
+- Strong opioids: Alfentanil, Buprenorphine, Diamorphine hydrochloride, Dihydrocodeine tartrate, Fentanyl, Hydromorphone hydrochloride, Meptazinol, Methadone hydrochloride, Morphine sulfate, Oxycodone hydrochloride, Tramadol hydrochloride
+ - Anti-muscarinics: Aclidinium bromide, Alverine citrate, Bupropion hydrochloride, Minims atropine sulfate, Atropine sulfate, Minims cyclopentolate hydrochloride, Cyclopentolate hydrochloride, Dicycloverine hydrochloride, Erdosteine, Fesoterodine fumarate, Flavoxate hydrochloride, Glycopyrrolate bromide, Hyoscine, Procyclidine hydrochloride, Propantheline bromide, Propiverine hydrochloride
+ - Anti-emetics: Betahistine hydrochloride, Cyclizine, Domperidone, Granisetron hydrochloride, Levomepromazine hydrochloride, Metoclopramide hydrochloride, Ondansetron hydrochloride, Prochlorperazine maleate
+ - Anxiolytics: Buspirone hydrochloride, Chlordiazepoxide hydrochloride, Clobazam, Clomethiazole, Clonazepam, Diazepam, Loprazolam, Lorazepam, Lormetazepam, Midazolam, Nitrazepam, Temazepam, Trazodone hydrochloride, Zopiclone, Zolpidem tartrate
 
@@
 - v1 is my first attempt at compiling the codes, without a clinician checking over the codes
+- v2 includes amendments suggested by Sarah Mills, I've labelled these with 'SM' in the text
 
@@
 
-All formulations make sense to me. Query `Alfentanil 0.1% nasal spray`?
+- Alfentanil – I’ve included all formulations, including an “Alfentanil 0.1% nasal spray”
+- SM: agree
 
@@
 
-Keep only the analgesics section (0407), not the opioid dependence section (0410) - the latter uses buprenorphine hydrochloride and a buprenorphine hcl/naloxone combination.
+- •	Burprenorphine – Keep only the analgesics section (0407), not the opioid dependence section (0410) - the latter uses buprenorphine hydrochloride and a buprenorphine hcl/naloxone combination. 
+- SM: agree
 
@@
 
-Included both section cough suppressants and analgesics.
+ - I’ve included both the cough suppressants and analgesics sections
+ - SM: agree
 
@@
 
-Keep only the analgesics subsection (0407), not the general anaesthesia and local anaesthesia sections.
+- Keep only the analgesics subsection (0407), not the general anaesthesia and local anaesthesia sections.
+- SM: Agree (patches, which are in the analgesics subsection, should be included, and infusion bags/fentanyl citrate, which is in the anaesthesia sections, should be excluded)
 
@@
 
-Kept the cough suppressants section (0309) and analgesics (0407), not drug dependence (0410).
+- Kept the cough suppressants section (0309) and analgesics (0407), not drug dependence (0410).
+- SM: agree
 
@@
 
-Query gel, intrasite gel?
+- I’ve kept all formulations, including several gels/intrasite gels? 
+- SM: agree, gels often used for fungating tumours
 
@@
 
-Query combination: Oxycodone hydrochloride/naloxone hydrochloride?
+- I’ve kept all formulations, except for the combination “Oxycodone hydrochloride/naloxone hydrochloride” 
+- SM: agree
 
@@
 
-Keep only paragraph code 030102 for antimuscarinic bronchodilators, not compound bronchodilator preparations 030104 (Aclidinium bromide/formoterol)
+- Keep only paragraph code 030102 for antimuscarinic bronchodilators, not compound bronchodilator preparations 030104 (Aclidinium bromide/formoterol)
+- SM: agree
 
@@
 
-included combination Alverine citrate/simethicone, as well as Alverine citrate compound preparations (Sterculia 62% / Alverine 0.5% granules and Spasmonal Fibre granules)
+- SM: not familiar with combination Alverine citrate/simethicone, or Alverine citrate compound preparations (Sterculia 62% / Alverine 0.5% granules and Spasmonal Fibre granules)
+- therefore, removed those combinations, only kept preparations with only alverine citrate
 
@@
   meds = "alverine citrate", 
-  bnf_to_add = search_bnf(keyword="alverine citrate")
+  bnf_to_add = search_bnf(keyword="alverine citrate") %>% filter(bnf_chemical_substance == "Alverine citrate")  # only Alverine citrate
   )
@@
 
-Query whether only eye drops, or oral/parenteral is also relevant here? Also "Co-phenotrope (Diphenox hydrochloride/atropine sulfate)" for acute diarrhea?
+- Query whether only eye drops, or oral/parenteral is also relevant here? Also "Co-phenotrope (Diphenox hydrochloride/atropine sulfate)" for acute diarrhea?
+- SM: I’d restrict to eye drops (it’s an off licence use but recognosed for PEOLC use)
 
@@
     # minims already included by an above entry
-    filter(!str_detect(bnf_product, regex("minims", ignore_case=TRUE)))
+    filter(!str_detect(bnf_product, regex("minims", ignore_case=TRUE))) %>%
+    filter(bnf_chemical_substance_code == "1105000B0")  # eye drops only
   )
@@
 ### dicycloverine hydrochloride
+
+- I’ve included all formulations, including “Dicycloverine hydrochloride compound preparations (Kolanticon gel)”
+- SM:agree
 
@@
 
+- glycopyrrolate bromide (glycopyrronium bromide in BNF) – this spans multiple chapters – gastro-intestinal (oral preparations), respiratory (inhalation), skin (topical, powder), and general anaesthesia (parenteral) – which of these are relevant?
+(the anaesthesia section has a combination of Neostigmine metilsulfite/glycopyrronium bromide also)
+
+- SM: in my experience only the oral and sub cutaneous form is used in palliative care.
+Definitely wouldn’t include the inhalation and topical/powder versions as they are used for non-palliative considitons.
+- SM: don't include the Neostigmine metilsulfite/glycopyrronium bromide combination
+
 ```{r}
@@
   meds = "glycopyrrolate bromide", 
-  bnf_to_add = search_bnf(keyword="glycopyrr.* bromide")
+  bnf_to_add = search_bnf(keyword="glycopyrr.* bromide") %>%
+    filter(bnf_chemical_substance_code %in% c("0102000L0", "1501030G0"))  # oral & parenteral, respectively
   )
@@
 
-Query:this occurs in 2 forms, butylbromide (antispasmodic) and hydrobromide (nausea and vertigo) – are both relevant? And in all formulations (oral, injectable, transdermal patch)
+- Query:this occurs in 2 forms, butylbromide (antispasmodic) and hydrobromide (nausea and vertigo) – are both relevant? And in all formulations (oral, injectable, transdermal patch)
+- SM: all formulations and forms are used for PEOLC
 
@@
 
-Keeping cyclizine hydrochloride and lactate forms, but not the combination "Morphine tartrate and cyclizine tartrate" in the Analgesics section (0407)
+- Keeping cyclizine hydrochloride and lactate forms, but not the combination "Morphine tartrate and cyclizine tartrate" in the Analgesics section (0407)
+- SM: I’d include the combination version
 
@@
   meds = "cyclizine", 
-  bnf_to_add = search_bnf(keyword="cyclizine") %>%
-    filter(bnf_section_code != "0407")
+  bnf_to_add = search_bnf(keyword="cyclizine")  # don't filter out the combination with morphine tartrate
   )
@@
 
-Keeping only systemic clonazepam (BNF subparagraph “control of epilepsy”) and not parenteral (Clonazepam 1mg/1ml inj ampoules and diluent; subparagraph “drugs used in status epilepticus”)
+- Keeping only systemic clonazepam (BNF subparagraph “control of epilepsy”) and not parenteral (Clonazepam 1mg/1ml inj ampoules and diluent; subparagraph “drugs used in status epilepticus”)
+- SM: sometimes parenteral can be used in terminal agitation or in brain tumours with seizures, so would include
 
@@
   meds = "clonazepam", 
-  bnf_to_add = search_bnf(keyword="clonazepam") %>%
-    filter(bnf_paragraph_code == "040801")
+  bnf_to_add = search_bnf(keyword="clonazepam")  # don't remove parenteral
   )
@@
 
-Keeping all formulations, oral solution/suspension, rectal solution tube, suppositories, injection ampoules
+- Keeping all formulations, oral solution/suspension, rectal solution tube, suppositories, injection ampoules
+- SM: agree
 
@@
 
-Keeping all formulations, tablets, oral solution/suspension, injection ampoules
+- Keeping all formulations, tablets, oral solution/suspension, injection ampoules
+- SM: agree
 
@@
 
-keeping only BNF section 0408 “antiepileptic drugs” and not 1501 ”General anaesthesia”; of 0408, keeping both midazolam maleate (tablets, oral liquid, oral syringe) and midazolam hydrochloride (oral syringe, oral solution); (note: 1501 has injection ampoules and infusion vials, as well as oral solution)
+- keeping only BNF section 0408 “antiepileptic drugs” and not 1501 ”General anaesthesia”; of 0408, keeping both midazolam maleate (tablets, oral liquid, oral syringe) and midazolam hydrochloride (oral syringe, oral solution); (note: 1501 has injection ampoules and infusion vials, as well as oral solution)
+- SM: agree
 
@@
 
-This picks up flunitrazepam (Rohypnol) also, but I didn't include that.
+- This picks up flunitrazepam (Rohypnol) also, but I didn't include that.
+- SM: agree
 
@@
 ```{r}
-write_csv(jic_medicines, file = "./bnf/bnf_codes_just_in_case_v1.csv")  # v1 before SM checked
+# write_csv(jic_medicines, file = "./bnf/bnf_codes_just_in_case_v1.csv")  # v1 before SM checked
+write_csv(jic_medicines, file = "./bnf/bnf_codes_just_in_case_v2.csv")  # v2 after SM checked
 ```

### 2025-10-27 place_of_death_report_2025.qmd
quick pared-down version of the full script for producing graphs of the latest figures for deaths at home in 2025

---
title: "Place of death trend in Scotland, 2025"
author: "Jan Savinc"
format: html
editor_options: 
  chunk_output_type: console
---

# Introduction

As of 2025-03-03, the NRS Covid-specific updates on weekly death registrations are no longer published. The new publication at https://www.nrscotland.gov.uk/publications/deaths-registered-weekly-in-scotland/ is updated weekly, and Table 7 contains the breakdown by place of death ("location").

# Preliminaries

## Load packages

```{r}
library(tidyverse)
library(readxl)  # for reading excel files
library(curl)  # for downloading files
library(janitor)  # for cleaning datasets quickly
library(rvest)  # for scraping NRS website
library(ISOweek)  # for dealing with ISO 8601 weeks
```

## Load functions

```{r}
source("./functions.R")
```

## Find latest file

```{r}
url_nrs_weekly_deaths <- "https://www.nrscotland.gov.uk/publications/deaths-registered-weekly-in-scotland/"

all_nrs_links <-
  read_html(url_nrs_weekly_deaths) %>%
  html_nodes("a") %>%
  html_attr("href") %>%  # find all links on the page
  tibble(links = .)

nrs_filename_pattern <- "weekly-deaths-.*\\.xlsx"

weekly_deaths_url_relative <-
  all_nrs_links %>%
  filter(str_detect(links, pattern = nrs_filename_pattern)) %>%
  filter(!str_detect(links, pattern = "monthly")) %>%  # remove monthly data
  filter(!str_detect(links, pattern = "excess")) %>%  # remove excess deaths data
  filter(!str_detect(links, pattern = "influenz")) %>%  # remove influenza data
  pull(links)

stopifnot(length(weekly_deaths_url_relative)==1)  # stop if more than one URL found

## get absolute URL
weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative, base = url_nrs_weekly_deaths)
```

## Download file

```{r}
date_string <- strftime(today(), format = "%F")
path_today <- file.path("./downloaded_data", date_string)
path_to_downloaded_file <- file.path(path_today, basename(weekly_deaths_url_absolute))

if (!dir.exists(path_today)) dir.create(path_today)

# download.file(url = weekly_deaths_url_absolute, destfile = path_to_downloaded_file, quiet = TRUE)
todays_file <- curl_fetch_disk(weekly_deaths_url_absolute, path = path_to_downloaded_file)
```

## Read file

```{r}
deaths_by_pod_raw <- 
  read_excel(path = path_to_downloaded_file, sheet = "7", skip = 4) %>%
  clean_names()
```

## Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
latest_data_modified_date <- strftime(todays_file$modified, format="%F")
latest_year <- year(latest_data_modified_date)

dir_outputs <- file.path("./outputs", latest_year, latest_data_modified_date)
if (!dir.exists(dir_outputs)) dir.create(dir_outputs, recursive = TRUE)
```

# Wrangle data

## Load previous data

```{r}
deaths_2015_to_2019 <- 
  read_csv("./historical_data/2020-04-30_weekly-deaths-by-location-2015-2019.csv", skip = 2) %>%
  rename(year = 1) %>%  # rename first column
  mutate(
    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
  ) %>%
  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
  drop_na(`1`) %>%
  pivot_longer(cols = all_of(as.character(1:53)), values_to = "n_deaths", names_to = "week_number") %>%
  mutate(week_number=as.integer(week_number)) %>%
  recode_place_of_death()

## construct column names for the 2020 data
headers_in_nrs_data_2020 <-
  c(
    "place_of_death",
    "cause_of_death",
    paste0("week_", 1:53),
    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
    "remove_1",
    "remove_2"
  )

deaths_2020 <- read_excel(
  path = "./historical_data/2022-01-06_covid-deaths-21-data-week-52.xlsx",
  sheet = "Table 3 (2020)",
  col_names = headers_in_nrs_data_2020,
  range = cell_rows(6:125)
) %>%  # the data are inrows 6:125
  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "n_deaths", values_drop_na = TRUE) %>%  # convert to long data format
  mutate(
    week_number = as.integer(parse_number(week_number)),  # extract week number
    place_of_death =  # find best match in the reference names used in other datasets
      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)],  # the reference is in the variable order_of_place_of_death_levels
    year = 2020,
    dataset = str_remove_all(dataset , "Registered deaths\\: "),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
    ) %>%
  filter(dataset!="Difference") %>%
  pivot_wider(names_from = dataset, values_from = n_deaths) %>%
  rename(deaths_mean = `five year average`, n_deaths = `2020`) %>%
  recode_place_of_death()

cell_ranges_for_causes_of_death_data_2021 <- list(
  "All" = "A6:V58",
  "Care home" = "A113:V165",
  "Home & other non-institution" = "A220:V272",
  "Hospital" = "A327:V379",
  "Other" = "A434:V486"
)

deaths_2021 <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data_2021,
    .f = ~readxl::read_excel("./historical_data/2022-06-29_covid-deaths-22-data-week-25.xlsx", sheet = "8", range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "n_deaths") %>%
  # filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "n_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = n_deaths) %>%
  filter(!is.na(n_deaths)) %>%  # remove blank entries - these are for the current year but for dates in the future
  recode_place_of_death() %>%
  mutate(cause_of_death = if_else(cause_of_death=="All causes", "All", cause_of_death))

cell_ranges_for_causes_of_death_data_2022 <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

deaths_2022 <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data_2022,
    .f = ~readxl::read_excel("./historical_data/2023-08-02_covid-deaths-22-data-final.xlsx", sheet = "8", range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "n_deaths") %>%
  # filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "n_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = n_deaths) %>%
  filter(!is.na(n_deaths)) %>%  # remove blank entries - these are for the current year but for dates in the future
  recode_place_of_death() %>%
  mutate(cause_of_death = if_else(cause_of_death=="All causes", "All", cause_of_death))
```

## Convert new data to tidy (long) format

```{r}
deaths_by_pod_clean <-
  deaths_by_pod_raw %>%
  pivot_longer(cols = all_locations_all_causes:other_institution_covid_19_mentioned, names_to = c("place_of_death", "type_of_cause"), names_pattern = "(all_locations|care_home|hospital|home_non_institution|other_institution)\\_(all_causes|covid_19_mentioned)", values_to = "n_deaths") %>%
  recode_place_of_death() %>%
  mutate(
    year = as.integer(registration_year),
    cause_of_death = case_when(
      type_of_cause == "all_causes" ~ "All",
      type_of_cause == "covid_19_mentioned" ~ "Covid-19 mentioned",
      TRUE ~ NA_character_
    ),
    date_w_c = as_date(week_beginning)
  ) %>% 
  select(-c(type_of_cause, week_beginning, registration_year))
```

## Merge new data and old

```{r}
deaths_current <-
  bind_rows(
    deaths_2020 %>%
      filter(cause_of_death == "All"),
    deaths_2021 %>%
      filter(cause_of_death == "All"),
    deaths_2022 %>%
      filter(cause_of_death == "All"),
    deaths_by_pod_clean %>%
      filter(cause_of_death == "All")
  )

deaths_range_2015_2019 <-
  deaths_2015_to_2019 %>%
  filter(year != "average") %>%
  filter(!(year!=2015 & week_number==53)) %>%  # remove entries for w53 in years where there wasn't one, #ISOweekproblems
  group_by(place_of_death, week_number) %>%
  summarise(
    cause_of_death = "All",
    year_range = "2015-2019",
    mean_deaths = mean(n_deaths),
    min_deaths = min(n_deaths),
    max_deaths = max(n_deaths),
    .groups = "drop"
    )
```

## Annual totals so far

```{r}
deaths_annual <-
  bind_rows(
    deaths_2015_to_2019 %>% filter(year!="average" & place_of_death!="average") %>% mutate(year=as.integer(year), cause_of_death="All"),
    deaths_current %>% filter(cause_of_death=="All")
  ) %>%
  count(year, place_of_death, cause_of_death, wt = n_deaths, name = "n_deaths") %>%
  group_by(year, place_of_death != "All") %>%
  mutate(  # add percentage
    prop_deaths = n_deaths / sum(n_deaths)
  ) %>%
  ungroup %>%
  mutate(year = if_else(  # rename current year so it's clear when the data end
    year == max(deaths_current$year),
    paste0(year, " up to W", deaths_current %>% filter(year==max(deaths_current$year)) %>% pull(week_number) %>% max),
    as.character(year)
  ))
```

# Analyses

## Home deaths plot

```{r}
(fig_weekly_home_deaths_2015_2019_range <-
  deaths_current %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(deaths_range_2015_2019, join_by(place_of_death, cause_of_death, week_number)) %>%
  ggplot(aes(x = as_date(date_w_c))) +
  add_vertical_lines_for_start_of_year(years = 2020:2025) +
  geom_text(
    data = tibble(year = 2020:2025) %>%
      mutate(date_w_c = map_vec(.x = year, ~compute_start_date_from_week_number(week_number = 1, year_number = .x))),
    aes(x = date_w_c, y = 30, label = paste0("'",str_sub(year,3,4))), hjust = -0.5, angle = -30
  ) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks") +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 6), limits = c(0,NA)) +
  geom_line(aes(y = n_deaths)) +
  geom_line(aes(y = mean_deaths), linetype = "dotted", colour = "grey20") +
  geom_ribbon(aes(ymin = min_deaths, ymax = max_deaths), fill = "#4477aa", alpha = 0.5) +
  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # facet_grid(~place_of_death) +
  # scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
  #                     labels = c("All deaths", 
  #                                "Non Covid-19 deaths"),
  #                     guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
  #                     ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)",
       # subtitle = date_range_text,
       # caption = paste0(c(death_registrations_text,historic_period_2022_text,source_nrs_text), collapse="\n")
       ))

fig_weekly_home_deaths_2015_2019_range %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_weekly_home_deaths_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png", ".svg"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")
```

## Annual deaths

```{r}
# uncomment to check that the assertion about Other being ~0.5% is true
# deaths_annual %>%
#   filter(place_of_death=="Other")

(fig_proportion_deaths_by_place_annual <- 
  deaths_annual %>%
  filter(place_of_death!="All") %>%
  filter(!str_starts(year, "2025")) %>%
  filter(place_of_death != "Other") %>%
  ggplot(aes(x=year, y=prop_deaths, colour=place_of_death, shape=place_of_death)) +
  geom_point(size = rel(3)) +
  geom_line(aes(group = place_of_death)) +
  geom_text(
    data = . %>% filter(!year %in% as.character(2015:2018)) %>%
      mutate(y = if_else(place_of_death=="Care home", prop_deaths - 0.03, prop_deaths + 0.03)),
    aes(
      x = year,
      y = y,
      label = scales::label_percent(accuracy = 0.1)(prop_deaths)
    ), 
    show.legend = FALSE
  ) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "top") +
  scale_colour_manual(values = viridisLite::viridis(n=4, option = "D")[1:3]) +
  # scale_colour_manual(values = RColorBrewer::brewer.pal(3, "Dark2")) +
  # scale_colour_manual(values = scales::pal_dichromat(name = "Categorical.12")(3)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, NA)) +
  labs(x = NULL, y = "Deaths (%)", caption = '"Other institutions" represent ~0.5% of annual deaths and are not shown.'))

fig_proportion_deaths_by_place_annual %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_deaths_by_place_annual", latest_data_modified_date), collapse = "_")), extensions = c(".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
```
